### 2016-09-09 write-ups/PLDI/paper.tex
first commit

%
% LaTeX template for prepartion of submissions to PLDI'16
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'16 web site.
% 
\documentclass[nocopyrightspace]{sigplanconf-pldi16}
% \documentclass[pldi-cameraready]{sigplanconf-pldi16}

%
% the following standard packages may be helpful, but are not required
%
\usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink

\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{calc}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{color}
\usepackage{xspace}
\usepackage{array}
\usepackage{multirow}
\usepackage{varwidth}
\usepackage{tabularx}  % for 'tabularx' environment and 'X' column type
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{xcolor,colortbl}
\usepackage{graphicx}

\usepackage{booktabs}
\usepackage{subfig}

\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

%\newcommand{\todo}[1]{}
%\renewcommand{\todo}[1]{{\color{red} TODO: {#1}}}
\newcolumntype{C}[1]{>{\centering}m{#1}}
%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

%\newcommand{\ADComment}[1]{\textcolor{green}{AD: #1}}
%\newcommand{\TSComment}[1]{\textcolor{blue}{TS: #1}}
%\newcommand{\COULDCUT}[1]{\textcolor{blue}{#1}}

\newcommand{\myfig}{Fig.~}
\newcommand{\myfiglong}{Figure~}
\newcommand{\myfigs}{Figs.~}
\newcommand{\mytab}{Tab.~}
\newcommand{\mytablong}{Table~}
\newcommand{\mysec}{Sec.~}
\newcommand{\myalg}{Alg.~}
\newcommand{\nvidia}{Nvidia\xspace}
\newcommand{\totalcombinations}{70\xspace}
\newcommand{\positivecombinations}{55\xspace}
\newcommand{\cbedot}{\emph{cbe-dot}\xspace}
\newcommand{\cbeht}{\emph{cbe-ht}\xspace}
\newcommand{\ctoctree}{\emph{ct-octree}\xspace}
\newcommand{\tpotm}{\emph{tpo-tm}\xspace}
\newcommand{\sdkred}{\emph{sdk-red}\xspace}
\newcommand{\sdkrednf}{\emph{sdk-red-nf}\xspace}
\newcommand{\lsbh}{\emph{ls-bh}\xspace}
\newcommand{\lsbhnf}{\emph{ls-bh-nf}\xspace}
\newcommand{\cubscan}{\emph{cub-scan}\xspace}
\newcommand{\cubscannf}{\emph{cub-scan-nf}\xspace}
\newcommand{\heuristiccombinations}{eight\xspace}
\newcommand{\effectivethreshold}{5\%\xspace}
\definecolor{Gray}{gray}{0.85}

% lstlisting macros
\lstset{
  language=C, % choose the language of the code
  showspaces=false, % show spaces adding particular underscores
  showstringspaces=false, % underline spaces within strings
  showtabs=false, % show tabs within strings adding particular underscores
  tabsize=8, % sets default tabsize to 2 spaces
  captionpos=b, % sets the caption-position to bottom
  mathescape=true, % activates special behaviour of the dollar sign
  basicstyle=\scriptsize\tt,
  columns=fullfelxible,
  %commentstyle=\rmfamily\itshape,
  morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then},
  escapeinside={(*@}{@*)}
}

\newcommand{\code}[1]{\lstset{basicstyle=\tt}\lstinline!#1!\lstset{basicstyle=\scriptsize\tt}}

\newcommand{\MP}{\textsf{MP}}
\newcommand{\LB}{\textsf{LB}}
\newcommand{\SB}{\textsf{SB}}
\newcommand{\noisethreshold}{3\xspace}
\newcommand{\TotalLitmusTests}{196.6\xspace}
\newcommand{\loadop}{\textsf{ld}\xspace}
\newcommand{\storeop}{\textsf{st}\xspace}

\newcommand{\Td}[2]{{#1}_{#2}}
\newcommand{\Tdl}[3]{\langle \Td{#1}{#2}, #3 \rangle}
\newcommand{\Tdlsigma}[4]{\Tdl{#1}{#2}{#4@#3}}

\newcommand{\randstrsolo}{\emph{rand-str}\xspace}
\newcommand{\sysstrsolo}{\emph{sys-str}\xspace}
\newcommand{\nostrsolo}{\emph{no-str}\xspace}
\newcommand{\cachestrsolo}{\emph{cache-str}\xspace}

\newcommand{\randstr}[1]{\emph{rand-str}#1}
\newcommand{\sysstr}[1]{\emph{sys-str}#1}
\newcommand{\nostr}[1]{\emph{no-str}#1}
\newcommand{\cachestr}[1]{\emph{cache-str}#1}

\newcommand{\NumGPUs}{seven\xspace}
\newcommand{\NumApplications}{ten\xspace}
\newcommand{\NumApplicationsUnique}{seven\xspace}

\begin{document}

\title{Responsive GPU Workgroup Management in the Presence of
  Persistent Kernels}

%
% any author declaration will be ignored  when using 'pldi' option (for double blind review)
%

\authorinfo{}
{\makebox{} \\
}
{}

%\authorinfo{Tyler Sorensen}
%{\makebox{Imperial College London, UK} \\
%}
%{t.sorensen15@imperial.ac.uk}

%\authorinfo{Alastair F. Donaldson}
%{\makebox{Imperial College London, UK} \\
%}
%{alastair.donaldson@imperial.ac.uk}

\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}\label{sec:intro}

GPUs are highly parallel co-processors that are now found on many
devices, from the most powerful supercomputers to mobile phones. While
originally designed to accelerate graphics applications, GPUs are now
used in a wide range of applications, including molecular simulations
and weather forecasting.

New application domains for GPUs are enabled by general purpose
programming languages for GPUs. The current most popular such
languages being CUDA and OpenCL. These languages allow developers to
write C-like programs which may be dispatched and executed on a GPU
device. GPUs have a hierarchical programming model; groups of threads
are called \emph{workgroups} and intra-workgroup threads interact
differently than inter-workgroup threads.

GPUs are massively parallel and employ many threads to execute a
program. Traditionally, applications accelerated by GPUs are
data-parallel, requiring little interaction between threads. A small
set of bulk synchronisation primitives are provided and any
inter-thread communication must be performed around these primitives.

While this traditional way to program GPUs has proved useful, it also
has limitations for some applications, in particular applications that
expose a dynamic amount of parallelism throughout execution. Such
applications are said to have irregular parallelism (we call such
applications \emph{IP applications}). Efficient parallel computation
of such applications requires finer grained synchronisation than what
GPU languages currently provide as primitives.

In an effort to achieve high performance for IP applications, new GPU
programming paradigms have been developed that include custom GPU
synchronisation constructs. Here we focus on two:

\begin{enumerate}
\item {\bf Inter-workgroup barrier}: Workgroups are able to
  efficiently synchronise frequently using a custom inter-workgroup
  barrier that is not provided as GPU primitive.

\item {\bf Work stealing}: Computation is divided into tasks that are
  stored in workgroup-local queues. If a workgroup runs out of local
  tasks to compute, it is able to steal a task from a different
  workgroup.
\end{enumerate}

Both paradigms fit into what are called \emph{persistent kernels}, the
defining characteristic being that persistent kernels use a relatively
small number of long-running threads to perform computation. In
contrast, traditional GPU programs employ many short running threads
perform computation. In this work, the persistent kernels we consider
use either work stealing or inter-workgroup barrier paradigms.

Persistent kernels rely on strong forward progress properties (FPP)
between workgroups; that is, workgroups must make forward progress to
the other workgroups executing. In the inter-workgroup barrier, all
workgroups spin at a barrier instruction waiting for all workgroups to
arrive. If one (or more) workgroups stop making forward progress, the
application will deadlock as the live workgroups will wait
indefinitely for the halted workgroups. Less obviously, in the work
stealing paradigm, the workgroups terminate only when \emph{all} tasks
have been completed. If a workgroup takes a task, but stops making
forward progress before the task is completed, the application will
deadlock as the live workgroups will spin waiting for the halted
workgroup to finish its task.

The current GPU programming models generally do not provide strong
enough inter-workgroup FPP to allow persistent kernels. This is
because GPUs do not provide workgroup preemption. That is, GPU
programming models allow GPU programs to be launched with many more
workgroups than the available hardware resources are able run
concurrently. The GPU runtime will schedule as many workgroups as the
available resources can provide, and the remaining workgroups will
wait until hardware resources become available, e.g.  when one of the
scheduled workgroups finishes execution. There is no mechanism to swap
workgroups from 'executing' to 'waiting' and vice versa. Workgroups
that have been scheduled and have not finished execution are called
\emph{occupant}.

Previous work experimenting with persistent kernels has required
\emph{a priori} knowledge of the number of workgroups that the GPU
hardware is able to execute the program with concurrently. This
number, called the \emph{maximal launch} or \emph{occupancy bound}, is
a function of both a GPU chip and a compiled program. Because occupant
workgroups are never preempted, if the GPU program is launched with a
number of workgroups which doesn't exceed the occupancy bound, then
all workgroups will be occupant and the strong inter-workgroup FPP
required by persistent kernels are provided. Using the occupancy
bound, previous work has successfully implemented persistent kernels
and shown them to be performant on current GPUs. Recent work has even
shown that the occupancy bound can be accurately discovered
programitically, removing the need for any \emph{a priori} knowledge
about the GPU.

Despite the promising performance and ease of programming persistent
kernels with occupant workgroups, GPU language specification
committees are reluctant to commit to the inter-workgroup FPP
currently provided by occupant workgroups. That is, even if current
GPUs provide strong FPP between workgroups, future GPUs may not. Thus
the performance gains enabled by persistent kernels may not be
possible under future GPUs.

We believe that two fundamental issues stopping committees from
committing to strong inter-workgroup FPP are:

\begin{itemize}
\item {\bf Responsive GPU multi-tasking} A high priority GPU task
  (e.g. graphics) may need to quickly gain access to the GPU. To do
  this, some of the resources will have to be taken from the
  persistent kernel in order to run the high priority task. This will
  reduce the number of occupant workgroups executing the persistent
  kernel.

\item {\bf Energy throttling} Some GPUs, especially in mobile devices,
  may wish to monitor the energy consumed (i.e. heat output) of the
  GPU chip.  A GPU which is consuming too much energy may be throttled
  by limiting the amount of resources (and thus the number of occupant
  workgroups) to conserve battery life.
\end{itemize}

Both issues are a case of the GPU runtime needing to adjust the
resources used by the persistent kernel during execution. This causes
occupant workgroups to potentially be preempted, which may lead to
deadlock in persistent kernels. A less severe, but still undesirable
issue is if the runtime determines more GPU resources are available
during a persistent kernel execution. In this case, the persistent
kernel will be under-utilising GPU resources.

In this work, we propose a solution that unites persistent kernels
with the need for GPU runtimes to dynamically (and quickly) adjust the
resources available to the persistent kernel. We propose a new GPU
programming paradigm called \emph{dynamic persistent kernels}, or
DPK. DPKs provide strong inter-workgroup FPP while still allowing the
GPU runtime to adjust the number of workgroups executing the DPK in a
timely manner, enabling responsive GPU multi-tasking or energy
throttling.

The key insight behind DPK, is that many persistent kernels contain
points in their execution where workgroups can easily be either
terminated or started. We propose several new instructions to mark
these entry and exit points. These instructions serve as hooks for the
GPU runtime to adjust the amount of resources available to the DPK,
and thus the number of occupant workgroups the DPK can rely on to have
strong FPPs.

For example, many applications using the inter-workgroup barrier
paradigm do not carry workgroup-local information across calls to the
inter-workgroup barrier. Thus the number of workgroups can be adjusted
at every call to the inter-workgroup barrier. Many applications using
the work-stealing paradigm do not carry state between task
computation; thus workgroups can be created or terminated immediately
after ever task computation.

While we envision support for DPKs to ultimately be implemented in
proprietary GPU drivers, we are able to provide a prototype
implementation in software using the strong inter-workgroup FPP
available on current GPUs. This allows us to test the feasibility of
our approach and as well get upper bounds for the performance cost and
the responsiveness of multi-tasking.

\end{document}

### 2016-09-15 write-ups/PLDI/paper.tex

@@
 
-%\newcommand{\ADComment}[1]{\textcolor{green}{AD: #1}}
+\newcommand{\ADComment}[1]{\textcolor{green}{AD: #1}}
 %\newcommand{\TSComment}[1]{\textcolor{blue}{TS: #1}}
@@
 
-GPUs are highly parallel co-processors that are now found on many
-devices, from the most powerful supercomputers to mobile phones. While
+Graphics processing units (GPUs) are highly parallel co-processors that are now incorporated into many
+devices, ranging from the world's most powerful supercomputers to battery-powered devices such as tablets and mobile phones. While
 originally designed to accelerate graphics applications, GPUs are now
 used in a wide range of applications, including molecular simulations
-and weather forecasting.
+and weather forecasting. \ADComment{Expand the list of application areas a bit more; mention their attractiveness due to energy-efficiency.}
 
 New application domains for GPUs are enabled by general purpose
-programming languages for GPUs. The current most popular such
-languages being CUDA and OpenCL. These languages allow developers to
-write C-like programs which may be dispatched and executed on a GPU
-device. GPUs have a hierarchical programming model; groups of threads
-are called \emph{workgroups} and intra-workgroup threads interact
-differently than inter-workgroup threads.
+programming languages with associate compilation tools that target GPU architectures.
+The current most popular GPU programming languages are CUDA and OpenCL. These languages allow developers to
+write C-like programs that can be dispatched and executed on a GPU
+device.
 
-GPUs are massively parallel and employ many threads to execute a
-program. Traditionally, applications accelerated by GPUs are
+GPUs are massively parallel, employing many threads to execute a
+program.  These threads are organised hierarchically: groups of threads
+are called \emph{workgroups}, and threads interact differently depending on whether they reside in the same workgroup or not.  Traditionally, applications accelerated by GPUs are
 data-parallel, requiring little interaction between threads. A small
 set of bulk synchronisation primitives are provided and any
-inter-thread communication must be performed around these primitives.
+inter-thread communication must be performed around these primitives.  \ADComment{Do we want to make the point that this is traditionally limited to the intra-workgroup case?}
 
@@
 applications \emph{IP applications}). Efficient parallel computation
-of such applications requires finer grained synchronisation than what
-GPU languages currently provide as primitives.
+of IP applications requires finer-grained synchronisation than that
+which is provided by GPU language primitives.
 
@@
 programming paradigms have been developed that include custom GPU
-synchronisation constructs. Here we focus on two:
+synchronisation constructs. Here we focus on two \ADComment{The way this is worded it is unclear whether you are describing two programming paradigms, or two synchronisation constructs}:
 
@@
 contrast, traditional GPU programs employ many short running threads
-perform computation. In this work, the persistent kernels we consider
+to perform computation. In this work, the persistent kernels we consider
 use either work stealing or inter-workgroup barrier paradigms.
 
-Persistent kernels rely on strong forward progress properties (FPP)
-between workgroups; that is, workgroups must make forward progress to
-the other workgroups executing. In the inter-workgroup barrier, all
+Persistent kernels rely on strong forward progress properties (FPPs)
+between workgroups; that is, workgroups must make forward progress with respect to
+the other workgroups that are executing. In the case of the inter-workgroup barrier, all
 workgroups spin at a barrier instruction waiting for all workgroups to
@@
 application will deadlock as the live workgroups will wait
-indefinitely for the halted workgroups. Less obviously, in the work
+indefinitely for the halted workgroups \ADComment{Can we re-word the preceding to make it clearer that the workgroups have been stalled by a scheduler: they have not actually terminated their execution by reaching the end of the kernel}.
+Less obviously, in the work
 stealing paradigm, the workgroups terminate only when \emph{all} tasks
@@
 The current GPU programming models generally do not provide strong
-enough inter-workgroup FPP to allow persistent kernels. This is
+enough inter-workgroup FPP to allow persistent kernels. \ADComment{This might be a good place to cite the specs.}  This is
 because GPUs do not provide workgroup preemption. That is, GPU
@@
 \emph{occupant}.
+\ADComment{Note: it's ironic that lack of preemption is what leads to the problem of workgroups being scheduled in waves, and yet also what means we might not be able to guarantee the properties of the occupancy-bound execution model in the future.}
 
@@
 the performance gains enabled by persistent kernels may not be
-possible under future GPUs.
+safely achievable for future GPUs.
 
@@
   may wish to monitor the energy consumed (i.e. heat output) of the
-  GPU chip.  A GPU which is consuming too much energy may be throttled
+  GPU chip.  A GPU that is consuming too much energy may be throttled
   by limiting the amount of resources (and thus the number of occupant
@@
 occupant workgroups to potentially be preempted, which may lead to
-deadlock in persistent kernels. A less severe, but still undesirable
+deadlock in persistent kernels.
+\ADComment{It might be good to illustrate the previous points with an example; e.g.\ we could envisage a kernel that is being executed concurrently by 16 workgroups.
+Some (4 say) have reached an inter workgroup barrier and are awaiting the others.
+But then a graphics tasks causes the 12 workgroups that have not yet reached the barrier to be preempted, and in anticipation of further graphics processing requirements the scheduler decides not to reduce the workgroup allocation for the kernel to 4 indefinitely.
+The 4 workgroups waiting at the barrier cannot now make progress.
+Energy throttling example is analogous, except that the count is reduced from 16 to 12 due to low battery.}
+A less severe, but still undesirable
 issue is if the runtime determines more GPU resources are available
@@
 kernel will be under-utilising GPU resources.
+\ADComment{I'm not sure I get this one.
+Do you mean here that at the point of kernel launch there may be e.g.
+only 4 workgroup slots available, say due to low battery, but then later on (e.g., due to a phone being charged) there may be scope for more workgroups to run in parallel?}
+
+\ADComment{We should mention that GPU watchdogs are the rather drastic solution to this problem, to keep the overall system responsive, but that they do not help from the point of view of IP applications themselves.}
 
@@
 strong FPPs.
+\ADComment{We should make it clear early on that some care on the part of the programmer is required to make this work, and that if the programmer does not insert sufficient yield points, the GPU watchdog may still have to kill a kernel.}
 

### 2016-10-12 src/ckill_only/host/main.cpp
added a first example that uses both ckill and cfork

@@ -9,6 +9,7 @@
 #define CL_INT_TYPE cl_int
 #define ATOMIC_CL_INT_TYPE cl_int
 #define MY_CL_GLOBAL 
+#define CL_UCHAR_TYPE cl_uchar
 
 #include "profile.h"
 #include "base/commandlineflags.h"
@@ -71,6 +72,7 @@ void list_devices() {
 int get_kernels(CL_Execution &exec) {
 	int ret = CL_SUCCESS;
 	exec.exec_kernels["mega_kernel"] = cl::Kernel(exec.exec_program, "mega_kernel", &ret);
+	cl_float4 x;
 	return ret;
 }
 
@@ -106,7 +108,7 @@ int main(int argc, char *argv[]) {
 	exec.exec_queue = queue;
 
 	// Should be built into the cmake file. Haven't thought of how to do this yet.
-	err = exec.compile_kernel(kernel_file, "C:/Users/Tyler/Documents/GitHub/GPU_workgroup_managment/src/uvm_tests/test1/include/rt_device/");
+	err = exec.compile_kernel(kernel_file, "C:/Users/Tyler/Documents/GitHub/GPU_workgroup_managment/src/uvm_tests/test1/include/rt_device/", "C:/Users/Tyler/Documents/GitHub/GPU_workgroup_managment/src/first_resize/common/");
 
 	check_ocl(err);
 

### 2016-10-12 src/first_resize/host/main.cpp

@@ -8,21 +8,20 @@
 // Should be in a directory somewhere probably. Or defined in CMake.
 #define CL_INT_TYPE cl_int
 #define ATOMIC_CL_INT_TYPE cl_int
-#define MY_CL_GLOBAL 
 #define CL_UCHAR_TYPE cl_uchar
+#define MY_CL_GLOBAL
 
-#include "profile.h"
 #include "base/commandlineflags.h"
 #include "opencl/opencl.h"
 #include "cl_execution.h"
 #include "discovery.h"
 #include "cl_communicator.h"
+#include "../common/restoration_ctx.h"
 #include "cl_scheduler.h"
 #include "kernel_ctx.h"
 #include "iw_barrier.h"
 
-DEFINE_string(input, "", "Input path");
-DEFINE_string(output, "", "Output path");
+
 DEFINE_int32(platform_id, 0, "OpenCL platform ID to use");
 DEFINE_int32(device_id, 0, "OpenCL device ID to use");
 DEFINE_bool(list, false, "List OpenCL platforms and devices");
@@ -71,7 +70,7 @@ void list_devices() {
 	}
 }
 
-int get_kernels(CL_Execution &exec) {
+int get_app_kernels(CL_Execution &exec) {
 	int ret = CL_SUCCESS;
 	exec.exec_kernels["mega_kernel"] = cl::Kernel(exec.exec_program, "mega_kernel", &ret);
 	return ret;
@@ -80,11 +79,6 @@ int get_kernels(CL_Execution &exec) {
 int main(int argc, char *argv[]) {
 
 	flags::ParseCommandLineFlags(&argc, &argv, true);
-
-	if (FLAGS_input.empty() || FLAGS_output.empty()) {
-		printf("Input and output files not specified.");
-		exit(0);
-	}
 
 	CL_Execution exec;
 	int err = 0;
@@ -114,11 +108,11 @@ int main(int argc, char *argv[]) {
 	exec.exec_queue = queue;
 
 	// Should be built into the cmake file. Haven't thought of how to do this yet.
-	err = exec.compile_kernel(kernel_file, "C:/Users/Tyler/Documents/GitHub/GPU_workgroup_managment/src/uvm_tests/test1/include/rt_device/");
[…]

### 2016-10-12 src/single_kernel/host/main.cpp

@@ -9,6 +9,7 @@
 #define CL_INT_TYPE cl_int
 #define ATOMIC_CL_INT_TYPE cl_int
 #define MY_CL_GLOBAL 
+#define CL_UCHAR_TYPE cl_uchar
 
 #include "profile.h"
 #include "base/commandlineflags.h"

### 2016-10-12 src/uvm_tests/test1/include/rt/cl_execution.h

@@ -113,7 +113,7 @@ class CL_Execution {
   
   
   //roughly from IWOCL tutorial (needs attribution)
-  int compile_kernel(const char* kernel_file, const char * kernel_include) {
+  int compile_kernel(const char* kernel_file, const char * kernel_include, char * extra_include = "") {
     int ret = CL_SUCCESS;
     exec_program = cl::Program(exec_context, loadProgram(kernel_file));
     
@@ -126,13 +126,19 @@ class CL_Execution {
     //Include the rt_device sources
     options << "-I" << kernel_include << " ";
 
+	if (strcmp(extra_include, "") != 0) {
+		options << "-I" << extra_include << " ";
+	}
+
 	//Define the int and atomic int type
 	options << "-D" << "CL_INT_TYPE=int" << " ";
+
+	options << "-D" << "CL_UCHAR_TYPE=uchar" << " ";
 
 	options << "-D" << "ATOMIC_CL_INT_TYPE=atomic_int" << " ";
 
 	options << "-D" << "MY_CL_GLOBAL=__global" << " ";
-    
+  
     //Needed so we know to include Nvidia atomics
     options << check_atomics();
     

### 2016-10-12 src/uvm_tests/test1/include/rt/cl_scheduler.h

@@ -13,6 +13,8 @@ void mk_init_scheduler_ctx(CL_Execution *exec, CL_Scheduler_ctx *s_ctx) {
   s_ctx->pool_lock = (cl_int*)clSVMAlloc(exec->exec_context(), CL_MEM_READ_WRITE, sizeof(cl_int), 4);
   s_ctx->groups_to_kill = (cl_int*)clSVMAlloc(exec->exec_context(), CL_MEM_READ_WRITE, sizeof(cl_int), 4);
   s_ctx->persistent_flag = (cl_int*)clSVMAlloc(exec->exec_context(), CL_MEM_SVM_FINE_GRAIN_BUFFER | CL_MEM_SVM_ATOMICS, sizeof(cl_int), 4);
+  s_ctx->r_ctx_arr = (Restoration_ctx*)clSVMAlloc(exec->exec_context(), CL_MEM_READ_WRITE, sizeof(Restoration_ctx) * MAX_P_GROUPS, 4);
+
 
   *(s_ctx->persistent_flag) = PERSIST_TASK_UNDEF;
   *(s_ctx->groups_to_kill) = 0;
@@ -20,7 +22,9 @@ void mk_init_scheduler_ctx(CL_Execution *exec, CL_Scheduler_ctx *s_ctx) {
 
   for (int i = 0; i < MAX_P_GROUPS; i++) {
 	  s_ctx->task_array[i] = TASK_WAIT;
+	  s_ctx->r_ctx_arr[i].target = 0;
   }
+
   
   *(s_ctx->scheduler_flag) = DEVICE_SCHEDULER_INIT;
   *(s_ctx->available_workgroups) = 0;
@@ -35,6 +39,7 @@ void free_scheduler_ctx(CL_Execution *exec, CL_Scheduler_ctx *s_ctx) {
   clSVMFree(exec->exec_context(), s_ctx->pool_lock);
   clSVMFree(exec->exec_context(), s_ctx->groups_to_kill);
   clSVMFree(exec->exec_context(), s_ctx->persistent_flag);
+  clSVMFree(exec->exec_context(), s_ctx->r_ctx_arr);
 
 }
 
@@ -55,6 +60,8 @@ int set_scheduler_args(cl::Kernel *k, CL_Scheduler_ctx *s_ctx, int &arg_index) {
   err |= clSetKernelArgSVMPointer((*k)(), arg_index, s_ctx->groups_to_kill);
   arg_index++;
[…]

### 2016-10-12 src/uvm_tests/test1/include/rt_common/cl_scheduler.h

@@ -1,5 +1,7 @@
 #pragma once
 
+#include "restoration_ctx.h"
+
 // For kernel code, define INT_TYPE as int, for host code cl_int
 #ifndef CL_INT_TYPE
 #error "CL_INT_TYPE not defined"
@@ -46,5 +48,7 @@ typedef struct {
   MY_CL_GLOBAL ATOMIC_CL_INT_TYPE * groups_to_kill;
   
   MY_CL_GLOBAL ATOMIC_CL_INT_TYPE * persistent_flag;
+
+  MY_CL_GLOBAL Restoration_ctx * r_ctx_arr;
   
 } CL_Scheduler_ctx;

### 2016-10-12 src/scheduler_rt/rt_host/cl_execution.h
did some moving around of my simple test codes and the scheduler runtime code

@@ -113,12 +113,14 @@ class CL_Execution {
   
   
   //roughly from IWOCL tutorial (needs attribution)
-  int compile_kernel(const char* kernel_file, const char * kernel_include, char * extra_include = "") {
+  int compile_kernel(const char* kernel_file, const char * kernel_include, const char * extra_include = "") {
     int ret = CL_SUCCESS;
     exec_program = cl::Program(exec_context, loadProgram(kernel_file));
     
     std::stringstream options;
     options.setf(std::ios::fixed, std::ios::floatfield);
+
+	
     
     //set compiler options here, example below 
     //options << " -cl-fast-relaxed-math";
@@ -129,6 +131,8 @@ class CL_Execution {
 	if (strcmp(extra_include, "") != 0) {
 		options << "-I" << extra_include << " ";
 	}
+
+	std::cout << options.str() << std::endl;
 
 	//Define the int and atomic int type
 	options << "-D" << "CL_INT_TYPE=int" << " ";

### 2016-10-12 kernel_merge/include/ProcessNonPersistentKernelVisitor.h
Extended kernel merge tool to generate mega kernel similar to hand-crafted one from Tyler.  In the process I made the discovery protocol take a scratchpad argument, and changed 'graphics' to 'non-persistent' in the code base.

@@ -24,7 +24,7 @@ public:
 
   bool VisitCallExpr(CallExpr *CE);
 
-  void EmitRewrittenText();
+  void EmitRewrittenText(std::ostream & out);
 
   KernelInfo GetKI() {
     return this->KI;

### 2016-10-12 kernel_merge/src/KernelMerge.cpp

@@ -1,3 +1,5 @@
+#include <fstream>
+
 #include "clang/Tooling/CommonOptionsParser.h"
 #include "clang/Rewrite/Core/Rewriter.h"
 #include "clang/Frontend/TextDiagnosticPrinter.h"
@@ -35,17 +37,26 @@ int main(int argc, const char **argv) {
   ProcessNonPersistentKernelVisitor NonPersistentVisitor(AUs[0].get());
   ProcessPersistentKernelVisitor PersistentVisitor(AUs[1].get());
 
-  llvm::outs() << "#include \"discovery.cl\"\n";
-  llvm::outs() << "#include \"kernel_ctx.cl\"\n";
-  llvm::outs() << "#include \"cl_scheduler.cl\"\n";
-  llvm::outs() << "#include \"iw_barrier.cl\"\n";
-  llvm::outs() << "\n";
-  NonPersistentVisitor.EmitRewrittenText();
-  llvm::outs() << "\n";
-  PersistentVisitor.EmitRewrittenText();
-  llvm::outs() << "\n";
-  llvm::outs() << "\n";
-  llvm::outs() << "kernel void mega_kernel("
+  std::ofstream RestorationContextH;
+  RestorationContextH.open("restoration_ctx.h");
+  RestorationContextH << "#pragma once\n";
+  RestorationContextH << PersistentVisitor.GetRestorationCtx();
+  RestorationContextH.close();
+
+  std::ofstream Merged;
+  Merged.open("merged.cl");
+
+  Merged << "#include \"discovery.cl\"\n";
+  Merged << "#include \"kernel_ctx.cl\"\n";
+  Merged << "#include \"cl_scheduler.cl\"\n";
+  Merged << "#include \"iw_barrier.cl\"\n";
+  Merged << "\n";
+  NonPersistentVisitor.EmitRewrittenText(Merged);
+  Merged << "\n";
+  PersistentVisitor.EmitRewrittenText(Merged);
+  Merged << "\n";
+  Merged << "\n";
+  Merged << "kernel void mega_kernel("
                << NonPersistentVisitor.GetKI().OriginalParameterText << ", "
                << PersistentVisitor.GetKI().OriginalParameterText << ", "
                << "__global IW_barrier * __bar, "
@@ -54,70 +65,24 @@ int main(int argc, const char **argv) {
                << "__global Kernel_ctx * persistent_kernel_ctx, "
[…]

### 2016-10-12 kernel_merge/src/ProcessNonPersistentKernelVisitor.cpp

@@ -28,14 +28,14 @@ bool ProcessNonPersistentKernelVisitor::VisitCallExpr(CallExpr *CE) {
 }
 
 
-void ProcessNonPersistentKernelVisitor::EmitRewrittenText() {
+void ProcessNonPersistentKernelVisitor::EmitRewrittenText(std::ostream & out) {
   const RewriteBuffer *RewriteBuf =
     RW.getRewriteBufferFor(AU->getSourceManager().getMainFileID());
   if (!RewriteBuf) {
     errs() << "Nothing was re-written\n";
     exit(1);
   }
-  llvm::outs() << std::string(RewriteBuf->begin(), RewriteBuf->end());
+  out << std::string(RewriteBuf->begin(), RewriteBuf->end());
 }
 
 void ProcessNonPersistentKernelVisitor::ProcessKernelFunction(FunctionDecl *D) {

### 2016-10-12 kernel_merge/src/ProcessPersistentKernelVisitor.cpp

@@ -25,26 +25,9 @@ void ProcessPersistentKernelVisitor::ProcessWhileStmt(WhileStmt *S) {
 
   auto condition = RW.getRewrittenText(S->getCond()->getSourceRange());
   RW.ReplaceText(S->getCond()->getSourceRange(), "true");
-  RW.InsertTextAfterToken(CS->getLBracLoc(), "\nswitch(__restoration_ctx->target) {\ncase 0:\nif(!(" + condition + ")) { break; }\n");
-  unsigned counter = 1;
-
-  for (auto s : CS->body()) {
-    CallExpr *CE = dyn_cast<CallExpr>(s);
-    if (!CE) {
-      continue;
-    }
-    if ("resizing_global_barrier" == CE->getCalleeDecl()->getAsFunction()->getNameAsString()) {
-      std::stringstream strstr;
-      strstr << "\ncase " << counter << ":\n";
-      strstr << "__restoration_ctx->target = 0;\n";
-
-      SourceLocation SemiLoc = clang::arcmt::trans::findSemiAfterLocation(CE->getLocEnd(), AU->getASTContext());
-      RW.InsertTextAfterToken(SemiLoc, strstr.str());
-      counter++;
-    }
-  }
-
+  RW.InsertTextAfterToken(CS->getLBracLoc(), "\nswitch(__restoration_ctx->target) {\ncase 0:\nif(!(" + condition + ")) { return; }\n");
   RW.InsertTextBefore(CS->getRBracLoc(), "}");
+
 }
 
 bool ProcessPersistentKernelVisitor::VisitCallExpr(CallExpr *CE) {
@@ -59,8 +42,19 @@ bool ProcessPersistentKernelVisitor::VisitCallExpr(CallExpr *CE) {
     RW.InsertTextBefore(CE->getSourceRange().getBegin(), "k_");
   }
   if (name == "resizing_global_barrier") {
+    ForkPointCounter++;
     assert(CE->getNumArgs() == 1);
-    RW.ReplaceText(CE->getSourceRange(), "global_barrier(__bar, __k_ctx)");
+    std::stringstream sstr;
+    sstr << "{ Restoration_ctx to_fork; to_fork.target = " << ForkPointCounter << "; ";
+    for (auto DS : DeclsToRestore) {
+      for (auto D : DS->decls()) {
+        VarDecl *VD = dyn_cast<VarDecl>(D);
+        sstr << "to_fork." << VD->getNameAsString() << " = " << VD->getNameAsString() << "; ";
+      }
+    }
[…]

### 2016-10-12 kernel_merge/stubs/premerge/global_barrier.h

@@ -2,5 +2,6 @@
 #define GLOBAL_BARRIER_H
 
 void resizing_global_barrier();
+void global_barrier();
 
 #endif

### 2016-10-13 write-ups/PLDI/paper.tex
Notes from chat today.

@@
 \section{Introduction}\label{sec:intro}
+
+\ADComment{Points to make: It is important to be able to execute
+  long-running, cooperative applications on GPUs (give examples).
+  However, resource sharing and resource flexibility is also important
+  (give examples, including sharing resources with graphics, sharing
+  resources between compute taks, and being flexible about resources
+  under energy constraints), and these are not supported by today's
+  GPU platforms: long-running applications cause display freezes, etc.
+  Resource-based preemption, whereby workgroup execution is postponed
+  to free up resources for other purposes (or to reduce resources to
+  save energy) can potentially enable resource sharing.  There is
+  evidence in the research literature that efficient resource sharing
+  for GPUs is possible (cite academic works), and signs from industry
+  that this is coming in future architectures.  [Explain more clearly
+    what we mean by resource-based preemption.]  Unfortunately,
+  arbitrary resource-based preemption is fundamentally incompatible
+  with cooperative applications---only trivial examples of cooperation
+  are safe [reduction, load balancing via atomic increment]; the HSA
+  model of guaranteeing preemption in reverse order of IDs allows some
+  cooperative applications to be implemented.  Still, in the general
+  case even a mutex cannot be used safely.}
+
+\ADComment{Say that we believe the problem can only be solved through
+  programming language innovations with appropriate runtime support.
+  We have observed inter-workgroup barriers and work stealing to be
+  the main use cases for cooperation, and so in this paper propose two
+  new language primitives, ckill and cfork, designed to meet the needs
+  of these primitives.  [Describe them a bit.]}
+
+\ADComment{Move on to our main contributions}
+
+\ADComment{This is what we had before.}
 
@@
 
+\section{Motivating example: work stealing}
+
+\ADComment{Use the work stealing example to both introduce OpenCL and
+  also make our key point: that arbitrary preemption will not work.}
+
+\section{The ckill and cfork primitives}
+
+\ADComment{Remind the reader that these are inspired by work stealing
+  and barrier synchonization.  In \mysec\ref{ } we describe the
+  semantics of kernels that use these primitives.  In \mysec\ref{ } we
+  show how work stealing can be implemented.  In \mysec\ref{ } we show
+  how a barrier can be implemented.}
+
+\subsection{Semantics of the new primitives}
+
+\ADComment{General semantics: let $N$ be the maximum number of
+  workgroups the GPU of interest can support [gloss over or deal with
+    issue that this is kernel-dependent].  A kernel is launched with
+  $M$ workgroups, for some $0 < M \leq N$.  Initially these are all
+  scheduled, and are guaranteed to make independent forward progress.}
+
+\ADComment{The purpose of ckill is for a workgroup to volunteer to
+  quit execution of the kernel if the resources that it is using
+  (e.g.\ the associated compute unit) are required for other tasks.  A
+  ckill statement must be reached under workgroup-uniform control
+  flow.  When a workgroup reaches ckill, the workgroup either
+  terminates execution (because the environment requires the
+  workgroup's resources), or continues execution as if the ckill
+  statement were a no=op).  From a semantics perspective this is a
+  non-deterministic choice: an application that invokes ckill should
+  be prepared for either outcome.  In practice, the effect of ckill
+  depends on environmental conditions.}
+
+\ADComment{The purpose of cfork is for a workgroup to indicate that it
+  would be useful to have additional workgroups join the computation,
+  and to specify that the program point associated with the cfork
+  statement would be a suitable place for a new workgroup to start
+  executing.  A cfork statement must also be reached under
+  workgroup-uniform control flow.  If $C$ workgroups are executing
+  when a workgroup reaches a cfork statement, with $0 < C \leq M$,
+  then between 0 and $M-C$ new workgroups commence execution from the
+  following statement.  From a semantics perspective, the number of
+  workgroups that join the computation is nondeterministic, within the
+  permitted range.  In practice, how many workgroups join depends on
+  environmental conditions.}
+
+\ADComment{TODO: describe persistent variables.}
+
+\ADComment{TODO: explain that ckill and cfork atomically update the
+  kernel context (ids).}
+
+\ADComment{TODO: somewhere we need to explain the constraints this
+  places on programming: how ids work, what one can assume about them,
+  etc.}
+
+\ADComment{Now describe the more refined semantics where IDs are
+  relinquished in decreasing order; explain that from a formal
+  perspective thie refines the original semantics.}
+
+\ADComment{In the above, explain at some point how a scheduler can
+  exploit cfork and ckill to utilise the GPU resources.}
+
+\subsection{Preemption-aware work stealing}
+
+\ADComment{Show how work stealing can now be made safe.}
+
+\subsection{A preemption-aware global barrier}
+
+\ADComment{Show how the global barrier can be written in a manner that
+  maintains contiguous ids using the refined semantics.}
+
+\section{Application-specific optimisations}
+
+\ADComment{Explain that for global barriers, we explain how some new
+  primitives can lead to a better implementation.  Explain that for
+  work stealing there are various ways the existing ckill and cfork
+  primitives can be used.}
+
+\subsection{An optimised barrier implementation}
+
+\ADComment{Remark that the barrier of \mysec\ref{sec:naivebarrier} may
+  be inefficient, discuss the use of query primitives to make this
+  more efficient.}
+
+\subsection{Variations on work stealing}
+
+\ADComment{Three cases here: always ckill/cfork, ckill/cfork only when
+  queue is empty, and have a ``mediator'' workgroup to manage things.}
+
+\section{A mega kernel-based prototype}
+
+\ADComment{Explain carefully that doing this properly would require
+  driver support, etc., and then explain what we've actually done.}
+
+\section{Experimental evaluation}
+
+\ADComment{TODO: write up notes from whiteboard chat.}
+
+\section{Related work}
+
+\section{Conclusions and future work}
+
+\ADComment{Next-generation GPUs and drivers for evaluation,
+  implementation in a real driver, migration of work between multiple
+  GPUs.}
+
 \end{document}

### 2016-10-13 src/test_suite/octree/lbabp.cc
Fix the octree example (new name of header)

@@ -1,7 +1,7 @@
 #include <stdio.h>
 
 #include "opencl/opencl.h"
-#include "test_suite/octree/octree_host_dev_shared.h"
+#include "test_suite/octree/shared_host_side.h"
 #include "test_suite/octree/lbabp.h"
 #include "test_suite/octree/helper.h"
 

### 2016-10-13 src/test_suite/octree/octree.cc
Update octree example with OpenCL2.0 atomic operations

@@ -117,7 +117,7 @@ bool Octree::run(unsigned int threads, unsigned int blocks, LBMethod method, int
   kernelOss << kernelFile.rdbuf();
 
   cl::Program program(context, kernelOss.str());
-  err = program.build(devices);
+  err = program.build(devices, "-cl-std=CL2.0");
   if (err != CL_SUCCESS) {
     std::string buildlog;
     buildlog = program.getBuildInfo<CL_PROGRAM_BUILD_LOG>(devices[0], &err);

### 2016-10-13 src/test_suite/octree/octree.cc
Remove useless 'isStatic' argument, since we only implement dynamic (i.e., work-stealing) version

@@ -154,8 +154,7 @@ bool Octree::run(unsigned int threads, unsigned int blocks, LBMethod method, int
   kernel.setArg(7, treeSize);
   kernel.setArg(8, particlesDone);
   kernel.setArg(9, maxChildren);
-  kernel.setArg(10, (int)0);
-  kernel.setArg(11, stealAttempts);
+  kernel.setArg(10, stealAttempts);
 
   cl::NDRange local_size(threads);
   cl::NDRange global_size(blocks * threads);

### 2016-10-13 src/test_suite/octree/octree.cc
Prepare octree example for cfork/ckill: num_pool is constant
The number of pools used to be retrieved from get_num_groups(),
but this will not be constant anymore when we use cfork/ckill, so
we now need to pass it as a constant argument of makeOctree().

@@ -155,6 +155,7 @@ bool Octree::run(unsigned int threads, unsigned int blocks, LBMethod method, int
   kernel.setArg(8, particlesDone);
   kernel.setArg(9, maxChildren);
   kernel.setArg(10, stealAttempts);
+  kernel.setArg(11, blocks);
 
   cl::NDRange local_size(threads);
   cl::NDRange global_size(blocks * threads);

### 2016-10-14 src/test_suite/octree/octree.cc
Inline the initOctree kernel in the makeOctree one

@@ -130,17 +130,7 @@ bool Octree::run(unsigned int threads, unsigned int blocks, LBMethod method, int
 
   lbws.setQueueSize(context, queue, program, 256, blocks);
 
-  cl::Kernel kernel(program, "initOctree", &err);
-  checkErr(err, "kernel constructor for initOctree");
-  kernel.setArg(0, lbws.deviceptr());
-  kernel.setArg(1, lbws.getMaxl());
-  kernel.setArg(2, stealAttempts);
-  kernel.setArg(3, treeSize);
-  kernel.setArg(4, particlesDone);
-  kernel.setArg(5, numParticles);
-
-  queue.enqueueNDRangeKernel(kernel, cl::NullRange, cl::NDRange(1), cl::NDRange(1), NULL, &event);
-  event.wait();
+  cl::Kernel kernel;
 
   kernel = cl::Kernel(program, "makeOctree", &err);
   checkErr(err, "kernel constructor for makeOctree");

### 2016-11-07 src/test_suite/connect_four/common/connect_four_data.h
Inline common header in device code to be able to use kernel_merge

@@ -18,7 +18,7 @@ const CL_INT_TYPE MINUS_INF = -666;
 /* The actual level of lookahead is controlled by the "maxlevel" flag,
  * MAX_LOOKAHEAD is a hard-coded upper-bound. */
 const CL_INT_TYPE MAX_LOOKAHEAD = 7;
-const CL_INT_TYPE NUM_NODE = 10000;
+const CL_INT_TYPE NUM_NODE = 20000;
 
 /*---------------------------------------------------------------------------*/
 

### 2016-11-07 src/host_skeleton/skel.cpp
got pannotia working

@@ -116,6 +116,9 @@ void reset_barrier(CL_Execution *exec, cl::Buffer d_bar) {
 		h_bar.barrier_flags[i] = 0;
 	}
 	h_bar.phase = 0;
+	h_bar.counter = 0;
+	h_bar.sense = 0;
+
 
 	int err = exec->exec_queue.enqueueWriteBuffer(d_bar, CL_TRUE, 0, sizeof(IW_barrier), &h_bar);
 	check_ocl(err);
@@ -526,6 +529,9 @@ void run_merged(CL_Execution *exec) {
 		cl_comm.print_summary_file((FLAGS_output_summary + it).c_str());
 		cl_comm.print_summary();
 		times.push_back(cl_comm.get_persistent_time());
+		//cout << "Check value: " << *(s_ctx.check_value) << endl;
+		cout << "Persistent Error: " << error_persistent << endl;
+		cout << "Non-persistent Error: " << error_non_persistent << endl;
 		cout << endl;
 		CL_Communicator::my_sleep(1000);
 	}
@@ -535,7 +541,7 @@ void run_merged(CL_Execution *exec) {
 	clean_non_persistent_task(exec);
 
 	cout << endl << "error non persistent: " << error_non_persistent << endl;
-	cout << endl << "error persistent: " << error_non_persistent << endl << endl;
+	cout << endl << "error persistent: " << error_persistent << endl << endl;
 
 	cout << "stats for persistent tasks" << endl;
 	cout << "Total time: " << CL_Communicator::reduce_times_ms(times) << " ms" << endl;

### 2016-11-07 src/host_skeleton/skel.cpp
Add exec argument in function checking non persistent task

@@ -56,6 +56,8 @@ using namespace std;
 #include "../graph_apps/bc/bc.h"
 #elif defined PERSISTENT_OCTREE
 #include "octree.h"
+#elif defined PERSISTENT_CONNECT_FOUR
+#include "connect_four.h"
 #else
 #error "No persistent task macro defined? like PERSISTENT_XYZ (check your CMakeLists.txt)"
 #endif
@@ -183,11 +185,11 @@ void run_non_persistent(CL_Execution *exec) {
 		times.push_back(end - begin);
 
 		// check the result
-		if (!check_non_persistent_task()) {
+		if (!check_non_persistent_task(exec)) {
 			error = 1;
 		}
 
-		reset_non_persistent();
+		reset_non_persistent(exec);
 	}
 
 	clean_non_persistent_task(exec);
@@ -397,13 +399,13 @@ void execute_merged_iteration(CL_Execution *exec, CL_Communicator &cl_comm, int 
 			}
 
 			// Reset non-persistent task
-			reset_non_persistent();
+			reset_non_persistent(exec);
 
 			// Launch the task
 			cl_comm.send_task_synchronous(workgroups_for_non_persistent, "first");
 
 			// check the result
-			if (!check_non_persistent_task()) {
+			if (!check_non_persistent_task(exec)) {
 				error = 1;
 			}
 			begin = end;
@@ -505,7 +507,7 @@ void run_merged(CL_Execution *exec) {
 		reset_barrier(exec, d_bar);
 		reset_persistent_task(exec);
 		restart_scheduler(&s_ctx);
-		reset_non_persistent();
+		reset_non_persistent(exec);
 		reset_persistent_task(exec);
 		if (i == 0) {
 			cl_comm.set_record_groups_time_data(true);

### 2016-11-07 src/non_persistent_kernels/matmult/matmult.h

@@ -107,6 +107,8 @@ void init_non_persistent_app(CL_Execution *exec) {
   /* Matrix C */
   c_mem_size = sizeof(cl_int) * FLAGS_A_row * FLAGS_B_col;
   d_C = cl::Buffer(exec->exec_context, CL_MEM_WRITE_ONLY, c_mem_size);
+  err = exec->exec_queue.enqueueFillBuffer(d_C, 0, 0, c_mem_size);
+  check_ocl(err);
 
   free(h_A);
   free(h_B);
@@ -129,28 +131,31 @@ int set_non_persistent_app_args(int arg_index, cl::Kernel k) {
   check_ocl(k.setArg(arg_index++, FLAGS_B_row));
   check_ocl(k.setArg(arg_index++, FLAGS_B_col));
   check_ocl(k.setArg(arg_index++, d_C));
-
   return arg_index;
 }
 
 /*---------------------------------------------------------------------------*/
 
-void reset_non_persistent() {
+void reset_non_persistent(CL_Execution *exec) {
+  cl_int err;
+  err = exec->exec_queue.enqueueFillBuffer(d_C, 0, 0, c_mem_size);
+  check_ocl(err);
   return;
 }
 
 /*---------------------------------------------------------------------------*/
 
-bool check_non_persistent_task() {
+bool check_non_persistent_task(CL_Execution *exec) {
   cl_int err = 0;
   /* h_C is allocated in init */
-  /* err = exec.exec_queue.enqueueReadBuffer(d_C, CL_TRUE, 0, c_mem_size, h_C); */
-  /* check_ocl(err); */
+  err = exec->exec_queue.enqueueReadBuffer(d_C, CL_TRUE, 0, c_mem_size, h_C);
+  check_ocl(err);
 
-  /* int hash = hash_mat(h_C, FLAGS_A_row, FLAGS_B_col); */
-  cout << "Cannot compute Hash since cannot read buffer :( see with Tyler to add exec argument in check_non_persistent " << endl;
+  int hash = hash_mat(h_C, FLAGS_A_row, FLAGS_B_col);
+  /* The hash for a matrix of 100 with seed 1234 is -161045286 */
+  int ref_hash = -161045286;
 
-  return 1;
+  return (hash == ref_hash);
 }
 
 /*---------------------------------------------------------------------------*/

### 2016-11-07 src/non_persistent_kernels/reduce/reduce.h

@@ -49,11 +49,11 @@ int set_non_persistent_app_args(int arg_index, cl::Kernel k) {
 	return arg_index;
 }
 
-void reset_non_persistent() {
+void reset_non_persistent(CL_Execution *exec) {
 	*graphics_result = INT_MAX;
 }
 
-bool check_non_persistent_task() {
+bool check_non_persistent_task(CL_Execution *exec) {
 	return arr_min == *graphics_result;
 }
 

### 2016-11-07 src/host_skeleton/skel.cpp
Revert passing argument exec in skel interface for non persistent

@@ -188,11 +188,11 @@ void run_non_persistent(CL_Execution *exec) {
 		times.push_back(end - begin);
 
 		// check the result
-		if (!check_non_persistent_task(exec)) {
+		if (!check_non_persistent_task()) {
 			error = 1;
 		}
 
-		reset_non_persistent(exec);
+		reset_non_persistent();
 	}
 
 	clean_non_persistent_task(exec);
@@ -402,13 +402,13 @@ void execute_merged_iteration(CL_Execution *exec, CL_Communicator &cl_comm, int 
 			}
 
 			// Reset non-persistent task
-			reset_non_persistent(exec);
+			reset_non_persistent();
 
 			// Launch the task
 			cl_comm.send_task_synchronous(workgroups_for_non_persistent, "first");
 
 			// check the result
-			if (!check_non_persistent_task(exec)) {
+			if (!check_non_persistent_task()) {
 				error = 1;
 			}
 			begin = end;
@@ -510,7 +510,7 @@ void run_merged(CL_Execution *exec) {
 		reset_barrier(exec, d_bar);
 		reset_persistent_task(exec);
 		restart_scheduler(&s_ctx);
-		reset_non_persistent(exec);
+		reset_non_persistent();
 		reset_persistent_task(exec);
 		if (i == 0) {
 			cl_comm.set_record_groups_time_data(true);

### 2016-11-07 src/non_persistent_kernels/matmult/matmult.h

@@ -136,26 +136,27 @@ int set_non_persistent_app_args(int arg_index, cl::Kernel k) {
 
 /*---------------------------------------------------------------------------*/
 
-void reset_non_persistent(CL_Execution *exec) {
-  cl_int err;
-  err = exec->exec_queue.enqueueFillBuffer(d_C, 0, 0, c_mem_size);
-  check_ocl(err);
-  return;
+void reset_non_persistent() {
+  /* cl_int err; */
+  /* err = exec->exec_queue.enqueueFillBuffer(d_C, 0, 0, c_mem_size); */
+  /* check_ocl(err); */
+  /* return; */
 }
 
 /*---------------------------------------------------------------------------*/
 
-bool check_non_persistent_task(CL_Execution *exec) {
-  cl_int err = 0;
-  /* h_C is allocated in init */
-  err = exec->exec_queue.enqueueReadBuffer(d_C, CL_TRUE, 0, c_mem_size, h_C);
-  check_ocl(err);
+bool check_non_persistent_task() {
+  /* cl_int err = 0; */
+  /* /\* h_C is allocated in init *\/ */
+  /* err = exec->exec_queue.enqueueReadBuffer(d_C, CL_TRUE, 0, c_mem_size, h_C); */
+  /* check_ocl(err); */
 
-  int hash = hash_mat(h_C, FLAGS_A_row, FLAGS_B_col);
-  /* The hash for a matrix of 100 with seed 1234 is -161045286 */
-  int ref_hash = -161045286;
+  /* int hash = hash_mat(h_C, FLAGS_A_row, FLAGS_B_col); */
+  /* /\* The hash for a matrix of 100 with seed 1234 is -161045286 *\/ */
+  /* int ref_hash = -161045286; */
 
-  return (hash == ref_hash);
+  /* return (hash == ref_hash); */
+  return 0;
 }
 
 /*---------------------------------------------------------------------------*/

### 2016-11-07 src/non_persistent_kernels/reduce/reduce.h

@@ -49,11 +49,11 @@ int set_non_persistent_app_args(int arg_index, cl::Kernel k) {
 	return arg_index;
 }
 
-void reset_non_persistent(CL_Execution *exec) {
+void reset_non_persistent() {
 	*graphics_result = INT_MAX;
 }
 
-bool check_non_persistent_task(CL_Execution *exec) {
+bool check_non_persistent_task() {
 	return arr_min == *graphics_result;
 }
 

### 2016-11-07 src/non_persistent_kernels/matmult/matmult.h
Add host-side matrix mult

@@ -18,6 +18,8 @@ size_t c_mem_size;
 cl::Buffer d_C;
 cl_int *h_C;
 
+int h_hash;
+
 /*---------------------------------------------------------------------------*/
 
 const char* non_persistent_app_name() {
@@ -61,6 +63,29 @@ int hash_mat(cl_int *M, int num_row, int num_col)
     col++;
   }
   return hash;
+}
+
+/*---------------------------------------------------------------------------*/
+
+void host_matrix_multiply(cl_int *A, int A_row, int A_col,
+                         cl_int *B, int B_row, int B_col,
+                         cl_int *C)
+{
+  int C_row = A_row;
+  int C_col = B_col;
+
+  if (A_col != B_row) {
+    return;
+  }
+
+  for (int r = 0 ; r < C_row; r++) {
+    for (int c = 0; c < C_col; c++) {
+      C[(r * C_row) + c] = 0;
+      for (int i = 0; i < A_col; i++) {
+        C[(r * C_row) + c] += A[(r * A_row) + i] * B[(i * B_row) + c];
+      }
+    }
+  }
 }
 
 /*---------------------------------------------------------------------------*/
@@ -114,10 +139,22 @@ void init_non_persistent_app(CL_Execution *exec) {
   free(h_B);
 
   /* Allocate and keep host buffer for C */
-  h_C = (cl_int *)malloc(c_mem_size);
+  h_C = (cl_int *)calloc(FLAGS_A_row * FLAGS_B_col, sizeof(cl_int));
   if (h_C == NULL) {
+    cout << "calloc failed" << endl;
     exit(EXIT_FAILURE);
   }
+
+  /* Compute C to obtain hash */
+  host_matrix_multiply(h_A, FLAGS_A_row, FLAGS_A_col,
+                       h_B, FLAGS_B_row, FLAGS_B_col,
+                       h_C);
+
+  h_hash = hash_mat(h_C, FLAGS_A_row, FLAGS_B_col);
+
+  cout << "matmult: host side hash: " << h_hash << endl;
+
+  memset(h_C, 0, c_mem_size);
 }
 
 /*---------------------------------------------------------------------------*/

### 2016-11-07 src/non_persistent_kernels/matmult/matmult.h
Use svm_C

@@ -15,8 +15,7 @@ size_t b_mem_size;
 cl::Buffer d_B;
 
 size_t c_mem_size;
-cl::Buffer d_C;
-cl_int *h_C;
+cl_int *svm_C;
 
 int h_hash;
 
@@ -131,30 +130,27 @@ void init_non_persistent_app(CL_Execution *exec) {
 
   /* Matrix C */
   c_mem_size = sizeof(cl_int) * FLAGS_A_row * FLAGS_B_col;
-  d_C = cl::Buffer(exec->exec_context, CL_MEM_WRITE_ONLY, c_mem_size);
-  err = exec->exec_queue.enqueueFillBuffer(d_C, 0, 0, c_mem_size);
-  check_ocl(err);
-
-  free(h_A);
-  free(h_B);
-
-  /* Allocate and keep host buffer for C */
-  h_C = (cl_int *)calloc(FLAGS_A_row * FLAGS_B_col, sizeof(cl_int));
-  if (h_C == NULL) {
-    cout << "calloc failed" << endl;
+  svm_C = (cl_int *)clSVMAlloc(exec->exec_context(), CL_MEM_SVM_FINE_GRAIN_BUFFER, c_mem_size, 4);
+  if (svm_C == NULL) {
+    printf("clSVMAlloc failed\n");
     exit(EXIT_FAILURE);
   }
+
+  /* Allocate and keep host buffer for C */
 
   /* Compute C to obtain hash */
   host_matrix_multiply(h_A, FLAGS_A_row, FLAGS_A_col,
                        h_B, FLAGS_B_row, FLAGS_B_col,
-                       h_C);
+                       svm_C);
 
-  h_hash = hash_mat(h_C, FLAGS_A_row, FLAGS_B_col);
+  h_hash = hash_mat(svm_C, FLAGS_A_row, FLAGS_B_col);
 
   cout << "matmult: host side hash: " << h_hash << endl;
 
-  memset(h_C, 0, c_mem_size);
+  memset(svm_C, 0, c_mem_size);
+
+  free(h_A);
+  free(h_B);
 }
 
 /*---------------------------------------------------------------------------*/
@@ -167,7 +163,7 @@ int set_non_persistent_app_args(int arg_index, cl::Kernel k) {
   check_ocl(k.setArg(arg_index++, d_B));
   check_ocl(k.setArg(arg_index++, FLAGS_B_row));
[…]

### 2016-11-07 src/non_persistent_kernels/matmult/matmult.h
Add hashing and SVM variables in device code (matmult)

@@ -17,7 +17,9 @@ cl::Buffer d_B;
 size_t c_mem_size;
 cl_int *svm_C;
 
-int h_hash;
+int ref_hash;
+cl_int *svm_counter;
+cl_int *svm_hash;
 
 /*---------------------------------------------------------------------------*/
 
@@ -136,18 +138,21 @@ void init_non_persistent_app(CL_Execution *exec) {
     exit(EXIT_FAILURE);
   }
 
-  /* Allocate and keep host buffer for C */
-
   /* Compute C to obtain hash */
   host_matrix_multiply(h_A, FLAGS_A_row, FLAGS_A_col,
                        h_B, FLAGS_B_row, FLAGS_B_col,
                        svm_C);
 
-  h_hash = hash_mat(svm_C, FLAGS_A_row, FLAGS_B_col);
-
-  cout << "matmult: host side hash: " << h_hash << endl;
+  ref_hash = hash_mat(svm_C, FLAGS_A_row, FLAGS_B_col);
 
   memset(svm_C, 0, c_mem_size);
+
+  /* counter and hash */
+  svm_counter = (cl_int *)clSVMAlloc(exec->exec_context(), CL_MEM_SVM_FINE_GRAIN_BUFFER, sizeof(cl_int), 4);
+  svm_hash = (cl_int *)clSVMAlloc(exec->exec_context(), CL_MEM_SVM_FINE_GRAIN_BUFFER, sizeof(cl_int), 4);
+
+  *svm_counter = 0;
+  *svm_hash = 0;
 
   free(h_A);
   free(h_B);
@@ -164,37 +169,31 @@ int set_non_persistent_app_args(int arg_index, cl::Kernel k) {
   check_ocl(k.setArg(arg_index++, FLAGS_B_row));
   check_ocl(k.setArg(arg_index++, FLAGS_B_col));
   check_ocl(clSetKernelArgSVMPointer(k(), arg_index++, svm_C));
+  check_ocl(clSetKernelArgSVMPointer(k(), arg_index++, svm_counter));
+  check_ocl(clSetKernelArgSVMPointer(k(), arg_index++, svm_hash));
   return arg_index;
 }
 
 /*---------------------------------------------------------------------------*/
 
 void reset_non_persistent() {
-  /* cl_int err; */
-  /* err = exec->exec_queue.enqueueFillBuffer(d_C, 0, 0, c_mem_size); */
[…]

### 2016-11-07 src/run_test.py
Layout for python script for running tests

@@ -1,0 +1,92 @@
+from __future__ import print_function
+import sys
+import os
+import subprocess
+
+def fake_placeholder(x):
+    return
+
+EXE_PATH     = ""
+DATA_PATH    = ""
+ITERS        = 1
+PRINT        = fake_placeholder
+DATA_PRINT   = fake_placeholder
+NAME_OF_CHIP = ""
+
+PROGRAMS = {
+    "pannotia_color"
+}
+
+PROGRAM_DATA = {
+    "pannotia_color" : [ {
+        "input" : os.path.join("inputs", "color", "ecology1.graph"),
+        "solution" : os.path.join("solutions", "color_ecology.txt") } ]
+}
+
+def run_suite():
+    for p in PROGRAMS:
+        for d in PROGRAM_DATA[p]:
+            exe = os.path.join(EXE_PATH, p)
+            graph_in = os.path.join(DATA_PATH, d["input"])
+            graph_sol = os.path.join(DATA_PATH, d["solution"])
+            cmd = [exe, "--graph_file", graph_in, "--graph_solution_file", graph_sol]
+            print("WOULD RUN: ", cmd)
+            # p_obj = subprocess.Popen(cmd, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
+            # ret_code = p_obj.wait()
+            # sout, serr = p_obj.communicate()
+            # print("* stdout:")
+            # print(sout)
+            # print("* stderr:")
+            # print(serr)
+            # if (ret_code != 0):
+            #     print("Error running " + " ".join(cmd))
+            #     exit(ret_code)
+
+def main():
+
+    global EXE_PATH
+    global DATA_PATH
+    global PRINT
+    global DATA_PRINT
+    global NAME_OF_CHIP
+
+    # if len(sys.argv) != 5:
+    #     print "Please provide the following arguments:"
+    #     print "path to executables, path to data, name of run, name of chip"
+    #     return 1
+
+    if len(sys.argv) != 3:
+        print("Please provide the following arguments:")
+        print("path to executables, path to data")
+        return 1
+
+    EXE_PATH = sys.argv[1]
+    DATA_PATH = sys.argv[2]
+
+    run_suite()
+
+    #NAME_OF_CHIP = sys.argv[4]
+    # log_file = sys.argv[3] + "_exec.log"
+    # print "recording all to " + log_file
[…]

### 2016-11-08 write-ups/PLDI/paper.tex
Started on paper intro based on notes with Tyler.

@@
 \section{Introduction}\label{sec:intro}
+
+\ADComment{Be sure to say that a GPU \emph{kernel} is.}
+
+Although GPU acceleration of general-purpose computation has focused
+on regular data-parallel algorithms, many data-parallel algorithms are
+\emph{irregular} \ADComment{say what that means}.  There is growing
+interest in accelerating irregular data parallel algorithms on GPUs,
+including \ADComment{name some algorithms and say why they matter},
+with several recent notable successes~\cite{...}.
+
+By their very nature, irregular algorithms demand synchronization
+between the workgroups that execute a GPU kernel.  For example, work
+stealing algorithms require the ability for one workgroup to steal
+from the queue of another workgroup, which can be safely achieved by
+protecting the queue of each workgroup using a mutex.  Breadth-first
+graph traversal uses a frontier-based approach that requires a barrier
+synchronization between all workgroups to allow computation to proceed
+from one frontier to the next.  Inter-workgroup synchronization
+involves \emph{blocking} between workgroups, e.g.\ workgroup A may be
+blocked until workgroup B releases a mutex, or blocked until all other
+workgroups arrive at an inter-workgroup barrier.  Thus, irregular data
+parallel algorithms that rely on inter-workgroup synchronization are
+\emph{blocking algorithms}.\footnote{\ADComment{Perhaps make the point
+    that inter-workgroup synchronization is possible in very rare
+    cases without blocking, e.g. threadfence reduction.}}
+
+To operate correctly, a blocking concurrent algorithm requires
+\emph{fair} scheduling of threads.  In the context of irregular data
+parallel algorithms accelerated on a GPU, this translates to requiring
+fair scheduling of workgroups.  Without fairness, a workgroup may fail
+to make progress, leading to starvation.  For example, if workgroup B
+holds a mutex and workgroup A is spinning, waiting for the mutex to be
+released, an unfair scheduler can forever schedule A, so that B never
+releases a mutex.  Similarly, an unfair scheduler can induce
+starvation if workgroup A is at an inter-workgroup barrier, spinning
+until other workgroups arrive.
+
+Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
+and HSA~\cite{...}---specify very few constraints on the manner in
+which workgroups are scheduled.  In particular, they do \emph{not}
+guarantee fair scheduling, and fair scheduling is not observed in
+practice.  Roughly speaking, each workgroup executing a GPU kernel is
+mapped to a hardware \emph{compute unit}.  If a kernel is executed
+with more workgroups than there are compute units then at least two
+workgroups must share a single compute unit.  The simplest way to
+handle this is an \emph{occupancy-bound} execution model, whereby once
+a workgroup has commenced execution on a compute unit (is has become
+\emph{occupant}), the workgroup has exclusive access to the compute
+unit until it has finished execution, after which a subsequent
+workgroup can be bound to the compute unit.  There is empirical
+evidence that the occupancy-bound execution model is widely
+implemented in practice~\cite{...}.
+
+Clearly, the occupancy-bound execution model does not provide fair
+scheduling between workgroups in general: if all compute units are
+occupied then a not-yet-occupant workgroup will not be scheduled until
+some occupant workgroup has completed execution.  Yet the execution
+model \emph{does} provide fair scheduling between occupant workgroups,
+which are bound to separate compute units that operate in parallel.
+Current implementations of irregular data parallel algorithms on GPUs
+assume this occupancy-bound execution model, exploiting it by
+scheduling no more workgroups than there are available compute units.
+Because current GPUs do provide the occupancy-bound execution model in
+practice, this strategy allows starvation-free blocking
+synchronization between workgroups.
+
+However, none of the current GPU programming models mandate the
+occupancy-bound execution model, and there are clear reasons why this
+model is not actually desirable from the point-of-view of GPU vendors
+nor end users.  First, the execution model does not enable
+multi-tasking, because a workgroup effectively \emph{owns} a compute
+unit until the workgroup has completed execution.  If a long-running
+kernel occupies all workgroups, the GPU cannot be used for other
+tasks, such as graphics rendering or shorter, higher-priority
+computations, in the meantime.  Precisely because of this problem,
+operating systems employ a GPU \emph{watchdog} to heuristically detect
+and kill long-running computations on the GPU that is being used for
+rendering.  If one disables this watchdog and launches a long-running
+kernel, the display becomes unresponsive for the duration of kernel
+execution.  Second, \emph{energy throttling} is an important concern
+for energy-constrained devices, such as smartphones and tablets.  In
+the context of mobile GPU computing, it is desirable for a GPU driver
+to be able to power down one or more compute units if the battery
+level is low or if the user puts the device into a low power state.
+Implementing energy throttling requires being able to suspend the
+execution of a workgroup, power down the compute unit to which the
+workgroup was bound, and schedule the workgroup for later execution on
+one of the other compute units when it becomes free.
+
+Our assessment, informed by discussions with a number of industrial
+practitioners who have been involved in the OpenCL and/or HSA
+standardisation efforts~\cite{CitePersonalCommunication}, is that GPU
+vendors are reluctant to commit to the occupancy-bound execution model
+that they currently implement, due to the need for multi-tasking and
+energy throttling, yet do not wish to guarantee fair scheduling, due
+to the runtime overhead associated with supporting full preemption of
+workgroups, e.g.\ based on time-slicing.  Vendors instead wish to
+retain the essence of the occupancy-bound execution model, which is
+simple to implement efficiently, providing preemption for the special
+cases associated with multi-tasking for high-priority jobs (such as
+graphics) and energy throttling.  In particular, they do not want to
+commit to fair scheduling of a workgroup once it has been preempted.
+As argued above, without fairness it is not possible to implement
+non-blocking algorithms, including data parallel irregular algorithms,
+safely.
+
+  Doesn't fascilitate energy throttling or
+  multi-tasking. Multi-tasking is important esp. for responsive
+  systems. (Say something colourful here, its an enabler. A use case
+  for today and many use cases for tomorrow.)
+
+  Our Proposal: Non-blocking GPU programs (majority of kernels today)
+  can be exucuted however the GPU scheduler sees fit. Algorithm that
+  require a degree of fairness (e.g. blocking algorithms) must be
+  written a small set of language extensions, and launched in a
+  cooperative mode. This indicates that the scheduler must interact
+  with these kernels in a special manner via the language
+  extensionts. These kernels are written with language extensions that
+  allow designated points at which a workgroup can be de-scheduled or
+  re-scheduled to be identified.
+
+  In traditonal co-operative multi-tasking, a task reliquishes control
+  and regains control in a stateful manner. We propose a lighter
+  weight model we believe to suitible for the type of blocking
+  algorithms we are aware of for GPUs. At a point where the algorithm
+  could proceed with fewer workgroups, a workgroup can offer itself
+  to be killed. This indicates to the scheduler that the algorithm
+  would happily proceed with the workgroup continuing, but is prepared
+  for the workgroup to be destroyed for good. In addition, the programmer
+  identifies designated points where it would be beneificial to join the
+  computation if available. On offering to fork, the algorithm indicates
+  that it is prepared to proceed with the existing workgroups, but is able
+  to take advantage of one or more additional workgroups commencing at the
+  given program point. Inheriting the state of the invoking workgroup.
+
+  The scheduler agrees to scheduler the workgroups fairly, but is at
+  liberty to eliminate workgroups for good when they offer themselves
+  to be killed. This allows the schduler to repond to changing depands
+  on the GPU, e.g. to implement energy throttling.
+
+  *Note: its not really completely fair, its occupancy bound
+
+  This programming model creates a contract between the scheduler
+  and cooperative kernel:
+
+  Functional requirements: Scheduler agrees that cooperative kernels
+  have fairly scheduled workgroups. The Cooperative kernel is written
+  in a way that can tolerate threads leaving or joining at the
+  designated areas (e.g. a thread should not opt to be killed if it
+  contains local work to do, and a programmer should be prepared for
+  the number of executing workgroups to change dynamically during
+  computation).
+
+  Non-Functional: The cooperative kernel must ensure that offer_kill
+  is executed frequently enough such that the system can accomodate
+  soft-real time constraints (e.g. allowing a smooth framerate for
+  graphics or throttling the energy before the mobile battery drains).
+  To enable the cooperative execute efficiently, the scheduler should
+  adopt a policy that avoids under utililasion of hardware resources
+  for the cooperative kernel (e.g. the scheduler should not kill
+  workgroups unless they are required).
+
+  The programming model is backwards compatible because semantically
+  it is fine for these things to be ignored and GPUs already give
+  occupant bound fairness.
+
+  Our contributions:
+
+  - Programming model in detail (semantics of new operations)
+
+  - Prototype implementation of the programming model using OpenCL 2.0
+    advance features to build the scheudler and implement a clang
+    based compiler to transform input programs using the language
+    extensions to lower-level programs that interact with the
+    scheduler. Our prototype simulates multiple independent kernels
+    into a mega-kernel. A vendor specific implementation could achieve
+    higher efficiency through modications to the GPU driver that may
+    take advantage of hardware features. (e.g. mega kernel might not
+    be good to compile). Its reasonable to think our proof of concept
+    provides a lower bound on how well we can do.
+
+  - We ported a large set of blocking algorithms to our programming
+    model. We evaluate co-operative scheduling with a cooperative
+    kernel and non-coperative kernel. GPUs from multiple vendors and
+    access various things.
+
+  - Our results are promising.
+
+  - In the evaluation, make a note that the applications we worked on
+    Namely if they use only a barrier or work-stealing idiom, are
+    straightforward to adopt to our programming model.
 

### 2016-11-08 write-ups/PLDI/paper.tex
Got a bit further with intro.

@@
 fair scheduling of workgroups.  Without fairness, a workgroup may fail
-to make progress, leading to starvation.  For example, if workgroup B
-holds a mutex and workgroup A is spinning, waiting for the mutex to be
-released, an unfair scheduler can forever schedule A, so that B never
-releases a mutex.  Similarly, an unfair scheduler can induce
-starvation if workgroup A is at an inter-workgroup barrier, spinning
-until other workgroups arrive.
+to make progress, leading to starvation.  For example, if workgroup A
+holds a mutex and workgroup B is spin-waiting for the mutex to be
+released, an unfair scheduler can forever schedule B, so that A never
+releases the mutex.  Similarly, an unfair scheduler can induce
+starvation if workgroup A is at an inter-workgroup barrier awaiting
+the arrival of other workgroups, by allowing A to spin forever so that
+the other workgroups do not progress to the barrier.
 
@@
 evidence that the occupancy-bound execution model is widely
-implemented in practice~\cite{...}.
+implemented by today's GPU devices and drivers~\cite{...}, and it is
+evidently a simple model to implement efficiently.
 
@@
 occupied then a not-yet-occupant workgroup will not be scheduled until
-some occupant workgroup has completed execution.  Yet the execution
-model \emph{does} provide fair scheduling between occupant workgroups,
-which are bound to separate compute units that operate in parallel.
+some occupant workgroup completes execution.  Yet the execution
+model \emph{does} provide fair scheduling between \emph{occupant} workgroups,
+because they are bound to separate compute units that operate in parallel.
 Current implementations of irregular data parallel algorithms on GPUs
@@
 
-However, none of the current GPU programming models mandate the
-occupancy-bound execution model, and there are clear reasons why this
-model is not actually desirable from the point-of-view of GPU vendors
-nor end users.  First, the execution model does not enable
+However, none of the current GPU programming models actually mandate
+the occupancy-bound execution model, and there are clear reasons why
+this model is not actually desirable from the point-of-view of GPU
+vendors nor end users.  First, the execution model does not enable
 multi-tasking, because a workgroup effectively \emph{owns} a compute
@@
 execution.  Second, \emph{energy throttling} is an important concern
-for energy-constrained devices, such as smartphones and tablets.  In
-the context of mobile GPU computing, it is desirable for a GPU driver
-to be able to power down one or more compute units if the battery
-level is low or if the user puts the device into a low power state.
-Implementing energy throttling requires being able to suspend the
-execution of a workgroup, power down the compute unit to which the
-workgroup was bound, and schedule the workgroup for later execution on
-one of the other compute units when it becomes free.
+for battery-powered devices, such as smartphones and tablets.  It is
+desirable for a mobile GPU driver to be able to power down one or more
+compute units if the battery level is low or if the user puts the
+device into a low power state.  Implementing energy throttling
+requires being able to suspend the execution of a workgroup, power
+down the compute unit to which the workgroup was bound, and schedule
+the workgroup for later execution on a different compute unit.
 
@@
 retain the essence of the occupancy-bound execution model, which is
-simple to implement efficiently, providing preemption for the special
-cases associated with multi-tasking for high-priority jobs (such as
+simple to implement efficiently, but provide preemption for the special
+cases associated with multi-tasking for high-priority jobs (most crucially
 graphics) and energy throttling.  In particular, they do not want to
@@
 
-  Doesn't fascilitate energy throttling or
-  multi-tasking. Multi-tasking is important esp. for responsive
-  systems. (Say something colourful here, its an enabler. A use case
-  for today and many use cases for tomorrow.)
+\paragraph{Our proposal}
 
@@
 
-  Non-Functional: The cooperative kernel must ensure that offer_kill
+  Non-Functional: The cooperative kernel must ensure that offer\_kill
   is executed frequently enough such that the system can accomodate
@@
     straightforward to adopt to our programming model.
+
+  Doesn't fascilitate energy throttling or
+  multi-tasking. Multi-tasking is important esp. for responsive
+  systems. (Say something colourful here, its an enabler. A use case
+  for today and many use cases for tomorrow.)
 

### 2016-11-08 src/run_test.py
Add script for running batch of experiments run_test.py

@@ -14,14 +14,60 @@ DATA_PRINT   = fake_placeholder
 NAME_OF_CHIP = ""
 
 PROGRAMS = {
-    "pannotia_color"
+    "pannotia_color",
+    #"pannotia_mis",
+    #"pannotia_bc",
+    #"pannotia_sssp"
 }
 
 PROGRAM_DATA = {
-    "pannotia_color" : [ {
-        "input" : os.path.join("inputs", "color", "ecology1.graph"),
-        "solution" : os.path.join("solutions", "color_ecology.txt") } ]
+
+    "pannotia_color" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
+                           "solution" : os.path.join("solutions", "color_ecology.txt") },
+                         # { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
+                         #   "solution" : os.path.join("solutions", "color_G3_circuit.txt") }
+    ],
+
+    "pannotia_mis" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
+                         "solution" : os.path.join("solutions", "color_ecology.txt") },
+                       { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
+                         "solution" : os.path.join("solutions", "color_G3_circuit.txt") } ],
+
+    "pannotia_bc" : [ { "input" : os.path.join("inputs", "bc", "1k_128k.gr"),
+                        "solution" : os.path.join("solutions", "bc_1k_128k.gr") },
+                      #{ "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
+                      #  "solution" : os.path.join("solutions", "bc_2k_1M.gr") }
+    ],
+
+    "pannotia_sssp" : [ { "input" : os.path.join("inputs", "sssp", "USA-road-d.NW.gr"),
+                          "solution" : os.path.join("solutions", "sssp_usa.txt") } ]
 }
+
+def my_print(file_handle, data):
+    print(data)
+    file_handle.write(data + os.linesep)
+
+def exec_cmd(cmd, extr_args=[], flag=""):
+    finalcmd = cmd + extr_args
+    PRINT("====================== RUN COMMAND =======================")
+    PRINT(flag)
+    for s in finalcmd:
+        PRINT(s)
+    PRINT("----------------------------------------------------------")
+    p_obj = subprocess.Popen(cmd, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
[…]

### 2016-11-08 src/run_test.py
Enable all pannotia tests

@@ -15,17 +15,17 @@ NAME_OF_CHIP = ""
 
 PROGRAMS = {
     "pannotia_color",
-    #"pannotia_mis",
-    #"pannotia_bc",
-    #"pannotia_sssp"
+    "pannotia_mis",
+    "pannotia_bc",
+    "pannotia_sssp"
 }
 
 PROGRAM_DATA = {
 
     "pannotia_color" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
                            "solution" : os.path.join("solutions", "color_ecology.txt") },
-                         # { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
-                         #   "solution" : os.path.join("solutions", "color_G3_circuit.txt") }
+                         { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
+                           "solution" : os.path.join("solutions", "color_G3_circuit.txt") }
     ],
 
     "pannotia_mis" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
@@ -35,8 +35,8 @@ PROGRAM_DATA = {
 
     "pannotia_bc" : [ { "input" : os.path.join("inputs", "bc", "1k_128k.gr"),
                         "solution" : os.path.join("solutions", "bc_1k_128k.gr") },
-                      #{ "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
-                      #  "solution" : os.path.join("solutions", "bc_2k_1M.gr") }
+                      { "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
+                       "solution" : os.path.join("solutions", "bc_2k_1M.gr") }
     ],
 
     "pannotia_sssp" : [ { "input" : os.path.join("inputs", "sssp", "USA-road-d.NW.gr"),

### 2016-11-08 src/pannotia/graph_apps/bc/bc.h
Update pannotia test with matmult non-persistent

@@ -190,7 +190,7 @@ bool diff_solution_file_int(float * a, const char * solution_fname, int v) {
 			ret = false;
 			break;
 		}
-		if (my_abs(compare - a[i]) > 1) { // .001 as a threshold seems to work.
+		if (my_abs(compare - a[i]) > 8) { // .001 as a threshold seems to work.
 			printf("ERROR %d found %f expected %f\n", i, compare, a[i]);
 			ret = false;
 			break;

### 2016-11-08 src/run_test.py
Add collection of stat files

@@ -2,12 +2,15 @@ from __future__ import print_function
 import sys
 import os
 import subprocess
+import shutil
+import glob
 
 def fake_placeholder(x):
     return
 
 EXE_PATH     = ""
 DATA_PATH    = ""
+STAT_PATH    = ""
 ITERS        = 1
 PRINT        = fake_placeholder
 DATA_PRINT   = fake_placeholder
@@ -23,24 +26,31 @@ PROGRAMS = {
 PROGRAM_DATA = {
 
     "pannotia_color" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
-                           "solution" : os.path.join("solutions", "color_ecology.txt") },
+                           "solution" : os.path.join("solutions", "color_ecology.txt"),
+                           "stat" : "color_ecology" },
                          { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
-                           "solution" : os.path.join("solutions", "color_G3_circuit.txt") }
+                           "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
+                           "stat" : "color_G3_circuit" }
     ],
 
     "pannotia_mis" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
-                         "solution" : os.path.join("solutions", "color_ecology.txt") },
+                         "solution" : os.path.join("solutions", "color_ecology.txt"),
+                         "stat" : "mis_ecology" },
                        { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
-                         "solution" : os.path.join("solutions", "color_G3_circuit.txt") } ],
+                         "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
+                         "stat" : "mis_G3_circuit" } ],
 
     "pannotia_bc" : [ { "input" : os.path.join("inputs", "bc", "1k_128k.gr"),
-                        "solution" : os.path.join("solutions", "bc_1k_128k.gr") },
+                        "solution" : os.path.join("solutions", "bc_1k_128k.txt"),
+                        "stat" : "bc_1k_128k" },
                       { "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
-                       "solution" : os.path.join("solutions", "bc_2k_1M.gr") }
[…]

### 2016-11-08 src/run_test.py
Add flag for frequency of non-persistent task

@@ -8,13 +8,14 @@ import glob
 def fake_placeholder(x):
     return
 
-EXE_PATH     = ""
-DATA_PATH    = ""
-STAT_PATH    = ""
-ITERS        = 1
-PRINT        = fake_placeholder
-DATA_PRINT   = fake_placeholder
-NAME_OF_CHIP = ""
+EXE_PATH      = ""
+DATA_PATH     = ""
+STAT_PATH     = ""
+ITERS         = 1
+PRINT         = fake_placeholder
+DATA_PRINT    = fake_placeholder
+NAME_OF_CHIP  = ""
+NON_PRST_FREQ = ""
 
 PROGRAMS = {
     "pannotia_color",
@@ -104,11 +105,14 @@ def run_suite():
             exe = os.path.join(EXE_PATH, p)
             graph_in = os.path.join(DATA_PATH, d["input"])
             graph_sol = os.path.join(DATA_PATH, d["solution"])
-            cmd = [exe, "--graph_file", graph_in, "--graph_solution_file", graph_sol]
-            exec_cmd(cmd, ["--run_persistent", "2", "--threads_per_wg", "128"], "== standalone")
-            # standalone execution does not produce stat files, so nothing to collect here
-            exec_cmd(cmd, ["--skip_tasks", "1", "--merged_iterations", "2", "--threads_per_wg", "128"], "== merged without task")
-            # exec_cmd(cmd, ["--merged_iterations", "1", "--threads_per_wg", "128"], "== merged without task")
+            cmd = [exe, "--non_persistent_frequency", NON_PRST_FREQ, "--graph_file", graph_in, "--graph_solution_file", graph_sol, "--threads_per_wg", "128"]
+            # standalone, which does not produce stat files, so nothing to collect here
+            # exec_cmd(cmd, ["--run_persistent", "2"], "== standalone")
+            # merged without tasks, stats to collect afterwards
+            # exec_cmd(cmd, ["--skip_tasks", "1", "--merged_iterations", "2"], "== merged without task")
+            # collect_stats(d)
+            # merged
+            exec_cmd(cmd, ["--merged_iterations", "1"], "== merged")
             collect_stats(d)
 
 def main():
@@ -119,10 +123,11 @@ def main():
     global PRINT
     global DATA_PRINT
     global NAME_OF_CHIP
+    global NON_PRST_FREQ
 
-    if len(sys.argv) != 6:
+    if len(sys.argv) != 7:
         print("Please provide the following arguments:")
[…]

### 2016-11-08 write-ups/PLDI/paper.tex
Basically there with intro, given to Paul Thomson to read.

--- 2016-10-13 write-ups/PLDI/paper.tex
@@
 
+\newcommand{\offerfork}{\mathsf{offer\_fork}}
+\newcommand{\offerkill}{\mathsf{offer\_kill}}
+
 \begin{document}
 
-\title{Responsive GPU Workgroup Management in the Presence of
-  Persistent Kernels}
+\title{Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs}
 
@@
 \section{Introduction}\label{sec:intro}
+
+\paragraph{The needs of irregular data-parallel algorithms}
+Acceleration of general-purpose computations on graphics processing
+units (GPUs) has tended to focus on \emph{regular} data-parallel
+algorithms, for which work to be processed can be evenly split between
+the workgroups that execute a GPU kernel ahead of time.  However, many
+interesting data-parallel algorithms are \emph{irregular}: the amount
+of work to be processed is unknown ahead of time and may be determined
+by the computation itself.  There is growing interest in accelerating irregular
+data-parallel algorithms on GPUs, particularly algorithms that
+traverse and manipulate graphs, with several recent notable
+successes~\cite{...}.
+
+Irregular algorithms usually require \emph{blocking synchronization}
+between workgroups, to balance load and communicate intermediate
+results.  For example, work stealing algorithms require each workgroup
+to maintain a queue, which is typically protected by a mutex to enable
+stealing by other workgroups.  As another example, breadth-first graph
+traversal uses a frontier-based approach that requires a barrier
+synchronization between all workgroups to achieve consensus that
+exploration of a frontier has been accomplished.  \ADComment{Perhaps
+  make the point that inter-workgroup synchronization is possible in
+  very rare cases without blocking, e.g. threadfence reduction.}
+
+To operate correctly, a blocking concurrent algorithm requires
+\emph{fair} scheduling of threads.  In the context of irregular
+data-parallel algorithms accelerated on a GPU, this translates to
+requiring fair scheduling of workgroups.  Without fairness, a
+workgroup may fail to make progress, leading to starvation.  For
+example, if workgroup $A$ holds a mutex and workgroup $B$ is
+spin-waiting for the mutex to be released, an unfair scheduler can
+forever schedule $B$, so that $A$ never releases the mutex.
+Similarly, an unfair scheduler can induce starvation if workgroup $A$
+is at an inter-workgroup barrier awaiting the arrival of other
+workgroups, by allowing A to spin forever so that the other workgroups
+do not progress to the barrier.
+
+\paragraph{Occupancy-bound execution}
+Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
+and HSA~\cite{...}---specify very few constraints on the manner in
+which workgroups are scheduled.  In particular, they do \emph{not}
+specify that workgroups must be fairly scheduled, and implementations
+of these programming models do not provide fair scheduling in
+practice.  Roughly speaking, each workgroup executing a GPU kernel is
+mapped to a hardware \emph{compute unit}.  If a kernel is executed
+with more workgroups than there are compute units then at least two
+workgroups must share a single compute unit.  The simplest way to
+handle this is via an \emph{occupancy-bound} execution model, whereby
+once a workgroup has commenced execution on a compute unit (is has
+become \emph{occupant}), the workgroup has exclusive access to the
+compute unit until it has finished execution, after which a subsequent
+workgroup can be bound to the compute unit.  Experiments suggest that
+the occupancy-bound execution model is widely implemented by today's
+GPU devices and drivers~\cite{...}, and it is evidently a simple model
+for device vendors to implement efficiently.
+
+The occupancy-bound execution model does not guarantee fair scheduling
+between workgroups: if all compute units are occupied then a
+not-yet-occupant workgroup will not be scheduled until some occupant
+workgroup completes execution.  Yet the execution model \emph{does}
+provide fair scheduling between \emph{occupant} workgroups, because
+they are bound to separate compute units that operate in parallel.
+Current implementations of irregular data parallel algorithms on GPUs
+assume that the target GPU device provides the occupancy-bound
+execution model, which they exploit by scheduling no more workgroups
+than there are available compute units~\cite{...}.  This works today
+because \emph{today}'s GPUs seem to provide the occupancy-bound
+execution model.
+
+\paragraph{Resistance to occupancy-bound execution}
+
+Despite its practical prevalence, none of the current GPU programming
+models actually mandate occupancy-bound execution.  Further, there are
+clear reasons why this model is undesirable from the point-of-view of
+GPU vendors and end users.  First, the execution model does not enable
+multi-tasking, because a workgroup effectively \emph{owns} a compute
+unit until the workgroup has completed execution.  If a long-running
+kernel occupies all workgroups, the GPU cannot be used for other
+tasks, such as graphics rendering, in the meantime.  Precisely because
+of this problem, operating systems employ a GPU \emph{watchdog} to
+heuristically detect and kill long-running computations on the GPU
+being used for display rendering.  If this watchdog is disabled,
+execution of a long-running kernel causes the display to freeze for
+the duration of kernel execution.  Second, \emph{energy throttling} is
+an important concern for battery-powered devices, such as smartphones
+and tablets.  It is desirable for a mobile GPU driver to be able to
+power down one or more compute units if the battery level is low or if
+the user puts the device into a low power state.  Implementing energy
+throttling requires being able to suspend the execution of a
+workgroup, power down the compute unit to which the workgroup was
+bound, and schedule the workgroup for later execution on a different
+compute unit.
+
+Our assessment, informed by discussions with a number of industrial
+practitioners who have been involved in the OpenCL and/or HSA
+standardisation efforts~\cite{CitePersonalCommunication}, is that
+\textbf{(1)} GPU vendors will not commit to the occupancy-bound
+execution model that they currently implement, due to the need for
+multi-tasking and energy throttling, yet \textbf{(2)} GPU vendors will
+not guarantee fair scheduling, due to the runtime overhead associated
+with supporting full preemption of workgroups, e.g.\ based on
+time-slicing.  Vendors instead wish to retain the essence of the
+occupancy-bound execution model, which is simple to implement
+efficiently, and provide preemption for the special cases associated
+with multi-tasking for high-priority jobs (most crucially graphics)
+and energy throttling.  In particular, they do not want to commit to
+fair scheduling of a workgroup once it has been preempted.
+
+\paragraph{Our proposal}
+%
+To summarise: blocking algorithms, such as most irregular algorithms,
+demand a degree of fair scheduling of workgroups, but for good reasons
+GPU vendors will not commit to the guarantees of even the
+occupancy-bound execution model.
+
+We propose \emph{cooperative kernels}, an extension to the GPU
+programming model that aims to resolve this impasse.  A GPU kernel
+that requires workgroups to be fairly scheduled is identified as a
+\emph{cooperative} kernel, and written using two additional language
+primitives, $\offerkill$ and $\offerfork$, which must be placed
+carefully by the programmer.
+
+At a point where the cooperative kernel could proceed with fewer
+workgroups, a workgroup can execute $\offerkill$ to sacrifice itself to
+the scheduler.  This indicates to the scheduler that the workgroup
+would ideally continue with its execution, but that if required the
+scheduler may destroy the workgroup for good: the cooperative kernel
+is prepared to deal with either scenario.
+
+At a point where the cooperative kernel could benefit from having
+additional workgroups join the computation, a workgroup can execute
+$\offerfork$.  This indicates to the scheduler that the cooperative
+kernel is prepared to proceed with the existing set of workgroups, but
+is able to take advantage of having one or more additional workgroups
+commencing execution at the $\offerfork$ program point, inheriting the
+state of the workgroup that executed $\offerfork$.
+
+The use of $\offerfork$ and $\offerkill$ creates a contract between
+the scheduler and (the programmer of the) cooperative kernel, with
+both functional and non-functional requirements.  Functionally: the
+scheduler must guarantee that the workgroups executing a cooperative
+kernel are fairly scheduled; the cooperative kernel must be robust to
+workgroups leaving and joining the computation when the $\offerkill$
+and $\offerfork$ functions, respectively, are called.
+Non-functionally: a cooperative kernel must ensure that $\offerkill$
+is executed frequently enough that the scheduler can accommodate
+soft-real time constraints, e.g.\ allowing a smooth frame-rate for
+graphics or allowing compute units to be powered down quickly when
+required; the scheduler should respond to $\offerkill$ and
+$\offerfork$ calls in a manner that avoids under-utilisation of
+hardware resources by the cooperative kernel, e.g.\ the scheduler
+should not kill workgroups unless they are required for another
+purpose.
+
+Cooperative kernels have several appealing properties.  First, they
+accommodate the needs of blocking algorithms, including irregular
+data-parallel algorithms: such algorithms can be implemented naturally
+as cooperative kernels \ADComment{boast that we did a good set}, so
+that fair scheduling between workgroups is guaranteed.  Second, they
+have no impact on the development of regular (non-cooperative
+kernels): these can be programmed exactly as they are now.  Third, the
+programming model is backwards compatible: a non-blocking algorithm
+implementation that works on today's GPU platforms can be upgraded to
+use our language extensions for cooperative scheduling, and will
+behave exactly as it does at present if the new $\offerkill$ and
+$\offerfork$ functions are simply ignored.  Fourth, the programming
+model is easy to implement on top of the occupancy-bound execution
+model that current GPU platforms provide; we present a
+proof-of-concept implementation on top of OpenCL 2.0 that requires no
+special-purpose hardware or driver support.  Fifth, our results
+suggest that cooperative kernels can provide efficient multi-tasking
+of cooperative and non-cooperative tasks: our experiments across a
+range of GPUs from two vendors show very good response times for
+scheduling short, non-cooperative kernels concurrently with
+long-running cooperative kernels.  These performance results are with
+respect to our proof-of-concept, portable implementation; a vendor
+specific implementation of cooperative kernels could likely yield
+better performance still through device-specific driver and compiler
+support.
+
+In summary, our main contributions are:
+
+\begin{enumerate}
+
+\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms  \ADComment{Maybe big up to say that we give semantics to the new operations.}
+
+\item A portable implementation of cooperative kernels on top of the OpenCL 2.0 programming model
+
+\item The porting of \ADComment{TODO} irregular graph algorithms \ADComment{tighten} to the cooperative kernels model
+
+\item \ADComment{The experiments}
+
+\end{enumerate}
+
+\ADComment{Text to maybe work in.}
+
+In traditional co-operative multi-tasking, a task relinquishes control
+and regains control in a stateful manner. We propose a lighter weight
+model we believe to suitible for the type of blocking algorithms we
+are aware of for GPUs.
+
+The scheduler agrees to scheduler the workgroups fairly, but is at
+liberty to eliminate workgroups for good when they offer themselves
+to be killed. This allows the schduler to repond to changing depands
+on the GPU, e.g. to implement energy throttling.
+
+Prototype implementation of the programming model using OpenCL 2.0
+advance features to build the scheudler and implement a clang based
+compiler to transform input programs using the language extensions to
+lower-level programs that interact with the scheduler. Our prototype
+simulates multiple independent kernels into a mega-kernel. A vendor
+specific implementation could achieve higher efficiency through
+modications to the GPU driver that may take advantage of hardware
+features. (e.g. mega kernel might not be good to compile). Its
+reasonable to think our proof of concept provides a lower bound on how
+well we can do.
+
+We ported a large set of blocking algorithms to our programming
+model. We evaluate co-operative scheduling with a cooperative kernel
+and non-coperative kernel. GPUs from multiple vendors and access
+various things.
+
+Our results are promising.
+
+In the evaluation, make a note that the applications we worked on
+Namely if they use only a barrier or work-stealing idiom, are
+straightforward to adopt to our programming model.
+
+\section{Motivating example: work stealing}
+
+\ADComment{Use the work stealing example to both introduce OpenCL and
+  also make our key point: that arbitrary preemption will not work.}
+
+\section{The ckill and cfork primitives}
+
+\ADComment{Remind the reader that these are inspired by work stealing
+  and barrier synchonization.  In \mysec\ref{ } we describe the
+  semantics of kernels that use these primitives.  In \mysec\ref{ } we
+  show how work stealing can be implemented.  In \mysec\ref{ } we show
+  how a barrier can be implemented.}
+
+\subsection{Semantics of the new primitives}
+
+\ADComment{General semantics: let $N$ be the maximum number of
+  workgroups the GPU of interest can support [gloss over or deal with
+    issue that this is kernel-dependent].  A kernel is launched with
+  $M$ workgroups, for some $0 < M \leq N$.  Initially these are all
+  scheduled, and are guaranteed to make independent forward progress.}
+
+\ADComment{The purpose of ckill is for a workgroup to volunteer to
+  quit execution of the kernel if the resources that it is using
+  (e.g.\ the associated compute unit) are required for other tasks.  A
+  ckill statement must be reached under workgroup-uniform control
+  flow.  When a workgroup reaches ckill, the workgroup either
+  terminates execution (because the environment requires the
+  workgroup's resources), or continues execution as if the ckill
+  statement were a no=op).  From a semantics perspective this is a
+  non-deterministic choice: an application that invokes ckill should
+  be prepared for either outcome.  In practice, the effect of ckill
+  depends on environmental conditions.}
+
+\ADComment{The purpose of cfork is for a workgroup to indicate that it
+  would be useful to have additional workgroups join the computation,
+  and to specify that the program point associated with the cfork
+  statement would be a suitable place for a new workgroup to start
+  executing.  A cfork statement must also be reached under
+  workgroup-uniform control flow.  If $C$ workgroups are executing
+  when a workgroup reaches a cfork statement, with $0 < C \leq M$,
+  then between 0 and $M-C$ new workgroups commence execution from the
+  following statement.  From a semantics perspective, the number of
+  workgroups that join the computation is nondeterministic, within the
+  permitted range.  In practice, how many workgroups join depends on
+  environmental conditions.}
+
+\ADComment{TODO: describe persistent variables.}
+
+\ADComment{TODO: explain that ckill and cfork atomically update the
+  kernel context (ids).}
+
+\ADComment{TODO: somewhere we need to explain the constraints this
+  places on programming: how ids work, what one can assume about them,
+  etc.}
+
+\ADComment{Now describe the more refined semantics where IDs are
+  relinquished in decreasing order; explain that from a formal
+  perspective thie refines the original semantics.}
+
+\ADComment{In the above, explain at some point how a scheduler can
+  exploit cfork and ckill to utilise the GPU resources.}
+
+\subsection{Preemption-aware work stealing}
+
+\ADComment{Show how work stealing can now be made safe.}
+
+\subsection{A preemption-aware global barrier}
+
+\ADComment{Show how the global barrier can be written in a manner that
+  maintains contiguous ids using the refined semantics.}
+
+\section{Application-specific optimisations}
+
+\ADComment{Explain that for global barriers, we explain how some new
+  primitives can lead to a better implementation.  Explain that for
+  work stealing there are various ways the existing ckill and cfork
+  primitives can be used.}
+
+\subsection{An optimised barrier implementation}
+
+\ADComment{Remark that the barrier of \mysec\ref{sec:naivebarrier} may
+  be inefficient, discuss the use of query primitives to make this
+  more efficient.}
+
+\subsection{Variations on work stealing}
+
+\ADComment{Three cases here: always ckill/cfork, ckill/cfork only when
+  queue is empty, and have a ``mediator'' workgroup to manage things.}
+
+\section{A mega kernel-based prototype}
+
+\ADComment{Explain carefully that doing this properly would require
+  driver support, etc., and then explain what we've actually done.}
+
+\section{Experimental evaluation}
+
+\ADComment{TODO: write up notes from whiteboard chat.}
+
+\section{Related work}
+
+\section{Conclusions and future work}
+
+\ADComment{Next-generation GPUs and drivers for evaluation,
+  implementation in a real driver, migration of work between multiple
+  GPUs.}
+
+\section{Old intro}
 
@@
 
-\section{Motivating example: work stealing}
-
-\ADComment{Use the work stealing example to both introduce OpenCL and
-  also make our key point: that arbitrary preemption will not work.}
-
-\section{The ckill and cfork primitives}
-
-\ADComment{Remind the reader that these are inspired by work stealing
-  and barrier synchonization.  In \mysec\ref{ } we describe the
-  semantics of kernels that use these primitives.  In \mysec\ref{ } we
-  show how work stealing can be implemented.  In \mysec\ref{ } we show
-  how a barrier can be implemented.}
-
-\subsection{Semantics of the new primitives}
-
-\ADComment{General semantics: let $N$ be the maximum number of
-  workgroups the GPU of interest can support [gloss over or deal with
-    issue that this is kernel-dependent].  A kernel is launched with
-  $M$ workgroups, for some $0 < M \leq N$.  Initially these are all
-  scheduled, and are guaranteed to make independent forward progress.}
-
-\ADComment{The purpose of ckill is for a workgroup to volunteer to
-  quit execution of the kernel if the resources that it is using
-  (e.g.\ the associated compute unit) are required for other tasks.  A
-  ckill statement must be reached under workgroup-uniform control
-  flow.  When a workgroup reaches ckill, the workgroup either
-  terminates execution (because the environment requires the
-  workgroup's resources), or continues execution as if the ckill
-  statement were a no=op).  From a semantics perspective this is a
-  non-deterministic choice: an application that invokes ckill should
-  be prepared for either outcome.  In practice, the effect of ckill
-  depends on environmental conditions.}
-
-\ADComment{The purpose of cfork is for a workgroup to indicate that it
-  would be useful to have additional workgroups join the computation,
-  and to specify that the program point associated with the cfork
-  statement would be a suitable place for a new workgroup to start
-  executing.  A cfork statement must also be reached under
-  workgroup-uniform control flow.  If $C$ workgroups are executing
-  when a workgroup reaches a cfork statement, with $0 < C \leq M$,
-  then between 0 and $M-C$ new workgroups commence execution from the
-  following statement.  From a semantics perspective, the number of
-  workgroups that join the computation is nondeterministic, within the
-  permitted range.  In practice, how many workgroups join depends on
-  environmental conditions.}
-
-\ADComment{TODO: describe persistent variables.}
-
-\ADComment{TODO: explain that ckill and cfork atomically update the
-  kernel context (ids).}
-
-\ADComment{TODO: somewhere we need to explain the constraints this
-  places on programming: how ids work, what one can assume about them,
-  etc.}
-
-\ADComment{Now describe the more refined semantics where IDs are
-  relinquished in decreasing order; explain that from a formal
-  perspective thie refines the original semantics.}
-
-\ADComment{In the above, explain at some point how a scheduler can
-  exploit cfork and ckill to utilise the GPU resources.}
-
-\subsection{Preemption-aware work stealing}
-
-\ADComment{Show how work stealing can now be made safe.}
-
-\subsection{A preemption-aware global barrier}
-
-\ADComment{Show how the global barrier can be written in a manner that
-  maintains contiguous ids using the refined semantics.}
-
-\section{Application-specific optimisations}
-
-\ADComment{Explain that for global barriers, we explain how some new
-  primitives can lead to a better implementation.  Explain that for
-  work stealing there are various ways the existing ckill and cfork
-  primitives can be used.}
-
-\subsection{An optimised barrier implementation}
-
-\ADComment{Remark that the barrier of \mysec\ref{sec:naivebarrier} may
-  be inefficient, discuss the use of query primitives to make this
-  more efficient.}
-
-\subsection{Variations on work stealing}
-
-\ADComment{Three cases here: always ckill/cfork, ckill/cfork only when
-  queue is empty, and have a ``mediator'' workgroup to manage things.}
-
-\section{A mega kernel-based prototype}
-
-\ADComment{Explain carefully that doing this properly would require
-  driver support, etc., and then explain what we've actually done.}
-
-\section{Experimental evaluation}
-
-\ADComment{TODO: write up notes from whiteboard chat.}
-
-\section{Related work}
-
-\section{Conclusions and future work}
-
-\ADComment{Next-generation GPUs and drivers for evaluation,
-  implementation in a real driver, migration of work between multiple
-  GPUs.}
-
 \end{document}

### 2016-11-09 write-ups/PLDI/paper.tex

@@
 
-\section{The ckill and cfork primitives}
+\section{The Cooperative Kernels Programming Model}
+
+We describe our cooperative kernels programming model as an extension
+to the industry-standard OpenCL programming model for heterogenous
+many-core architectures.  However, the concept behind our programming
+model is more general, and could be applied to extend other GPU
+programming models, such as CUDA and HSA.
+
+\subsection{Semantics of Cooperative Kernels}
+
+As with a regular OpenCL kernel, a cooperative kernel is launched by
+the host application, with a specified number of workgroups, each
+consisting of a specified number of threads.  The kernel's behaviour
+is described by a kernel function, written in OpenCL C but using the
+additional language extensions that we present below, and a number of
+parameters to the kernel function are passed by the host application.
+
+\paragraph{Restored variables}
+
+\ADComment{TODO; perhaps say that kernel parameters are immutable.}
+
+\paragraph{Initial workroups}
+
+If the host application launches a cooperative kernel specifying that
+$N$ workgroups should execute the kernel, then when the kernel
+commences execution it should be executed by some number of workgroups
+$M$, with $1\leq M \leq N$.  The requested number of workgroups is
+called the \emph{workgroup limit} for the cooperative kernel, and $M$
+is the number of \emph{active workgroups}.
+
+The initial workgroups have consecutive workgroup ids in the range
+$[0, M-1]$.
+
+\paragraph{Semantics for $\offerkill$}
+
+Our first additional language primitive, $\offerkill$, is a
+workgroup-level function.  That is, similar to the OpenCL
+intra-workgroup $\mathsf{barrier}$ primitive, if $\offerkill$ appears
+inside conditional code then it must be executed by all threads in a
+workgroup, with all threads having executed the same number of
+iterations of any enclosing loops.
+
+Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$,
+i.e.\ the workgroup's id is not the largest among the active
+workgroups, then $\offerkill$ is a no-op.  Otherwise, we must have $m
+= M-1$, in which case there is a non-deterministic choice between (a)
+$\offerkill$ executing as a no-op, or (b) execution of the workgroup
+ceasing and the number of active workgroups $M$ being decremented by
+one.  Case (b) corresponds to the scheduler accepting the workgroup's
+offer to cease execution, allowing the scheduler to schedule some
+other task.
+
+The semantics of $\offerkill$ is atomic.  That is, if multiple
+workgroups reach $\offerkill$ simultaneously, the semantics is that
+the workgroups should each execute $\offerkill$ in some well-defined
+but arbitrary sequential order.  \ADComment{Should we say something
+  about linearisability?}
+
+\paragraph{Semantics for $\offerfork$}
+
+Our second language primitive, $\offerfork$, is also a workgroup-level
+function.
+
+Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  An
+integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
+$k$ new workgroups are spawned with consecutive ids in the range $[M,
+  M+k-1]$, and the active workgroup count $M$ is incremented by $k$.
+The $k$ new workgroups commence execution at the program point
+immediately following the $\offerfork$ call.  Each thread in each new
+workgroup has an arbitrary local state, except that:
+
+\begin{itemize}
+
+\item the threads have access to the parameters originally passed to
+  the cooperative kernel by the host application (recall that these
+  are immutable);
+
+\item for each variable $v$ in the set of restored variables
+  identified for the cooperative kernel, every thread's copy of $v$ is
+  initialised to the value that thread 0 in workgroup $m$ held for $v$
+  at the point of the $\offerfork$ call.
+
+\end{itemize}
+
+Notice that $k=0$ is always a legitimate choice for the number of
+workgroups to be spawned by an $\offerfork$ call, and in particular
+that if the number of active workgrouups $M$ is equal to the workgroup
+limit $N$, $k=0$ is guaranteed.
+
+\ADComment{Remark on our decision to have every thread inherit the state of the master thread from workgroup $m$.}
+
+\subsection{Non-functional Requirements of Cooperative Kernels}
+
+If a cooperative kernel is launched with $M$ workgroups, the
+
+However, unlike a regular kernel, if $M$ workgroups are requested for execution of a cooperative kernel.
+
+A \emph{cooperative kernel} is launched with a desired workgroup
+count, $M$, where $M > 0$.
 

### 2016-11-09 write-ups/PLDI/paper.tex
Progress.

--- 2016-11-08 write-ups/PLDI/paper.tex
@@
 \newcommand{\offerkill}{\mathsf{offer\_kill}}
+\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
+\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
+\newcommand{\getgroupid}[1]{\mathsf{get\_group\_id(#1)}}
+\newcommand{\getnumgroups}[1]{\mathsf{get\_num\_groups(#1)}}
 
@@
 
-\section{The ckill and cfork primitives}
+\section{The Cooperative Kernels Programming Model}
+
+We describe our cooperative kernels programming model as an extension
+to the industry-standard OpenCL programming model for heterogenous
+many-core architectures.  However, the cooperative kernels concept is
+more general, and could be applied to extend other GPU programming
+models, such as CUDA and HSA.  \ADComment{However, unclear that we could implement it on top of these, but that's not really the point.}
+
+\subsection{Semantics of Cooperative Kernels}
+
+As with a regular OpenCL kernel, a cooperative kernel is launched by
+the host application, with a specified number of workgroups, each
+consisting of a specified number of threads.  The kernel's behaviour
+is described by a kernel function, written in OpenCL C but using the
+additional language extensions that we present below, and a number of
+parameters to the kernel function are passed by the host application.
+
+The parameters to a cooperative kernel are \emph{immutable}, as if
+they were qualified with the \texttt{const} keyword (though pointer
+parameters can refer to mutable data).
+
+A regular OpenCL kernel can be executed by a multi-dimensional grid of
+workgroups, called an ND-range.  In our current proposal, the ND-range
+for a cooperative kernel must be one-dimensional.  \ADComment{Say why?
+  And mention that none of the applications that we deem would benefit
+  from cooperative kernels need a multi-dimensional ND-range.}
+
+As in regular OpenCL, each workgroup can be multi-dimensional.
+
+\paragraph{Restored variables}
+
+A variable declared in the root scope of the cooperative kernel can
+optionally be annotated with a new \texttt{restore} qualifier.  By
+annotating a variable with \texttt{restore}, the programmer indicates
+that when new workgroups are spawned via $\offerkill$, the threads in
+these workgroups should obtain a well-defined initial value for the
+variable, obtained from the workgroup that executed $\offerkill$.  We
+detail the semantics for this when we describe $\offerkill$ below.
+
+Only private memory variables can be annotated with \texttt{restore},
+which may include pointers to local or global memory so long as the
+pointer variables themselves are stored in private memory.  In
+particular, variables and arrays in local memory cannot be annotated
+with \texttt{restore}.  \ADComment{The preceding is what we have
+  implemented, but actually it is pretty clear how we could handle
+  local memory restoration if we wanted to; it's just that none of our
+  examples needed it.}
+
+\paragraph{Active workroups}
+
+If the host application launches a cooperative kernel specifying that
+$N$ workgroups should execute the kernel, then when the kernel
+commences execution it should be executed by some number of workgroups
+$M$, with $1\leq M \leq N$.  The requested number of workgroups is
+called the \emph{workgroup limit} for the cooperative kernel, and $M$
+is the number of \emph{active workgroups}.  As explained below, the
+number $M$ of active workgroups may change during the lifetime of a
+cooperative kernel.
+
+Fair scheduling is guaranteed between active workgroups.  That is, if
+some thread in an active workgroup is enabled then eventually some
+thread in the active workgroup is guaranteed to execute an
+instruction.  Note that our programming model does not mandate that
+threads within a workgroup must be fairly scheduled, and indeed this
+is often not the case in GPU programming models, because threads are
+often organised into \emph{warps} or \emph{wavefronts} within a
+workgroup, exhibiting lock-step predicated execution that is
+inherently unfair.
+
+The initial workgroups have consecutive workgroup ids in the range
+$[0, M-1]$.
+
+The OpenCL primitive $\getgroupid{0}$ retrieves the id of a
+workgroup as usual.  When executed by a cooperative kernel, the OpenCL
+primitive $\getnumgroups{0}$ returns $M$, the current number
+of active workgroups.  Because the ND-ranges of cooperative kernels
+are one-dimensional, $\getgroupid{\mathit{i}}$ and
+$\getnumgroups{\mathit{i}}$ return 0 and 1, respectively, for $i >
+0$.
+
+\ADComment{What happens to $M$ when a workgroup reaches the end of a
+  cooperative kernel?  Feels like $M$ should decrease, but what does
+  this mean for barrier-only kernels?  Do they need to account for the
+  number of active workgroups changing during the final barrier
+  interval?  An alternative is to say that there is an implicit global barrier at kernel exit.}
+
+\paragraph{Semantics for $\offerkill$}
+
+Our first additional language primitive, $\offerkill$, is a
+workgroup-level function: similar to the OpenCL intra-workgroup
+$\mathsf{barrier}$ primitive, if $\offerkill$ appears inside
+conditional code then it must be executed by \emph{all} threads in a
+workgroup, with all threads having executed the same number of
+iterations of any enclosing loops.
+
+Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$,
+i.e.\ the workgroup's id is not the largest among the active
+workgroups, then $\offerkill$ is a no-op.  If instead $m = M-1$, there
+is a non-deterministic choice between (a) $\offerkill$ executing as a
+no-op, or (b) execution of the workgroup ceasing and the number of
+active workgroups $M$ being decremented by one.  Case (b) corresponds
+to the scheduler accepting the workgroup's offer to cease execution,
+allowing the scheduler to schedule some other task on the compute unit
+on which the workgroup was executing.
+
+The semantics of $\offerkill$ is atomic.  That is, if multiple
+workgroups reach $\offerkill$ simultaneously, the semantics is as if
+each workgroup executes $\offerkill$ in some arbitrary sequential
+order.  \ADComment{Should we say something about linearisability?}
+
+\paragraph{Semantics for $\offerfork$}
+
+Our second new language primitive, $\offerfork$, is also a workgroup-level
+function.
+
+Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  An
+integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
+$k$ new workgroups are spawned with consecutive ids in the range
+$[M, M+k-1]$, and the active workgroup count $M$ is incremented by
+$k$.
+
+The $k$ new workgroups commence execution at the program point
+immediately following the $\offerfork$ call.  The private variables of
+each thread in a new workgroup are all uninitialised, as are any local
+variables and arrays shared across the workgroup; reading from these
+locations without initialising them is an undefined behaviour.  There
+are two exceptions to this:
+
+\begin{itemize}
+
+\item because the parameters to a cooperative kernel are immutable,
+  the new threads have access to these parameters and can safely read
+  from them;
+
+\item for each variable $v$ annoated with \texttt{restore}, every new
+  thread's copy of $v$ is initialised to the value that thread 0 in
+  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.
+
+\end{itemize}
+
+The second exception, for restored variables, can be stated more
+formally as follows.  Let $\mathit{WG}[x][y].v$ denote the value
+stored in variable $v$ of thread $y$ in workgroup $x$.  Then for each
+new workgroup $i \in [M, M+k-1]$ and each thread $j$ in workgroup $i$,
+$\mathit{WG}[i][j].v$ is initialised to $\mathit{WG}[m][0].v$ for each
+variable $v$ annotated with \texttt{restore}.
+
+Notice that $k=0$ is always a legitimate choice for the number of
+workgroups to be spawned by an $\offerfork$ call, and if the number of
+active workgrouups $M$ is equal to the workgroup limit $N$, $k=0$ is
+guaranteed.
+
+\ADComment{Remark on our decision to have every thread inherit the
+  state of the master thread from workgroup $m$.}
+
+\paragraph{Global barriers}
+
+Many irregular algorithms require synchronisation across workgroups.
+This can be achieved via a \emph{global barrier}, but a global barrier
+is not provided as a primitive in OpenCL, CUDA or HSA.  This is
+because, as discussed in Section~\ref{ }, an inter-workgroup barrier
+requires fair scheduling of workgroups.  As a result, past works have
+either split computation into multiple separate kernels, so that the
+end of one kernel and the start of the next kernel implicitly provides
+a global barrier~\cite{ }, or have assumed the occupancy-bound
+execution model to build a global barrier that works reliably on
+today's GPU platforms.
+
+Because workgroups of a cooperative kernel are fairly scheduled, a
+global barrier primitive can be provided.  We specify two such primitives: $\globalbarrier()$
+and $\resizingglobalbarrier()$.
+
+A $\globalbarrier()$ is a kernel-level function: if it appears in
+conditional code then it must be reached by \emph{all} threads
+executing the cooperative kernel.  On reaching a $\globalbarrier()$, a
+thread waits until all threads have arrived at the barrier.  Once all
+threads have arrived, the threads may proceed past the barrier with
+the guarantee that all global memory accesses issued before the
+barrier have completed.
+
+A $\resizingglobalbarrier()$ is identical to a global barrier, except
+that, semantically, every workgroup executes $\offerkill$ on reaching
+the barrier, and every workgroup executes $\offerfork$ on leaving the
+barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent to:
+%
+\lstset{basicstyle=\tt}
+\begin{lstlisting}
+  $\offerkill()$;
+  $\globalbarrier()$;
+  $\offerfork()$;
+\end{lstlisting}
+\lstset{basicstyle=\scriptsize\tt}
+
+The $\globalbarrier()$ primitive can be implemented directly using the
+atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
+but the implementation is involved, hence why we include this function
+as a primitive.  As noted above, the $\resizingglobalbarrier()$
+primitive is, semantically, synactic sugar for a sequence of other
+primitive calls.  We include it as a primitive because (a) we have
+found that global barriers often provide a natural point for
+cooperation with the scheduler, and (b) identifying a known point at
+which workgroups are required to synchronize and at which workgroups
+can be safely killed and forked enables a custom implementation of
+$\resizingglobalbarrier()$ that provides the semantics described
+above, but is significantly more efficient than the above sequence of
+calls.  We discuss our efficient implementation of
+$\resizingglobalbarrier$ in detail in Section~\ref{...}.
+
+\subsection{Backwards compatibility}
+
+\ADComment{Perhaps this would fit better under ``programming guidelines'' -- not sure.}
+
+\ADComment{Make the point here that if an irregular algorithm is
+  written as a cooperative kernel then it will execute just as safely
+  as it does today on current hardware if: (1) the \texttt{restore}
+  annotation and $\offerkill$/$\offerfork$ calls are defined away to
+  nothing, (2) $\resizingglobalbarrier$ is redefined to
+  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
+  that assumes the occupancy-bound execution model is provided.
+If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}
+
+\subsection{Programming guidelines}
+
+An OpenCL programmer writing a cooperative kernel should pay careful
+attention to the following details.
+
+\paragraph{Keeping at least one workgroup alive}
+
+The programer is responsible for keeping at least one workgroup alive.
+In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.
+This can be guarded against by wrapping each $\offerkill$ call in a conditional statement:
+
+\lstset{basicstyle=\tt}
+\begin{lstlisting}
+  if($\getgroupid{0}$ != 0) $\offerkill$;
+\end{lstlisting}
+\lstset{basicstyle=\scriptsize\tt}
+
+We could, alternatively, have formulated the semantics of $\offerkill$
+so that workgroup 0 is never killed.  However, there might be a use
+case for a cooperative kernel reaching a point where it would be
+acceptable for the kernel to exit, although desirable for some
+remaining computation to be performed if competing workloads allow it.
+
+\paragraph{A changing number of workgroups}
+
+In regular OpenCL, $\getnumgroups{0}$ returns a runtime-constant
+value: once the number of executing workgroups has been identified, it
+is guaranteed not to change.
+
+In contrast, the number of workgroups executing a cooperative kernel
+can vary due to workgroups executing $\offerkill$, $\offerfork$ and
+$\resizingglobalbarrier$.  The programmer must therefore write their
+cooperative kernel in a manner that is robust to changes in the value
+returned by $\getnumgroups{0}$.
+
+The main impact of this relates to partitioning of data to be
+processed.
+
+\ADComment{Tyler, can we give an example of a data processing loop
+  that would have to be changed?}
+
+\paragraph{Cooperative kernels that only use barriers}
+
+\ADComment{Need to account for the issue that $M$ might decrease as
+  workgroups actually exit the kernel.}
+
+If a cooperative kernel does not directly issue $\offerkill$ and
+$\offerfork$ calls then the only possibilty for the number of
+workgroups to change is due to $\resizingglobalbarrier$ being invoked.
+
+In this case, because the resizing barrier is a kernel-level function,
+the statements executed by the cooperative kernel can be divided into
+\emph{resizing barrier intervals}: a resizing barrier interval starts
+at a resizing barrier or the kernel entry point, and ends at a
+resizing barrier or a kernel exit point.  At any moment of execution,
+all threads are guaranteed to be executing within the same resizing
+barrier interval.  As a result, the number of workgroups executing the
+kernel is guaratneed to remain unchanged within each resizing barrier
+interval, so the result returned by $\getnumgroups{0}$ is guaratneed
+to be identical on repeated calls during such an interval.  This
+guarantee may be exploited for efficiently when writing a cooperative
+kernel that uses global barriers but does not call $\offerfork$ and
+$\offerkill$ directly.
+
+\subsection{Non-functional Requirements of Cooperative Kernels}
+
+The semantics presented in Section~\ref{...} describe the envelope of
+behaviour that a developer of a cooperative kernel should be prepared
+for.
+
+However, the aim of cooperative kernels is to find a balance that
+allows \emph{efficient execution} of long-running kernels
+(e.g.\ implementing irregular data-parallel algorithms), and
+\emph{responsive multi-tasking} so that the GPU can be shared between
+one or more cooperative kernels and a number of short, high-priority
+tasks with soft real-time constraints.
+
+To achieve this balance, both an implementation of the ooperative
+kernels model, and the programmer of a cooperative kernel, must strive
+to meet a number of non-functional requirements, which we outline
+here.
+
+\paragraph{Sufficient $\offerkill$ calls}
+
+Recall that the purpose of $\offerkill$ is to provide the scheduler
+with an opportunity to destroy a workgroup in order to schedule
+higher-priority tasks.  Because these higher-priority tasks, such as
+graphics rendering, may have soft real-time constraints, the scheduler
+relies on the cooperative kernel to execute $\offerkill$ sufficiently
+frequently that the scheduler can meet such real-time constraints.
+
+\ADComment{Got here.}
+
+\subsection{Preemption-aware work stealing}
 
@@
   how a barrier can be implemented.}
-
-\subsection{Semantics of the new primitives}
-
-\ADComment{General semantics: let $N$ be the maximum number of
-  workgroups the GPU of interest can support [gloss over or deal with
-    issue that this is kernel-dependent].  A kernel is launched with
-  $M$ workgroups, for some $0 < M \leq N$.  Initially these are all
-  scheduled, and are guaranteed to make independent forward progress.}
-
-\ADComment{The purpose of ckill is for a workgroup to volunteer to
-  quit execution of the kernel if the resources that it is using
-  (e.g.\ the associated compute unit) are required for other tasks.  A
-  ckill statement must be reached under workgroup-uniform control
-  flow.  When a workgroup reaches ckill, the workgroup either
-  terminates execution (because the environment requires the
-  workgroup's resources), or continues execution as if the ckill
-  statement were a no=op).  From a semantics perspective this is a
-  non-deterministic choice: an application that invokes ckill should
-  be prepared for either outcome.  In practice, the effect of ckill
-  depends on environmental conditions.}
-
-\ADComment{The purpose of cfork is for a workgroup to indicate that it
-  would be useful to have additional workgroups join the computation,
-  and to specify that the program point associated with the cfork
-  statement would be a suitable place for a new workgroup to start
-  executing.  A cfork statement must also be reached under
-  workgroup-uniform control flow.  If $C$ workgroups are executing
-  when a workgroup reaches a cfork statement, with $0 < C \leq M$,
-  then between 0 and $M-C$ new workgroups commence execution from the
-  following statement.  From a semantics perspective, the number of
-  workgroups that join the computation is nondeterministic, within the
-  permitted range.  In practice, how many workgroups join depends on
-  environmental conditions.}
-
-\ADComment{TODO: describe persistent variables.}
-
-\ADComment{TODO: explain that ckill and cfork atomically update the
-  kernel context (ids).}
-
-\ADComment{TODO: somewhere we need to explain the constraints this
-  places on programming: how ids work, what one can assume about them,
-  etc.}
-
-\ADComment{Now describe the more refined semantics where IDs are
-  relinquished in decreasing order; explain that from a formal
-  perspective thie refines the original semantics.}
-
-\ADComment{In the above, explain at some point how a scheduler can
-  exploit cfork and ckill to utilise the GPU resources.}
-
-\subsection{Preemption-aware work stealing}
 

### 2016-11-09 src/test_suite/connect_four/common/connect_four_data.h
Add host code interface and update device code to be able to use kernel merge

@@ -18,7 +18,6 @@ const CL_INT_TYPE MINUS_INF = -666;
 /* The actual level of lookahead is controlled by the "maxlevel" flag,
  * MAX_LOOKAHEAD is a hard-coded upper-bound. */
 const CL_INT_TYPE MAX_LOOKAHEAD = 7;
-const CL_INT_TYPE NUM_NODE = 20000;
 
 /*---------------------------------------------------------------------------*/
 

### 2016-11-09 src/test_suite/connect_four/connect_four.h

@@ -1,0 +1,356 @@
+#include "connect_four_data.h"
+#include <math.h>
+
+/*---------------------------------------------------------------------------*/
+
+DEFINE_string(restoration_ctx_path, "test_suite/connect_four", "Path to restoration context");
+DEFINE_string(merged_kernel_file, "test_suite/connect_four/device/merged.cl", "path to the merged mega kernel file");
+DEFINE_string(persistent_kernel_file, "test_suite/connect_four/device/standalone.cl", "path to the standalone mega kernel file");
+
+/*---------------------------------------------------------------------------*/
+
+DEFINE_int32(pools, 10, "Number of task pools");
+// Pool size is quite arbitrary, may be tuned
+DEFINE_int32(pool_size, 20, "Size of a task pool");
+DEFINE_int32(maxlevel, 2, "Max level of look-ahead");
+
+DEFINE_string(board_file, "test_suite/connect_four/board.txt", "Path to board description");
+
+/*---------------------------------------------------------------------------*/
+// global vars
+
+cl_uchar *h_board;
+cl::Buffer d_board;
+Node *h_nodes;
+cl::Buffer d_nodes;
+cl::Buffer d_node_head;
+cl_int num_node;
+cl::Buffer d_task_pool;
+Task *h_task_pool;
+cl::Buffer d_task_pool_lock;
+cl::Buffer d_task_pool_head;
+cl::Buffer d_next_move_value;
+cl::Buffer d_root_done;
+cl::Buffer d_debug_int;
+cl::Buffer d_debug_board;
+size_t board_mem_size;
+
+/*===========================================================================*/
+// helper functions
+
+void print_board(cl_uchar *board)
+{
+  printf(" - - - - - - -\n");
+
+  for (int i = 0; i < NUM_ROW; i++) {
+    for (int j = 0; j < NUM_COL; j++) {
+      switch (board[(i * 7) + j]) {
+      case EMPTY:
+        printf(" .");
+        break;
+      case HUMAN:
+        printf(" o");
+        break;
+      case COMPUTER:
+        printf(" x");
+        break;
+      default:
+        printf(" ?");
+        break;
+      }
+    }
+    printf("\n");
+  }
+
[…]

### 2016-11-09 write-ups/PLDI/paper.tex
Revisions discussed with Sreepathi.

@@
 
+\newcommand{\restore}{\texttt{restore}}
+
 \newcommand{\code}[1]{\lstset{basicstyle=\tt}\lstinline!#1!\lstset{basicstyle=\scriptsize\tt}}
@@
 
+\ADComment{Think about whether ``restore'' is best name.}
+
+\ADComment{Address issue of placing constructs manually --
+  transactional semantics and barriers.  Argue why it's easy to do this.}
+
+\ADComment{Mention dynamic parallelism ane explain why it's no good.}
+
 \paragraph{The needs of irregular data-parallel algorithms}
@@
 data-parallel algorithms on GPUs, particularly algorithms that
-traverse and manipulate graphs, with several recent notable
+traverse and manipulate linked data structures, with several recent notable
 successes~\cite{...}.
@@
 between workgroups, to balance load and communicate intermediate
-results.  For example, work stealing algorithms require each workgroup
+results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
+As another example, work stealing algorithms require each workgroup
 to maintain a queue, which is typically protected by a mutex to enable
-stealing by other workgroups.  As another example, breadth-first graph
-traversal uses a frontier-based approach that requires a barrier
-synchronization between all workgroups to achieve consensus that
-exploration of a frontier has been accomplished.  \ADComment{Perhaps
+stealing by other workgroups.  \ADComment{Perhaps
   make the point that inter-workgroup synchronization is possible in
@@
 data-parallel algorithms accelerated on a GPU, this translates to
-requiring fair scheduling of workgroups.  Without fairness, a
+requiring fair scheduling of workgroups.  \ADComment{Shorten} Without fairness, a
 workgroup may fail to make progress, leading to starvation.  For
@@
 \paragraph{Occupancy-bound execution}
+\ADComment{Add footnote to acknowledge multiple wgs per cu}
+\ADComment{Say that current GPU schedulers are not required to be fair.}
 Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
@@
 handle this is via an \emph{occupancy-bound} execution model, whereby
-once a workgroup has commenced execution on a compute unit (is has
+once a workgroup has commenced execution on a compute unit (it has
 become \emph{occupant}), the workgroup has exclusive access to the
@@
 execution model, which they exploit by scheduling no more workgroups
-than there are available compute units~\cite{...}.  This works today
+than there are available compute units~\cite{...}.  This works
 because \emph{today}'s GPUs seem to provide the occupancy-bound
@@
 
-\paragraph{Resistance to occupancy-bound execution}
+\paragraph{Resistance to guaranteeing occupancy-bound execution}
 
@@
 unit until the workgroup has completed execution.  If a long-running
-kernel occupies all workgroups, the GPU cannot be used for other
-tasks, such as graphics rendering, in the meantime.  Precisely because
+kernel occupies all compute units, the GPU cannot be used for other
+tasks, such as graphics rendering, in the meantime.  \ADComment{Could cut or move to later:} Precisely because
 of this problem, operating systems employ a GPU \emph{watchdog} to
@@
 an important concern for battery-powered devices, such as smartphones
-and tablets.  It is desirable for a mobile GPU driver to be able to
+and tablets.  In the future, it will be desirable for a mobile GPU driver to be able to
 power down one or more compute units if the battery level is low or if
 the user puts the device into a low power state.  Implementing energy
-throttling requires being able to suspend the execution of a
+throttling would require being able to suspend the execution of a
 workgroup, power down the compute unit to which the workgroup was
@@
 not guarantee fair scheduling, due to the runtime overhead associated
-with supporting full preemption of workgroups, e.g.\ based on
-time-slicing.  Vendors instead wish to retain the essence of the
+with supporting full preemption of workgroups~\cite{ISCAPAPER}.  Vendors instead wish to retain the essence of the
 occupancy-bound execution model, which is simple to implement
-efficiently, and provide preemption for the special cases associated
+efficiently, and provide preemption for \ADComment{shorten} the special cases associated
 with multi-tasking for high-priority jobs (most crucially graphics)
@@
 fair scheduling of a workgroup once it has been preempted.
+
+\ADComment{Kepler preemption support?}
 
@@
 is able to take advantage of having one or more additional workgroups
-commencing execution at the $\offerfork$ program point, inheriting the
-state of the workgroup that executed $\offerfork$.
+commencing execution at the $\offerfork$ program point.
 
@@
 the scheduler and (the programmer of the) cooperative kernel, with
-both functional and non-functional requirements.  Functionally: the
+both functional and non-functional requirements.  Functionally, the
 scheduler must guarantee that the workgroups executing a cooperative
-kernel are fairly scheduled; the cooperative kernel must be robust to
-workgroups leaving and joining the computation when the $\offerkill$
-and $\offerfork$ functions, respectively, are called.
-Non-functionally: a cooperative kernel must ensure that $\offerkill$
+kernel are fairly scheduled, while the cooperative kernel must be robust to
+workgroups leaving and joining the computation in response to $\offerkill$
+and $\offerfork$.
+Non-functionally, a cooperative kernel must ensure that $\offerkill$
 is executed frequently enough that the scheduler can accommodate
@@
 graphics or allowing compute units to be powered down quickly when
-required; the scheduler should respond to $\offerkill$ and
+required.  The scheduler should respond to $\offerkill$ and
 $\offerfork$ calls in a manner that avoids under-utilisation of
@@
 
-Cooperative kernels have several appealing properties.  First, they
-accommodate the needs of blocking algorithms, including irregular
-data-parallel algorithms: such algorithms can be implemented naturally
+Cooperative kernels have several appealing properties:
+
+\begin{enumerate}
+
+\item They
+accommodate the needs of the irregular
+data-parallel algorithms that have been successfully accelerated using GPUs: such algorithms can be implemented naturally
 as cooperative kernels \ADComment{boast that we did a good set}, so
-that fair scheduling between workgroups is guaranteed.  Second, they
+that fair scheduling between workgroups is guaranteed.
+
+\item They
 have no impact on the development of regular (non-cooperative
-kernels): these can be programmed exactly as they are now.  Third, the
+kernels): these can be programmed exactly as they are now.
+
+\item The
 programming model is backwards compatible: a non-blocking algorithm
@@
 behave exactly as it does at present if the new $\offerkill$ and
-$\offerfork$ functions are simply ignored.  Fourth, the programming
+$\offerfork$ functions are simply ignored.
+
+\item The programming
 model is easy to implement on top of the occupancy-bound execution
@@
 proof-of-concept implementation on top of OpenCL 2.0 that requires no
-special-purpose hardware or driver support.  Fifth, our results
+special-purpose hardware or driver support.
+
+\item Our results
 suggest that cooperative kernels can provide efficient multi-tasking
@@
 support.
+
+\end{enumerate}
 
@@
 
+\ADComment{A natural exit does not decrement $M$.}
+
 As with a regular OpenCL kernel, a cooperative kernel is launched by
@@
 
+\ADComment{Perhaps remove this because a driver-based approach could handle this.}
 A regular OpenCL kernel can be executed by a multi-dimensional grid of
@@
 A variable declared in the root scope of the cooperative kernel can
-optionally be annotated with a new \texttt{restore} qualifier.  By
-annotating a variable with \texttt{restore}, the programmer indicates
-that when new workgroups are spawned via $\offerkill$, the threads in
+optionally be annotated with a new \restore qualifier.  By
+annotating a variable with \restore, the programmer indicates
+that when new workgroups are spawned via $\offerfork$, the threads in
 these workgroups should obtain a well-defined initial value for the
-variable, obtained from the workgroup that executed $\offerkill$.  We
-detail the semantics for this when we describe $\offerkill$ below.
+variable, obtained from the workgroup that executed $\offerfork$.  We
+detail the semantics for this when we describe $\offerfork$ below.
 
-Only private memory variables can be annotated with \texttt{restore},
+\ADComment{Remove?}
+Only private memory variables can be annotated with \restore,
 which may include pointers to local or global memory so long as the
@@
 particular, variables and arrays in local memory cannot be annotated
-with \texttt{restore}.  \ADComment{The preceding is what we have
+with \restore.  \ADComment{The preceding is what we have
   implemented, but actually it is pretty clear how we could handle
@@
 \paragraph{Semantics for $\offerkill$}
+
+\ADComment{Don't allow zero to be killed.}
+
+\ADComment{Explain that the ``last one dies'' semantics is convenient because it maintains contiguous ids, but other options are possible and may be more efficient sometimes.}
 
@@
 
-\item for each variable $v$ annoated with \texttt{restore}, every new
+\item for each variable $v$ annoated with \restore, every new
   thread's copy of $v$ is initialised to the value that thread 0 in
@@
 
-The second exception, for restored variables, can be stated more
-formally as follows.  Let $\mathit{WG}[x][y].v$ denote the value
-stored in variable $v$ of thread $y$ in workgroup $x$.  Then for each
-new workgroup $i \in [M, M+k-1]$ and each thread $j$ in workgroup $i$,
-$\mathit{WG}[i][j].v$ is initialised to $\mathit{WG}[m][0].v$ for each
-variable $v$ annotated with \texttt{restore}.
-
 Notice that $k=0$ is always a legitimate choice for the number of
 workgroups to be spawned by an $\offerfork$ call, and if the number of
-active workgrouups $M$ is equal to the workgroup limit $N$, $k=0$ is
+active workgroups $M$ is equal to the workgroup limit $N$, $k=0$ is
 guaranteed.
@@
 because, as discussed in Section~\ref{ }, an inter-workgroup barrier
-requires fair scheduling of workgroups.  As a result, past works have
-either split computation into multiple separate kernels, so that the
-end of one kernel and the start of the next kernel implicitly provides
-a global barrier~\cite{ }, or have assumed the occupancy-bound
-execution model to build a global barrier that works reliably on
-today's GPU platforms.
+requires fair scheduling of workgroups.
 
@@
 found that global barriers often provide a natural point for
-cooperation with the scheduler, and (b) identifying a known point at
-which workgroups are required to synchronize and at which workgroups
-can be safely killed and forked enables a custom implementation of
+cooperation with the scheduler, and (b) a custom implementation of
 $\resizingglobalbarrier()$ that provides the semantics described
-above, but is significantly more efficient than the above sequence of
+above can be made significantly more efficient than the above sequence of
 calls.  We discuss our efficient implementation of
 $\resizingglobalbarrier$ in detail in Section~\ref{...}.
-
-\subsection{Backwards compatibility}
-
-\ADComment{Perhaps this would fit better under ``programming guidelines'' -- not sure.}
-
-\ADComment{Make the point here that if an irregular algorithm is
-  written as a cooperative kernel then it will execute just as safely
-  as it does today on current hardware if: (1) the \texttt{restore}
-  annotation and $\offerkill$/$\offerfork$ calls are defined away to
-  nothing, (2) $\resizingglobalbarrier$ is redefined to
-  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
-  that assumes the occupancy-bound execution model is provided.
-If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}
 
@@
 \begin{lstlisting}
-  if($\getgroupid{0}$ != 0) $\offerkill$;
+  if($\getgroupid{0}$ != 0) $\offerkill$();
 \end{lstlisting}
@@
 
-The main impact of this relates to partitioning of data to be
-processed.
-
-\ADComment{Tyler, can we give an example of a data processing loop
-  that would have to be changed?}
-
 \paragraph{Cooperative kernels that only use barriers}
@@
 
+\ADComment{Shorten sentences.}
 In this case, because the resizing barrier is a kernel-level function,
@@
 interval, so the result returned by $\getnumgroups{0}$ is guaratneed
-to be identical on repeated calls during such an interval.  This
-guarantee may be exploited for efficiently when writing a cooperative
-kernel that uses global barriers but does not call $\offerfork$ and
-$\offerkill$ directly.
+to be identical on repeated calls during such an interval.
+
+This guarantee may be exploited by algorithms that peform strided data
+processing.  \ADComment{Tyler, can we give an example of a data
+  processing loop that would have to be changed?}
 
@@
 
+\ADComment{Sree mentioned TimeGraph paper.}
+
 \paragraph{Sufficient $\offerkill$ calls}
@@
 
+\ADComment{Forward reference to experiments with display freezing.}
+
 \ADComment{Got here.}
+
+\ADComment{Generosity: initial number of workgroups, and responding to offer fork.}
+
+\subsection{Alternative Semantic Choices}
+
+\ADComment{Explain here where we could have done things differently.}
+
+\ADComment{What are difficultiese with other alternatives and why we didn't use them.}
+
+\ADComment{Allowing zero to be killed.}
+
+\subsection{Backwards Compatibility}
+
+\ADComment{Perhaps this would fit better under ``programming guidelines'' -- not sure.}
+
+\ADComment{Make the point here that if an irregular algorithm is
+  written as a cooperative kernel then it will execute just as safely
+  as it does today on current hardware if: (1) the \restore
+  annotation and $\offerkill$/$\offerfork$ calls are defined away to
+  nothing, (2) $\resizingglobalbarrier$ is redefined to
+  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
+  that assumes the occupancy-bound execution model is provided.
+If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}
 

### 2016-11-09 src/non_persistent_kernels/matmult/matmult.h
Add message for matrix multiply dimensions

@@ -102,6 +102,9 @@ void init_non_persistent_app(CL_Execution *exec) {
     FLAGS_B_col = FLAGS_matdim;
   }
 
+  printf("Matrix multiply input dimensions: A (%d * %d), B (%d * %d)\n",
+         FLAGS_A_row, FLAGS_A_col, FLAGS_B_row, FLAGS_B_col);
+
   if (FLAGS_A_col != FLAGS_B_row) {
     cout << "Error: incompatile matrix size (A col: " << FLAGS_A_col;
     cout << ", B row: " << FLAGS_B_row << ")" << endl;

### 2016-11-09 src/test_suite/final_octree/octree.h
Serialize settings of kernel args

@@ -122,62 +122,27 @@ void init_persistent_app_for_real(CL_Execution *exec, int occupancy) {
 
 }
 
+/*---------------------------------------------------------------------------*/
+
 void set_persistent_app_args_for_real(int arg_index, cl::Kernel k) {
   // Set args for persistent kernel
-  int err = 0;
-
-  err = k.setArg(arg_index, randdata);
-  arg_index++;
-  check_ocl(err);
-
-  err = k.setArg(arg_index, particles);
-  arg_index++;
-  check_ocl(err);
-
-  err = k.setArg(arg_index, newParticles);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, tree);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, FLAGS_numParticles);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, treeSize);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, particlesDone);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, FLAGS_maxChildren);
-  arg_index++;
-  check_ocl(err);
-
-  err = k.setArg(arg_index, max_workgroups);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, deq);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, dh);
-  arg_index++;
-  check_ocl(err);
-  err = k.setArg(arg_index, maxlength);
-  arg_index++;
-  check_ocl(err);
-
-  /* frompart */
-  err = k.setArg(arg_index, NULL);
-  arg_index++;
-  check_ocl(err);
-
-  /* topart */
-  err = k.setArg(arg_index, NULL);
-  arg_index++;
-  check_ocl(err);
-
+  check_ocl(k.setArg(arg_index++, randdata));
+  check_ocl(k.setArg(arg_index++, particles));
+  check_ocl(k.setArg(arg_index++, newParticles));
+  check_ocl(k.setArg(arg_index++, tree));
+  check_ocl(k.setArg(arg_index++, FLAGS_numParticles));
+  check_ocl(k.setArg(arg_index++, treeSize));
+  check_ocl(k.setArg(arg_index++, particlesDone));
+  check_ocl(k.setArg(arg_index++, FLAGS_maxChildren));
[…]

### 2016-11-09 write-ups/PLDI/paper.tex
Progress.

@@
 
+\newcommand{\NumAlgorithms}{8}
+
 \begin{document}
@@
 
-\ADComment{Address issue of placing constructs manually --
-  transactional semantics and barriers.  Argue why it's easy to do this.}
-
-\ADComment{Mention dynamic parallelism ane explain why it's no good.}
+\ADComment{Open to attack: why not just buy another GPU?}
 
@@
 to maintain a queue, which is typically protected by a mutex to enable
-stealing by other workgroups.  \ADComment{Perhaps
-  make the point that inter-workgroup synchronization is possible in
-  very rare cases without blocking, e.g. threadfence reduction.}
+stealing by other workgroups.
 
@@
 data-parallel algorithms accelerated on a GPU, this translates to
-requiring fair scheduling of workgroups.  \ADComment{Shorten} Without fairness, a
+requiring fair scheduling of workgroups.  Without fairness, a
 workgroup may fail to make progress, leading to starvation.  For
-example, if workgroup $A$ holds a mutex and workgroup $B$ is
-spin-waiting for the mutex to be released, an unfair scheduler can
-forever schedule $B$, so that $A$ never releases the mutex.
-Similarly, an unfair scheduler can induce starvation if workgroup $A$
-is at an inter-workgroup barrier awaiting the arrival of other
-workgroups, by allowing A to spin forever so that the other workgroups
-do not progress to the barrier.
+example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
+another workgroup $B$ to spin-wait forever for the mutex to be
+released.  Similarly, an unfair scheduler can a workgroup to spin-wait
+at a global barrier so that other workgroups do not make progress
+towards the barrier.
 
-\paragraph{Occupancy-bound execution}
-\ADComment{Add footnote to acknowledge multiple wgs per cu}
-\ADComment{Say that current GPU schedulers are not required to be fair.}
-Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
-and HSA~\cite{...}---specify very few constraints on the manner in
-which workgroups are scheduled.  In particular, they do \emph{not}
-specify that workgroups must be fairly scheduled, and implementations
-of these programming models do not provide fair scheduling in
-practice.  Roughly speaking, each workgroup executing a GPU kernel is
-mapped to a hardware \emph{compute unit}.  If a kernel is executed
-with more workgroups than there are compute units then at least two
-workgroups must share a single compute unit.  The simplest way to
-handle this is via an \emph{occupancy-bound} execution model, whereby
-once a workgroup has commenced execution on a compute unit (it has
-become \emph{occupant}), the workgroup has exclusive access to the
-compute unit until it has finished execution, after which a subsequent
-workgroup can be bound to the compute unit.  Experiments suggest that
-the occupancy-bound execution model is widely implemented by today's
-GPU devices and drivers~\cite{...}, and it is evidently a simple model
-for device vendors to implement efficiently.
+\paragraph{A degree of fairness: occupancy-bound execution}
+GPU schedulers are \emph{not} fair. the current GPU
+programming models---OpenCL~\cite{...}, CUDA~\cite{...}  and
+HSA~\cite{...}---specify almost no guarantees regarding scheduling of
+workgroups, and implementations of these programming models do not
+provide fair scheduling in practice.  Roughly speaking, each workgroup
+executing a GPU kernel is mapped to a hardware \emph{compute
+  unit}.\footnote{In practice the situation can be more complex: if a
+  kernel is sufficiently simple and if workgroups are small enough,
+  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
+%
+If a kernel is executed with more workgroups than there are compute
+units then at least two workgroups must share a single compute unit.
+The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
+execution model~\cite{...}.  With this model, once a workgroup has commenced execution on a
+compute unit (it has become \emph{occupant}), the workgroup has
+exclusive access to the compute unit until it has finished execution,
+after which a subsequent workgroup can be bound to the compute unit.
+Experiments suggest that the occupancy-bound execution model is widely
+implemented by today's GPU devices and drivers~\cite{...}.  It is
+evidently a simple model for device vendors to implement efficiently.
 
@@
 assume that the target GPU device provides the occupancy-bound
-execution model, which they exploit by scheduling no more workgroups
+execution model.  They exploit the model by scheduling no more workgroups
 than there are available compute units~\cite{...}.  This works
@@
 kernel occupies all compute units, the GPU cannot be used for other
-tasks, such as graphics rendering, in the meantime.  \ADComment{Could cut or move to later:} Precisely because
+tasks, such as graphics rendering, in the meantime.  Precisely because
 of this problem, operating systems employ a GPU \emph{watchdog} to
 heuristically detect and kill long-running computations on the GPU
-being used for display rendering.  If this watchdog is disabled,
-execution of a long-running kernel causes the display to freeze for
-the duration of kernel execution.  Second, \emph{energy throttling} is
+being used for display rendering~\cite{...}.  Second, \emph{energy throttling} is
 an important concern for battery-powered devices, such as smartphones
-and tablets.  In the future, it will be desirable for a mobile GPU driver to be able to
+and tablets~\cite{...}.  In the future, it will be desirable for a mobile GPU driver to be able to
 power down one or more compute units if the battery level is low or if
@@
 practitioners who have been involved in the OpenCL and/or HSA
-standardisation efforts~\cite{CitePersonalCommunication}, is that
-\textbf{(1)} GPU vendors will not commit to the occupancy-bound
-execution model that they currently implement, due to the need for
-multi-tasking and energy throttling, yet \textbf{(2)} GPU vendors will
-not guarantee fair scheduling, due to the runtime overhead associated
-with supporting full preemption of workgroups~\cite{ISCAPAPER}.  Vendors instead wish to retain the essence of the
-occupancy-bound execution model, which is simple to implement
-efficiently, and provide preemption for \ADComment{shorten} the special cases associated
-with multi-tasking for high-priority jobs (most crucially graphics)
-and energy throttling.  In particular, they do not want to commit to
-fair scheduling of a workgroup once it has been preempted.
+standardisation efforts~\cite{CitePersonalCommunication}, is that (1)
+GPU vendors will not commit to the occupancy-bound execution model
+that they currently implement, due to the need for multi-tasking and
+energy throttling, yet (2) GPU vendors will not guarantee fair
+scheduling, due to the runtime overhead associated with supporting
+full preemption of workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors
+instead wish to retain the essence of the occupancy-bound execution
+model, which is simple to implement efficiently, and provide
+support for preemption only in key special cases.
 
-\ADComment{Kepler preemption support?}
-
-\paragraph{Our proposal}
+\paragraph{Our proposal: cooperative kernels}
 %
@@
 that requires workgroups to be fairly scheduled is identified as a
-\emph{cooperative} kernel, and written using two additional language
-primitives, $\offerkill$ and $\offerfork$, which must be placed
-carefully by the programmer.
+\emph{cooperative} kernel, and is written using two additional
+language primitives, $\offerkill$ and $\offerfork$, which must be
+placed carefully by the programmer.
 
 At a point where the cooperative kernel could proceed with fewer
-workgroups, a workgroup can execute $\offerkill$ to sacrifice itself to
-the scheduler.  This indicates to the scheduler that the workgroup
-would ideally continue with its execution, but that if required the
-scheduler may destroy the workgroup for good: the cooperative kernel
-is prepared to deal with either scenario.
+workgroups, a workgroup can execute $\offerkill$, offering to
+sacrifice itself to the scheduler.  This indicates to the scheduler
+that the workgroup would ideally continue with its execution, but that
+if required the scheduler may destroy the workgroup for good: the
+cooperative kernel is prepared to deal with either scenario.
 
@@
 is able to take advantage of having one or more additional workgroups
-commencing execution at the $\offerfork$ program point.
+commence execution directly after the $\offerfork$ program point.
 
@@
 scheduler must guarantee that the workgroups executing a cooperative
-kernel are fairly scheduled, while the cooperative kernel must be robust to
-workgroups leaving and joining the computation in response to $\offerkill$
-and $\offerfork$.
-Non-functionally, a cooperative kernel must ensure that $\offerkill$
-is executed frequently enough that the scheduler can accommodate
-soft-real time constraints, e.g.\ allowing a smooth frame-rate for
-graphics or allowing compute units to be powered down quickly when
-required.  The scheduler should respond to $\offerkill$ and
-$\offerfork$ calls in a manner that avoids under-utilisation of
-hardware resources by the cooperative kernel, e.g.\ the scheduler
-should not kill workgroups unless they are required for another
-purpose.
+kernel are fairly scheduled, while the cooperative kernel must be
+robust to workgroups leaving and joining the computation in response
+to $\offerkill$ and $\offerfork$.  Non-functionally, a cooperative
+kernel must ensure that $\offerkill$ is executed frequently enough
+that the scheduler can accommodate soft-real time constraints,
+e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
+units to be powered down quickly when required.  The scheduler should
+respond to $\offerkill$ and $\offerfork$ calls in a manner that avoids
+under-utilisation of hardware resources by the cooperative kernel,
+e.g.\ the scheduler should not kill workgroups unless they are
+required for another purpose, and should facilitate forking of
+additional workgroups when possible.
 
-Cooperative kernels have several appealing properties:
+The cooperative kernels programming model has several appealing
+properties:
 
@@
 
-\item They
-accommodate the needs of the irregular
-data-parallel algorithms that have been successfully accelerated using GPUs: such algorithms can be implemented naturally
-as cooperative kernels \ADComment{boast that we did a good set}, so
-that fair scheduling between workgroups is guaranteed.
+\item By providing fair scheduling between workgroups, cooperative
+  kernels meet the needs of blocking algorithms, including irregular
+  data-parallel algorithms.
 
-\item They
-have no impact on the development of regular (non-cooperative
-kernels): these can be programmed exactly as they are now.
+\item The model has no impact on the development of regular
+  (non-cooperative) kernels: these can be programmed exactly as they
+  are now, as can graphics shading routines written using APIs such as
+  OpenGL, Vulkan and DirectX.
 
-\item The
-programming model is backwards compatible: a non-blocking algorithm
-implementation that works on today's GPU platforms can be upgraded to
-use our language extensions for cooperative scheduling, and will
+\item The model is backwards-compatible: a non-blocking algorithm
+  implementation that works on today's GPU platforms can be upgraded
+  to use our language extensions for cooperative scheduling, and will
   behave exactly as it does at present if the new $\offerkill$ and
@@
 
-\item The programming
-model is easy to implement on top of the occupancy-bound execution
-model that current GPU platforms provide; we present a
-proof-of-concept implementation on top of OpenCL 2.0 that requires no
-special-purpose hardware or driver support.
+\item The model can be implemented on top of the occupancy-bound
+  execution model that current GPU platforms provide---we present a
+  proof-of-concept implementation on top of OpenCL 2.0 that requires
+  no special-purpose hardware or driver support.
 
-\item Our results
-suggest that cooperative kernels can provide efficient multi-tasking
-of cooperative and non-cooperative tasks: our experiments across a
-range of GPUs from two vendors show very good response times for
-scheduling short, non-cooperative kernels concurrently with
-long-running cooperative kernels.  These performance results are with
-respect to our proof-of-concept, portable implementation; a vendor
-specific implementation of cooperative kernels could likely yield
-better performance still through device-specific driver and compiler
-support.
+\item Our experiments across four GPUs from two vendors show that our
+  portable, proof-of-concept implementation of cooperative kernels can
+  provide efficient multi-tasking of cooperative and non-cooperative
+  tasks.  A GPU vendor implementation could likely provide even better
+  efficiency, by leveraging platform-specific hardware, driver and
+  compiler features.
 
@@
 
+Placing the $\offerkill$ and $\offerfork$ primitives requires manual
+effort and care, but our experience porting a representative set of
+GPU-accelerated irregular algorithms to use cooperative kernels
+suggests that this is straightforward in practice.  We discuss this in
+detail in Section~\ref{sec:portingalgorithms}.
+
 In summary, our main contributions are:
 
-\begin{enumerate}
+\begin{itemize}
 
-\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms  \ADComment{Maybe big up to say that we give semantics to the new operations.}
+\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms.
+We describe the semantics of the programming model in detail (Section~\ref{sec:cooperativekernels}).
 
-\item A portable implementation of cooperative kernels on top of the OpenCL 2.0 programming model
+\item A proof-of-concept portable implementation of cooperative
+  kernels on top of the OpenCL 2.0 programming model
+  (Section~\ref{sec:implementation}).
 
-\item The porting of \ADComment{TODO} irregular graph algorithms \ADComment{tighten} to the cooperative kernels model
+\item An experience report porting \NumAlgorithms{} irregular GPU
+  algorithms to use cooperative kernels, and experimental results
+  assessing the performance overhead associated with cooperation and
+  the responsiveness to high priority tasks that cooperation enables,
+  across four GPUs from two vendors (Section~\ref{sec:experiments}).
 
-\item \ADComment{The experiments}
+\end{itemize}
 
-\end{enumerate}
+We begin by providing an overview of the OpenCL programming model and
+two motivating examples of irregular algorithms (Section~\ref{sec:background}).
+At the end of the paper we discuss related work (Section~\ref{sec:relatedwork}) and avenues for future work (Section~\ref{sec:conclusion}).
 
-\ADComment{Text to maybe work in.}
-
-In traditional co-operative multi-tasking, a task relinquishes control
-and regains control in a stateful manner. We propose a lighter weight
-model we believe to suitible for the type of blocking algorithms we
-are aware of for GPUs.
-
-The scheduler agrees to scheduler the workgroups fairly, but is at
-liberty to eliminate workgroups for good when they offer themselves
-to be killed. This allows the schduler to repond to changing depands
-on the GPU, e.g. to implement energy throttling.
-
-Prototype implementation of the programming model using OpenCL 2.0
-advance features to build the scheudler and implement a clang based
-compiler to transform input programs using the language extensions to
-lower-level programs that interact with the scheduler. Our prototype
-simulates multiple independent kernels into a mega-kernel. A vendor
-specific implementation could achieve higher efficiency through
-modications to the GPU driver that may take advantage of hardware
-features. (e.g. mega kernel might not be good to compile). Its
-reasonable to think our proof of concept provides a lower bound on how
-well we can do.
-
-We ported a large set of blocking algorithms to our programming
-model. We evaluate co-operative scheduling with a cooperative kernel
-and non-coperative kernel. GPUs from multiple vendors and access
-various things.
-
-Our results are promising.
-
-In the evaluation, make a note that the applications we worked on
-Namely if they use only a barrier or work-stealing idiom, are
-straightforward to adopt to our programming model.
-
-\section{Motivating example: work stealing}
+\section{Background and Examples}\label{sec:background}
 
@@
 
-\section{The Cooperative Kernels Programming Model}
+\section{The Cooperative Kernels Programming Model}\label{sec:cooperativekernels}
 
 We describe our cooperative kernels programming model as an extension
-to the industry-standard OpenCL programming model for heterogenous
+to the industry-standard OpenCL programming model for heterogeneous
 many-core architectures.  However, the cooperative kernels concept is
 more general, and could be applied to extend other GPU programming
-models, such as CUDA and HSA.  \ADComment{However, unclear that we could implement it on top of these, but that's not really the point.}
+models, such as CUDA and HSA.
 
-\subsection{Semantics of Cooperative Kernels}
+We describe the semantics of cooperative kernels
+(Section~\ref{sec:semantics}), then summarise the important
+differences between programming with regular vs.\ cooperative kernels
+(Section~\ref{sec:programmingguidelines}) and the non-functional
+properties that a developer of a cooperative kernel, and the
+implementer of a scheduler for cooperative kernels, should strive for
+(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
+kernels has been guided by the applications that we have studied
+(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
+have taken particular semantic decisions where other reasonable
+alternatives were possible (Section~\ref{sec:semanticalternatives}).
+We conclude the section by discussing the backwards-compatible nature
+of cooperative kernels with respect to blocking algorithms designed to
+run on current GPU platforms
+(Section~\ref{sec:backwardscompatibility}).
+
+\subsection{Semantics of Cooperative Kernels}\label{sec:semantics}
 
@@
 
-\item for each variable $v$ annoated with \restore, every new
+\item for each variable $v$ annotated with \restore, every new
   thread's copy of $v$ is initialised to the value that thread 0 in
@@
 is not provided as a primitive in OpenCL, CUDA or HSA.  This is
-because, as discussed in Section~\ref{ }, an inter-workgroup barrier
+because, as discussed in Section~\ref{sec:intro}, an inter-workgroup barrier
 requires fair scheduling of workgroups.
@@
 as a primitive.  As noted above, the $\resizingglobalbarrier()$
-primitive is, semantically, synactic sugar for a sequence of other
+primitive is, semantically, syntactic sugar for a sequence of other
 primitive calls.  We include it as a primitive because (a) we have
@@
 calls.  We discuss our efficient implementation of
-$\resizingglobalbarrier$ in detail in Section~\ref{...}.
+$\resizingglobalbarrier$ in detail in Section~\ref{sec:resizingbarrier}.
 
-\subsection{Programming guidelines}
+\subsection{Programming Guidelines}\label{sec:programmingguidelines}
 
@@
 
-The programer is responsible for keeping at least one workgroup alive.
+The programmer is responsible for keeping at least one workgroup alive.
 In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.
@@
 If a cooperative kernel does not directly issue $\offerkill$ and
-$\offerfork$ calls then the only possibilty for the number of
+$\offerfork$ calls then the only possibility for the number of
 workgroups to change is due to $\resizingglobalbarrier$ being invoked.
@@
 barrier interval.  As a result, the number of workgroups executing the
-kernel is guaratneed to remain unchanged within each resizing barrier
-interval, so the result returned by $\getnumgroups{0}$ is guaratneed
+kernel is guaranteed to remain unchanged within each resizing barrier
+interval, so the result returned by $\getnumgroups{0}$ is guaranteed
 to be identical on repeated calls during such an interval.
 
-This guarantee may be exploited by algorithms that peform strided data
+This guarantee may be exploited by algorithms that perform strided data
 processing.  \ADComment{Tyler, can we give an example of a data
@@
 
-\subsection{Non-functional Requirements of Cooperative Kernels}
+\paragraph{Placing the new constructs}
 
-The semantics presented in Section~\ref{...} describe the envelope of
+\ADComment{TODO: needs work.}
+.  The irregular
+algorithms we encountered either employ work stealing or operate on
+graph data structures.  Work stealing algorithms have a transactional
+flavour, whereby tasks are repeatedly pulled from a queue and
+processed by a workgroup.  The point of task completion---the end of a transaction---is a natural place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to indicate that
+
+\subsection{Non-Functional Requirements of Cooperative Kernels}\label{sec:nonfunctional}
+
+The semantics presented in Section~\ref{sec:semantics} describe the envelope of
 behaviour that a developer of a cooperative kernel should be prepared
@@
 
-To achieve this balance, both an implementation of the ooperative
+To achieve this balance, both an implementation of the cooperative
 kernels model, and the programmer of a cooperative kernel, must strive
@@
 
-\subsection{Alternative Semantic Choices}
+\subsection{Alternative Semantic Choices}\label{sec:semanticalternatives}
 
@@
 
-\ADComment{What are difficultiese with other alternatives and why we didn't use them.}
+\ADComment{What are difficulties with other alternatives and why we didn't use them.}
 
@@
 
-\subsection{Backwards Compatibility}
+\subsection{Backwards Compatibility}\label{sec:backwardscompatibility}
 
@@
 
-\subsection{Preemption-aware work stealing}
-
-\ADComment{Remind the reader that these are inspired by work stealing
-  and barrier synchonization.  In \mysec\ref{ } we describe the
-  semantics of kernels that use these primitives.  In \mysec\ref{ } we
-  show how work stealing can be implemented.  In \mysec\ref{ } we show
-  how a barrier can be implemented.}
-
-\ADComment{Show how work stealing can now be made safe.}
-
-\subsection{A preemption-aware global barrier}
-
-\ADComment{Show how the global barrier can be written in a manner that
-  maintains contiguous ids using the refined semantics.}
-
-\section{Application-specific optimisations}
-
-\ADComment{Explain that for global barriers, we explain how some new
-  primitives can lead to a better implementation.  Explain that for
-  work stealing there are various ways the existing ckill and cfork
-  primitives can be used.}
-
-\subsection{An optimised barrier implementation}
-
-\ADComment{Remark that the barrier of \mysec\ref{sec:naivebarrier} may
-  be inefficient, discuss the use of query primitives to make this
-  more efficient.}
-
-\subsection{Variations on work stealing}
-
-\ADComment{Three cases here: always ckill/cfork, ckill/cfork only when
-  queue is empty, and have a ``mediator'' workgroup to manage things.}
-
-\section{A mega kernel-based prototype}
+\section{Prototype Implementation}\label{sec:implementation}
 
@@
 
-\section{Experimental evaluation}
+\ADComment{Give details of how host and device interact.}
 
-\ADComment{TODO: write up notes from whiteboard chat.}
+\ADComment{Explain kernel merge tool.}
 
-\section{Related work}
+\subsection{An Efficient Resizing Barrier}\label{sec:resizingbarrier}
 
-\section{Conclusions and future work}
+\ADComment{Optimised barrier implementation.}
+
+\section{Applications and Experiments}\label{sec:experiments}
+
+\subsection{Porting Irregular GPU Algorithms to Cooperative Kernels}\label{sec:portingalgorithms}
+
+\ADComment{Among other things, make sure we address the point that placing the primitives was pretty easy.}
+
+\section{Related Work}\label{sec:relatedwork}
+
+\section{Conclusions and Future Work}\label{sec:conclusion}
 
@@
 
-\section{Old intro}
+\section{Notes}
 
-\ADComment{Points to make: It is important to be able to execute
-  long-running, cooperative applications on GPUs (give examples).
-  However, resource sharing and resource flexibility is also important
-  (give examples, including sharing resources with graphics, sharing
-  resources between compute taks, and being flexible about resources
-  under energy constraints), and these are not supported by today's
-  GPU platforms: long-running applications cause display freezes, etc.
-  Resource-based preemption, whereby workgroup execution is postponed
-  to free up resources for other purposes (or to reduce resources to
-  save energy) can potentially enable resource sharing.  There is
-  evidence in the research literature that efficient resource sharing
-  for GPUs is possible (cite academic works), and signs from industry
-  that this is coming in future architectures.  [Explain more clearly
-    what we mean by resource-based preemption.]  Unfortunately,
-  arbitrary resource-based preemption is fundamentally incompatible
-  with cooperative applications---only trivial examples of cooperation
-  are safe [reduction, load balancing via atomic increment]; the HSA
-  model of guaranteeing preemption in reverse order of IDs allows some
-  cooperative applications to be implemented.  Still, in the general
-  case even a mutex cannot be used safely.}
+  \ADComment{Perhaps
+  make the point that inter-workgroup synchronization is possible in
+  very rare cases without blocking, e.g. threadfence reduction.}
 
-\ADComment{Say that we believe the problem can only be solved through
-  programming language innovations with appropriate runtime support.
-  We have observed inter-workgroup barriers and work stealing to be
-  the main use cases for cooperation, and so in this paper propose two
-  new language primitives, ckill and cfork, designed to meet the needs
-  of these primitives.  [Describe them a bit.]}
+\ADComment{If this watchdog is disabled,
+execution of a long-running kernel causes the display to freeze for
+the duration of kernel execution.}
 
-\ADComment{Move on to our main contributions}
+\ADComment{Kepler preemption support?  Mention in related work?}
 
-\ADComment{This is what we had before.}
+\ADComment{Mention dynamic parallelism and explain why it's no good.}
 
-Graphics processing units (GPUs) are highly parallel co-processors that are now incorporated into many
-devices, ranging from the world's most powerful supercomputers to battery-powered devices such as tablets and mobile phones. While
-originally designed to accelerate graphics applications, GPUs are now
-used in a wide range of applications, including molecular simulations
-and weather forecasting. \ADComment{Expand the list of application areas a bit more; mention their attractiveness due to energy-efficiency.}
-
-New application domains for GPUs are enabled by general purpose
-programming languages with associate compilation tools that target GPU architectures.
-The current most popular GPU programming languages are CUDA and OpenCL. These languages allow developers to
-write C-like programs that can be dispatched and executed on a GPU
-device.
-
-GPUs are massively parallel, employing many threads to execute a
-program.  These threads are organised hierarchically: groups of threads
-are called \emph{workgroups}, and threads interact differently depending on whether they reside in the same workgroup or not.  Traditionally, applications accelerated by GPUs are
-data-parallel, requiring little interaction between threads. A small
-set of bulk synchronisation primitives are provided and any
-inter-thread communication must be performed around these primitives.  \ADComment{Do we want to make the point that this is traditionally limited to the intra-workgroup case?}
-
-While this traditional way to program GPUs has proved useful, it also
-has limitations for some applications, in particular applications that
-expose a dynamic amount of parallelism throughout execution. Such
-applications are said to have irregular parallelism (we call such
-applications \emph{IP applications}). Efficient parallel computation
-of IP applications requires finer-grained synchronisation than that
-which is provided by GPU language primitives.
+\section{Remainder of old intro}
 

### 2016-11-09 src/lonestar/graph_apps/bfs/bfs.h
added lonestar apps bfs and sssp

@@ -1,0 +1,371 @@
+#include "portable_endian.h"
+
+// Could probably go into a pannotia header
+DEFINE_string(graph_file, "", "Path to the graph_file");
+DEFINE_string(graph_output, "", "Path to output the graph result");
+DEFINE_string(graph_solution_file, "pannotia/solutions/sssp_usa.txt", "the path the mega kernel file");
+
+
+// ---
+DEFINE_string(restoration_ctx_path, "lonestar/graph_apps/bfs", "Path to restoration context");
+DEFINE_string(merged_kernel_file, "lonestar/graph_apps/bfs/device/merged.cl", "the path the mega kernel file");
+DEFINE_string(persistent_kernel_file, "lonestar/graph_apps/bfs/device/standalone.cl", "the path the mega kernel file");
+
+#define MYINFINITY 1000000000
+
+typedef struct {
+
+	cl_uint nnodes, nedges;
+	cl_uint *noutgoing, *psrc, *edgessrcdst;
+
+} Graph;
+
+cl::Buffer d_dist, d_in_wl, d_in_index, d_out_wl, d_out_index,  d_g_edgessrcdst, d_g_psrc;
+cl_uint g_nnodes;
+Graph hgraph;
+cl_uint *zero_array;
+cl_uint *read_array;
+const float wl_mult = .25;
+cl_int* h_wl;
+cl_int h_index;
+
+void * safe_malloc(size_t x) {
+	void * tmp = malloc(x);
+	if (!tmp) {
+		check_ocl(-1);
+	}
+	return tmp;
+}
+
+unsigned allocOnHost(Graph * g) {
+  g->edgessrcdst = (cl_uint *) safe_malloc((g->nedges+1) * sizeof(unsigned int)); // First entry acts as null.
+  g->psrc = (cl_uint *) calloc(g->nnodes+1, sizeof(unsigned int));           // Init to null.
+  g->psrc[g->nnodes] = g->nedges;                                            // Last entry points to end of edges, to avoid thread divergence in drelax.
+  g->noutgoing = (cl_uint *) calloc(g->nnodes, sizeof(unsigned int));        // Init to 0.
+  return 0;
+}
+
[…]

### 2016-11-09 src/lonestar/graph_apps/bfs/restoration_ctx.h

@@ -1,0 +1,10 @@
+#pragma once
+typedef struct {
+  CL_UCHAR_TYPE target;
+  CL_INT_TYPE iteration;
+  MY_CL_GLOBAL CL_INT_TYPE * in_wl;
+  MY_CL_GLOBAL CL_INT_TYPE * out_wl;
+  MY_CL_GLOBAL CL_INT_TYPE * in_index;
+  MY_CL_GLOBAL CL_INT_TYPE * out_index;
+} Restoration_ctx;
+

### 2016-11-09 src/lonestar/graph_apps/sssp/sssp.h

@@ -1,0 +1,382 @@
+#include "portable_endian.h"
+
+// Could probably go into a pannotia header
+DEFINE_string(graph_file, "", "Path to the graph_file");
+DEFINE_string(graph_output, "", "Path to output the graph result");
+DEFINE_string(graph_solution_file, "pannotia/solutions/sssp_usa.txt", "the path the mega kernel file");
+
+
+// ---
+DEFINE_string(restoration_ctx_path, "lonestar/graph_apps/sssp", "Path to restoration context");
+DEFINE_string(merged_kernel_file, "lonestar/graph_apps/sssp/device/merged.cl", "the path the mega kernel file");
+DEFINE_string(persistent_kernel_file, "lonestar/graph_apps/sssp/device/standalone.cl", "the path the mega kernel file");
+
+#define MYINFINITY 1000000000
+
+typedef struct {
+
+	cl_uint nnodes, nedges;
+	cl_uint *noutgoing, *psrc, *edgessrcdst, *edgessrcwt;
+
+} Graph;
+
+cl::Buffer d_dist, d_in_wl, d_in_index, d_out_wl, d_out_index,  d_g_edgessrcdst, d_g_psrc, d_g_edgessrcwt;
+cl_uint g_nnodes;
+Graph hgraph;
+cl_uint *zero_array;
+cl_uint *read_array;
+const float wl_mult = .25;
+cl_int* h_wl;
+cl_int h_index;
+
+void * safe_malloc(size_t x) {
+	void * tmp = malloc(x);
+	if (!tmp) {
+		check_ocl(-1);
+	}
+	return tmp;
+}
+
+unsigned allocOnHost(Graph * g) {
+  g->edgessrcdst = (cl_uint *) safe_malloc((g->nedges+1) * sizeof(unsigned int)); // First entry acts as null.
+  g->psrc = (cl_uint *) calloc(g->nnodes+1, sizeof(unsigned int));           // Init to null.
+  g->psrc[g->nnodes] = g->nedges;                                            // Last entry points to end of edges, to avoid thread divergence in drelax.
+  g->noutgoing = (cl_uint *) calloc(g->nnodes, sizeof(unsigned int));        // Init to 0.
[…]

### 2016-11-09 src/host_skeleton/skel.cpp
and the rest of the junnk to make lonestar work

@@ -54,6 +54,10 @@ using namespace std;
 #include "../graph_apps/sssp/sssp.h"
 #elif defined PERSISTENT_PANNOTIA_BC
 #include "../graph_apps/bc/bc.h"
+#elif defined PERSISTENT_LONESTAR_BFS
+#include "../lonestar/graph_apps/bfs/bfs.h"
+#elif defined PERSISTENT_LONESTAR_SSSP
+#include "../lonestar/graph_apps/sssp/sssp.h"
 #elif defined PERSISTENT_OCTREE
 #include "octree.h"
 #elif defined PERSISTENT_CONNECT_FOUR
@@ -242,6 +246,8 @@ void run_persistent(CL_Execution *exec) {
 		                           file::Path(FLAGS_scheduler_rt_path),
 		                           file::Path(FLAGS_restoration_ctx_path),
 		                           FLAGS_use_query_barrier);
+
+	
 	check_ocl(err);
 	exec->exec_kernels["persistent"] = cl::Kernel(exec->exec_program, persistent_kernel_name(), &err);
 	check_ocl(err);
@@ -253,12 +259,15 @@ void run_persistent(CL_Execution *exec) {
 	arg_index = set_persistent_app_args_for_occupancy(arg_index, exec->exec_kernels["persistent"]);
 	err = exec->exec_queue.flush();
 	check_ocl(err);
-
+	
 	cl::Buffer d_bar(exec->exec_context, CL_MEM_READ_WRITE, sizeof(IW_barrier));
 	reset_barrier(exec, d_bar);
+	
 	err = exec->exec_kernels["persistent"].setArg(arg_index, d_bar);
 	arg_index++;
 	check_ocl(err);
+
+	
 
 	cl::Buffer d_ctx_mem(exec->exec_context, CL_MEM_READ_WRITE, sizeof(Discovery_ctx));
 	reset_discovery(exec, d_ctx_mem, true);

### 2016-11-09 src/host_skeleton/skel.cpp
fixed skel so that the default is 128 threads per workgroup

@@ -35,7 +35,7 @@ DEFINE_int32(use_query_barrier, 0, "flag to use regular ckill barrier or query c
 
 
 DEFINE_int32(run_non_persistent, 0, "Run only the non persistent task (solo run). If greater than 0, specifies the iterations");
-DEFINE_int32(threads_per_wg, 256, "Threads per workgroup for non persistent task solo runs");
+DEFINE_int32(threads_per_wg, 128, "Threads per workgroup for non persistent task solo runs");
 DEFINE_int32(num_wgs, 8, "Workgroups for non persistent task solo runs");
 DEFINE_int32(merged_iterations, 1, "Iterations to run the merged kernel for");
 

### 2016-11-09 src/run_test.py
Add configurations of non-persistent task, run for all configurations

@@ -15,13 +15,12 @@ ITERS         = 1
 PRINT         = fake_placeholder
 DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""
-NON_PRST_FREQ = ""
 
 PROGRAMS = {
     "pannotia_color",
-    "pannotia_mis",
-    "pannotia_bc",
-    "pannotia_sssp"
+    # "pannotia_mis",
+    # "pannotia_bc",
+    # "pannotia_sssp"
 }
 
 PROGRAM_DATA = {
@@ -29,9 +28,9 @@ PROGRAM_DATA = {
     "pannotia_color" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
                            "solution" : os.path.join("solutions", "color_ecology.txt"),
                            "stat" : "color_ecology" },
-                         { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
-                           "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
-                           "stat" : "color_G3_circuit" }
+                         # { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
+                         #   "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
+                         #   "stat" : "color_G3_circuit" }
     ],
 
     "pannotia_mis" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
@@ -54,13 +53,21 @@ PROGRAM_DATA = {
                           "stat" : "sssp_usa_road" } ]
 }
 
+# TODO: make it per machine ?
+MATMULT_CONFIG = [
+    # This is for Hugues laptop
+    { "freq" : "70", "matdim" : "160", "name" : "light" },
+    { "freq" : "40", "matdim" : "160", "name" : "medium" },
+    { "freq" : "40", "matdim" : "260", "name" : "high" },
+]
+
 def my_print(file_handle, data):
     print(data)
     file_handle.write(data + os.linesep)
 
 def exec_cmd(cmd, extr_args=[], flag=""):
     finalcmd = cmd + extr_args
-    PRINT("====================== RUN COMMAND =======================")
+    PRINT("====================== RUN START =========================")
     PRINT(flag)
     for s in finalcmd:
         PRINT(s)
@@ -75,7 +82,7 @@ def exec_cmd(cmd, extr_args=[], flag=""):
     PRINT("----------------------------------------------------------")
     str_ret_code = "Return code: " + str(ret_code)
     PRINT(str_ret_code)
[…]

### 2016-11-10 write-ups/PLDI/paper.tex
Worked on background.

@@
 \newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
-\newcommand{\getgroupid}[1]{\mathsf{get\_group\_id(#1)}}
-\newcommand{\getnumgroups}[1]{\mathsf{get\_num\_groups(#1)}}
+\newcommand{\getgroupid}{\mathsf{get\_group\_id}}
+\newcommand{\getnumgroups}{\mathsf{get\_num\_groups}}
+\newcommand{\getlocalid}{\mathsf{get\_local\_id}}
+\newcommand{\getglobalid}{\mathsf{get\_global\_id}}
+\newcommand{\getlocalsize}{\mathsf{get\_global\_id}}
 
@@
 
-\ADComment{Use the work stealing example to both introduce OpenCL and
-  also make our key point: that arbitrary preemption will not work.}
+We use two worked examples---work stealing and breadth-first graph
+search---to provide background on the OpenCL programming model and
+irregular data-parallel algorithms.
+
+\begin{figure}
+\ADComment{Tyler/Hugues: please insert workstealing example.}
+\begin{lstlisting}
+TODO
+\end{lstlisting}
+\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
+\end{figure}
+
+\begin{figure}
+\ADComment{Tyler/Hugues: please insert graph traversal example.}
+\begin{lstlisting}
+TODO
+\end{lstlisting}
+\caption{An OpenCL graph search algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
+\end{figure}
+
+\paragraph{OpenCL}
+An OpenCL program is divided into \emph{host} and \emph{device}
+components.  A host application runs on the CPU and launches one or
+more \emph{kernels} that run on accelerator devices---GPUs in the
+context of this paper.  A kernel is writen in OpenCL C, based on C99.
+All threads executing a kernel execute the same entry function with
+identical arguments, but a thread may query its unique thread id in
+order to access different data or follow different control flow paths
+to other threads.
+
+Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
+kernel that uses a work stealing queue.  \ADComment{Describe some
+  aspects of the example, saying that the mutex is not a primitive,
+  and can be coded up using the memory model but is not safe due to
+  the execution model.  Make it clear that the lock function uses busy-waiting.}  Figure~\ref{fig:graphsearch} shows the
+skeleton of a graph traversal algorithm.  \ADComment{Describe some
+  aspects of the example, saying that the global barrier is not a
+  primitive, and can be coded up using the memory model but is not
+  safe due to the execution model.  Make it clear that the barrier uses busy-waiting.}
+
+\paragraph{Execution model}
+
+The threads executing a kernel are hierarchically organised.  A kernel
+is executed by a number of \emph{workgroups}, and each workgroup
+consists of a fixed number of \emph{threads}.  Within a workgroup,
+threads may also be organised into \emph{subgroups} (termed
+\emph{warps} and \emph{wavefronts} on NVIDIA and AMD architectures,
+respectively).
+
+Two barrier primitives are provided in OpenCL.  A workgroup barrier
+synchronises all threads in a workgroup and ensures memory
+consistency.  This barrier is a \emph{workgroup-level} function: it
+must be encountered uniformly by all or none of the threads of a
+workgroup, thus can only appear in conditional code if all threads in
+a workgroup that reach the barrier agree on the guards of all
+enclosing conditional statements.  A subgroup barrier works
+analogously at the subgroup level.  As discussed below, a
+\emph{global} barrier (synchronising all threads running a kernel) is
+not provided as a primitive.
+
+OpenCL specifically makes no guarantees about fair scheduling between
+workgroups, stating~\cite{...}: \emph{``A conforming implementation
+  may choose to serialize the workgroups so a correct algorithm cannot
+  assume that workgroups will execute in parallel.  There is no safe
+  and portable way to synchronize across the independent execution of
+  workgroups since once in the work-pool, they can execute in any
+  order.''}  OpenCL is not alone in this regard: CUDA similarly
+provides no guarantees~\cite{...}, and HSA guarantees only that
+\ADComment{Tyler, can you state the HSA guarantee?}~\cite{...}.
+
+Revisiting the examples of Figures~\ref{fig:workstealing}
+and~\ref{fig:graphsearch}: the mutex function of
+Figure~\ref{fig:workstealing} will cause starvation if a workgroup
+that holds the mutex is not scheduled again and so does not release
+the mutex.  A similar issue affects the global barrier function of
+Figure~\ref{fig:graphsearch}.  Because the barrier is implemented
+using busy-waiting, a set of workgroups waiting at the barrier may be
+continuously scheduled, starving a workgroup that has not yet arrived
+at the barrier and meaning that no workgroup proceeds past the
+barrier.
+
+\paragraph{Id functions}
+
+Builtin functions $\getlocalid()$ and $\getglobalid()$ provide a
+thread with an id that is unique within its workgroup and unique
+across all workgroups, respectively.  Each workgroup has an id,
+obtained via $\getgroupid()$.  The number of threads per workgroup and
+the number of workgroups are obtained via $\getlocalsize()$ and
+$\getnumgroups()$.  A thread's local and global id are related via the
+equation $\getglobalid() = \getgroupid() \times \getlocalsize() +
+\getlocalid()$.
+
+In general, an OpenCL kernel is organised as a multi-dimensional grid
+of multi-dimensional workgroups, and the above id functions each take
+an integer dimension argument.  For simplicity, we consider only the
+one-dimensional case throughout the paper, which captures all the
+GPU-accelerated irregular algorithms we are aware of, and use
+e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
+
+\paragraph{Memory model}
+
+An OpenCL kernel has access to four memory spaces.  Shared virtual
+memory (SVM) is accessible to all threads, and to the host
+application.  Atomic operations on this memory space enable
+fine-grained communication between host and device.  Our
+implementation of cooperative kernels
+(Section~\ref{sec:implementation}) depends on this recent OpenCL
+feature.  Global memory is shared among all threads executing a
+kernel, but can only be safely accessed by the host application before
+or after kernel execution.  Local memory is shared among all threads
+in the same workgroup: each workgroup has its own portion of local
+memory, which can be used for fast intra-workgroup communication.
+Each thread has an allocation of very fast private memory, where
+function-local variables are allocated.
+
+Fine-grained communication between threads, and also with the host in
+the case of SVM, is enabled via a set of atomic data types and
+operations, based on the C11 memory model~\cite{...}.
+\ADComment{Perhaps say more.}  \ADComment{Can we illustrate using the
+  examples?}
 
@@
 
-Our first additional language primitive, $\offerkill$, is a
-workgroup-level function: similar to the OpenCL intra-workgroup
-$\mathsf{barrier}$ primitive, if $\offerkill$ appears inside
-conditional code then it must be executed by \emph{all} threads in a
-workgroup, with all threads having executed the same number of
-iterations of any enclosing loops.
+Similar to the OpemncL workgroup $\mathsf{barrier}$ primitive, our first additional language primitive, $\offerkill$, is a
+workgroup-level function (see Section~\ref{sec:background}).
 

### 2016-11-10 write-ups/PLDI/paper.tex
Progress and operational semantics.

@@
 \usepackage{graphicx}
+\usepackage{mathpartir}
 
@@
 
-\newcommand{\restore}{\texttt{restore}}
+\newcommand{\transmit}{\mathsf{transmit}}
 
@@
 \section{Introduction}\label{sec:intro}
-
-\ADComment{Think about whether ``restore'' is best name.}
 
@@
 
-We use two worked examples---work stealing and breadth-first graph
-search---to provide background on the OpenCL programming model and
-irregular data-parallel algorithms.
+We use two examples of irregular data-parallel algorithms---a work stealing queue and breadth-first graph
+search---to introduce the industry-standard OpenCL programming model on which we base cooperative kernels.
 
@@
 
-\paragraph{OpenCL}
+\paragraph{OpenCL kernels}
+
 An OpenCL program is divided into \emph{host} and \emph{device}
@@
 We describe our cooperative kernels programming model as an extension
-to the industry-standard OpenCL programming model for heterogeneous
-many-core architectures.  However, the cooperative kernels concept is
-more general, and could be applied to extend other GPU programming
-models, such as CUDA and HSA.
+to OpenCL.  However, the cooperative kernels concept is more general,
+and could be applied to extend other GPU programming models, such as
+CUDA and HSA.
 
@@
 (described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
-have taken particular semantic decisions where other reasonable
-alternatives were possible (Section~\ref{sec:semanticalternatives}).
+might have taken different, equally reasonable, semantic decisions (Section~\ref{sec:semanticalternatives}).
 We conclude the section by discussing the backwards-compatible nature
@@
 
-\ADComment{A natural exit does not decrement $M$.}
-
-As with a regular OpenCL kernel, a cooperative kernel is launched by
-the host application, with a specified number of workgroups, each
-consisting of a specified number of threads.  The kernel's behaviour
-is described by a kernel function, written in OpenCL C but using the
-additional language extensions that we present below, and a number of
-parameters to the kernel function are passed by the host application.
-
-The parameters to a cooperative kernel are \emph{immutable}, as if
+As with a regular OpenCL kernel (see Section~\ref{sec:background}), a
+cooperative kernel is launched by the host application.  The host
+application passes parameters to the kernel and specifies a desired
+number of workgroups, each consisting of a specified number of
+threads.  The parameters to a cooperative kernel are \emph{immutable}, as if
 they were qualified with the \texttt{const} keyword (though pointer
@@
 
-\ADComment{Perhaps remove this because a driver-based approach could handle this.}
-A regular OpenCL kernel can be executed by a multi-dimensional grid of
-workgroups, called an ND-range.  In our current proposal, the ND-range
-for a cooperative kernel must be one-dimensional.  \ADComment{Say why?
-  And mention that none of the applications that we deem would benefit
-  from cooperative kernels need a multi-dimensional ND-range.}
-
-As in regular OpenCL, each workgroup can be multi-dimensional.
-
-\paragraph{Restored variables}
+A cooperative kernel is written in OpenCL C plus the following
+language extensions: $\transmit$, a
+new qualifier on the local variables of a thread; $\offerkill$ and
+$\offerfork$, the key functions that enable cooperative scheduling;
+$\globalbarrier$ and $\resizingglobalbarrier$, two primitives to allow
+inter-workgroup synchronisation.
+
+We now explain the semantics of these language extensions, after which
+we explain how the work stealing and graph traveral algorithms of
+Section~\ref{sec:background} can be turned into cooperative kernels
+that use the extensions.
+
+\paragraph{Transmitted variables}
 
 A variable declared in the root scope of the cooperative kernel can
-optionally be annotated with a new \restore qualifier.  By
-annotating a variable with \restore, the programmer indicates
-that when new workgroups are spawned via $\offerfork$, the threads in
-these workgroups should obtain a well-defined initial value for the
-variable, obtained from the workgroup that executed $\offerfork$.  We
-detail the semantics for this when we describe $\offerfork$ below.
+optionally be annotated with a new $\transmit$ qualifier.  By
+annotating a variable with $\transmit$, the programmer indicates that
+when a workgroup spawns new workgroups by calling $\offerfork$, the
+workgroup should transmit the current value of this variable to serve
+as an initial value for the new workgroups.  We detail the semantics
+for this when we describe $\offerfork$ below.
 
 \ADComment{Remove?}
-Only private memory variables can be annotated with \restore,
+Only private memory variables can be annotated with $\transmit$,
 which may include pointers to local or global memory so long as the
@@
 particular, variables and arrays in local memory cannot be annotated
-with \restore.  \ADComment{The preceding is what we have
+with $\transmit$.  \ADComment{The preceding is what we have
   implemented, but actually it is pretty clear how we could handle
@@
 
-\item for each variable $v$ annotated with \restore, every new
+\item for each variable $v$ annotated with $\transmit$, every new
   thread's copy of $v$ is initialised to the value that thread 0 in
@@
 
+\paragraph{Examples}
+
+\ADComment{TODO}
+
 \subsection{Programming Guidelines}\label{sec:programmingguidelines}
@@
   written as a cooperative kernel then it will execute just as safely
-  as it does today on current hardware if: (1) the \restore
+  as it does today on current hardware if: (1) the $\transmit$
   annotation and $\offerkill$/$\offerfork$ calls are defined away to
@@
 \section{Notes}
+
+\ADComment{A natural exit does not decrement $M$.  We haven't made a point of this, and I don't think we should right now, but we should remember to think this through carefully.}
 
@@
 
+\section{Semantics}
+
+Thread state: $(l, \mathit{ss})$, where $l \in L$ is the thread's
+local state, and $\mathit{ss}$ is a semi-colon-separated sequence of
+program statements that the thread remains to execute.
+
+Workgroup state: $((l_1, \mathit{ss}_1), \dots, (l_d,
+\mathit{ss}_d))$, where each $(l_i, \mathit{ss}_i)$ is a thread state.
+
+Kernel state: an $N$-tuple $(\sigma, (w_1, w_2, \dots, w_M, \bot,
+\dots, \bot))$, where each $w_i$ is a workgroup state, which are
+followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.
+
+We treat thread-level semantics abstractly by assuming a transition
+relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
+\mathit{ss}'))$ that is defined unless $\mathit{ss} =
+\mathsf{special}(); \mathit{tt}$, where $\mathit{special}$ is one of
+$\offerkill$, $\offerfork$, $\globalbarrier$ or
+$\resizingglobalbarrier$.
+
+To abstractly account for the interaction between global barrier
+synchronization and the GPU memory model, we assume a $\mathsf{sync}$
+that can be applied to the shared state $\sigma$, or to the shared
+state and a thead local state.  In the first case,
+$\mathsf{sync}(\sigma)$ returns an updated shared state that reflects
+all in-flight memory stores having completed.  In the second case,
+$\mathsf{sync}(\sigma, l)$ returns a thread local state reflecting all
+in-flight memory loads issued by the thread having completed.
+
+\begin{figure*}
+\begin{center}
+
+\[
+\inferrule{
+w_i(j) = (l, \mathit{ss})
+\\
+(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l', \mathit{ss}'))
+\\
+w_i' = w_i[j \mapsto (l', \mathit{ss}')]
+}
+{
+(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma', (\dots, w_i', \dots))
+}
+(\textsc{Thread-Step})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_i(j) = (l_j, \offerkill();\mathit{ss})
+\\
+w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+}
+{
+(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
+}
+(\textsc{Kill-No-Op})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
+}
+{
+(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
+}
+(\textsc{Kill})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
+\\
+w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+}
+{
+(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
+}
+(\textsc{Fork-No-Op})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
+\\
+w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+\\
+k \in [0, N - M]
+\\
+\forall a \in [1, k] \;.\; w_{M+a} = ((l_1, \mathit{ss}), \dots, (l_1, \mathit{ss}))
+}
+{
+(\sigma, (\dots, w_i, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_M, w_{M+1}, \dots, w_{M+k}, \bot, \dots, \bot))
+}
+(\textsc{Fork})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
+\\
+\forall i \;.\;\forall j\;.\; l_{i,j}' = \mathsf{sync}(\sigma, l_{i,j})
+\\
+\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}', \mathit{ss})
+\\
+\sigma' = \mathsf{sync}(\sigma)
+}
+{
+(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
+}
+(\textsc{Barrier})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
+\\
+\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \offerkill(); \globalbarrier(); \offerfork();\mathit{ss})
+}
+{
+(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
+}
+(\textsc{Resizing-Barrier})
+\]
+
+\end{center}
+
+\caption{Abstract operational semantics for our cooperative kernels language extensions}
+
+\end{figure*}
+
 \end{document}

### 2016-11-10 write-ups/PLDI/paper.tex
Progress

@@
 
-\paragraph{Id functions}
+\paragraph{Identity and size functions}
 
@@
 In general, an OpenCL kernel is organised as a multi-dimensional grid
-of multi-dimensional workgroups, and the above id functions each take
+of multi-dimensional workgroups, and the above functions each take
 an integer dimension argument.  For simplicity, we consider only the
@@
 Section~\ref{sec:background} can be turned into cooperative kernels
-that use the extensions.
+that use the extensions.  We present the semantics intuitively, using English.
+In Appendix~\ref{appendix:semantics} we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.
 
@@
 when a workgroup spawns new workgroups by calling $\offerfork$, the
-workgroup should transmit the current value of this variable to serve
-as an initial value for the new workgroups.  We detail the semantics
+workgroup should transmit its current value for the variable to serve
+as an initial value for the variable for the threads of the new workgroups.  We detail the semantics
 for this when we describe $\offerfork$ below.
 
-\ADComment{Remove?}
-Only private memory variables can be annotated with $\transmit$,
-which may include pointers to local or global memory so long as the
-pointer variables themselves are stored in private memory.  In
-particular, variables and arrays in local memory cannot be annotated
-with $\transmit$.  \ADComment{The preceding is what we have
-  implemented, but actually it is pretty clear how we could handle
-  local memory restoration if we wanted to; it's just that none of our
-  examples needed it.}
-
 \paragraph{Active workroups}
 
-If the host application launches a cooperative kernel specifying that
-$N$ workgroups should execute the kernel, then when the kernel
-commences execution it should be executed by some number of workgroups
-$M$, with $1\leq M \leq N$.  The requested number of workgroups is
-called the \emph{workgroup limit} for the cooperative kernel, and $M$
-is the number of \emph{active workgroups}.  As explained below, the
-number $M$ of active workgroups may change during the lifetime of a
-cooperative kernel.
+If the host application launches a cooperative kernel requesting $N$
+workgroups, this indicates that the kernel should be executed with a
+maximum of $N$ workgroups, and that as many workgroups as possible, up
+to this limit, are desired.  However, the scheduler may initially
+schedule fewer than $N$ workgroups, and as explained below the number
+of workgroups that execute the cooperative kernel can change during
+the lifetime of the kernel.
+
+The workgroups executing the kernel are called \emph{active
+  workgroups}, and $M$ denotes the number of active workgroups.
+Initially, the number of active workgroups $M$ is initialised to some
+integer between $1$ and $N$.  The initial active workgroups have
+consecutive workgroup ids in the range $[0, M-1]$.
+
+The $\getgroupid()$ function retrieves the id of a workgroup as usual
+(see Section~\ref{sec:background}).  When executed by a cooperative
+kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
+active workgroups.  This is in contrast to $\getnumgroups()$ for
+regular kernels, which returns the fixed number of workgroups that
+were requested at kernel launch time.
 
@@
 
-The initial workgroups have consecutive workgroup ids in the range
-$[0, M-1]$.
-
-The OpenCL primitive $\getgroupid{0}$ retrieves the id of a
-workgroup as usual.  When executed by a cooperative kernel, the OpenCL
-primitive $\getnumgroups{0}$ returns $M$, the current number
-of active workgroups.  Because the ND-ranges of cooperative kernels
-are one-dimensional, $\getgroupid{\mathit{i}}$ and
-$\getnumgroups{\mathit{i}}$ return 0 and 1, respectively, for $i >
-0$.
-
-\ADComment{What happens to $M$ when a workgroup reaches the end of a
-  cooperative kernel?  Feels like $M$ should decrease, but what does
-  this mean for barrier-only kernels?  Do they need to account for the
-  number of active workgroups changing during the final barrier
-  interval?  An alternative is to say that there is an implicit global barrier at kernel exit.}
-
 \paragraph{Semantics for $\offerkill$}
 
-\ADComment{Don't allow zero to be killed.}
-
-\ADComment{Explain that the ``last one dies'' semantics is convenient because it maintains contiguous ids, but other options are possible and may be more efficient sometimes.}
-
-Similar to the OpemncL workgroup $\mathsf{barrier}$ primitive, our first additional language primitive, $\offerkill$, is a
-workgroup-level function (see Section~\ref{sec:background}).
-
-Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$,
-i.e.\ the workgroup's id is not the largest among the active
-workgroups, then $\offerkill$ is a no-op.  If instead $m = M-1$, there
-is a non-deterministic choice between (a) $\offerkill$ executing as a
-no-op, or (b) execution of the workgroup ceasing and the number of
-active workgroups $M$ being decremented by one.  Case (b) corresponds
-to the scheduler accepting the workgroup's offer to cease execution,
-allowing the scheduler to schedule some other task on the compute unit
-on which the workgroup was executing.
+The $\offerkill$ primitive allows the cooperative kernel to return
+compute units to the scheduler by offering to sacrifice workgroups.
+The idea is that while having the scheduler be permitted to terminate
+the execution of workgroups in an arbitary fashion would be drastic,
+the kernel may contain identifiable points of execution at which a
+workgroup could gracefully leave the computation.  We give examples of
+these for work stealing and graph traversal below.
+
+Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
+$\offerkill$, is a workgroup-level function---it must be encountered
+uniformly by all threads in a workgroup (see
+Section~\ref{sec:background}).
+
+Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$
+or $M=1$, i.e.\ the workgroup's id is not the largest among the active
+workgroups, or if only a single workgroup is executing the cooperative
+kernel, then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
+M-1$, there is a non-deterministic choice between (a) $\offerkill$
+executing as a no-op, or (b) execution of the workgroup ceasing and
+the number of active workgroups $M$ being decremented by one.  Case
+(b) corresponds to the scheduler accepting the workgroup's offer to
+cease execution, allowing the scheduler to schedule some other task on
+the compute unit on which the workgroup was executing.
 
@@
 each workgroup executes $\offerkill$ in some arbitrary sequential
-order.  \ADComment{Should we say something about linearisability?}
+order.
+
+In this formulation of cooperative kernels, only the active workgroup
+with the largest id can be killed, and workgroup 0 can never be
+killed.  We discuss the rationale for these choices, as well as
+possible alternative choices, in
+Section~\ref{sec:semanticalternatives}.
 
@@
 
-Our second new language primitive, $\offerfork$, is also a workgroup-level
-function.
+The $\offerfork$ primitive allows the cooperative kernel to indicate
+to the scheduler that it could benefit from, and is prepared to
+handle, additional workgroups joining the computation.  Recall that
+the cooperative kernel was launched with a limit of $N$ workgroups,
+indicating the number of workgroups that would ideally execute the
+kernel.  However, the number of active workgroups, $M$, may be smaller
+than $N$, either because (due to competing workloads) the scheduler
+did not provide $N$ workgroups initially, or because the kernel has
+given up some number of workgroups via $\offerkill$ calls.  Through
+$\offerfork$, the kernel and scheduler can work together to allow new
+workgroups to join the computation at an appropriate point and with
+appropriate state.
+
+Like $\offerkill$, $\offerfork$ is a workgroup-level function.
 
@@
 integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
-$k$ new workgroups are spawned with consecutive ids in the range
-$[M, M+k-1]$, and the active workgroup count $M$ is incremented by
-$k$.
+$k$ new workgroups are spawned with consecutive ids in the range $[M,
+  M+k-1]$, and the active workgroup count $M$ is incremented by $k$.
 
 The $k$ new workgroups commence execution at the program point
-immediately following the $\offerfork$ call.  The private variables of
-each thread in a new workgroup are all uninitialised, as are any local
-variables and arrays shared across the workgroup; reading from these
-locations without initialising them is an undefined behaviour.  There
-are two exceptions to this:
+immediately following the $\offerfork$ call.  The variables that
+describe the state of a thread are all uninitialised for the threads
+in the new workgroups; reading from these variables without first
+initialising them is an undefined behaviour.  There are two exceptions
+to this:
 
@@
 \item because the parameters to a cooperative kernel are immutable,
-  the new threads have access to these parameters and can safely read
-  from them;
+  the new threads have access to these parameters as part of their
+  local state and can safely read from them;
 
@@
 \end{itemize}
+
+In effect, thread 0 of the spawning workgroup transmits the relevant
+portion of its local state to the new threads.  We discuss in
+Section~\ref{sec:semanticalternatives} the rationale for having all
+new threads obtain this local state portion from thread 0, as well as
+our reasons for selecting which variables to transmit via annotations
+rather than transmitting the entire thread state.
 
@@
 guaranteed.
-
-\ADComment{Remark on our decision to have every thread inherit the
-  state of the master thread from workgroup $m$.}
 
@@
 
+\ADComment{Allowing arbitrary workgroups to be killed (not just the largest workgroup)}
+
+\ADComment{Having thread 0 transmit its state, and transmitting only annotated variables.}
+
 \subsection{Backwards Compatibility}\label{sec:backwardscompatibility}
@@
   GPUs.}
+
+\appendix
+
+\section{Operational Semantics for Cooperative Kernels}\label{appendix:semantics}
+
+\ADComment{TODO: write it up properly.}
+
+\ADComment{TODO: remark that we do not model transmit faithfully.}
+
+Thread state: $(l, \mathit{ss})$, where $l \in L$ is the thread's
+local state, and $\mathit{ss}$ is a semi-colon-separated sequence of
+program statements that the thread remains to execute.
+
+Workgroup state: $((l_1, \mathit{ss}_1), \dots, (l_d,
+\mathit{ss}_d))$, where each $(l_i, \mathit{ss}_i)$ is a thread state.
+
+Kernel state: an $N$-tuple $(\sigma, (w_1, w_2, \dots, w_M, \bot,
+\dots, \bot))$, where each $w_i$ is a workgroup state, which are
+followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.
+
+We treat thread-level semantics abstractly by assuming a transition
+relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
+\mathit{ss}'))$ that is defined unless $\mathit{ss} =
+\mathsf{special}(); \mathit{tt}$, where $\mathit{special}$ is one of
+$\offerkill$, $\offerfork$, $\globalbarrier$ or
+$\resizingglobalbarrier$.
+
+To abstractly account for the interaction between global barrier
+synchronization and the GPU memory model, we assume a $\mathsf{sync}$
+that can be applied to the shared state $\sigma$, or to the shared
+state and a thead local state.  In the first case,
+$\mathsf{sync}(\sigma)$ returns an updated shared state that reflects
+all in-flight memory stores having completed.  In the second case,
+$\mathsf{sync}(\sigma, l)$ returns a thread local state reflecting all
+in-flight memory loads issued by the thread having completed.
+
+\begin{figure*}
+\begin{center}
+
+\[
+\inferrule{
+w_i(j) = (l, \mathit{ss})
+\\
+(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l', \mathit{ss}'))
+\\
+w_i' = w_i[j \mapsto (l', \mathit{ss}')]
+}
+{
+(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma', (\dots, w_i', \dots))
+}
+(\textsc{Thread-Step})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_i(j) = (l_j, \offerkill();\mathit{ss})
+\\
+w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+}
+{
+(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
+}
+(\textsc{Kill-No-Op})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
+}
+{
+(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
+}
+(\textsc{Kill})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
+\\
+w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+}
+{
+(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
+}
+(\textsc{Fork-No-Op})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
+\\
+w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+\\
+k \in [0, N - M]
+\\
+\forall a \in [1, k] \;.\; w_{M+a} = ((l_1, \mathit{ss}), \dots, (l_1, \mathit{ss}))
+}
+{
+(\sigma, (\dots, w_i, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_M, w_{M+1}, \dots, w_{M+k}, \bot, \dots, \bot))
+}
+(\textsc{Fork})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
+\\
+\forall i \;.\;\forall j\;.\; l_{i,j}' = \mathsf{sync}(\sigma, l_{i,j})
+\\
+\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}', \mathit{ss})
+\\
+\sigma' = \mathsf{sync}(\sigma)
+}
+{
+(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
+}
+(\textsc{Barrier})
+\]
+
+\medskip
+
+\[
+\inferrule{
+\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
+\\
+\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \offerkill(); \globalbarrier(); \offerfork();\mathit{ss})
+}
+{
+(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
+}
+(\textsc{Resizing-Barrier})
+\]
+
+\end{center}
+
+\caption{Abstract operational semantics for our cooperative kernels language extensions}
+
+\end{figure*}
 
@@
 
-\section{Semantics}
-
-Thread state: $(l, \mathit{ss})$, where $l \in L$ is the thread's
-local state, and $\mathit{ss}$ is a semi-colon-separated sequence of
-program statements that the thread remains to execute.
-
-Workgroup state: $((l_1, \mathit{ss}_1), \dots, (l_d,
-\mathit{ss}_d))$, where each $(l_i, \mathit{ss}_i)$ is a thread state.
-
-Kernel state: an $N$-tuple $(\sigma, (w_1, w_2, \dots, w_M, \bot,
-\dots, \bot))$, where each $w_i$ is a workgroup state, which are
-followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.
-
-We treat thread-level semantics abstractly by assuming a transition
-relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
-\mathit{ss}'))$ that is defined unless $\mathit{ss} =
-\mathsf{special}(); \mathit{tt}$, where $\mathit{special}$ is one of
-$\offerkill$, $\offerfork$, $\globalbarrier$ or
-$\resizingglobalbarrier$.
-
-To abstractly account for the interaction between global barrier
-synchronization and the GPU memory model, we assume a $\mathsf{sync}$
-that can be applied to the shared state $\sigma$, or to the shared
-state and a thead local state.  In the first case,
-$\mathsf{sync}(\sigma)$ returns an updated shared state that reflects
-all in-flight memory stores having completed.  In the second case,
-$\mathsf{sync}(\sigma, l)$ returns a thread local state reflecting all
-in-flight memory loads issued by the thread having completed.
-
-\begin{figure*}
-\begin{center}
-
-\[
-\inferrule{
-w_i(j) = (l, \mathit{ss})
-\\
-(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l', \mathit{ss}'))
-\\
-w_i' = w_i[j \mapsto (l', \mathit{ss}')]
-}
-{
-(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma', (\dots, w_i', \dots))
-}
-(\textsc{Thread-Step})
-\]
-
-\medskip
-
-\[
-\inferrule{
-\forall j \;.\; w_i(j) = (l_j, \offerkill();\mathit{ss})
-\\
-w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
-}
-{
-(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
-}
-(\textsc{Kill-No-Op})
-\]
-
-\medskip
-
-\[
-\inferrule{
-\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
-}
-{
-(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
-}
-(\textsc{Kill})
-\]
-
-\medskip
-
-\[
-\inferrule{
-\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
-\\
-w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
-}
-{
-(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
-}
-(\textsc{Fork-No-Op})
-\]
-
-\medskip
-
-\[
-\inferrule{
-\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
-\\
-w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
-\\
-k \in [0, N - M]
-\\
-\forall a \in [1, k] \;.\; w_{M+a} = ((l_1, \mathit{ss}), \dots, (l_1, \mathit{ss}))
-}
-{
-(\sigma, (\dots, w_i, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_M, w_{M+1}, \dots, w_{M+k}, \bot, \dots, \bot))
-}
-(\textsc{Fork})
-\]
-
-\medskip
-
-\[
-\inferrule{
-\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
-\\
-\forall i \;.\;\forall j\;.\; l_{i,j}' = \mathsf{sync}(\sigma, l_{i,j})
-\\
-\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}', \mathit{ss})
-\\
-\sigma' = \mathsf{sync}(\sigma)
-}
-{
-(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
-}
-(\textsc{Barrier})
-\]
-
-\medskip
-
-\[
-\inferrule{
-\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
-\\
-\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \offerkill(); \globalbarrier(); \offerfork();\mathit{ss})
-}
-{
-(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
-}
-(\textsc{Resizing-Barrier})
-\]
-
-\end{center}
-
-\caption{Abstract operational semantics for our cooperative kernels language extensions}
-
-\end{figure*}
-
 \end{document}

### 2016-11-10 write-ups/PLDI/paper.tex
Progress.

@@
 
-In effect, thread 0 of the spawning workgroup transmits the relevant
-portion of its local state to the new threads.  We discuss in
-Section~\ref{sec:semanticalternatives} the rationale for having all
-new threads obtain this local state portion from thread 0, as well as
-our reasons for selecting which variables to transmit via annotations
-rather than transmitting the entire thread state.
+In effect, thread 0 of the forking workgroup transmits the relevant
+portion of its local state to the threads of the forked workgroups.
+We discuss in Section~\ref{sec:semanticalternatives} the rationale for
+having all new threads obtain this local state portion from thread 0,
+as well as our reasons for selecting which variables to transmit via
+annotations rather than transmitting the entire thread state.
 
@@
 
+\ADComment{Maybe refer back to motivating examples?}
 Many irregular algorithms require synchronisation across workgroups.
@@
 Because workgroups of a cooperative kernel are fairly scheduled, a
-global barrier primitive can be provided.  We specify two such primitives: $\globalbarrier()$
-and $\resizingglobalbarrier()$.
-
-A $\globalbarrier()$ is a kernel-level function: if it appears in
+global barrier primitive can be provided.  We specify two such primitives: $\globalbarrier$
+and $\resizingglobalbarrier$.
+
+Our $\globalbarrier$ primitive is a kernel-level function: if it appears in
 conditional code then it must be reached by \emph{all} threads
@@
 
-A $\resizingglobalbarrier()$ is identical to a global barrier, except
-that, semantically, every workgroup executes $\offerkill$ on reaching
-the barrier, and every workgroup executes $\offerfork$ on leaving the
-barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent to:
+The $\resizingglobalbarrier$ primitive is also a kernel-level
+function.  It is identical to $\globalbarrier$, except that it caters
+for cooperation with the scheduler: by issuing a
+$\resizingglobalbarrier$ the programmer indicates that the cooperative
+kernel is prepared to proceed after the barrier with fewer workgroups,
+or with additional workgroups joining the computation.
+Semantically, it is as if every workgroup executes $\offerkill$ on reaching the
+barrier, and every workgroup executes $\offerfork$ on leaving the
+barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent
+to: \ADComment{Does the order of these three calls matter, semantically?}
 %
@@
 
-The $\globalbarrier()$ primitive can be implemented directly using the
-atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
-but the implementation is involved, hence why we include this function
-as a primitive.  As noted above, the $\resizingglobalbarrier()$
-primitive is, semantically, syntactic sugar for a sequence of other
-primitive calls.  We include it as a primitive because (a) we have
-found that global barriers often provide a natural point for
-cooperation with the scheduler, and (b) a custom implementation of
-$\resizingglobalbarrier()$ that provides the semantics described
-above can be made significantly more efficient than the above sequence of
+The $\globalbarrier$ primitive can be implemented following any
+inter-workgroup barrier design, e.g.~\cite{...}, and the atomic
+operations provided by the OpenCL 2.0 memory model enable a
+memory-safe implementation~\cite{...}.  However, implementing such a
+barrier is involved, hence why we include this function as a
+primitive.  As noted above, the $\resizingglobalbarrier$ primitive
+is, semantically, syntactic sugar for a sequence of other primitive
+calls.  We include it as a primitive because (a) we have found that
+global barriers often provide a natural point for cooperation with the
+scheduler, and (b) a custom implementation of
+$\resizingglobalbarrier()$ that provides the semantics described above
+can be made significantly more efficient than the above sequence of
 calls.  We discuss our efficient implementation of
-$\resizingglobalbarrier$ in detail in Section~\ref{sec:resizingbarrier}.
+$\resizingglobalbarrier$ in detail in
+Section~\ref{sec:resizingbarrier}.
 
@@
 
-\ADComment{TODO}
+\ADComment{TODO.  We want to show the examples from Section~\ref{sec:background} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}
 
@@
 attention to the following details.
-
-\paragraph{Keeping at least one workgroup alive}
-
-The programmer is responsible for keeping at least one workgroup alive.
-In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.
-This can be guarded against by wrapping each $\offerkill$ call in a conditional statement:
-
-\lstset{basicstyle=\tt}
-\begin{lstlisting}
-  if($\getgroupid{0}$ != 0) $\offerkill$();
-\end{lstlisting}
-\lstset{basicstyle=\scriptsize\tt}
-
-We could, alternatively, have formulated the semantics of $\offerkill$
-so that workgroup 0 is never killed.  However, there might be a use
-case for a cooperative kernel reaching a point where it would be
-acceptable for the kernel to exit, although desirable for some
-remaining computation to be performed if competing workloads allow it.
 
@@
 
-\ADComment{Allowing zero to be killed.}
+\ADComment{Allowing zero to be killed; the following is some text we had before:}
+
+\paragraph{Keeping at least one workgroup alive}
+
+The programmer is responsible for keeping at least one workgroup alive.
+In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.
+This can be guarded against by wrapping each $\offerkill$ call in a conditional statement:
+
+\lstset{basicstyle=\tt}
+\begin{lstlisting}
+  if($\getgroupid{0}$ != 0) $\offerkill$();
+\end{lstlisting}
+\lstset{basicstyle=\scriptsize\tt}
+
+We could, alternatively, have formulated the semantics of $\offerkill$
+so that workgroup 0 is never killed.  However, there might be a use
+case for a cooperative kernel reaching a point where it would be
+acceptable for the kernel to exit, although desirable for some
+remaining computation to be performed if competing workloads allow it.
 

### 2016-11-10 write-ups/PLDI/paper.tex
Edits with Hugues.

@@
 
-Our $\globalbarrier$ primitive is a kernel-level function: if it appears in
-conditional code then it must be reached by \emph{all} threads
-executing the cooperative kernel.  On reaching a $\globalbarrier()$, a
-thread waits until all threads have arrived at the barrier.  Once all
-threads have arrived, the threads may proceed past the barrier with
-the guarantee that all global memory accesses issued before the
-barrier have completed.
+Our $\globalbarrier$ primitive is a kernel-level function: if it
+appears in conditional code then it must be reached by \emph{all}
+threads executing the cooperative kernel.  On reaching a
+$\globalbarrier()$, a thread waits until all threads have arrived at
+the barrier.  Once all threads have arrived, the threads may proceed
+past the barrier with the guarantee that all global memory accesses
+issued before the barrier have completed.  The $\globalbarrier$
+primitive can be implemented following any inter-workgroup barrier
+design, e.g.~\cite{...}, and the atomic operations provided by the
+OpenCL 2.0 memory model enable a memory-safe
+implementation~\cite{...}.  However, implementing such a barrier is
+involved, hence why we include this function as a primitive.
 
@@
 or with additional workgroups joining the computation.
-Semantically, it is as if every workgroup executes $\offerkill$ on reaching the
-barrier, and every workgroup executes $\offerfork$ on leaving the
-barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent
-to: \ADComment{Does the order of these three calls matter, semantically?}
-%
+
+Semantically, when all threads have reached $\resizingglobalbarrier$,
+the number of active workgroups, $M$, is set to a new value, $M'$ say.
+If $M' = M$ then the active workgroups remain unchanged and proceed to
+execute after the barrier.  If $M' < M$, workgroups $[M', M-1]$ are
+killed.  If $M' > M$ then $M'-M$ new workgroups join the computation,
+as if they were forked from workgroup 0.  In particular, the
+$\transmit$-annotated local state of thread 0 in workgroup 0 is
+transmitted to the threads of the new workgroups.  The change in
+number of active workgroups from $M$ to $M'$ is atomic.
+
+The semantics of $\resizingglobalbarrier$ can be modelled via the following sequence of calls:
+
 \lstset{basicstyle=\tt}
 \begin{lstlisting}
+  $\globalbarrier()$;
+  if($\getgroupid{0}$ != 0) $\offerfork$();
+  $\globalbarrier()$;
   $\offerkill()$;
   $\globalbarrier()$;
-  $\offerfork()$;
 \end{lstlisting}
@@
 
-The $\globalbarrier$ primitive can be implemented following any
-inter-workgroup barrier design, e.g.~\cite{...}, and the atomic
-operations provided by the OpenCL 2.0 memory model enable a
-memory-safe implementation~\cite{...}.  However, implementing such a
-barrier is involved, hence why we include this function as a
-primitive.  As noted above, the $\resizingglobalbarrier$ primitive
-is, semantically, syntactic sugar for a sequence of other primitive
-calls.  We include it as a primitive because (a) we have found that
-global barriers often provide a natural point for cooperation with the
-scheduler, and (b) a custom implementation of
-$\resizingglobalbarrier()$ that provides the semantics described above
-can be made significantly more efficient than the above sequence of
-calls.  We discuss our efficient implementation of
-$\resizingglobalbarrier$ in detail in
-Section~\ref{sec:resizingbarrier}.
+The enclosing $\globalbarrier$ calls make the operation appear to be
+atomic.  The middle $\globalbarrier$ is required to ensure that if
+$M'\geq M$, the workgroups with ids smaller than $M$ are left intact.
+
+Because $\resizingglobalbarrier$ can be implemented in the above
+manner, we do not regard it conceptually as a primitive of the
+cooperative kernels programming model.  However, we discuss in
+Section~\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
+implemented significantly more efficiently if it is treated as a
+primitive operation that interacts directly with the scheduler.
 
@@
 
-An OpenCL programmer writing a cooperative kernel should pay careful
-attention to the following details.
-
-\paragraph{A changing number of workgroups}
-
-In regular OpenCL, $\getnumgroups{0}$ returns a runtime-constant
-value: once the number of executing workgroups has been identified, it
-is guaranteed not to change.
-
-In contrast, the number of workgroups executing a cooperative kernel
-can vary due to workgroups executing $\offerkill$, $\offerfork$ and
-$\resizingglobalbarrier$.  The programmer must therefore write their
-cooperative kernel in a manner that is robust to changes in the value
-returned by $\getnumgroups{0}$.
+We remark on some issues that programmer working with cooperative
+kernels should be aware of.
+
+\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
+the value returned by $\getnumgroups()$ is not fixed during the
+lifetime of a cooperative kernel.  Although $\getnumgroups()$ has
+atomic semantics, and so can be called at any time to yield a
+well-defined value, the value returned by two successive calls may
+differ if workgroups have executed $\offerkill$, $\offerfork$ or
+$\resizingglobalbarrier$ in-between.  The programmer must therefore
+write their cooperative kernel in a manner that is robust to changes
+in the value returned by $\getnumgroups()$.
 
@@
 
-\ADComment{Need to account for the issue that $M$ might decrease as
-  workgroups actually exit the kernel.}
-
-If a cooperative kernel does not directly issue $\offerkill$ and
-$\offerfork$ calls then the only possibility for the number of
+The situation is more stable if a cooperative kernel does not directly
+issue $\offerkill$ and $\offerfork$ calls directly.
+
+ then the only possibility for the number of
 workgroups to change is due to $\resizingglobalbarrier$ being invoked.
@@
 
-Kernel state: an $N$-tuple $(\sigma, (w_1, w_2, \dots, w_M, \bot,
-\dots, \bot))$, where each $w_i$ is a workgroup state, which are
+Kernel state: a pair $(\sigma, (w_1, w_2, \dots, w_M, \bot, \dots, \bot))$, where $\sigma$ is the shared state,
+each $w_i$ is a workgroup state, which are
 followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.
@@
 \mathit{ss}'))$ that is defined unless $\mathit{ss} =
-\mathsf{special}(); \mathit{tt}$, where $\mathit{special}$ is one of
+\mathsf{special}(); \mathit{tt}$, where $\mathsf{special}$ is one of
 $\offerkill$, $\offerfork$, $\globalbarrier$ or
@@
 To abstractly account for the interaction between global barrier
-synchronization and the GPU memory model, we assume a $\mathsf{sync}$
-that can be applied to the shared state $\sigma$, or to the shared
-state and a thead local state.  In the first case,
-$\mathsf{sync}(\sigma)$ returns an updated shared state that reflects
-all in-flight memory stores having completed.  In the second case,
-$\mathsf{sync}(\sigma, l)$ returns a thread local state reflecting all
-in-flight memory loads issued by the thread having completed.
+synchronization and the GPU memory model, we assume a function
+$\mathsf{sync}$ that is applied to a global state.  This returns a new
+global state in which the shared state and thread local states have
+been updated to reflect the fact that all stores to memory and loads
+from memory to thread-local storate have completed.
 
@@
 (\textsc{Kill})
-\]
-
-\medskip
-
-\[
-\inferrule{
-\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
-\\
-w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
-}
-{
-(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
-}
-(\textsc{Fork-No-Op})
 \]
@@
 \\
-\forall i \;.\;\forall j\;.\; l_{i,j}' = \mathsf{sync}(\sigma, l_{i,j})
-\\
-\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}', \mathit{ss})
-\\
-\sigma' = \mathsf{sync}(\sigma)
+\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \mathit{ss})
 }
 {
-(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
+(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow \mathsf{sync}(\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
 }
@@
 \\
-\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \offerkill(); \globalbarrier(); \offerfork();\mathit{ss})
+\forall i \neq 0 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
+\\
+\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
 }

### 2016-11-10 write-ups/PLDI/paper.tex
Tweak.
Fixes suggested by Hugues.
Add work-stealing example

@@
 %\newcommand{\TSComment}[1]{\textcolor{blue}{TS: #1}}
+\newcommand{\HEComment}[1]{\textcolor{orange}{HE: #1}}
 %\newcommand{\COULDCUT}[1]{\textcolor{blue}{#1}}
@@
 \ADComment{Tyler/Hugues: please insert workstealing example.}
+
+\HEComment{standard version: one task pool per workgroup, the number of
+workgroup is a constant. So we access our pool by pool\_id ==
+get\_group\_id(0). When we try to steal, we can look into other pools
+which are indexed by (pool\_id + delta) modulo get\_num\_workgroups(0).
+}
+
 \begin{lstlisting}
-TODO
+__kernel connect_four(__global uchar *base_board,
+                      const int lookahead,
+                      __global atomic_int *best_move,
+                      __global int *all_done,
+                      __global Node *tree,
+                      __global Task *pools,
+                      __global Mutex *pool_m)
+{
+  // init
+  if (get_global_id(0) == 0) {
+    init_first_tasks(tree, pools);
+  }
+  global_barrier();
+
+  // main loop
+  while (true) {
+    // obtain a task, or terminate if all work is done
+    int pool_id = get_group_id(0);
+    Task *t = pop_or_steal(pools, pool_m, pool_id);
+    if (t == NULL) {
+      if (*all_done) { return; } else { continue; }
+    }
+
+    // perform task, may create new tasks
+    Node *node = tree[t->node_index];
+    value = node_value(base_board, node);
+    if (winning(value) || node->level == lookahead) {
+      // we reached a leaf
+      update_parent_value(node, tree, best_move, all_done);
+    } else {
+      // create_children() pushes or donates new tasks
+      create_children(node, tree, pools, pool_m, pool_id);
+    }
+  }
+}
+\end{lstlisting}
+
+\HEComment{ cooperative version: add kill/fork, and must pass the number
+of task pools explicitely since get\_num\_workgroups() cannot be used to
+know the number of pools. }
+
+\begin{lstlisting}
+__kernel connect_four(__global uchar *base_board,
+                      const int lookahead,
+                      __global atomic_int *best_move,
+                      __global int *all_done,
+                      __global Node *tree,
+                      __global Task *pools,
+                      __global Mutex *pool_m,
+                      const int num_pools)
+{
+  // init
+  if (get_global_id(0) == 0) {
+    init_first_tasks(tree, pools, num_pools);
+  }
+  global_barrier();
+
+  // main loop
+  while (true) {
+    offer_kill();
+    offer_fork();
+
+    // obtain a task, or terminate if all work is done
+    int pool_id = get_group_id(0) % num_pools;
+    Task *t = pop_or_steal(pools, pool_m, pool_id, num_pools);
+    if (t == NULL) {
+      if (*all_done) { return; } else { continue; }
+    }
+
+    // perform task, may create new tasks
+    Node *node = tree[t->node_index];
+    value = node_value(base_board, node);
+    if (winning(value) || node->level == lookahead) {
+      // we reached a leaf
+      update_parent_value(node, tree, best_move, all_done);
+    } else {
+      // create_children() pushes or donates new tasks
+      create_children(node, tree, pools, pool_m, pool_id, num_pools);
+    }
+  }
+}
 \end{lstlisting}
@@
 \\
+\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
+\\
 \forall i \neq 0 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
-\\
-\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
 }

### 2016-11-10 src/host_skeleton/skel.cpp
Typo: fix a comment

@@ -237,7 +237,7 @@ int amd_check(int v) {
 	return v;
 }
 
-// Just for running the non persistent task
+// Just for running the persistent task
 void run_persistent(CL_Execution *exec) {
 	printf("Running persistent app %s\n", persistent_app_name());
 	printf("Running %d iterations\n", FLAGS_run_persistent);
@@ -247,7 +247,7 @@ void run_persistent(CL_Execution *exec) {
 		                           file::Path(FLAGS_restoration_ctx_path),
 		                           FLAGS_use_query_barrier);
 
-	
+
 	check_ocl(err);
 	exec->exec_kernels["persistent"] = cl::Kernel(exec->exec_program, persistent_kernel_name(), &err);
 	check_ocl(err);
@@ -259,15 +259,15 @@ void run_persistent(CL_Execution *exec) {
 	arg_index = set_persistent_app_args_for_occupancy(arg_index, exec->exec_kernels["persistent"]);
 	err = exec->exec_queue.flush();
 	check_ocl(err);
-	
+
 	cl::Buffer d_bar(exec->exec_context, CL_MEM_READ_WRITE, sizeof(IW_barrier));
 	reset_barrier(exec, d_bar);
-	
+
 	err = exec->exec_kernels["persistent"].setArg(arg_index, d_bar);
 	arg_index++;
 	check_ocl(err);
 
-	
+
 
 	cl::Buffer d_ctx_mem(exec->exec_context, CL_MEM_READ_WRITE, sizeof(Discovery_ctx));
 	reset_discovery(exec, d_ctx_mem, true);

### 2016-11-10 write-ups/PLDI/paper.tex
Structure.

--- 2016-11-10 write-ups/PLDI/paper.tex
@@
 The situation is more stable if a cooperative kernel does not directly
-issue $\offerkill$ and $\offerfork$ calls directly.
-
- then the only possibility for the number of
-workgroups to change is due to $\resizingglobalbarrier$ being invoked.
-
-\ADComment{Shorten sentences.}
-In this case, because the resizing barrier is a kernel-level function,
-the statements executed by the cooperative kernel can be divided into
-\emph{resizing barrier intervals}: a resizing barrier interval starts
-at a resizing barrier or the kernel entry point, and ends at a
-resizing barrier or a kernel exit point.  At any moment of execution,
-all threads are guaranteed to be executing within the same resizing
-barrier interval.  As a result, the number of workgroups executing the
-kernel is guaranteed to remain unchanged within each resizing barrier
-interval, so the result returned by $\getnumgroups{0}$ is guaranteed
-to be identical on repeated calls during such an interval.
-
-This guarantee may be exploited by algorithms that perform strided data
-processing.  \ADComment{Tyler, can we give an example of a data
+issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
+only calls to $\resizingglobalbarrier$ can cause the number of active
+workgroups to change and, as discussed in Section~\ref{sec:semantics},
+this function causes synchronisation between all threads and causes
+the number of active workgroups to change atomically.  At any point
+during execution, the threads of a kernel are executing between some
+pair of resizing barrier calls, which we call a \emph{resizing barrier
+  intervals} (here we consider the kernel entry and exit points
+conceptually to be special cases of resizing barriers).  The number of
+workgroups executing the kernel is thus constant within each resizing
+barrier interval, i.e.\ the result returned by repeated calls to
+$\getnumgroups()$ will be identical during such an interval.
+
+This guarantee can be exploited by algorithms that perform strided
+data processing.  \ADComment{Tyler, can we give an example of a data
   processing loop that would have to be changed?}
@@
 
-\ADComment{TODO: needs work.}
-.  The irregular
-algorithms we encountered either employ work stealing or operate on
-graph data structures.  Work stealing algorithms have a transactional
-flavour, whereby tasks are repeatedly pulled from a queue and
-processed by a workgroup.  The point of task completion---the end of a transaction---is a natural place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to indicate that
+It is up to the programmmer to decide where to place our new language
+constructs.  This must be done bearing in mind both the semantics of
+the constructs, as detailed in Section~\ref{ }, as well as the
+non-functional requirements of cooperative kernels that we describe in
+Section~\ref{sec:nonfunctional}.
+
+We note, however, that the programming model and associated language
+constructs were inspired by the requirements of the irregular GPU
+algorithms proposed in the literature, and that at least for these
+algorithms it is usually easy to identify where to place the
+constructs.
+
+The irregular algorithms we encountered either employ work stealing or
+operate on graph data structures.
+
+Work stealing algorithms have a transactional flavour, whereby tasks
+are repeatedly pulled from a queue and processed by a workgroup.  The
+point of task completion---the end of a transaction---is a natural
+place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to
+indicate both that the workgroup could gracefully exit computation,
+but also that there is more work to be processed and that further
+workgroups would be useful in order to process this work.  Such
+algorithms usually require little state to persist between tasks;
+typically only the pointers to the queue data structures need to be
+pre-initialised, by annotating them with $\transmit$.
+\ADComment{Refer back to cooperative versions of examples in
+  Section~\ref{sec:semantics}.}
+
+The level-by-level approach employed by many graph algorithms tends to
+involve global barrier synchronisation.  In some cases, a global
+barrier is required in the midst of a transaction, in which case
+$\globalbarrier$, but not $\resizingglobalbarrier$, is suitable for
+achieving synchronization without disturbing the set of active
+workgroups.  But often it is the case that on completing a level of
+the graph algorithm, the next level could be processed by more or
+fewer workgroups, something which $\resizingglobalbarrier$
+facilitates.  Again, only a small amount of state is relevant to new
+workgroups joining the computation, such as the value of the level
+being processed; this relevant state must be annotated with
+$\transmit$.
 
@@
 for.
-
+%
 However, the aim of cooperative kernels is to find a balance that
@@
 
-\ADComment{Among other things, make sure we address the point that placing the primitives was pretty easy.}
+\ADComment{Describe the algorithms that we ported.}
+
+\ADComment{Give stats on how many variables had to be marked with
+  $\transmit$ (i.e., occurred in the restoration context), how many barriers we changed to resizing, how many offer kill and offer fork calls we included.}
+
+\ADComment{Make sure we address the point that placing the primitives was pretty easy.}
+
+\section{The GPUs we Tested}
+
+\ADComment{Describe the platforms.  Explain why we could not try other
+  platforms.}
+
+\section{Determining Representative Workloads}
+
+\ADComment{Better section title?}
+
+\ADComment{Hugues, can you please explain your experimental
+  methodology for working out required response times here?}
+
+\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
+
+\paragraph{Results}  \ADComment{Discuss the results.}
+
+\section{The Overhad of Moving to Cooperative Kernels}
+
+\ADComment{Results for stand-alone vs.\ cooperative without co-scheduling.}
+
+\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
+
+\paragraph{Results}  \ADComment{Discuss the results.}
+
+\section{Performance and Responsiveness of Cooperative Scheduling}
+
+\ADComment{Preamble}
+
+\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
+
+\paragraph{Results}  \ADComment{Discuss the results.}
 
@@
 \forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
+\\
+$M > 0$
 }
@@
 \\
+\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\mathit{ss})
+\\
 \forall i \neq 0 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
-\\
-\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
 }

### 2016-11-10 src/run_test.py
Better naming of log files

@@ -5,6 +5,13 @@ import subprocess
 import shutil
 import glob
 
+###############################################################################
+# TODO:
+# - remove printing to file since we have summary
+# - use regexp to grab occupancy of mergedskiptask
+# - reuse occupancy when running standalone
+###############################################################################
+
 def fake_placeholder(x):
     return
 
@@ -17,9 +24,9 @@ DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""
 
 PROGRAMS = {
-    "pannotia_color",
+    # "pannotia_color",
     # "pannotia_mis",
-    # "pannotia_bc",
+    "pannotia_bc",
     # "pannotia_sssp"
 }
 
@@ -28,9 +35,9 @@ PROGRAM_DATA = {
     "pannotia_color" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
                            "solution" : os.path.join("solutions", "color_ecology.txt"),
                            "stat" : "color_ecology" },
-                         # { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
-                         #   "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
-                         #   "stat" : "color_G3_circuit" }
+                         { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
+                           "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
+                           "stat" : "color_G3_circuit" }
     ],
 
     "pannotia_mis" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
@@ -43,9 +50,9 @@ PROGRAM_DATA = {
     "pannotia_bc" : [ { "input" : os.path.join("inputs", "bc", "1k_128k.gr"),
                         "solution" : os.path.join("solutions", "bc_1k_128k.txt"),
                         "stat" : "bc_1k_128k" },
-                      { "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
-                        "solution" : os.path.join("solutions", "bc_2k_1M.txt"),
-                        "stat" : "bc_2k_1M" }
+                      # { "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
+                      #   "solution" : os.path.join("solutions", "bc_2k_1M.txt"),
+                      #   "stat" : "bc_2k_1M" }
[…]

### 2016-11-10 write-ups/PLDI/paper.tex
Fix.
Started to add references.

@@
 
-\section{The GPUs we Tested}
+\subsection{The GPUs we Tested}
 
@@
 
-\section{Determining Representative Workloads}
+\subsection{Determining Representative Workloads}
 
@@
 
-\section{The Overhad of Moving to Cooperative Kernels}
+\subsection{The Overhad of Moving to Cooperative Kernels}
 
@@
 
-\section{Performance and Responsiveness of Cooperative Scheduling}
+\subsection{Performance and Responsiveness of Cooperative Scheduling}
 
@@
 \section{Related Work}\label{sec:relatedwork}
+
+\cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
+  SMs, not to workgroups, so that tasks taht would benefit from being
+  processed by the same SM can be scheduled to the same SM.  Has
+  little to do with preemption, but more to do with irregular
+  computation.}
+
+\cite{DBLP:conf/isca/TanasicGCRNV14} \ADComment{I have not read, but
+  it's about enabling preemptive multiprogramming, so we should read
+  it carefully.}
+
+\cite{DBLP:conf/ppopp/Muyan-OzcelikO16} \ADComment{I have not read; we
+  should understand this one carefully.}
+
+\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
+  implement preemption, i.e.\ shows that it can be achieved.
+  Simulator only preemption.  Not sure whether their approach would
+  give fair scheduling.}
+
+\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
+  keep a GPU busy by flooding it with work.  Would absolutely not
+  stand up to preemption.}
+
+\cite{DBLP:conf/usenix/KatoLRI11} \ADComment{Idea seems to be to
+  accept that GPUs do not offer epreemption, but find a reasonable way
+  to schedule tasks to maximise responsiveness or throughput.}
 
@@
   GPUs.}
+
+\clearpage
+
+\bibliographystyle{abbrvnat}
+\bibliography{tyler}
 

### 2016-11-10 write-ups/PLDI/paper.tex
Chat with Tyler and Hugues.

@@
 
+\ADComment{Points to make are: there is spinning, either on mutex implementation or main loop, so fairness is required.
+Note OpenCL features: kernel keyword, global qualifier, ``get'' functions.
+Note that pop\_or\_steal is executed by all, but internally the master manipulates the queue.
+t is a single task, but is chunky enough that all threads can process part of it.}
+
 \begin{lstlisting}
-__kernel connect_four(__global uchar *base_board,
-                      const int lookahead,
-                      __global atomic_int *best_move,
-                      __global int *all_done,
-                      __global Node *tree,
-                      __global Task *pools,
-                      __global Mutex *pool_m)
-{
-  // init
-  if (get_global_id(0) == 0) {
-    init_first_tasks(tree, pools);
-  }
-  global_barrier();
-
-  // main loop
-  while (true) {
-    // obtain a task, or terminate if all work is done
-    int pool_id = get_group_id(0);
-    Task *t = pop_or_steal(pools, pool_m, pool_id);
-    if (t == NULL) {
-      if (*all_done) { return; } else { continue; }
-    }
-
-    // perform task, may create new tasks
-    Node *node = tree[t->node_index];
-    value = node_value(base_board, node);
-    if (winning(value) || node->level == lookahead) {
-      // we reached a leaf
-      update_parent_value(node, tree, best_move, all_done);
-    } else {
-      // create_children() pushes or donates new tasks
-      create_children(node, tree, pools, pool_m, pool_id);
+kernel work_stealing(global Task *pools) {
+  while (more_work(pools)) {
+    int pool_id = get_group_id();
+    Task *t = pop_or_steal(pools, pool_id);
+    if (t) {
+      // task processing may create more work
+      process_task(t, pools, pool_id);
     }
@@
   $\globalbarrier()$;
-  if($\getgroupid{0}$ != 0) $\offerfork$();
+  if($\getgroupid()$ == 0) $\offerfork$();
   $\globalbarrier()$;
@@
 
-\ADComment{Explain kernel merge tool.}
+\ADComment{Explain kernel merge tool briefly.  Explain library briefly.}
+
+\ADComment{Give most attention to interesting performance and safety issues, e.g.\ reverse ticket lock, efficient barrier, the query functions...}
+
+\ADComment{Host-device interaction through SVM.  Make clear that this is probably the most aggressive use of OpenCL 2.0 ever.}
 
@@
 \ADComment{Optimised barrier implementation.}
+
+\subsection{Vendor Implementation Defects}
+
+\ADComment{Briefly rant about our pain.
+Compiler hang, compiler crashes, deadlocks, defunct processes.
+Caution: don't make it sound so terrible that you can't trust our results.
+Gently acknowledge that our implementation may be imperfect.}
 
@@
   platforms.}
+
+\ADComment{Tyler and Hugues, please write.}
 
@@
 
-To abstractly account for the interaction between global barrier
+To abstractly accounts for the interaction between global barrier
 synchronization and the GPU memory model, we assume a function
@@
 \\
-$M > 0$
+M > 0
 }
@@
 {
-(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow \mathsf{sync}(\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
+(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow \mathsf{sync}(\sigma , (w_1', \dots, w_M', \bot, \dots, \bot))
 }
@@
 \\
-\forall j\;.\;w_0'(j) = (l_{0,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\mathit{ss})
+\forall j\;.\;w_1'(j) = (l_{1,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\mathit{ss})
 \\
-\forall i \neq 0 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
+\forall i \neq 1 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
 }

### 2016-11-10 src/run_test.py
Add retrieval of nb of workgroups from skiptask run to parametrise standalone run

@@ -4,6 +4,7 @@ import os
 import subprocess
 import shutil
 import glob
+import re
 
 ###############################################################################
 # TODO:
@@ -113,6 +114,17 @@ def collect_stats(d, prefix):
     # move all potential stat files to the stats dir
     mv_wildcard(prefix + "*", stat_dir)
 
+def extract_finalsize(filename):
+    finalsize = -1
+    re_finalsize = re.compile(', [0-9]+ occupancy, [0-9]+ final size')
+    with open(filename) as f:
+        for line in f:
+            match = re_finalsize.search(line)
+            if match != None:
+                finalsize = match.group().split()[3]
+                break
+    return finalsize
+
 def run_suite():
     for p in PROGRAMS:
         for d in PROGRAM_DATA[p]:
@@ -125,10 +137,30 @@ def run_suite():
             cmd = cmd + ["--graph_file", graph_in]
             cmd = cmd + ["--graph_solution_file", graph_sol]
             cmd = cmd + ["--threads_per_wg", "128"]
+            # indicate very high number of workgroups to finally obtain occupancy
+            cmd = cmd + ["--num_wgs", "1000"]
             cmd = cmd + ["--skip_tasks", "1"]
             prefix = d["stat"] + "_skiptask"
-            exec_cmd(cmd, prefix)
-            # grab occupancy
+            record_stdout = prefix + "_stdout.txt"
+            exec_cmd(cmd, prefix, record_stdout)
+            # grab finalsize (min(occupancy, nb of workgroups))
+            finalsize = extract_finalsize(record_stdout)
+            collect_stats(d, prefix)
+            if finalsize == -1:
+                PRINT("Could not find finalsize after run of merged skiptask")
+                exit(-1)
+
+            # RUN: standalone
+            cmd = [exe]
+            cmd = cmd + ["--graph_file", graph_in]
+            cmd = cmd + ["--graph_solution_file", graph_sol]
+            cmd = cmd + ["--threads_per_wg", "128"]
+            cmd = cmd + ["--run_persistent", "1"]
+            cmd = cmd + ["--num_wgs", finalsize]
+            prefix = d["stat"] + "_standalone"
+            record_stdout = prefix + "_stdout.txt"
+            exec_cmd(cmd, prefix, record_stdout)
+            collect_stats(d, prefix)
 
             for c in MATMULT_CONFIG:
                 # Merged
@@ -136,30 +168,13 @@ def run_suite():
[…]

### 2016-11-10 write-ups/PLDI/paper.tex
Fixed font size.
added graph traversal example

@@
 %\newcommand{\COULDCUT}[1]{\textcolor{blue}{#1}}
+\newcommand{\TSComment}[1]{\textcolor{purple}{TS: #1}}
 
@@
 \ADComment{Tyler/Hugues: please insert graph traversal example.}
+\TSComment{Hows this?}
+
 \begin{lstlisting}
-TODO
+kernel graph_traversal(global graph *g, 
+                       global nodes *n0, global nodes *n1) {
+
+  int level = 0;
+  global nodes * in_nodes = n0;
+  global nodes * out_nodes = n1;
+
+  int tid = get_global_id();
+  int stride = get_global_size();
+
+  while(in_nodes.size > 0) {
+
+    for (int i = tid; i < nodes.size; i += stride) {
+      process_node(g, in_nodes[i], out_nodes, level);
+    }
+
+    level++;
+    swap(in_nodes, out_nodes);
+    reset(out_nodes);
+    global_barrier();
+  }
+}
 \end{lstlisting}
-\caption{An OpenCL graph search algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
+\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
 \end{figure}

### 2016-11-10 write-ups/PLDI/paper.tex
added sigplanconf.cls
Add description of workstealing

@@
 % 
-\documentclass[nocopyrightspace]{sigplanconf-pldi16}
+\documentclass[nocopyrightspace,10pt]{sigplanconf}
 % \documentclass[pldi-cameraready]{sigplanconf-pldi16}
@@
 
-We use two examples of irregular data-parallel algorithms---a work stealing queue and breadth-first graph
-search---to introduce the industry-standard OpenCL programming model on which we base cooperative kernels.
-
+We use two examples of irregular data-parallel algorithms---a work
+stealing queue and breadth-first graph search---to introduce the
+industry-standard OpenCL programming model on which we base cooperative
+kernels.
+
+\paragraph{OpenCL concepts}
+An OpenCL program is typically executed by many \emph{threads} running
+in parallel. Threads are organized into \emph{workgroups} of equal
+size. The memory is organized in three levels: \emph{global} memory is
+accessible by any thread, \emph{local} memory is shared at the workgroup
+level (i.e., threads of different workgroups cannot access the same
+local memory), and \emph{private} memory is dedicated to thread-level
+storage and cannot be shared between threads. These concepts are
+illustrated over the two examples.
+
+\paragraph{Work stealing}
+Work stealing is generally used for computational work that can be
+splitted into independently processable tasks. The processing of a given
+task may create an arbitrary number of new tasks in such a way that it
+is typically hard to know the total number of tasks in advance. Work
+stealing enables the dynamic balance of task processing over several
+processing units, OpenCL workgroups in our case. Each workgroup has an
+associated task pool from which it obtains tasks to process and to where
+it stores newly created tasks. When a workgroup's pool becomes empty, a
+workgroup---rather than terminating---tries to \emph{steal} a task from
+an other pool. Hence, workgroups keep on processing tasks until all the
+work is done.
+
+\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation of
+a work stealing application. All threads start their execution at the
+function marked with the $\mathsf{kernel}$ keyword. They receive a
+pointer to the task pools that reside in global memory, and which were
+initialized by the host program beforehand to contain the initial
+tasks. Each thread can identify the workgroup it belongs to via the
+function \getgroupid, and OpenCL provides a series of primitives in the
+same vein to access for instance the thread global id, or its local id
+within a workgroup. Here the group id is stored into the variable
+$\mathsf(pool\_id)$ (which is, by default, in private memory) and used
+to access the relevant task pool. The $\mathsf(pop\_or\_steal)$ function
+is called by all threads of a workgroup, although inside only the thread
+with a local id equal to $0$, also called \emph{master thread} of the
+workgroup, will try to pop or steal a task from the pools. The master
+thread uses local memory inside $\mathsf(pop\_or\_steal)$ to ensure that
+the returned value is valid for all threads of its workgroup. If a task
+was obtained, then all threads of the workgroup can participate in the
+processing of the task inside the $\mathsf{process\_task}$
+function. \HEComment{Manage to say that a task is considered big enough
+to benefit from parallel computation} If some new tasks are created
+during the processing, it is again the master thread who will be in
+charge of pushing them to the pool. Although not depicted here,
+concurrent accesses to each pool is guarded by a mutex which is
+implemented by atomic compare and swap operations on global
+memory. Henceforth, master threads may be spinning on mutexes, and
+workgroups are also spinning on the main loop, so a fair scheduling of
+workgroups is desirable.
 \begin{figure}
-\ADComment{Tyler/Hugues: please insert workstealing example.}
-
-\HEComment{standard version: one task pool per workgroup, the number of
-workgroup is a constant. So we access our pool by pool\_id ==
-get\_group\_id(0). When we try to steal, we can look into other pools
-which are indexed by (pool\_id + delta) modulo get\_num\_workgroups(0).
-}
-
-\ADComment{Points to make are: there is spinning, either on mutex implementation or main loop, so fairness is required.
-Note OpenCL features: kernel keyword, global qualifier, ``get'' functions.
-Note that pop\_or\_steal is executed by all, but internally the master manipulates the queue.
-t is a single task, but is chunky enough that all threads can process part of it.}
+
+\ADComment{Points to make are: there is spinning, either on mutex
+implementation or main loop, so fairness is required.  Note OpenCL
+features: kernel keyword, global qualifier, ``get'' functions.  Note
+that pop\_or\_steal is executed by all, but internally the master
+manipulates the queue.  t is a single task, but is chunky enough that
+all threads can process part of it.}
 
@@
 
-\HEComment{ cooperative version: add kill/fork, and must pass the number
-of task pools explicitely since get\_num\_workgroups() cannot be used to
-know the number of pools. }
-
-\begin{lstlisting}
-__kernel connect_four(__global uchar *base_board,
-                      const int lookahead,
-                      __global atomic_int *best_move,
-                      __global int *all_done,
-                      __global Node *tree,
-                      __global Task *pools,
-                      __global Mutex *pool_m,
-                      const int num_pools)
-{
-  // init
-  if (get_global_id(0) == 0) {
-    init_first_tasks(tree, pools, num_pools);
-  }
-  global_barrier();
-
-  // main loop
-  while (true) {
-    offer_kill();
-    offer_fork();
-
-    // obtain a task, or terminate if all work is done
-    int pool_id = get_group_id(0) % num_pools;
-    Task *t = pop_or_steal(pools, pool_m, pool_id, num_pools);
-    if (t == NULL) {
-      if (*all_done) { return; } else { continue; }
-    }
-
-    // perform task, may create new tasks
-    Node *node = tree[t->node_index];
-    value = node_value(base_board, node);
-    if (winning(value) || node->level == lookahead) {
-      // we reached a leaf
-      update_parent_value(node, tree, best_move, all_done);
-    } else {
-      // create_children() pushes or donates new tasks
-      create_children(node, tree, pools, pool_m, pool_id, num_pools);
-    }
-  }
-}
-\end{lstlisting}
 \caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}

### 2016-11-10 write-ups/PLDI/paper.tex

@@
 \begin{figure}
-\ADComment{Tyler/Hugues: please insert graph traversal example.}
-\TSComment{Hows this?}
 
 \begin{lstlisting}
-kernel graph_traversal(global graph *g, 
+kernel graph_app(global graph *g, 
                  global nodes *n0, global nodes *n1) {
@@
 OpenCL specifically makes no guarantees about fair scheduling between
-workgroups, stating~\cite{...}: \emph{``A conforming implementation
-  may choose to serialize the workgroups so a correct algorithm cannot
-  assume that workgroups will execute in parallel.  There is no safe
-  and portable way to synchronize across the independent execution of
-  workgroups since once in the work-pool, they can execute in any
-  order.''}  OpenCL is not alone in this regard: CUDA similarly
-provides no guarantees~\cite{...}, and HSA guarantees only that
-\ADComment{Tyler, can you state the HSA guarantee?}~\cite{...}.
+workgroups \TSComment{This should explicitly say \emph{workgroups
+    executing the same kernel}, otherwise HSA has some weirdness},
+stating~\cite{...}: \emph{``A conforming implementation may choose to
+  serialize the workgroups so a correct algorithm cannot assume that
+  workgroups will execute in parallel.  There is no safe and portable
+  way to synchronize across the independent execution of workgroups
+  since once in the work-pool, they can execute in any order.''}
+OpenCL is not alone in this regard: CUDA similarly provides no
+guarantees~\cite{...}.
+
+HSA provides limited, one-way guarantees,
+stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
+  for values written by work-group B without deadlock provided ... A
+  comes after B in work-group flattened ID order}. However, this
+one-way communication is not sufficient so support either the
+work-stealing or graph application examples of
+Figure~\ref{fig:graphsearch} and \ref{fig:workstealing}. As described
+above, both require \emph{symmetric} communication between
+threads. \ADComment{Tyler, can you state the HSA guarantee?},
+\TSComment{How is this?}~\cite{...}.
 
@@
 primitive can be implemented following any inter-workgroup barrier
-design, e.g.~\cite{...}, and the atomic operations provided by the
-OpenCL 2.0 memory model enable a memory-safe
+design, \TSComment{This isn't exactly true. For cooperative kernels,
+  the barriers must be able to handle a growing or shrinking number of
+  workgroups.} e.g.~\cite{...}, and the atomic operations provided by
+the OpenCL 2.0 memory model enable a memory-safe
 implementation~\cite{...}.  However, implementing such a barrier is
@@
 
+\begin{figure}
+
+\begin{lstlisting}
+kernel graph_app(global graph *g, 
+                 global nodes *n0, global nodes *n1) {
+
+(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
+(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
+(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
+
+  while(in_nodes.size > 0) {
+
+(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
+(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
+
+    for (int i = tid; i < nodes.size; i += stride) {
+      process_node(g, in_nodes[i], out_nodes, level);
+    }
+
+    level++;
+    swap(in_nodes, out_nodes);
+    reset(out_nodes);
+(*@\label{line:cgraph:resizing}@*)    resizing_global_barrier();
+  }
+}
+\end{lstlisting}
+\caption{Cooperative kernel port for Figure~\ref{fig:graphsearch}}\label{fig:cgraphsearch}
+\end{figure}
+
 \ADComment{TODO.  We want to show the examples from Section~\ref{sec:background} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}
+
+In practice, we found that transforming traditional blocking GPU
+kernels into cooperative kernels was straightforward. For example,
+Figure~\ref{fig:cgraphsearch} shows the graph traversal example of
+Figure~\ref{fig:graphsearch} transformed into a cooperative kernel.
+On line~\ref{line:cgraph:resizing}, we change the original barrier
+into a resizing barrier. Several variables are marked to be
+transmitted in the case of workgroups joining at the resizing barrier
+(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
+\ref{line:cgraph:transmit3}): {\tt level} must be restored so that
+new workgroups know which frontier they are processing; {\tt
+  in\_nodes} and {\tt out\_nodes} must be restored so that new
+workgroups know which of the node arrays are input and output. Lastly,
+the static work distribution of the original kernel is no longer valid
+in a cooperative kernel. This is because the stride (which is based on
+$M$) may change after each resizing barrier
+call. To fix this, we re-distribute the work after each resizing
+barrier call by recomputing the thread id and stride
+(lines~\ref{line:cgraph:rechunking1} and
+\ref{line:cgraph:rechunking2}).
+
+\TSComment{Can Hugues talk about work stealing here?}
 
@@
 the value returned by $\getnumgroups()$ is not fixed during the
-lifetime of a cooperative kernel.  Although $\getnumgroups()$ has
-atomic semantics, and so can be called at any time to yield a
-well-defined value, the value returned by two successive calls may
-differ if workgroups have executed $\offerkill$, $\offerfork$ or
-$\resizingglobalbarrier$ in-between.  The programmer must therefore
-write their cooperative kernel in a manner that is robust to changes
-in the value returned by $\getnumgroups()$.
+lifetime of a cooperative kernel \TSComment{Neither is {\tt
+    get\_global\_size}, is that worth mentioning?}.  Although
+$\getnumgroups()$ has atomic semantics, and so can be called at any
+time to yield a well-defined value, the value returned by two
+successive calls may differ if workgroups have executed $\offerkill$,
+$\offerfork$ or $\resizingglobalbarrier$ in-between.  The programmer
+must therefore write their cooperative kernel in a manner that is
+robust to changes in the value returned by $\getnumgroups()$.
 
@@
 data processing.  \ADComment{Tyler, can we give an example of a data
-  processing loop that would have to be changed?}
+  processing loop that would have to be changed?} \TSComment{Done,
+its in figure~\ref{fig:cgraphsearch}}
 

### 2016-11-10 write-ups/PLDI/paper.tex
added cooperative example of graph traversal and some text. Also added quote by HSA

@@
 
-We use two examples of irregular data-parallel algorithms---a work
-stealing queue and breadth-first graph search---to introduce the
-industry-standard OpenCL programming model on which we base cooperative
-kernels.
-
-\paragraph{OpenCL concepts}
-An OpenCL program is typically executed by many \emph{threads} running
-in parallel. Threads are organized into \emph{workgroups} of equal
-size. The memory is organized in three levels: \emph{global} memory is
-accessible by any thread, \emph{local} memory is shared at the workgroup
-level (i.e., threads of different workgroups cannot access the same
-local memory), and \emph{private} memory is dedicated to thread-level
-storage and cannot be shared between threads. These concepts are
-illustrated over the two examples.
-
-\paragraph{Work stealing}
-Work stealing is generally used for computational work that can be
-splitted into independently processable tasks. The processing of a given
-task may create an arbitrary number of new tasks in such a way that it
-is typically hard to know the total number of tasks in advance. Work
-stealing enables the dynamic balance of task processing over several
-processing units, OpenCL workgroups in our case. Each workgroup has an
-associated task pool from which it obtains tasks to process and to where
-it stores newly created tasks. When a workgroup's pool becomes empty, a
-workgroup---rather than terminating---tries to \emph{steal} a task from
-an other pool. Hence, workgroups keep on processing tasks until all the
-work is done.
-
-\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation of
-a work stealing application. All threads start their execution at the
-function marked with the $\mathsf{kernel}$ keyword. They receive a
-pointer to the task pools that reside in global memory, and which were
-initialized by the host program beforehand to contain the initial
-tasks. Each thread can identify the workgroup it belongs to via the
-function \getgroupid, and OpenCL provides a series of primitives in the
-same vein to access for instance the thread global id, or its local id
-within a workgroup. Here the group id is stored into the variable
-$\mathsf(pool\_id)$ (which is, by default, in private memory) and used
-to access the relevant task pool. The $\mathsf(pop\_or\_steal)$ function
-is called by all threads of a workgroup, although inside only the thread
-with a local id equal to $0$, also called \emph{master thread} of the
-workgroup, will try to pop or steal a task from the pools. The master
-thread uses local memory inside $\mathsf(pop\_or\_steal)$ to ensure that
-the returned value is valid for all threads of its workgroup. If a task
-was obtained, then all threads of the workgroup can participate in the
-processing of the task inside the $\mathsf{process\_task}$
-function. \HEComment{Manage to say that a task is considered big enough
-to benefit from parallel computation} If some new tasks are created
-during the processing, it is again the master thread who will be in
-charge of pushing them to the pool. Although not depicted here,
-concurrent accesses to each pool is guarded by a mutex which is
-implemented by atomic compare and swap operations on global
-memory. Henceforth, master threads may be spinning on mutexes, and
-workgroups are also spinning on the main loop, so a fair scheduling of
-workgroups is desirable.
+We use two examples of irregular data-parallel algorithms---a work stealing queue and breadth-first graph
+search---to introduce the industry-standard OpenCL programming model on which we base cooperative kernels.
+
 \begin{figure}
-
-\ADComment{Points to make are: there is spinning, either on mutex
-implementation or main loop, so fairness is required.  Note OpenCL
-features: kernel keyword, global qualifier, ``get'' functions.  Note
-that pop\_or\_steal is executed by all, but internally the master
-manipulates the queue.  t is a single task, but is chunky enough that
-all threads can process part of it.}
+\ADComment{Tyler/Hugues: please insert workstealing example.}
+
+\HEComment{standard version: one task pool per workgroup, the number of
+workgroup is a constant. So we access our pool by pool\_id ==
+get\_group\_id(0). When we try to steal, we can look into other pools
+which are indexed by (pool\_id + delta) modulo get\_num\_workgroups(0).
+}
+
+\ADComment{Points to make are: there is spinning, either on mutex implementation or main loop, so fairness is required.
+Note OpenCL features: kernel keyword, global qualifier, ``get'' functions.
+Note that pop\_or\_steal is executed by all, but internally the master manipulates the queue.
+t is a single task, but is chunky enough that all threads can process part of it.}
 
@@
 
+\HEComment{ cooperative version: add kill/fork, and must pass the number
+of task pools explicitely since get\_num\_workgroups() cannot be used to
+know the number of pools. }
+
+\begin{lstlisting}
+__kernel connect_four(__global uchar *base_board,
+                      const int lookahead,
+                      __global atomic_int *best_move,
+                      __global int *all_done,
+                      __global Node *tree,
+                      __global Task *pools,
+                      __global Mutex *pool_m,
+                      const int num_pools)
+{
+  // init
+  if (get_global_id(0) == 0) {
+    init_first_tasks(tree, pools, num_pools);
+  }
+  global_barrier();
+
+  // main loop
+  while (true) {
+    offer_kill();
+    offer_fork();
+
+    // obtain a task, or terminate if all work is done
+    int pool_id = get_group_id(0) % num_pools;
+    Task *t = pop_or_steal(pools, pool_m, pool_id, num_pools);
+    if (t == NULL) {
+      if (*all_done) { return; } else { continue; }
+    }
+
+    // perform task, may create new tasks
+    Node *node = tree[t->node_index];
+    value = node_value(base_board, node);
+    if (winning(value) || node->level == lookahead) {
+      // we reached a leaf
+      update_parent_value(node, tree, best_move, all_done);
+    } else {
+      // create_children() pushes or donates new tasks
+      create_children(node, tree, pools, pool_m, pool_id, num_pools);
+    }
+  }
+}
+\end{lstlisting}
 \caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}

### 2016-11-10 write-ups/PLDI/paper.tex
Fxi latex compilation and move paragraph

--- 2016-11-10 write-ups/PLDI/paper.tex
@@
 
-\paragraph{OpenCL concepts}
-An OpenCL program is typically executed by many \emph{threads} running
-in parallel. Threads are organized into \emph{workgroups} of equal
-size. The memory is organized in three levels: \emph{global} memory is
-accessible by any thread, \emph{local} memory is shared at the workgroup
-level (i.e., threads of different workgroups cannot access the same
-local memory), and \emph{private} memory is dedicated to thread-level
-storage and cannot be shared between threads. These concepts are
-illustrated over the two examples.
+% \paragraph{OpenCL concepts}
+% An OpenCL program is typically executed by many \emph{threads} running
+% in parallel. Threads are organized into \emph{workgroups} of equal
+% size. The memory is organized in three levels: \emph{global} memory is
+% accessible by any thread, \emph{local} memory is shared at the workgroup
+% level (i.e., threads of different workgroups cannot access the same
+% local memory), and \emph{private} memory is dedicated to thread-level
+% storage and cannot be shared between threads. These concepts are
+% illustrated over the two examples.
+
+\ADComment{Points to make are: there is spinning, either on mutex
+implementation or main loop, so fairness is required.  Note OpenCL
+features: kernel keyword, global qualifier, ``get'' functions.  Note
+that pop\_or\_steal is executed by all, but internally the master
+manipulates the queue.  t is a single task, but is chunky enough that
+all threads can process part of it.}
+
+\begin{figure}
+
+\begin{lstlisting}
+kernel work_stealing(global Task *pools) {
+  while (more_work(pools)) {
+    int pool_id = get_group_id();
+    Task *t = pop_or_steal(pools, pool_id);
+    if (t) {
+      // task processing may create more work
+      process_task(t, pools, pool_id);
+    }
+  }
+}
+\end{lstlisting}
+
+\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
+\end{figure}
+
+\begin{figure}
+
+\begin{lstlisting}
+kernel graph_app(global graph *g, 
+                 global nodes *n0, global nodes *n1) {
+
+  int level = 0;
+  global nodes * in_nodes = n0;
+  global nodes * out_nodes = n1;
+
+  int tid = get_global_id();
+  int stride = get_global_size();
+
+  while(in_nodes.size > 0) {
+
+    for (int i = tid; i < nodes.size; i += stride) {
+      process_node(g, in_nodes[i], out_nodes, level);
+    }
+
+    level++;
+    swap(in_nodes, out_nodes);
+    reset(out_nodes);
+    global_barrier();
+  }
+}
+\end{lstlisting}
+\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
+\end{figure}
+
+\paragraph{OpenCL kernels}
+
+An OpenCL program is divided into \emph{host} and \emph{device}
+components.  A host application runs on the CPU and launches one or
+more \emph{kernels} that run on accelerator devices---GPUs in the
+context of this paper.  A kernel is writen in OpenCL C, based on C99.
+All threads executing a kernel execute the same entry function with
+identical arguments, but a thread may query its unique thread id in
+order to access different data or follow different control flow paths
+to other threads.
+
+Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
+kernel that uses a work stealing queue.  \ADComment{Describe some
+  aspects of the example, saying that the mutex is not a primitive,
+  and can be coded up using the memory model but is not safe due to
+  the execution model.  Make it clear that the lock function uses
+  busy-waiting.}
 
@@
 tasks. Each thread can identify the workgroup it belongs to via the
-function \getgroupid, and OpenCL provides a series of primitives in the
+function $\getgroupid$, and OpenCL provides a series of primitives in the
 same vein to access for instance the thread global id, or its local id
@@
 workgroups is desirable.
-\begin{figure}
-
-\ADComment{Points to make are: there is spinning, either on mutex
-implementation or main loop, so fairness is required.  Note OpenCL
-features: kernel keyword, global qualifier, ``get'' functions.  Note
-that pop\_or\_steal is executed by all, but internally the master
-manipulates the queue.  t is a single task, but is chunky enough that
-all threads can process part of it.}
-
-\begin{lstlisting}
-kernel work_stealing(global Task *pools) {
-  while (more_work(pools)) {
-    int pool_id = get_group_id();
-    Task *t = pop_or_steal(pools, pool_id);
-    if (t) {
-      // task processing may create more work
-      process_task(t, pools, pool_id);
-    }
-  }
-}
-\end{lstlisting}
-
-\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
-\end{figure}
-
-\begin{figure}
-
-\begin{lstlisting}
-kernel graph_app(global graph *g, 
-                 global nodes *n0, global nodes *n1) {
-
-  int level = 0;
-  global nodes * in_nodes = n0;
-  global nodes * out_nodes = n1;
-
-  int tid = get_global_id();
-  int stride = get_global_size();
-
-  while(in_nodes.size > 0) {
-
-    for (int i = tid; i < nodes.size; i += stride) {
-      process_node(g, in_nodes[i], out_nodes, level);
-    }
-
-    level++;
-    swap(in_nodes, out_nodes);
-    reset(out_nodes);
-    global_barrier();
-  }
-}
-\end{lstlisting}
-\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
-\end{figure}
-
-\paragraph{OpenCL kernels}
-
-An OpenCL program is divided into \emph{host} and \emph{device}
-components.  A host application runs on the CPU and launches one or
-more \emph{kernels} that run on accelerator devices---GPUs in the
-context of this paper.  A kernel is writen in OpenCL C, based on C99.
-All threads executing a kernel execute the same entry function with
-identical arguments, but a thread may query its unique thread id in
-order to access different data or follow different control flow paths
-to other threads.
-
-Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
-kernel that uses a work stealing queue.  \ADComment{Describe some
-  aspects of the example, saying that the mutex is not a primitive,
-  and can be coded up using the memory model but is not safe due to
-  the execution model.  Make it clear that the lock function uses busy-waiting.}  Figure~\ref{fig:graphsearch} shows the
+
+Figure~\ref{fig:graphsearch} shows the
 skeleton of a graph traversal algorithm.  \ADComment{Describe some

### 2016-11-10 write-ups/PLDI/paper.tex
Add paragraph on adaptation of work stealing

@@
 
-  while(in_nodes.size > 0) {
+(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
 
@@
 
+(*@\label{line:graph:swap}@*)    swap(in_nodes, out_nodes);
+(*@\label{line:graph:gb1}@*)    global_barrier();
+(*@\label{line:graph:reset}@*)    reset(out_nodes);
     level++;
-    swap(in_nodes, out_nodes);
-    reset(out_nodes);
-    global_barrier();
+(*@\label{line:graph:gb2}@*)    global_barrier();
   }
@@
 
-Figure~\ref{fig:graphsearch} shows the
-skeleton of a graph traversal algorithm.  \ADComment{Describe some
+\ADComment{Describe some
   aspects of the example, saying that the global barrier is not a
   primitive, and can be coded up using the memory model but is not
-  safe due to the execution model.  Make it clear that the barrier uses busy-waiting.}
+  safe due to the execution model.  Make it clear that the barrier
+  uses busy-waiting.}
+
+\paragraph{Graph traversal} Figure~\ref{fig:graphsearch} shows the
+skeleton of a frontier based graph traversal algorithm (this approach
+has been shown to be efficient on GPUs in~\ref{...}). The kernel is
+given three arguments in global memory: a graph structure, and two
+arrays that can contain graph nodes. Initially, {\tt n0} contains the
+starting nodes to process. Threads keep three local variables: {\tt
+  level} which indicates the frontier level which is currently being
+processed, {\tt in\_nodes} and {\tt out\_nodes} which point to
+distinct node arrays. {\tt in\_nodes} contains the nodes to be
+processed for the current frontier, while {\tt out\_nodes} contains
+the nodes to be processed during the next frontier.
+
+The application iterates as long as the current frontier contains
+nodes to process (line~\ref{line:graph:iterate}). At each frontier,
+the nodes to be processed are evenly distributed between the executing
+threads. This is done through \emph{stride} based processing. A thread
+retrieves its global thread id and the total number of threads
+(obtained through the {\tt get\_global\_id()} and {\tt
+  get\_global\_size()} OpenCL primitives). A thread then processes
+each node associated with id (i.e.\ indices with its thread id value
+added to any integer multiple of the total number of threads up to the
+list size). A thread processes a node via the {\tt process\_node}
+function, which may (a) push nodes to process in the next frontier to
+the {\tt out\_nodes} list and (b) use the frontier level in
+computation. After processing the frontier, the threads swap their
+local node array pointers (line~\ref{line:graph:swap}), i.e.\ the {\tt
+  out\_nodes} of the current frontier becomes the input of the next
+frontier.
+
+At this point, the GPU threads must wait for all other threads to
+finishing processing the frontier before they can continue. To achieve
+this, we use a global barrier construct
+(line~\ref{line:graph:gb1}). After all threads reach this point, the
+output node array is reset (line~\ref{line:graph:reset}) and the level
+is incremented. Again, threads must wait until the output node is
+reset at a global barrier (line~\ref{line:graph:gb2}). At this point
+threads can continue to the next frontier.
+
+The global barrier used in this application is not provided as a GPU
+primitive, though many previous works have shown that such a global
+barrier can be implemented~\cite{...}. These implementations use well
+defined atomic memory operations to ensure memory consistency and use
+spinning to ensure threads wait at the barrier for the remaining
+threads to reach the barrier. Even though such barriers appear to run
+reliably on current GPUs, they violate GPU language specifications,
+which do not provide fair scheduling of workgroups. These barriers
+require fair workgroup scheduling in order to function correctly. That
+is, \emph{all} workgroups must be fairly scheduled in order to reach
+the barrier, otherwise, the workgroups waiting at the barrier will not
+make progress.
 
@@
 
+For the work stealing example, there is no state to transmit since a
+computation is entirely parametrized by a task, which is retrieved from
+a task pool located in global memory. Care must be taken on the dynamic
+change of the number of workgroups though: we cannot assume that the
+total number of task pools is equal to $\getnumgroups$. Thus, the number
+of pools must be given as an argument, and the $\mathsf{pool}$ argument
+must be obtained from the group id modulo the total number of
+pools. This total number of pools must also be passed to
+$\mathsf{pop\_or\_steal}$, since when a workgroup wants to steal it must
+access other pool indexes, which are bounded by the total number of
+pools.
+
 \subsection{Programming Guidelines}\label{sec:programmingguidelines}

### 2016-11-10 src/run_test.py
Add configurations of number of workgroups for the non-persistent

@@ -6,13 +6,6 @@ import shutil
 import glob
 import re
 
-###############################################################################
-# TODO:
-# - remove printing to file since we have summary
-# - use regexp to grab occupancy of mergedskiptask
-# - reuse occupancy when running standalone
-###############################################################################
-
 def fake_placeholder(x):
     return
 
@@ -25,10 +18,10 @@ DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""
 
 PROGRAMS = {
-    # "pannotia_color",
-    # "pannotia_mis",
+    "pannotia_color",
+    "pannotia_mis",
     "pannotia_bc",
-    # "pannotia_sssp"
+    "pannotia_sssp"
 }
 
 PROGRAM_DATA = {
@@ -46,14 +39,15 @@ PROGRAM_DATA = {
                          "stat" : "mis_ecology" },
                        { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
                          "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
-                         "stat" : "mis_G3_circuit" } ],
+                         "stat" : "mis_G3_circuit" }
+    ],
 
     "pannotia_bc" : [ { "input" : os.path.join("inputs", "bc", "1k_128k.gr"),
                         "solution" : os.path.join("solutions", "bc_1k_128k.txt"),
                         "stat" : "bc_1k_128k" },
-                      # { "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
-                      #   "solution" : os.path.join("solutions", "bc_2k_1M.txt"),
-                      #   "stat" : "bc_2k_1M" }
+                      { "input" : os.path.join("inputs", "bc", "2k_1M.gr"),
+                        "solution" : os.path.join("solutions", "bc_2k_1M.txt"),
+                        "stat" : "bc_2k_1M" }
     ],
 
     "pannotia_sssp" : [ { "input" : os.path.join("inputs", "sssp", "USA-road-d.NW.gr"),
@@ -61,7 +55,7 @@ PROGRAM_DATA = {
                           "stat" : "sssp_usa_road" } ]
 }
 
-# TODO: make it per machine ?
+# TODO: make it per machine
 MATMULT_CONFIG = [
     # This is for Hugues laptop
     { "freq" : "70", "matdim" : "160", "name" : "light" },
@@ -163,21 +157,28 @@ def run_suite():
             collect_stats(d, prefix)
 
             for c in MATMULT_CONFIG:
-                # Merged
-                cmd = [exe]
[…]

### 2016-11-10 write-ups/PLDI/paper.tex
added write-up about the graph traversal algorithm in section 2

@@
 
-For the work stealing example, there is no state to transmit since a
-computation is entirely parametrized by a task, which is retrieved from
-a task pool located in global memory. Care must be taken on the dynamic
-change of the number of workgroups though: we cannot assume that the
-total number of task pools is equal to $\getnumgroups$. Thus, the number
-of pools must be given as an argument, and the $\mathsf{pool}$ argument
-must be obtained from the group id modulo the total number of
-pools. This total number of pools must also be passed to
-$\mathsf{pop\_or\_steal}$, since when a workgroup wants to steal it must
-access other pool indexes, which are bounded by the total number of
-pools.
-
 \subsection{Programming Guidelines}\label{sec:programmingguidelines}

### 2016-11-10 src/run_test.py
Add iterations

@@ -12,7 +12,7 @@ def fake_placeholder(x):
 EXE_PATH      = ""
 DATA_PATH     = ""
 STAT_PATH     = ""
-ITERS         = 1
+ITERATIONS    = "10"
 PRINT         = fake_placeholder
 DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""
@@ -134,6 +134,7 @@ def run_suite():
             # indicate very high number of workgroups to finally obtain occupancy
             cmd = cmd + ["--num_wgs", "1000"]
             cmd = cmd + ["--skip_tasks", "1"]
+            cmd = cmd + ["--merged_iterations", ITERATIONS]
             prefix = d["stat"] + "_skiptask"
             record_stdout = prefix + "_stdout.txt"
             exec_cmd(cmd, prefix, record_stdout)
@@ -149,7 +150,7 @@ def run_suite():
             cmd = cmd + ["--graph_file", graph_in]
             cmd = cmd + ["--graph_solution_file", graph_sol]
             cmd = cmd + ["--threads_per_wg", "128"]
-            cmd = cmd + ["--run_persistent", "1"]
+            cmd = cmd + ["--run_persistent", ITERATIONS]
             cmd = cmd + ["--num_wgs", finalsize]
             prefix = d["stat"] + "_standalone"
             record_stdout = prefix + "_stdout.txt"
@@ -162,6 +163,7 @@ def run_suite():
                     cmd = cmd + ["--graph_file", graph_in]
                     cmd = cmd + ["--graph_solution_file", graph_sol]
                     cmd = cmd + ["--threads_per_wg", "128"]
+                    cmd = cmd + ["--merged_iterations", ITERATIONS]
                     # indicate very high number of workgroups to finally obtain occupancy
                     cmd = cmd + ["--num_wgs", "1000"]
                     cmd = cmd + ["--non_persistent_frequency", c["freq"]]
@@ -186,6 +188,7 @@ def main():
     global DATA_PATH
     global STAT_PATH
     global PRINT
+    global ITERATIONS
     global DATA_PRINT
     global NAME_OF_CHIP
     global MATMULT_CONFIG

### 2016-11-10 write-ups/PLDI/paper.tex

--- 2016-11-10 write-ups/PLDI/paper.tex
@@
 
+\begin{figure}
+
+\begin{lstlisting}
+kernel work_stealing(global Task *pools, const int num_pools) {
+  while (more_work(pools)) {
+    int pool_id = get_group_id() % num_pools;
+    Task *t = pop_or_steal(pools, pool_id, num_pools);
+    if (t) {
+      // task processing may create more work
+      process_task(t, pools, pool_id);
+    }
+  }
+}
+\end{lstlisting}
+
+\caption{Cooperative version of work stealing: the number of workpools
+must be explicitely passed as arguments since the number of workgroups
+is not a constant anymore}\label{fig:workstealing-cooperative}
+\end{figure}
+
 \subsection{Programming Guidelines}\label{sec:programmingguidelines}

### 2016-11-10 write-ups/PLDI/paper.tex
Edits.

@@
 
-Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
-kernel that uses a work stealing queue.  \ADComment{Describe some
+\paragraph{Work stealing example}
+  \ADComment{Describe some
   aspects of the example, saying that the mutex is not a primitive,
@@
   busy-waiting.}  
-
-\paragraph{Work stealing}
-Work stealing is generally used for computational work that can be
-splitted into independently processable tasks. The processing of a given
-task may create an arbitrary number of new tasks in such a way that it
-is typically hard to know the total number of tasks in advance. Work
-stealing enables the dynamic balance of task processing over several
-processing units, OpenCL workgroups in our case. Each workgroup has an
-associated task pool from which it obtains tasks to process and to where
-it stores newly created tasks. When a workgroup's pool becomes empty, a
-workgroup---rather than terminating---tries to \emph{steal} a task from
-an other pool. Hence, workgroups keep on processing tasks until all the
-work is done.
+Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
+kernel that uses a work stealing queue.
+%
+Work stealing is useful when a computation can be split into multiple
+tasks that can be processed independently, but where the number of
+tasks is not known ahead of time because processing a given task may
+create an arbitrary number of new tasks. Work stealing enables dynamic
+balancing of tasks across several processing units, OpenCL workgroups
+in our case. Each workgroup has an associated task pool from which it
+obtains tasks to process, and to which it stores newly created
+tasks. When a workgroup's pool becomes empty, a workgroup---rather
+than terminating---tries to \emph{steal} a task from an other
+pool. Hence, workgroups keep on processing tasks until all the work is
+done.
 
@@
 one-way communication is not sufficient so support either the
-work-stealing or graph application examples of
+work stealing or graph application examples of
 Figure~\ref{fig:graphsearch} and \ref{fig:workstealing}. As described
@@
 at every call to the inter-workgroup barrier. Many applications using
-the work-stealing paradigm do not carry state between task
+the work stealing paradigm do not carry state between task
 computation; thus workgroups can be created or terminated immediately

### 2016-11-10 write-ups/PLDI/paper.tex
Progress on the background section - great examples guys\!

@@
 % 
-\documentclass[nocopyrightspace,10pt]{sigplanconf}
+\documentclass[numbers,nocopyrightspace,10pt]{sigplanconf}
 % \documentclass[pldi-cameraready]{sigplanconf-pldi16}
@@
 
-\section{Background and Examples}\label{sec:background}
+\section{Background and Motivating Examples}\label{sec:background}
 
@@
 
-\ADComment{Points to make are: there is spinning, either on mutex
-implementation or main loop, so fairness is required.  Note OpenCL
-features: kernel keyword, global qualifier, ``get'' functions.  Note
-that pop\_or\_steal is executed by all, but internally the master
-manipulates the queue.  t is a single task, but is chunky enough that
-all threads can process part of it.}
-
 \begin{figure}
@@
 All threads executing a kernel execute the same entry function with
-identical arguments, but a thread may query its unique thread id in
-order to access different data or follow different control flow paths
-to other threads.
+identical arguments, but a thread may invoke the $\getglobalid()$
+function to query its unique thread id in order to access different
+data or follow different control flow paths to other threads.
+
+The threads executing a kernel are hierarchically organised.  A kernel
+is executed by a number of \emph{workgroups}, and each workgroup
+consists of a fixed number of threads.  Builtin functions
+$\getlocalid()$ and $\getgroupid()$ return a thread's local id within
+its workgroup and the id of the workgroup, respectively.\footnote{In general, an OpenCL kernel is organised as a multi-dimensional grid
+of multi-dimensional workgroups, and the above functions each take
+an integer dimension argument.  For simplicity, we consider only the
+one-dimensional case throughout the paper, which captures all the
+GPU-accelerated irregular algorithms we are aware of, and use
+e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
+}  The number
+of threads per workgroup and the number of workgroups are obtained via
+$\getlocalsize()$ and $\getnumgroups()$.  A thread's local and global
+id are related via the equation $\getglobalid() = \getgroupid() \times
+\getlocalsize() + \getlocalid()$.
+
+\ADComment{Maybe cut?} Within a workgroup, threads may also be
+organised into \emph{subgroups} (termed \emph{warps} and
+\emph{wavefronts} on NVIDIA and AMD architectures, respectively).
+
+\paragraph{Memory spaces and memory model}
+An OpenCL kernel has access to four memory spaces.  Shared virtual
+memory (SVM) is accessible to all threads, and to the host
+application.  Global memory is shared among all threads executing a
+kernel, but can only be safely accessed by the host application before
+or after kernel execution.  Local memory is shared among all threads
+in the same workgroup: each workgroup has its own portion of local
+memory, which can be used for fast intra-workgroup communication.
+Each thread has an allocation of very fast private memory, where
+function-local variables are allocated.
+
+Fine-grained communication between threads, and also with the host in
+the case of SVM, is enabled via a set of atomic data types and
+operations, based on the C11 memory model~\cite{...}.  In particular,
+atomic SVM operations enable fine-grained communication between host
+and device.  Our implementation of cooperative kernels
+(Section~\ref{sec:implementation}) depends on this recent OpenCL
+feature.
 
 \paragraph{Work stealing example}
-  \ADComment{Describe some
-  aspects of the example, saying that the mutex is not a primitive,
-  and can be coded up using the memory model but is not safe due to
-  the execution model.  Make it clear that the lock function uses
-  busy-waiting.}  
 Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
@@
 %
-Work stealing is useful when a computation can be split into multiple
+The work stealing paradigm is useful when a computation can be split into multiple
 tasks that can be processed independently, but where the number of
@@
 
-\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation of
-a work stealing application. All threads start their execution at the
+\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation
+of a work stealing algorithm. All threads start their execution at the
 function marked with the $\mathsf{kernel}$ keyword. They receive a
-pointer to the task pools that reside in global memory, and which were
-initialized by the host program beforehand to contain the initial
+pointer to the task pools, which reside in global memory, and which
+were initialized by the host program beforehand to contain the initial
 tasks. Each thread can identify the workgroup it belongs to via the
-function $\getgroupid$, and OpenCL provides a series of primitives in the
-same vein to access for instance the thread global id, or its local id
-within a workgroup. Here the group id is stored into the variable
-$\mathsf(pool\_id)$ (which is, by default, in private memory) and used
-to access the relevant task pool. The $\mathsf(pop\_or\_steal)$ function
-is called by all threads of a workgroup, although inside only the thread
-with a local id equal to $0$, also called \emph{master thread} of the
-workgroup, will try to pop or steal a task from the pools. The master
-thread uses local memory inside $\mathsf(pop\_or\_steal)$ to ensure that
-the returned value is valid for all threads of its workgroup. If a task
-was obtained, then all threads of the workgroup can participate in the
-processing of the task inside the $\mathsf{process\_task}$
-function. \HEComment{Manage to say that a task is considered big enough
-to benefit from parallel computation} If some new tasks are created
-during the processing, it is again the master thread who will be in
-charge of pushing them to the pool. Although not depicted here,
-concurrent accesses to each pool is guarded by a mutex which is
-implemented by atomic compare and swap operations on global
-memory. Henceforth, master threads may be spinning on mutexes, and
-workgroups are also spinning on the main loop, so a fair scheduling of
-workgroups is desirable.
+function $\getgroupid()$.  Here the group id is stored into the
+variable $\mathsf{pool\_id}$ (which is, by default, in private memory)
+and used to access the relevant task pool. The
+$\mathsf{pop\_or\_steal}$ function is called by all threads of a
+workgroup, although inside only the thread with a local id equal to
+$0$, also called the \emph{master thread} of the workgroup,
+\ADComment{Hugues, I changed; please check:} will try to pop a task
+from the workgroup's pool, or steal a task from a different pool. The
+master thread uses local memory inside $\mathsf{pop\_or\_steal}$ to
+ensure that the returned value is valid for all threads of its
+workgroup. If a task was obtained, then all threads of the workgroup
+participate in the processing of the task inside the
+$\mathsf{process\_task}$ function. \HEComment{Manage to say that a
+  task is considered big enough to benefit from parallel computation}
+\ADComment{I think that would be too much detail given the space we
+  have, so I would not delve into this. Remove both comments if you
+  agree.} Processing this task may lead to the creation of further
+tasks, which the master thread will push to the workgroup's pool.
+
+Although not depicted here, concurrent accesses to each pool inside
+$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
+mutex which is implemented using atomic compare and swap operations on
+global memory.  As a result, the kernel presents two opportunities for
+spin-waiting: spinning to obtain a mutex, and spinning in the main
+kernel loop to obtain a task to process.  The algorithm thus depends
+on a fair scheduler, which, as discussed in Section~\ref{sec:intro},
+is not provided by current GPU programming models.  In
+Section~\ref{sec:cooperativeexamples} we show how the example can be
+updated to use our cooperative threads programming model to resolve
+the scheduling issue.
 
@@
 
-\paragraph{Graph traversal} Figure~\ref{fig:graphsearch} shows the
+\paragraph{Graph traversal example} Figure~\ref{fig:graphsearch} shows the
 skeleton of a frontier based graph traversal algorithm (this approach
@@
 
-The threads executing a kernel are hierarchically organised.  A kernel
-is executed by a number of \emph{workgroups}, and each workgroup
-consists of a fixed number of \emph{threads}.  Within a workgroup,
-threads may also be organised into \emph{subgroups} (termed
-\emph{warps} and \emph{wavefronts} on NVIDIA and AMD architectures,
-respectively).
-
 Two barrier primitives are provided in OpenCL.  A workgroup barrier
@@
 OpenCL specifically makes no guarantees about fair scheduling between
-workgroups \TSComment{This should explicitly say \emph{workgroups
-    executing the same kernel}, otherwise HSA has some weirdness},
-stating~\cite{...}: \emph{``A conforming implementation may choose to
-  serialize the workgroups so a correct algorithm cannot assume that
-  workgroups will execute in parallel.  There is no safe and portable
-  way to synchronize across the independent execution of workgroups
-  since once in the work-pool, they can execute in any order.''}
-OpenCL is not alone in this regard: CUDA similarly provides no
-guarantees~\cite{...}.
+workgroups executing the same kernel, stating~\cite{...}: \emph{``A
+  conforming implementation may choose to serialize the workgroups so
+  a correct algorithm cannot assume that workgroups will execute in
+  parallel.  There is no safe and portable way to synchronize across
+  the independent execution of workgroups since once in the work-pool,
+  they can execute in any order.''}  OpenCL is not alone in this
+regard: CUDA similarly provides no guarantees~\cite{...}.
 
@@
 work stealing or graph application examples of
-Figure~\ref{fig:graphsearch} and \ref{fig:workstealing}. As described
+Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
 above, both require \emph{symmetric} communication between
@@
 barrier.
-
-\paragraph{Identity and size functions}
-
-Builtin functions $\getlocalid()$ and $\getglobalid()$ provide a
-thread with an id that is unique within its workgroup and unique
-across all workgroups, respectively.  Each workgroup has an id,
-obtained via $\getgroupid()$.  The number of threads per workgroup and
-the number of workgroups are obtained via $\getlocalsize()$ and
-$\getnumgroups()$.  A thread's local and global id are related via the
-equation $\getglobalid() = \getgroupid() \times \getlocalsize() +
-\getlocalid()$.
-
-In general, an OpenCL kernel is organised as a multi-dimensional grid
-of multi-dimensional workgroups, and the above functions each take
-an integer dimension argument.  For simplicity, we consider only the
-one-dimensional case throughout the paper, which captures all the
-GPU-accelerated irregular algorithms we are aware of, and use
-e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
-
-\paragraph{Memory model}
-
-An OpenCL kernel has access to four memory spaces.  Shared virtual
-memory (SVM) is accessible to all threads, and to the host
-application.  Atomic operations on this memory space enable
-fine-grained communication between host and device.  Our
-implementation of cooperative kernels
-(Section~\ref{sec:implementation}) depends on this recent OpenCL
-feature.  Global memory is shared among all threads executing a
-kernel, but can only be safely accessed by the host application before
-or after kernel execution.  Local memory is shared among all threads
-in the same workgroup: each workgroup has its own portion of local
-memory, which can be used for fast intra-workgroup communication.
-Each thread has an allocation of very fast private memory, where
-function-local variables are allocated.
-
-Fine-grained communication between threads, and also with the host in
-the case of SVM, is enabled via a set of atomic data types and
-operations, based on the C11 memory model~\cite{...}.
-\ADComment{Perhaps say more.}  \ADComment{Can we illustrate using the
-  examples?}
 

### 2016-11-10 src/run_test.py
updated the script to include lonestar applications, recover from errors, etc

@@ -5,6 +5,7 @@ import subprocess
 import shutil
 import glob
 import re
+import platform
 
 def fake_placeholder(x):
     return
@@ -12,51 +13,106 @@ def fake_placeholder(x):
 EXE_PATH      = ""
 DATA_PATH     = ""
 STAT_PATH     = ""
-ITERATIONS    = "10"
+ITERATIONS    = "1"
 PRINT         = fake_placeholder
 DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""
+PLATFORM_ID   = "0"
+IS_AMD        = "0"
 
 PROGRAMS = {
-    "pannotia_color",
-    "pannotia_mis",
-    "pannotia_bc",
-    "pannotia_sssp"
+    "lonestar_sssp"
+#    "lonestar_bfs",
+#    "pannotia_color",
+#    "pannotia_mis",
+#    "pannotia_bc",
+#    "pannotia_sssp" 
 }
 
 PROGRAM_DATA = {
 
-    "pannotia_color" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
-                           "solution" : os.path.join("solutions", "color_ecology.txt"),
-                           "stat" : "color_ecology" },
-                         { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
-                           "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
-                           "stat" : "color_G3_circuit" }
+    "pannotia_color" : [ { "input" : os.path.join("pannotia","inputs", "color", "ecology1.graph"),
+                           "solution" : os.path.join("pannotia","solutions", "color_ecology.txt"),
+                           "stat" : "color_ecology",
+                           "suite" : "pannotia"},
+                         { "input" : os.path.join("pannotia","inputs", "color", "G3_circuit.graph"),
+                           "solution" : os.path.join("pannotia","solutions", "color_G3_circuit.txt"),
+                           "stat" : "color_G3_circuit",
+                           "suite" : "pannotia"}
     ],
 
-    "pannotia_mis" : [ { "input" : os.path.join("inputs", "color", "ecology1.graph"),
-                         "solution" : os.path.join("solutions", "color_ecology.txt"),
-                         "stat" : "mis_ecology" },
-                       { "input" : os.path.join("inputs", "color", "G3_circuit.graph"),
-                         "solution" : os.path.join("solutions", "color_G3_circuit.txt"),
[…]

### 2016-11-10 src/run_test.py
Fix typo in matmult config names

@@ -26,7 +26,7 @@ PROGRAMS = {
 #    "pannotia_color",
 #    "pannotia_mis",
 #    "pannotia_bc",
-#    "pannotia_sssp" 
+#    "pannotia_sssp"
 }
 
 PROGRAM_DATA = {
@@ -66,7 +66,7 @@ PROGRAM_DATA = {
                           "stat" : "sssp_usa_road",
                           "suite" : "pannotia"}
     ],
-    
+
     "lonestar_sssp" : [ { "input" : os.path.join("lonestar", "inputs","rmat22.gr"),
                         "solution" : os.path.join("lonestar","solutions", "sssp_rmat22.txt"),
                         "stat" : "sssp_rmat22",
@@ -92,7 +92,7 @@ PROGRAM_DATA = {
                         "solution" : os.path.join("lonestar","solutions", "bfs_USA_W.txt"),
                         "stat" : "bfs_USA-road-d.W",
                         "suite" : "lonestar"}
-                        
+
     ],
 }
 
@@ -180,7 +180,7 @@ def optional_debug():
     if platform.system() == "Windows":
         return "Debug"
     return ""
-    
+
 
 def run_suite():
     for p in PROGRAMS:
@@ -286,7 +286,7 @@ def main():
     if NAME_OF_CHIP == "HD5500":
         MATMULT_CONFIG = MATMULT_CONFIG_HD5500
     elif NAME_OF_CHIP == "HD520":
-	    MATMULT_CONFIG = MATMULT_CONFIG_HD5500
+	    MATMULT_CONFIG = MATMULT_CONFIG_HD520
 	else:
 	    print("Cannot find a matmult for your chip! Exiting")
 		exit(0)

### 2016-11-11 write-ups/PLDI/paper.tex
Smoothing and surgery.

@@
   morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then},
-  escapeinside={(*@}{@*)}
+  escapeinside={(*@}{@*)},
+  numbers=left
 }
@@
 \newcommand{\getlocalsize}{\mathsf{get\_global\_id}}
+\newcommand{\getglobalsize}{\mathsf{get\_global\_size}}
+
+\newcommand{\keyword}[1]{\mathsf{#1}}
 
@@
 
-We use two examples of irregular data-parallel algorithms---a work
-stealing queue and breadth-first graph search---to introduce the
-industry-standard OpenCL programming model on which we base cooperative
-kernels.
+We provide background on the industry-standard OpenCL programming
+model on which we base cooperative kernels (\mysec\ref{sec:opencl}),
+and illustrate OpenCL using two irregular data-parallel algorithms---a
+work stealing queue and breadth-first graph search---that also
+pinpoint the problems associated with unfair scheduling of workgroups
+and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}.
+
+\subsection{OpenCL Background}
 
@@
 % illustrated over the two examples.
-
-\begin{figure}
-
-\begin{lstlisting}
-kernel work_stealing(global Task *pools) {
-  while (more_work(pools)) {
-    int pool_id = get_group_id();
-    Task *t = pop_or_steal(pools, pool_id);
-    if (t) {
-      // task processing may create more work
-      process_task(t, pools, pool_id);
-    }
-  }
-}
-\end{lstlisting}
-
-\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
-\end{figure}
-
-\begin{figure}
-
-\begin{lstlisting}
-kernel graph_app(global graph *g, 
-                 global nodes *n0, global nodes *n1) {
-
-  int level = 0;
-  global nodes * in_nodes = n0;
-  global nodes * out_nodes = n1;
-
-  int tid = get_global_id();
-  int stride = get_global_size();
-
-(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
-
-    for (int i = tid; i < nodes.size; i += stride) {
-      process_node(g, in_nodes[i], out_nodes, level);
-    }
-
-(*@\label{line:graph:swap}@*)    swap(in_nodes, out_nodes);
-(*@\label{line:graph:gb1}@*)    global_barrier();
-(*@\label{line:graph:reset}@*)    reset(out_nodes);
-    level++;
-(*@\label{line:graph:gb2}@*)    global_barrier();
-  }
-}
-\end{lstlisting}
-\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
-\end{figure}
-
-\paragraph{OpenCL kernels}
 
@@
 id are related via the equation $\getglobalid() = \getgroupid() \times
-\getlocalsize() + \getlocalid()$.
-
-\ADComment{Maybe cut?} Within a workgroup, threads may also be
-organised into \emph{subgroups} (termed \emph{warps} and
-\emph{wavefronts} on NVIDIA and AMD architectures, respectively).
+\getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize()$, which is equal to $\getlocalsize()\times\getnumgroups()$.
+
+The execution of all threads in a workgroup can be synchronised via a
+workgroup barrier, which also ensures memory consistency.  This
+barrier is a \emph{workgroup-level} function: it must be encountered
+uniformly by all or none of the threads of a workgroup, thus can only
+appear in conditional code if all threads in a workgroup that reach
+the barrier agree on the guards of all enclosing conditional
+statements.  As discussed below, a \emph{global} barrier
+(synchronising all threads running a kernel) is \emph{not} provided as
+a primitive.
 
 \paragraph{Memory spaces and memory model}
-An OpenCL kernel has access to four memory spaces.  Shared virtual
-memory (SVM) is accessible to all threads, and to the host
-application.  Global memory is shared among all threads executing a
+An OpenCL kernel has access to four memory spaces.  \emph{Shared virtual
+memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
 kernel, but can only be safely accessed by the host application before
-or after kernel execution.  Local memory is shared among all threads
-in the same workgroup: each workgroup has its own portion of local
-memory, which can be used for fast intra-workgroup communication.
-Each thread has an allocation of very fast private memory, where
+or after kernel execution.  Each workgroup has its own portion of \emph{local},
+which can be used for fast intra-workgroup communication, and every thread has
+an allocation of very fast \emph{private} memory, where
 function-local variables are allocated.
@@
 the case of SVM, is enabled via a set of atomic data types and
-operations, based on the C11 memory model~\cite{...}.  In particular,
+operations, based on the C11 memory model~\cite{...}.  \ADComment{Could cut:} In particular,
 atomic SVM operations enable fine-grained communication between host
@@
 
+\paragraph{Execution model}
+
+OpenCL specifically makes no guarantees about fair scheduling between
+workgroups executing the same kernel, stating~\cite{...}: \emph{``A
+  conforming implementation may choose to serialize the workgroups so
+  a correct algorithm cannot assume that workgroups will execute in
+  parallel.  There is no safe and portable way to synchronize across
+  the independent execution of workgroups since once in the work-pool,
+  they can execute in any order.''}  OpenCL is not alone in this
+regard: CUDA similarly provides no guarantees~\cite{...}.
+
+HSA provides limited, one-way guarantees,
+stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
+  for values written by work-group B without deadlock provided ... A
+  comes after B in work-group flattened ID order}. However, this
+one-way communication is not sufficient so support either the
+work stealing or graph application examples of
+Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
+above, both require \emph{symmetric} communication between
+threads.
+
+\subsection{Motivating Examples}\label{sec:openclexamples}
+
+\begin{figure}
+
+\begin{lstlisting}
+kernel work_stealing(global Task * pools) {
+  while (more_work(pools)) {
+    int pool_id = get_group_id();
+    Task * t = pop_or_steal(pools, pool_id);
+    if (t) {
+      // task processing may create more work
+      process_task(t, pools, pool_id);
+    }
+  }
+}
+\end{lstlisting}
+
+\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
+\end{figure}
+
+\begin{figure}
+
+\begin{lstlisting}
+kernel graph_app(global graph * g, 
+                 global nodes * n0, global nodes * n1) {
+  int level = 0;
+  global nodes * in_nodes = n0;
+  global nodes * out_nodes = n1;
+
+  int tid = get_global_id();
+  int stride = get_global_size();
+
+(*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
+
+    for (int i = tid; i < nodes.size; i += stride) {
+      process_node(g, in_nodes[i], out_nodes, level);
+    }
+
+(*@\label{line:graph:swap}@*)    swap(in_nodes, out_nodes);
+(*@\label{line:graph:gb1}@*)    global_barrier();
+(*@\label{line:graph:reset}@*)    reset(out_nodes);
+    level++;
+(*@\label{line:graph:gb2}@*)    global_barrier();
+  }
+}
+\end{lstlisting}
+\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
+\end{figure}
+
 \paragraph{Work stealing example}
-Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
-kernel that uses a work stealing queue.
 %
@@
 than terminating---tries to \emph{steal} a task from an other
-pool. Hence, workgroups keep on processing tasks until all the work is
+pool. Workgroups keep on processing tasks until all the work is
 done.
 
-\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation
-of a work stealing algorithm. All threads start their execution at the
+\myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
+kernel that uses a work stealing queue. All threads start their execution at the
 function marked with the $\mathsf{kernel}$ keyword. They receive a
@@
 $\mathsf{pop\_or\_steal}$ function is called by all threads of a
-workgroup, although inside only the thread with a local id equal to
+workgroup.  Inside $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to
 $0$, also called the \emph{master thread} of the workgroup,
@@
 workgroup. If a task was obtained, then all threads of the workgroup
-participate in the processing of the task inside the
+participate in the processing of the task via the
 $\mathsf{process\_task}$ function. \HEComment{Manage to say that a
@@
 $\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
-mutex which is implemented using atomic compare and swap operations on
+mutex, implemented using atomic compare and swap operations on
 global memory.  As a result, the kernel presents two opportunities for
@@
 on a fair scheduler, which, as discussed in Section~\ref{sec:intro},
-is not provided by current GPU programming models.  In
-Section~\ref{sec:cooperativeexamples} we show how the example can be
-updated to use our cooperative threads programming model to resolve
-the scheduling issue.
-
-\ADComment{Describe some
-  aspects of the example, saying that the global barrier is not a
-  primitive, and can be coded up using the memory model but is not
-  safe due to the execution model.  Make it clear that the barrier
-  uses busy-waiting.}
+is not provided by current GPU programming models.
 
 \paragraph{Graph traversal example} Figure~\ref{fig:graphsearch} shows the
-skeleton of a frontier based graph traversal algorithm (this approach
-has been shown to be efficient on GPUs in~\ref{...}). The kernel is
+skeleton of a frontier-based graph traversal algorithm; such algorithms have
+been shown to be execute efficiently on GPUs~\cite{...}.
+The kernel is
 given three arguments in global memory: a graph structure, and two
-arrays that can contain graph nodes. Initially, {\tt n0} contains the
-starting nodes to process. Threads keep three local variables: {\tt
-  level} which indicates the frontier level which is currently being
-processed, {\tt in\_nodes} and {\tt out\_nodes} which point to
-distinct node arrays. {\tt in\_nodes} contains the nodes to be
-processed for the current frontier, while {\tt out\_nodes} contains
+arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
+starting nodes to process. Threads keep three local variables: $\keyword{level}$, which indicates the frontier level currently being
+processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$, which point to
+distinct node arrays. The nodes to be
+processed for the current frontier are contained in $\keyword{in\_nodes}$, while $\keyword{out\_nodes}$ contains
 the nodes to be processed during the next frontier.
@@
 retrieves its global thread id and the total number of threads
-(obtained through the {\tt get\_global\_id()} and {\tt
-  get\_global\_size()} OpenCL primitives). A thread then processes
-each node associated with id (i.e.\ indices with its thread id value
-added to any integer multiple of the total number of threads up to the
-list size). A thread processes a node via the {\tt process\_node}
+(obtained through the $\getglobalid()$ and $\getglobalsize()$ primitives). A thread then processes
+each node associated with its id, i.e.\ nodes at indices that are a multiple of the total number of threads, plus the thread's id. A thread processes a node via the $\keyword{process\_node}$
 function, which may (a) push nodes to process in the next frontier to
-the {\tt out\_nodes} list and (b) use the frontier level in
-computation. After processing the frontier, the threads swap their
-local node array pointers (line~\ref{line:graph:swap}), i.e.\ the {\tt
-  out\_nodes} of the current frontier becomes the input of the next
+$\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
+the computation. \ADComment{Tyler: should swap take the addresses of the pointers?} After processing the frontier, the threads swap their
+local node array pointers (line~\ref{line:graph:swap}), i.e.\ the $\keyword{out\_nodes}$ of the current frontier becomes the $\keyword{in\_nodes}$ of the next
 frontier.
@@
 barrier can be implemented~\cite{...}. These implementations use well
-defined atomic memory operations to ensure memory consistency and use
-spinning to ensure threads wait at the barrier for the remaining
-threads to reach the barrier. Even though such barriers appear to run
-reliably on current GPUs, they violate GPU language specifications,
-which do not provide fair scheduling of workgroups. These barriers
-require fair workgroup scheduling in order to function correctly. That
-is, \emph{all} workgroups must be fairly scheduled in order to reach
-the barrier, otherwise, the workgroups waiting at the barrier will not
-make progress.
-
-\paragraph{Execution model}
-
-Two barrier primitives are provided in OpenCL.  A workgroup barrier
-synchronises all threads in a workgroup and ensures memory
-consistency.  This barrier is a \emph{workgroup-level} function: it
-must be encountered uniformly by all or none of the threads of a
-workgroup, thus can only appear in conditional code if all threads in
-a workgroup that reach the barrier agree on the guards of all
-enclosing conditional statements.  A subgroup barrier works
-analogously at the subgroup level.  As discussed below, a
-\emph{global} barrier (synchronising all threads running a kernel) is
-not provided as a primitive.
-
-OpenCL specifically makes no guarantees about fair scheduling between
-workgroups executing the same kernel, stating~\cite{...}: \emph{``A
-  conforming implementation may choose to serialize the workgroups so
-  a correct algorithm cannot assume that workgroups will execute in
-  parallel.  There is no safe and portable way to synchronize across
-  the independent execution of workgroups since once in the work-pool,
-  they can execute in any order.''}  OpenCL is not alone in this
-regard: CUDA similarly provides no guarantees~\cite{...}.
-
-HSA provides limited, one-way guarantees,
-stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
-  for values written by work-group B without deadlock provided ... A
-  comes after B in work-group flattened ID order}. However, this
-one-way communication is not sufficient so support either the
-work stealing or graph application examples of
-Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
-above, both require \emph{symmetric} communication between
-threads. \ADComment{Tyler, can you state the HSA guarantee?},
-\TSComment{How is this?}~\cite{...}.
-
-Revisiting the examples of Figures~\ref{fig:workstealing}
-and~\ref{fig:graphsearch}: the mutex function of
-Figure~\ref{fig:workstealing} will cause starvation if a workgroup
-that holds the mutex is not scheduled again and so does not release
-the mutex.  A similar issue affects the global barrier function of
-Figure~\ref{fig:graphsearch}.  Because the barrier is implemented
-using busy-waiting, a set of workgroups waiting at the barrier may be
-continuously scheduled, starving a workgroup that has not yet arrived
-at the barrier and meaning that no workgroup proceeds past the
-barrier.
-
-\section{The Cooperative Kernels Programming Model}\label{sec:cooperativekernels}
+defined atomic memory operations to ensure memory consistency, and
+employ spinning to ensure threads wait at the barrier until all
+threads have arrived.  This spinning behaviour requires fair
+scheduling between workgroups for the barrier to operate correctly,
+and as a result the barrier is \emph{not} guaranteed to work correctly
+under current GPU programming models.  Such barriers appear to run
+reliably on current GPUs for kernels that are executed with no more
+workgroups than there are compute units; this is due to the fairness
+of the occupancy-bound execution model that current GPUs appear to
+provide.  But, as discussed in \mysec\ref{sec:intro}, this
+occupancy-bound model is not mandated by language standards or vendor
+implementations, and is unlikely to be guaranteed in the future.
+
+In Section~\ref{sec:cooperativeexamples} we show how the work stealing
+and graph traversal examples of Figures~\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
+updated to use our cooperative kernels programming model to resolve
+the scheduling issue.
+
+\section{Cooperative Kernels}\label{sec:cooperativekernels}
 
@@
 
-As with a regular OpenCL kernel (see Section~\ref{sec:background}), a
+As with a regular OpenCL kernel (see Section~\ref{sec:opencl}), a
 cooperative kernel is launched by the host application.  The host
@@
 we explain how the work stealing and graph traveral algorithms of
-Section~\ref{sec:background} can be turned into cooperative kernels
+Section~\ref{sec:openclexamples} can be turned into cooperative kernels
 that use the extensions.  We present the semantics intuitively, using English.
@@
 The $\getgroupid()$ function retrieves the id of a workgroup as usual
-(see Section~\ref{sec:background}).  When executed by a cooperative
+(see Section~\ref{sec:opencl}).  When executed by a cooperative
 kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
@@
 uniformly by all threads in a workgroup (see
-Section~\ref{sec:background}).
+Section~\ref{sec:opencl}).
 
@@
 
-\ADComment{TODO.  We want to show the examples from Section~\ref{sec:background} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}
+\ADComment{TODO.  We want to show the examples from Section~\ref{sec:openclexamples} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}
 

### 2016-11-11 write-ups/PLDI/paper.tex
Address get_global_size comment

@@
 
-\subsection{OpenCL Background}
+\subsection{OpenCL Background}\label{sec:opencl}
 
@@
 the value returned by $\getnumgroups()$ is not fixed during the
-lifetime of a cooperative kernel \TSComment{Neither is {\tt
-    get\_global\_size}, is that worth mentioning?}.  Although
+lifetime of a cooperative kernel.  Although
 $\getnumgroups()$ has atomic semantics, and so can be called at any
@@
 successive calls may differ if workgroups have executed $\offerkill$,
-$\offerfork$ or $\resizingglobalbarrier$ in-between.  The programmer
+$\offerfork$ or $\resizingglobalbarrier$ in-between.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
+a function of $\getnumgroups()$, the value returned by $\getglobalsize()$ is also subject to change.
+The programmer
 must therefore write their cooperative kernel in a manner that is
-robust to changes in the value returned by $\getnumgroups()$.
+robust to changes in the values returned by these functions.
 
@@
 barrier interval, i.e.\ the result returned by repeated calls to
-$\getnumgroups()$ will be identical during such an interval.
+$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
 

### 2016-11-11 write-ups/PLDI/paper.tex
Good progress with Section 3, but all a bit in flux right now.

@@
 We describe the semantics of cooperative kernels
-(Section~\ref{sec:semantics}), then summarise the important
+(Section~\ref{sec:semantics}) and summarise the important
 differences between programming with regular vs.\ cooperative kernels
-(Section~\ref{sec:programmingguidelines}) and the non-functional
+by showing how the motivating examples of \mysec\ref{sec:openclexamples} can be adapted to use cooperative kernels
+(Section~\ref{sec:programmingguidelines}).
+We then focus on the non-functional
 properties that a developer of a cooperative kernel, and the
@@
 (Section~\ref{sec:nonfunctional}).  The semantics of cooperative
-kernels has been guided by the applications that we have studied
+kernels has been guided by the practical applications we have studied
 (described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
-might have taken different, equally reasonable, semantic decisions (Section~\ref{sec:semanticalternatives}).
+might have taken different and also reasonable semantic decisions (Section~\ref{sec:semanticalternatives}).
 We conclude the section by discussing the backwards-compatible nature
@@
 
-We now explain the semantics of these language extensions, after which
-we explain how the work stealing and graph traveral algorithms of
-Section~\ref{sec:openclexamples} can be turned into cooperative kernels
-that use the extensions.  We present the semantics intuitively, using English.
-In Appendix~\ref{appendix:semantics} we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.
+We now explain the semantics of these language extensions.  We present the semantics intuitively, using English.
+In Appendix~\ref{appendix:semantics} (see supplementary material) we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.
 
@@
 
-\ADComment{Maybe refer back to motivating examples?}
 Many irregular algorithms require synchronisation across workgroups.
-This can be achieved via a \emph{global barrier}, but a global barrier
-is not provided as a primitive in OpenCL, CUDA or HSA.  This is
-because, as discussed in Section~\ref{sec:intro}, an inter-workgroup barrier
-requires fair scheduling of workgroups.
+This can be achieved via a \emph{global barrier}, as illustrated in the graph traversal example of \mysec\ref{sec:openclexamples} and \myfig\ref{fig:graphsearch}.
+Because a global barrier requires fair scheduling of workgroups it cannot be safely implemented in OpenCL, CUDA or HSA, and thus is not provided as a primitive in any of these programming models.
 
@@
 
-\lstset{basicstyle=\tt}
+\lstset{basicstyle=\tt,numbers=none}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\lstset{basicstyle=\scriptsize\tt}
+\lstset{basicstyle=\scriptsize\tt,numbers=left}
 
@@
 
-\paragraph{Examples}
+\subsection{Programming With Coopeative Kernels}\label{sec:programmingguidelines}
+
+\ADComment{Revise: we're folding this into the overview of the examples.}
+
+We remark on some issues that programmer working with cooperative
+kernels should be aware of.
+
+\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
+the value returned by $\getnumgroups()$ is not fixed during the
+lifetime of a cooperative kernel.  Although
+$\getnumgroups()$ has atomic semantics, and so can be called at any
+time to yield a well-defined value, the value returned by two
+successive calls may differ if workgroups have executed $\offerkill$,
+$\offerfork$ or $\resizingglobalbarrier$ in-between.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
+a function of $\getnumgroups()$, the value returned by $\getglobalsize()$ is also subject to change.
+The programmer
+must therefore write their cooperative kernel in a manner that is
+robust to changes in the values returned by these functions.
+
+\paragraph{Cooperative kernels that only use barriers}
+
+The situation is more stable if a cooperative kernel does not directly
+issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
+only calls to $\resizingglobalbarrier$ can cause the number of active
+workgroups to change and, as discussed in Section~\ref{sec:semantics},
+this function causes synchronisation between all threads and causes
+the number of active workgroups to change atomically.  At any point
+during execution, the threads of a kernel are executing between some
+pair of resizing barrier calls, which we call a \emph{resizing barrier
+  intervals} (here we consider the kernel entry and exit points
+conceptually to be special cases of resizing barriers).  The number of
+workgroups executing the kernel is thus constant within each resizing
+barrier interval, i.e.\ the result returned by repeated calls to
+$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
+
+This guarantee can be exploited by algorithms that perform strided
+data processing.  \ADComment{Tyler, can we give an example of a data
+  processing loop that would have to be changed?} \TSComment{Done,
+its in figure~\ref{fig:cgraphsearch}}
+
+\paragraph{Placing the new constructs}
+
+It is up to the programmmer to decide where to place our new language
+constructs.  This must be done bearing in mind both the semantics of
+the constructs, as detailed in Section~\ref{ }, as well as the
+non-functional requirements of cooperative kernels that we describe in
+Section~\ref{sec:nonfunctional}.
+
+We note, however, that the programming model and associated language
+constructs were inspired by the requirements of the irregular GPU
+algorithms proposed in the literature, and that at least for these
+algorithms it is usually easy to identify where to place the
+constructs.
+
+The irregular algorithms we encountered either employ work stealing or
+operate on graph data structures.
+
+Work stealing algorithms have a transactional flavour, whereby tasks
+are repeatedly pulled from a queue and processed by a workgroup.  The
+point of task completion---the end of a transaction---is a natural
+place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to
+indicate both that the workgroup could gracefully exit computation,
+but also that there is more work to be processed and that further
+workgroups would be useful in order to process this work.  Such
+algorithms usually require little state to persist between tasks;
+typically only the pointers to the queue data structures need to be
+pre-initialised, by annotating them with $\transmit$.
+\ADComment{Refer back to cooperative versions of examples in
+  Section~\ref{sec:semantics}.}
+
+The level-by-level approach employed by many graph algorithms tends to
+involve global barrier synchronisation.  In some cases, a global
+barrier is required in the midst of a transaction, in which case
+$\globalbarrier$, but not $\resizingglobalbarrier$, is suitable for
+achieving synchronization without disturbing the set of active
+workgroups.  But often it is the case that on completing a level of
+the graph algorithm, the next level could be processed by more or
+fewer workgroups, something which $\resizingglobalbarrier$
+facilitates.  Again, only a small amount of state is relevant to new
+workgroups joining the computation, such as the value of the level
+being processed; this relevant state must be annotated with
+$\transmit$.
+
+\ADComment{Fold this in:}
+
+Using coopeative kernels requires some programmer effort
+
+\ADComment{Lifted:} , after which
+we explain how the work stealing and graph traveral algorithms of
+Section~\ref{sec:openclexamples} can be turned into cooperative kernels
+that use the extensions.
 
@@
 
-\subsection{Programming Guidelines}\label{sec:programmingguidelines}
-
-We remark on some issues that programmer working with cooperative
-kernels should be aware of.
-
-\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
-the value returned by $\getnumgroups()$ is not fixed during the
-lifetime of a cooperative kernel.  Although
-$\getnumgroups()$ has atomic semantics, and so can be called at any
-time to yield a well-defined value, the value returned by two
-successive calls may differ if workgroups have executed $\offerkill$,
-$\offerfork$ or $\resizingglobalbarrier$ in-between.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
-a function of $\getnumgroups()$, the value returned by $\getglobalsize()$ is also subject to change.
-The programmer
-must therefore write their cooperative kernel in a manner that is
-robust to changes in the values returned by these functions.
-
-\paragraph{Cooperative kernels that only use barriers}
-
-The situation is more stable if a cooperative kernel does not directly
-issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
-only calls to $\resizingglobalbarrier$ can cause the number of active
-workgroups to change and, as discussed in Section~\ref{sec:semantics},
-this function causes synchronisation between all threads and causes
-the number of active workgroups to change atomically.  At any point
-during execution, the threads of a kernel are executing between some
-pair of resizing barrier calls, which we call a \emph{resizing barrier
-  intervals} (here we consider the kernel entry and exit points
-conceptually to be special cases of resizing barriers).  The number of
-workgroups executing the kernel is thus constant within each resizing
-barrier interval, i.e.\ the result returned by repeated calls to
-$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
-
-This guarantee can be exploited by algorithms that perform strided
-data processing.  \ADComment{Tyler, can we give an example of a data
-  processing loop that would have to be changed?} \TSComment{Done,
-its in figure~\ref{fig:cgraphsearch}}
-
-\paragraph{Placing the new constructs}
-
-It is up to the programmmer to decide where to place our new language
-constructs.  This must be done bearing in mind both the semantics of
-the constructs, as detailed in Section~\ref{ }, as well as the
-non-functional requirements of cooperative kernels that we describe in
-Section~\ref{sec:nonfunctional}.
-
-We note, however, that the programming model and associated language
-constructs were inspired by the requirements of the irregular GPU
-algorithms proposed in the literature, and that at least for these
-algorithms it is usually easy to identify where to place the
-constructs.
-
-The irregular algorithms we encountered either employ work stealing or
-operate on graph data structures.
-
-Work stealing algorithms have a transactional flavour, whereby tasks
-are repeatedly pulled from a queue and processed by a workgroup.  The
-point of task completion---the end of a transaction---is a natural
-place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to
-indicate both that the workgroup could gracefully exit computation,
-but also that there is more work to be processed and that further
-workgroups would be useful in order to process this work.  Such
-algorithms usually require little state to persist between tasks;
-typically only the pointers to the queue data structures need to be
-pre-initialised, by annotating them with $\transmit$.
-\ADComment{Refer back to cooperative versions of examples in
-  Section~\ref{sec:semantics}.}
-
-The level-by-level approach employed by many graph algorithms tends to
-involve global barrier synchronisation.  In some cases, a global
-barrier is required in the midst of a transaction, in which case
-$\globalbarrier$, but not $\resizingglobalbarrier$, is suitable for
-achieving synchronization without disturbing the set of active
-workgroups.  But often it is the case that on completing a level of
-the graph algorithm, the next level could be processed by more or
-fewer workgroups, something which $\resizingglobalbarrier$
-facilitates.  Again, only a small amount of state is relevant to new
-workgroups joining the computation, such as the value of the level
-being processed; this relevant state must be annotated with
-$\transmit$.
-
 \subsection{Non-Functional Requirements of Cooperative Kernels}\label{sec:nonfunctional}

### 2016-11-11 write-ups/PLDI/paper.tex
Surgery

@@
 
-\ADComment{Revise: we're folding this into the overview of the examples.}
-
-We remark on some issues that programmer working with cooperative
-kernels should be aware of.
+Working with cooperative kernels requires some manual effort on the
+part of the programmer: they must place the $\transmit$ qualifier and
+use the new language primitives described in
+\mysec\ref{sec:semantics}, and account for slight changes in the OpenCL semantics induced by these new features.
+
+The programming model and associated language constructs were inspired
+by the requirements of the irregular GPU algorithms proposed in the
+literature, and that at least for these algorithms it is usually easy
+to identify where to place the constructs.
+
+We discuss the main semantic change induced by our programming
+model---that the number of workgroups executing a kernel varies
+dynamically---and illustrate our programming model and the impact of
+this semantic change by adapting the work stealing and graph traversal
+algorithms of \mysec\ref{sec:openclexamples} to use cooperative
+kernels.
 
@@
 robust to changes in the values returned by these functions.
+
+\paragraph{Adapting work stealing}
+
+\myfiglong\ref{fig:workstealing-cooperative} shows a cooperative
+version of the work stealing example of \myfig\ref{...} that we
+discussed in \mysec\ref{sec:openclexamples}.
+\ADComment{TODO: go through with Hugues; where are the offer functions?}
+
+For the work stealing example, there is no state to transmit since a
+computation is entirely parametrized by a task, which is retrieved from
+a task pool located in global memory. Care must be taken on the dynamic
+change of the number of workgroups though: we cannot assume that the
+total number of task pools is equal to $\getnumgroups$. Thus, the number
+of pools must be given as an argument, and the $\mathsf{pool}$ argument
+must be obtained from the group id modulo the total number of
+pools. This total number of pools must also be passed to
+$\mathsf{pop\_or\_steal}$, since when a workgroup wants to steal it must
+access other pool indexes, which are bounded by the total number of
+pools.
+
+\begin{figure}
+
+\begin{lstlisting}
+kernel work_stealing(global Task *pools, const int num_pools) {
+  while (more_work(pools)) {
+    int pool_id = get_group_id() % num_pools;
+    Task *t = pop_or_steal(pools, pool_id, num_pools);
+    if (t) {
+      // task processing may create more work
+      process_task(t, pools, pool_id);
+    }
+  }
+}
+\end{lstlisting}
+
+\caption{Cooperative version of the work stealing example from \myfig\ref{fig:XXX}; the number of workpools
+must be explicitly passed as arguments since the number of workgroups
+is not a constant anymore}\label{fig:workstealing-cooperative}
+\end{figure}
 
@@
 This guarantee can be exploited by algorithms that perform strided
-data processing.  \ADComment{Tyler, can we give an example of a data
-  processing loop that would have to be changed?} \TSComment{Done,
-its in figure~\ref{fig:cgraphsearch}}
-
-\paragraph{Placing the new constructs}
-
-It is up to the programmmer to decide where to place our new language
-constructs.  This must be done bearing in mind both the semantics of
-the constructs, as detailed in Section~\ref{ }, as well as the
-non-functional requirements of cooperative kernels that we describe in
-Section~\ref{sec:nonfunctional}.
-
-We note, however, that the programming model and associated language
-constructs were inspired by the requirements of the irregular GPU
-algorithms proposed in the literature, and that at least for these
-algorithms it is usually easy to identify where to place the
-constructs.
-
-The irregular algorithms we encountered either employ work stealing or
-operate on graph data structures.
-
-Work stealing algorithms have a transactional flavour, whereby tasks
-are repeatedly pulled from a queue and processed by a workgroup.  The
-point of task completion---the end of a transaction---is a natural
-place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to
-indicate both that the workgroup could gracefully exit computation,
-but also that there is more work to be processed and that further
-workgroups would be useful in order to process this work.  Such
-algorithms usually require little state to persist between tasks;
-typically only the pointers to the queue data structures need to be
-pre-initialised, by annotating them with $\transmit$.
-\ADComment{Refer back to cooperative versions of examples in
-  Section~\ref{sec:semantics}.}
-
-The level-by-level approach employed by many graph algorithms tends to
-involve global barrier synchronisation.  In some cases, a global
-barrier is required in the midst of a transaction, in which case
-$\globalbarrier$, but not $\resizingglobalbarrier$, is suitable for
-achieving synchronization without disturbing the set of active
-workgroups.  But often it is the case that on completing a level of
-the graph algorithm, the next level could be processed by more or
-fewer workgroups, something which $\resizingglobalbarrier$
-facilitates.  Again, only a small amount of state is relevant to new
-workgroups joining the computation, such as the value of the level
-being processed; this relevant state must be annotated with
-$\transmit$.
-
-\ADComment{Fold this in:}
-
-Using coopeative kernels requires some programmer effort
-
-\ADComment{Lifted:} , after which
-we explain how the work stealing and graph traveral algorithms of
-Section~\ref{sec:openclexamples} can be turned into cooperative kernels
-that use the extensions.
-
-\begin{figure}
-
-\begin{lstlisting}
-kernel graph_app(global graph *g, 
-                 global nodes *n0, global nodes *n1) {
-
-(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
-(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
-(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
-
-  while(in_nodes.size > 0) {
-
-(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
-(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
-
-    for (int i = tid; i < nodes.size; i += stride) {
-      process_node(g, in_nodes[i], out_nodes, level);
-    }
-
-    level++;
-    swap(in_nodes, out_nodes);
-    reset(out_nodes);
-(*@\label{line:cgraph:resizing}@*)    resizing_global_barrier();
-  }
-}
-\end{lstlisting}
-\caption{Cooperative kernel port for Figure~\ref{fig:graphsearch}}\label{fig:cgraphsearch}
-\end{figure}
-
-\ADComment{TODO.  We want to show the examples from Section~\ref{sec:openclexamples} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}
-
-In practice, we found that transforming traditional blocking GPU
-kernels into cooperative kernels was straightforward. For example,
-Figure~\ref{fig:cgraphsearch} shows the graph traversal example of
-Figure~\ref{fig:graphsearch} transformed into a cooperative kernel.
-On line~\ref{line:cgraph:resizing}, we change the original barrier
+data processing.  We now illustrate this using the graph traversal motivating example.
+
+\paragraph{Adapting graph traversal} \ADComment{Tyler, the original had two global barriers; the port has just one resizing barrier.  Is this right?}
+\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the graph traversal kernel
+of \myfig\ref{fig:graphsearch}, introduced in
+\mysec\ref{sec:openclexamples}.  
+On line~\ref{line:cgraph:resizing}, we change the original global barrier
 into a resizing barrier. Several variables are marked to be
@@
 (lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
-\ref{line:cgraph:transmit3}): {\tt level} must be restored so that
-new workgroups know which frontier they are processing; {\tt
-  in\_nodes} and {\tt out\_nodes} must be restored so that new
+\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so that
+new workgroups know which frontier they are processing; $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so that new
 workgroups know which of the node arrays are input and output. Lastly,
@@
 \ref{line:cgraph:rechunking2}).
-
-\TSComment{Can Hugues talk about work stealing here?}
-
-For the work stealing example, there is no state to transmit since a
-computation is entirely parametrized by a task, which is retrieved from
-a task pool located in global memory. Care must be taken on the dynamic
-change of the number of workgroups though: we cannot assume that the
-total number of task pools is equal to $\getnumgroups$. Thus, the number
-of pools must be given as an argument, and the $\mathsf{pool}$ argument
-must be obtained from the group id modulo the total number of
-pools. This total number of pools must also be passed to
-$\mathsf{pop\_or\_steal}$, since when a workgroup wants to steal it must
-access other pool indexes, which are bounded by the total number of
-pools.
+This example exploits the fact that the cooperative kernel does not issue $\offerkill$ nor $\offerfork$ directly: \ADComment{Actually I'm confused and need to discuss :-)}
 
@@
 \begin{lstlisting}
-kernel work_stealing(global Task *pools, const int num_pools) {
-  while (more_work(pools)) {
-    int pool_id = get_group_id() % num_pools;
-    Task *t = pop_or_steal(pools, pool_id, num_pools);
-    if (t) {
-      // task processing may create more work
-      process_task(t, pools, pool_id);
-    }
+kernel graph_app(global graph *g, 
+                 global nodes *n0, global nodes *n1) {
+
+(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
+(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
+(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
+
+  while(in_nodes.size > 0) {
+
+(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
+(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
+
+    for (int i = tid; i < nodes.size; i += stride) {
+      process_node(g, in_nodes[i], out_nodes, level);
+    }
+
+    level++;
+    swap(in_nodes, out_nodes);
+    reset(out_nodes);
+(*@\label{line:cgraph:resizing}@*)    resizing_global_barrier();
   }
@@
 \end{lstlisting}
-
-\caption{Cooperative version of work stealing: the number of workpools
-must be explicitely passed as arguments since the number of workgroups
-is not a constant anymore}\label{fig:workstealing-cooperative}
+\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing global barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
 \end{figure}
+
+In \mysec\ref{ } we describe the set of irregular GPU algorithms used
+in our experiments, which largely captures the irregular blocking
+algorithms that are available as open source GPU kernels.  These all
+employ either work stealing or operate on graph data structures.
+Like the simplified example of \myfig\ref{ }, the work stealing algorithms have a transactional flavour
+and require little or no state to be carried between transactions.
+Thus the point at which a workgroup is ready to process a new task is a natural point to place $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.
+%
+The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
+most level-by-level graph algorithms.  These algorithms tend to
+require more state to be annotated with $\transmit$, and in some cases
+a global barrier is required in the midst of a transaction, in which
+case $\globalbarrier$, but not $\resizingglobalbarrier$, is suitable
+for achieving synchronization without disturbing the set of active
+workgroups.  But often it is the case that on completing a level of
+the graph algorithm, the next level could be processed by more or
+fewer workgroups, something which $\resizingglobalbarrier$
+facilitates.
+
+In summary, our experience so far indicates that cooperative kernels
+is a good fit for current irregular algorithms.  Of cousre this is, in
+part, because we designed the model with these algorithms in mind, but
+we argue that the fit is natural not just for the examples that we
+study, but more generally for at least these two patterns of irregular
+algorithm.
 

### 2016-11-11 write-ups/PLDI/paper.tex
Fixed missing refs.

@@
 
-In Section~\ref{sec:cooperativeexamples} we show how the work stealing
+In Section~\ref{sec:programmingguidelines} we show how the work stealing
 and graph traversal examples of Figures~\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
@@
 \myfiglong\ref{fig:workstealing-cooperative} shows a cooperative
-version of the work stealing example of \myfig\ref{...} that we
+version of the work stealing example of \myfig\ref{fig:workstealing} that we
 discussed in \mysec\ref{sec:openclexamples}.
@@
 
-\caption{Cooperative version of the work stealing example from \myfig\ref{fig:XXX}; the number of workpools
+\caption{Cooperative version of the work stealing example from \myfig\ref{fig:workstealing}; the number of workpools
 must be explicitly passed as arguments since the number of workgroups
@@
 
-In \mysec\ref{ } we describe the set of irregular GPU algorithms used
+In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
 in our experiments, which largely captures the irregular blocking
@@
 employ either work stealing or operate on graph data structures.
-Like the simplified example of \myfig\ref{ }, the work stealing algorithms have a transactional flavour
+Like the simplified example of \myfig\ref{fig:workstealing-cooperative}, the work stealing algorithms have a transactional flavour
 and require little or no state to be carried between transactions.

### 2016-11-11 write-ups/PLDI/paper.tex
Use section macros.

@@
 suggests that this is straightforward in practice.  We discuss this in
-detail in Section~\ref{sec:portingalgorithms}.
+detail in \mysec\ref{sec:portingalgorithms}.
 
@@
 \item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms.
-We describe the semantics of the programming model in detail (Section~\ref{sec:cooperativekernels}).
+We describe the semantics of the programming model in detail (\mysec\ref{sec:cooperativekernels}).
 
@@
   kernels on top of the OpenCL 2.0 programming model
-  (Section~\ref{sec:implementation}).
+  (\mysec\ref{sec:implementation}).
 
@@
   the responsiveness to high priority tasks that cooperation enables,
-  across four GPUs from two vendors (Section~\ref{sec:experiments}).
+  across four GPUs from two vendors (\mysec\ref{sec:experiments}).
 
@@
 We begin by providing an overview of the OpenCL programming model and
-two motivating examples of irregular algorithms (Section~\ref{sec:background}).
-At the end of the paper we discuss related work (Section~\ref{sec:relatedwork}) and avenues for future work (Section~\ref{sec:conclusion}).
+two motivating examples of irregular algorithms (\mysec\ref{sec:background}).
+At the end of the paper we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).
 
@@
 and device.  Our implementation of cooperative kernels
-(Section~\ref{sec:implementation}) depends on this recent OpenCL
+(\mysec\ref{sec:implementation}) depends on this recent OpenCL
 feature.
@@
 kernel loop to obtain a task to process.  The algorithm thus depends
-on a fair scheduler, which, as discussed in Section~\ref{sec:intro},
+on a fair scheduler, which, as discussed in \mysec\ref{sec:intro},
 is not provided by current GPU programming models.
@@
 
-In Section~\ref{sec:programmingguidelines} we show how the work stealing
+In \mysec\ref{sec:programmingguidelines} we show how the work stealing
 and graph traversal examples of Figures~\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
@@
 We describe the semantics of cooperative kernels
-(Section~\ref{sec:semantics}) and summarise the important
+(\mysec\ref{sec:semantics}) and summarise the important
 differences between programming with regular vs.\ cooperative kernels
 by showing how the motivating examples of \mysec\ref{sec:openclexamples} can be adapted to use cooperative kernels
-(Section~\ref{sec:programmingguidelines}).
+(\mysec\ref{sec:programmingguidelines}).
 We then focus on the non-functional
@@
 implementer of a scheduler for cooperative kernels, should strive for
-(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
+(\mysec\ref{sec:nonfunctional}).  The semantics of cooperative
 kernels has been guided by the practical applications we have studied
-(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
-might have taken different and also reasonable semantic decisions (Section~\ref{sec:semanticalternatives}).
+(described in \mysec\ref{sec:portingalgorithms}), and we discuss several cases where we
+might have taken different and also reasonable semantic decisions (\mysec\ref{sec:semanticalternatives}).
 We conclude the section by discussing the backwards-compatible nature
@@
 run on current GPU platforms
-(Section~\ref{sec:backwardscompatibility}).
+(\mysec\ref{sec:backwardscompatibility}).
 
@@
 
-As with a regular OpenCL kernel (see Section~\ref{sec:opencl}), a
+As with a regular OpenCL kernel (see \mysec\ref{sec:opencl}), a
 cooperative kernel is launched by the host application.  The host
@@
 The $\getgroupid()$ function retrieves the id of a workgroup as usual
-(see Section~\ref{sec:opencl}).  When executed by a cooperative
+(see \mysec\ref{sec:opencl}).  When executed by a cooperative
 kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
@@
 uniformly by all threads in a workgroup (see
-Section~\ref{sec:opencl}).
+\mysec\ref{sec:opencl}).
 
@@
 possible alternative choices, in
-Section~\ref{sec:semanticalternatives}.
+\mysec\ref{sec:semanticalternatives}.
 
@@
 portion of its local state to the threads of the forked workgroups.
-We discuss in Section~\ref{sec:semanticalternatives} the rationale for
+We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
 having all new threads obtain this local state portion from thread 0,
@@
 cooperative kernels programming model.  However, we discuss in
-Section~\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
+\mysec\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
 implemented significantly more efficiently if it is treated as a
@@
 only calls to $\resizingglobalbarrier$ can cause the number of active
-workgroups to change and, as discussed in Section~\ref{sec:semantics},
+workgroups to change and, as discussed in \mysec\ref{sec:semantics},
 this function causes synchronisation between all threads and causes
@@
 
-The semantics presented in Section~\ref{sec:semantics} describe the envelope of
+The semantics presented in \mysec\ref{sec:semantics} describe the envelope of
 behaviour that a developer of a cooperative kernel should be prepared

### 2016-11-11 write-ups/PLDI/paper.tex
Use figure macros.

@@
 
-\paragraph{Graph traversal example} Figure~\ref{fig:graphsearch} shows the
+\paragraph{Graph traversal example} \mylongfig\ref{fig:graphsearch} shows the
 skeleton of a frontier-based graph traversal algorithm; such algorithms have
@@
 In \mysec\ref{sec:programmingguidelines} we show how the work stealing
-and graph traversal examples of Figures~\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
+and graph traversal examples of \myfigs\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
 updated to use our cooperative kernels programming model to resolve

### 2016-11-11 write-ups/PLDI/paper.tex
Oops.

@@
 
-\paragraph{Graph traversal example} \mylongfig\ref{fig:graphsearch} shows the
+\paragraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
 skeleton of a frontier-based graph traversal algorithm; such algorithms have

### 2016-11-11 write-ups/PLDI/paper.tex
Tweaks.

@@
 pinpoint the problems associated with unfair scheduling of workgroups
-and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}.
+and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}).
 
@@
 $0$, also called the \emph{master thread} of the workgroup,
-\ADComment{Hugues, I changed; please check:} will try to pop a task
+will try to pop a task
 from the workgroup's pool, or steal a task from a different pool. The
@@
 participate in the processing of the task via the
-$\mathsf{process\_task}$ function. \HEComment{Manage to say that a
-  task is considered big enough to benefit from parallel computation}
-\ADComment{I think that would be too much detail given the space we
-  have, so I would not delve into this. Remove both comments if you
-  agree.} Processing this task may lead to the creation of further
+$\mathsf{process\_task}$ function.  Processing this task may lead to the creation of further
 tasks, which the master thread will push to the workgroup's pool.
@@
 kernels model, and the programmer of a cooperative kernel, must strive
-to meet a number of non-functional requirements, which we outline
-here.
-
-\ADComment{Sree mentioned TimeGraph paper.}
+to meet the following non-functional requirements.
 
@@
 
-Recall that the purpose of $\offerkill$ is to provide the scheduler
-with an opportunity to destroy a workgroup in order to schedule
+The purpose of $\offerkill$ is to provide the scheduler with an
+opportunity to destroy a workgroup in order to schedule
 higher-priority tasks.  Because these higher-priority tasks, such as
@@
 
-\ADComment{Forward reference to experiments with display freezing.}
+Using work stealing as an example: in the kernel of \myfig\ref{...}, a
+workgroup offers itself to the scheduler after processing each task.
+If tasks turn out to be very short then it might turn out to be
+reasonable to invoke $\offerkill$ less frequently, after every $k$
+tasks for some appropriate $k$.  On the other hand, if individual
+tasks may be very time-consuming then it may be necessary to rewrite
+the algorithm so work on a larger number of shorter tasks, so as to
+achieve a higher rate of calls to $\offerfork$.
+
+Getting this non-functional requirement right depends on the
+constraints of the tasks to be co-scheduled with a cooperative kernel,
+on properties of the cooperative kernel itself, and on the efficiency
+of the GPU that is used for execution.  In \mysec\ref{...} we conduct
+experiments to understand the response rate that would be required to
+co-schedule graphics rendering with a cooperative kernel in order to
+achieve a smooth frame rate, and we evalaute responsiveness in
+practice in \mysec\ref{...}.
+
+\paragraph{Sufficient workgroups for the cooperative kernel}
+
+Recall that the cooperative kernel would ideally utilise the $N$
+workgroups requested upon kernel launch, in order to conduct its
+processing efficiently.  The scheduler should thus aim to provide with
+cooperative kernel with as many workgroups, up to this number, as
+other constraints allow.  As a result, the scheduler should generally
+accept an $\offerkill$ only if a comput unit is required for another
+task, and should respond positively to $\offerfork$ calls by
+scheduling new workgroups if compute units are available.  Our
+implementation of cooperative scheduling (\mysec\ref{ }) is as
+generous as possible to the cooperative kernel in this regard.
+However, these are not hard-and-fast rules: a scheduler might employ
+heuristics based on historical workloads to avoid thrashing.  For
+example, if graphics calls are known to arrive at a certain rate then
+the scheduler might decide that it is worth accepting an $\offerkill$
+from the cooperative kernel on the assumption that a graphics task
+will be requested momentarily.  Similarly, the scheduler might provide
+some, but not all, compute units to the cooperative kernel in response
+to an $\offerfork$ call if it has reason to believe that the remaining
+compute units are likely to be useful for other high priority tasks
+that are likely to arrive imminently.
+
+\subsection{Alternative Semantic Choices}\label{sec:semanticalternatives}
+
+\paragraph{Killing order}
+
+We opted for a semantics whereby only the active workgroup with the
+highest id can be killed.  This has an appealing property: it means
+that the ids of active workgroups are contiguous, which is important
+for processing of contiguous data.  The cooperative graph traversal
+algorithm of \myfig\ref{ } illustrates this: the algorithm is prepared
+for $\getglobalsize()$ to change after each resizing barrier call, but
+depends on the fact that $\getglobalid()$ returns a contiguous range
+of thread ids.
 
@@
 
-\ADComment{Generosity: initial number of workgroups, and responding to offer fork.}
-
-\subsection{Alternative Semantic Choices}\label{sec:semanticalternatives}
+\paragraph{Keeping one workgroup alive}
 

### 2016-11-11 write-ups/PLDI/paper.tex
Fixed some refs.

@@
 \ref{line:cgraph:rechunking2}).
-This example exploits the fact that the cooperative kernel does not issue $\offerkill$ nor $\offerfork$ directly: \ADComment{Actually I'm confused and need to discuss :-)}
+\ADComment{Tyler: can you please add the line number references in the following?} This example exploits the fact that the cooperative kernel does not issue $\offerkill$ nor $\offerfork$ directly: the value of $\keyword{stride}$ obtained from $\getglobalsize()$ at line~\ref{...} is stable until the next resizing barrier at line~\ref{...}, and in particular can be safely used inside the $\keyword{for}$ loop.
 
@@
 
-\subsection{Non-Functional Requirements of Cooperative Kernels}\label{sec:nonfunctional}
+\subsection{Non-Functional Requirements}\label{sec:nonfunctional}
 
@@
 
-Using work stealing as an example: in the kernel of \myfig\ref{...}, a
+Using work stealing as an example: in the kernel of \myfig\ref{fig:workstealing-cooperative}, a
 workgroup offers itself to the scheduler after processing each task.
@@
 on properties of the cooperative kernel itself, and on the efficiency
-of the GPU that is used for execution.  In \mysec\ref{...} we conduct
+of the GPU that is used for execution.  In \mysec\ref{sec:representativeworkloads} we conduct
 experiments to understand the response rate that would be required to
@@
 achieve a smooth frame rate, and we evalaute responsiveness in
-practice in \mysec\ref{...}.
+practice in \mysec\ref{sec:responsiveness}.
 
@@
 scheduling new workgroups if compute units are available.  Our
-implementation of cooperative scheduling (\mysec\ref{ }) is as
+implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
 generous as possible to the cooperative kernel in this regard.
@@
 for processing of contiguous data.  The cooperative graph traversal
-algorithm of \myfig\ref{ } illustrates this: the algorithm is prepared
+algorithm of \myfig\ref{fig:cgraphsearch} illustrates this: the algorithm is prepared
 for $\getglobalsize()$ to change after each resizing barrier call, but
@@
 
-\ADComment{Got here.}
+\ADComment{TODO: Discuss downside, which is that if WG with high id is
+  doing a chunky task, a WG with a lower id may be ready to be killed
+  but cannot be taken.  Work stealing algorithms, which do not depend
+  on thread id, could take advantage of this.}
 
@@
 
-\ADComment{Explain here where we could have done things differently.}
-
-\ADComment{What are difficulties with other alternatives and why we didn't use them.}
-
-\ADComment{Allowing zero to be killed; the following is some text we had before:}
-
-\paragraph{Keeping at least one workgroup alive}
+\ADComment{TODO: To simplify programming, avoiding the need to wrap all kills in conditionals, we always keep one workgroup alive.  This fits well with our decision to kill WGs in decreasing order.}
+
+\ADComment{TODO: There are possible use cases for allowing all WGs to be killed, though, and we might wish to allow them.}
+
+\paragraph{Transmission of partial state from a single thread}
+
+\ADComment{TODO: explain why we have thread 0 transmit its state, and transmitting only annotated variables.}
+
+\ADComment{Something old to maybe work in:}
 
@@
 remaining computation to be performed if competing workloads allow it.
-
-\ADComment{Allowing arbitrary workgroups to be killed (not just the largest workgroup)}
-
-\ADComment{Having thread 0 transmit its state, and transmitting only annotated variables.}
 
@@
 
-\subsection{Determining Representative Workloads}
+\subsection{Determining Representative Workloads}\label{sec:representativeworkloads}
 
@@
 
-\subsection{Performance and Responsiveness of Cooperative Scheduling}
+\subsection{Performance and Responsiveness of Cooperative Scheduling}\label{sec:responsiveness}
 

### 2016-11-11 write-ups/PLDI/paper.tex
Spelling

@@
 more \emph{kernels} that run on accelerator devices---GPUs in the
-context of this paper.  A kernel is writen in OpenCL C, based on C99.
+context of this paper.  A kernel is written in OpenCL C, based on C99.
 All threads executing a kernel execute the same entry function with
@@
 The idea is that while having the scheduler be permitted to terminate
-the execution of workgroups in an arbitary fashion would be drastic,
+the execution of workgroups in an arbitrary fashion would be drastic,
 the kernel may contain identifiable points of execution at which a
@@
 
-\subsection{Programming With Coopeative Kernels}\label{sec:programmingguidelines}
+\subsection{Programming With Cooperative Kernels}\label{sec:programmingguidelines}
 
@@
 For the work stealing example, there is no state to transmit since a
-computation is entirely parametrized by a task, which is retrieved from
+computation is entirely parameterised by a task, which is retrieved from
 a task pool located in global memory. Care must be taken on the dynamic
@@
 In summary, our experience so far indicates that cooperative kernels
-is a good fit for current irregular algorithms.  Of cousre this is, in
+is a good fit for current irregular algorithms.  Of course this is, in
 part, because we designed the model with these algorithms in mind, but
@@
 co-schedule graphics rendering with a cooperative kernel in order to
-achieve a smooth frame rate, and we evalaute responsiveness in
+achieve a smooth frame rate, and we evaluate responsiveness in
 practice in \mysec\ref{sec:responsiveness}.
@@
 other constraints allow.  As a result, the scheduler should generally
-accept an $\offerkill$ only if a comput unit is required for another
+accept an $\offerkill$ only if a compute unit is required for another
 task, and should respond positively to $\offerfork$ calls by
@@
 
-\subsection{The Overhad of Moving to Cooperative Kernels}
+\subsection{The Overhead of Moving to Cooperative Kernels}
 
@@
 \cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
-  SMs, not to workgroups, so that tasks taht would benefit from being
+  SMs, not to workgroups, so that tasks that would benefit from being
   processed by the same SM can be scheduled to the same SM.  Has

### 2016-11-11 write-ups/PLDI/paper.tex
More semantic differences.

@@
 
-\ADComment{TODO: Discuss downside, which is that if WG with high id is
-  doing a chunky task, a WG with a lower id may be ready to be killed
-  but cannot be taken.  Work stealing algorithms, which do not depend
-  on thread id, could take advantage of this.}
+A disadvantage of this decision is that it may provide sub-optimal
+responsiveness from the point of view of the scheduler.  Suppose the
+scheduler requires an additional compute unit, but the active thread
+with the larget id is processing some computationally intensive work
+and will take a while to reach $\offerkill$.  Our semantics means that
+the scheuduler cannot take advantage of the fact that another active
+workgroup may invoke $\offerkill$ sooner.
+
+Cooperative kernels that do not require contiguous thread ids might me
+more suited to a semantics in which workgroups can be killed in any
+order, but where workgroup ids (and thus thread global ids) are not
+guaranteed to be contiguous.
 

### 2016-11-11 write-ups/PLDI/paper.tex
Add line references for workstealing examples

--- 2016-11-11 write-ups/PLDI/paper.tex
@@
 \begin{lstlisting}
-kernel work_stealing(global Task * pools) {
+(*@\label{line:wksteal:kernelfunc}@*) kernel work_stealing(global Task * pools) {
   while (more_work(pools)) {
-    int pool_id = get_group_id();
-    Task * t = pop_or_steal(pools, pool_id);
+(*@\label{line:wksteal:getgroupid}@*)    int pool_id = get_group_id();
+(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(pools, pool_id);
     if (t) {
       // task processing may create more work
-      process_task(t, pools, pool_id);
+(*@\label{line:wksteal:processtask}@*)      process_task(t, pools, pool_id);
     }
@@
 %
-The work stealing paradigm is useful when a computation can be split into multiple
-tasks that can be processed independently, but where the number of
-tasks is not known ahead of time because processing a given task may
-create an arbitrary number of new tasks. Work stealing enables dynamic
-balancing of tasks across several processing units, OpenCL workgroups
-in our case. Each workgroup has an associated task pool from which it
-obtains tasks to process, and to which it stores newly created
-tasks. When a workgroup's pool becomes empty, a workgroup---rather
-than terminating---tries to \emph{steal} a task from an other
-pool. Workgroups keep on processing tasks until all the work is
-done.
+The work stealing paradigm is useful when a computation can be split
+into multiple tasks that can be processed independently, but where the
+number of tasks is not known ahead of time because processing a given
+task may create an arbitrary number of new tasks. Work stealing enables
+dynamic balancing of tasks across several processing units, OpenCL
+workgroups in our case. Each workgroup has an associated task pool from
+which it obtains tasks to process, and to which it stores newly created
+tasks. When a workgroup's pool becomes empty, a workgroup---rather than
+terminating---tries to \emph{steal} a task from an other
+pool. Workgroups keep on processing tasks until all the work is done.
 
 \myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
-kernel that uses a work stealing queue. All threads start their execution at the
-function marked with the $\mathsf{kernel}$ keyword. They receive a
-pointer to the task pools, which reside in global memory, and which
-were initialized by the host program beforehand to contain the initial
-tasks. Each thread can identify the workgroup it belongs to via the
-function $\getgroupid()$.  Here the group id is stored into the
-variable $\mathsf{pool\_id}$ (which is, by default, in private memory)
-and used to access the relevant task pool. The
-$\mathsf{pop\_or\_steal}$ function is called by all threads of a
-workgroup.  Inside $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to
-$0$, also called the \emph{master thread} of the workgroup,
-will try to pop a task
-from the workgroup's pool, or steal a task from a different pool. The
-master thread uses local memory inside $\mathsf{pop\_or\_steal}$ to
-ensure that the returned value is valid for all threads of its
-workgroup. If a task was obtained, then all threads of the workgroup
-participate in the processing of the task via the
-$\mathsf{process\_task}$ function.  Processing this task may lead to the creation of further
-tasks, which the master thread will push to the workgroup's pool.
+kernel that uses a work stealing queue. All threads start their
+execution at the function marked with the $\mathsf{kernel}$ keyword
+(line~\ref{line:wksteal:kernelfunc}). They receive a pointer to the task
+pools, which reside in global memory, and which were initialized by the
+host program beforehand to contain the initial tasks. Each thread can
+identify the workgroup it belongs to via the function $\getgroupid()$
+(line~\ref{line:wksteal:getgroupid}). Here the group id is stored into
+the variable $\mathsf{pool\_id}$ (which is, by default, in private
+memory) and used to access the relevant task pool. The
+$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
+is called by all threads of a workgroup.  Inside
+$\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
+also called the \emph{master thread} of the workgroup, will try to pop a
+task from the workgroup's pool, or steal a task from a different
+pool. The master thread uses local memory inside
+$\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
+all threads of its workgroup. If a task was obtained, then all threads
+of the workgroup participate in the processing of the task via the
+$\mathsf{process\_task}$ function (line~\ref{line:wksteal:processtask}).
+Processing this task may lead to the creation of further tasks, which
+the master thread will push to the workgroup's pool.
 

### 2016-11-11 write-ups/PLDI/paper.tex
More semantic differences.

--- 2016-11-11 write-ups/PLDI/paper.tex
@@
 Semantically, when all threads have reached $\resizingglobalbarrier$,
-the number of active workgroups, $M$, is set to a new value, $M'$ say.
+the number of active workgroups, $M$, is set to a new value, $M'$ say, with $M' > 0$.
 If $M' = M$ then the active workgroups remain unchanged and proceed to
@@
 
-\ADComment{TODO: To simplify programming, avoiding the need to wrap all kills in conditionals, we always keep one workgroup alive.  This fits well with our decision to kill WGs in decreasing order.}
-
-\ADComment{TODO: There are possible use cases for allowing all WGs to be killed, though, and we might wish to allow them.}
-
-\paragraph{Transmission of partial state from a single thread}
-
-\ADComment{TODO: explain why we have thread 0 transmit its state, and transmitting only annotated variables.}
-
-\ADComment{Something old to maybe work in:}
-
-The programmer is responsible for keeping at least one workgroup alive.
-In particular, if all workgroups execute $\offerkill$, the scheduler is at liberty to destroy all workgroups, causing the cooperative kernel to exit.
-This can be guarded against by wrapping each $\offerkill$ call in a conditional statement:
-
+Our semantics dictate that the workgroup with id 0 will not be killed
+if it invokes $\offerkill$.  This avoids the possibility of the
+cooperative kernel terminating early due to the programmer
+inadvertantly allowing all workgroups to be killed, and the decision
+to keep workgroup 0 alive fits well with our choice to kill workgroups
+in descending order of id.
+
+However, there might be a use case for a cooperative kernel reaching a
+point where it would be acceptable for the kernel to exit, although
+desirable for some remaining computation to be performed if competing
+workloads allow it.
+In this case, a semantics where all workgroups can be killed via $\offerkill$ would be appropriate, and the programmer would need to guard each $\offerkill$ with an id check in cases where killing all workgroups would be unacceptable.
+For example:
+%
 \lstset{basicstyle=\tt,numbers=none}
@@
 \lstset{basicstyle=\scriptsize\tt,numbers=left}
-
-We could, alternatively, have formulated the semantics of $\offerkill$
-so that workgroup 0 is never killed.  However, there might be a use
-case for a cooperative kernel reaching a point where it would be
-acceptable for the kernel to exit, although desirable for some
-remaining computation to be performed if competing workloads allow it.
+%
+would ensure that at least workgroup 0 is kept alive.
+
+\paragraph{Transmission of partial state from a single thread}
+
+\ADComment{TODO: explain why we have thread 0 transmit its state, and transmitting only annotated variables.}
 

### 2016-11-11 write-ups/PLDI/paper.tex
More semantic differences.

@@
 
-\ADComment{TODO: explain why we have thread 0 transmit its state, and transmitting only annotated variables.}
+Recall from the semantics of $\offerfork$ that newly forked workgroups
+inherit the variable valuation associated with thread 0 of the forking
+workgroup, but only for $\transmit$-annotated variables.  Alternative
+choices here would be to have forked workgroups inherit values for
+\emph{all} variables from the forking workgroup, and to have thread
+$i$ in the forking workgroup provide the valuation for thread $i$ in
+each spawned workgroup, rather than having thread 0 transmit the
+valuation to all new threads.
+
+\ADComment{TODO: justify our choice: uniform variables and efficiency.}
 

### 2016-11-11 write-ups/PLDI/paper.tex
Add better discussion with line references in cooperative version of work stealing

--- 2016-11-11 write-ups/PLDI/paper.tex
@@
 \begin{lstlisting}
-(*@\label{line:wksteal:kernelfunc}@*) kernel work_stealing(global Task * pools) {
+(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * pools) {
+(*@\label{line:wksteal:getgroupid}@*)  int pool_id = get_group_id();
   while (more_work(pools)) {
-(*@\label{line:wksteal:getgroupid}@*)    int pool_id = get_group_id();
 (*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(pools, pool_id);
@@
 discussed in \mysec\ref{sec:openclexamples}.
-\ADComment{TODO: go through with Hugues; where are the offer functions?}
 
@@
 computation is entirely parameterised by a task, which is retrieved from
-a task pool located in global memory. Care must be taken on the dynamic
-change of the number of workgroups though: we cannot assume that the
-total number of task pools is equal to $\getnumgroups$. Thus, the number
-of pools must be given as an argument, and the $\mathsf{pool}$ argument
-must be obtained from the group id modulo the total number of
-pools. This total number of pools must also be passed to
-$\mathsf{pop\_or\_steal}$, since when a workgroup wants to steal it must
-access other pool indexes, which are bounded by the total number of
-pools.
+a task pool located in global memory. In the main loop, before trying to
+obtain a task, we make the workgroup offer itself to be forked or killed
+(lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
+that a workgroup may be killed even if its associated task pool is not
+empty, since remaining tasks could eventually be stealed by other
+workgroups.  Care must be taken on computing pool identifiers: the
+original version in \myfig\ref{fig:workstealing} can rely on the fact
+that there are as many pools as workgroups, but here the number of
+workgroups may change during execution. Therefore, the number of pools
+must now be explicitely given as an argument
+(line~\ref{line:cwksteal:argnumpool}) and used as an upper bound of pool
+indexes when scanning pools in $\mathsf{more\_work}$
+(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
+(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
+workgroup's pool index (line~\ref{line:cwksteal:getgroupid}) must not
+only be bounded by the number of pools but also be moved inside the main
+loop since a forked workgroup that starts to execute after $\offerfork$
+must re-compute its pool index.
 
@@
 \begin{lstlisting}
-kernel work_stealing(global Task *pools, const int num_pools) {
-  while (more_work(pools)) {
-    int pool_id = get_group_id() % num_pools;
-    Task *t = pop_or_steal(pools, pool_id, num_pools);
+kernel work_stealing(global Task * pools,
+(*@\label{line:cwksteal:argnumpool}@*)                     const int num_pools) {
+(*@\label{line:cwksteal:morework}@*)  while (more_work(pools, num_pools)) {
+(*@\label{line:cwksteal:offerfork}@*)    offer_fork();
+(*@\label{line:cwksteal:offerkill}@*)    offer_kill();
+(*@\label{line:cwksteal:getgroupid}@*)    int pool_id = get_group_id() % num_pools;
+(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(pools, pool_id, num_pools);
     if (t) {
       // task processing may create more work
-      process_task(t, pools, pool_id);
+(*@\label{line:cwksteal:processtask}@*)      process_task(t, pools, pool_id);
     }

### 2016-11-11 write-ups/PLDI/paper.tex

@@
 \lstset{basicstyle=\scriptsize\tt,numbers=left}
-
-We could, alternatively, have formulated the semantics of $\offerkill$
-so that workgroup 0 is never killed.  However, there might be a use
-case for a cooperative kernel reaching a point where it would be
-acceptable for the kernel to exit, although desirable for some
-remaining computation to be performed if competing workloads allow it.
+%
+would ensure that at least workgroup 0 is kept alive.
+
+\paragraph{Transmission of partial state from a single thread}
+
+Recall from the semantics of $\offerfork$ that newly forked workgroups
+inherit the variable valuation associated with thread 0 of the forking
+workgroup, but only for $\transmit$-annotated variables.  Alternative
+choices here would be to have forked workgroups inherit values for
+\emph{all} variables from the forking workgroup, and to have thread
+$i$ in the forking workgroup provide the valuation for thread $i$ in
+each spawned workgroup, rather than having thread 0 transmit the
+valuation to all new threads.
+
+\ADComment{TODO: justify our choice: uniform variables and efficiency.}
 

### 2016-11-11 src/test_suite/final_octree/octree.h
Almost working

@@ -1,11 +1,7 @@
-#include "octree_common.h"
-
-/*---------------------------------------------------------------------------*/
-
 DEFINE_int32(numParticles, 10000, "number of particles to treat");
 DEFINE_int32(maxChildren, 20, "maximum number of children");
-DEFINE_int32(threads, 64, "number of threads");
-//DEFINE_int32(num_iterations, 300, "number of iterations");
+DEFINE_int32(threads, 128, "number of threads");
+DEFINE_int32(pool_size, 300, "size of task pools");
 
 // ---
 DEFINE_string(restoration_ctx_path, "test_suite/final_octree/", "Path to restoration context");
@@ -13,22 +9,30 @@ DEFINE_string(merged_kernel_file, "test_suite/final_octree/device/merged.cl", "p
 DEFINE_string(persistent_kernel_file, "test_suite/final_octree/device/standalone.cl", "path to the standalone mega kernel file");
 
 /*---------------------------------------------------------------------------*/
+
+typedef struct {
+  cl_float4 middle;
+  cl_bool flip;
+  cl_uint end;
+  cl_uint beg;
+  cl_uint treepos;
+} Task;
+
+/*---------------------------------------------------------------------------*/
 // global vars
 
 const unsigned int MAXTREESIZE = 11000000;
-const unsigned int maxlength = 256;
-int max_workgroups = 0;
+int num_workgroups = 0;
 
-//cl::Buffer d_num_iterations;
-cl::Buffer randdata;
 cl::Buffer particles;
 cl::Buffer newParticles;
 cl::Buffer tree;
 cl::Buffer treeSize;
 cl::Buffer particlesDone;
-cl::Buffer stealAttempts;
-cl::Buffer deq;
-cl::Buffer dh;
+cl::Buffer pools;
+cl::Buffer task_pool_lock;
+cl::Buffer pool_head;
+Task *h_pools = NULL;
 
 /*---------------------------------------------------------------------------*/
 // Specific to octree
@@ -40,6 +44,8 @@ double genrand_real1(void);
 
 void generate_particles(CL_Execution *exec)
 {
+  printf("Enter %s%\n", __func__);
+
   cl_float4* lparticles = new cl_float4[FLAGS_numParticles];
   char fname[256];
[…]

### 2016-11-11 src/test_suite/final_octree/octree_common.h

@@ -9,10 +9,3 @@ typedef struct {
 } Task;
 
 /*---------------------------------------------------------------------------*/
-
-typedef struct {
-  ATOMIC_CL_INT_TYPE tail;
-  ATOMIC_CL_INT_TYPE head;
-} DequeHeader;
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/final_octree/octree.h
Seems to be working for good

@@ -1,7 +1,7 @@
-DEFINE_int32(numParticles, 10000, "number of particles to treat");
-DEFINE_int32(maxChildren, 20, "maximum number of children");
+DEFINE_int32(numParticles, 5000000, "number of particles to treat");
+DEFINE_int32(maxChildren, 10, "maximum number of children");
 DEFINE_int32(threads, 128, "number of threads");
-DEFINE_int32(pool_size, 300, "size of task pools");
+DEFINE_int32(pool_size, 2000, "size of task pools");
 
 // ---
 DEFINE_string(restoration_ctx_path, "test_suite/final_octree/", "Path to restoration context");
@@ -21,7 +21,7 @@ typedef struct {
 /*---------------------------------------------------------------------------*/
 // global vars
 
-const unsigned int MAXTREESIZE = 11000000;
+const unsigned int MAXTREESIZE = 50000000;
 int num_workgroups = 0;
 
 cl::Buffer particles;

### 2016-11-11 src/test_suite/final_octree/octree.h
Fix number of workgroups wrt occupancy

@@ -116,7 +116,7 @@ void reset_persistent_task(CL_Execution *exec) {
 // Empty for Pannotia apps
 void init_persistent_app_for_real(CL_Execution *exec, int occupancy) {
   printf("Enter %s%\n", __func__);
-  num_workgroups = occupancy - 1;
+  num_workgroups = occupancy;
 
   particles = cl::Buffer(exec->exec_context, CL_MEM_READ_WRITE, sizeof(cl_float4) * FLAGS_numParticles);
   newParticles = cl::Buffer(exec->exec_context, CL_MEM_READ_WRITE, sizeof(cl_float4) * FLAGS_numParticles);

### 2016-11-11 src/test_suite/final_octree/octree.h
Remove debugging prints

@@ -44,8 +44,6 @@ double genrand_real1(void);
 
 void generate_particles(CL_Execution *exec)
 {
-  printf("Enter %s%\n", __func__);
-
   cl_float4* lparticles = new cl_float4[FLAGS_numParticles];
   char fname[256];
   snprintf(fname, 256, "octreecacheddata-%dparticles.dat", FLAGS_numParticles);
@@ -94,7 +92,6 @@ const char* persistent_kernel_name() {
 /*---------------------------------------------------------------------------*/
 
 void reset_persistent_task(CL_Execution *exec) {
-  printf("Enter %s%\n", __func__);
   // re-write 0 to the CL buffers, etc
   int err = 0;
 
@@ -108,14 +105,12 @@ void reset_persistent_task(CL_Execution *exec) {
   check_ocl(err);
 
   generate_particles(exec);
-  printf("Exit %s%\n", __func__);
 }
 
 /*---------------------------------------------------------------------------*/
 
 // Empty for Pannotia apps
 void init_persistent_app_for_real(CL_Execution *exec, int occupancy) {
-  printf("Enter %s%\n", __func__);
   num_workgroups = occupancy;
 
   particles = cl::Buffer(exec->exec_context, CL_MEM_READ_WRITE, sizeof(cl_float4) * FLAGS_numParticles);
@@ -140,7 +135,6 @@ void init_persistent_app_for_real(CL_Execution *exec, int occupancy) {
 /*---------------------------------------------------------------------------*/
 
 void set_persistent_app_args_for_real(int arg_index, cl::Kernel k) {
-  printf("Enter %s%\n", __func__);
   // Set args for persistent kernel
 
   check_ocl(k.setArg(arg_index++, particles));
@@ -165,12 +159,10 @@ void set_persistent_app_args_for_real(int arg_index, cl::Kernel k) {
 
 void init_persistent_app_for_occupancy(CL_Execution *exec)
 {
-  printf("Enter %s%\n", __func__);
   // nothing to do for octree
 }
 
 int set_persistent_app_args_for_occupancy(int arg_index, cl::Kernel k) {
-  printf("Enter %s%\n", __func__);
   // Set dummy args for persistent kernel
   int err = 0;
   int num_args_octree = 14;
@@ -183,7 +175,6 @@ int set_persistent_app_args_for_occupancy(int arg_index, cl::Kernel k) {
 }
 
 void output_persistent_solution(const char *fname, CL_Execution *exec) {
[…]

### 2016-11-11 src/test_suite/final_octree/octree.h
Comment code used for debugging pool heads

@@ -215,18 +215,18 @@ bool check_persistent_task(CL_Execution *exec) {
   std::cout << "OCTREE: Particles in tree: " << sum << " (" << FLAGS_numParticles << ") [" << hparticlesDone << "]" << std::endl;
   delete(htree);
 
-  // for debug
-  cl_int *h_pool_head = (cl_int *)calloc(num_workgroups, sizeof(cl_int));
-  if (h_pool_head == NULL) {
-    cout << "calloc failed" << endl;
-    exit(EXIT_FAILURE);
-  }
-  err = exec->exec_queue.enqueueReadBuffer(pool_head, CL_TRUE, 0, num_workgroups * sizeof(cl_uint), h_pool_head);
-  check_ocl(err);
-  for (int i = 0; i < num_workgroups; i++) {
-    printf("pool head %.3d: %u\n", i, h_pool_head[i]);
-  }
-  free(h_pool_head);
+  // // For debug
+  // cl_int *h_pool_head = (cl_int *)calloc(num_workgroups, sizeof(cl_int));
+  // if (h_pool_head == NULL) {
+  //   cout << "calloc failed" << endl;
+  //   exit(EXIT_FAILURE);
+  // }
+  // err = exec->exec_queue.enqueueReadBuffer(pool_head, CL_TRUE, 0, num_workgroups * sizeof(cl_uint), h_pool_head);
+  // check_ocl(err);
+  // for (int i = 0; i < num_workgroups; i++) {
+  //   printf("pool head %.3d: %u\n", i, h_pool_head[i]);
+  // }
+  // free(h_pool_head);
 
   return (sum == FLAGS_numParticles) && (hparticlesDone == FLAGS_numParticles);
 }

### 2016-11-11 src/test_suite/final_octree/octree_common.h
Update merged and standalone kernels, remove old header (useless now)

@@ -1,11 +1,0 @@
-/*---------------------------------------------------------------------------*/
-
-typedef struct {
-  CL_FLOAT4_TYPE middle;
-  CL_BOOL_TYPE flip;
-  CL_UINT_TYPE end;
-  CL_UINT_TYPE beg;
-  CL_UINT_TYPE treepos;
-} Task;
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/interface_octree/common/octree_types.h
Remove old versions of octree

@@ -1,19 +1,0 @@
-
-/*---------------------------------------------------------------------------*/
-
-typedef struct {
-  CL_FLOAT4_TYPE middle;
-  CL_BOOL_TYPE flip;
-  CL_UINT_TYPE end;
-  CL_UINT_TYPE beg;
-  CL_UINT_TYPE treepos;
-} Task;
-
-/*---------------------------------------------------------------------------*/
-
-typedef struct {
-  ATOMIC_CL_INT_TYPE tail;
-  ATOMIC_CL_INT_TYPE head;
-} DequeHeader;
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/interface_octree/common/restoration_ctx.h

@@ -1,5 +1,0 @@
-#pragma once
-typedef struct {
-  CL_UCHAR_TYPE target;
-} Restoration_ctx;
-

### 2016-11-11 src/test_suite/interface_octree/host/octree.h

@@ -1,245 +1,0 @@
-#include "octree_types.h"
-
-/*---------------------------------------------------------------------------*/
-
-DEFINE_int32(numParticles, 10000, "number of particles to treat");
-DEFINE_int32(maxChildren, 20, "maximum number of children");
-DEFINE_int32(threads, 64, "number of threads");
-//DEFINE_int32(num_iterations, 300, "number of iterations");
-
-// see some octree types definitions in common/octree.h
-
-// ---
-DEFINE_string(restoration_ctx_path, "test_suite/interface_octree/common", "Path to restoration context");
-DEFINE_string(merged_kernel_file, "test_suite/interface_octree/device/merged.cl", "path to the merged mega kernel file");
-DEFINE_string(persistent_kernel_file, "test_suite/interface_octree/device/standalone.cl", "the path the mega kernel file");
-
-/*---------------------------------------------------------------------------*/
-// global vars
-
-const unsigned int MAXTREESIZE = 11000000;
-const unsigned int maxlength = 256;
-int max_workgroups = 0;
-
-//cl::Buffer d_num_iterations;
-cl::Buffer randdata;
-cl::Buffer particles;
-cl::Buffer newParticles;
-cl::Buffer tree;
-cl::Buffer treeSize;
-cl::Buffer particlesDone;
-cl::Buffer stealAttempts;
-cl::Buffer deq;
-cl::Buffer dh;
-
-/*---------------------------------------------------------------------------*/
-// Specific to octree
-
-// genrand_real1() is defined in rand.cc
-double genrand_real1(void);
-
-/*---------------------------------------------------------------------------*/
-
-void generate_particles(CL_Execution *exec)
-{
-  cl_float4* lparticles = new cl_float4[FLAGS_numParticles];
-  char fname[256];
-  snprintf(fname, 256, "octreecacheddata-%dparticles.dat", FLAGS_numParticles);
-  FILE* f = fopen(fname, "rb");
-  if (!f) {
-    cout << "Generating and caching data" << endl;
-
-    int clustersize = 100;
-    for (int i = 0; i < (FLAGS_numParticles / clustersize); i++) {
[…]

### 2016-11-11 src/test_suite/mega_octree/common/octree.h

@@ -1,18 +1,0 @@
-/*---------------------------------------------------------------------------*/
-
-typedef struct {
-  CL_FLOAT4_TYPE middle;
-  CL_BOOL_TYPE flip;
-  CL_UINT_TYPE end;
-  CL_UINT_TYPE beg;
-  CL_UINT_TYPE treepos;
-} Task;
-
-/*---------------------------------------------------------------------------*/
-
-typedef struct {
-  ATOMIC_CL_INT_TYPE tail;
-  ATOMIC_CL_INT_TYPE head;
-} DequeHeader;
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/mega_octree/common/restoration_ctx.h

@@ -1,6 +1,0 @@
-#pragma once
-
-typedef struct {
-  CL_INT_TYPE target;
-  CL_INT_TYPE sense;
-} Restoration_ctx;

### 2016-11-11 src/test_suite/mega_octree/host/main.cpp

@@ -1,486 +1,0 @@
-#include <atomic>
-#include <iostream>
-#include <fstream>
-#include <string>
-#include<iostream>
-#include <vector>
-#include <limits.h>
-
-// Must be loaded early because it defines CL_XXX_TYPE
-#include "cl_types.h"
-
-#include "base/commandlineflags.h"
-#include "opencl/opencl.h"
-#include "cl_execution.h"
-#include "discovery.h"
-#include "cl_communicator.h"
-#include "../common/restoration_ctx.h"
-#include "../common/octree.h"
-#include "cl_scheduler.h"
-#include "kernel_ctx.h"
-#include "iw_barrier.h"
-#include "sense_reversal_barrier.h"
-#include "base/file.h"
-
-DEFINE_int32(platform_id, 0, "OpenCL platform ID to use");
-DEFINE_int32(device_id, 0, "OpenCL device ID to use");
-DEFINE_bool(list, false, "List OpenCL platforms and devices");
-DEFINE_string(scheduler_rt_path, "scheduler_rt/rt_device", "Path to scheduler runtime includes");
-DEFINE_string(restoration_ctx_path, "test_suite/mega_octree/common/", "Path to restoration context");
-DEFINE_int32(non_persistent_wgs, 2, "ratio of workgroups to send to non-persistent task. Special values are (-1) to send all but one workgroup and (-2) to send one workgroup");
-DEFINE_int32(skip_tasks, 0, "flag to say if non persistent tasks should be skipped: 0 - don't skip, 1 - skip");
-
-/*===========================================================================*/
-// specific to octree
-
-DEFINE_int32(numParticles, 100000, "number of particles to treat");
-DEFINE_int32(maxChildren, 20, "maximum number of children");
-DEFINE_int32(threads, 64, "number of threads");
-DEFINE_int32(num_iterations, 300, "number of iterations");
-static const unsigned int MAXTREESIZE = 11000000;
-
-// see some octree types definitions in common/octree.h
-
-/*===========================================================================*/
-
-using namespace std;
-
-const char *kernel_file = XSTR(KERNEL_FILE);
-
[…]

### 2016-11-11 src/test_suite/mega_octree_km/common/restoration_ctx.h

@@ -1,9 +1,0 @@
-#pragma once
-typedef struct {
-  CL_UCHAR_TYPE target;
-  unsigned int local_id;
-  unsigned int local_size;
-  unsigned int localStealAttempts;
-  CL_INT_TYPE i;
-} Restoration_ctx;
-

### 2016-11-11 src/test_suite/mega_octree_km/host/main.cpp

@@ -1,489 +1,0 @@
-#include <atomic>
-#include <iostream>
-#include <fstream>
-#include <string>
-#include<iostream>
-#include <vector>
-#include <limits.h>
-
-// Must be loaded early because it defines CL_XXX_TYPE
-#include "scheduler_rt/rt_common/cl_types.h"
-
-#include "base/commandlineflags.h"
-#include "opencl/opencl.h"
-#include "cl_execution.h"
-#include "discovery.h"
-#include "cl_communicator.h"
-#include "../common/restoration_ctx.h"
-#include "../common/octree.h"
-#include "cl_scheduler.h"
-#include "kernel_ctx.h"
-#include "iw_barrier.h"
-#include "base/file.h"
-
-DEFINE_int32(platform_id, 0, "OpenCL platform ID to use");
-DEFINE_int32(device_id, 0, "OpenCL device ID to use");
-DEFINE_bool(list, false, "List OpenCL platforms and devices");
-DEFINE_string(scheduler_rt_path, "scheduler_rt/rt_device", "Path to scheduler runtime includes");
-DEFINE_string(restoration_ctx_path, "test_suite/mega_octree_km/common/", "Path to restoration context");
-DEFINE_int32(non_persistent_wgs, 2, "ratio of workgroups to send to non-persistent task. Special values are (-1) to send all but one workgroup and (-2) to send one workgroup");
-DEFINE_int32(skip_tasks, 0, "flag to say if non persistent tasks should be skipped: 0 - don't skip, 1 - skip");
-
-/*===========================================================================*/
-// specific to octree
-
-DEFINE_int32(numParticles, 100000, "number of particles to treat");
-DEFINE_int32(maxChildren, 20, "maximum number of children");
-DEFINE_int32(threads, 64, "number of threads");
-DEFINE_int32(num_iterations, 300, "number of iterations");
-static const unsigned int MAXTREESIZE = 11000000;
-
-// see some octree types definitions in common/octree.h
-
-/*===========================================================================*/
-
-using namespace std;
-
-const char *kernel_file = XSTR(KERNEL_FILE);
-
-//From IWOCL tutorial (needs attribution)
[…]

### 2016-11-11 src/test_suite/octree/helper.cc

@@ -1,14 +1,0 @@
-#include <iostream>
-
-#include "opencl/opencl.h"
-
-/*---------------------------------------------------------------------------*/
-
-void checkErr(cl_int err, const char * name) {
-  if (err != CL_SUCCESS) {
-    std::cerr << "ERROR: " << name  << " (" << err << ")" << std::endl;
-    exit(EXIT_FAILURE);
-  }
-}
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/octree/helper.h

@@ -1,9 +1,0 @@
-/* requires:
-   #include <CL/cl.hpp>
-*/
-
-/*---------------------------------------------------------------------------*/
-
-void checkErr(cl_int err, const char * name);
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/octree/main.cc

@@ -1,50 +1,0 @@
-#include <stdio.h>
-
-#include "opencl/opencl.h"
-#include "test_suite/octree/shared_host_side.h"
-#include "test_suite/octree/octree.h"
-
-/*---------------------------------------------------------------------------*/
-
-int main(int argc, char* argv[]) {
-
-  if (argc!=5) {
-    printf("\nUsage:\t./octreepart threads blocks particleCount maxChildren\n\n");
-    return 1;
-  }
-
-  int threads = atoi(argv[1]);
-  if (threads<=0||threads>512) {
-    printf("Threads must be between 1 and 128\n");
-    return 1;
-  }
-
-  int blocks = atoi(argv[2]);
-  if (blocks<=0||blocks>512) {
-    printf("Blocks must be between 1 and 512\n");
-    return 1;
-  }
-
-  int particleCount = atoi(argv[3]);
-  if (particleCount<=0||particleCount>50000000) {
-    printf("particleCount must be between 1 and 5000000\n");
-    return 1;
-  }
-
-  int maxChildren = atoi(argv[4]);
-  if (maxChildren<=0||maxChildren>100) {
-    printf("maxChildren must be between 1 and 100\n");
-    return 1;
-  }
-
-  Octree o;
-  o.run(threads, blocks, maxChildren, particleCount);
-
-  printf("Threads: %d Blocks: %d Method: dynamic ParticleCount: %d maxChildren: %d "
-         "MaxMem: %d Time: %f\n",
-         threads, blocks, particleCount, maxChildren,
-         o.getMaxMem(), o.getTime());
-  float err = o.printStats();
-
-  return 0;
-}

### 2016-11-11 src/test_suite/octree/octree.cc

@@ -1,269 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <iostream>
-#include <fstream>
-#include <sstream>
-#include <string>
-#include <streambuf>
-
-#include "opencl/opencl.h"
-#include "test_suite/octree/shared_host_side.h"
-#include "test_suite/octree/octree.h"
-#include "test_suite/octree/helper.h"
-
-/*---------------------------------------------------------------------------*/
-
-float Octree::printStats()
-{
-  // Hugues: all stats are retrieved from device memory at the end of
-  // Octree::run().
-
-  if(htreeSize >= MAXTREESIZE) {
-    printf("Tree too large!\n");
-    return -1;
-  }
-
-  unsigned int sum = 0;
-  for(unsigned int i = 0; i < htreeSize; i++) {
-    //printf("%d\n",htree[i]);
-    if (htree[i] & 0x80000000) {
-      sum += htree[i] & 0x7fffffff;
-    }
-  }
-
-  printf("Tree size: %d\n", htreeSize);
-  printf("Particles in tree: %d (%d) [%d]\n", sum, numParticles, hparticlesDone);
-  printf("Steal attempts: %d\n", hstealAttempts);
-
-  float numf = numParticles;
-  float sumf = sum;
-
-  return (numf - sumf) / numf;
-}
-
-/*---------------------------------------------------------------------------*/
-
-bool Octree::run(unsigned int threads, unsigned int blocks, int maxChildren, int numParticles)
-{
-  this->numParticles = numParticles;
-
-  // Initiate OpenCL
-  cl_int err;
-  cl::Event event;
-
-  // platform
-  std::vector<cl::Platform> platforms;
-  cl::Platform::get(&platforms);
-  if (platforms.size() == 0) {
-    printf("Platform size is 0\n");
-    exit (EXIT_FAILURE);
-  }
-
-  // context
-  cl_context_properties properties[] = { CL_CONTEXT_PLATFORM, (cl_context_properties)(platforms[0])(), 0};
-  cl::Context context(CL_DEVICE_TYPE_GPU, properties);
-
-  // device
-  std::vector<cl::Device> devices = context.getInfo<CL_CONTEXT_DEVICES>();
-
-  // command queue
-  cl::CommandQueue queue(context, devices[0], CL_QUEUE_PROFILING_ENABLE, &err);
-  checkErr(err, "cl::CommandQueue::CommandQueue()");
-
[…]

### 2016-11-11 src/test_suite/octree/octree.h

@@ -1,37 +1,0 @@
-/* requires:
-   #include "lbabp.h"
- */
-
-/*---------------------------------------------------------------------------*/
-
-class Octree
-{
-	static const unsigned int MAXTREESIZE = 11000000;
-	unsigned int numParticles;
-	cl::Buffer particles;
-	cl::Buffer newParticles;
-
-	cl::Buffer tree;
-	cl::Buffer treeSize;
-	cl::Buffer particlesDone;
-  cl::Buffer stealAttempts;
-
-	float totalTime;
-
-        /* stats */
-        int maxMem;
-        unsigned int* htree;
-        unsigned int htreeSize;
-        unsigned int hparticlesDone;
-        unsigned int hstealAttempts;
-
-public:
-
-	void generateParticles(cl::CommandQueue queue);
-	bool run(unsigned int threads, unsigned int blocks, int maxChildren, int numParticles);
-	float printStats();
-	float getTime();
-	int getMaxMem();
-};
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/octree/shared_host_side.h

@@ -1,38 +1,0 @@
-/* requires:
-   #include <CL/cl.hpp>
-*/
-
-/*
- * Host types must be kept in sync with device counterparts, which are
- * declared in the kernel file
- */
-
-/*---------------------------------------------------------------------------*/
-/* Task */
-
-typedef struct {
-  cl_float4 middle;
-  bool flip;
-  unsigned int end;
-  unsigned int beg;
-  unsigned int treepos;
-} Task;
-
-/*---------------------------------------------------------------------------*/
-/* class DLBABP */
-
-typedef struct {
-  volatile int tail;
-  volatile int head;
-} DequeHeader;
-
-/*---------------------------------------------------------------------------*/
-/* DLBABP */
-
-typedef struct {
-  Task *deq;
-  DequeHeader* dh;
-  unsigned int maxlength;
-} DLBABP;
-
-/*---------------------------------------------------------------------------*/

### 2016-11-11 src/test_suite/final_octree/octree.h
Make octree stable

@@ -1,7 +1,7 @@
 DEFINE_int32(numParticles, 5000000, "number of particles to treat");
 DEFINE_int32(maxChildren, 10, "maximum number of children");
 DEFINE_int32(threads, 128, "number of threads");
-DEFINE_int32(pool_size, 2000, "size of task pools");
+DEFINE_int32(pool_size, 10000, "size of task pools");
 
 // ---
 DEFINE_string(restoration_ctx_path, "test_suite/final_octree/", "Path to restoration context");

### 2016-11-11 write-ups/PLDI/paper.tex
fixed cooperative kernel to use 2 barriers and added line numbers

@@
 
-\paragraph{Adapting graph traversal} \ADComment{Tyler, the original had two global barriers; the port has just one resizing barrier.  Is this right?}
-\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the graph traversal kernel
-of \myfig\ref{fig:graphsearch}, introduced in
-\mysec\ref{sec:openclexamples}.  
-On line~\ref{line:cgraph:resizing}, we change the original global barrier
-into a resizing barrier. Several variables are marked to be
-transmitted in the case of workgroups joining at the resizing barrier
+\paragraph{Adapting graph traversal} 
+\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
+graph traversal kernel of \myfig\ref{fig:graphsearch}, introduced in
+\mysec\ref{sec:openclexamples}.  On lines~\ref{line:cgraph:resizing1}
+and ~\ref{line:cgraph:resizing2}, we change the original global
+barriers into a resizing barriers. Several variables are marked to be
+transmitted in the case of workgroups joining at the resizing barriers
 (lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
-\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so that
-new workgroups know which frontier they are processing; $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so that new
-workgroups know which of the node arrays are input and output. Lastly,
-the static work distribution of the original kernel is no longer valid
-in a cooperative kernel. This is because the stride (which is based on
-$M$) may change after each resizing barrier
+\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so
+that new workgroups know which frontier they are processing;
+$\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so
+that new workgroups know which of the node arrays are input and
+output. Lastly, the static work distribution of the original kernel is
+no longer valid in a cooperative kernel. This is because the stride
+(which is based on $M$) may change after each resizing barrier
 call. To fix this, we re-distribute the work after each resizing
@@
 (lines~\ref{line:cgraph:rechunking1} and
-\ref{line:cgraph:rechunking2}).
-\ADComment{Tyler: can you please add the line number references in the following?} This example exploits the fact that the cooperative kernel does not issue $\offerkill$ nor $\offerfork$ directly: the value of $\keyword{stride}$ obtained from $\getglobalsize()$ at line~\ref{...} is stable until the next resizing barrier at line~\ref{...}, and in particular can be safely used inside the $\keyword{for}$ loop.
+\ref{line:cgraph:rechunking2}). This example exploits the fact that
+the cooperative kernel does not issue $\offerkill$ nor $\offerfork$
+directly: the value of $\keyword{stride}$ obtained from
+$\getglobalsize()$ at line~\ref{line:cgraph:rechunking2} is stable
+until the next resizing barrier at line~\ref{line:cgraph:resizing1},
+and in particular can be safely used inside the $\keyword{for}$ loop.
 
@@
 
+    swap(&in_nodes, &out_nodes);
+(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
+    reset(out_nodes);
     level++;
-    swap(in_nodes, out_nodes);
-    reset(out_nodes);
-(*@\label{line:cgraph:resizing}@*)    resizing_global_barrier();
+(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
   }

### 2016-11-11 src/test_suite/connect_four/connect_four.h
Better dimensionning of pools

@@ -9,9 +9,8 @@ DEFINE_string(persistent_kernel_file, "test_suite/connect_four/device/standalone
 
 /*---------------------------------------------------------------------------*/
 
-DEFINE_int32(pools, 10, "Number of task pools");
 // Pool size is quite arbitrary, may be tuned
-DEFINE_int32(pool_size, 20, "Size of a task pool");
+DEFINE_int32(pool_size, 1000, "Size of a task pool");
 DEFINE_int32(maxlevel, 2, "Max level of look-ahead");
 
 DEFINE_string(board_file, "test_suite/connect_four/board.txt", "Path to board description");
@@ -34,6 +33,7 @@ cl::Buffer d_root_done;
 cl::Buffer d_debug_int;
 cl::Buffer d_debug_board;
 size_t board_mem_size;
+cl_int num_workgroups;
 
 /*===========================================================================*/
 // helper functions
@@ -156,18 +156,18 @@ void reset_persistent_task(CL_Execution *exec) {
   check_ocl(err);
 
   // tasks
-  h_task_pool = (Task *)calloc(FLAGS_pools * FLAGS_pool_size, sizeof(Task));
+  h_task_pool = (Task *)calloc(num_workgroups * FLAGS_pool_size, sizeof(Task));
   if (h_task_pool == NULL) {
     cout << "calloc failed" << endl;
     exit(EXIT_FAILURE);
   }
-  err = exec->exec_queue.enqueueWriteBuffer(d_task_pool, CL_TRUE, 0, FLAGS_pools * FLAGS_pool_size * sizeof(Task), h_task_pool);
+  err = exec->exec_queue.enqueueWriteBuffer(d_task_pool, CL_TRUE, 0, num_workgroups * FLAGS_pool_size * sizeof(Task), h_task_pool);
   check_ocl(err);
 
-  err = exec->exec_queue.enqueueFillBuffer(d_task_pool_lock, 0, 0, FLAGS_pools * sizeof(cl_int));
+  err = exec->exec_queue.enqueueFillBuffer(d_task_pool_lock, 0, 0, num_workgroups * sizeof(cl_int));
   check_ocl(err);
 
-  err = exec->exec_queue.enqueueFillBuffer(d_task_pool_head, 0, 0, FLAGS_pools * sizeof(cl_int));
[…]

### 2016-11-11 src/test_suite/connect_four/connect_four.h
Comment out printing for debugging

@@ -270,55 +270,55 @@ bool check_persistent_task(CL_Execution *exec) {
   print_board(h_board);
 
   // Nodes
-  Node *h_nodes = (Node *)calloc(num_node, sizeof(Node));
-  if (h_nodes == NULL) {
-    cout << "calloc failed" << endl;
-    exit(EXIT_FAILURE);
-  }
-  err = exec->exec_queue.enqueueReadBuffer(d_nodes, CL_TRUE, 0, num_node * sizeof(Node), h_nodes);
-  check_ocl(err);
+  // Node *h_nodes = (Node *)calloc(num_node, sizeof(Node));
+  // if (h_nodes == NULL) {
+  //   cout << "calloc failed" << endl;
+  //   exit(EXIT_FAILURE);
+  // }
+  // err = exec->exec_queue.enqueueReadBuffer(d_nodes, CL_TRUE, 0, num_node * sizeof(Node), h_nodes);
+  // check_ocl(err);
 
-  limit = num_node < 70 ? num_node : 70;
-  cout << "Nodes (limited to " << limit << "): " << endl;
-  for (int i = 0; i < limit; i++) {
-    printf("   [%2.2d]", i);
-    printf(" l:%+d", h_nodes[i].level);
-    printf(" p:%+2.2d", h_nodes[i].parent);
-    printf(" v:%+3.3d", h_nodes[i].value);
-    // moves
-    printf(" m[");
-    for (int j = 0; j < h_nodes[i].level; j++) {
-      printf(" %d", h_nodes[i].moves[j]);
-    }
-    printf(" ]\n");
-  }
-  cout << endl;
+  // limit = num_node < 70 ? num_node : 70;
+  // cout << "Nodes (limited to " << limit << "): " << endl;
+  // for (int i = 0; i < limit; i++) {
+  //   printf("   [%2.2d]", i);
+  //   printf(" l:%+d", h_nodes[i].level);
+  //   printf(" p:%+2.2d", h_nodes[i].parent);
+  //   printf(" v:%+3.3d", h_nodes[i].value);
+  //   // moves
+  //   printf(" m[");
+  //   for (int j = 0; j < h_nodes[i].level; j++) {
+  //     printf(" %d", h_nodes[i].moves[j]);
+  //   }
+  //   printf(" ]\n");
+  // }
+  // cout << endl;
 
   // print task pools and head
[…]

### 2016-11-11 write-ups/PLDI/paper.tex
Abstract

--- 2016-11-11 write-ups/PLDI/paper.tex
@@
 \begin{abstract}
+There is growing interest in accelerating irregular data-parallel
+algorithms on GPUs.  These algorithms are typically \emph{blocking},
+so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
+do not mandate fair scheduling, and GPU schedulers are inherently
+unfair in practice.  Current irregular algorithms only work on today's
+GPUs by identifying hardware compute units and scheduling workgroups
+exclusively to these compute units.  As well as being non-portable,
+this is unacceptable because the GPU cannot be used in the meantime
+for high-priority tasks with soft real-time constraints,
+such as graphics rendering.
+
+We propose \emph{cooperative kernels}, an extension to the traditional
+GPU programming model geared towards writing blocking algorithms.
+Cooperative kernels enable fair scheduling of workgroups executing a
+blocking algorithm, as well as co-scheduling of other tasks on the
+GPU, via a small set of language extensions that create a contract
+between the programmer and the scheduler.  We describe our cooperative
+kernels programming model and its semantics in detail, and our
+implementation of a portable prototype of cooperative kernels on top
+of OpenCL 2.0.  We evaluate the technique by porting a set of
+irregular work stealing and graph algorithms to our programming model
+and evaluate the performance and responsiveness of the approach across
+four GPUs from two vendors.  Because our implementation exploits no
+vendor-specific hardware, driver or compiler support, our encouraging
+experimental results provide a lower-bound on the efficiency with which
+cooperative kernels can be implemented in practice.
 
@@
 \section{Introduction}\label{sec:intro}
-
-\ADComment{Open to attack: why not just buy another GPU?}
 
@@
 
-\ADComment{TODO: justify our choice: uniform variables and efficiency.}
+We opted for transmitting only selected varibles based on the
+observation that many of a thread's private variables are dead at the
+point of issuing $\offerfork$ or $\resizingglobalbarrier$, thus it
+would be wasteful to transmit them.  A live variable analysis could
+instead be employed to over-approximate the variables that might be
+accessed by newly arriving workgroups, so that these are automatically
+transmitted.
+
+In all cases, we found that a variable that needed to be transmitted
+had the property of being uniform across the workgroup.  That is,
+despite each thread having its own copy of the variable, each thread
+is in agreement on the variable's value.  As an example, the
+$\keyword{level}$, $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$
+variables used in \myfig\ref{fig:cgraphsearch} are all stored in thread-private
+memory, but all threads in a workgroup agree on the values of these
+variables at each $\resizingglobalbarrier$ call.  As a result,
+transmitting the thread 0's valuation of the annotated variables is
+equivalent to (and more efficient than) transmitting values on a
+thread-by-thread basis.  We have not yet encountered a real-world
+example where our current semantics would not suffice.
 
@@
 
-\ADComment{Perhaps this would fit better under ``programming guidelines'' -- not sure.}
-
-\ADComment{Make the point here that if an irregular algorithm is
-  written as a cooperative kernel then it will execute just as safely
-  as it does today on current hardware if: (1) the $\transmit$
-  annotation and $\offerkill$/$\offerfork$ calls are defined away to
-  nothing, (2) $\resizingglobalbarrier$ is redefined to
-  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
-  that assumes the occupancy-bound execution model is provided.
-If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}
+A goal of coopeative kernels is to allow traditional, non-blocking GPU
+tasks, such as graphics shaders (in programming models such as OpenGL,
+Vulkan and DirectX) and regular data-parallel computational kernels,
+to be co-scheduled with blocking irregular algorithms.  It is
+important to note that our new model has no effect on the manner in
+which these traditional GPU workloads are programmed.
+
+Furthermore, if our cooperative language extensions are
+ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
+$\transmit$ annotations are ignored, and $\globalbarrier$ and
+$\resizingglobalbarrier$ calls are redirected to an existing global
+barrier implementation (so that no resizing occurs in the latter
+case)---then cooperative kernel will behave just like its
+non-cooperative counterpart would behave.  As a result, a developer
+for whom today's occupancy-bound execution model suffices can upgrade
+their kernel to a cooperative form and retain backwards-compatibility
+with today's GPU platforms by pre-processing away our language
+extensions.
 

### 2016-11-11 write-ups/PLDI/paper.tex
Cited a bunch of graph papers.

@@
 of work to be processed is unknown ahead of time and may be determined
-by the computation itself.  There is growing interest in accelerating irregular
+by the computation itself.  The last decade has seen growing interest in accelerating irregular
 data-parallel algorithms on GPUs, particularly algorithms that
-traverse and manipulate linked data structures, with several recent notable
-successes~\cite{...}.
+traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16}.
+\ADComment{Tyler, can you add relevant workstealing papers if not here already?}
 
@@
 GPU schedulers are \emph{not} fair. the current GPU
-programming models---OpenCL~\cite{...}, CUDA~\cite{...}  and
-HSA~\cite{...}---specify almost no guarantees regarding scheduling of
+programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{...}  and
+HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
 workgroups, and implementations of these programming models do not
@@
 OpenCL specifically makes no guarantees about fair scheduling between
-workgroups executing the same kernel, stating~\cite{...}: \emph{``A
+workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
   conforming implementation may choose to serialize the workgroups so
@@
   the independent execution of workgroups since once in the work-pool,
-  they can execute in any order.''}  OpenCL is not alone in this
+  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} OpenCL is not alone in this
 regard: CUDA similarly provides no guarantees~\cite{...}.

### 2016-11-11 src/test_suite/final_octree/octree.h
Update defaults values

@@ -1,7 +1,7 @@
-DEFINE_int32(numParticles, 5000000, "number of particles to treat");
+DEFINE_int32(numParticles, 3000000, "number of particles to treat");
 DEFINE_int32(maxChildren, 10, "maximum number of children");
 DEFINE_int32(threads, 128, "number of threads");
-DEFINE_int32(pool_size, 10000, "size of task pools");
+DEFINE_int32(pool_size, 20000, "size of task pools");
 
 // ---
 DEFINE_string(restoration_ctx_path, "test_suite/final_octree/", "Path to restoration context");

### 2016-11-11 write-ups/PLDI/paper.tex

@@
 GPU schedulers are \emph{not} fair. the current GPU
-programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{...}  and
+programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
 HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
@@
 The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
-execution model~\cite{...}.  With this model, once a workgroup has commenced execution on a
+execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}.
+\ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
 compute unit (it has become \emph{occupant}), the workgroup has
@@
 Experiments suggest that the occupancy-bound execution model is widely
-implemented by today's GPU devices and drivers~\cite{...}.  It is
+implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
 evidently a simple model for device vendors to implement efficiently.
@@
 execution model.  They exploit the model by scheduling no more workgroups
-than there are available compute units~\cite{...}.  This works
+than there are available compute units~\cite{...}. \ADComment{Tyler, please pick best cite for previous claim.} This works
 because \emph{today}'s GPUs seem to provide the occupancy-bound
@@
 heuristically detect and kill long-running computations on the GPU
-being used for display rendering~\cite{...}.  Second, \emph{energy throttling} is
+being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.  Second, \emph{energy throttling} is
 an important concern for battery-powered devices, such as smartphones
-and tablets~\cite{...}.  In the future, it will be desirable for a mobile GPU driver to be able to
+and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
 power down one or more compute units if the battery level is low or if
@@
 practitioners who have been involved in the OpenCL and/or HSA
-standardisation efforts~\cite{CitePersonalCommunication}, is that (1)
-GPU vendors will not commit to the occupancy-bound execution model
-that they currently implement, due to the need for multi-tasking and
-energy throttling, yet (2) GPU vendors will not guarantee fair
-scheduling, due to the runtime overhead associated with supporting
-full preemption of workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors
-instead wish to retain the essence of the occupancy-bound execution
-model, which is simple to implement efficiently, and provide
-support for preemption only in key special cases.
+standardisation efforts
+(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
+and various engineers from GPU vendors), is that (1) GPU vendors will
+not commit to the occupancy-bound execution model that they currently
+implement, due to the need for multi-tasking and energy throttling,
+yet (2) GPU vendors will not guarantee fair scheduling, due to the
+runtime overhead associated with supporting full preemption of
+workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
+retain the essence of the occupancy-bound execution model, which is
+simple to implement efficiently, and provide support for preemption
+only in key special cases.
 
@@
 the case of SVM, is enabled via a set of atomic data types and
-operations, based on the C11 memory model~\cite{...}.  \ADComment{Could cut:} In particular,
-atomic SVM operations enable fine-grained communication between host
-and device.  Our implementation of cooperative kernels
-(\mysec\ref{sec:implementation}) depends on this recent OpenCL
-feature.
+operations, based on the C11 memory model~\cite{C11}.
 
@@
 
-\paragraph{Work stealing example}
+\paragraph{Work stealing example}  \ADComment{Tyler/Hugues: can you please say in the first paragraph that this has been implemented on GPUs and cite the right papers?}
 %
@@
 skeleton of a frontier-based graph traversal algorithm; such algorithms have
-been shown to be execute efficiently on GPUs~\cite{...}.
+been shown to be execute efficiently on GPUs~\cite{...}. \ADComment{Tyler, can you please work out which of the graph algorithm papers is the right one to cite here?}
 The kernel is
@@
 The global barrier used in this application is not provided as a GPU
-primitive, though many previous works have shown that such a global
-barrier can be implemented~\cite{...}. These implementations use well
+primitive, though previous works have shown that such a global
+barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}.
+\ADComment{Are there additional papers we should be citing on GPU barriers?
+E.g., is the sense reversal idea part of XF or is that from another paper?}  These implementations use well
 defined atomic memory operations to ensure memory consistency, and
@@
 issued before the barrier have completed.  The $\globalbarrier$
-primitive can be implemented following any inter-workgroup barrier
-design, \TSComment{This isn't exactly true. For cooperative kernels,
-  the barriers must be able to handle a growing or shrinking number of
-  workgroups.} e.g.~\cite{...}, and the atomic operations provided by
+primitive can be implemented by adapting an inter-workgroup barrier
+design, e.g.~\cite{XF}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
 the OpenCL 2.0 memory model enable a memory-safe
-implementation~\cite{...}.  However, implementing such a barrier is
+implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  However, implementing such a barrier is
 involved, hence why we include this function as a primitive.
@@
 
+\ADComment{Kepler preemption support?  Mention in related work?}
+
+\ADComment{Mention dynamic parallelism and explain why it's no good.}
+
 \cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
@@
 
-\section{Notes}
-
-\ADComment{A natural exit does not decrement $M$.  We haven't made a point of this, and I don't think we should right now, but we should remember to think this through carefully.}
-
-  \ADComment{Perhaps
-  make the point that inter-workgroup synchronization is possible in
-  very rare cases without blocking, e.g. threadfence reduction.}
-
-\ADComment{If this watchdog is disabled,
-execution of a long-running kernel causes the display to freeze for
-the duration of kernel execution.}
-
-\ADComment{Kepler preemption support?  Mention in related work?}
-
-\ADComment{Mention dynamic parallelism and explain why it's no good.}
-
-\section{Remainder of old intro}
-
-In an effort to achieve high performance for IP applications, new GPU
-programming paradigms have been developed that include custom GPU
-synchronisation constructs. Here we focus on two \ADComment{The way this is worded it is unclear whether you are describing two programming paradigms, or two synchronisation constructs}:
-
-\begin{enumerate}
-\item {\bf Inter-workgroup barrier}: Workgroups are able to
-  efficiently synchronise frequently using a custom inter-workgroup
-  barrier that is not provided as GPU primitive.
-
-\item {\bf Work stealing}: Computation is divided into tasks that are
-  stored in workgroup-local queues. If a workgroup runs out of local
-  tasks to compute, it is able to steal a task from a different
-  workgroup.
-\end{enumerate}
-
-Both paradigms fit into what are called \emph{persistent kernels}, the
-defining characteristic being that persistent kernels use a relatively
-small number of long-running threads to perform computation. In
-contrast, traditional GPU programs employ many short running threads
-to perform computation. In this work, the persistent kernels we consider
-use either work stealing or inter-workgroup barrier paradigms.
-
-Persistent kernels rely on strong forward progress properties (FPPs)
-between workgroups; that is, workgroups must make forward progress with respect to
-the other workgroups that are executing. In the case of the inter-workgroup barrier, all
-workgroups spin at a barrier instruction waiting for all workgroups to
-arrive. If one (or more) workgroups stop making forward progress, the
-application will deadlock as the live workgroups will wait
-indefinitely for the halted workgroups \ADComment{Can we re-word the preceding to make it clearer that the workgroups have been stalled by a scheduler: they have not actually terminated their execution by reaching the end of the kernel}.
-Less obviously, in the work
-stealing paradigm, the workgroups terminate only when \emph{all} tasks
-have been completed. If a workgroup takes a task, but stops making
-forward progress before the task is completed, the application will
-deadlock as the live workgroups will spin waiting for the halted
-workgroup to finish its task.
-
-The current GPU programming models generally do not provide strong
-enough inter-workgroup FPP to allow persistent kernels. \ADComment{This might be a good place to cite the specs.}  This is
-because GPUs do not provide workgroup preemption. That is, GPU
-programming models allow GPU programs to be launched with many more
-workgroups than the available hardware resources are able run
-concurrently. The GPU runtime will schedule as many workgroups as the
-available resources can provide, and the remaining workgroups will
-wait until hardware resources become available, e.g.  when one of the
-scheduled workgroups finishes execution. There is no mechanism to swap
-workgroups from 'executing' to 'waiting' and vice versa. Workgroups
-that have been scheduled and have not finished execution are called
-\emph{occupant}.
-\ADComment{Note: it's ironic that lack of preemption is what leads to the problem of workgroups being scheduled in waves, and yet also what means we might not be able to guarantee the properties of the occupancy-bound execution model in the future.}
-
-Previous work experimenting with persistent kernels has required
-\emph{a priori} knowledge of the number of workgroups that the GPU
-hardware is able to execute the program with concurrently. This
-number, called the \emph{maximal launch} or \emph{occupancy bound}, is
-a function of both a GPU chip and a compiled program. Because occupant
-workgroups are never preempted, if the GPU program is launched with a
-number of workgroups which doesn't exceed the occupancy bound, then
-all workgroups will be occupant and the strong inter-workgroup FPP
-required by persistent kernels are provided. Using the occupancy
-bound, previous work has successfully implemented persistent kernels
-and shown them to be performant on current GPUs. Recent work has even
-shown that the occupancy bound can be accurately discovered
-programitically, removing the need for any \emph{a priori} knowledge
-about the GPU.
-
-Despite the promising performance and ease of programming persistent
-kernels with occupant workgroups, GPU language specification
-committees are reluctant to commit to the inter-workgroup FPP
-currently provided by occupant workgroups. That is, even if current
-GPUs provide strong FPP between workgroups, future GPUs may not. Thus
-the performance gains enabled by persistent kernels may not be
-safely achievable for future GPUs.
-
-We believe that two fundamental issues stopping committees from
-committing to strong inter-workgroup FPP are:
-
-\begin{itemize}
-\item {\bf Responsive GPU multi-tasking} A high priority GPU task
-  (e.g. graphics) may need to quickly gain access to the GPU. To do
-  this, some of the resources will have to be taken from the
-  persistent kernel in order to run the high priority task. This will
-  reduce the number of occupant workgroups executing the persistent
-  kernel.
-
-\item {\bf Energy throttling} Some GPUs, especially in mobile devices,
-  may wish to monitor the energy consumed (i.e. heat output) of the
-  GPU chip.  A GPU that is consuming too much energy may be throttled
-  by limiting the amount of resources (and thus the number of occupant
-  workgroups) to conserve battery life.
-\end{itemize}
-
-Both issues are a case of the GPU runtime needing to adjust the
-resources used by the persistent kernel during execution. This causes
-occupant workgroups to potentially be preempted, which may lead to
-deadlock in persistent kernels.
-\ADComment{It might be good to illustrate the previous points with an example; e.g.\ we could envisage a kernel that is being executed concurrently by 16 workgroups.
-Some (4 say) have reached an inter workgroup barrier and are awaiting the others.
-But then a graphics tasks causes the 12 workgroups that have not yet reached the barrier to be preempted, and in anticipation of further graphics processing requirements the scheduler decides not to reduce the workgroup allocation for the kernel to 4 indefinitely.
-The 4 workgroups waiting at the barrier cannot now make progress.
-Energy throttling example is analogous, except that the count is reduced from 16 to 12 due to low battery.}
-A less severe, but still undesirable
-issue is if the runtime determines more GPU resources are available
-during a persistent kernel execution. In this case, the persistent
-kernel will be under-utilising GPU resources.
-\ADComment{I'm not sure I get this one.
-Do you mean here that at the point of kernel launch there may be e.g.
-only 4 workgroup slots available, say due to low battery, but then later on (e.g., due to a phone being charged) there may be scope for more workgroups to run in parallel?}
-
-\ADComment{We should mention that GPU watchdogs are the rather drastic solution to this problem, to keep the overall system responsive, but that they do not help from the point of view of IP applications themselves.}
-
-In this work, we propose a solution that unites persistent kernels
-with the need for GPU runtimes to dynamically (and quickly) adjust the
-resources available to the persistent kernel. We propose a new GPU
-programming paradigm called \emph{dynamic persistent kernels}, or
-DPK. DPKs provide strong inter-workgroup FPP while still allowing the
-GPU runtime to adjust the number of workgroups executing the DPK in a
-timely manner, enabling responsive GPU multi-tasking or energy
-throttling.
-
-The key insight behind DPK, is that many persistent kernels contain
-points in their execution where workgroups can easily be either
-terminated or started. We propose several new instructions to mark
-these entry and exit points. These instructions serve as hooks for the
-GPU runtime to adjust the amount of resources available to the DPK,
-and thus the number of occupant workgroups the DPK can rely on to have
-strong FPPs.
-\ADComment{We should make it clear early on that some care on the part of the programmer is required to make this work, and that if the programmer does not insert sufficient yield points, the GPU watchdog may still have to kill a kernel.}
-
-For example, many applications using the inter-workgroup barrier
-paradigm do not carry workgroup-local information across calls to the
-inter-workgroup barrier. Thus the number of workgroups can be adjusted
-at every call to the inter-workgroup barrier. Many applications using
-the work stealing paradigm do not carry state between task
-computation; thus workgroups can be created or terminated immediately
-after ever task computation.
-
-While we envision support for DPKs to ultimately be implemented in
-proprietary GPU drivers, we are able to provide a prototype
-implementation in software using the strong inter-workgroup FPP
-available on current GPUs. This allows us to test the feasibility of
-our approach and as well get upper bounds for the performance cost and
-the responsiveness of multi-tasking.
-
 \end{document}

### 2016-11-11 src/non_persistent_kernels/matmult/matmult.h
Bound values of matrices

@@ -42,7 +42,7 @@ int rand_fill(cl_int *M, int num_row, int num_col, int seed)
     for (int col = 0; col < num_col; col++) {
       seed = seed * seed;
       seed = (seed / 1000) % 1000000;
-      M[(row * num_col) + col] = (cl_int)seed;
+      M[(row * num_col) + col] = (cl_int)(seed % 10);
     }
   }
 
@@ -51,15 +51,15 @@ int rand_fill(cl_int *M, int num_row, int num_col, int seed)
 
 /*---------------------------------------------------------------------------*/
 
-int hash_mat(cl_int *M, int num_row, int num_col)
+cl_int hash_mat(cl_int *M, cl_int num_row, cl_int num_col)
 {
   // hash the diagonal using djb2, see
   // http://www.cse.yorku.ca/~oz/hash.html
-  int hash = 5381;
-  int row = 0;
-  int col = 0;
+  cl_int hash = 5381;
+  cl_int row = 0;
+  cl_int col = 0;
   while (row < num_row && col < num_col) {
-    hash = (hash * 33) + M[(row * num_col) + col];
+    hash = (hash * (cl_int)33) + M[(row * num_col) + col];
     row++;
     col++;
   }

### 2016-11-11 src/test_suite/final_octree/octree.h
Update defaults

@@ -1,5 +1,5 @@
 DEFINE_int32(numParticles, 3000000, "number of particles to treat");
-DEFINE_int32(maxChildren, 10, "maximum number of children");
+DEFINE_int32(maxChildren, 20, "maximum number of children");
 DEFINE_int32(threads, 128, "number of threads");
 DEFINE_int32(pool_size, 20000, "size of task pools");
 

### 2016-11-11 write-ups/PLDI/paper.tex

@@
 
-\paragraph{Work stealing example}  \ADComment{Tyler/Hugues: can you please say in the first paragraph that this has been implemented on GPUs and cite the right papers?}
+\paragraph{Work stealing example}
 %
@@
 terminating---tries to \emph{steal} a task from an other
-pool. Workgroups keep on processing tasks until all the work is done.
+pool. Workgroups keep on processing tasks until all the work is
+done. Previous work addresses the feasibility and efficiency of work
+stealing on GPUs~ref\cite{dlb-web,TPO10}.
 
@@
 primitive can be implemented by adapting an inter-workgroup barrier
-design, e.g.~\cite{XF}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
+design, e.g.~\cite{XF10}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
 the OpenCL 2.0 memory model enable a memory-safe

### 2016-11-11 src/host_skeleton/skel.cpp
worked on good parameters for octree

@@ -543,6 +543,8 @@ void run_merged(CL_Execution *exec) {
 		//cout << "Check value: " << *(s_ctx.check_value) << endl;
 		cout << "Persistent Error: " << error_persistent << endl;
 		cout << "Non-persistent Error: " << error_non_persistent << endl;
+		cout << "check value: " << *s_ctx.check_value << endl;
+
 		cout << endl;
 		CL_Communicator::my_sleep(1000);
 	}

### 2016-11-11 src/test_suite/final_octree/octree.h

@@ -1,7 +1,7 @@
 DEFINE_int32(numParticles, 3000000, "number of particles to treat");
-DEFINE_int32(maxChildren, 20, "maximum number of children");
+DEFINE_int32(maxChildren, 4, "maximum number of children");
 DEFINE_int32(threads, 128, "number of threads");
-DEFINE_int32(pool_size, 20000, "size of task pools");
+DEFINE_int32(pool_size, 50000, "size of task pools");
 
 // ---
 DEFINE_string(restoration_ctx_path, "test_suite/final_octree/", "Path to restoration context");

### 2016-11-11 src/pannotia/graph_apps/color/color.h
fixed a bug in color

@@ -106,22 +106,40 @@ void reset_persistent_task(CL_Execution *exec) {
 int set_persistent_app_args_for_occupancy(int arg_index, cl::Kernel k) {
 	// Set args for persistent kernel
 	int err = k.setArg(arg_index, row_d);
+	check_ocl(err);
+
 	arg_index++;
 	err |= k.setArg(arg_index, col_d);
+	check_ocl(err);
+
 	arg_index++;
-	err = k.setArg(arg_index, node_value_d);
+	err |= k.setArg(arg_index, node_value_d);
+	check_ocl(err);
+
 	arg_index++;
 	err |= k.setArg(arg_index, color_d);
+	check_ocl(err);
+
 	arg_index++;
-	err = k.setArg(arg_index, stop_d1);
+	err |= k.setArg(arg_index, stop_d1);
+	check_ocl(err);
+
 	arg_index++;
 	err |= k.setArg(arg_index, stop_d2);
+	check_ocl(err);
+
 	arg_index++;
-	err = k.setArg(arg_index, max_d);
+	err |= k.setArg(arg_index, max_d);
+	check_ocl(err);
+
 	arg_index++;
 	err |= k.setArg(arg_index, num_nodes);
+
+	check_ocl(err);
+
 	arg_index++;
 	err |= k.setArg(arg_index, num_edges);
+	check_ocl(err);
 	arg_index++;
 	check_ocl(err);
 

### 2016-11-11 src/run_test.py

@@ -30,11 +30,11 @@ EXIT_THREAD      = 0
 
 PROGRAMS = {
 #    "lonestar_sssp"
-    "lonestar_bfs",
+#    "lonestar_bfs",
     "pannotia_color",
-    "pannotia_mis",
-    "pannotia_bc",
-    "pannotia_sssp"
+#    "pannotia_mis",
+#    "pannotia_bc",
+#    "pannotia_sssp"
 }
 
 PROGRAM_DATA = {

### 2016-11-11 src/host_skeleton/skel.cpp
skeleton print a stdout summary because python pipes don't get along with these applications

@@ -26,6 +26,7 @@ DEFINE_string(output_timestamp_executing_groups, "timestamp_executing_groups", "
 DEFINE_string(output_timestamp_non_persistent, "timestamp_non_persistent", "Path to output timestamps for non persistent tasks");
 DEFINE_string(output_non_persistent_duration, "non_persistent_duration", "Path to output duration results for non persistent tasks");
 DEFINE_string(output_summary, "summary", "output file for summary of results");
+DEFINE_string(output_summary2, "summary2", "output file for summary of results (2nd edition, used in scripts)");
 DEFINE_int32(non_persistent_wgs, 2, "ratio of workgroups to send to non-persistent task. Special values are (-1) to send all but one workgroup and (-2) to send one workgroup");
 DEFINE_int32(skip_tasks, 0, "flag to say if non persistent tasks should be skipped: 0 - don't skip, 1 - skip");
 DEFINE_int32(non_persistent_frequency, 100, "frequency in milliseconds to launch non_persistent tasks");
@@ -74,6 +75,14 @@ using namespace std;
 #else
 #error "No non-persistent task macro defined ? like NON_PERSISTENT_XYZ (check your CMakeLists.txt)"
 #endif
+
+void print_string_to_file(const char * fname, std::string to_print) {
+	FILE * fp = fopen(fname, "w");
+	if (!fp) { printf("ERROR: unable to open file %s\n", fname); }
+	fprintf(fp, "%s\n", to_print.c_str());
+
+	fclose(fp);
+}
 
 //From IWOCL tutorial (needs attribution)
 unsigned getDeviceList(std::vector<std::vector<cl::Device> >& devices)
@@ -239,8 +248,14 @@ int amd_check(int v) {
 
 // Just for running the persistent task
 void run_persistent(CL_Execution *exec) {
-	printf("Running persistent app %s\n", persistent_app_name());
-	printf("Running %d iterations\n", FLAGS_run_persistent);
+	std::stringstream summ_str1;
+	//printf("Running persistent app %s\n", persistent_app_name());
+	//printf("Running %d iterations\n", FLAGS_run_persistent);
+
+	summ_str1 << "Running persistent app " << persistent_app_name() << "\n";
[…]

### 2016-11-11 src/run_test.py

@@ -190,6 +190,8 @@ def exec_cmd(cmd, prefix="", record_file=""):
     cmd = cmd + ["--output_non_persistent_duration", prefix + "_non_persistent_duration"]
     cmd = cmd + ["--output_timestamp_executing_groups", prefix + "_timestamp_executing_groups"]
     cmd = cmd + ["--output_timestamp_non_persistent", prefix + "_timestamp_non_persistent"]
+    cmd = cmd + ["--output_summary2", record_file]
+
 
     if prefix in CHECK_POINT_DATA:
         print(prefix + " found in checkpoint")
@@ -198,25 +200,25 @@ def exec_cmd(cmd, prefix="", record_file=""):
     #time.sleep(30)
     EXIT_THREAD = 0
     local_time_begin = time.time()
-    thread = Thread(target = const_print_time)
-    thread.start()
+    #thread = Thread(target = const_print_time)
+    #thread.start()
 
     PRINT("====================== RUN START =========================")
     PRINT(prefix)
     for s in cmd:
         PRINT(s)
     PRINT("----------------------------------------------------------")
-    p_obj = subprocess.Popen(cmd, stdout=subprocess.PIPE, stderr=subprocess.PIPE)
+    p_obj = subprocess.Popen(cmd)
     ret_code = p_obj.wait()
-    sout, serr = p_obj.communicate()
-    PRINT("* stdout:")
-    PRINT(sout.decode())
-    if record_file != "":
-        f = open(record_file, "w")
-        f.write(sout.decode())
-        f.close()
-    PRINT("* stderr:")
-    PRINT(serr.decode())
+    #sout, serr = p_obj.communicate()
+    #PRINT("* stdout:")
+    #PRINT(sout.decode())
+    #if record_file != "":
+    #    f = open(record_file, "w")
+    #    f.write(sout.decode())
+    #    f.close()
+    #PRINT("* stderr:")
+    #PRINT(serr.decode())
     local_time_end = time.time()
     PRINT("----------------------------------------------------------")
     str_ret_code = "Return code: " + str(ret_code)
@@ -225,7 +227,7 @@ def exec_cmd(cmd, prefix="", record_file=""):
     PRINT("time since start: " + str(local_time_end - TIME_BEGIN) + " seconds")
     PRINT("====================== RUN END ===========================")
     EXIT_THREAD = 1
-    thread.join();
+    #thread.join();
     if (ret_code != 0):
[…]

### 2016-11-11 src/scheduler_rt/rt_host/cl_communicator.h

@@ -196,7 +196,7 @@ class CL_Communicator {
 				if (groups == 0) {
 					break;
 				}
-				spinwait_microsec(1);
+				my_sleep(1);
 			}
 
 			persistent_end = gettime_chrono();

### 2016-11-11 src/run_test.py
crippled the checkpointing (it was buggy anways) and set iterations back to 10

@@ -17,7 +17,7 @@ def fake_placeholder(x):
 EXE_PATH      = ""
 DATA_PATH     = ""
 STAT_PATH     = ""
-ITERATIONS    = "2"
+ITERATIONS    = "10"
 PRINT         = fake_placeholder
 DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""
@@ -158,18 +158,18 @@ def write_to_checkpoint(data):
     f.write(str(time.time() - TIME_BEGIN) + "\n")
     f.close()
 
-def get_check_point_data():
-    global CHECK_POINT_DATA
-    global TIME_BEGIN
-    #pdb.set_trace()
-    if (os.path.isfile(CHECK_POINT_FILE)):
-        print("HI THERE")
-        f = open(CHECK_POINT_FILE, "r")
-        CHECK_POINT_DATA = f.readlines()
-        CHECK_POINT_DATA = [x.replace("\n", "") for x in CHECK_POINT_DATA]
-        checked_time = float(CHECK_POINT_DATA[len(CHECK_POINT_DATA) - 1])
-        TIME_BEGIN = TIME_BEGIN - checked_time
-        f.close()
+#def get_check_point_data():
+#    global CHECK_POINT_DATA
+#    global TIME_BEGIN
+#    #pdb.set_trace()
+#    if (os.path.isfile(CHECK_POINT_FILE)):
+#        print("HI THERE")
+#        f = open(CHECK_POINT_FILE, "r")
+#        CHECK_POINT_DATA = f.readlines()
+#        CHECK_POINT_DATA = [x.replace("\n", "") for x in CHECK_POINT_DATA]
+#        checked_time = float(CHECK_POINT_DATA[len(CHECK_POINT_DATA) - 1])
+#        TIME_BEGIN = TIME_BEGIN - checked_time
+#        f.close()
 
 def const_print_time():
     global EXIT_THREAD

### 2016-11-11 write-ups/PLDI/paper.tex
Related work sketch.

--- 2016-11-11 write-ups/PLDI/paper.tex
@@
 
+Lots of work on irregular algorithms (assumes fairness), and persistent threads (assumes OB model, which assumes a degree of fairness).  Our contribution can enable these to be safe.
+
+Work on multi-tasking to improve throughput, etc., but this does not relate to fairness and wouldn't help blocking algorithms.
+
+Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
+
+Dynamic parallelism could be extended to give barriers, but doesn't work with work stealing (less general).  It is in recent OpenCL.
+
 \ADComment{Kepler preemption support?  Mention in related work?}
-
-\ADComment{Mention dynamic parallelism and explain why it's no good.}
 

### 2016-11-12 write-ups/PLDI/paper.tex
Comments from Paul.

@@
 \newcommand{\getglobalid}{\mathsf{get\_global\_id}}
-\newcommand{\getlocalsize}{\mathsf{get\_global\_id}}
+\newcommand{\getlocalsize}{\mathsf{get\_local\_size}}
 \newcommand{\getglobalsize}{\mathsf{get\_global\_size}}
@@
 another workgroup $B$ to spin-wait forever for the mutex to be
-released.  Similarly, an unfair scheduler can a workgroup to spin-wait
-at a global barrier so that other workgroups do not make progress
+released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
+indefinitely at a global barrier so that other workgroups do not make progress
 towards the barrier.
@@
 
-\item An experience report porting \NumAlgorithms{} irregular GPU
+\item An experience report of porting \NumAlgorithms{} irregular GPU
   algorithms to use cooperative kernels, and experimental results
@@
 memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
-kernel, but can only be safely accessed by the host application before
-or after kernel execution.  Each workgroup has its own portion of \emph{local},
-which can be used for fast intra-workgroup communication, and every thread has
+kernel.  Each workgroup has its own portion of \emph{local} memory,
+which can be used for fast intra-workgroup communication.  Every thread has
 an allocation of very fast \emph{private} memory, where
@@
 
-Fine-grained communication between threads, and also with the host in
-the case of SVM, is enabled via a set of atomic data types and
-operations, based on the C11 memory model~\cite{C11}.
+Communication between threads in the same workgroup can be achieved
+using the workgroup barrier construct described above.  Finer-grained
+communication within a workgroup, as well as inter-workgroup
+communication and communication with the host while the kernel is
+running, is enabled by a set of atomic data types and operations.  In
+particular, fine-grained host/device communication is via atomic
+operations on SVM.
 
@@
   comes after B in work-group flattened ID order}. However, this
-one-way communication is not sufficient so support either the
-work stealing or graph application examples of
-Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
-above, both require \emph{symmetric} communication between
+one-way communication is not sufficient to support blocking algorithms that use
+mutexes and inter-workgroup barriers, both of which require \emph{symmetric} communication between
 threads.
@@
 
-(*@\label{line:graph:swap}@*)    swap(in_nodes, out_nodes);
+(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
 (*@\label{line:graph:gb1}@*)    global_barrier();
@@
 on a fair scheduler, which, as discussed in \mysec\ref{sec:intro},
-is not provided by current GPU programming models.
+is not guaranteed by current GPU programming models.
 
@@
 the nodes to be processed are evenly distributed between the executing
-threads. This is done through \emph{stride} based processing. A thread
-retrieves its global thread id and the total number of threads
-(obtained through the $\getglobalid()$ and $\getglobalsize()$ primitives). A thread then processes
-each node associated with its id, i.e.\ nodes at indices that are a multiple of the total number of threads, plus the thread's id. A thread processes a node via the $\keyword{process\_node}$
+threads. This is done through \emph{stride} based processing. 
+%
+In this case, the stride is the total number of threads, obtained via 
+$\getglobalsize()$.  Each thread processes nodes with the following sequence of indices:
+\code{$\keyword{tid}$+$\keyword{stride}$*0},
+\code{$\keyword{tid}$+$\keyword{stride}$*1},
+\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
+A thread processes a node via the $\keyword{process\_node}$
 function, which may (a) push nodes to process in the next frontier to
 $\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
-the computation. \ADComment{Tyler: should swap take the addresses of the pointers?} After processing the frontier, the threads swap their
+the computation. After processing the frontier, the threads swap their
 local node array pointers (line~\ref{line:graph:swap}), i.e.\ the $\keyword{out\_nodes}$ of the current frontier becomes the $\keyword{in\_nodes}$ of the next
@@
 threads have arrived.  This spinning behaviour requires fair
-scheduling between workgroups for the barrier to operate correctly,
-and as a result the barrier is \emph{not} guaranteed to work correctly
-under current GPU programming models.  Such barriers appear to run
+scheduling between workgroups for the barrier to operate correctly.
+Again, fairness is not guaranteed by current GPU programming models.
+
+The mutexes and barriers required by these two examples appear to run
 reliably on current GPUs for kernels that are executed with no more
 workgroups than there are compute units; this is due to the fairness
-of the occupancy-bound execution model that current GPUs appear to
-provide.  But, as discussed in \mysec\ref{sec:intro}, this
-occupancy-bound model is not mandated by language standards or vendor
-implementations, and is unlikely to be guaranteed in the future.
+of the occupancy-bound execution model that current GPUs have been
+shown, experimentally, to provide.  But, as discussed in
+\mysec\ref{sec:intro}, this occupancy-bound model is not mandated by
+language standards or vendor implementations, and is unlikely to be
+guaranteed in the future.
 

### 2016-11-12 write-ups/PLDI/paper.tex
More feedback from Paul.

@@
 number of workgroups, each consisting of a specified number of
-threads.  The parameters to a cooperative kernel are \emph{immutable}, as if
+threads.  Unlike in a regular kernel, the parameters to a cooperative kernel are \emph{immutable}, as if
 they were qualified with the \texttt{const} keyword (though pointer
@@
 Initially, the number of active workgroups $M$ is initialised to some
-integer between $1$ and $N$.  The initial active workgroups have
+integer between $1$ and $N$---to ensure that at least one workgroup is scheduled, the scheduler will postpone execution of the kernel until a some compute unit becomes available.
+The initial active workgroups have
 consecutive workgroup ids in the range $[0, M-1]$.

### 2016-11-12 src/host_skeleton/skel.cpp
more tweaks to script

@@ -320,7 +320,7 @@ void run_persistent(CL_Execution *exec) {
 		reset_barrier(exec, d_bar);
 		reset_persistent_task(exec);
 		restart_scheduler(&s_ctx);
-		CL_Communicator::my_sleep(1000);
+		CL_Communicator::my_sleep(5000);
 		err = exec->exec_queue.flush();
 		check_ocl(err);
 		err = exec->exec_queue.finish();
@@ -599,7 +599,7 @@ void run_merged(CL_Execution *exec) {
 		
 
 		cout << endl;
-		CL_Communicator::my_sleep(1000);
+		CL_Communicator::my_sleep(5000);
 	}
 
 	free_scheduler_ctx(exec, &s_ctx);

### 2016-11-12 src/run_test.py
Add matmult config for carrot (radeon r7)

@@ -142,6 +142,13 @@ MATMULT_CONFIG_HD520 = [
     { "freq" : "40", "matdim" : "260", "name" : "high" },
 ]
 
+MATMULT_CONFIG_RADEON_R7 = [
+    # This is for carrot
+    { "freq" : "70", "matdim" : "277", "name" : "light" },
+    { "freq" : "40", "matdim" : "277", "name" : "medium" },
+    { "freq" : "40", "matdim" : "431", "name" : "high" },
+]
+
 def my_print(file_handle, data):
     print(data)
     file_handle.write(data + os.linesep)

### 2016-11-12 write-ups/PLDI/paper.tex
added a table about the applications

@@
 \newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
+
+\newcommand\crule[3][black]{\textcolor{#1}{\rule{#2}{#3}}}
 
@@
 \newcommand{\effectivethreshold}{5\%\xspace}
-\definecolor{Gray}{gray}{0.85}
+\definecolor{Gray3}{gray}{0.9}
+\definecolor{Gray2}{gray}{0.75}
+\definecolor{Gray1}{gray}{0.6}
 
@@
 
+\begin{table}
+\small
+\centering
+\begin{tabular}{ l c c c }
+app & resizing barriers & ckill - cfork & tranmit vars\\
+\hline
+\rowcolor{Gray1}
+color & 2 / 2 & 0 - 0 & 4\\
+\rowcolor{Gray1}
+mis & 3 / 3 & 0 - 0 & 0\\
+\rowcolor{Gray1}
+bc & 3 / 6 & 0 - 0 & 3\\
+\rowcolor{Gray1}
+p-sssp & 3 / 3 & 0 - 0 & 0 \\
+\rowcolor{Gray2}
+bfs & 2 / 2 & 0 - 0  & 4 \\
+\rowcolor{Gray2}
+l-sssp & 2 / 2 & 0 - 0  & 4 \\
+\rowcolor{Gray3}
+octree & 0 / 0 & 1 - 1 & 0\\
+\rowcolor{Gray3}
+game & 0 / 0 & 1 - 1 & 0\\
+\end{tabular} \\
+\vspace{.2cm}
+\crule[Gray1]{.2cm}{.2cm} - Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} - Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} - work stealing
+\caption{Blocking GPU applications investigated}
+\label{tab:applications}
+\end{table}
+
 \ADComment{Describe the algorithms that we ported.}
@@
 \ADComment{Give stats on how many variables had to be marked with
-  $\transmit$ (i.e., occurred in the restoration context), how many barriers we changed to resizing, how many offer kill and offer fork calls we included.}
+  $\transmit$ (i.e., occurred in the restoration context), how many
+  barriers we changed to resizing, how many offer kill and offer fork
+  calls we included.}  \TSComment{check out
+  Table~\ref{tab:applications}. The barrier column shows how many of
+  the original barriers were ported to resizing barriers. In all cases
+  except bc we converted all of them. The only reason why bc is
+  different is because barriers occur deeper in the call stack and our
+  prototype cannot handle that. The algorithm itself is compatable
+  (i.e. changing these barriers is as easy as any other, there is
+  nothing special about them.)}
 

### 2016-11-12 src/run_test.py
added Iris matrix configurations to the script

@@ -30,11 +30,11 @@ EXIT_THREAD      = 0
 
 PROGRAMS = {
 #    "lonestar_sssp"
-    "lonestar_bfs",
+#    "lonestar_bfs",
     "pannotia_color",
-    "pannotia_mis",
-    "pannotia_bc",
-    "pannotia_sssp"
+#    "pannotia_mis",
+#    "pannotia_bc",
+#    "pannotia_sssp"
 }
 
 PROGRAM_DATA = {
@@ -140,6 +140,12 @@ MATMULT_CONFIG_HD520 = [
     { "freq" : "70", "matdim" : "160", "name" : "light" },
     { "freq" : "40", "matdim" : "160", "name" : "medium" },
     { "freq" : "40", "matdim" : "260", "name" : "high" },
+]
+
+MATMULT_CONFIG_IRIS = [
+    { "freq" : "70", "matdim" : "290", "name" : "light" },
+    { "freq" : "40", "matdim" : "290", "name" : "medium" },
+    { "freq" : "40", "matdim" : "445", "name" : "high" },
 ]
 
 def my_print(file_handle, data):
@@ -383,10 +389,12 @@ def main():
     if NAME_OF_CHIP == "HD5500":
         MATMULT_CONFIG = MATMULT_CONFIG_HD5500
     elif NAME_OF_CHIP == "HD520":
-	MATMULT_CONFIG = MATMULT_CONFIG_HD520
+        MATMULT_CONFIG = MATMULT_CONFIG_HD520
+    elif NAME_OF_CHIP == "IRIS":
+        MATMULT_CONFIG = MATMULT_CONFIG_IRIS
     else:
         print("Cannot find a matmult for your chip! Exiting")
-	exit(0)
+        exit(0)
     PLATFORM_ID = sys.argv[6]
     IS_AMD = sys.argv[7]
     log_file = sys.argv[4] + ".log"

### 2016-11-12 write-ups/PLDI/paper.tex
Add paragraph on the GPU we tested

@@
 
+\HEComment{This is more or less already done in section
+\ref{sec:programmingguidelines} , maybe no need to rewrite here?}
+
 \ADComment{Give stats on how many variables had to be marked with
@@
 
+\HEComment{the reason we need SVM fine grain and atomics may be better
+discussed directly in the section on implementation?}
+
+Our implementation approach requires two optional features of OpenCL
+$2.0$, namely SVM fine-grained buffers and SVM atomics, which are
+currently not widely supported. \nvidia chips are ruled out since
+\nvidia does not support OpenCL $2.0$. AMD support SVM atomics only
+their Kaveri family of chips and only on Linux, requiring to compile the
+latest driver.  \HEComment{do we need to mention the struggle for AMD ?}
+Among the GPUs available to us, only 4 support our requirements: Intel
+HD 520, HD 5500 and IRIS 6100, and AMD Radeon R7.
+
 \ADComment{Describe the platforms.  Explain why we could not try other
 platforms.}
-
-\ADComment{Tyler and Hugues, please write.}
 

### 2016-11-12 src/run_test.py
fixed a bug in the script

@@ -28,14 +28,14 @@ CHECK_POINT_DATA = []
 TIME_BEGIN       = 0.0
 EXIT_THREAD      = 0
 
-PROGRAMS = {
+PROGRAMS = [
 #    "lonestar_sssp"
 #    "lonestar_bfs",
     "pannotia_color",
-#    "pannotia_mis",
-#    "pannotia_bc",
-#    "pannotia_sssp"
-}
+    "pannotia_mis",
+    "pannotia_bc",
+    "pannotia_sssp"
+]
 
 PROGRAM_DATA = {
 
@@ -52,12 +52,12 @@ PROGRAM_DATA = {
     ],
 
     "pannotia_mis" : [ { "input" : os.path.join("pannotia","inputs", "color", "ecology1.graph"),
-                         "solution" : os.path.join("pannotia","solutions", "color_ecology.txt"),
+                         "solution" : os.path.join("pannotia","solutions", "mis_ecology.txt"),
                          "stat" : "mis_ecology",
                          "suite" : "pannotia",
                          "query_barrier" : [0,1]},
                        { "input" : os.path.join("pannotia","inputs", "color", "G3_circuit.graph"),
-                         "solution" : os.path.join("pannotia","solutions", "color_G3_circuit.txt"),
+                         "solution" : os.path.join("pannotia","solutions", "mis_G3_circuit.txt"),
                          "stat" : "mis_G3_circuit",
                          "suite" : "pannotia",
                          "query_barrier" : [0,1]}

### 2016-11-13 write-ups/PLDI/paper.tex
Finished addressing Paul's comments.

--- 2016-11-12 write-ups/PLDI/paper.tex
@@
 
-Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$
+Suppose a workgroup with id $m$ executes $\offerkill$.  
+If the workgroup has the highest largest id of all active workgroups then it can be killed by the scheduler, except that workgroup 0 can never be killed (to ensure that the kernel computation does not terminate early).
+More formally, if $m < M-1$
 or $M=1$, i.e.\ the workgroup's id is not the largest among the active
@@
 executing as a no-op, or (b) execution of the workgroup ceasing and
-the number of active workgroups $M$ being decremented by one.  Case
+the number of active workgroups $M$ being atomically decremented by one.  Case
 (b) corresponds to the scheduler accepting the workgroup's offer to
@@
 
-The semantics of $\offerkill$ is atomic.  That is, if multiple
-workgroups reach $\offerkill$ simultaneously, the semantics is as if
-each workgroup executes $\offerkill$ in some arbitrary sequential
-order.
-
-In this formulation of cooperative kernels, only the active workgroup
-with the largest id can be killed, and workgroup 0 can never be
-killed.  We discuss the rationale for these choices, as well as
-possible alternative choices, in
-\mysec\ref{sec:semanticalternatives}.
-
 \paragraph{Semantics for $\offerfork$}
 
-The $\offerfork$ primitive allows the cooperative kernel to indicate
+The $\offerfork$ primitive (also a workgroup-level function) allows the cooperative kernel to indicate
 to the scheduler that it could benefit from, and is prepared to
@@
 
-Like $\offerkill$, $\offerfork$ is a workgroup-level function.
-
-Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  An
-integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
+Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  Then the following occurs: an
+integer $k$ in the range $[0, N-M]$ is chosen by the scheduler;
 $k$ new workgroups are spawned with consecutive ids in the range $[M,
-  M+k-1]$, and the active workgroup count $M$ is incremented by $k$.
+  M+k-1]$; the active workgroup count $M$ is atomically incremented by $k$.
 
@@
 Semantically, when all threads have reached $\resizingglobalbarrier$,
-the number of active workgroups, $M$, is set to a new value, $M'$ say, with $M' > 0$.
+the number of active workgroups, $M$, is atomically set to a new value, $M'$ say, with $M' > 0$.
 If $M' = M$ then the active workgroups remain unchanged and proceed to
@@
 
-The enclosing $\globalbarrier$ calls make the operation appear to be
-atomic.  The middle $\globalbarrier$ is required to ensure that if
-$M'\geq M$, the workgroups with ids smaller than $M$ are left intact.
+The enclosing $\globalbarrier$ calls ensure that the change in number
+of active workgroups from $M$ to $M'$ occurs entirely within the
+resizing barrier, so that from a programmer's perspective, $M$ changes atomically.
+The middle $\globalbarrier$ is required to ensure that forking occurs before killing, so that while the barrier may kill a workgroup with id $i$, it should not also fork a new workgroup with id $i$.
+That is, workgroups $[0, \mathit{min}(M, M') - 1]$ should be left intact.
 
@@
 by the requirements of the irregular GPU algorithms proposed in the
-literature, and that at least for these algorithms it is usually easy
-to identify where to place the constructs.
+literature.  For these algorithms, it was usually easy to identify
+where to place the constructs.
 
@@
 the value returned by $\getnumgroups()$ is not fixed during the
-lifetime of a cooperative kernel.  Although
-$\getnumgroups()$ has atomic semantics, and so can be called at any
-time to yield a well-defined value, the value returned by two
-successive calls may differ if workgroups have executed $\offerkill$,
-$\offerfork$ or $\resizingglobalbarrier$ in-between.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
-a function of $\getnumgroups()$, the value returned by $\getglobalsize()$ is also subject to change.
+lifetime of a cooperative kernel: it may change due to workgroups executing
+$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
+a function of $\getnumgroups()$, the value it returns is also subject to change.
 The programmer
@@
 robust to changes in the values returned by these functions.
+In our experience this either involves avoiding the use of $\getnumgroups()$ at all, or writing a kernel where the only opportunity for the number of groups to change comes from $\resizingglobalbarrier()$, in which case the number of active groups does offer some stability, as discussed below.
 
@@
 that a workgroup may be killed even if its associated task pool is not
-empty, since remaining tasks could eventually be stealed by other
+empty, since remaining tasks could eventually be stolen by other
 workgroups.  Care must be taken on computing pool identifiers: the
@@
 only calls to $\resizingglobalbarrier$ can cause the number of active
-workgroups to change and, as discussed in \mysec\ref{sec:semantics},
-this function causes synchronisation between all threads and causes
-the number of active workgroups to change atomically.  At any point
+workgroups to change.  At any point
 during execution, the threads of a kernel are executing between some
 pair of resizing barrier calls, which we call a \emph{resizing barrier
-  intervals} (here we consider the kernel entry and exit points
+  interval} (here we consider the kernel entry and exit points
 conceptually to be special cases of resizing barriers).  The number of
@@
 generous as possible to the cooperative kernel in this regard.
+
 However, these are not hard-and-fast rules: a scheduler might employ
@@
 that are likely to arrive imminently.
+With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.
 
@@
 
+\ADComment{Sreepathi mentioned the NCCL framework.}
+
 Lots of work on irregular algorithms (assumes fairness), and persistent threads (assumes OB model, which assumes a degree of fairness).  Our contribution can enable these to be safe.

### 2016-11-13 write-ups/PLDI/paper.tex
Some related work text.

@@
 
+\paragraph{Irregular algorithms and persistent kernels}
+%
+There has been a lot of work on accelerating blocking irregular
+algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
+programming style for long-running kernels.  These approaches rely on
+the occupancy-bound execution model, and aim to take ownership of the
+GPU by scheduling as many workgroups as the GPU has compute units,
+relying on fair scheduling between these workgroups.  These approaches
+render the GPU unavailable for other tasks, and are not future-proof
+because (as discussed throughout the paper) the occupancy-bound
+execution model is not guaranteed by GPU programming models and is
+unlikely to be guaranteed in practice by future GPU platforms.
+
+\ADComment{Now say that we could update these approaches to use cooperation.}
+
 \ADComment{Sreepathi mentioned the NCCL framework.}

### 2016-11-13 src/test_suite/connect_four/connect_four.h
Fix compilation on Linux

@@ -124,10 +124,22 @@ void init_persistent_app_for_occupancy(CL_Execution *exec)
 
 int set_persistent_app_args_for_occupancy(int arg_index, cl::Kernel k) {
   // Set dummy args for persistent kernel
-  int num_args = 14;
-  for (int i = 0; i < num_args; i++) {
-    check_ocl(k.setArg(arg_index++, NULL));
-  }
+
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, 0));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, 0));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, 0));
+  check_ocl(k.setArg(arg_index++, 0));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, NULL));
+  check_ocl(k.setArg(arg_index++, NULL));
+
   return arg_index;
 }
 

### 2016-11-13 src/test_suite/connect_four/connect_four.h
Add meaningful checking of persistent computation for connect_four

@@ -346,6 +346,16 @@ bool check_persistent_task(CL_Execution *exec) {
   }
   printf("\n");
 
+  // check values for the board committed in the git repo
+  bool check = false;
+  check |= h_next_move_value[0] == (cl_int)0;
+  check |= h_next_move_value[1] == (cl_int)1;
+  check |= h_next_move_value[2] == (cl_int)PLUS_INF;
+  check |= h_next_move_value[3] == (cl_int)2;
+  check |= h_next_move_value[4] == (cl_int)2;
+  check |= h_next_move_value[5] == (cl_int)0;
+  check |= h_next_move_value[6] == (cl_int)0;
+
   // Root done
   cl_int h_root_done = 0;
   err = exec->exec_queue.enqueueReadBuffer(d_root_done, CL_TRUE, 0, sizeof(cl_int), &h_root_done);
@@ -364,6 +374,7 @@ bool check_persistent_task(CL_Execution *exec) {
 
   // cout << "Debug int: " << h_debug_int << endl;
 
+  return check;
 }
 
 /*---------------------------------------------------------------------------*/

### 2016-11-13 src/run_test.py
Update matmult config for hugues laptop, this is the one used for experiments

@@ -137,9 +137,9 @@ MATMULT_CONFIG_HD5500 = [
 
 MATMULT_CONFIG_HD520 = [
     # This is for Hugues laptop
-    { "freq" : "70", "matdim" : "160", "name" : "light" },
-    { "freq" : "40", "matdim" : "160", "name" : "medium" },
-    { "freq" : "40", "matdim" : "260", "name" : "high" },
+    { "freq" : "70", "matdim" : "200", "name" : "light" },
+    { "freq" : "40", "matdim" : "200", "name" : "medium" },
+    { "freq" : "40", "matdim" : "322", "name" : "high" },
 ]
 
 MATMULT_CONFIG_IRIS = [
@@ -191,7 +191,7 @@ def const_print_time():
 
 def exec_cmd(cmd, prefix="", record_file=""):
     global EXIT_THREAD
-    
+
     cmd = cmd + ["--output_summary", prefix + "_summary"]
     cmd = cmd + ["--output_non_persistent_duration", prefix + "_non_persistent_duration"]
     cmd = cmd + ["--output_timestamp_executing_groups", prefix + "_timestamp_executing_groups"]
@@ -236,7 +236,7 @@ def exec_cmd(cmd, prefix="", record_file=""):
     if (ret_code != 0):
         print("Error running " + " ".join(cmd))
         return ret_code
-    
+
     write_to_checkpoint(prefix)
     return 0
 
@@ -310,7 +310,7 @@ def run_suite():
                     if finalsize == -1:
                         PRINT("Could not find finalsize after run of merged skiptask")
                         continue
-            
+
 
                 # RUN: standalone
                 if (q == 0):
@@ -404,7 +404,7 @@ def main():
     log_file_handle = ""
     if (os.path.isfile(log_file)):
         log_file_handle = open(log_file, "a")
-        
+
     log_file_handle = open(log_file, "w")
     PRINT = lambda x : my_print(log_file_handle,x)
 

### 2016-11-13 write-ups/PLDI/paper.tex

@@
 \begin{lstlisting}
-(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * pools) {
-(*@\label{line:wksteal:getgroupid}@*)  int pool_id = get_group_id();
-  while (more_work(pools)) {
-(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(pools, pool_id);
+(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
+(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
+  while (more_work(queues)) {
+(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
     if (t) {
       // task processing may create more work
-(*@\label{line:wksteal:processtask}@*)      process_task(t, pools, pool_id);
+(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
     }
@@
 number of tasks is not known ahead of time because processing a given
-task may create an arbitrary number of new tasks. Work stealing enables
-dynamic balancing of tasks across several processing units, OpenCL
-workgroups in our case. Each workgroup has an associated task pool from
-which it obtains tasks to process, and to which it stores newly created
-tasks. When a workgroup's pool becomes empty, a workgroup---rather than
-terminating---tries to \emph{steal} a task from an other
-pool. Workgroups keep on processing tasks until all the work is
-done. Previous work addresses the feasibility and efficiency of work
-stealing on GPUs~ref\cite{dlb-web,TPO10}.
+task may create an arbitrary number of new tasks. Work stealing
+enables dynamic balancing of tasks across several processing units,
+OpenCL workgroups in our case. Each workgroup has an associated task
+queue from which it obtains tasks to process, and to which it stores
+newly created tasks. When a workgroup's queue becomes empty, the
+workgroup---rather than terminating---tries to \emph{steal} a task
+from an other queue. Workgroups keep on processing tasks until all the
+work is done. Previous work addresses the feasibility and efficiency
+of work stealing on GPUs~ref\cite{dlb-web,TPO10}.
 
@@
 (line~\ref{line:wksteal:kernelfunc}). They receive a pointer to the task
-pools, which reside in global memory, and which were initialized by the
+queues, which reside in global memory, and which were initialized by the
 host program beforehand to contain the initial tasks. Each thread can
@@
 (line~\ref{line:wksteal:getgroupid}). Here the group id is stored into
-the variable $\mathsf{pool\_id}$ (which is, by default, in private
-memory) and used to access the relevant task pool. The
+the variable $\mathsf{queue\_id}$ (which is, by default, in private
+memory) and used to access the relevant task queue. The
 $\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
@@
 also called the \emph{master thread} of the workgroup, will try to pop a
-task from the workgroup's pool, or steal a task from a different
-pool. The master thread uses local memory inside
+task from the workgroup's queue, or steal a task from a different
+queue. The master thread uses local memory inside
 $\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
@@
 Processing this task may lead to the creation of further tasks, which
-the master thread will push to the workgroup's pool.
-
-Although not depicted here, concurrent accesses to each pool inside
+the master thread will push to the workgroup's queue.
+
+Although not depicted here, concurrent accesses to queues inside
 $\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
-mutex, implemented using atomic compare and swap operations on
-global memory.  As a result, the kernel presents two opportunities for
-spin-waiting: spinning to obtain a mutex, and spinning in the main
+mutex per queue, implemented using atomic compare and swap operations
+on global memory.  As a result, the kernel presents two opportunities
+for spin-waiting: spinning to obtain a mutex, and spinning in the main
 kernel loop to obtain a task to process.  The algorithm thus depends
-on a fair scheduler, which, as discussed in \mysec\ref{sec:intro},
-is not guaranteed by current GPU programming models.
+on a fair scheduler, which, as discussed in \mysec\ref{sec:intro}, is
+not guaranteed by current GPU programming models.
 
@@
 
-For the work stealing example, there is no state to transmit since a
-computation is entirely parameterised by a task, which is retrieved from
-a task pool located in global memory. In the main loop, before trying to
-obtain a task, we make the workgroup offer itself to be forked or killed
+In this example there is no state to transmit since a computation is
+entirely parameterised by a task, which is retrieved from a queue
+located in global memory. In the main loop, before trying to obtain a
+task, we make the workgroup offer itself to be forked or killed
 (lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
-that a workgroup may be killed even if its associated task pool is not
-empty, since remaining tasks could eventually be stolen by other
-workgroups.  Care must be taken on computing pool identifiers: the
+that a workgroup may be killed even if its associated task queue is
+not empty, since remaining tasks could eventually be stolen by other
+workgroups.  Care must be taken on computing queue identifiers: the
 original version in \myfig\ref{fig:workstealing} can rely on the fact
-that there are as many pools as workgroups, but here the number of
-workgroups may change during execution. Therefore, the number of pools
-must now be explicitely given as an argument
-(line~\ref{line:cwksteal:argnumpool}) and used as an upper bound of pool
-indexes when scanning pools in $\mathsf{more\_work}$
+that there are as many queues as workgroups, but here the number of
+workgroups may change during execution. Therefore, the number of
+queues must now be explicitely given as an argument
+(line~\ref{line:cwksteal:argnbqueue}) and used as an upper bound of
+queue indexes when scanning queues in $\mathsf{more\_work}$
 (line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
 (line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
-workgroup's pool index (line~\ref{line:cwksteal:getgroupid}) must not
-only be bounded by the number of pools but also be moved inside the main
-loop since a forked workgroup that starts to execute after $\offerfork$
-must re-compute its pool index.
+workgroup's queue index (line~\ref{line:cwksteal:getgroupid}) must now
+be moved inside the main loop since a forked workgroup that starts to
+execute after $\offerfork$ must re-compute its group id.
 
@@
 \begin{lstlisting}
-kernel work_stealing(global Task * pools,
-(*@\label{line:cwksteal:argnumpool}@*)                     const int num_pools) {
-(*@\label{line:cwksteal:morework}@*)  while (more_work(pools, num_pools)) {
+kernel work_stealing(global Task * queues,
+(*@\label{line:cwksteal:argnbqueue}@*)                     const int nb_queues) {
+(*@\label{line:cwksteal:morework}@*)  while (more_work(queues, nb_queues)) {
 (*@\label{line:cwksteal:offerfork}@*)    offer_fork();
 (*@\label{line:cwksteal:offerkill}@*)    offer_kill();
-(*@\label{line:cwksteal:getgroupid}@*)    int pool_id = get_group_id() % num_pools;
-(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(pools, pool_id, num_pools);
+(*@\label{line:cwksteal:getgroupid}@*)    int queue_id = get_group_id();
+(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id, nb_queues);
     if (t) {
       // task processing may create more work
-(*@\label{line:cwksteal:processtask}@*)      process_task(t, pools, pool_id);
+(*@\label{line:cwksteal:processtask}@*)      process_task(t, queues, queue_id);
     }
@@
 
-\caption{Cooperative version of the work stealing example from \myfig\ref{fig:workstealing}; the number of workpools
-must be explicitly passed as arguments since the number of workgroups
-is not a constant anymore}\label{fig:workstealing-cooperative}
+\caption{Cooperative version of the work stealing kernel of
+  \myfig~\ref{fig:workstealing}, using $\offerfork$ and $\offerkill$,
+  and passing the total number of queues
+  explicitely}\label{fig:workstealing-cooperative}
 \end{figure}

### 2016-11-13 write-ups/PLDI/paper.tex
Just basic ideas

--- 2016-11-13 write-ups/PLDI/paper.tex
@@
 
-Suppose a workgroup with id $m$ executes $\offerkill$.  
-If the workgroup has the highest largest id of all active workgroups then it can be killed by the scheduler, except that workgroup 0 can never be killed (to ensure that the kernel computation does not terminate early).
-More formally, if $m < M-1$
+Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$
 or $M=1$, i.e.\ the workgroup's id is not the largest among the active
@@
 executing as a no-op, or (b) execution of the workgroup ceasing and
-the number of active workgroups $M$ being atomically decremented by one.  Case
+the number of active workgroups $M$ being decremented by one.  Case
 (b) corresponds to the scheduler accepting the workgroup's offer to
@@
 
+The semantics of $\offerkill$ is atomic.  That is, if multiple
+workgroups reach $\offerkill$ simultaneously, the semantics is as if
+each workgroup executes $\offerkill$ in some arbitrary sequential
+order.
+
+In this formulation of cooperative kernels, only the active workgroup
+with the largest id can be killed, and workgroup 0 can never be
+killed.  We discuss the rationale for these choices, as well as
+possible alternative choices, in
+\mysec\ref{sec:semanticalternatives}.
+
 \paragraph{Semantics for $\offerfork$}
 
-The $\offerfork$ primitive (also a workgroup-level function) allows the cooperative kernel to indicate
+The $\offerfork$ primitive allows the cooperative kernel to indicate
 to the scheduler that it could benefit from, and is prepared to
@@
 
-Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  Then the following occurs: an
-integer $k$ in the range $[0, N-M]$ is chosen by the scheduler;
+Like $\offerkill$, $\offerfork$ is a workgroup-level function.
+
+Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  An
+integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
 $k$ new workgroups are spawned with consecutive ids in the range $[M,
-  M+k-1]$; the active workgroup count $M$ is atomically incremented by $k$.
+  M+k-1]$, and the active workgroup count $M$ is incremented by $k$.
 
@@
 Semantically, when all threads have reached $\resizingglobalbarrier$,
-the number of active workgroups, $M$, is atomically set to a new value, $M'$ say, with $M' > 0$.
+the number of active workgroups, $M$, is set to a new value, $M'$ say, with $M' > 0$.
 If $M' = M$ then the active workgroups remain unchanged and proceed to
@@
 
-The enclosing $\globalbarrier$ calls ensure that the change in number
-of active workgroups from $M$ to $M'$ occurs entirely within the
-resizing barrier, so that from a programmer's perspective, $M$ changes atomically.
-The middle $\globalbarrier$ is required to ensure that forking occurs before killing, so that while the barrier may kill a workgroup with id $i$, it should not also fork a new workgroup with id $i$.
-That is, workgroups $[0, \mathit{min}(M, M') - 1]$ should be left intact.
+The enclosing $\globalbarrier$ calls make the operation appear to be
+atomic.  The middle $\globalbarrier$ is required to ensure that if
+$M'\geq M$, the workgroups with ids smaller than $M$ are left intact.
 
@@
 by the requirements of the irregular GPU algorithms proposed in the
-literature.  For these algorithms, it was usually easy to identify
-where to place the constructs.
+literature, and that at least for these algorithms it is usually easy
+to identify where to place the constructs.
 
@@
 the value returned by $\getnumgroups()$ is not fixed during the
-lifetime of a cooperative kernel: it may change due to workgroups executing
-$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
-a function of $\getnumgroups()$, the value it returns is also subject to change.
+lifetime of a cooperative kernel.  Although
+$\getnumgroups()$ has atomic semantics, and so can be called at any
+time to yield a well-defined value, the value returned by two
+successive calls may differ if workgroups have executed $\offerkill$,
+$\offerfork$ or $\resizingglobalbarrier$ in-between.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
+a function of $\getnumgroups()$, the value returned by $\getglobalsize()$ is also subject to change.
 The programmer
@@
 robust to changes in the values returned by these functions.
-In our experience this either involves avoiding the use of $\getnumgroups()$ at all, or writing a kernel where the only opportunity for the number of groups to change comes from $\resizingglobalbarrier()$, in which case the number of active groups does offer some stability, as discussed below.
 
@@
 that a workgroup may be killed even if its associated task pool is not
-empty, since remaining tasks could eventually be stolen by other
+empty, since remaining tasks could eventually be stealed by other
 workgroups.  Care must be taken on computing pool identifiers: the
@@
 only calls to $\resizingglobalbarrier$ can cause the number of active
-workgroups to change.  At any point
+workgroups to change and, as discussed in \mysec\ref{sec:semantics},
+this function causes synchronisation between all threads and causes
+the number of active workgroups to change atomically.  At any point
 during execution, the threads of a kernel are executing between some
 pair of resizing barrier calls, which we call a \emph{resizing barrier
-  interval} (here we consider the kernel entry and exit points
+  intervals} (here we consider the kernel entry and exit points
 conceptually to be special cases of resizing barriers).  The number of
@@
 generous as possible to the cooperative kernel in this regard.
-
 However, these are not hard-and-fast rules: a scheduler might employ
@@
 that are likely to arrive imminently.
-With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.
 
@@
 
+\HEComment{the reason we need SVM fine grain and atomics may be better
+discussed directly in the section on implementation?}
+
+Our implementation approach requires two optional features of OpenCL
+$2.0$, namely SVM fine-grained buffers and SVM atomics, which are
+currently not widely supported. \nvidia chips are ruled out since
+\nvidia does not support OpenCL $2.0$. AMD support SVM atomics only
+their Kaveri family of chips and only on Linux, requiring to compile the
+latest driver.  \HEComment{do we need to mention the struggle for AMD ?}
+Among the GPUs available to us, only 4 support our requirements: Intel
+HD 520, HD 5500 and IRIS 6100, and AMD Radeon R7.
+
 \ADComment{Describe the platforms.  Explain why we could not try other
@@
 
-\ADComment{Tyler and Hugues, please write.}
-
 \subsection{Determining Representative Workloads}\label{sec:representativeworkloads}
@@
 \ADComment{Better section title?}
+\HEComment{Dimensioning Non-persistent Workloads}
+
+To assess how cooperative blocking algorithms can be, we fire lightweight
+but still meaningful workloads that requires workgroups to be taken from
+the blocking algorithms.
+
+We tried to measure how much GPU computation is required to maintain
+smooth animation on the screen during typical desktop usage. We model
+these transient \HEComment{no sure about meaning if transient here}
+using
+
+The idea is
+to run an infinite loop that launches a given kernel
 
@@
 \section{Related Work}\label{sec:relatedwork}
-
-\ADComment{Sreepathi mentioned the NCCL framework.}
 

### 2016-11-13 src/host_skeleton/skel.cpp
got things working for octree

@@ -320,7 +320,7 @@ void run_persistent(CL_Execution *exec) {
 		reset_barrier(exec, d_bar);
 		reset_persistent_task(exec);
 		restart_scheduler(&s_ctx);
-		CL_Communicator::my_sleep(5000);
+		CL_Communicator::my_sleep(1000);
 		err = exec->exec_queue.flush();
 		check_ocl(err);
 		err = exec->exec_queue.finish();
@@ -599,7 +599,7 @@ void run_merged(CL_Execution *exec) {
 		
 
 		cout << endl;
-		CL_Communicator::my_sleep(5000);
+		CL_Communicator::my_sleep(1000);
 	}
 
 	free_scheduler_ctx(exec, &s_ctx);

### 2016-11-13 src/run_test.py

@@ -31,10 +31,11 @@ EXIT_THREAD      = 0
 PROGRAMS = [
 #    "lonestar_sssp"
 #    "lonestar_bfs",
-    "pannotia_color",
-    "pannotia_mis",
-    "pannotia_bc",
-    "pannotia_sssp"
+#    "pannotia_color",
+#    "pannotia_mis",
+#    "pannotia_bc",
+#    "pannotia_sssp",
+    "final_octree"
 ]
 
 PROGRAM_DATA = {
@@ -115,6 +116,10 @@ PROGRAM_DATA = {
                         "query_barrier" : [0,1]}
 
     ],
+    "final_octree" : [ { "stat" : "octree",
+                   "suite" : os.path.join("test_suite","final_octree"),
+                   "query_barrier" : [0]},
+                 ],
 }
 
 MATMULT_CONFIG = []
@@ -192,7 +197,7 @@ def const_print_time():
 
 def exec_cmd(cmd, prefix="", record_file=""):
     global EXIT_THREAD
-
+    
     cmd = cmd + ["--output_summary", prefix + "_summary"]
     cmd = cmd + ["--output_non_persistent_duration", prefix + "_non_persistent_duration"]
     cmd = cmd + ["--output_timestamp_executing_groups", prefix + "_timestamp_executing_groups"]
@@ -203,7 +208,7 @@ def exec_cmd(cmd, prefix="", record_file=""):
         print(prefix + " found in checkpoint")
         return 0;
 
-    time.sleep(30)
+    #time.sleep(30)
     EXIT_THREAD = 0
     local_time_begin = time.time()
     #thread = Thread(target = const_print_time)
@@ -237,7 +242,7 @@ def exec_cmd(cmd, prefix="", record_file=""):
     if (ret_code != 0):
         print("Error running " + " ".join(cmd))
         return ret_code
-
+    
     write_to_checkpoint(prefix)
     return 0
 
@@ -274,18 +279,34 @@ def optional_query(q):
         return "_query"
     return ""
 
+def optional_graph(d, graph_in):
+    if "input" in d:
+        return ["--graph_file", graph_in]
+    return []
+
+def optional_solution(d, graph_sol):
+    if "solution" in d:
+        return ["--graph_solution_file", graph_sol]
+    return []
+
 def run_suite():
     for p in PROGRAMS:
         for d in PROGRAM_DATA[p]:
             for q in d["query_barrier"]:
[…]

### 2016-11-13 src/run_test.py
added iris configuration. Seemed to have gotten lost somewhere

@@ -147,6 +147,13 @@ MATMULT_CONFIG_HD520 = [
     { "freq" : "40", "matdim" : "322", "name" : "high" },
 ]
 
+MATMULT_CONFIG_IRIS = [
+    # This is for purple laptop
+    { "freq" : "70", "matdim" : "290", "name" : "light" },
+    { "freq" : "40", "matdim" : "290", "name" : "medium" },
+    { "freq" : "40", "matdim" : "445", "name" : "high" },
+]
+
 MATMULT_CONFIG_RADEON_R7 = [
     # This is for carrot
     { "freq" : "70", "matdim" : "277", "name" : "light" },

### 2016-11-13 src/run_test.py
got connect four working

@@ -35,7 +35,8 @@ PROGRAMS = [
 #    "pannotia_mis",
 #    "pannotia_bc",
 #    "pannotia_sssp",
-    "final_octree"
+    "final_octree",
+    "connect_four"
 ]
 
 PROGRAM_DATA = {
@@ -118,6 +119,10 @@ PROGRAM_DATA = {
     ],
     "final_octree" : [ { "stat" : "octree",
                    "suite" : os.path.join("test_suite","final_octree"),
+                   "query_barrier" : [0]},
+                 ],
+    "connect_four" : [ { "stat" : "connect_four",
+                   "suite" : os.path.join("test_suite","connect_four"),
                    "query_barrier" : [0]},
                  ],
 }

### 2016-11-13 write-ups/PLDI/paper.tex
First try on sizing non-persistent tasks

@@
 
-\subsection{Determining Representative Workloads}\label{sec:representativeworkloads}
-
-\ADComment{Better section title?}
-\HEComment{Dimensioning Non-persistent Workloads}
-
-To assess how cooperative blocking algorithms can be, we fire lightweight
-but still meaningful workloads that requires workgroups to be taken from
-the blocking algorithms.
-
-We tried to measure how much GPU computation is required to maintain
-smooth animation on the screen during typical desktop usage. We model
-these transient \HEComment{no sure about meaning if transient here}
-using
-
-The idea is
-to run an infinite loop that launches a given kernel
-
-\ADComment{Hugues, can you please explain your experimental
-  methodology for working out required response times here?}
+\subsection{Sizing Non-Persistent Workloads}\label{sec:representativeworkloads}
+
+\HEComment{given the paper title, maybe ``non-blocking'' would be more
+  appropriate than ``non-persistent''?}
+
+Cooperative kernels enables other \emph{non-persistent} workloads to
+access GPU resources without stopping the blocking algorithm
+computation entirely. We size these workloads to represent
+approximatively the workload required by a system to maintain smooth
+screen animations. We repetitively ran a kernel with a controlled
+workload and detected from which workload size the screen animation
+started to jitter. Given the kernel workload and its scheduling
+frequency, we can infer what is the workload and frequency needed by
+the system to maintain a smooth animation.
+
+Based on this approach, we measured workloads required by the system
+to maintain smooth animations of three types: (\emph{light}) emphasis
+of desktop icons under the mouse pointer, (\emph{medium}) window
+dragging over the desktop, and (\emph{heavy}) WebGL animations in a
+browser. \mytab~\ref{tab:nonpersistentworkload} illustrates our
+results. In order to simulate such non-persistent workloads alongside
+kernels of blocking algorithms, we use a matrix multiplication kernel
+where matrix dimensions are tuned to trigger the relevant execution
+duration.
+
+\begin{table}
+\small
+\centering
+\begin{tabular}{ l c c }
+workload & duration (ms) & period (ms)\\
+\hline
+light & 3 & 70\\
+medium & 3 & 40\\
+heavy & 10 & 40\\
+\end{tabular}
+\caption{Caracteristics of non-persitent workloads}
+\label{tab:nonpersistentworkload}
+\end{table}
 

### 2016-11-13 write-ups/PLDI/paper.tex
typo
Edit "the gpu we tested"

@@
 \begin{tabular}{ l c c c }
-app & resizing barriers & ckill - cfork & tranmit vars\\
+app & resizing barriers & ckill - cfork & transmit vars\\
 \hline
@@
 
+\ADComment{Describe the platforms.  Explain why we could not try other
+platforms.}
+
 \HEComment{the reason we need SVM fine grain and atomics may be better
@@
 currently not widely supported. \nvidia chips are ruled out since
-\nvidia does not support OpenCL $2.0$. AMD support SVM atomics only
-their Kaveri family of chips and only on Linux, requiring to compile the
-latest driver.  \HEComment{do we need to mention the struggle for AMD ?}
-Among the GPUs available to us, only 4 support our requirements: Intel
-HD 520, HD 5500 and IRIS 6100, and AMD Radeon R7.
-
-\ADComment{Describe the platforms.  Explain why we could not try other
-platforms.}
+\nvidia does not support OpenCL $2.0$. Intel drivers meet our
+requirements on Windows, but not on Linux, although we tried the
+upstream driver that requires a patched Linux kernel. AMD support SVM
+atomics only their Kaveri family of chips and only on Linux, requiring
+to compile the latest driver. Among the GPUs available to us, only 4
+turned out to provide our requirements: Intel HD 520, HD 5500 and IRIS
+6100, and AMD Radeon R7.
+
+\HEComment{To decide: do we mention carrot AMD, in which terms?}
 

### 2016-11-13 write-ups/PLDI/paper.tex
Edit section on sizing non-persistent kernels

@@
 approximatively the workload required by a system to maintain smooth
-screen animations. We repetitively ran a kernel with a controlled
-workload and detected from which workload size the screen animation
-started to jitter. Given the kernel workload and its scheduling
-frequency, we can infer what is the workload and frequency needed by
-the system to maintain a smooth animation.
-
-Based on this approach, we measured workloads required by the system
-to maintain smooth animations of three types: (\emph{light}) emphasis
-of desktop icons under the mouse pointer, (\emph{medium}) window
-dragging over the desktop, and (\emph{heavy}) WebGL animations in a
-browser. \mytab~\ref{tab:nonpersistentworkload} illustrates our
-results. In order to simulate such non-persistent workloads alongside
-kernels of blocking algorithms, we use a matrix multiplication kernel
-where matrix dimensions are tuned to trigger the relevant execution
-duration.
+screen animations.
+
+\paragraph{Experimental setup} \ADComment{Describe the data we are
+  gathering.}  We repetitively run a kernel with a controlled workload
+and detect from which workload size the screen animation starts to
+jitter. Given the kernel workload and its scheduling frequency, we
+deduce what is the workload and frequency needed by the system to
+maintain a smooth animation. We use this approach to measure required
+by the system to maintain smooth animations of three types:
+(\emph{light}) emphasis of desktop icons under the mouse pointer,
+(\emph{medium}) window dragging over the desktop, and (\emph{heavy})
+WebGL animations in a browser.
 
@@
 
+\paragraph{Results} \ADComment{Discuss the results.}
+\mytab~\ref{tab:nonpersistentworkload} illustrates the caracteristics
+found for the three types of workloads. For the medium and heavy
+workloads, the 40ms period corresponds to an execution every $25^{th}$
+of seconds which coincide with the human persistence of vision. The
+3ms execution duration of both light and medium configurations may
+indicate a lower bound of GPU execution time for system workloads.
+
+\subsection{The Overhead of Moving to Cooperative Kernels}
+
+\ADComment{Results for stand-alone vs.\ cooperative without co-scheduling.}
+
 \paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
@@
 
-\subsection{The Overhead of Moving to Cooperative Kernels}
-
-\ADComment{Results for stand-alone vs.\ cooperative without co-scheduling.}
-
-\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
-
-\paragraph{Results}  \ADComment{Discuss the results.}
-
 \subsection{Performance and Responsiveness of Cooperative Scheduling}\label{sec:responsiveness}
@@
 \ADComment{Preamble}
+
+In order to simulate non-persistent workloads alongside kernels of
+blocking algorithms, we use a matrix multiplication kernel where
+matrix dimensions are tuned to trigger the relevant execution
+duration.
 

### 2016-11-13 src/run_test.py
fixed up the script for the new lonestar and bc approach

@@ -84,35 +84,25 @@ PROGRAM_DATA = {
                           "query_barrier" : [0,1]}
     ],
 
-    "lonestar_sssp" : [ { "input" : os.path.join("lonestar", "inputs","rmat22.gr"),
-                        "solution" : os.path.join("lonestar","solutions", "sssp_rmat22.txt"),
-                        "stat" : "sssp_rmat22",
+    "lonestar_sssp" : [ { "input" : os.path.join("lonestar", "inputs","rmat20.gr"),
+                        "solution" : os.path.join("lonestar","solutions", "sssp_rmat20.txt"),
+                        "stat" : "sssp_rmat20",
                         "suite" : "lonestar",
                         "query_barrier" : [0,1]},
-                      { "input" : os.path.join( "lonestar", "inputs", "r4-2e23.gr"),
-                        "solution" : os.path.join("lonestar","solutions", "sssp_r4-2e23.txt"),
-                        "stat" : "sssp_r4-2e23",
+                      { "input" : os.path.join( "lonestar", "inputs", "USA-road-d.NY.gr"),
+                        "solution" : os.path.join("lonestar","solutions", "sssp_usa_ny.txt"),
+                        "stat" : "sssp_usa_ny",
                         "suite" : "lonestar",
                         "query_barrier" : [0,1]},
-                   #   { "input" : os.path.join( "lonestar", "inputs", "USA-road-d.W.gr"),
-                   #     "solution" : os.path.join("lonestar","solutions", "sssp_USA_W.txt"),
-                   #     "stat" : "sssp_USA-road-d.W",
-                   #     "suite" : "lonestar",
-                   #     "query_barrier" : [0,1]}
     ],
     "lonestar_bfs" : [ { "input" : os.path.join( "lonestar", "inputs", "rmat22.gr"),
                         "solution" : os.path.join("lonestar","solutions", "bfs_rmat22.txt"),
                         "stat" : "bfs_rmat22",
                         "suite" : "lonestar",
                         "query_barrier" : [0,1]},
-                      { "input" : os.path.join( "lonestar", "inputs", "r4-2e23.gr"),
-                        "solution" : os.path.join("lonestar","solutions", "bfs_r4-2e23.txt"),
-                        "stat" : "bfs_r4-2e23",
-                        "suite" : "lonestar",
-                        "query_barrier" : [0,1]},
[…]

### 2016-11-13 write-ups/PLDI/paper.tex
Shortened abstract.

@@
 do not mandate fair scheduling, and GPU schedulers are inherently
-unfair in practice.  Current irregular algorithms only work on today's
-GPUs by identifying hardware compute units and scheduling workgroups
-exclusively to these compute units.  As well as being non-portable,
-this is unacceptable because the GPU cannot be used in the meantime
-for high-priority tasks with soft real-time constraints,
-such as graphics rendering.
-
-We propose \emph{cooperative kernels}, an extension to the traditional
-GPU programming model geared towards writing blocking algorithms.
-Cooperative kernels enable fair scheduling of workgroups executing a
-blocking algorithm, as well as co-scheduling of other tasks on the
-GPU, via a small set of language extensions that create a contract
-between the programmer and the scheduler.  We describe our cooperative
-kernels programming model and its semantics in detail, and our
-implementation of a portable prototype of cooperative kernels on top
-of OpenCL 2.0.  We evaluate the technique by porting a set of
-irregular work stealing and graph algorithms to our programming model
-and evaluate the performance and responsiveness of the approach across
-four GPUs from two vendors.  Because our implementation exploits no
-vendor-specific hardware, driver or compiler support, our encouraging
-experimental results provide a lower-bound on the efficiency with which
-cooperative kernels can be implemented in practice.
+unfair in practice.  This issue is currently circumvented by
+exploiting scheduling quirks particular to today's GPUs in a manner
+that does not allow the GPU to be shared with other workloads (such as
+graphics rendering tasks).  We propose \emph{cooperative kernels}, an
+extension to the traditional GPU programming model geared towards
+writing blocking algorithms.  Workgroups of a cooperative kernel are
+fairly scheduled, supporting blocking algorithms, and co-scheduling of
+other tasks is supported via a small set of language extensions that
+create a contract between the programmer and the scheduler.  We
+describe the semantics of the cooperative kernels programming model
+and discuss our portable prototype implementation.  We evaluate the
+technique by porting a set of irregular work stealing and graph
+algorithms to our programming model and assessing the performance and
+responsiveness of the approach across several GPUs.  Because our
+implementation exploits no vendor-specific hardware, driver or
+compiler support, our encouraging experimental results provide a
+lower-bound on the efficiency with which cooperative kernels can be
+implemented in practice.
 
@@
 unlikely to be guaranteed in practice by future GPU platforms.
-
+%
 \ADComment{Now say that we could update these approaches to use cooperation.}
 
+\paragraph{GPU multitasking}
+%
+\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
+  keep a GPU busy by flooding it with work.  Would absolutely not
+  stand up to preemption.}
+
+\cite{DBLP:conf/usenix/KatoLRI11} \ADComment{Idea seems to be to
+  accept that GPUs do not offer epreemption, but find a reasonable way
+  to schedule tasks to maximise responsiveness or throughput.}
+
+\paragraph{Preemption support for GPUs}
+%
+\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
+  implement preemption, i.e.\ shows that it can be achieved.
+  Simulator only preemption.  Not sure whether their approach would
+  give fair scheduling.}
+%
+Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
+%
+\ADComment{Kepler preemption support?  Mention in related work?}
+
 \ADComment{Sreepathi mentioned the NCCL framework.}
 
-Lots of work on irregular algorithms (assumes fairness), and persistent threads (assumes OB model, which assumes a degree of fairness).  Our contribution can enable these to be safe.
-
-Work on multi-tasking to improve throughput, etc., but this does not relate to fairness and wouldn't help blocking algorithms.
-
-Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
-
 Dynamic parallelism could be extended to give barriers, but doesn't work with work stealing (less general).  It is in recent OpenCL.
-
-\ADComment{Kepler preemption support?  Mention in related work?}
 
@@
   should understand this one carefully.}
-
-\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
-  implement preemption, i.e.\ shows that it can be achieved.
-  Simulator only preemption.  Not sure whether their approach would
-  give fair scheduling.}
-
-\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
-  keep a GPU busy by flooding it with work.  Would absolutely not
-  stand up to preemption.}
-
-\cite{DBLP:conf/usenix/KatoLRI11} \ADComment{Idea seems to be to
-  accept that GPUs do not offer epreemption, but find a reasonable way
-  to schedule tasks to maximise responsiveness or throughput.}
 

### 2016-11-13 write-ups/PLDI/paper.tex
Shortening intro.

@@
 \paragraph{The needs of irregular data-parallel algorithms}
-Acceleration of general-purpose computations on graphics processing
-units (GPUs) has tended to focus on \emph{regular} data-parallel
-algorithms, for which work to be processed can be evenly split between
-the workgroups that execute a GPU kernel ahead of time.  However, many
+%Acceleration of general-purpose computations on graphics processing
+%units (GPUs) has tended to focus on \emph{regular} data-parallel
+%algorithms, for which work to be processed can be evenly split between
+%the workgroups that execute a GPU kernel ahead of time.  However,
+Many
 interesting data-parallel algorithms are \emph{irregular}: the amount
@@
 between workgroups, to balance load and communicate intermediate
-results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
-As another example, work stealing algorithms require each workgroup
+results.  Many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
+Work stealing algorithms require each workgroup
 to maintain a queue, which is typically protected by a mutex to enable
@@
 To operate correctly, a blocking concurrent algorithm requires
-\emph{fair} scheduling of threads.  In the context of irregular
-data-parallel algorithms accelerated on a GPU, this translates to
-requiring fair scheduling of workgroups.  Without fairness, a
+\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
 workgroup may fail to make progress, leading to starvation.  For
@@
 HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
-workgroups, and implementations of these programming models do not
+workgroups, and current GPUs do not
 provide fair scheduling in practice.  Roughly speaking, each workgroup

### 2016-11-13 write-ups/PLDI/paper.tex
Add references to cooperative multitasking

--- 2016-11-13 write-ups/PLDI/paper.tex
@@
 
+\HEComment{cooperative multitasking: OS: RISC OS, IBM CICS, Windows
+  before 95, old MacOS, maybe Oberon? TRON?. In GPU context: some
+  paper mention ``cooperative'' multitasking but they simply refer to
+  the fact that kernels should not compute for too long in order to
+  let other have access to it. Also, many prog languages implements
+  variations of coroutines, We can see blocking channels of CSP (or
+  Golang more recently) as yielding (sending a message is yielding
+  computation to the process which receives it)}
+
+Cooperative multitasking was offered in ancient operating systems
+(e.g. Windows before the 95 edition) and is still used nowadays in
+RISC OS for instance. In the context of GPU resources, cooperative
+multitasking seems to refer to sizing kernels such that they do not
+compute for too long in order to yield the whole GPU
+regularly~\cite{adriaens2012case}. Our contribution present a much
+more flexible cooperative multitasking, where the long-running
+application is never totally stopped.
+
+\HEComment{ RISC OS see :
+\url{http://www.riscos.info/index.php/Preemptive_multitasking}
+}
+
+% ###
+% Cooperative multitasking for GPU-accelerated grid systems
+
+% http://onlinelibrary.wiley.com/doi/10.1002/cpe.1722/full
+
+% @article {CPE:CPE1722,
+% author = {Ino, Fumihiko and Ogita, Akihiro and Oita, Kentaro and Hagihara, Kenichi},
+% title = {Cooperative multitasking for GPU-accelerated grid systems},
+% journal = {Concurrency and Computation: Practice and Experience},
+% volume = {24},
+% number = {1},
+% publisher = {John Wiley & Sons, Ltd},
+% issn = {1532-0634},
+% url = {http://dx.doi.org/10.1002/cpe.1722},
+% doi = {10.1002/cpe.1722},
+% pages = {96--107},
+% keywords = {multitasking, GPU, CUDA, grid computing},
+% year = {2012},
+% }
+
+% Not super interesting: they just divide blocking computation in
+% smaller independent kernels, and fires them at regular intervals to
+% let the system be responsive.
+
 \paragraph{Irregular algorithms and persistent kernels}

### 2016-11-13 src/run_test.py
fixed the script

@@ -17,7 +17,7 @@ def fake_placeholder(x):
 EXE_PATH      = ""
 DATA_PATH     = ""
 STAT_PATH     = ""
-ITERATIONS    = "10"
+ITERATIONS    = "1"
 PRINT         = fake_placeholder
 DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""
@@ -29,14 +29,14 @@ TIME_BEGIN       = 0.0
 EXIT_THREAD      = 0
 
 PROGRAMS = [
-#    "lonestar_sssp"
-#    "lonestar_bfs",
+    "lonestar_sssp",
+    "lonestar_bfs",
 #    "pannotia_color",
 #    "pannotia_mis",
-#    "pannotia_bc",
+    "pannotia_bc",
 #    "pannotia_sssp",
-    "final_octree",
-    "connect_four"
+#    "final_octree",
+#    "connect_four"
 ]
 
 PROGRAM_DATA = {
@@ -189,7 +189,7 @@ def const_print_time():
     i = 0
     local_time_begin = time.time()
     while (EXIT_THREAD != 1) and (time.time() - local_time_begin < 1200):
-        time.sleep(10)
+        time.sleep(20)
         print("time update: " + str(i) + " - " + str(time.time() - local_time_begin))
         i = i + 1
     if (EXIT_THREAD == 1):
@@ -292,7 +292,7 @@ def optional_solution(d, graph_sol):
     return []
 
 def special_case(app, npconfig):
-    if app in ["pannotia_bc","lonestar_sssp", "lonestar_bfs"] && CHIP_NAME in ["IRIS", "HD5500"] && npconfig == "npwg_all_but_one":
+    if app in ["pannotia_bc","lonestar_sssp", "lonestar_bfs"] and NAME_OF_CHIP in ["IRIS", "HD5500"] and npconfig == "npwg_all_but_one":
         return True
     return False
 

### 2016-11-13 src/run_test.py
fixed script again

@@ -17,7 +17,7 @@ def fake_placeholder(x):
 EXE_PATH      = ""
 DATA_PATH     = ""
 STAT_PATH     = ""
-ITERATIONS    = "1"
+ITERATIONS    = "10"
 PRINT         = fake_placeholder
 DATA_PRINT    = fake_placeholder
 NAME_OF_CHIP  = ""

### 2016-11-13 write-ups/PLDI/paper.tex
More intro shortening.

@@
 between workgroups, to balance load and communicate intermediate
-results.  Many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level.
-Work stealing algorithms require each workgroup
+results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level, and 
+work stealing algorithms require each workgroup
 to maintain a queue, which is typically protected by a mutex to enable
@@
 workgroup may fail to make progress, leading to starvation.  For
-example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
-another workgroup $B$ to spin-wait forever for the mutex to be
+example, if one workgroup holds a mutex, an unfair scheduler may cause
+another workgroup to spin-wait forever for the mutex to be
 released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
-indefinitely at a global barrier so that other workgroups do not make progress
-towards the barrier.
+indefinitely at a global barrier so that other workgroups do not reach the barrier.
 
@@
 %
-If a kernel is executed with more workgroups than there are compute
-units then at least two workgroups must share a single compute unit.
-The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
+The simplest way for a GPU driver to handle 
+the case where more workgroups are launched than there are compute
+units is via an \emph{occupancy-bound}
 execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}.
@@
 workgroup completes execution.  Yet the execution model \emph{does}
-provide fair scheduling between \emph{occupant} workgroups, because
-they are bound to separate compute units that operate in parallel.
-Current implementations of irregular data parallel algorithms on GPUs
-assume that the target GPU device provides the occupancy-bound
-execution model.  They exploit the model by scheduling no more workgroups
-than there are available compute units~\cite{...}. \ADComment{Tyler, please pick best cite for previous claim.} This works
-because \emph{today}'s GPUs seem to provide the occupancy-bound
-execution model.
+provide fair scheduling between \emph{occupant} workgroups, which are
+bound to separate compute units that operate in parallel.  Current GPU
+implementations of irregular data parallel algorithms assume the
+occupancy-bound execution model, which they exploit by launching 
+no more workgroups than there are available compute
+units~\cite{...}. \ADComment{Tyler, please pick best cite for previous
+  claim.} This works because \emph{today}'s GPUs seem to provide the
+occupancy-bound execution model.
 
@@
 models actually mandate occupancy-bound execution.  Further, there are
-clear reasons why this model is undesirable from the point-of-view of
-GPU vendors and end users.  First, the execution model does not enable
+clear reasons why this model is undesirable.
+% from the point-of-view of
+%GPU vendors and end users.
+First, the execution model does not enable
 multi-tasking, because a workgroup effectively \emph{owns} a compute
@@
 kernel occupies all compute units, the GPU cannot be used for other
-tasks, such as graphics rendering, in the meantime.  Precisely because
-of this problem, operating systems employ a GPU \emph{watchdog} to
-heuristically detect and kill long-running computations on the GPU
-being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.  Second, \emph{energy throttling} is
+tasks, such as graphics rendering, in the meantime.
+%Precisely because
+%of this problem, operating systems employ a GPU \emph{watchdog} to
+%heuristically detect and kill long-running computations on the GPU
+%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
+Second, \emph{energy throttling} is
 an important concern for battery-powered devices, such as smartphones
@@
 workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
-retain the essence of the occupancy-bound execution model, which is
-simple to implement efficiently, and provide support for preemption
+retain the essence of the occupancy-bound model, which is
+simple to implement efficiently, supporting preemption
 only in key special cases.
@@
 \emph{cooperative} kernel, and is written using two additional
-language primitives, $\offerkill$ and $\offerfork$, which must be
-placed carefully by the programmer.
+language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.
 
@@
 workgroups, a workgroup can execute $\offerkill$, offering to
-sacrifice itself to the scheduler.  This indicates to the scheduler
-that the workgroup would ideally continue with its execution, but that
-if required the scheduler may destroy the workgroup for good: the
-cooperative kernel is prepared to deal with either scenario.
+sacrifice itself to the scheduler.  This indicates that the workgroup
+would ideally continue with its execution but that, if required, the
+scheduler may destroy the workgroup for good: the cooperative kernel
+is prepared to deal with either scenario.
 
@@
 additional workgroups join the computation, a workgroup can execute
-$\offerfork$.  This indicates to the scheduler that the cooperative
+$\offerfork$.  This indicates that the cooperative
 kernel is prepared to proceed with the existing set of workgroups, but
@@
 The use of $\offerfork$ and $\offerkill$ creates a contract between
-the scheduler and (the programmer of the) cooperative kernel, with
-both functional and non-functional requirements.  Functionally, the
+the scheduler and (the programmer of the) cooperative kernel.
+Functionally, the
 scheduler must guarantee that the workgroups executing a cooperative
@@
 \item The model has no impact on the development of regular
-  (non-cooperative) kernels: these can be programmed exactly as they
-  are now, as can graphics shading routines written using APIs such as
-  OpenGL, Vulkan and DirectX.
-
-\item The model is backwards-compatible: a non-blocking algorithm
-  implementation that works on today's GPU platforms can be upgraded
-  to use our language extensions for cooperative scheduling, and will
-  behave exactly as it does at present if the new $\offerkill$ and
-  $\offerfork$ functions are simply ignored.
+  (non-cooperative) compute and graphics kernels: these can be programmed exactly as they
+  are now.
+%, as can graphics shading routines written using APIs such as
+%OpenGL, Vulkan and DirectX.
+
+\item The model is backwards-compatible: if $\offerkill$ and
+  $\offerfork$ are ignored, a cooperative blocking kernel will behave
+  exactly as a regular blocking kernel does on current GPUs.
 
@@
 
-\item Our experiments across four GPUs from two vendors show that our
-  portable, proof-of-concept implementation of cooperative kernels can
-  provide efficient multi-tasking of cooperative and non-cooperative
-  tasks.  A GPU vendor implementation could likely provide even better
-  efficiency, by leveraging platform-specific hardware, driver and
-  compiler features.
+\item Our experiments with a range of applications using three GPUs show that the model can enable efficient multi-tasking of cooperative and non-cooperative
+  tasks.
 
@@
 
-\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms.
-We describe the semantics of the programming model in detail (\mysec\ref{sec:cooperativekernels}).
+\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).
 
@@
   algorithms to use cooperative kernels, and experimental results
-  assessing the performance overhead associated with cooperation and
-  the responsiveness to high priority tasks that cooperation enables,
-  across four GPUs from two vendors (\mysec\ref{sec:experiments}).
+  using three GPUs assessing the overhead associated with cooperative
+  kernels, and the responsiveness of the multi-tasking that the
+  approach enables (\mysec\ref{sec:experiments}).
 

### 2016-11-14 write-ups/PLDI/paper.tex
Add expe set up on standalone vs skip task

@@
 
-\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
+\paragraph{Experimental setup} \ADComment{Describe the data we are
+  gathering.}  The introduction of cooperative scheduling primitives
+introduces an overhead even if no killing, forking or resizing ever
+happen. We measure this overhead by comparing execution times of the
+original kernel and the cooperative one, forcing the scheduler to
+never modify the number of active workgroups. Since the cooperative
+kernel contains the non-persistent task code anyway, its occupancy
+bound may be less than the original kernel; we make sure to use this
+occupancy bound as the number of workgroups for the original kernel in
+order to compare relevant execution times.
 

### 2016-11-14 write-ups/PLDI/paper.tex
Add comment
Add more references for cooperative multitasking

@@
 
+\HEComment{Also include this occupancy reduction as an overhead?}
+
 \paragraph{Results}  \ADComment{Discuss the results.}
@@
 (e.g. Windows before the 95 edition) and is still used nowadays in
-RISC OS for instance. In the context of GPU resources, cooperative
-multitasking seems to refer to sizing kernels such that they do not
-compute for too long in order to yield the whole GPU
-regularly~\cite{adriaens2012case}. Our contribution present a much
-more flexible cooperative multitasking, where the long-running
-application is never totally stopped.
+RISC OS for instance. Many programming languages offer cooperative
+multitasking via variations of coroutines, e.g. coroutine objects in
+Python\footnote{https://docs.python.org/3/reference/datamodel.html#coroutine-objects}}
+or Lua\footnote{https://www.lua.org/pil/9.1.html}, or Ruby's fiber
+\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}. In the
+context of GPU resources, cooperative multitasking seems to refer to
+sizing kernels such that they do not compute for too long in order to
+yield the whole GPU regularly~\cite{adriaens2012case}. Our
+contribution present a much more flexible cooperative multitasking,
+where the long-running application is never totally stopped.
 

### 2016-11-14 write-ups/PLDI/paper.tex
Fix compilation

--- 2016-11-14 write-ups/PLDI/paper.tex
@@
 (e.g. Windows before the 95 edition) and is still used nowadays in
-RISC OS for instance. In the context of GPU resources, cooperative
-multitasking seems to refer to sizing kernels such that they do not
-compute for too long in order to yield the whole GPU
-regularly~\cite{adriaens2012case}. Our contribution present a much
-more flexible cooperative multitasking, where the long-running
-application is never totally stopped.
+RISC OS for instance. Many programming languages offer cooperative
+multitasking via variations of coroutines, e.g. coroutine objects in
+Python\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
+or Lua\footnote{\url{https://www.lua.org/pil/9.1.html}}, or Ruby's
+fiber \footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}. In
+the context of GPU resources, cooperative multitasking seems to refer
+to sizing kernels such that they do not compute for too long in order
+to yield the whole GPU regularly~\cite{adriaens2012case}. Our
+contribution present a much more flexible cooperative multitasking,
+where the long-running application is never totally stopped.
 

### 2016-11-14 write-ups/PLDI/paper.tex
Add transmit as a keyword
Add table and figure for overhead

@@
   %commentstyle=\rmfamily\itshape,
-  morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then},
+  morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then,transmit},
   escapeinside={(*@}{@*)},
@@
 \begin{tabular}{ l c c }
-workload & duration (ms) & period (ms)\\
+Workload & Duration (ms) & Period (ms)\\
 \hline
@@
 
-\paragraph{Experimental setup} \ADComment{Describe the data we are
-  gathering.}  The introduction of cooperative scheduling primitives
-introduces an overhead even if no killing, forking or resizing ever
-happen. We measure this overhead by comparing execution times of the
-original kernel and the cooperative one, forcing the scheduler to
-never modify the number of active workgroups. Since the cooperative
-kernel contains the non-persistent task code anyway, its occupancy
-bound may be less than the original kernel; we make sure to use this
-occupancy bound as the number of workgroups for the original kernel in
-order to compare relevant execution times.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% chip: hd520
+% overall - 
+% geomean: 1.19835906125
+% max ('color', 'ecology', 1.749620120141443)
+% ----- 
+% barrier 
+% max - ('color', 'ecology', 1.749620120141443)
+% geomean: 1.18027192048
+% ----- 
+% workstealing
+% max - ('octree', 'NA', 1.4165788173996423)
+% geomean: 1.4165788174
+% ----- 
+% chip: iris
+% overall - 
+% geomean: 1.07926294253
+% max ('sssp', 'usa_ny', 1.3661910830147503)
+% ----- 
+% barrier 
+% max - ('sssp', 'usa_ny', 1.3661910830147503)
+% geomean: 1.07122874447
+% ----- 
+% workstealing
+% max - ('octree', 'NA', 1.2333896396396398)
+% geomean: 1.11616906144
+% ----- 
+% chip: hd5500
+% overall - 
+% geomean: 1.15074583378
+% max ('bc', '1k_128k', 1.4466039296624815)
+% ----- 
+% barrier 
+% max - ('bc', '1k_128k', 1.4466039296624815)
+% geomean: 1.16116637952
+% ----- 
+% workstealing
+% max - ('octree', 'NA', 1.1814511945151949)
+% geomean: 1.1049985913
+% ----- 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\small
+\centering
+\begin{tabular}{ l c c c }
+Chip & HD520 & HD5500 & IRIS\\
+\hline
+overall gm  & $1.20$ & $1.15$ & $1.08$ \\
+overall max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
+
+\hline
+barrier gm & $1.18$ & $1.16$ & $1.07$ \\
+barrier max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
+
+\hline
+wk.steal. gm & $1.42$ & $1.10$ & $1.12$ \\
+wk.steal. max & $1.42^{\diamond}$ & $1.18^{\diamond}$ & $1.23^{\diamond}$ \\
+
+\hline
+\end{tabular}\\
+{\footnotesize
+$^{\ast}$color ecology, $^{\dagger}$bc 1k\_128k, $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
+}
+\caption{Overhead of cooperative kernels: execution time slowdown}
+\label{tab:overhead}
+\end{table}
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{cooperative_overhead.pdf}
+\caption{Overhead: ratio of execution time}\label{fig:overhead}
+\end{figure}
+
+\paragraph{Experimental setup}
+\ADComment{Describe the data we are gathering.}
+The introduction of cooperative scheduling primitives introduces an
+overhead even if no killing, forking or resizing ever happen. We
+measure this overhead by comparing execution times of the original
+kernel and the cooperative one, forcing the scheduler to never modify
+the number of active workgroups. Since the cooperative kernel contains
+the non-persistent task code anyway, its occupancy bound may be less
+than the original kernel; we make sure to use this occupancy bound as
+the number of workgroups for the original kernel in order to compare
+relevant execution times.
 
@@
 \ADComment{Preamble}
+
+voc: blocking and non-blocking ?
+
+durable / transient
+
+Cooperative scheduling enables to concurrently run non-persistent kernels alongside blocking ones.
+In terms of performance, we are interested to see whether the variations of workgroups for the blocking kernel.
+
+Its performance corresponds
+
+Performance of cooperative scheduling corresponds to execution time of cooperative kernels compared to the serial
+
+Cooperative scheduling can improve performances
 
@@
 duration.
+
+we compare :
+
+time to execute original kernel
+
+time to execute non-persistent tasks
+
+time to execute both in cooperation
+
+also, response time for non-persistent
+
+use matmult to represent system workload, ref to previous section
 

### 2016-11-14 write-ups/PLDI/paper.tex
Edit overhead

@@
 \centering
-\begin{tabular}{ l c c c }
-Chip & HD520 & HD5500 & IRIS\\
+\begin{tabular}{ l l l l }
+Chip & HD520 & HD5500 & Iris\\
 \hline
-overall gm  & $1.20$ & $1.15$ & $1.08$ \\
+overall geo. mean  & $1.20$ & $1.15$ & $1.08$ \\
 overall max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
@@
 \hline
-barrier gm & $1.18$ & $1.16$ & $1.07$ \\
+barrier geo. mean & $1.18$ & $1.16$ & $1.07$ \\
 barrier max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
@@
 \hline
-wk.steal. gm & $1.42$ & $1.10$ & $1.12$ \\
+wk.steal. geo. mean & $1.42$ & $1.10$ & $1.12$ \\
 wk.steal. max & $1.42^{\diamond}$ & $1.18^{\diamond}$ & $1.23^{\diamond}$ \\
@@
 \ADComment{Describe the data we are gathering.}
-The introduction of cooperative scheduling primitives introduces an
-overhead even if no killing, forking or resizing ever happen. We
-measure this overhead by comparing execution times of the original
-kernel and the cooperative one, forcing the scheduler to never modify
-the number of active workgroups. Since the cooperative kernel contains
-the non-persistent task code anyway, its occupancy bound may be less
-than the original kernel; we make sure to use this occupancy bound as
-the number of workgroups for the original kernel in order to compare
-relevant execution times.
-
-\HEComment{Also include this occupancy reduction as an overhead?}
+
+Cooperative scheduling primitives entails an overhead even if no
+killing, forking or resizing ever happen. We measure this overhead via
+the execution time ratio between the original kernel and the
+cooperative one, forcing the scheduler to never modify the number of
+active workgroups. Since the cooperative kernel contains the
+non-persistent task code anyway, its occupancy bound may be less than
+the original kernel; we make sure to use this occupancy bound as the
+number of workgroups for the original kernel in order to compare
+relevant execution times. \HEComment{Do we need to mention the
+  possible occupancy reduction as an overhead?}
 
 \paragraph{Results} \ADComment{Discuss the results.}
+\mytab~\ref{tab:overhead} lists overhead slowdown ratios, both
+geometric mean and maximum value. The overhead is below a 2 times
+slowdown even in the worst case. We consider these results as
+acceptable, given that our SVM-based scheduler performance is likely a
+far upper bound of what a scheduler implemented directly in the
+runtime could offer.
 

### 2016-11-14 write-ups/PLDI/paper.tex
Edit mostly comments

@@
 
-\HEComment{the reason we need SVM fine grain and atomics may be better
-discussed directly in the section on implementation?}
+\HEComment{I consider that the reason why we need SVM fine grain and
+  atomics is discussed in the implementation section}
 
@@
 \HEComment{given the paper title, maybe ``non-blocking'' would be more
-  appropriate than ``non-persistent''?}
+  appropriate than ``non-persistent''? Or ``non-cooperative'' ?}
 
@@
 
-\paragraph{Results} \ADComment{Discuss the results.}
+\paragraph{Results}
 \mytab~\ref{tab:nonpersistentworkload} illustrates the caracteristics
@@
 \subsection{The Overhead of Moving to Cooperative Kernels}
-
-\ADComment{Results for stand-alone vs.\ cooperative without co-scheduling.}
 
@@
 \paragraph{Experimental setup}
-\ADComment{Describe the data we are gathering.}
-
 Cooperative scheduling primitives entails an overhead even if no
@@
 
-\paragraph{Results} \ADComment{Discuss the results.}
+\paragraph{Results}
 \mytab~\ref{tab:overhead} lists overhead slowdown ratios, both

### 2016-11-14 write-ups/PLDI/paper.tex
Edit overhead

@@
 on properties of the cooperative kernel itself, and on the efficiency
-of the GPU that is used for execution.  In \mysec\ref{sec:representativeworkloads} we conduct
+of the GPU that is used for execution.  In \mysec\ref{sec:sizingnoncoop} we conduct
 experiments to understand the response rate that would be required to
@@
 
-\subsection{Sizing Non-Persistent Workloads}\label{sec:representativeworkloads}
-
-\HEComment{given the paper title, maybe ``non-blocking'' would be more
-  appropriate than ``non-persistent''? Or ``non-cooperative'' ?}
-
-Cooperative kernels enables other \emph{non-persistent} workloads to
-access GPU resources without stopping the blocking algorithm
-computation entirely. We size these workloads to represent
+\subsection{Sizing Non-Cooperative Kernels}\label{sec:sizingnoncoop}
+
+A cooperative kernel can yield workgroups to let a non-cooperative,
+transient one execute. We size non-cooperative kernels to represent
 approximatively the workload required by a system to maintain smooth
@@
 
-\paragraph{Experimental setup} \ADComment{Describe the data we are
-  gathering.}  We repetitively run a kernel with a controlled workload
-and detect from which workload size the screen animation starts to
-jitter. Given the kernel workload and its scheduling frequency, we
-deduce what is the workload and frequency needed by the system to
-maintain a smooth animation. We use this approach to measure required
-by the system to maintain smooth animations of three types:
-(\emph{light}) emphasis of desktop icons under the mouse pointer,
-(\emph{medium}) window dragging over the desktop, and (\emph{heavy})
-WebGL animations in a browser.
+\paragraph{Experimental setup} We repetitively launch a kernel with a
+controlled workload and detect from which workload size the screen
+animation starts to jitter. Given the kernel workload and its
+scheduling frequency, we can deduce what is the workload and frequency
+needed by the system to maintain a smooth animation. We use this
+approach to measure the system workloads needed to maintain smooth
+animations of three types: (\emph{light}) emphasis of desktop icons
+under the mouse pointer, (\emph{medium}) window dragging over the
+desktop, and (\emph{heavy}) WebGL animations in a browser.
 
@@
 \end{tabular}
-\caption{Caracteristics of non-persitent workloads}
-\label{tab:nonpersistentworkload}
+\caption{Caracteristics of non-cooperative workloads}
+\label{tab:noncooperativeworkload}
 \end{table}
@@
 \paragraph{Results}
-\mytab~\ref{tab:nonpersistentworkload} illustrates the caracteristics
+\mytab~\ref{tab:noncooperativeworkload} illustrates the caracteristics
 found for the three types of workloads. For the medium and heavy
 workloads, the 40ms period corresponds to an execution every $25^{th}$
-of seconds which coincide with the human persistence of vision. The
+of seconds, which coincides with the human persistence of vision. The
 3ms execution duration of both light and medium configurations may
@@
 Cooperative scheduling primitives entails an overhead even if no
-killing, forking or resizing ever happen. We measure this overhead via
-the execution time ratio between the original kernel and the
+killing, forking or resizing never happens. We measure this overhead
+via the execution time slowdown between the original kernel and the
 cooperative one, forcing the scheduler to never modify the number of
 active workgroups. Since the cooperative kernel contains the
-non-persistent task code anyway, its occupancy bound may be less than
-the original kernel; we make sure to use this occupancy bound as the
-number of workgroups for the original kernel in order to compare
+non-cooperative kernel code anyway, its occupancy bound may be less
+than the original kernel; we make sure to use this occupancy bound as
+the number of workgroups for the original kernel in order to compare
 relevant execution times. \HEComment{Do we need to mention the
@@
 \paragraph{Results}
-\mytab~\ref{tab:overhead} lists overhead slowdown ratios, both
-geometric mean and maximum value. The overhead is below a 2 times
-slowdown even in the worst case. We consider these results as
-acceptable, given that our SVM-based scheduler performance is likely a
-far upper bound of what a scheduler implemented directly in the
-runtime could offer.
+\mytab~\ref{tab:overhead} lists both geometric mean and maximum value
+of measured slowdowns. The overhead is below a 2 times slowdown even
+in the worst case, and quite light in general. We consider these
+results as encouraging, given that our SVM-based scheduler performance
+is likely a far upper bound of what a scheduler implemented directly
+in the runtime could offer.
 

### 2016-11-14 write-ups/PLDI/paper.tex
added some sample images
added one more figure
Shortening

@@
 
-\item The model is backwards-compatible: if $\offerkill$ and
-  $\offerfork$ are ignored, a cooperative blocking kernel will behave
-  exactly as a regular blocking kernel does on current GPUs.
-
-\item The model can be implemented on top of the occupancy-bound
-  execution model that current GPU platforms provide---we present a
-  proof-of-concept implementation on top of OpenCL 2.0 that requires
+\item The model is backwards-compatible: if $\offerkill$/$\offerfork$ are ignored, a cooperative kernel will behave
+  exactly as a regular kernel does on current GPUs.
+
+\item Cooperative kernels can be implemented over the occupancy-bound
+  execution model that current GPUs provide---we present a
+  proof-of-concept implementation that requires
   no special-purpose hardware or driver support.
@@
 
-Placing the $\offerkill$ and $\offerfork$ primitives requires manual
-effort and care, but our experience porting a representative set of
+Placing the new primitives requires manual effort, but our experience porting a representative set of
 GPU-accelerated irregular algorithms to use cooperative kernels
-suggests that this is straightforward in practice.  We discuss this in
-detail in \mysec\ref{sec:portingalgorithms}.
+(\mysec\ref{sec:portingalgorithms}) suggests that this is straightforward in practice.
 
@@
 
-\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).
+\item \emph{Cooperative kernels}, an extended GPU programming model that support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).
 
 \item A proof-of-concept portable implementation of cooperative
-  kernels on top of the OpenCL 2.0 programming model
+  kernels on top OpenCL 2.0
   (\mysec\ref{sec:implementation}).
 
-\item An experience report of porting \NumAlgorithms{} irregular GPU
-  algorithms to use cooperative kernels, and experimental results
-  using three GPUs assessing the overhead associated with cooperative
-  kernels, and the responsiveness of the multi-tasking that the
-  approach enables (\mysec\ref{sec:experiments}).
+\item Experimental results assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).
 
@@
 
-We provide background on the industry-standard OpenCL programming
-model on which we base cooperative kernels (\mysec\ref{sec:opencl}),
-and illustrate OpenCL using two irregular data-parallel algorithms---a
-work stealing queue and breadth-first graph search---that also
-pinpoint the problems associated with unfair scheduling of workgroups
-and motivate our cooperative threads approach (\mysec\ref{sec:openclexamples}).
+We outline the industry-standard OpenCL programming model on which we
+base cooperative kernels (\mysec\ref{sec:opencl}), and illustrate
+OpenCL and the scheduling requirements of irregular algorithms using two examples: a work stealing queue and frontier-based graph traversal
+(\mysec\ref{sec:openclexamples}).
 
@@
 All threads executing a kernel execute the same entry function with
-identical arguments, but a thread may invoke the $\getglobalid()$
-function to query its unique thread id in order to access different
-data or follow different control flow paths to other threads.
-
-The threads executing a kernel are hierarchically organised.  A kernel
-is executed by a number of \emph{workgroups}, and each workgroup
-consists of a fixed number of threads.  Builtin functions
+identical arguments.  A thread can call $\getglobalid()$
+to obtain its unique id, to access distinct data or follow different control flow paths to other threads.
+
+The threads executing a kernel are divided into a number of \emph{workgroups}, each consisting of a fixed number of threads.  Builtin functions
 $\getlocalid()$ and $\getgroupid()$ return a thread's local id within
-its workgroup and the id of the workgroup, respectively.\footnote{In general, an OpenCL kernel is organised as a multi-dimensional grid
-of multi-dimensional workgroups, and the above functions each take
-an integer dimension argument.  For simplicity, we consider only the
-one-dimensional case throughout the paper, which captures all the
+its workgroup and the id of the workgroup, respectively.\footnote{For simplicity we restrict attention to one-dimensional kernels, which captures all the
 GPU-accelerated irregular algorithms we are aware of, and use
@@
 id are related via the equation $\getglobalid() = \getgroupid() \times
-\getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize()$, which is equal to $\getlocalsize()\times\getnumgroups()$.
-
-The execution of all threads in a workgroup can be synchronised via a
+\getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize() = \getlocalsize()\times\getnumgroups()$.
+
+Execution of the threads in a workgroup can be synchronised via a
 workgroup barrier, which also ensures memory consistency.  This
-barrier is a \emph{workgroup-level} function: it must be encountered
-uniformly by all or none of the threads of a workgroup, thus can only
+barrier is a \emph{workgroup-level} function: it can only
 appear in conditional code if all threads in a workgroup that reach
 the barrier agree on the guards of all enclosing conditional
-statements.  As discussed below, a \emph{global} barrier
-(synchronising all threads running a kernel) is \emph{not} provided as
+statements.  A \emph{global} barrier
+(synchronising all threads of a kernel) is \emph{not} provided as
 a primitive.
@@
 memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
-kernel.  Each workgroup has its own portion of \emph{local} memory,
-which can be used for fast intra-workgroup communication.  Every thread has
+kernel.  Each workgroup has its own portion of \emph{local} memory for fast intra-workgroup communication.  Every thread has
 an allocation of very fast \emph{private} memory, where
@@
 
-Communication between threads in the same workgroup can be achieved
-using the workgroup barrier construct described above.  Finer-grained
+Communication within a workgroup can be achieved
+using a workgroup barrier.  Finer-grained
 communication within a workgroup, as well as inter-workgroup
@@
 workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
-  conforming implementation may choose to serialize the workgroups so
-  a correct algorithm cannot assume that workgroups will execute in
-  parallel.  There is no safe and portable way to synchronize across
+  conforming implementation may choose to serialize the workgroups 
+%so
+%  a correct algorithm cannot assume that workgroups will execute in
+%  parallel.  
+\dots There is no safe and portable way to synchronize across
   the independent execution of workgroups since once in the work-pool,
-  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} OpenCL is not alone in this
-regard: CUDA similarly provides no guarantees~\cite{...}.
+  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} CUDA similarly provides no guarantees~\cite{...}.
 
@@
 
+\begin{figure*}
+\includegraphics[width=.7\columnwidth]{iris_octree_NA.pdf} 
+\includegraphics[width=.7\columnwidth]{hd520_mis_ecology.pdf}
+\includegraphics[width=.7\columnwidth]{hd5500_bfs_rmat22.pdf}
+\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
+\end{figure*}
+
 \section{Related Work}\label{sec:relatedwork}

### 2016-11-14 write-ups/PLDI/paper.tex
Shortenings.

@@
 data-parallel algorithms on GPUs, particularly algorithms that
-traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16}.
-\ADComment{Tyler, can you add relevant workstealing papers if not here already?}
+traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10}.
 
@@
 %
-The work stealing paradigm is useful when a computation can be split
-into multiple tasks that can be processed independently, but where the
-number of tasks is not known ahead of time because processing a given
-task may create an arbitrary number of new tasks. Work stealing
-enables dynamic balancing of tasks across several processing units,
-OpenCL workgroups in our case. Each workgroup has an associated task
-queue from which it obtains tasks to process, and to which it stores
-newly created tasks. When a workgroup's queue becomes empty, the
-workgroup---rather than terminating---tries to \emph{steal} a task
-from an other queue. Workgroups keep on processing tasks until all the
-work is done. Previous work addresses the feasibility and efficiency
-of work stealing on GPUs~ref\cite{dlb-web,TPO10}.
+Work stealing enables dynamic balancing of tasks across several
+processing units, and is useful when the number of tasks to be
+processed in parallel is dynamic, because processing
+one task may create an arbitrary number of new tasks.  In the context
+of GPUs, each workgroup has an associated queue from which it
+obtains tasks to process, and to which it stores newly created
+tasks. If a workgroup's queue is empty, the workgroup tries to \emph{steal} a task from an other
+queue. Workgroups keep on processing tasks until all the work is
+done. Previous work addresses the feasibility and efficiency of work
+stealing on GPUs~\cite{dlb-web,TPO10}.
 
@@
 (line~\ref{line:wksteal:kernelfunc}). They receive a pointer to the task
-queues, which reside in global memory, and which were initialized by the
-host program beforehand to contain the initial tasks. Each thread can
-identify the workgroup it belongs to via the function $\getgroupid()$
+queues, in global memory, initialized by the
+host to contain the initial tasks. Each thread identifies
+the workgroup it belongs to via the function $\getgroupid()$
 (line~\ref{line:wksteal:getgroupid}). Here the group id is stored into
@@
 $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
-also called the \emph{master thread} of the workgroup, will try to pop a
+the \emph{master thread}, will try to pop a
 task from the workgroup's queue, or steal a task from a different
@@
 all threads of its workgroup. If a task was obtained, then all threads
-of the workgroup participate in the processing of the task via the
-$\mathsf{process\_task}$ function (line~\ref{line:wksteal:processtask}).
+of the workgroup participate in the processing of the task via 
+$\mathsf{process\_task}$ (line~\ref{line:wksteal:processtask}).
 Processing this task may lead to the creation of further tasks, which

### 2016-11-14 write-ups/PLDI/paper.tex
Finished shortening background.

@@
 \code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
-A thread processes a node via the $\keyword{process\_node}$
-function, which may (a) push nodes to process in the next frontier to
+A thread processes a node via $\keyword{process\_node}$, which may (a) push nodes to process in the next frontier to
 $\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
 the computation. After processing the frontier, the threads swap their
-local node array pointers (line~\ref{line:graph:swap}), i.e.\ the $\keyword{out\_nodes}$ of the current frontier becomes the $\keyword{in\_nodes}$ of the next
-frontier.
+node array pointers (line~\ref{line:graph:swap}).
 
@@
 primitive, though previous works have shown that such a global
-barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}.
-\ADComment{Are there additional papers we should be citing on GPU barriers?
-E.g., is the sense reversal idea part of XF or is that from another paper?}  These implementations use well
-defined atomic memory operations to ensure memory consistency, and
-employ spinning to ensure threads wait at the barrier until all
-threads have arrived.  This spinning behaviour requires fair
+barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16},
+based on CPU barrier designs~\cite[ch. 17]{HS08}.
+ These barriers employ spinning to ensure threads wait at the barrier until all
+threads have arrived, which requires fair
 scheduling between workgroups for the barrier to operate correctly.
@@
 shown, experimentally, to provide.  But, as discussed in
-\mysec\ref{sec:intro}, this occupancy-bound model is not mandated by
-language standards or vendor implementations, and is unlikely to be
-guaranteed in the future.
-
+\mysec\ref{sec:intro}, this model is not endorsed
+language standards or vendor implementations, and is unlikely to be respected in the future.
+%
 In \mysec\ref{sec:programmingguidelines} we show how the work stealing

### 2016-11-14 write-ups/PLDI/paper.tex
Shortening.

@@
 number of workgroups, each consisting of a specified number of
-threads.  Unlike in a regular kernel, the parameters to a cooperative kernel are \emph{immutable}, as if
-they were qualified with the \texttt{const} keyword (though pointer
+threads.  Unlike in a regular kernel, the parameters to a cooperative kernel are immutable (though pointer
 parameters can refer to mutable data).
 
-A cooperative kernel is written in OpenCL C plus the following
-language extensions: $\transmit$, a
-new qualifier on the local variables of a thread; $\offerkill$ and
-$\offerfork$, the key functions that enable cooperative scheduling;
-$\globalbarrier$ and $\resizingglobalbarrier$, two primitives to allow
-inter-workgroup synchronisation.
-
-We now explain the semantics of these language extensions.  We present the semantics intuitively, using English.
+Cooperative kernels are written using the following 
+extensions: $\transmit$, a qualifier on the variables of a
+thread; $\offerkill$ and $\offerfork$, the key functions that enable
+cooperative scheduling; and $\globalbarrier$ and $\resizingglobalbarrier$
+primitives for inter-workgroup synchronisation.
+%
+We present the semantics of these constructs intuitively, using English.
 In Appendix~\ref{appendix:semantics} (see supplementary material) we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.
@@
 A variable declared in the root scope of the cooperative kernel can
-optionally be annotated with a new $\transmit$ qualifier.  By
-annotating a variable with $\transmit$, the programmer indicates that
-when a workgroup spawns new workgroups by calling $\offerfork$, the
-workgroup should transmit its current value for the variable to serve
-as an initial value for the variable for the threads of the new workgroups.  We detail the semantics
-for this when we describe $\offerfork$ below.
+optionally be annotated with a new $\transmit$ qualifier.  Annotating
+a variable $v$ with $\transmit$ means that when a workgroup spawns new
+workgroups by calling $\offerfork$, the workgroup should transmit its
+current value for $v$ to the threads of the new workgroups, to serve
+as an initial value for $v$.  We detail the semantics for this when we
+describe $\offerfork$ below.
 
@@
 
-The workgroups executing the kernel are called \emph{active
-  workgroups}, and $M$ denotes the number of active workgroups.
-Initially, the number of active workgroups $M$ is initialised to some
-integer between $1$ and $N$---to ensure that at least one workgroup is scheduled, the scheduler will postpone execution of the kernel until a some compute unit becomes available.
-The initial active workgroups have
-consecutive workgroup ids in the range $[0, M-1]$.
-
-The $\getgroupid()$ function retrieves the id of a workgroup as usual
-(see \mysec\ref{sec:opencl}).  When executed by a cooperative
+The number of \emph{active workgroups}---workgroups executing the
+kernel---is denoted $M$.  The active workgroups
+have consecutive workgroup ids in the range $[0, M-1]$.
+Initially, at least one workgroup is active;
+if necessary the scheduler must postpone execution of the kernel until
+a some compute unit becomes available.
+
+When executed by a cooperative
 kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
@@
 regular kernels, which returns the fixed number of workgroups that
-were requested at kernel launch time.
-
-Fair scheduling is guaranteed between active workgroups.  That is, if
+were requested at kernel launch time (see \mysec\ref{sec:opencl}).
+
+Fair scheduling \emph{is} guaranteed between active workgroups.  That is, if
 some thread in an active workgroup is enabled then eventually some
 thread in the active workgroup is guaranteed to execute an
-instruction.  Note that our programming model does not mandate that
-threads within a workgroup must be fairly scheduled, and indeed this
-is often not the case in GPU programming models, because threads are
-often organised into \emph{warps} or \emph{wavefronts} within a
-workgroup, exhibiting lock-step predicated execution that is
-inherently unfair.
+instruction.
+
+%Note that our programming model does not mandate that
+%threads within a workgroup must be fairly scheduled, and indeed this
+%is often not the case in GPU programming models, because threads are
+%often organised into \emph{warps} or \emph{wavefronts} within a
+%workgroup, exhibiting lock-step predicated execution that is
+%inherently unfair.
 
@@
 compute units to the scheduler by offering to sacrifice workgroups.
-The idea is that while having the scheduler be permitted to terminate
-the execution of workgroups in an arbitrary fashion would be drastic,
-the kernel may contain identifiable points of execution at which a
-workgroup could gracefully leave the computation.  We give examples of
-these for work stealing and graph traversal below.
+The idea is as follows: allowing the scheduler to arbitrarily and abruptly terminate execution
+of workgroups might be drastic, yet the kernel
+may contain specific program points at which a workgroup could
+\emph{gracefully} leave the computation.  We give examples of these
+for work stealing and graph traversal in \mysec\ref{sec:programmingguidelines}.
 
@@
 More formally, if $m < M-1$
-or $M=1$, i.e.\ the workgroup's id is not the largest among the active
-workgroups, or if only a single workgroup is executing the cooperative
-kernel, then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
-M-1$, there is a non-deterministic choice between (a) $\offerkill$
-executing as a no-op, or (b) execution of the workgroup ceasing and
-the number of active workgroups $M$ being atomically decremented by one.  Case
-(b) corresponds to the scheduler accepting the workgroup's offer to
-cease execution, allowing the scheduler to schedule some other task on
+or $M=1$ then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
+M-1$, the scheduler can choose to ignore the offer, so that $\offerkill$
+executes as a no-op, or accept the offer, so that execution of the workgroup ceases and
+the number of active workgroups $M$ is atomically decremented by one.  In the second case, the scheduler can assign some other task to
 the compute unit on which the workgroup was executing.

### 2016-11-14 write-ups/PLDI/paper.tex
Shortening and polishing.

@@
 
-The $\offerfork$ primitive (also a workgroup-level function) allows the cooperative kernel to indicate
-to the scheduler that it could benefit from, and is prepared to
-handle, additional workgroups joining the computation.  Recall that
-the cooperative kernel was launched with a limit of $N$ workgroups,
-indicating the number of workgroups that would ideally execute the
-kernel.  However, the number of active workgroups, $M$, may be smaller
+Recall that
+a desired limit of $N$ workgroups was specified when the cooperative kernel was launched, but that the number of active workgroups, $M$, may be smaller
 than $N$, either because (due to competing workloads) the scheduler
 did not provide $N$ workgroups initially, or because the kernel has
-given up some number of workgroups via $\offerkill$ calls.  Through
-$\offerfork$, the kernel and scheduler can work together to allow new
+given up some workgroups via $\offerkill$ calls.  Through the
+$\offerfork$ primitive (also a workgroup-level function), the kernel and scheduler can collaborate to allow new
 workgroups to join the computation at an appropriate point and with
@@
 Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  Then the following occurs: an
-integer $k$ in the range $[0, N-M]$ is chosen by the scheduler;
+integer $k \in [0, N-M]$ is chosen by the scheduler;
 $k$ new workgroups are spawned with consecutive ids in the range $[M,
@@
 initialising them is an undefined behaviour.  There are two exceptions
-to this:
-
-\begin{itemize}
-
-\item because the parameters to a cooperative kernel are immutable,
-  the new threads have access to these parameters as part of their
-  local state and can safely read from them;
-
-\item for each variable $v$ annotated with $\transmit$, every new
-  thread's copy of $v$ is initialised to the value that thread 0 in
-  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.
-
-\end{itemize}
-
-In effect, thread 0 of the forking workgroup transmits the relevant
+to this: (1) because the parameters to a cooperative kernel are
+immutable, the new threads have access to these parameters as part of
+their local state and can safely read from them; (2) for each variable
+$v$ annotated with $\transmit$, every new thread's copy of $v$ is
+initialised to the value that thread 0 in workgroup $m$ held for $v$
+at the point of the $\offerfork$ call.
+%
+In effect, thread 0 of the forking workgroup \emph{transmits} the relevant
 portion of its local state to the threads of the forked workgroups.
-We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
-having all new threads obtain this local state portion from thread 0,
-as well as our reasons for selecting which variables to transmit via
-annotations rather than transmitting the entire thread state.
+
+%We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
+%having all new threads obtain this local state portion from thread 0,
+%as well as our reasons for selecting which variables to transmit via
+%annotations rather than transmitting the entire thread state.
 

### 2016-11-14 write-ups/PLDI/paper.tex
Shortening and polishing.

@@
 
-Because workgroups of a cooperative kernel are fairly scheduled, a
-global barrier primitive can be provided.  We specify two such primitives: $\globalbarrier$
+Workgroups of a \emph{cooperative} kernel \emph{are} fairly scheduled, so a
+global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
 and $\resizingglobalbarrier$.
@@
 
-Semantically, when all threads have reached $\resizingglobalbarrier$,
-the number of active workgroups, $M$, is atomically set to a new value, $M'$ say, with $M' > 0$.
-If $M' = M$ then the active workgroups remain unchanged and proceed to
-execute after the barrier.  If $M' < M$, workgroups $[M', M-1]$ are
-killed.  If $M' > M$ then $M'-M$ new workgroups join the computation,
+When all threads have reached $\resizingglobalbarrier$,
+the number of active workgroups, $M$, is atomically set to a new value, $M'$ say, with $0 < M' \leq N$.
+If $M' = M$ then the active workgroups remain unchanged.  If $M' < M$, workgroups $[M', M-1]$ are
+killed.  If $M' > M$ then $M'-M$ new workgroups join the computation after the barrier,
 as if they were forked from workgroup 0.  In particular, the
 $\transmit$-annotated local state of thread 0 in workgroup 0 is
-transmitted to the threads of the new workgroups.  The change in
-number of active workgroups from $M$ to $M'$ is atomic.
+transmitted to the threads of the new workgroups.
 
@@
 resizing barrier, so that from a programmer's perspective, $M$ changes atomically.
-The middle $\globalbarrier$ is required to ensure that forking occurs before killing, so that while the barrier may kill a workgroup with id $i$, it should not also fork a new workgroup with id $i$.
-That is, workgroups $[0, \mathit{min}(M, M') - 1]$ should be left intact.
+The middle $\globalbarrier$ ensures that forking occurs before killing, so that while the barrier may kill a workgroup with id $i$, it should not also fork a new workgroup with id $i$.
+That is, workgroups $[0, \textrm{min}(M, M') - 1]$ should be left intact.
 
 Because $\resizingglobalbarrier$ can be implemented in the above
-manner, we do not regard it conceptually as a primitive of the
-cooperative kernels programming model.  However, we discuss in
+manner, we do not regard it \emph{conceptually} as a primitive of our
+programming model.  However, we discuss in
 \mysec\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be

### 2016-11-14 write-ups/PLDI/paper.tex
Shortening.
Minor edits on section 5

@@
 \mysec\ref{sec:semantics}, and account for slight changes in the OpenCL semantics induced by these new features.
-
-The programming model and associated language constructs were inspired
-by the requirements of the irregular GPU algorithms proposed in the
-literature.  For these algorithms, it was usually easy to identify
-where to place the constructs.
 
@@
 Cooperative scheduling primitives entails an overhead even if no
-killing, forking or resizing never happens. We measure this overhead
-via the execution time slowdown between the original kernel and the
+killing, forking or resizing happen. We measure this overhead via the
+execution time slowdown between the original kernel and the
 cooperative one, forcing the scheduler to never modify the number of
@@
 \mytab~\ref{tab:overhead} lists both geometric mean and maximum value
-of measured slowdowns. The overhead is below a 2 times slowdown even
-in the worst case, and quite light in general. We consider these
-results as encouraging, given that our SVM-based scheduler performance
-is likely a far upper bound of what a scheduler implemented directly
-in the runtime could offer.
+of measured overheads. The slowdown is below 2 even in the worst case,
+and closer to 1 in general. We consider these results as encouraging,
+especially since our prototype performance is likely a far upper bound
+of what a scheduler implemented directly in the runtime could achieve.
 

### 2016-11-14 write-ups/PLDI/paper.tex
Edit workstealing: remove nb_queues

@@
 not empty, since remaining tasks could eventually be stolen by other
-workgroups.  Care must be taken on computing queue identifiers: the
-original version in \myfig\ref{fig:workstealing} can rely on the fact
-that there are as many queues as workgroups, but here the number of
-workgroups may change during execution. Therefore, the number of
-queues must now be explicitely given as an argument
-(line~\ref{line:cwksteal:argnbqueue}) and used as an upper bound of
-queue indexes when scanning queues in $\mathsf{more\_work}$
-(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
-(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
-workgroup's queue index (line~\ref{line:cwksteal:getgroupid}) must now
-be moved inside the main loop since a forked workgroup that starts to
-execute after $\offerfork$ must re-compute its group id.
+workgroups. Since $\offerfork$ may be the entry point of a workgroup,
+the queue id must now be computed after it. In particular, the queue
+id cannot be transmitted since we want a newly spawned workgroup to
+read its own queue and not the one of the forking workgroup.
+
+% The queue id cannot be transmitted since it corresponds to the
+% workgroup id.
 
@@
 \begin{lstlisting}
-kernel work_stealing(global Task * queues,
-(*@\label{line:cwksteal:argnbqueue}@*)                     const int nb_queues) {
-(*@\label{line:cwksteal:morework}@*)  while (more_work(queues, nb_queues)) {
+kernel work_stealing(global Task * queues) {
+(*@\label{line:cwksteal:morework}@*)  while (more_work(queues)) {
 (*@\label{line:cwksteal:offerfork}@*)    offer_fork();
@@
 (*@\label{line:cwksteal:getgroupid}@*)    int queue_id = get_group_id();
-(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id, nb_queues);
+(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
     if (t) {

### 2016-11-14 write-ups/PLDI/paper.tex
Edit figure caption: remove nb_queues
Polishing and shortening.

--- 2016-11-14 write-ups/PLDI/paper.tex
@@
 
-Working with cooperative kernels requires some manual effort on the
-part of the programmer: they must place the $\transmit$ qualifier and
-use the new language primitives described in
-\mysec\ref{sec:semantics}, and account for slight changes in the OpenCL semantics induced by these new features.
-
-The programming model and associated language constructs were inspired
-by the requirements of the irregular GPU algorithms proposed in the
-literature.  For these algorithms, it was usually easy to identify
-where to place the constructs.
-
 We discuss the main semantic change induced by our programming
 model---that the number of workgroups executing a kernel varies
-dynamically---and illustrate our programming model and the impact of
-this semantic change by adapting the work stealing and graph traversal
-algorithms of \mysec\ref{sec:openclexamples} to use cooperative
-kernels.
+dynamically---and illustrate the new language constructs and the
+impact of this semantic change by adapting the work stealing and graph
+traversal algorithms of \mysec\ref{sec:openclexamples} to use
+cooperative kernels.
 
@@
 the value returned by $\getnumgroups()$ is not fixed during the
-lifetime of a cooperative kernel: it may change due to workgroups executing
-$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
-a function of $\getnumgroups()$, the value it returns is also subject to change.
-The programmer
-must therefore write their cooperative kernel in a manner that is
+lifetime of a cooperative kernel: it corresponds to the active group count $M$ and may change due to workgroups executing
+$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Being a function of $\getnumgroups()$, the value returned by $\getglobalsize()$
+is also subject to change.
+A cooperative kernel must therefore be written in a manner that is
 robust to changes in the values returned by these functions.
-In our experience this either involves avoiding the use of $\getnumgroups()$ at all, or writing a kernel where the only opportunity for the number of groups to change comes from $\resizingglobalbarrier()$, in which case the number of active groups does offer some stability, as discussed below.
+
+In general, their volatility means that use of these functions should be avoided.
+However, the situation is more stable if a cooperative kernel does not directly
+issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
+only calls to $\resizingglobalbarrier$ can cause the number of active
+workgroups to change.  At any point
+during execution, the threads of a kernel are executing between some
+pair of resizing barrier calls, which we call a \emph{resizing barrier
+  interval} (here we consider the kernel entry and exit points
+conceptually to be special cases of resizing barriers).  The number of
+workgroups executing the kernel is thus constant within each resizing
+barrier interval, i.e.\ the result returned by repeated calls to
+$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
+%
+This guarantee can be exploited by algorithms that perform strided
+data processing, which we illustrate below using the graph traversal
+example.
 
@@
 \myfiglong\ref{fig:workstealing-cooperative} shows a cooperative
-version of the work stealing example of \myfig\ref{fig:workstealing} that we
-discussed in \mysec\ref{sec:openclexamples}.
-
+version of the work stealing example of \myfig\ref{fig:workstealing} from \mysec\ref{sec:openclexamples}.
+%
 In this example there is no state to transmit since a computation is
@@
 task, we make the workgroup offer itself to be forked or killed
-(lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
+(lines~\ref{line:cwksteal:offerfork}-\ref{line:cwksteal:offerkill}). Note
 that a workgroup may be killed even if its associated task queue is
-not empty, since remaining tasks could eventually be stolen by other
+not empty, since the remaining tasks will be stolen by other
 workgroups.  Care must be taken on computing queue identifiers: the
@@
 
-\paragraph{Cooperative kernels that only use barriers}
-
-The situation is more stable if a cooperative kernel does not directly
-issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
-only calls to $\resizingglobalbarrier$ can cause the number of active
-workgroups to change.  At any point
-during execution, the threads of a kernel are executing between some
-pair of resizing barrier calls, which we call a \emph{resizing barrier
-  interval} (here we consider the kernel entry and exit points
-conceptually to be special cases of resizing barriers).  The number of
-workgroups executing the kernel is thus constant within each resizing
-barrier interval, i.e.\ the result returned by repeated calls to
-$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
-
-This guarantee can be exploited by algorithms that perform strided
-data processing.  We now illustrate this using the graph traversal motivating example.
-
 \paragraph{Adapting graph traversal} 
 \myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
-graph traversal kernel of \myfig\ref{fig:graphsearch}, introduced in
+graph traversal kernel of \myfig\ref{fig:graphsearch} from
 \mysec\ref{sec:openclexamples}.  On lines~\ref{line:cgraph:resizing1}
@@
 $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so
-that new workgroups know which of the node arrays are input and
+that new workgroups know which of the node arrays to use for input and
 output. Lastly, the static work distribution of the original kernel is
@@
 
+\paragraph{Patterns for irregular algorithms}
 In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
@@
 algorithms that are available as open source GPU kernels.  These all
-employ either work stealing or operate on graph data structures.
+employ either work stealing or operate on graph data structures, and placing our new constructs follows a common, easy-to-follow pattern in each case.
+
 Like the simplified example of \myfig\ref{fig:workstealing-cooperative}, the work stealing algorithms have a transactional flavour
@@
 Thus the point at which a workgroup is ready to process a new task is a natural point to place $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.
-%
+
 The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
 most level-by-level graph algorithms.  These algorithms tend to
-require more state to be annotated with $\transmit$, and in some cases
+require some state to be transmitted, and in some cases
 a global barrier is required in the midst of a transaction, in which
@@
 
-In summary, our experience so far indicates that cooperative kernels
-is a good fit for current irregular algorithms.  Of course this is, in
-part, because we designed the model with these algorithms in mind, but
-we argue that the fit is natural not just for the examples that we
-study, but more generally for at least these two patterns of irregular
-algorithm.
+Our practical experience suggests that cooperative kernels are a good
+fit for at least these two classes of irregular algorithms, which is
+of course partly because we were inspired by these algorithms in
+designing the model.
 
@@
 However, the aim of cooperative kernels is to find a balance that
-allows \emph{efficient execution} of long-running kernels
-(e.g.\ implementing irregular data-parallel algorithms), and
-\emph{responsive multi-tasking} so that the GPU can be shared between
-one or more cooperative kernels and a number of short, high-priority
-tasks with soft real-time constraints.
-
-To achieve this balance, both an implementation of the cooperative
+allows \emph{efficient} execution of algorithms that require fair scheduling, and
+\emph{responsive} multi-tasking, so that the GPU can be shared between
+cooperative kernels and other shorter tasks with soft real-time constraints.
+%
+To achieve this balance, an implementation of the cooperative
 kernels model, and the programmer of a cooperative kernel, must strive
@@
 opportunity to destroy a workgroup in order to schedule
-higher-priority tasks.  Because these higher-priority tasks, such as
-graphics rendering, may have soft real-time constraints, the scheduler
-relies on the cooperative kernel to execute $\offerkill$ sufficiently
-frequently that the scheduler can meet such real-time constraints.
-
-Using work stealing as an example: in the kernel of \myfig\ref{fig:workstealing-cooperative}, a
-workgroup offers itself to the scheduler after processing each task.
-If tasks turn out to be very short then it might turn out to be
-reasonable to invoke $\offerkill$ less frequently, after every $k$
-tasks for some appropriate $k$.  On the other hand, if individual
-tasks may be very time-consuming then it may be necessary to rewrite
-the algorithm so work on a larger number of shorter tasks, so as to
-achieve a higher rate of calls to $\offerfork$.
+higher-priority tasks.  The scheduler relies on the cooperative kernel
+to execute $\offerkill$ sufficiently frequently that soft real-time
+constraints of other workloads can be met.
+%
+Using our work stealing example: in
+\myfig\ref{fig:workstealing-cooperative} a workgroup offers itself to
+the scheduler after processing each task.  If tasks are sufficiently
+fast to process then the scheduler will have ample opportunities to
+de-schedule workgroups.  But if tasks can be very time-consuming to
+process then it might be necessary to rewrite the algorithm so that
+tasks are shorter and more numerous, to achieve a higher rate of calls
+to $\offerkill$.
 
@@
 workgroups requested upon kernel launch, in order to conduct its
-processing efficiently.  The scheduler should thus aim to provide with
-cooperative kernel with as many workgroups, up to this number, as
-other constraints allow.  As a result, the scheduler should generally
-accept an $\offerkill$ only if a compute unit is required for another
-task, and should respond positively to $\offerfork$ calls by
-scheduling new workgroups if compute units are available.  Our
+processing efficiently.  The scheduler should thus aim to provide the
+cooperative kernel $N$ workgroups if other constraints allow it,
+by accepting an $\offerkill$ only if a compute unit is required for another
+task, and responding positively to $\offerfork$ calls if compute units are available.  Our
 implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
@@
 
+\ADComment{This could be shorter.}
 However, these are not hard-and-fast rules: a scheduler might employ

### 2016-11-14 write-ups/PLDI/paper.tex

@@
 that a workgroup may be killed even if its associated task queue is
-not empty, since the remaining tasks will be stolen by other
-workgroups.  Care must be taken on computing queue identifiers: the
-original version in \myfig\ref{fig:workstealing} can rely on the fact
-that there are as many queues as workgroups, but here the number of
-workgroups may change during execution. Therefore, the number of
-queues must now be explicitely given as an argument
-(line~\ref{line:cwksteal:argnbqueue}) and used as an upper bound of
-queue indexes when scanning queues in $\mathsf{more\_work}$
-(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
-(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
-workgroup's queue index (line~\ref{line:cwksteal:getgroupid}) must now
-be moved inside the main loop since a forked workgroup that starts to
-execute after $\offerfork$ must re-compute its group id.
+not empty, since remaining tasks will be stolen by other
+workgroups. Since $\offerfork$ may be the entry point of a workgroup,
+the queue id must now be computed after it (line~\ref{line:cwksteal:getgroupid}). In particular, the queue
+id cannot be transmitted since we want a newly spawned workgroup to
+read its own queue and not the one of the forking workgroup.
+
+% The queue id cannot be transmitted since it corresponds to the
+% workgroup id.
+>>>>>>> 65caf661ed1c47ada958a674a0605c42b495e409
 
@@
 \caption{Cooperative version of the work stealing kernel of
-  \myfig~\ref{fig:workstealing}, using $\offerfork$ and $\offerkill$,
-  and passing the total number of queues
-  explicitely}\label{fig:workstealing-cooperative}
+  \myfig~\ref{fig:workstealing}, using $\offerfork$ and
+  $\offerkill$}\label{fig:workstealing-cooperative}
 \end{figure}

### 2016-11-14 write-ups/PLDI/paper.tex
Oops
Tweaks.

@@
 % workgroup id.
->>>>>>> 65caf661ed1c47ada958a674a0605c42b495e409
 
@@
 
+\ADComment{Move this to the appendix.}
+
 \paragraph{Killing order}
@@
 A goal of coopeative kernels is to allow traditional, non-blocking GPU
-tasks, such as graphics shaders (in programming models such as OpenGL,
-Vulkan and DirectX) and regular data-parallel computational kernels,
+tasks (e.g.\ graphics shaders in OpenGL, Vulkan and DirectX, and regular data-parallel computational kernels)
 to be co-scheduled with blocking irregular algorithms.  It is
@@
 barrier implementation (so that no resizing occurs in the latter
-case)---then cooperative kernel will behave just like its
+case)---then a cooperative kernel will behave just like its
 non-cooperative counterpart would behave.  As a result, a developer

### 2016-11-15 write-ups/PLDI/paper.tex
Add paragraph on porting irregular GPU applications

@@
 \begin{tabular}{ l c c c }
-app & resizing barriers & ckill - cfork & transmit vars\\
+App. & resizing barriers & kill - fork & transmit vars\\
 \hline
@@
 
+\HEComment{Add citations for pannotia, lonestar, workstealing}
+
+\mytablong~\ref{tab:applications} gives an overview of the 8 irregular
+algorithms that we ported to cooperative kernels. Among them, 6 are
+graph algorithms, based on Pannotia and Lonestar GPU application suite,
+using global barriers; we indicate how many barriers become resizing
+ones out of the total number of barriers, and how many variables need to
+be transmitted. In all cases all barriers were converted, except for bc
+which contains barriers deeper in the call stack, a case not supported
+by our prototype. Still, these barriers could in principle also be
+converted.  The remaining 2 algorithms are workstealing applications:
+each required the addition of $\offerfork$ and $\offerkill$ at the start
+of the main loop, and no variables needed to be transmitted. For all
+these examples the porting to cooperative kernels was pretty
+straightforward and did not require to bend the original algorithm.
+
 \ADComment{Describe the algorithms that we ported.}
-
-\HEComment{This is more or less already done in section
-\ref{sec:programmingguidelines} , maybe no need to rewrite here?}
 
@@
   barriers we changed to resizing, how many offer kill and offer fork
-  calls we included.}  \TSComment{check out
-  Table~\ref{tab:applications}. The barrier column shows how many of
-  the original barriers were ported to resizing barriers. In all cases
-  except bc we converted all of them. The only reason why bc is
-  different is because barriers occur deeper in the call stack and our
-  prototype cannot handle that. The algorithm itself is compatable
-  (i.e. changing these barriers is as easy as any other, there is
-  nothing special about them.)}
+  calls we included.}
+
+  \TSComment{check out Table~\ref{tab:applications}. The barrier column
+  shows how many of the original barriers were ported to resizing
+  barriers. In all cases except bc we converted all of them. The only
+  reason why bc is different is because barriers occur deeper in the
+  call stack and our prototype cannot handle that. The algorithm itself
+  is compatable (i.e. changing these barriers is as easy as any other,
+  there is nothing special about them.)}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Review along section 5
Add comments
Pass over 5.1-5.4 done, looking really good.

--- 2016-11-14 write-ups/PLDI/paper.tex
@@
 
+\HEComment{I consider that the reason why we need SVM fine grain and
+  atomics is discussed in the implementation section}
+
 \ADComment{Explain carefully that doing this properly would require
@@
 
-\subsection{Porting Irregular GPU Algorithms to Cooperative Kernels}\label{sec:portingalgorithms}
+We discuss our experience porting irregular algorithms to cooperative kernels (\mysec\ref{sec:portingalgorithms}) and discuss the GPUs on which we evaluate these applications (\mysec\ref{sec:gpus}).
+For these GPUs, we report on experiments to determine non-cooperative workloads that model the requirements of various graphics rendering tasks (\mysec\ref{sec:sizingnoncoop}).
+We then examine the overhead associated with moving to cooperative kernels when multitasking is \emph{not} required (\mysec\ref{sec:overhead}), as well as the responsiveness and throughput observed when a cooperative kernel is multi-tasked with non-cooperative workloads (\mysec\ref{sec:responsiveness}).
+
+\subsection{Porting Irregular GPU Algorithms}\label{sec:portingalgorithms}
 
@@
 \centering
-\begin{tabular}{ l c c c }
-app & resizing barriers & ckill - cfork & transmit vars\\
+\begin{tabular}{ l c c c c }
+App. & resizing barriers & kill & fork & transmit vars\\
 \hline
 \rowcolor{Gray1}
-color & 2 / 2 & 0 - 0 & 4\\
+color & 2 / 2 & 0 & 0 & 4\\
 \rowcolor{Gray1}
-mis & 3 / 3 & 0 - 0 & 0\\
+mis & 3 / 3 & 0 & 0 & 0\\
 \rowcolor{Gray1}
-bc & 3 / 6 & 0 - 0 & 3\\
+bc & 3 / 6 & 0 & 0 & 3\\
 \rowcolor{Gray1}
-p-sssp & 3 / 3 & 0 - 0 & 0 \\
+p-sssp & 3 / 3 & 0 & 0 & 0 \\
 \rowcolor{Gray2}
-bfs & 2 / 2 & 0 - 0  & 4 \\
+bfs & 2 / 2 & 0 & 0  & 4 \\
 \rowcolor{Gray2}
-l-sssp & 2 / 2 & 0 - 0  & 4 \\
+l-sssp & 2 / 2 & 0 & 0  & 4 \\
 \rowcolor{Gray3}
-octree & 0 / 0 & 1 - 1 & 0\\
+octree & 0 / 0 & 1 & 1 & 0\\
 \rowcolor{Gray3}
-game & 0 / 0 & 1 - 1 & 0\\
+game & 0 / 0 & 1 & 1 & 0\\
 \end{tabular} \\
 \vspace{.2cm}
-\crule[Gray1]{.2cm}{.2cm} - Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} - Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} - work stealing
+\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
 \caption{Blocking GPU applications investigated}
@@
 
-\ADComment{Describe the algorithms that we ported.}
-
-\HEComment{This is more or less already done in section
-\ref{sec:programmingguidelines} , maybe no need to rewrite here?}
-
-\ADComment{Give stats on how many variables had to be marked with
-  $\transmit$ (i.e., occurred in the restoration context), how many
-  barriers we changed to resizing, how many offer kill and offer fork
-  calls we included.}  \TSComment{check out
-  Table~\ref{tab:applications}. The barrier column shows how many of
-  the original barriers were ported to resizing barriers. In all cases
-  except bc we converted all of them. The only reason why bc is
-  different is because barriers occur deeper in the call stack and our
-  prototype cannot handle that. The algorithm itself is compatable
-  (i.e. changing these barriers is as easy as any other, there is
-  nothing special about them.)}
-
-\ADComment{Make sure we address the point that placing the primitives was pretty easy.}
-
-\subsection{The GPUs we Tested}
-
-\ADComment{Describe the platforms.  Explain why we could not try other
-platforms.}
-
-\HEComment{I consider that the reason why we need SVM fine grain and
-  atomics is discussed in the implementation section}
-
-Our implementation approach requires two optional features of OpenCL
-$2.0$, namely SVM fine-grained buffers and SVM atomics, which are
-currently not widely supported. \nvidia chips are ruled out since
-\nvidia does not support OpenCL $2.0$. Intel drivers meet our
-requirements on Windows, but not on Linux, although we tried the
-upstream driver that requires a patched Linux kernel. AMD support SVM
+\ADComment{It might be good to add a LOC column to \mytab\ref{tab:applications} to indicate how large the original versions of these applicaitons were.}
+
+\ADComment{I thought we had some examples where we actually wanted a non-resizing barrier; we say this earlier in the paper so we need to smooth.}
+
+\ADComment{Missing: a discussion of what we did not port.}
+
+\ADComment{We need to say something about the number of data sets
+  associated with each application, so that we know the total number
+  of application/data-set pairs.}
+
+\mytablong\ref{tab:applications} gives an overview of the 8 irregular
+algorithms that we ported to cooperative kernels. Among them, 6 are
+graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...} GPU application suites,
+using global barriers.  We indicate how many of the original number of barriers are changed to resizing barriers, and how many variables need to
+be transmitted.  In all cases all barriers were converted, except for bc
+which contains barriers deeper in the call stack, a case not supported
+by our prototype. Still, these barriers could in principle also be
+converted.  The remaining 2 algorithms are workstealing applications:
+each required the addition of $\offerfork$ and $\offerkill$ at the start
+of the main loop, and no variables needed to be transmitted. For all
+these examples the porting to cooperative kernels was similarly straightforward to the porting illustrated in \mysec\ref{sec:programmingguidelines} for our running examples, and did not require reshaping of the original algorithm.
+
+\subsection{The GPUs we Tested}\label{sec:gpus}
+
+Our prototype implementation (\mysec\ref{sec:implementation}) requires two optional features of OpenCL
+2.0: SVM fine-grained buffers and SVM atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at present, since
+these vendors do not yet support OpenCL $2.0$, equivalent functionality is not provided by \nvidia's CUDA. Intel drivers meet our
+requirements on Windows, but not on Linux (we confirmed this by trying the
+upstream driver that requires a patched Linux kernel). AMD support SVM
 atomics only their Kaveri family of chips and only on Linux, requiring
-to compile the latest driver. Among the GPUs available to us, only 4
-turned out to provide our requirements: Intel HD 520, HD 5500 and IRIS
-6100, and AMD Radeon R7.
-
-\HEComment{To decide: do we mention carrot AMD, in which terms?}
+to compile the latest driver.
+
+\ADComment{Check:} Among the GPUs available to us, four met our
+requirements: Intel HD 520, HD 5500 and Iris 6100, and AMD Radeon R7.
+However, AMD's Linux drivers for OpenCL suffer from a known defect
+whereby long-running kernels lead to defunct processes that the OS
+cannot kill.  We observed this on our AMD platform with the latest
+driver, and simultaneously obtained alarming results on this platform
+when using SVM atomics: modifying a kernel to include a simple
+host/device handshake through SVM slowed down kernel execution by an
+order of magnitude in some cases.  This behaviour, combined with the
+defunt processes issue, lead us to suspect that the support we require
+for cooperative kernels is not yet mature enough within AMD's drivers,
+which we plan to discuss with engineers at AMD.
+
+We thus present experimental results for the Intel GPUs, whose characteristics are summarised in \mytab\ref{tab:chipstested}.
+
+\begin{table}
+\small
+\centering
+\begin{tabular}{ l l l l }
+Chip & HD520 & HD5500 & Iris\\
+\hline
+\# compute units  & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
+Driver   & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
+Host CPU & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
+\end{tabular}
+\caption{The Intel GPU platforms used in our evaluation}
+\label{tab:chipstested}
+\end{table}
 
@@
 
-A cooperative kernel can yield workgroups to let a non-cooperative,
-transient one execute. We size non-cooperative kernels to represent
-approximatively the workload required by a system to maintain smooth
-screen animations.
-
-\paragraph{Experimental setup} We repetitively launch a kernel with a
+\ADComment{Refer back to nonfunctional requirements section, \mysec\ref{sec:nonfunctional}.}
+
+Enabling rendering of smooth graphics in parallel with execution of irregular algorithms is an important use case for our approach.
+Because our prototype implementation is based on a mega-kernel that takes over the entire GPU (see \mysec\ref{sec:implementation}), we cannot assess this use case directly.
+
+As a proxy for genuine graphics rendering, we devised the following
+procedure to determine non-cooperative OpenCL workloads that simulate
+the computational intensity of various graphics rendering workloads.
+%
+\ADComment{I think this is good but could be clearer and it would be helpful to discuss.}
+We repetitively launch a kernel with a
 controlled workload and detect from which workload size the screen
-animation starts to jitter. Given the kernel workload and its
+animation starts to jitter. \ADComment{Be clear that this is manual.} Given the kernel workload and its
 scheduling frequency, we can deduce what is the workload and frequency
@@
 \centering
-\begin{tabular}{ l c c }
+\begin{tabular}{ l r r }
 Workload & Duration (ms) & Period (ms)\\
@@
 \end{tabular}
-\caption{Caracteristics of non-cooperative workloads}
+\caption{\ADComment{Clearer caption?} Characteristics of non-cooperative kernels workload}
 \label{tab:noncooperativeworkload}
@@
 
-\subsection{The Overhead of Moving to Cooperative Kernels}
+\subsection{The Overhead of Moving to Cooperative Kernels}\label{sec:overhead}
 
@@
 }
-\caption{Overhead of cooperative kernels: execution time slowdown}
+\caption{Execution time slowdown associated with cooperative scheduling mode without multi-tasking}
 \label{tab:overhead}
@@
 \paragraph{Experimental setup}
-Cooperative scheduling primitives entails an overhead even if no
-killing, forking or resizing happen. We measure this overhead via the
-execution time slowdown between the original kernel and the
-cooperative one, forcing the scheduler to never modify the number of
-active workgroups. Since the cooperative kernel contains the
-non-cooperative kernel code anyway, its occupancy bound may be less
-than the original kernel; we make sure to use this occupancy bound as
-the number of workgroups for the original kernel in order to compare
-relevant execution times. \HEComment{Do we need to mention the
-  possible occupancy reduction as an overhead?}
+Invoking the cooperative scheduling primitives incurs some overhead
+even if no killing, forking or resizing actually occurs, because the cooperative kernel still needs to interact with the scheduler to determine this.
+We assess this overhead by measuring the
+ slowdown in execution time between the original and cooperative versions of a kernel, forcing the scheduler to never modify the number of
+active workgroups in the cooperative case.
+
+Recall that our mega kernel-based implementation merges the code for a
+cooperative kernel with that of the non-cooperative task to be
+co-scheduled.  This blow-up in code size can have a negative impact on
+compiliation, e.g.\ due to higher register pressure, reducing the
+occupancy for the merged kernel---the maximum number of workgroups
+that the GPU driver will schedule.  This is an artifact of our
+proof-of-concept implementation strategy, and would not be a problem
+if our cooperative scheduling approach was implemented inside the GPU
+driver.  We thus launch both the original and cooperative versions of
+a kernel with the reduced occupancy bound in order to meaningfully
+compare execution times.
+
+\ADComment{How many times did we run each benchmark?}
 
 \paragraph{Results}
-\mytab~\ref{tab:overhead} lists both geometric mean and maximum value
-of measured overheads. The slowdown is below 2 even in the worst case,
-and closer to 1 in general. We consider these results as encouraging,
-especially since our prototype performance is likely a far upper bound
-of what a scheduler implemented directly in the runtime could achieve.
+For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
+maximum slowdown across all applications and data sets.  For the
+maximum slowdowns, we indicate which application and data set was
+responsible.  The slowdown is below 2 even in the worst case, and
+closer to 1 on average. We consider these results as encouraging,
+especially since the performance of our prototype could clearly be
+improved upon if cooperative scheduling was implemented directly in
+the GPU driver.
+
+\ADComment{Can we remark a bit more on the outliers?}
 
@@
 
-voc: blocking and non-blocking ?
-
-durable / transient
-
-Cooperative scheduling enables to concurrently run non-persistent kernels alongside blocking ones.
-In terms of performance, we are interested to see whether the variations of workgroups for the blocking kernel.
-
-Its performance corresponds
-
-Performance of cooperative scheduling corresponds to execution time of cooperative kernels compared to the serial
-
-Cooperative scheduling can improve performances
+\HEComment{Below, just random notes so far}
+
+Cooperative scheduling enables to concurrently run non-persistent
+kernels alongside blocking ones. In terms of performance, we are
+interested to see whether the variations of workgroups for the blocking
+kernel.
+
+Performance of cooperative scheduling corresponds to execution time of
+cooperative kernels compared to the serial
 

### 2016-11-15 write-ups/PLDI/paper.tex
Nearly finished writing up formal semantics.

@@
 
+\clearpage
+
 \appendix
@@
 
-\ADComment{TODO: write it up properly.}
-
-\ADComment{TODO: remark that we do not model transmit faithfully.}
-
-Thread state: $(l, \mathit{ss})$, where $l \in L$ is the thread's
-local state, and $\mathit{ss}$ is a semi-colon-separated sequence of
-program statements that the thread remains to execute.
-
-Workgroup state: $((l_1, \mathit{ss}_1), \dots, (l_d,
-\mathit{ss}_d))$, where each $(l_i, \mathit{ss}_i)$ is a thread state.
-
-Kernel state: a pair $(\sigma, (w_1, w_2, \dots, w_M, \bot, \dots, \bot))$, where $\sigma$ is the shared state,
-each $w_i$ is a workgroup state, which are
-followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.
-
-We treat thread-level semantics abstractly by assuming a transition
-relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
-\mathit{ss}'))$ that is defined unless $\mathit{ss} =
-\mathsf{special}(); \mathit{tt}$, where $\mathsf{special}$ is one of
-$\offerkill$, $\offerfork$, $\globalbarrier$ or
-$\resizingglobalbarrier$.
-
-To abstractly accounts for the interaction between global barrier
-synchronization and the GPU memory model, we assume a function
-$\mathsf{sync}$ that is applied to a global state.  This returns a new
-global state in which the shared state and thread local states have
-been updated to reflect the fact that all stores to memory and loads
-from memory to thread-local storate have completed.
+\newcommand{\myss}{\mathit{ss}}
+\newcommand{\Stmts}{\mathsf{Stmts}}
+\newcommand{\threadstates}{\mathsf{ThreadStates}}
+\newcommand{\sharedstates}{\mathsf{SharedStates}}
+\newcommand{\sync}{\mathsf{sync}}
+
+In \mysec\ref{sec:semantics} we presented the semantics of cooperative
+kernels relatively informally, using English, to provide the intuition
+behind our programmming model.  We now back this up with a more formal
+presentation as an operational semantics for an abstract GPU
+programming model.
+
+\paragraph{States}
+
+Let $L$ be a set of \emph{local states} that abstractly captures the
+private memory associated with a thread executing a GPU kernel.  Let
+$\Stmts$ denote the set of all possible statements that a
+thread can execute.  We do not detail the structure of these
+statements, except that we assume sequential composition of statements
+is provided by the $\code{;}$ separator, and that the $\offerkill$,
+$\offerfork$, $\globalbarrier$ and $\resizingglobalbarrier$ primitives
+from our cooperative kernels programming model are valid statements.
+
+A \emph{thread state} is then a pair $(l, \myss)$, where $l \in L$ and
+$\myss \in \Stmts$.  The $l$ component captures the valuation of all the
+thread's private memory, and the $\myss$ component captures the
+remaining statements to be executed by the thread.  Let $\threadstates$ denote the set of all thread states.
+
+Assuming that $d > 0$ threads per workgroup were requested on kernel launch, a \emph{workgroup state}
+is a $d$-tuple $((l_0, \myss_0), \dots, (l_{d-1}, \myss_{d-1}))$, where each $(l_i, \myss_i)$ is the thread state for the $i$th thread in the workgroup $(0\leq i < d$).
+
+Assuming that $N > 0$ was specified as the maximum number of
+workgroups that should execute the cooperative kernel, a \emph{kernel
+  state} is then a pair
+%
+\[
+(\sigma, (w_0, \dots, w_{M-1}, \underbrace{\bot, \dots,
+\bot}_{N-M}))\]
+%
+where: $\sigma$ represents the shared state of the kernel; $M
+\leq N$ is the number of \emph{active} workgroups; $w_i$ is the
+workgroup state for active workgroup $i$ ($0 \leq i < M$); and $N-M$
+occurrences of $\bot$ indicate \emph{absent} workgroups.  Let $\sharedstates$ denote the set of all possible shared states.
+
+\paragraph{Thread-level transitions}
+
+We leave the semantics for thread-level transitions abstract, assuming
+a binary relation $\rightarrow_{tau}$ on $\sharedstates{} \times
+\threadstates{}$.  If $(\sigma, (l, \myss)) \rightarrow_{\tau}
+(\sigma', (l', \myss'))$, this indicates that if a thread is in local
+state $(l, \myss)$, the thread can transition to local state $(l',
+\myss')$, changing the shared state from $\sigma$ to $\sigma'$ in the
+process.
+
+All we require is that $(\sigma, (l, \myss)) \rightarrow_{\tau}
+(\sigma', (l', \myss'))$ if $\myss$ has the form $\mathsf{special}();
+\mathit{tt}$, where $\mathsf{special}$ is one of $\offerkill$,
+$\offerfork$, $\globalbarrier$ or $\resizingglobalbarrier$.  This is
+because we shall specifically define the meaning of the new primitives
+introduced by our programming model.
+
+\paragraph{Memory synchronisation}
+
+GPUs are known to have relaxed memory models~\cite{ }.  To abstractly
+account for this, we assume that the shared state component $\sigma$
+is not simplyl a mapping from locations to values, but instead
+captures all the intricacies of GPU memory that can lead to this
+relaxed behaviour.  We also assume a function $\sync$ which, given a
+kernel state $\kappa$, returns a set of kernel states.  The idea is
+that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
+can be reached from $\kappa$ by stalling until all stores to memory
+and loads from memory to thread-local storage have completed.  All we
+require is that $\sync$ does not modify the number of active
+workgroups nor the component of a thread state that determines which
+statements remain to be executed.
 
@@
 \inferrule{
-w_i(j) = (l, \mathit{ss})
+w_i(j) = (l, \myss)
 \\
-(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l', \mathit{ss}'))
+(\sigma, (l, \myss)) \rightarrow_{\tau} (\sigma', (l', \myss'))
 \\
-w_i' = w_i[j \mapsto (l', \mathit{ss}')]
+w_i' = w_i[j \mapsto (l', \myss')]
 }
@@
 \inferrule{
-\forall j \;.\; w_i(j) = (l_j, \offerkill();\mathit{ss})
+\forall j \;.\; w_i(j) = (l_j, \offerkill();\myss)
 \\
-w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+w_i' = ((l_0, \myss), \dots, (l_{d-1}, \myss))
 }
@@
 \inferrule{
-\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
+\forall j \;.\; w_{M-1}(j) = (l_j, \offerkill();\myss)
 \\
@@
 {
-(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
+(\sigma, (\dots, w_{M-2}, w_{M-1}, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-2}, \bot, \bot, \dots, \bot))
 }
@@
 \inferrule{
-\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
+\forall j \;.\; w_i(j) = (l_j, \offerfork();\myss)
 \\
-w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
+w_i' = ((l_{0}, \myss), \dots, (l_{d-1}, \myss))
 \\
@@
 \\
-\forall a \in [1, k] \;.\; w_{M+a} = ((l_1, \mathit{ss}), \dots, (l_1, \mathit{ss}))
+\forall a \in [0, k - 1] \;.\; w_{M+a} = ((l_{0}, \myss), \dots, (l_{0}, \myss))
 }
 {
-(\sigma, (\dots, w_i, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_M, w_{M+1}, \dots, w_{M+k}, \bot, \dots, \bot))
+(\sigma, (\dots, w_i, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_{M-1}, w_{M}, \dots, w_{M+k-1}, \bot, \dots, \bot))
 }
@@
 \inferrule{
-\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
+\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\myss)
 \\
-\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \mathit{ss})
+\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \myss)
+\\
+\kappa \in \sync((\sigma , (w_{0}', \dots, w_{M-1}', \bot, \dots, \bot))
 }
 {
-(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow \mathsf{sync}(\sigma , (w_1', \dots, w_M', \bot, \dots, \bot))
+(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow \kappa)
 }
@@
 \inferrule{
-\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
+\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\myss)
 \\
-\forall j\;.\;w_1'(j) = (l_{1,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\mathit{ss})
+\forall j\;.\;w_{0}'(j) = (l_{1,j}, \globalbarrier(); \offerfork(); \globalbarrier(); \globalbarrier();\myss)
 \\
-\forall i \neq 1 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\mathit{ss})
+\forall i \neq 1 \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \globalbarrier(); \globalbarrier(); \offerkill(); \globalbarrier();\myss)
 }
 {
-(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
+(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow (\sigma, (w_{0}', \dots, w_{M-1}', \bot, \dots, \bot))
 }
@@
 
-\caption{Abstract operational semantics for our cooperative kernels language extensions}
+\caption{Abstract operational semantics for our cooperative kernels language extensions}\label{fig:semanticrules}
 
@@
 
+\paragraph{Operational semantics rules}
+
+\myfiglong\ref{fig:semanticrules} presents the rules of our
+operational semantics, defining a relation $\rightarrow$ on kernel
+states.
+
+Rule $\textsc{Thread-Step}$ defines the semantics for thread making a
+single execution step, delegating to the abstract $\rightarrow_{\tau}$
+relation to determine how the thread's local state and the shared
+state component change.
+
+Rule $\textsc{Kill-No-Op}$ reflects the fact that when a workgroup
+reaches $\offerkill$, this can always be treated as a no-op.  Whether
+a scheduler implementation accepts $\offerkill$ calls or not depends
+on competing workloads and how the scheduler has been designed to meet
+the non-functinoal requirements discussed in
+\mysec\ref{sec:nonfunctional}, but in general the programmer should
+always be prepared for the possiblity that a workgroup survives after
+calling $\offerkill$.
+
+The case where a workgroup's offer to be killed is accepted by the
+scheduler is captured by rule $\textsc{Kill}$.  Because we have
+adopted a semantics where workgroup 0 is never killed and where only
+the workgroup with the highest id can be killed, the rule only fires
+if $M > 0$ and workgroup $M-1$ has reached $\offerkill$.  The rule has
+the effect of replacing the workgroup state for $w_{M-1}$
+with $\bot$.
+
+Recall that $\offerkill$ is a workgroup-level function: the same
+syntactic $\offerfork$ call must be reached by all threads in a
+workgroup.  This is captured in our rules by requiring in both
+$\textsc{Kill-No-Op}$ and $\textsc{Kill}$ that every thread is ready
+to execute $\offerkill$ followed by an identical statement $\myss$.
+Neither rule applies until all threads in a workgroup reach
+$\offerfork$, and the workgroup gets stuck if multiple threads in a
+workgroiup reach $\offerfork$ with different subsequent statements.
+
+\ADComment{TODO: $\textsc{Fork}$}
+
+\ADComment{TODO: $\textsc{Barrier}$}
+
+\ADComment{TODO: $\textsc{Resizing-Barrier}$}
+
+\paragraph{Simplifications}
+
+\ADComment{All state is transmitted.}  \ADComment{Local memory is not
+  explicitly modeled but can be captured inside shared state.}
+\ADComment{Semantics for intra-workgroup barriers are ignored.}
+
 \end{document}

### 2016-11-15 write-ups/PLDI/paper.tex
A bit more.

@@
 
-\ADComment{TODO: $\textsc{Fork}$}
+The $\textsc{Fork}$ rule similarly requires all threads in a workgroup
+to reach $\offerfork$ with identical following statements.  A
+nondeterministic number of new workgroups, $k$, is selected to be
+forked, where $k \in [0, N-M]$.  Importantly, $k=0$ is always a
+legitimate choice, in which case $\offerfork$ has no effect on the
+number of workgroups that are executing.  In the case where $k > 0$,
+$k$ new workgroup states are created, where each workgroup inherits
+the local state of thread 0 in $w_i$, the workgroup executing the fork
+call.  After $\offerfork$, all threads in all workgroups, including
+the new threads, proceed to execute the sequence of statements $\myss$
+that followed $\offerfork$.
+
+A simplification here is that we do not model transmission of
+particular annotated variables from thread 0 of the forking workgroup,
+instead specifying that the entire local state of the workgroup is
+transmitted.  Extending the semantics to transmit only annotated
+variables would be straightforward but verbose, requiring the local
+memory component of a thread state to be split into two: the part of
+the local state to be transmitted (modelling $\transmit$-annotated
+variables), and the rest of the local state.
 
@@
 
-\paragraph{Simplifications}
-
-\ADComment{All state is transmitted.}  \ADComment{Local memory is not
+ \ADComment{Local memory is not
   explicitly modeled but can be captured inside shared state.}

### 2016-11-15 write-ups/PLDI/paper.tex
Related work on preemption.

@@
 
+\paragraph{Irregular algorithms and persistent kernels}
+
+There has been a lot of work on accelerating blocking irregular
+algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
+programming style for long-running kernels~\cite{...}.  These
+approaches rely on the occupancy-bound execution model, and aim to
+take ownership of the GPU by scheduling as many workgroups as the GPU
+has compute units, relying on fair scheduling between these
+workgroups.  This means the GPU is unavailable for other tasks, and
+the method is not future-proof because (as discussed throughout the
+paper) the occupancy-bound execution model is not guaranteed by GPU
+programming models and is unlikely to be guaranteed in practice by
+future GPU platforms.
+
+As we have demonstrated for a representative set of irregular
+algorithms, our cooperative kernels model allows blocking algorithms
+to be upgraded to run under a fair scheduler in a manner that, with
+appropriate care, facilitates responsive multi-tasking.
+
+\paragraph{Preemptive multitasking on GPUs}
+
+Hardware support for preemptive multitasking for \nvidia GPUs has been
+proposed, enabling full preemption as well as \emph{SM-draining},
+whereby workgroups occupying a symmetric multiprocessor (SM; a compute
+unit using our terminology) are drained until the SM becomes free for
+other tasks.  A follow-up work adds the notion of SM \emph{flushing},
+whereby a workgroup running on an SM can be cancelled and re-scheduled
+from scratch if the workgroup has not yet committed any visible
+side-effects to shared memory.  Both approaches have been evaluated
+using simulators, over sets of regular GPU kernels.
+
+The advantages of these approaches are that they do not require manual
+programmer effort, and the capability to fully preempt a workgroup
+guards against a rogue long-running kernel making the GPU unavailable
+for other tasks.  The drawbacks are that (a) because preemption is
+enabled by default, \emph{every} kernel takes a potential performance
+hit, and (b) the SM draining optimisation does not help in the case of
+blocking algorithms (not considered in these works), since a blocked
+workgroup cannot be drained.
+
+Our solution does not impose a performance penalty on non-cooperative
+kernels, which run in an unmodified manner, and careful placement of
+our new language primitives that exploit the structure of a
+cooperative kernel may enable more efficient multi-tasking than
+kernel-oblivious preemption.
+
+\paragraph{GPU multitasking}
+%
+\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
+  keep a GPU busy by flooding it with work.  Would absolutely not
+  stand up to preemption.}
+
+\cite{DBLP:conf/usenix/KatoLRI11} \ADComment{Idea seems to be to
+  accept that GPUs do not offer epreemption, but find a reasonable way
+  to schedule tasks to maximise responsiveness or throughput.}
+
+\paragraph{Preemption support for GPUs}
+%
+\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
+  implement preemption, i.e.\ shows that it can be achieved.
+  Simulator only preemption.  Not sure whether their approach would
+  give fair scheduling.}
+%
+Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
+%
+\ADComment{Kepler preemption support?  Mention in related work?}
+
+\ADComment{Sreepathi mentioned the NCCL framework.}
+
+Dynamic parallelism could be extended to give barriers, but doesn't work with work stealing (less general).  It is in recent OpenCL.
+
+\cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
+  SMs, not to workgroups, so that tasks that would benefit from being
+  processed by the same SM can be scheduled to the same SM.  Has
+  little to do with preemption, but more to do with irregular
+  computation.}
+
+\cite{DBLP:conf/isca/TanasicGCRNV14} \ADComment{I have not read, but
+  it's about enabling preemptive multiprogramming, so we should read
+  it carefully.}
+
+\cite{DBLP:conf/ppopp/Muyan-OzcelikO16} \ADComment{I have not read; we
+  should understand this one carefully.}
+
+\paragraph{Cooperative multitasking}
+
 \HEComment{cooperative multitasking: OS: RISC OS, IBM CICS, Windows
@@
 
-\paragraph{Irregular algorithms and persistent kernels}
-%
-There has been a lot of work on accelerating blocking irregular
-algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
-programming style for long-running kernels.  These approaches rely on
-the occupancy-bound execution model, and aim to take ownership of the
-GPU by scheduling as many workgroups as the GPU has compute units,
-relying on fair scheduling between these workgroups.  These approaches
-render the GPU unavailable for other tasks, and are not future-proof
-because (as discussed throughout the paper) the occupancy-bound
-execution model is not guaranteed by GPU programming models and is
-unlikely to be guaranteed in practice by future GPU platforms.
-%
-\ADComment{Now say that we could update these approaches to use cooperation.}
-
-\paragraph{GPU multitasking}
-%
-\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
-  keep a GPU busy by flooding it with work.  Would absolutely not
-  stand up to preemption.}
-
-\cite{DBLP:conf/usenix/KatoLRI11} \ADComment{Idea seems to be to
-  accept that GPUs do not offer epreemption, but find a reasonable way
-  to schedule tasks to maximise responsiveness or throughput.}
-
-\paragraph{Preemption support for GPUs}
-%
-\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
-  implement preemption, i.e.\ shows that it can be achieved.
-  Simulator only preemption.  Not sure whether their approach would
-  give fair scheduling.}
-%
-Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
-%
-\ADComment{Kepler preemption support?  Mention in related work?}
-
-\ADComment{Sreepathi mentioned the NCCL framework.}
-
-Dynamic parallelism could be extended to give barriers, but doesn't work with work stealing (less general).  It is in recent OpenCL.
-
-\cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
-  SMs, not to workgroups, so that tasks that would benefit from being
-  processed by the same SM can be scheduled to the same SM.  Has
-  little to do with preemption, but more to do with irregular
-  computation.}
-
-\cite{DBLP:conf/isca/TanasicGCRNV14} \ADComment{I have not read, but
-  it's about enabling preemptive multiprogramming, so we should read
-  it carefully.}
-
-\cite{DBLP:conf/ppopp/Muyan-OzcelikO16} \ADComment{I have not read; we
-  should understand this one carefully.}
-
 \section{Conclusions and Future Work}\label{sec:conclusion}

### 2016-11-15 write-ups/PLDI/paper.tex
Related work progress.

@@
 First, the execution model does not enable
-multi-tasking, because a workgroup effectively \emph{owns} a compute
+multitasking, because a workgroup effectively \emph{owns} a compute
 unit until the workgroup has completed execution.  If a long-running
@@
 not commit to the occupancy-bound execution model that they currently
-implement, due to the need for multi-tasking and energy throttling,
+implement, due to the need for multitasking and energy throttling,
 yet (2) GPU vendors will not guarantee fair scheduling, due to the
@@
 
-\item Our experiments with a range of applications using three GPUs show that the model can enable efficient multi-tasking of cooperative and non-cooperative
+\item Our experiments with a range of applications using three GPUs show that the model can enable efficient multitasking of cooperative and non-cooperative
   tasks.
@@
 allows \emph{efficient} execution of algorithms that require fair scheduling, and
-\emph{responsive} multi-tasking, so that the GPU can be shared between
+\emph{responsive} multitasking, so that the GPU can be shared between
 cooperative kernels and other shorter tasks with soft real-time constraints.
@@
 }
-\caption{Execution time slowdown associated with cooperative scheduling mode without multi-tasking}
+\caption{Execution time slowdown associated with cooperative scheduling mode without multitasking}
 \label{tab:overhead}
@@
 to be upgraded to run under a fair scheduler in a manner that, with
-appropriate care, facilitates responsive multi-tasking.
+appropriate care, facilitates responsive multitasking.
 
@@
 our new language primitives that exploit the structure of a
-cooperative kernel may enable more efficient multi-tasking than
+cooperative kernel may enable more efficient multitasking than
 kernel-oblivious preemption.
 
-\paragraph{GPU multitasking}
-%
-\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
-  keep a GPU busy by flooding it with work.  Would absolutely not
-  stand up to preemption.}
-
-\cite{DBLP:conf/usenix/KatoLRI11} \ADComment{Idea seems to be to
-  accept that GPUs do not offer epreemption, but find a reasonable way
-  to schedule tasks to maximise responsiveness or throughput.}
-
-\paragraph{Preemption support for GPUs}
-%
-\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
-  implement preemption, i.e.\ shows that it can be achieved.
-  Simulator only preemption.  Not sure whether their approach would
-  give fair scheduling.}
-%
-Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
+\paragraph{Optimizing GPU scheduling}
+%
+A number of works have considered how best to schedulel dynamic
+workloads on GPUs.  The Whippletree approach employs a persistent
+megakernel to schedule multiple, dynamic tasks in a manner that aims
+to best utilise GPU
+resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}.  The TimeGraph
+approach similarly aims to optimise scheduling of competing workloads,
+in the context of OpenGL, using GPU-CPU synchronisation so that the
+CPU can monitor and control the work that is scheduled on the
+GPU~\cite{DBLP:conf/usenix/KatoLRI11}.  To best utilise the compute
+units of a GPU for a particular kernel, SM-centric program
+transformations have been proposed, which aim to direct work to
+specific compute units, rather than to specific workgroups, based on
+the observation that in practice multiple workgroups may share the
+same compute unit, so that said workgrops might be in a good position
+to exploit memory caches associated with that compute
+unit~\cite{DBLP:conf/ics/WuCLSV15}.  None of these approaches tackle
+the problem of \emph{fair} scheduling on GPUs that our work addresses, thus they do not aid in safe deployment of blocking irregular algorithms.
+
 %
@@
 
-\cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
+\cite{} \ADComment{Basic idea: map tasks to
   SMs, not to workgroups, so that tasks that would benefit from being

### 2016-11-15 write-ups/PLDI/paper.tex
Add chip info for HD 520
Ideas for Section 4.

@@
 
+We envisage cooperative kernels being implemented as part of the
+runtime for a GPU implementation of OpenCL, requiring changes to the
+device driver allowing it to intercept our new primitives and schedule
+and deschedule workgroups accordingly.
+
+To experiment with our ideas on current GPUs, we have instead
+developed a prototype implementation that mocks up the required
+runtime support via a \emph{megakernel}, and exploits the
+occupancy-bound execution model that these GPUs provide.  To be clear:
+an aim of cooperative kernels is to avoid depending on the
+occupancy-bound model; our protoytpe exploits this model simply to
+allow us to experiment with current GPUs whose proprietary drivers we
+cannot change.
+
+\subsection{The megakernel approach}
+
+The mock-up works as follows.  Instead of multitasking multiple
+separate kernels, we merge a set of kernels into a megakernel---a
+single, monolithic kernel---and encode the scheduling logic inside
+this megakernel.  The megakernel is launched with as many workgroups
+as can run concurrently according to the occupancy-bound.  One
+workgroup takes the role of scheduler, and the remaining workgroups
+act as a pool of workers.  A worker repeatedly queries the scheduler
+to be assigned a task.  A task corresponds to executing a cooperative
+or non-cooperative kernel.  In the case where a task corresponds to a
+non-cooperative kernel, the workgroup executes the kernel function
+uninterrupted and awaits further work.  In the case of a cooperative
+kernel, the workgroup either starts from the kernel entry point,
+corresponding to the workgroup being an \emph{initial} workgroup of
+the kernel, or immediately jumps to a designated point within the
+cooperative kernel, corresponding to the workgroup being forked by an
+already executing workgroup.  In the latter case, the new workgroup
+also receives a struct containing the values of all relevant
+$\transmit$-annotated variables.
+
+\paragraph{Simplifying assumptions}
+
+\ADComment{One cooperative, one non-cooperative.
+Justification: sufficient to evaluate the key use case of a long-running irregular algorithm co-existing with a series of short-running tasks.
+Simplifieid task of implementing prototype, which we do not want to over-enginneer since it is only a mock-up.}
+
+\ADComment{Restrictions on kernel structure: variables at top of kernel are implicitly $\transmit$-annotated, and $\offerkill$, $\offerfork$ and $\resizingglobalbarrier$ calls must appear only in kernel entry function.
+Avoiding $\transmit$ allowed us to reuse the clang OpenCL front-end, and restricitng starting points for fork to be in main kernel allowed a ``goto''-based approach to redirecting control flow; threading this through procedure calls would be possible but laborious.
+Restricting kill points enables kill to be implemented as ``return''; exiting from a procedure call stack would require stack unwinding support, akin to throwing an exception.
+Again make point that all these issues are only relevant due to our mock-up.}
+
+\paragraph{Merging kernels}
+
+\ADComment{Explain kernel merge tool briefly.  Explain library briefly.}
+
+\subsection{Scheduler design}\label{sec:resizingbarrier}
+
+\ADComment{Give most attention to interesting performance and safety issues, e.g.\ reverse ticket lock, efficient barrier, the query functions...}
+
+\ADComment{The host and device components of the scheduler, the use of
+  SVM, and the host API for scheduling both sorts of kernel.}
+
 \HEComment{I consider that the reason why we need SVM fine grain and
@@
 
-\ADComment{Explain carefully that doing this properly would require
-  driver support, etc., and then explain what we've actually done.}
-
 \ADComment{Give details of how host and device interact.}
 
-\ADComment{Explain kernel merge tool briefly.  Explain library briefly.}
-
-\ADComment{Give most attention to interesting performance and safety issues, e.g.\ reverse ticket lock, efficient barrier, the query functions...}
-
 \ADComment{Host-device interaction through SVM.  Make clear that this is probably the most aggressive use of OpenCL 2.0 ever.}
 
-\subsection{An Efficient Resizing Barrier}\label{sec:resizingbarrier}
+\paragraph{An efficient resizing barrier}
 
@@
 
+\ADComment{Rewrite:} There is scope for combining the approaches by using cooperative scheduling...
+
 \paragraph{Optimizing GPU scheduling}

### 2016-11-15 write-ups/PLDI/paper.tex
Add lines of code for applications
Add data sets

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \centering
-\begin{tabular}{ l c c c c }
-App. & resizing barriers & kill & fork & transmit vars\\
+\begin{tabular}{ l c c c c c c}
+App. & resizing barr. & kill & fork & transmit & LoC & data sets\\
 \hline
 \rowcolor{Gray1}
-color & 2 / 2 & 0 & 0 & 4\\
+color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
 \rowcolor{Gray1}
-mis & 3 / 3 & 0 & 0 & 0\\
+mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
 \rowcolor{Gray1}
-bc & 3 / 6 & 0 & 0 & 3\\
+bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
 \rowcolor{Gray1}
-p-sssp & 3 / 3 & 0 & 0 & 0 \\
+p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
 \rowcolor{Gray2}
-bfs & 2 / 2 & 0 & 0  & 4 \\
+bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
 \rowcolor{Gray2}
-l-sssp & 2 / 2 & 0 & 0  & 4 \\
+l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
 \rowcolor{Gray3}
-octree & 0 / 0 & 1 & 1 & 0\\
+octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
 \rowcolor{Gray3}
-game & 0 / 0 & 1 & 1 & 0\\
+game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
 \end{tabular} \\
@@
 
-\ADComment{It might be good to add a LOC column to \mytab\ref{tab:applications} to indicate how large the original versions of these applicaitons were.}
-
 \ADComment{I thought we had some examples where we actually wanted a non-resizing barrier; we say this earlier in the paper so we need to smooth.}
@@
 \ADComment{Missing: a discussion of what we did not port.}
-
-\ADComment{We need to say something about the number of data sets
-  associated with each application, so that we know the total number
-  of application/data-set pairs.}
 
@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...} GPU application suites,
-using global barriers.  We indicate how many of the original number of barriers are changed to resizing barriers, and how many variables need to
-be transmitted.  In all cases all barriers were converted, except for bc
-which contains barriers deeper in the call stack, a case not supported
-by our prototype. Still, these barriers could in principle also be
-converted.  The remaining 2 algorithms are workstealing applications:
-each required the addition of $\offerfork$ and $\offerkill$ at the start
-of the main loop, and no variables needed to be transmitted. For all
-these examples the porting to cooperative kernels was similarly straightforward to the porting illustrated in \mysec\ref{sec:programmingguidelines} for our running examples, and did not require reshaping of the original algorithm.
+graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...}
+GPU application suites, using global barriers.  We indicate how many of
+the original number of barriers are changed to resizing barriers, and
+how many variables need to be transmitted.  In all cases all barriers
+were converted, except for bc which contains barriers deeper in the call
+stack, a case not supported by our prototype. Still, these barriers
+could in principle also be converted. The remaining 2 algorithms are
+workstealing applications: each required the addition of $\offerfork$
+and $\offerkill$ at the start of the main loop, and no variables needed
+to be transmitted.  For all these examples the porting to cooperative
+kernels was similarly straightforward to the porting illustrated in
+\mysec\ref{sec:programmingguidelines} for our running examples, and did
+not require reshaping of the original algorithm. Most graph applications
+are executed on 2 different data sets, such that in total we have 13
+application/data set pairs.
 
@@
 \hline
-\# compute units  & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
-Driver   & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
-Host CPU & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
+\# comp. units  & 24 & \ADComment{...} & \ADComment{...} \\
+Driver   & 20.19.15.4501 & \ADComment{...} & \ADComment{...} \\
+Host CPU & i3-6100U & \ADComment{...} & \ADComment{...} \\
 \end{tabular}

### 2016-11-15 write-ups/PLDI/paper.tex
Edit discussion on data sets
Add number of runs

@@
 \mysec\ref{sec:programmingguidelines} for our running examples, and did
-not require reshaping of the original algorithm. Most graph applications
-are executed on 2 different data sets, such that in total we have 13
-application/data set pairs.
+not require reshaping of the original algorithm. Most graph come with 2
+different data sets, leading to 13 application/data set pairs in total.
 
@@
 \hline
-\# comp. units  & 24 & \ADComment{...} & \ADComment{...} \\
-Driver   & 20.19.15.4501 & \ADComment{...} & \ADComment{...} \\
-Host CPU & i3-6100U & \ADComment{...} & \ADComment{...} \\
+\# comp. units  & 24 & 24 & 47 \\
+Driver   & 20.19.15.4501 & \ADComment{...} & 20.19.15.4463 \\
+Host CPU & i3-6100U & \ADComment{...} & i3-5157U \\
 \end{tabular}
@@
 
-\ADComment{How many times did we run each benchmark?}
-
 \paragraph{Results}
 For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
-maximum slowdown across all applications and data sets.  For the
-maximum slowdowns, we indicate which application and data set was
-responsible.  The slowdown is below 2 even in the worst case, and
-closer to 1 on average. We consider these results as encouraging,
-especially since the performance of our prototype could clearly be
-improved upon if cooperative scheduling was implemented directly in
-the GPU driver.
+maximum slowdown across all applications and data sets, each benchmark
+begin run 10 times. For the maximum slowdowns, we indicate which
+application and data set was responsible.  The slowdown is below 2 even
+in the worst case, and closer to 1 on average. We consider these results
+as encouraging, especially since the performance of our prototype could
+clearly be improved upon if cooperative scheduling was implemented
+directly in the GPU driver.
 

### 2016-11-15 write-ups/PLDI/paper.tex

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...} GPU application suites,
-using global barriers.  We indicate how many of the original number of barriers are changed to resizing barriers, and how many variables need to
-be transmitted.  In all cases all barriers were converted, except for bc
-which contains barriers deeper in the call stack, a case not supported
-by our prototype. Still, these barriers could in principle also be
-converted.  The remaining 2 algorithms are workstealing applications:
-each required the addition of $\offerfork$ and $\offerkill$ at the start
-of the main loop, and no variables needed to be transmitted. For all
-these examples the porting to cooperative kernels was similarly straightforward to the porting illustrated in \mysec\ref{sec:programmingguidelines} for our running examples, and did not require reshaping of the original algorithm.
+graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...}
+GPU application suites, using global barriers.  We indicate how many of
+the original number of barriers are changed to resizing barriers, and
+how many variables need to be transmitted.  In all cases all barriers
+were converted, except for bc which contains barriers deeper in the call
+stack, a case not supported by our prototype. Still, these barriers
+could in principle also be converted. The remaining 2 algorithms are
+workstealing applications: each required the addition of $\offerfork$
+and $\offerkill$ at the start of the main loop, and no variables needed
+to be transmitted.  For all these examples the porting to cooperative
+kernels was similarly straightforward to the porting illustrated in
+\mysec\ref{sec:programmingguidelines} for our running examples, and did
+not require reshaping of the original algorithm. Most graph come with 2
+different data sets, leading to 13 application/data set pairs in total.
 
@@
 \hline
-\# compute units  & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
-Driver   & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
-Host CPU & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
+\# comp. units  & 24 & 24 & 47 \\
+Driver   & 20.19.15.4501 & \ADComment{...} & 20.19.15.4463 \\
+Host CPU & i3-6100U & \ADComment{...} & i3-5157U \\
 \end{tabular}

### 2016-11-15 write-ups/PLDI/paper.tex

@@
 
-\ADComment{Rewrite:} There is scope for combining the approaches by using cooperative scheduling...
+There is scope for combining the approaches: if a
+cooperative kernel is insufficiently cooperative, i.e.\ it does not
+invoke $\offerkill$ enough, then hardware support for preemption could
+be invoked to prevent the kernel from denying other tasks access to
+the GPU indefinitely.
 
@@
 %
-A number of works have considered how best to schedulel dynamic
-workloads on GPUs.  The Whippletree approach employs a persistent
+A number of works have considered how best to schedule dynamic
+workloads on GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
+Among these, the Whippletree approach employs a persistent
 megakernel to schedule multiple, dynamic tasks in a manner that aims
@@
 
+\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{...}.
+This \emph{dynamic parallelism} can be exploited to implement a GPU-based scheduler: an initial scheduler kernel uses the feature to spawn additional kernels as required~\cite{ }.
+
+ is launched
+
 %
@@
 \ADComment{Sreepathi mentioned the NCCL framework.}
+
+\cite{TPO10}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Some more related work.

@@
 \centering
-\begin{tabular}{ l c c c c c c}
-App. & resizing barr. & kill & fork & transmit & LoC & data sets\\
+\begin{tabular}{ l c c c c }
+App. & resizing barriers & kill & fork & transmit vars\\
 \hline
 \rowcolor{Gray1}
-color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
+color & 2 / 2 & 0 & 0 & 4\\
 \rowcolor{Gray1}
-mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
+mis & 3 / 3 & 0 & 0 & 0\\
 \rowcolor{Gray1}
-bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
+bc & 3 / 6 & 0 & 0 & 3\\
 \rowcolor{Gray1}
-p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
+p-sssp & 3 / 3 & 0 & 0 & 0 \\
 \rowcolor{Gray2}
-bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
+bfs & 2 / 2 & 0 & 0  & 4 \\
 \rowcolor{Gray2}
-l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
+l-sssp & 2 / 2 & 0 & 0  & 4 \\
 \rowcolor{Gray3}
-octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
+octree & 0 / 0 & 1 & 1 & 0\\
 \rowcolor{Gray3}
-game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
+game & 0 / 0 & 1 & 1 & 0\\
 \end{tabular} \\
@@
 
+\ADComment{It might be good to add a LOC column to \mytab\ref{tab:applications} to indicate how large the original versions of these applicaitons were.}
+
 \ADComment{I thought we had some examples where we actually wanted a non-resizing barrier; we say this earlier in the paper so we need to smooth.}
@@
 \ADComment{Missing: a discussion of what we did not port.}
+
+\ADComment{We need to say something about the number of data sets
+  associated with each application, so that we know the total number
+  of application/data-set pairs.}
 
@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...}
-GPU application suites, using global barriers.  We indicate how many of
-the original number of barriers are changed to resizing barriers, and
-how many variables need to be transmitted.  In all cases all barriers
-were converted, except for bc which contains barriers deeper in the call
-stack, a case not supported by our prototype. Still, these barriers
-could in principle also be converted. The remaining 2 algorithms are
-workstealing applications: each required the addition of $\offerfork$
-and $\offerkill$ at the start of the main loop, and no variables needed
-to be transmitted.  For all these examples the porting to cooperative
-kernels was similarly straightforward to the porting illustrated in
-\mysec\ref{sec:programmingguidelines} for our running examples, and did
-not require reshaping of the original algorithm. Most graph come with 2
-different data sets, leading to 13 application/data set pairs in total.
+graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...} GPU application suites,
+using global barriers.  We indicate how many of the original number of barriers are changed to resizing barriers, and how many variables need to
+be transmitted.  In all cases all barriers were converted, except for bc
+which contains barriers deeper in the call stack, a case not supported
+by our prototype. Still, these barriers could in principle also be
+converted.  The remaining 2 algorithms are workstealing applications:
+each required the addition of $\offerfork$ and $\offerkill$ at the start
+of the main loop, and no variables needed to be transmitted. For all
+these examples the porting to cooperative kernels was similarly straightforward to the porting illustrated in \mysec\ref{sec:programmingguidelines} for our running examples, and did not require reshaping of the original algorithm.
 
@@
 \hline
-\# comp. units  & 24 & 24 & 47 \\
-Driver   & 20.19.15.4501 & \ADComment{...} & 20.19.15.4463 \\
-Host CPU & i3-6100U & \ADComment{...} & i3-5157U \\
+\# compute units  & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
+Driver   & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
+Host CPU & \ADComment{...} & \ADComment{...} & \ADComment{...} \\
 \end{tabular}
@@
 
+\ADComment{How many times did we run each benchmark?}
+
 \paragraph{Results}
 For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
-maximum slowdown across all applications and data sets, each benchmark
-begin run 10 times. For the maximum slowdowns, we indicate which
-application and data set was responsible.  The slowdown is below 2 even
-in the worst case, and closer to 1 on average. We consider these results
-as encouraging, especially since the performance of our prototype could
-clearly be improved upon if cooperative scheduling was implemented
-directly in the GPU driver.
+maximum slowdown across all applications and data sets.  For the
+maximum slowdowns, we indicate which application and data set was
+responsible.  The slowdown is below 2 even in the worst case, and
+closer to 1 on average. We consider these results as encouraging,
+especially since the performance of our prototype could clearly be
+improved upon if cooperative scheduling was implemented directly in
+the GPU driver.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Remove figure of overhead

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
-most level-by-level graph algorithms.  These algorithms tend to
-require some state to be transmitted, and in some cases
-a global barrier is required in the midst of a transaction, in which
-case $\globalbarrier$, but not $\resizingglobalbarrier$, is suitable
-for achieving synchronization without disturbing the set of active
-workgroups.  But often it is the case that on completing a level of
+most level-by-level graph algorithms.
+It is typically the case that on completing a level of
 the graph algorithm, the next level could be processed by more or
 fewer workgroups, something which $\resizingglobalbarrier$
-facilitates.
+facilitates.  As in \myfig\ref{fig:cgraphsearch}, some level-specific state must be transmitted to new workgroups.
 
@@
 \centering
-\begin{tabular}{ l c c c c c c}
+\begin{tabular}{ l r r r r r r}
 App. & resizing barr. & kill & fork & transmit & LoC & data sets\\
@@
 
-\ADComment{I thought we had some examples where we actually wanted a non-resizing barrier; we say this earlier in the paper so we need to smooth.}
-
-\ADComment{Missing: a discussion of what we did not port.}
-
 \mytablong\ref{tab:applications} gives an overview of the 8 irregular
@@
 
+From the Lonestar suite we omit two applications that use global barriers:
+dmr has been shown to be unstable across GPU vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially short-running so that it is not
+interesting for our multitasking approach.
+%
+These exceptions aside, we have handled all the available OpenCL
+implemenatinos of irregular algorithms that we are aware of, plus two
+easy-to-adapt CUDA algorithms (the workstealing examples).
+
 \subsection{The GPUs we Tested}\label{sec:gpus}
 
-Our prototype implementation (\mysec\ref{sec:implementation}) requires two optional features of OpenCL
-2.0: SVM fine-grained buffers and SVM atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at present, since
-these vendors do not yet support OpenCL $2.0$, equivalent functionality is not provided by \nvidia's CUDA. Intel drivers meet our
-requirements on Windows, but not on Linux (we confirmed this by trying the
-upstream driver that requires a patched Linux kernel). AMD support SVM
-atomics only their Kaveri family of chips and only on Linux, requiring
-to compile the latest driver.
-
-\ADComment{Check:} Among the GPUs available to us, four met our
+Our prototype implementation (\mysec\ref{sec:implementation}) requires
+two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
+atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at
+present, since these vendors do not yet support OpenCL $2.0$,
+equivalent functionality is not provided by \nvidia's CUDA. Intel
+drivers meet our requirements on Windows, but not on Linux (we
+confirmed this by trying the upstream driver that requires a patched
+Linux kernel). AMD support SVM atomics only their Kaveri family of
+chips and only on Linux, requiring to compile the latest driver.
+
+Among the GPUs available to us, four met our
 requirements: Intel HD 520, HD 5500 and Iris 6100, and AMD Radeon R7.
@@
 whereby long-running kernels lead to defunct processes that the OS
-cannot kill.  We observed this on our AMD platform with the latest
+cannot kill.\footnote{See \url{http://developer.amd.com/download/AMD_APP_SDK_Release_Notes_Developer.pdf}}  We observed this on our AMD platform with the latest
 driver, and simultaneously obtained alarming results on this platform
@@
 %
-\ADComment{I think this is good but could be clearer and it would be helpful to discuss.}
-We repetitively launch a kernel with a
-controlled workload and detect from which workload size the screen
-animation starts to jitter. \ADComment{Be clear that this is manual.} Given the kernel workload and its
-scheduling frequency, we can deduce what is the workload and frequency
-needed by the system to maintain a smooth animation. We use this
-approach to measure the system workloads needed to maintain smooth
-animations of three types: (\emph{light}) emphasis of desktop icons
-under the mouse pointer, (\emph{medium}) window dragging over the
-desktop, and (\emph{heavy}) WebGL animations in a browser.
+We designed a synthetic kernel that occupies all workgrouops of a
+given GPU for a parameterised length of time $t$.  We wrote a host
+program to repeatedly launch this kernel in an infinite loop.  We then
+searched for a maximum value for the parameter $t$ that allowed the
+synthetic kernel to execute without having an observable impact on
+graphics rendering, where ``observable'' was decided manually by one
+of the paper authors.  Having found the appropriate value for $t$, we
+ran the host application for $X$ seconds, measuring the time $Y < X$
+dedicated to GPU execution during this period, as well as the number
+of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
+experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
+spent using the GPU to render the display between kernel calls and the
+rate at which the OS requires the GPU for display rendering,
+respectively.
+
+We used this approach to measure the GPU availability required for
+three types of rendering: \emph{light}, whereby desktop icons are
+smoothly emphasised under the mouse pointer; \emph{medium}, whereby
+window dragging over the desktop is smoothly animated; and
+\emph{heavy}, which requires smooth animation of a WebGL shader in a
+browser.  For \emph{heavy} we used the Paper
+Planes\footnote{\url{https://www.chromeexperiments.com/experiment/paper-planes}}
+and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
+shaders running in a recent version of Firefox.
 
@@
 \begin{tabular}{ l r r }
-Workload & Duration (ms) & Period (ms)\\
+Workload & Duration (ms) & Rate (ms)\\
 \hline
@@
 \end{tabular}
-\caption{\ADComment{Clearer caption?} Characteristics of non-cooperative kernels workload}
+\caption{Computed estimates of the GPU requirements for three different classes of graphics rendering}
 \label{tab:noncooperativeworkload}
@@
 
-\paragraph{Results}
-\mytab~\ref{tab:noncooperativeworkload} illustrates the caracteristics
-found for the three types of workloads. For the medium and heavy
-workloads, the 40ms period corresponds to an execution every $25^{th}$
-of seconds, which coincides with the human persistence of vision. The
-3ms execution duration of both light and medium configurations may
-indicate a lower bound of GPU execution time for system workloads.
+\mytablong\ref{tab:noncooperativeworkload} illustrates the data
+computed for the three types of workloads across our GPUs.  The
+\emph{Duration} and \emph{Rate} columns report values corresponding to
+$(X-Y)/n$ and $X/n$ in the above discussion, respectively.  For the
+medium and heavy workloads, the 40ms rate corresponds to graphics
+rendering occuring 25 times per second, which coincides with the human
+persistence of vision. The 3ms execution duration of both light and
+medium configurations indicates that, as expected, GPU computation is
+cheaper for basic display rendering compared with rendering that uses
+complex shaders.
+
+We use these results to guide our evaluation of the responsiveness of
+cooperative scheduling in \mysec\ref{sec:responsiveness}.
 
@@
 
-\begin{figure}
-\includegraphics[width=\columnwidth]{cooperative_overhead.pdf}
-\caption{Overhead: ratio of execution time}\label{fig:overhead}
-\end{figure}
-
 \paragraph{Experimental setup}
@@
 For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
-maximum slowdown across all applications and data sets, each benchmark
-begin run 10 times. For the maximum slowdowns, we indicate which
+maximum slowdown across all applications and data sets, with averages and maxima computed over 10 runs per benchmark. For the maximum slowdowns, we indicate which
 application and data set was responsible.  The slowdown is below 2 even

### 2016-11-15 write-ups/PLDI/paper.tex

@@
 
-\ADComment{One cooperative, one non-cooperative.
-Justification: sufficient to evaluate the key use case of a long-running irregular algorithm co-existing with a series of short-running tasks.
-Simplifieid task of implementing prototype, which we do not want to over-enginneer since it is only a mock-up.}
-
-\ADComment{Restrictions on kernel structure: variables at top of kernel are implicitly $\transmit$-annotated, and $\offerkill$, $\offerfork$ and $\resizingglobalbarrier$ calls must appear only in kernel entry function.
-Avoiding $\transmit$ allowed us to reuse the clang OpenCL front-end, and restricitng starting points for fork to be in main kernel allowed a ``goto''-based approach to redirecting control flow; threading this through procedure calls would be possible but laborious.
-Restricting kill points enables kill to be implemented as ``return''; exiting from a procedure call stack would require stack unwinding support, akin to throwing an exception.
-Again make point that all these issues are only relevant due to our mock-up.}
+\ADComment{One cooperative, one non-cooperative.  Justification:
+  sufficient to evaluate the key use case of a long-running irregular
+  algorithm co-existing with a series of short-running tasks.
+  Simplifieid task of implementing prototype, which we do not want to
+  over-enginneer since it is only a mock-up.}
+
+\ADComment{Restrictions on kernel structure: variables at top of
+  kernel are implicitly $\transmit$-annotated, and $\offerkill$,
+  $\offerfork$ and $\resizingglobalbarrier$ calls must appear only in
+  kernel entry function.  Avoiding $\transmit$ allowed us to reuse the
+  clang OpenCL front-end, and restricitng starting points for fork to
+  be in main kernel allowed a ``goto''-based approach to redirecting
+  control flow; threading this through procedure calls would be
+  possible but laborious.  Restricting kill points enables kill to be
+  implemented as ``return''; exiting from a procedure call stack would
+  require stack unwinding support, akin to throwing an exception.
+  Again make point that all these issues are only relevant due to our
+  mock-up.}
 
@@
 
-\ADComment{Explain kernel merge tool briefly.  Explain library briefly.}
+\ADComment{Explain kernel merge tool briefly.  Explain library
+  briefly.}
+
+\TSComment{we should mention that non-cooperative kernels must be
+written in a persistent thread style}
 
@@
 
-\ADComment{Give most attention to interesting performance and safety issues, e.g.\ reverse ticket lock, efficient barrier, the query functions...}
+To enable multitasking through cooperative kernels, the runtime (in
+our case, the mega-kernel), must (a) track the state of all
+workgroups, i.e.\ if a workgroup is waiting or computing a kernel (b)
+maintain consistent context states for each kernel, i.e.\ the number
+of active workgroups computing a cooperative kernel and (c) provide a
+safe way for these states to be modified (according to the semantics
+of $\offerfork$ and $\offerkill$. We discuss how that is achieved
+here.
+
+\paragraph{Kernel context}
+
+When a workgroup executes a kernel (cooperative or otherwise), it must
+be able to query a consistent context to obtain a workgroup id value
+(starting from zero up to the number of executing workgroups) and the
+number of workgroups executing the kernel. In single kernel execution,
+the OpenCL primitives {\tt get\_num\_groups()} and {\tt
+  get\_group\_id()} provide this functionality natively. For multiple
+kernel execution, we must emulate these primitives, per kernel,
+explicitly through \emph{kernel contexts}. This structure contains:
+(a) the number of active groups, (b) the total number of groups asked
+for, and (c) a mapping from the native workgroup id to a new workgroup
+id valid for a particular kernel. Our framework requires the native
+kernel context (to obtain a global base workgroup id) as well as an
+additional distinct kernel context for each kernel running
+concurrently. \TSComment{a driver implementation would likely generate
+these on the fly}.
+
+\paragraph{Scheduler context}
+
+In order to dynamically manage workgroups executing cooperative
+kernels, our framework must (a) track the state of each workgroup and
+(b) provide a channel of communication from the scheduler to
+workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
+we use a \emph{Scheduler context} structure which contains a mapping
+from primitive workgroup ids to their state i.e.\ waiting or computing
+a kernel as well as an integer value describing the number of
+workgroups in each state. The scheduler can then send cooperative
+kernels a \emph{resource message}, indicating if workgroups executing
+the cooperative kernel need to (a) exit at $\offerkill$ or (b)
+transmit their state to forking workgroups at $\offerfork$. Thus, the
+scheduler context needs a communication channel for each cooperative
+kernel. In our case, the communication channel is implemented using
+global atomic variables and OpenCL atomic operations to ensure
+consistency.
+
+\paragraph{Launching kernels}
+
+As in native OpenCL, the host maintains the ability to launch kernels.
+In our framework, the host has the additional ability to request that
+a kernel be cooperative \emph{and} to specify the amount of
+\emph{resource} each kernel requires. In our implementation, the only
+resource we have explicit control over is the number of concurrently
+executing workgroups; implementations closer to metal might also
+consider local memory and registers as resource.
+
+To launch a kernel, the host sends a data packet to the GPU, which
+consists of: a kernel to execute, the inputs to the kernel, the amount
+of resource to use for the kernel (which includes the number of
+workgroups to execute the kernel), and a flag indicating if the kernel
+is cooperative. In our implementation, this communication channel
+between host and device is achieved through fine-grained SVM atomics.
+The kernel specified must be one of the kernels merged inside the
+mega-kernel. If there are large constant inputs to the kernel, these
+can be copied before mega-kernel launch to decrease the amount of SVM
+traffic.
+
+The GPU scheduler then receives the data packet from the host
+describing a kernel $K$ to be executed and now must make decisions
+about how to schedule the $K$. Consider the case where $K$ requires
+$W$ workgroups. The scheduler can query the scheduler context and if
+there is at least $W$ available workgroups, then $K$ can be scheduled
+immediately. If there are not enough available workgroups (say there
+are $W_a$ workgroups available), but a cooperative kernel $K_c$ is
+executing. The scheduler can signal $K_c$ through $K_c$'s
+communication channel in the scheduler context communication
+requesting that $K_c$ provide $W - W_a$ workgroups (via $\offerkill$)
+as soon as possible. The scheduler then waits until $W$ workgroups
+become available.
+
+When $W$ workgroups become available, the scheduler initialises a
+kernel context $KC$ for $K$, setting the number of active workgroups
+to $W$ and mapping available workgroups (located in the scheduler
+context) to their new $KC$ workgroup ids. The scheduler then sends $W$
+workgroups from the available workgroups to execute kernel $K$ with
+kernel context $KC$.
+
+\paragraph{$\offerkill$ implementation}
+
+\ADComment{Give most attention to interesting performance and safety
+  issues, e.g.\ reverse ticket lock, efficient barrier, the query
+  functions...}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Changes with Hugues.
worked on section 4

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
-There is scope for combining the approaches: if a
-cooperative kernel is insufficiently cooperative, i.e.\ it does not
-invoke $\offerkill$ enough, then hardware support for preemption could
-be invoked to prevent the kernel from denying other tasks access to
-the GPU indefinitely.
+\ADComment{Rewrite:} There is scope for combining the approaches by using cooperative scheduling...
 
@@
 %
-A number of works have considered how best to schedule dynamic
-workloads on GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
-Among these, the Whippletree approach employs a persistent
+A number of works have considered how best to schedulel dynamic
+workloads on GPUs.  The Whippletree approach employs a persistent
 megakernel to schedule multiple, dynamic tasks in a manner that aims
@@
 
-\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{...}.
-This \emph{dynamic parallelism} can be exploited to implement a GPU-based scheduler: an initial scheduler kernel uses the feature to spawn additional kernels as required~\cite{ }.
-
- is launched
-
 %
@@
 \ADComment{Sreepathi mentioned the NCCL framework.}
-
-\cite{TPO10}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Minor

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 browser.  For \emph{heavy} we used the Paper
-Planes\footnote{\url{https://www.chromeexperiments.com/experiment/paper-planes}}
+Planes\footnote{\url{https://paperplanes.world}}
 and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
@@
 co-scheduled.  This blow-up in code size can have a negative impact on
-compiliation, e.g.\ due to higher register pressure, reducing the
+compilation, e.g.\ due to higher register pressure, reducing the
 occupancy for the merged kernel---the maximum number of workgroups

### 2016-11-15 write-ups/PLDI/paper.tex
Related work stuff.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
+\begin{figure}
+\includegraphics[width=\columnwidth]{cooperative_overhead.pdf}
+\caption{Overhead: ratio of execution time}\label{fig:overhead}
+\end{figure}
+
 \paragraph{Experimental setup}
@@
 There has been a lot of work on accelerating blocking irregular
-algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
+algorithms using GPUs~\cite{...,TPO10}, and on the \emph{persistent threads}
 programming style for long-running kernels~\cite{...}.  These
@@
 side-effects to shared memory.  Both approaches have been evaluated
-using simulators, over sets of regular GPU kernels.
+using simulators, over sets of regular GPU kernels.  State-of-the-art \nvidia GPUs targeting data centers (where multi-tasking is essential) do support
+preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.
 
@@
 
-\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{...}.
-This \emph{dynamic parallelism} can be exploited to implement a GPU-based scheduler: an initial scheduler kernel uses the feature to spawn additional kernels as required~\cite{ }.
-
- is launched
-
-%
-\ADComment{Kepler preemption support?  Mention in related work?}
-
-\ADComment{Sreepathi mentioned the NCCL framework.}
+\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{CUDASpec,opencl2Spec}.
+This \emph{dynamic parallelism} can be exploited to implement a
+GPU-based scheduler, by having an initial scheduler kernel repeatedly
+spawn further kernels as required, according to some scheduling
+policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, a kernel
+that uses dynamic parallelism is still subject to the scheduling
+limitations of current GPU programming models, and in particular the
+problem that workgroups are not fairly scheduled.  Hence this feature
+does not offer a solution to the problem of deploying blocking
+algorithms on GPUs in a safe manner that we tackle with this work.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Related work.

@@
 unit using our terminology) are drained until the SM becomes free for
-other tasks.  A follow-up work adds the notion of SM \emph{flushing},
+other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  A follow-up work adds the notion of SM \emph{flushing},
 whereby a workgroup running on an SM can be cancelled and re-scheduled
 from scratch if the workgroup has not yet committed any visible
-side-effects to shared memory.  Both approaches have been evaluated
+side-effects to shared memory~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
 using simulators, over sets of regular GPU kernels.  State-of-the-art \nvidia GPUs targeting data centers (where multi-tasking is essential) do support
@@
 
-\cite{TPO10}
-
-Dynamic parallelism could be extended to give barriers, but doesn't work with work stealing (less general).  It is in recent OpenCL.
-
-\cite{} \ADComment{Basic idea: map tasks to
-  SMs, not to workgroups, so that tasks that would benefit from being
-  processed by the same SM can be scheduled to the same SM.  Has
-  little to do with preemption, but more to do with irregular
-  computation.}
-
-\cite{DBLP:conf/isca/TanasicGCRNV14} \ADComment{I have not read, but
-  it's about enabling preemptive multiprogramming, so we should read
-  it carefully.}
-
-\cite{DBLP:conf/ppopp/Muyan-OzcelikO16} \ADComment{I have not read; we
-  should understand this one carefully.}
-
 \paragraph{Cooperative multitasking}

### 2016-11-15 write-ups/PLDI/paper.tex
Citations

@@
 data-parallel algorithms on GPUs, particularly algorithms that
-traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10}.
+traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
 
@@
 GPU schedulers are \emph{not} fair. the current GPU
-programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
+programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
 HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
@@
 units is via an \emph{occupancy-bound}
-execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}.
-\ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
+execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
 compute unit (it has become \emph{occupant}), the workgroup has
@@
 Experiments suggest that the occupancy-bound execution model is widely
-implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
+implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
 evidently a simple model for device vendors to implement efficiently.
@@
 no more workgroups than there are available compute
-units~\cite{...}. \ADComment{Tyler, please pick best cite for previous
-  claim.} This works because \emph{today}'s GPUs seem to provide the
+units~\cite{owens-persistent}. This works because \emph{today}'s GPUs seem to provide the
 occupancy-bound execution model.
@@
   the independent execution of workgroups since once in the work-pool,
-  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} CUDA similarly provides no guarantees~\cite{...}.
+  they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.
 
@@
 skeleton of a frontier-based graph traversal algorithm; such algorithms have
-been shown to be execute efficiently on GPUs~\cite{...}. \ADComment{Tyler, can you please work out which of the graph algorithm papers is the right one to cite here?}
+been shown to be execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}. 
 The kernel is
@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...}
+graph algorithms, based on Pannotia~\cite{Pannotia} and Lonestar~\cite{BNP12}
 GPU application suites, using global barriers.  We indicate how many of
@@
 There has been a lot of work on accelerating blocking irregular
-algorithms using GPUs~\cite{...,TPO10}, and on the \emph{persistent threads}
-programming style for long-running kernels~\cite{...}.  These
+algorithms using GPUs, and on the \emph{persistent threads}
+programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
+These
 approaches rely on the occupancy-bound execution model, and aim to
-take ownership of the GPU by scheduling as many workgroups as the GPU
-has compute units, relying on fair scheduling between these
+take ownership of the GPU by scheduling as many workgroups as the occupancy bound allows, relying on fair scheduling between these
 workgroups.  This means the GPU is unavailable for other tasks, and
@@
 kernels, which run in an unmodified manner, and careful placement of
-our new language primitives that exploit the structure of a
-cooperative kernel may enable more efficient multitasking than
+our new language primitives can exploit the structure of a
+cooperative kernel, enabling more efficient multitasking than
 kernel-oblivious preemption.
@@
 
-\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{CUDASpec,opencl2Spec}.
+\paragraph{Dynamic parallelism} Recent versions of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
 This \emph{dynamic parallelism} can be exploited to implement a
@@
 
-\HEComment{cooperative multitasking: OS: RISC OS, IBM CICS, Windows
-  before 95, old MacOS, maybe Oberon? TRON?. In GPU context: some
-  paper mention ``cooperative'' multitasking but they simply refer to
-  the fact that kernels should not compute for too long in order to
-  let other have access to it. Also, many prog languages implements
-  variations of coroutines, We can see blocking channels of CSP (or
-  Golang more recently) as yielding (sending a message is yielding
-  computation to the process which receives it)}
-
-Cooperative multitasking was offered in ancient operating systems
-(e.g. Windows before the 95 edition) and is still used nowadays in
-RISC OS for instance. Many programming languages offer cooperative
-multitasking via variations of coroutines, e.g. coroutine objects in
+Cooperative multitasking was offered in older operating systems
+(e.g. Windows before the 95 edition) and is still used by some
+operating systems, such as RISC
+OS.\footnote{\url{http://www.riscos.info/index.php/Preemptive_multitasking}}
+Many programming languages offer cooperative multitasking via
+variations of coroutines, e.g. coroutine objects in
 Python\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
-or Lua\footnote{\url{https://www.lua.org/pil/9.1.html}}, or Ruby's
-fiber \footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}. In
-the context of GPU resources, cooperative multitasking seems to refer
-to sizing kernels such that they do not compute for too long in order
-to yield the whole GPU regularly~\cite{adriaens2012case}. Our
-contribution present a much more flexible cooperative multitasking,
-where the long-running application is never totally stopped.
-
-\HEComment{ RISC OS see :
-\url{http://www.riscos.info/index.php/Preemptive_multitasking}
-}
-
-% ###
-% Cooperative multitasking for GPU-accelerated grid systems
-
-% http://onlinelibrary.wiley.com/doi/10.1002/cpe.1722/full
-
-% @article {CPE:CPE1722,
-% author = {Ino, Fumihiko and Ogita, Akihiro and Oita, Kentaro and Hagihara, Kenichi},
-% title = {Cooperative multitasking for GPU-accelerated grid systems},
-% journal = {Concurrency and Computation: Practice and Experience},
-% volume = {24},
-% number = {1},
-% publisher = {John Wiley & Sons, Ltd},
-% issn = {1532-0634},
-% url = {http://dx.doi.org/10.1002/cpe.1722},
-% doi = {10.1002/cpe.1722},
-% pages = {96--107},
-% keywords = {multitasking, GPU, CUDA, grid computing},
-% year = {2012},
-% }
-
-% Not super interesting: they just divide blocking computation in
-% smaller independent kernels, and fires them at regular intervals to
-% let the system be responsive.
+or Lua,\footnote{\url{https://www.lua.org/pil/9.1.html}} or Ruby's
+\emph{fiber}.\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}
+
+In the context of GPUs, cooperative multitasking has been used to
+refer to the standard setup today, whereby the onus is on a GPU kernel
+to complete execution within a reasonable time budget and return
+control to the CPU application~\cite{adriaens2012case,CPE:CPE1722}.
+In this setting, the programmer is responsible for ensuring cooperation between the host application and GPU kernel so that a kernel does not run for too long.
+%
+This is in contrast to our notion of cooperative kernels, which is designed specifically to support the needs of long-running kernels that one might wish to multitask with other workloads.
 
 \section{Conclusions and Future Work}\label{sec:conclusion}
+
+\ADComment{Do citation tidy-up.}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Beginning of subsection on performance and responsiveness

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
-\ADComment{Preamble}
-
-\HEComment{Below, just random notes so far}
-
-Cooperative scheduling enables to concurrently run non-persistent
-kernels alongside blocking ones. In terms of performance, we are
-interested to see whether the variations of workgroups for the blocking
-kernel.
-
-Performance of cooperative scheduling corresponds to execution time of
-cooperative kernels compared to the serial
-
-In order to simulate non-persistent workloads alongside kernels of
-blocking algorithms, we use a matrix multiplication kernel where
-matrix dimensions are tuned to trigger the relevant execution
-duration.
+Cooperative scheduling enables to run a non-cooperative kernel alongside
+a cooperative one, we focus here on the performance impact of this
+multitasking and on the scheduling responsiveness. The performance is
+measured with respect to the performance of each kernel run in
+sequence. Responsiveness, i.e., how fast can a non-cooperative kernel
+can be scheduled, depends not only on the scheduler implementation, but
+also on how often the cooperative kernel offer workgroups to be killed.
+
+% In order to simulate non-persistent workloads alongside kernels of
+% blocking algorithms, we use a matrix multiplication kernel where
+% matrix dimensions are tuned to trigger the relevant execution
+% duration.
+
+\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
+
+For a given cooperative kernel and its data set, we make it run while
+regularly asking to schedule a non-cooperative kernel that simulates
+graphics rendering, according to the classes listed in
+\mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
+multiplication kernel as the non-cooperative workload, with matrix
+dimensions tailored for each GPU to reach the appropriate execution
+duration when this non-cooperative kernel have access to all available
+workgroups. Besides, we have different run where the non-cooperative
+kernel is scheduled with different number of workgroups: either 1, a
+quarter, a half or all but one workgroups among the ones available for
+the cooperative kernel. Moreover, in graph algorithms we try both the
+regular and the query barrier.
+
+Our experiments span 13 pairs of cooperative kernels and data sets, 3
+classes for the non-cooperative kernel, 4 quantities of workgroups
+claimed for the non-cooperative kernel and 2 variations of resizing
+barriers for graph algorithms, leading to 288 configurations.
+
+A run is characterized by the cooperative kernel, its data set, the
+non-cooperative kernel, its size, the number of workgroups we ask for
+the non-cooperative kernel. This forms a configuration.
+
+For each GPU, we first
 
@@
 
-\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
+run cooperative kernels and fire non-cooperative kernels regularly,
+asking for different number of workgroups
+
+explain gather time, execution time. related to responsiveness and
+performance
+
+explain query barrier, or mention the results with / without query
+barriers.
+
+recall 10 runs for each configurations. How many configurations ?
+
+Mention that some configuration failed.
+
+We check correctness (completion) of both kernels. Some errors
+sometimes.
 
 \paragraph{Results}  \ADComment{Discuss the results.}
+explain horizontal lines
+
+missing results for all but one in bfs rmat22 since errors
+
+also, high gather time corresponds to applications and data sets where
+resize/kill is triggered less often.
+
+Each configuration was run 10 times on each GPU.
 
@@
 There has been a lot of work on accelerating blocking irregular
-algorithms using GPUs~\cite{...,TPO10}, and on the \emph{persistent threads}
+algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
 programming style for long-running kernels~\cite{...}.  These
@@
 side-effects to shared memory.  Both approaches have been evaluated
-using simulators, over sets of regular GPU kernels.  State-of-the-art \nvidia GPUs targeting data centers (where multi-tasking is essential) do support
-preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.
+using simulators, over sets of regular GPU kernels.
 
@@
 
-\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{CUDASpec,opencl2Spec}.
-This \emph{dynamic parallelism} can be exploited to implement a
-GPU-based scheduler, by having an initial scheduler kernel repeatedly
-spawn further kernels as required, according to some scheduling
-policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, a kernel
-that uses dynamic parallelism is still subject to the scheduling
-limitations of current GPU programming models, and in particular the
-problem that workgroups are not fairly scheduled.  Hence this feature
-does not offer a solution to the problem of deploying blocking
-algorithms on GPUs in a safe manner that we tackle with this work.
+\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{...}.
+This \emph{dynamic parallelism} can be exploited to implement a GPU-based scheduler: an initial scheduler kernel uses the feature to spawn additional kernels as required~\cite{ }.
+
+ is launched
+
+%
+\ADComment{Kepler preemption support?  Mention in related work?}
+
+\ADComment{Sreepathi mentioned the NCCL framework.}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Squeezing.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 do not mandate fair scheduling, and GPU schedulers are inherently
-unfair in practice.  This issue is currently circumvented by
-exploiting scheduling quirks particular to today's GPUs in a manner
-that does not allow the GPU to be shared with other workloads (such as
-graphics rendering tasks).  We propose \emph{cooperative kernels}, an
-extension to the traditional GPU programming model geared towards
-writing blocking algorithms.  Workgroups of a cooperative kernel are
-fairly scheduled, supporting blocking algorithms, and co-scheduling of
-other tasks is supported via a small set of language extensions that
-create a contract between the programmer and the scheduler.  We
-describe the semantics of the cooperative kernels programming model
-and discuss our portable prototype implementation.  We evaluate the
-technique by porting a set of irregular work stealing and graph
-algorithms to our programming model and assessing the performance and
-responsiveness of the approach across several GPUs.  Because our
-implementation exploits no vendor-specific hardware, driver or
-compiler support, our encouraging experimental results provide a
-lower-bound on the efficiency with which cooperative kernels can be
-implemented in practice.
+unfair in practice.  Current approaches avoid this issue by exploiting
+scheduling quirks particular to today's GPUs in a manner that does not
+allow the GPU to be shared with other workloads (such as graphics
+rendering tasks).  We propose \emph{cooperative kernels}, an extension
+to the traditional GPU programming model geared towards writing
+blocking algorithms.  Workgroups of a cooperative kernel \emph{are}
+fairly scheduled, and multitasking is supported via a small set of
+language extensions that the kernel and scheduler use to cooperate.
+We describe the semantics of the cooperative kernels programming model
+and our prototype implementation.  We evaluate the approach by porting
+a set of irregular work stealing and graph algorithms to our
+programming model and examining their performance under multitasking
+across three GPUs.  Our implementation exploits no
+vendor-specific hardware, driver or compiler support, so our encouraging
+results provide a lower-bound on the efficiency with
+which cooperative kernels can be implemented in practice.
 
@@
 of work to be processed is unknown ahead of time and may be determined
-by the computation itself.  The last decade has seen growing interest in accelerating irregular
-data-parallel algorithms on GPUs, particularly algorithms that
+by the computation itself.  There is growing interest in accelerating such algorithms on GPUs, particularly algorithms that
 traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
@@
 To operate correctly, a blocking concurrent algorithm requires
-\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
+\emph{fair} scheduling of workgroups.  Without fairness, a
 workgroup may fail to make progress, leading to starvation.  For
@@
 \paragraph{A degree of fairness: occupancy-bound execution}
-GPU schedulers are \emph{not} fair. the current GPU
+GPU schedulers are \emph{not} fair. The current GPU
 programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
@@
 executing a GPU kernel is mapped to a hardware \emph{compute
-  unit}.\footnote{In practice the situation can be more complex: if a
-  kernel is sufficiently simple and if workgroups are small enough,
-  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
+  unit}.\footnote{In practice, depending on the kernel, multiple workgroups might map to the same compute unit; we ignore this in our current discussion.}
 %
@@
 compute unit (it has become \emph{occupant}), the workgroup has
-exclusive access to the compute unit until it has finished execution,
-after which a subsequent workgroup can be bound to the compute unit.
+exclusive access to the compute unit until it has finished execution.
 Experiments suggest that the occupancy-bound execution model is widely
@@
 bound to separate compute units that operate in parallel.  Current GPU
-implementations of irregular data parallel algorithms assume the
+implementations of blocking algorithms assume the
 occupancy-bound execution model, which they exploit by launching 
@@
 multitasking, because a workgroup effectively \emph{owns} a compute
-unit until the workgroup has completed execution.  If a long-running
-kernel occupies all compute units, the GPU cannot be used for other
-tasks, such as graphics rendering, in the meantime.
+unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
+tasks, such as graphics rendering.
 %Precisely because
@@
 Second, \emph{energy throttling} is
-an important concern for battery-powered devices, such as smartphones
-and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
-power down one or more compute units if the battery level is low or if
+an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
+power down some compute units if the battery level is low or if
 the user puts the device into a low power state.  Implementing energy
@@
 (including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
-and various engineers from GPU vendors), is that (1) GPU vendors will
+and various engineers from GPU vendors), is that (1) for the above reasons, GPU vendors will
 not commit to the occupancy-bound execution model that they currently
-implement, due to the need for multitasking and energy throttling,
+implement,
 yet (2) GPU vendors will not guarantee fair scheduling, due to the
 runtime overhead associated with supporting full preemption of
-workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
-retain the essence of the occupancy-bound model, which is
-simple to implement efficiently, supporting preemption
+workgroups.  Vendors instead wish to
+retain the essence of the simple occupancy-bound model, supporting preemption
 only in key special cases.
@@
 %
-To summarise: blocking algorithms, such as most irregular algorithms,
-demand a degree of fair scheduling of workgroups, but for good reasons
-GPU vendors will not commit to the guarantees of even the
+To summarise: blocking algorithms
+demand fair scheduling of workgroups, but for good reasons
+GPU vendors will not commit even to the guarantees of the
 occupancy-bound execution model.
@@
 We propose \emph{cooperative kernels}, an extension to the GPU
-programming model that aims to resolve this impasse.  A GPU kernel
-that requires workgroups to be fairly scheduled is identified as a
-\emph{cooperative} kernel, and is written using two additional
+programming model that aims to resolve this impasse.  A kernel
+that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
 language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.
@@
 sacrifice itself to the scheduler.  This indicates that the workgroup
-would ideally continue with its execution but that, if required, the
-scheduler may destroy the workgroup for good: the cooperative kernel
-is prepared to deal with either scenario.
-
+would ideally continue executing, but that if required the
+scheduler may destroy the workgroup for good.  The cooperative kernel
+must be prepared to deal with either scenario.
+%
 At a point where the cooperative kernel could benefit from having
@@
 
+\vspace{-1mm}
 \begin{enumerate}
@@
 \item Cooperative kernels can be implemented over the occupancy-bound
-  execution model that current GPUs provide---we present a
-  proof-of-concept implementation that requires
-  no special-purpose hardware or driver support.
+  execution model that current GPUs provide: our proof-of-concept implementation requires
+  no additional hardware or driver support.  Thus the model should be eay for vendors to implement directly.
 
@@
 \end{enumerate}
+\vspace{-1mm}
 
@@
 
+\vspace{-1mm}
 \begin{itemize}
@@
 \end{itemize}
-
-We begin by providing an overview of the OpenCL programming model and
-two motivating examples of irregular algorithms (\mysec\ref{sec:background}).
-At the end of the paper we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).
+\vspace{-1mm}
+
+We begin by providing background on OpenCL via two motivating examples (\mysec\ref{sec:background}).
+At the end we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).
 
@@
 
-GPUs are known to have relaxed memory models~\cite{ }.  To abstractly
+GPUs are known to have relaxed memory models~\cite{ABDGKPSW-2015}.  To abstractly
 account for this, we assume that the shared state component $\sigma$

### 2016-11-15 write-ups/PLDI/paper.tex
Squeeze

@@
 $\getlocalid()$ and $\getgroupid()$ return a thread's local id within
-its workgroup and the id of the workgroup, respectively.\footnote{For simplicity we restrict attention to one-dimensional kernels, which captures all the
-GPU-accelerated irregular algorithms we are aware of, and use
+its workgroup and the id of the workgroup, respectively.\footnote{For simplicity we assume 1D kernels, which captures all the
+irregular algorithms we know of, using
 e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
@@
   they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.
-
+%
 HSA provides limited, one-way guarantees,
@@
   for values written by work-group B without deadlock provided ... A
-  comes after B in work-group flattened ID order}. However, this
-one-way communication is not sufficient to support blocking algorithms that use
+  comes after B in work-group flattened ID order}. This is not sufficient to support blocking algorithms that use
 mutexes and inter-workgroup barriers, both of which require \emph{symmetric} communication between
@@
 Work stealing enables dynamic balancing of tasks across several
-processing units, and is useful when the number of tasks to be
-processed in parallel is dynamic, because processing
-one task may create an arbitrary number of new tasks.  In the context
+processing units. It is useful when the number of tasks to be
+processed is dynamic, due to one task creating an arbitrary number of new tasks.  In the context
 of GPUs, each workgroup has an associated queue from which it
@@
 tasks. If a workgroup's queue is empty, the workgroup tries to \emph{steal} a task from an other
-queue. Workgroups keep on processing tasks until all the work is
-done. Previous work addresses the feasibility and efficiency of work
+queue. Previous work addresses the feasibility and efficiency of work
 stealing on GPUs~\cite{dlb-web,TPO10}.
@@
 \myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
-kernel that uses a work stealing queue. All threads start their
-execution at the function marked with the $\mathsf{kernel}$ keyword
-(line~\ref{line:wksteal:kernelfunc}). They receive a pointer to the task
+kernel that uses a work stealing queue. Each thread receives a pointer to the task
 queues, in global memory, initialized by the
 host to contain the initial tasks. Each thread identifies
-the workgroup it belongs to via the function $\getgroupid()$
-(line~\ref{line:wksteal:getgroupid}). Here the group id is stored into
-the variable $\mathsf{queue\_id}$ (which is, by default, in private
-memory) and used to access the relevant task queue. The
+the workgroup it belongs to via $\getgroupid()$
+(line~\ref{line:wksteal:getgroupid}), which is used as a queue id and used to access the relevant task queue. The
 $\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
@@
 all threads of its workgroup. If a task was obtained, then all threads
-of the workgroup participate in the processing of the task via 
-$\mathsf{process\_task}$ (line~\ref{line:wksteal:processtask}).
-Processing this task may lead to the creation of further tasks, which
+of the workgroup participate in the processing of the task (line~\ref{line:wksteal:processtask}).
+This may lead to the creation of further tasks, which
 the master thread will push to the workgroup's queue.
@@
 mutex per queue, implemented using atomic compare and swap operations
-on global memory.  As a result, the kernel presents two opportunities
+on global memory.  The kernel presents two opportunities
 for spin-waiting: spinning to obtain a mutex, and spinning in the main
-kernel loop to obtain a task to process.  The algorithm thus depends
-on a fair scheduler, which, as discussed in \mysec\ref{sec:intro}, is
-not guaranteed by current GPU programming models.
+kernel loop to obtain a task.  The algorithm thus depends
+on a fair scheduler, which is not guaranteed by current GPU programming models.
 
@@
 arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
-starting nodes to process. Threads keep three local variables: $\keyword{level}$, which indicates the frontier level currently being
-processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$, which point to
-distinct node arrays. The nodes to be
-processed for the current frontier are contained in $\keyword{in\_nodes}$, while $\keyword{out\_nodes}$ contains
-the nodes to be processed during the next frontier.
+starting nodes to process. Private variable $\keyword{level}$ indicates the frontier level currently being
+processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ point to
+distinct node arrays recording the nodes to be processed during the current and next frontier, respectively.
 
@@
 At this point, the GPU threads must wait for all other threads to
-finishing processing the frontier before they can continue. To achieve
+finishing processing the frontier. To achieve
 this, we use a global barrier construct
@@
 output node array is reset (line~\ref{line:graph:reset}) and the level
-is incremented. Again, threads must wait until the output node is
-reset at a global barrier (line~\ref{line:graph:gb2}). At this point
-threads can continue to the next frontier.
+is incremented. The threads use another global barrier to wait until the output node is
+reset (line~\ref{line:graph:gb2}), after which the threads can continue to the next frontier.
 
@@
 scheduling between workgroups for the barrier to operate correctly.
-Again, fairness is not guaranteed by current GPU programming models.
-
-The mutexes and barriers required by these two examples appear to run
+
+The mutexes and barriers used by these two examples appear to run
 reliably on current GPUs for kernels that are executed with no more
-workgroups than there are compute units; this is due to the fairness
+workgroups than there are compute units.  This is due to the fairness
 of the occupancy-bound execution model that current GPUs have been

### 2016-11-15 write-ups/PLDI/paper.tex
Squeeze

@@
 
-We describe our cooperative kernels programming model as an extension
+We present our cooperative kernels programming model as an extension
 to OpenCL.  However, the cooperative kernels concept is more general,
 and could be applied to extend other GPU programming models, such as
-CUDA and HSA.
-
-We describe the semantics of cooperative kernels
-(\mysec\ref{sec:semantics}) and summarise the important
-differences between programming with regular vs.\ cooperative kernels
-by showing how the motivating examples of \mysec\ref{sec:openclexamples} can be adapted to use cooperative kernels
-(\mysec\ref{sec:programmingguidelines}).
-We then focus on the non-functional
-properties that a developer of a cooperative kernel, and the
-implementer of a scheduler for cooperative kernels, should strive for
-(\mysec\ref{sec:nonfunctional}).  The semantics of cooperative
-kernels has been guided by the practical applications we have studied
-(described in \mysec\ref{sec:portingalgorithms}), and we discuss several cases where we
-might have taken different and also reasonable semantic decisions (\mysec\ref{sec:semanticalternatives}).
-We conclude the section by discussing the backwards-compatible nature
-of cooperative kernels with respect to blocking algorithms designed to
-run on current GPU platforms
+CUDA and HSA.  We describe the semantics of the model
+(\mysec\ref{sec:semantics}), use our motivating examples to discuss
+programmability (\mysec\ref{sec:programmingguidelines}), outline
+important nonfunctional properties that the model requires to work
+successfully (\mysec\ref{sec:nonfunctional}) and explain that the
+model is backwards-compatible
 (\mysec\ref{sec:backwardscompatibility}).
+%
+In Appendices~\ref{appendix:semantics} and~\ref{appendix:semanticalternatives} (supplementary material)
+we present a formal semantics for cooperative kernels atop an abstract GPU programming model,
+and discuss several cases where we
+might have taken different and also reasonable semantic decisions.
 
@@
 With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.
-
-\subsection{Alternative Semantic Choices}\label{sec:semanticalternatives}
-
-\ADComment{Move this to the appendix.}
-
-\paragraph{Killing order}
-
-We opted for a semantics whereby only the active workgroup with the
-highest id can be killed.  This has an appealing property: it means
-that the ids of active workgroups are contiguous, which is important
-for processing of contiguous data.  The cooperative graph traversal
-algorithm of \myfig\ref{fig:cgraphsearch} illustrates this: the algorithm is prepared
-for $\getglobalsize()$ to change after each resizing barrier call, but
-depends on the fact that $\getglobalid()$ returns a contiguous range
-of thread ids.
-
-A disadvantage of this decision is that it may provide sub-optimal
-responsiveness from the point of view of the scheduler.  Suppose the
-scheduler requires an additional compute unit, but the active thread
-with the larget id is processing some computationally intensive work
-and will take a while to reach $\offerkill$.  Our semantics means that
-the scheuduler cannot take advantage of the fact that another active
-workgroup may invoke $\offerkill$ sooner.
-
-Cooperative kernels that do not require contiguous thread ids might me
-more suited to a semantics in which workgroups can be killed in any
-order, but where workgroup ids (and thus thread global ids) are not
-guaranteed to be contiguous.
-
-\paragraph{Keeping one workgroup alive}
-
-Our semantics dictate that the workgroup with id 0 will not be killed
-if it invokes $\offerkill$.  This avoids the possibility of the
-cooperative kernel terminating early due to the programmer
-inadvertantly allowing all workgroups to be killed, and the decision
-to keep workgroup 0 alive fits well with our choice to kill workgroups
-in descending order of id.
-
-However, there might be a use case for a cooperative kernel reaching a
-point where it would be acceptable for the kernel to exit, although
-desirable for some remaining computation to be performed if competing
-workloads allow it.
-In this case, a semantics where all workgroups can be killed via $\offerkill$ would be appropriate, and the programmer would need to guard each $\offerkill$ with an id check in cases where killing all workgroups would be unacceptable.
-For example:
-%
-\lstset{basicstyle=\tt,numbers=none}
-\begin{lstlisting}
-  if($\getgroupid{0}$ != 0) $\offerkill$();
-\end{lstlisting}
-\lstset{basicstyle=\scriptsize\tt,numbers=left}
-%
-would ensure that at least workgroup 0 is kept alive.
-
-\paragraph{Transmission of partial state from a single thread}
-
-Recall from the semantics of $\offerfork$ that newly forked workgroups
-inherit the variable valuation associated with thread 0 of the forking
-workgroup, but only for $\transmit$-annotated variables.  Alternative
-choices here would be to have forked workgroups inherit values for
-\emph{all} variables from the forking workgroup, and to have thread
-$i$ in the forking workgroup provide the valuation for thread $i$ in
-each spawned workgroup, rather than having thread 0 transmit the
-valuation to all new threads.
-
-We opted for transmitting only selected varibles based on the
-observation that many of a thread's private variables are dead at the
-point of issuing $\offerfork$ or $\resizingglobalbarrier$, thus it
-would be wasteful to transmit them.  A live variable analysis could
-instead be employed to over-approximate the variables that might be
-accessed by newly arriving workgroups, so that these are automatically
-transmitted.
-
-In all cases, we found that a variable that needed to be transmitted
-had the property of being uniform across the workgroup.  That is,
-despite each thread having its own copy of the variable, each thread
-is in agreement on the variable's value.  As an example, the
-$\keyword{level}$, $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$
-variables used in \myfig\ref{fig:cgraphsearch} are all stored in thread-private
-memory, but all threads in a workgroup agree on the values of these
-variables at each $\resizingglobalbarrier$ call.  As a result,
-transmitting the thread 0's valuation of the annotated variables is
-equivalent to (and more efficient than) transmitting values on a
-thread-by-thread basis.  We have not yet encountered a real-world
-example where our current semantics would not suffice.
 
@@
 
+\section{Alternative Semantic Choices}\label{appendix:semanticalternatives}
+
+The semantics of cooperative kernels has been guided by the practical
+applications we have studied (described in
+\mysec\ref{sec:portingalgorithms}).  We now discuss several cases
+where we might have taken different and also reasonable semantic
+decisions.
+
+\paragraph{Killing order}
+
+We opted for a semantics whereby only the active workgroup with the
+highest id can be killed.  This has an appealing property: it means
+that the ids of active workgroups are contiguous, which is important
+for processing of contiguous data.  The cooperative graph traversal
+algorithm of \myfig\ref{fig:cgraphsearch} illustrates this: the
+algorithm is prepared for $\getglobalsize()$ to change after each
+resizing barrier call, but depends on the fact that $\getglobalid()$
+returns a contiguous range of thread ids.
+
+A disadvantage of this decision is that it may provide sub-optimal
+responsiveness from the point of view of the scheduler.  Suppose the
+scheduler requires an additional compute unit, but the active thread
+with the larget id is processing some computationally intensive work
+and will take a while to reach $\offerkill$.  Our semantics means that
+the scheuduler cannot take advantage of the fact that another active
+workgroup may invoke $\offerkill$ sooner.
+
+Cooperative kernels that do not require contiguous thread ids might me
+more suited to a semantics in which workgroups can be killed in any
+order, but where workgroup ids (and thus thread global ids) are not
+guaranteed to be contiguous.
+
+\paragraph{Keeping one workgroup alive}
+
+Our semantics dictate that the workgroup with id 0 will not be killed
+if it invokes $\offerkill$.  This avoids the possibility of the
+cooperative kernel terminating early due to the programmer
+inadvertantly allowing all workgroups to be killed, and the decision
+to keep workgroup 0 alive fits well with our choice to kill workgroups
+in descending order of id.
+
+However, there might be a use case for a cooperative kernel reaching a
+point where it would be acceptable for the kernel to exit, although
+desirable for some remaining computation to be performed if competing
+workloads allow it.
+In this case, a semantics where all workgroups can be killed via $\offerkill$ would be appropriate, and the programmer would need to guard each $\offerkill$ with an id check in cases where killing all workgroups would be unacceptable.
+For example:
+%
+\lstset{basicstyle=\tt,numbers=none}
+\begin{lstlisting}
+  if($\getgroupid{0}$ != 0) $\offerkill$();
+\end{lstlisting}
+\lstset{basicstyle=\scriptsize\tt,numbers=left}
+%
+would ensure that at least workgroup 0 is kept alive.
+
+\paragraph{Transmission of partial state from a single thread}
+
+Recall from the semantics of $\offerfork$ that newly forked workgroups
+inherit the variable valuation associated with thread 0 of the forking
+workgroup, but only for $\transmit$-annotated variables.  Alternative
+choices here would be to have forked workgroups inherit values for
+\emph{all} variables from the forking workgroup, and to have thread
+$i$ in the forking workgroup provide the valuation for thread $i$ in
+each spawned workgroup, rather than having thread 0 transmit the
+valuation to all new threads.
+
+We opted for transmitting only selected varibles based on the
+observation that many of a thread's private variables are dead at the
+point of issuing $\offerfork$ or $\resizingglobalbarrier$, thus it
+would be wasteful to transmit them.  A live variable analysis could
+instead be employed to over-approximate the variables that might be
+accessed by newly arriving workgroups, so that these are automatically
+transmitted.
+
+In all cases, we found that a variable that needed to be transmitted
+had the property of being uniform across the workgroup.  That is,
+despite each thread having its own copy of the variable, each thread
+is in agreement on the variable's value.  As an example, the
+$\keyword{level}$, $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$
+variables used in \myfig\ref{fig:cgraphsearch} are all stored in thread-private
+memory, but all threads in a workgroup agree on the values of these
+variables at each $\resizingglobalbarrier$ call.  As a result,
+transmitting the thread 0's valuation of the annotated variables is
+equivalent to (and more efficient than) transmitting values on a
+thread-by-thread basis.  We have not yet encountered a real-world
+example where our current semantics would not suffice.
+
 \end{document}

### 2016-11-15 write-ups/PLDI/paper.tex
section 5.5 getting shape

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 do not mandate fair scheduling, and GPU schedulers are inherently
-unfair in practice.  Current approaches avoid this issue by exploiting
-scheduling quirks particular to today's GPUs in a manner that does not
-allow the GPU to be shared with other workloads (such as graphics
-rendering tasks).  We propose \emph{cooperative kernels}, an extension
-to the traditional GPU programming model geared towards writing
-blocking algorithms.  Workgroups of a cooperative kernel \emph{are}
-fairly scheduled, and multitasking is supported via a small set of
-language extensions that the kernel and scheduler use to cooperate.
-We describe the semantics of the cooperative kernels programming model
-and our prototype implementation.  We evaluate the approach by porting
-a set of irregular work stealing and graph algorithms to our
-programming model and examining their performance under multitasking
-across three GPUs.  Our implementation exploits no
-vendor-specific hardware, driver or compiler support, so our encouraging
-results provide a lower-bound on the efficiency with
-which cooperative kernels can be implemented in practice.
+unfair in practice.  This issue is currently circumvented by
+exploiting scheduling quirks particular to today's GPUs in a manner
+that does not allow the GPU to be shared with other workloads (such as
+graphics rendering tasks).  We propose \emph{cooperative kernels}, an
+extension to the traditional GPU programming model geared towards
+writing blocking algorithms.  Workgroups of a cooperative kernel are
+fairly scheduled, supporting blocking algorithms, and co-scheduling of
+other tasks is supported via a small set of language extensions that
+create a contract between the programmer and the scheduler.  We
+describe the semantics of the cooperative kernels programming model
+and discuss our portable prototype implementation.  We evaluate the
+technique by porting a set of irregular work stealing and graph
+algorithms to our programming model and assessing the performance and
+responsiveness of the approach across several GPUs.  Because our
+implementation exploits no vendor-specific hardware, driver or
+compiler support, our encouraging experimental results provide a
+lower-bound on the efficiency with which cooperative kernels can be
+implemented in practice.
 
@@
 of work to be processed is unknown ahead of time and may be determined
-by the computation itself.  There is growing interest in accelerating such algorithms on GPUs, particularly algorithms that
-traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
+by the computation itself.  The last decade has seen growing interest in accelerating irregular
+data-parallel algorithms on GPUs, particularly algorithms that
+traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10}.
 
@@
 To operate correctly, a blocking concurrent algorithm requires
-\emph{fair} scheduling of workgroups.  Without fairness, a
+\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
 workgroup may fail to make progress, leading to starvation.  For
@@
 \paragraph{A degree of fairness: occupancy-bound execution}
-GPU schedulers are \emph{not} fair. The current GPU
-programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
+GPU schedulers are \emph{not} fair. the current GPU
+programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
 HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
@@
 executing a GPU kernel is mapped to a hardware \emph{compute
-  unit}.\footnote{In practice, depending on the kernel, multiple workgroups might map to the same compute unit; we ignore this in our current discussion.}
+  unit}.\footnote{In practice the situation can be more complex: if a
+  kernel is sufficiently simple and if workgroups are small enough,
+  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
 %
@@
 units is via an \emph{occupancy-bound}
-execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
+execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}.
+\ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
 compute unit (it has become \emph{occupant}), the workgroup has
-exclusive access to the compute unit until it has finished execution.
+exclusive access to the compute unit until it has finished execution,
+after which a subsequent workgroup can be bound to the compute unit.
 Experiments suggest that the occupancy-bound execution model is widely
-implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
+implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
 evidently a simple model for device vendors to implement efficiently.
@@
 bound to separate compute units that operate in parallel.  Current GPU
-implementations of blocking algorithms assume the
+implementations of irregular data parallel algorithms assume the
 occupancy-bound execution model, which they exploit by launching 
 no more workgroups than there are available compute
-units~\cite{owens-persistent}. This works because \emph{today}'s GPUs seem to provide the
+units~\cite{...}. \ADComment{Tyler, please pick best cite for previous
+  claim.} This works because \emph{today}'s GPUs seem to provide the
 occupancy-bound execution model.
@@
 multitasking, because a workgroup effectively \emph{owns} a compute
-unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
-tasks, such as graphics rendering.
+unit until the workgroup has completed execution.  If a long-running
+kernel occupies all compute units, the GPU cannot be used for other
+tasks, such as graphics rendering, in the meantime.
 %Precisely because
@@
 Second, \emph{energy throttling} is
-an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
-power down some compute units if the battery level is low or if
+an important concern for battery-powered devices, such as smartphones
+and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
+power down one or more compute units if the battery level is low or if
 the user puts the device into a low power state.  Implementing energy
@@
 (including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
-and various engineers from GPU vendors), is that (1) for the above reasons, GPU vendors will
+and various engineers from GPU vendors), is that (1) GPU vendors will
 not commit to the occupancy-bound execution model that they currently
-implement,
+implement, due to the need for multitasking and energy throttling,
 yet (2) GPU vendors will not guarantee fair scheduling, due to the
 runtime overhead associated with supporting full preemption of
-workgroups.  Vendors instead wish to
-retain the essence of the simple occupancy-bound model, supporting preemption
+workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
+retain the essence of the occupancy-bound model, which is
+simple to implement efficiently, supporting preemption
 only in key special cases.
@@
 %
-To summarise: blocking algorithms
-demand fair scheduling of workgroups, but for good reasons
-GPU vendors will not commit even to the guarantees of the
+To summarise: blocking algorithms, such as most irregular algorithms,
+demand a degree of fair scheduling of workgroups, but for good reasons
+GPU vendors will not commit to the guarantees of even the
 occupancy-bound execution model.
@@
 We propose \emph{cooperative kernels}, an extension to the GPU
-programming model that aims to resolve this impasse.  A kernel
-that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
+programming model that aims to resolve this impasse.  A GPU kernel
+that requires workgroups to be fairly scheduled is identified as a
+\emph{cooperative} kernel, and is written using two additional
 language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.
@@
 sacrifice itself to the scheduler.  This indicates that the workgroup
-would ideally continue executing, but that if required the
-scheduler may destroy the workgroup for good.  The cooperative kernel
-must be prepared to deal with either scenario.
-%
+would ideally continue with its execution but that, if required, the
+scheduler may destroy the workgroup for good: the cooperative kernel
+is prepared to deal with either scenario.
+
 At a point where the cooperative kernel could benefit from having
@@
 
-\vspace{-1mm}
 \begin{enumerate}
@@
 \item Cooperative kernels can be implemented over the occupancy-bound
-  execution model that current GPUs provide: our proof-of-concept implementation requires
-  no additional hardware or driver support.  Thus the model should be eay for vendors to implement directly.
+  execution model that current GPUs provide---we present a
+  proof-of-concept implementation that requires
+  no special-purpose hardware or driver support.
 
@@
 \end{enumerate}
-\vspace{-1mm}
 
@@
 
-\vspace{-1mm}
 \begin{itemize}
@@
 \end{itemize}
-\vspace{-1mm}
-
-We begin by providing background on OpenCL via two motivating examples (\mysec\ref{sec:background}).
-At the end we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).
+
+We begin by providing an overview of the OpenCL programming model and
+two motivating examples of irregular algorithms (\mysec\ref{sec:background}).
+At the end of the paper we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).
 
@@
   the independent execution of workgroups since once in the work-pool,
-  they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.
+  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} CUDA similarly provides no guarantees~\cite{...}.
 
@@
 skeleton of a frontier-based graph traversal algorithm; such algorithms have
-been shown to be execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}. 
+been shown to be execute efficiently on GPUs~\cite{...}. \ADComment{Tyler, can you please work out which of the graph algorithm papers is the right one to cite here?}
 The kernel is
@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{Pannotia} and Lonestar~\cite{BNP12}
+graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...}
 GPU application suites, using global barriers.  We indicate how many of
@@
 
-\ADComment{Preamble}
-
-\HEComment{Below, just random notes so far}
-
-Cooperative scheduling enables to concurrently run non-persistent
-kernels alongside blocking ones. In terms of performance, we are
-interested to see whether the variations of workgroups for the blocking
-kernel.
-
-Performance of cooperative scheduling corresponds to execution time of
-cooperative kernels compared to the serial
-
-In order to simulate non-persistent workloads alongside kernels of
-blocking algorithms, we use a matrix multiplication kernel where
-matrix dimensions are tuned to trigger the relevant execution
-duration.
-
-we compare :
-
-time to execute original kernel
-
-time to execute non-persistent tasks
-
-time to execute both in cooperation
-
-also, response time for non-persistent
-
-use matmult to represent system workload, ref to previous section
-
-\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
+Cooperative scheduling enables to run a non-cooperative kernel alongside
+a cooperative one, we focus here on the performance impact of this
+multitasking and on the scheduling responsiveness. The performance is
+measured with respect to the performance of each kernel run in
+sequence. Responsiveness, i.e., how fast a non-cooperative kernel can be
+scheduled, depends not only on the scheduler implementation, but also on
+how often the cooperative kernel offer workgroups to be killed.
+
+% In order to simulate non-persistent workloads alongside kernels of
+% blocking algorithms, we use a matrix multiplication kernel where
+% matrix dimensions are tuned to trigger the relevant execution
+% duration.
+
+\paragraph{Experimental setup}
+For a given cooperative kernel and its data set, we make it run while
+regularly asking to schedule a non-cooperative kernel that simulates
+graphics rendering, according to the classes listed in
+\mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
+multiplication kernel as the non-cooperative workload, with matrix
+dimensions tailored for each GPU to reach the appropriate execution
+duration.  In separate runs, we ask the scheduler to launch this
+non-cooperative kernels with different number of workgroups: either 1, a
+quarter, a half or all but one workgroups among the ones available for
+the cooperative kernel. Moreover, for graph algorithms we try both the
+regular and the query barrier.
+
+Our experiments span 13 pairs of cooperative kernels and data sets, 3
+classes of non-cooperative kernel workloads and frequency, 4 quantities
+of workgroups claimed for the non-cooperative kernel and 2 variations of
+resizing barriers for graph algorithms, leading to 288
+configurations. Each of this configuration is run 10 times on each
+GPU. For each run, we record the execution time of the cooperative
+kernel. For each scheduling of the non-cooperative kernel during the
+run, we also record the \emph{gather time} needed by the scheduler to
+collect enough workgroups to launch the non-cooperative kernel, and the
+non-cooperative kernel execution time.
+
+Whenever a cooperative or non-cooperative kernel terminates, we verify
+that it has really performed its computation by comparing its results
+against the expected one. We encountered two kind of issues during the
+experiments: some configurations would lead to a system freeze or crash
+although they run fine on other GPUs. Although we don't claim that our
+programs are all correct, this issue seems to indicate a bug in the
+OpenCL compiler or driver of the problematic GPU. We also have some
+cooperative kernels that returned invalid results when the
+non-cooperative kernel takes all but one of the cooperative kernel
+workgroups, we decided to ignore these results.
 
 \paragraph{Results}  \ADComment{Discuss the results.}
+explain horizontal lines
+
+missing results for all but one in bfs rmat22 since errors
+
+also, high gather time corresponds to applications and data sets where
+resize/kill is triggered less often.
+
+Each configuration was run 10 times on each GPU.
 
@@
 There has been a lot of work on accelerating blocking irregular
-algorithms using GPUs, and on the \emph{persistent threads}
-programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
-These
+algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
+programming style for long-running kernels~\cite{...}.  These
 approaches rely on the occupancy-bound execution model, and aim to
-take ownership of the GPU by scheduling as many workgroups as the occupancy bound allows, relying on fair scheduling between these
+take ownership of the GPU by scheduling as many workgroups as the GPU
+has compute units, relying on fair scheduling between these
 workgroups.  This means the GPU is unavailable for other tasks, and
@@
 unit using our terminology) are drained until the SM becomes free for
-other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  A follow-up work adds the notion of SM \emph{flushing},
+other tasks.  A follow-up work adds the notion of SM \emph{flushing},
 whereby a workgroup running on an SM can be cancelled and re-scheduled
 from scratch if the workgroup has not yet committed any visible
-side-effects to shared memory~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
-using simulators, over sets of regular GPU kernels.  State-of-the-art \nvidia GPUs targeting data centers (where multi-tasking is essential) do support
-preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.
+side-effects to shared memory.  Both approaches have been evaluated
+using simulators, over sets of regular GPU kernels.
 
@@
 kernels, which run in an unmodified manner, and careful placement of
-our new language primitives can exploit the structure of a
-cooperative kernel, enabling more efficient multitasking than
+our new language primitives that exploit the structure of a
+cooperative kernel may enable more efficient multitasking than
 kernel-oblivious preemption.
@@
 
-\paragraph{Dynamic parallelism} Recent versions of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
-This \emph{dynamic parallelism} can be exploited to implement a
-GPU-based scheduler, by having an initial scheduler kernel repeatedly
-spawn further kernels as required, according to some scheduling
-policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, a kernel
-that uses dynamic parallelism is still subject to the scheduling
-limitations of current GPU programming models, and in particular the
-problem that workgroups are not fairly scheduled.  Hence this feature
-does not offer a solution to the problem of deploying blocking
-algorithms on GPUs in a safe manner that we tackle with this work.
+\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{...}.
+This \emph{dynamic parallelism} can be exploited to implement a GPU-based scheduler: an initial scheduler kernel uses the feature to spawn additional kernels as required~\cite{ }.
+
+ is launched
+
+%
+\ADComment{Kepler preemption support?  Mention in related work?}
+
+\ADComment{Sreepathi mentioned the NCCL framework.}
+
+\cite{TPO10}
+
+Dynamic parallelism could be extended to give barriers, but doesn't work with work stealing (less general).  It is in recent OpenCL.
+
+\cite{} \ADComment{Basic idea: map tasks to
+  SMs, not to workgroups, so that tasks that would benefit from being
+  processed by the same SM can be scheduled to the same SM.  Has
+  little to do with preemption, but more to do with irregular
+  computation.}
+
+\cite{DBLP:conf/isca/TanasicGCRNV14} \ADComment{I have not read, but
+  it's about enabling preemptive multiprogramming, so we should read
+  it carefully.}
+
+\cite{DBLP:conf/ppopp/Muyan-OzcelikO16} \ADComment{I have not read; we
+  should understand this one carefully.}
 
@@
 
-Cooperative multitasking was offered in older operating systems
-(e.g. Windows before the 95 edition) and is still used by some
-operating systems, such as RISC
-OS.\footnote{\url{http://www.riscos.info/index.php/Preemptive_multitasking}}
-Many programming languages offer cooperative multitasking via
-variations of coroutines, e.g. coroutine objects in
+\HEComment{cooperative multitasking: OS: RISC OS, IBM CICS, Windows
+  before 95, old MacOS, maybe Oberon? TRON?. In GPU context: some
+  paper mention ``cooperative'' multitasking but they simply refer to
+  the fact that kernels should not compute for too long in order to
+  let other have access to it. Also, many prog languages implements
+  variations of coroutines, We can see blocking channels of CSP (or
+  Golang more recently) as yielding (sending a message is yielding
+  computation to the process which receives it)}
+
+Cooperative multitasking was offered in ancient operating systems
+(e.g. Windows before the 95 edition) and is still used nowadays in
+RISC OS for instance. Many programming languages offer cooperative
+multitasking via variations of coroutines, e.g. coroutine objects in
 Python\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
-or Lua,\footnote{\url{https://www.lua.org/pil/9.1.html}} or Ruby's
-\emph{fiber}.\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}
-
-In the context of GPUs, cooperative multitasking has been used to
-refer to the standard setup today, whereby the onus is on a GPU kernel
-to complete execution within a reasonable time budget and return
-control to the CPU application~\cite{adriaens2012case,CPE:CPE1722}.
-In this setting, the programmer is responsible for ensuring cooperation between the host application and GPU kernel so that a kernel does not run for too long.
-%
-This is in contrast to our notion of cooperative kernels, which is designed specifically to support the needs of long-running kernels that one might wish to multitask with other workloads.
+or Lua\footnote{\url{https://www.lua.org/pil/9.1.html}}, or Ruby's
+fiber \footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}. In
+the context of GPU resources, cooperative multitasking seems to refer
+to sizing kernels such that they do not compute for too long in order
+to yield the whole GPU regularly~\cite{adriaens2012case}. Our
+contribution present a much more flexible cooperative multitasking,
+where the long-running application is never totally stopped.
+
+\HEComment{ RISC OS see :
+\url{http://www.riscos.info/index.php/Preemptive_multitasking}
+}
+
+% ###
+% Cooperative multitasking for GPU-accelerated grid systems
+
+% http://onlinelibrary.wiley.com/doi/10.1002/cpe.1722/full
+
+% @article {CPE:CPE1722,
+% author = {Ino, Fumihiko and Ogita, Akihiro and Oita, Kentaro and Hagihara, Kenichi},
+% title = {Cooperative multitasking for GPU-accelerated grid systems},
+% journal = {Concurrency and Computation: Practice and Experience},
+% volume = {24},
+% number = {1},
+% publisher = {John Wiley & Sons, Ltd},
+% issn = {1532-0634},
+% url = {http://dx.doi.org/10.1002/cpe.1722},
+% doi = {10.1002/cpe.1722},
+% pages = {96--107},
+% keywords = {multitasking, GPU, CUDA, grid computing},
+% year = {2012},
+% }
+
+% Not super interesting: they just divide blocking computation in
+% smaller independent kernels, and fires them at regular intervals to
+% let the system be responsive.
 
 \section{Conclusions and Future Work}\label{sec:conclusion}
-
-\ADComment{Do citation tidy-up.}
 
@@
 
-GPUs are known to have relaxed memory models~\cite{ABDGKPSW-2015}.  To abstractly
+GPUs are known to have relaxed memory models~\cite{ }.  To abstractly
 account for this, we assume that the shared state component $\sigma$

### 2016-11-15 write-ups/PLDI/paper.tex
Finished appendices.

@@
 %
-where: $\sigma$ represents the shared state of the kernel; $M
-\leq N$ is the number of \emph{active} workgroups; $w_i$ is the
-workgroup state for active workgroup $i$ ($0 \leq i < M$); and $N-M$
-occurrences of $\bot$ indicate \emph{absent} workgroups.  Let $\sharedstates$ denote the set of all possible shared states.
+where: $\sigma$ represents the shared state of the kernel; $M \leq N$
+is the number of \emph{active} workgroups; $w_i$ is the workgroup
+state for active workgroup $i$ ($0 \leq i < M$); and $N-M$ occurrences
+of $\bot$ indicate \emph{absent} workgroups.  Let $\sharedstates$
+denote the set of all possible shared states.  We regard
+workgroup-local storage as being part of the shared state of a kernel.
 
@@
 \forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \myss)
-\\
+\\\\
 \kappa \in \sync((\sigma , (w_{0}', \dots, w_{M-1}', \bot, \dots, \bot))
@@
 {
-(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow \kappa)
+(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow \kappa
 }
@@
 relation to determine how the thread's local state and the shared
-state component change.
+state component change.  For simplicity, this rule ignores the
+semantics of intra-workgroup barriers, which are not our focus here.
 
@@
 
-\ADComment{TODO: $\textsc{Barrier}$}
-
-\ADComment{TODO: $\textsc{Resizing-Barrier}$}
-
- \ADComment{Local memory is not
-  explicitly modeled but can be captured inside shared state.}
-\ADComment{Semantics for intra-workgroup barriers are ignored.}
+The $\textsc{Barrier}$ rule requires that \emph{every} thread
+executing the kernel reaches $\globalbarrier()$ an identical following
+statement.  This reflects that fact that $\globalbarrier()$ is a
+kernel-level function.  Each thread then skips over the
+$\globalbarrier()$ call, and the $\sync$ function is applied to yield
+a set of kernel states that can arise due to memory synchronization
+taking place.  An arbitary member of this set is selected as the next
+kernel state.
+
+Despite its apparent complexity, the $\textsc{Resizing-Barrier}$ rule simply implements the rewriting of $\resizingglobalbarrier$ in terms of $\globalbarrier$, $\offerkill$ and $\offerfork$ discussed in \mysec\ref{ }.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Added a placeholder appendix.
merging I guess

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
+\paragraph{Dynamically managing workgroups}
+
 The GPU scheduler then receives the data packet from the host
@@
 
-\paragraph{$\offerkill$ implementation}
+In the case where a cooperative kernel $K_c$ is executing and there
+are $W$ avaiable workgroups, the scheduler may decide to give $K_c$
+$W_k <= W$ additional workgroups via an $\offerfork$ command issued by
+$K_c$. To do this, scheduler signals $K_c$ through the $K_c$'s
+scheduler context communication channel to indicate the number of
+workgroups that are allowed to join at the next $\offerfork$ command.
+Then, when a workgroup $w$ of $K_c$ executes $\offerfork$, $w$
+observes $W_k$, and takes $W_k$ workgroups from the pool of available
+workgroups, provides them with transmit variables, target, and updated
+kernel context.
+\TSComment{this is sloppy}
+
+\paragraph{Efficient implementation}
+
+The interesting interactions with respect to performance and safety
+occur in the interaction of the scheduler and cooperative kernel. We
+discuss these details now.
+
+At the core of the scheduler is the workgroup pool, which maps each
+workgroup to a status and keeps track of the number of workgroups in
+each status. The workgroup pool can change in two ways: (1)
+\emph{assigning tasks}, where workgroups are changed from available to
+computing a kernel (2) \emph{becoming available}, where a workgroup
+exits a kernel computation (either through completion or $\offerkill$)
+and changes its status from computing a kernel to available.
+
+\emph{Assigning tasks} occurs either when the scheduler is launches a
+new kernel, or when a workgroup calls $\offerfork$ and observes that
+there are available resources. This creates a potential conflict
+between forking workgroups as well as the scheduler, as they may try
+to assign the same available workgroup to different tasks
+simultaneously. To fix this conflict, we introduce an
+\emph{assignment} mutex in the scheduler context.
+
+When the scheduler determines a new kernel has been launched, it spins
+waiting to acquire the assignment lock. Once acquired, it determines
+how to obtain the required number of workgroups (possibly by
+signalling a cooperative kernel to enable $\offkill$). The scheduler
+than waits until the required number of workgroups become available.
+Because it holds the assignment mutex, the scheduler knows that the
+number of available workgroups in the workgroup pool can only
+increase.
+
+When $\offerfork$ is executed, the calling workgroup has a single
+attempt to acquire the assignment mutex.  If it fails, the call
+returns without a fork. This provides a quick return of $\offerfork$
+when another actor is doing an assigning task. If the assignment mutex
+is acquired, then the calling workgroup can check the scheduler
+context to see if its allowed to fork by the scheduler and safely
+assign available workgroups from the workgroup pool in isolation from
+other actors that which to assign tasks. \TSComment{it also updates
+  the kernel context, and sends transmit variables while holding this
+  lock}
+
+As opposed to assignment tasks, availability tasks can be performed in
+a lock-free manner without conflict. This is because a workgroup only
+modifies its own state in the task pool and can atomically increment
+and decrement the required counters. Availability tasks are also able
+to be performed lock-free with assignment tasks (although accesses to
+the workgroup state and counters must be performed
+atomically). Availability tasks will only increase the number of
+available workgroups, and do not conflict with assignment tasks
+sending available workgroups to kernels.
+
+Although assignment tasks to the workgroup pool can be performed
+concurrently, concurrent calls to $\offerkill$ must be protected
+because workgroups that exit (a) modify the kernel context which
+causes a conflict and (b) the semantics of $\offerkill$ guarantee that
+only the workgroup with the highest id can exit. Rather than introduce
+a new mutex to protect $\offerkill$, we instead use the number of
+active groups from the kernel context.  The call to $\offerkill$
+begins by checking the calling workgroup id against the number of
+active groups for the kernel. If the calling workgroup is the last
+workgroup, it enters and is then able to check if it can be
+killed. Otherwise it immediately returns. If a workgroup determines
+it should be killed, it updates
+
+From this implementation, it falls out that $\offerkill$ and
+$\offerfork$ from the same workgroup can never succeed concurrently.
+If $\offerkill$ terminates the calling workgroup, then the scheduler
+is holding the assignment lock and will continue to hold it until 
+the call to $\offerkill$ finishes. Thus, any call to $\offerfork$
+during this time will not result in forking workgroups.
+
+\paragraph{Example}
 
@@
 browser.  For \emph{heavy} we used the Paper
-Planes\footnote{\url{https://paperplanes.world}}
+Planes\footnote{\url{https://www.chromeexperiments.com/experiment/paper-planes}}
 and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
@@
 co-scheduled.  This blow-up in code size can have a negative impact on
-compilation, e.g.\ due to higher register pressure, reducing the
+compiliation, e.g.\ due to higher register pressure, reducing the
 occupancy for the merged kernel---the maximum number of workgroups
@@
 
-Cooperative scheduling enables to run a non-cooperative kernel alongside
-a cooperative one, we focus here on the performance impact of this
-multitasking and on the scheduling responsiveness. The performance is
-measured with respect to the performance of each kernel run in
-sequence. Responsiveness, i.e., how fast a non-cooperative kernel can be
-scheduled, depends not only on the scheduler implementation, but also on
-how often the cooperative kernel offer workgroups to be killed.
-
-% In order to simulate non-persistent workloads alongside kernels of
-% blocking algorithms, we use a matrix multiplication kernel where
-% matrix dimensions are tuned to trigger the relevant execution
-% duration.
-
-\paragraph{Experimental setup}
-For a given cooperative kernel and its data set, we make it run while
-regularly asking to schedule a non-cooperative kernel that simulates
-graphics rendering, according to the classes listed in
-\mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
-multiplication kernel as the non-cooperative workload, with matrix
-dimensions tailored for each GPU to reach the appropriate execution
-duration.  In separate runs, we ask the scheduler to launch this
-non-cooperative kernels with different number of workgroups: either 1, a
-quarter, a half or all but one workgroups among the ones available for
-the cooperative kernel. Moreover, for graph algorithms we try both the
-regular and the query barrier.
-
-Our experiments span 13 pairs of cooperative kernels and data sets, 3
-classes of non-cooperative kernel workloads and frequency, 4 quantities
-of workgroups claimed for the non-cooperative kernel and 2 variations of
-resizing barriers for graph algorithms, leading to 288
-configurations. Each of this configuration is run 10 times on each
-GPU. For each run, we record the execution time of the cooperative
-kernel. For each scheduling of the non-cooperative kernel during the
-run, we also record the \emph{gather time} needed by the scheduler to
-collect enough workgroups to launch the non-cooperative kernel, and the
-non-cooperative kernel execution time.
-
-Whenever a cooperative or non-cooperative kernel terminates, we verify
-that it has really performed its computation by comparing its results
-against the expected one. We encountered two kind of issues during the
-experiments: some configurations would lead to a system freeze or crash
-although they run fine on other GPUs. Although we don't claim that our
-programs are all correct, this issue seems to indicate a bug in the
-OpenCL compiler or driver of the problematic GPU. We also have some
-cooperative kernels that returned invalid results when the
-non-cooperative kernel takes all but one of the cooperative kernel
-workgroups, we decided to ignore these results.
+\ADComment{Preamble}
+
+\HEComment{Below, just random notes so far}
+
+Cooperative scheduling enables to concurrently run non-persistent
+kernels alongside blocking ones. In terms of performance, we are
+interested to see whether the variations of workgroups for the blocking
+kernel.
+
+Performance of cooperative scheduling corresponds to execution time of
+cooperative kernels compared to the serial
+
+In order to simulate non-persistent workloads alongside kernels of
+blocking algorithms, we use a matrix multiplication kernel where
+matrix dimensions are tuned to trigger the relevant execution
+duration.
+
+we compare :
+
+time to execute original kernel
+
+time to execute non-persistent tasks
+
+time to execute both in cooperation
+
+also, response time for non-persistent
+
+use matmult to represent system workload, ref to previous section
+
+\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}
 
 \paragraph{Results}  \ADComment{Discuss the results.}
-explain horizontal lines
-
-missing results for all but one in bfs rmat22 since errors
-
-also, high gather time corresponds to applications and data sets where
-resize/kill is triggered less often.
-
-Each configuration was run 10 times on each GPU.
 

### 2016-11-15 write-ups/PLDI/paper.tex

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
+\section{Full Set of Graphs for Responsiveness Experiments}
+
+\ADComment{TODO: we need this.  Probably not the right title.}
+
 \end{document}

### 2016-11-15 write-ups/PLDI/paper.tex
fixed a compilation issue
Conclusion.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 by the computation itself.  There is growing interest in accelerating such algorithms on GPUs, particularly algorithms that
-traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
+traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
 
@@
 queue. Previous work addresses the feasibility and efficiency of work
-stealing on GPUs~\cite{dlb-web,TPO10}.
+stealing on GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}.
 
@@
 algorithms using GPUs, and on the \emph{persistent threads}
-programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
+programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
 These
@@
 
-\ADComment{Do citation tidy-up.}
-
-\ADComment{Next-generation GPUs and drivers for evaluation,
-  implementation in a real driver, migration of work between multiple
-  GPUs.}
+We have proposed \emph{cooperative kernels}, a small set of extensions
+to the GPUP programming model that allow long-running, blocking
+kernels to be fairly scheduled and to share GPU resources with other
+workloads via multitasking.  Our prototype implementation uses a
+megakernel to mock up the environment in which a cooperative kernel
+would run if native support for our programming model was provided.
+Our experimental results over three Intel GPUs show that the
+programming model is a good fit for the needs of current irregular
+algorithms that operate on graphs and use work stealing.  The
+performance that could be gained through a native implementation with
+dedicated driver support would be even better.
+
+There are three main avenues for future work at present: seeking out
+additional classes of irregular algorithms that would benefit from GPU
+acceleration using cooperartive kernels, investigating providing
+native driver support for cooperative kernels in an open source
+graphics driver, and integrating cooperative kernels into the back-end
+of a compiler that generates irregular OpenCL from a higher-level DSL,
+e.g.\ an OpenCL version of a recent compiler for graph algorithms that
+targets CUDA~\cite{DBLP:conf/oopsla/PaiP16}.  Longer-term, as more
+vendors adopt the advanced features of OpenCL that our approach
+requires, we will be in a position to assess the programming model
+across a wider range of devices.
 

### 2016-11-15 write-ups/PLDI/paper.tex

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 how to obtain the required number of workgroups (possibly by
-signalling a cooperative kernel to enable $\offkill$). The scheduler
+signalling a cooperative kernel to enable $\offerkill$). The scheduler
 than waits until the required number of workgroups become available.

### 2016-11-15 write-ups/PLDI/paper.tex
expe setup 5.5

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 by the computation itself.  There is growing interest in accelerating such algorithms on GPUs, particularly algorithms that
-traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
+traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
 
@@
 queue. Previous work addresses the feasibility and efficiency of work
-stealing on GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}.
+stealing on GPUs~\cite{dlb-web,TPO10}.
 
@@
 collect enough workgroups to launch the non-cooperative kernel, and the
-non-cooperative kernel execution time.
-
-Whenever a cooperative or non-cooperative kernel terminates, we verify
-that it has really performed its computation by comparing its results
-against the expected one. We encountered two kind of issues during the
-experiments: some configurations would lead to a system freeze or crash
-although they run fine on other GPUs. Although we don't claim that our
-programs are all correct, this issue seems to indicate a bug in the
-OpenCL compiler or driver of the problematic GPU. We also have some
-cooperative kernels that returned invalid results when the
-non-cooperative kernel takes all but one of the cooperative kernel
-workgroups, we decided to ignore these results.
+non-cooperative kernel execution time. We make sure all kernels perform
+their computations by comparing their result with expected results. We
+had to ignore some configurations where the non-cooperative kernel asks
+for all but one workgroups (namely bc, bfs and l-sssp on all chips, and
+color on Iris), and the game application on HD520, since they lead to
+system issues like GPU deadlock or blue screens---which we think are
+manifestations of driver bugs.
+
+% Whenever a cooperative or non-cooperative kernel terminates, we verify
+% that it has really performed its computation by comparing its results
+% against the expected one. We encountered two kind of issues during the
+% experiments: some configurations would lead to a system freeze or crash
+% although they run fine on other GPUs. Although we don't claim that our
+% programs are all correct, this issue seems to indicate a bug in the
+% OpenCL compiler or driver of the problematic GPU. We also have some
+% cooperative kernels that returned invalid results when the
+% non-cooperative kernel takes all but one of the cooperative kernel
+% workgroups, we decided to ignore these results.
 
@@
 algorithms using GPUs, and on the \emph{persistent threads}
-programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
+programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.
 These
@@
 
-We have proposed \emph{cooperative kernels}, a small set of extensions
-to the GPUP programming model that allow long-running, blocking
-kernels to be fairly scheduled and to share GPU resources with other
-workloads via multitasking.  Our prototype implementation uses a
-megakernel to mock up the environment in which a cooperative kernel
-would run if native support for our programming model was provided.
-Our experimental results over three Intel GPUs show that the
-programming model is a good fit for the needs of current irregular
-algorithms that operate on graphs and use work stealing.  The
-performance that could be gained through a native implementation with
-dedicated driver support would be even better.
-
-There are three main avenues for future work at present: seeking out
-additional classes of irregular algorithms that would benefit from GPU
-acceleration using cooperartive kernels, investigating providing
-native driver support for cooperative kernels in an open source
-graphics driver, and integrating cooperative kernels into the back-end
-of a compiler that generates irregular OpenCL from a higher-level DSL,
-e.g.\ an OpenCL version of a recent compiler for graph algorithms that
-targets CUDA~\cite{DBLP:conf/oopsla/PaiP16}.  Longer-term, as more
-vendors adopt the advanced features of OpenCL that our approach
-requires, we will be in a position to assess the programming model
-across a wider range of devices.
+\ADComment{Do citation tidy-up.}
+
+\ADComment{Next-generation GPUs and drivers for evaluation,
+  implementation in a real driver, migration of work between multiple
+  GPUs.}
 
@@
 %
-where: $\sigma$ represents the shared state of the kernel; $M \leq N$
-is the number of \emph{active} workgroups; $w_i$ is the workgroup
-state for active workgroup $i$ ($0 \leq i < M$); and $N-M$ occurrences
-of $\bot$ indicate \emph{absent} workgroups.  Let $\sharedstates$
-denote the set of all possible shared states.  We regard
-workgroup-local storage as being part of the shared state of a kernel.
+where: $\sigma$ represents the shared state of the kernel; $M
+\leq N$ is the number of \emph{active} workgroups; $w_i$ is the
+workgroup state for active workgroup $i$ ($0 \leq i < M$); and $N-M$
+occurrences of $\bot$ indicate \emph{absent} workgroups.  Let $\sharedstates$ denote the set of all possible shared states.
 
@@
 \forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \myss)
-\\\\
+\\
 \kappa \in \sync((\sigma , (w_{0}', \dots, w_{M-1}', \bot, \dots, \bot))
@@
 {
-(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow \kappa
+(\sigma, (w_{0}, \dots, w_{M-1}, \bot, \dots, \bot)) \rightarrow \kappa)
 }
@@
 relation to determine how the thread's local state and the shared
-state component change.  For simplicity, this rule ignores the
-semantics of intra-workgroup barriers, which are not our focus here.
+state component change.
 
@@
 
-The $\textsc{Barrier}$ rule requires that \emph{every} thread
-executing the kernel reaches $\globalbarrier()$ an identical following
-statement.  This reflects that fact that $\globalbarrier()$ is a
-kernel-level function.  Each thread then skips over the
-$\globalbarrier()$ call, and the $\sync$ function is applied to yield
-a set of kernel states that can arise due to memory synchronization
-taking place.  An arbitary member of this set is selected as the next
-kernel state.
-
-Despite its apparent complexity, the $\textsc{Resizing-Barrier}$ rule simply implements the rewriting of $\resizingglobalbarrier$ in terms of $\globalbarrier$, $\offerkill$ and $\offerfork$ discussed in \mysec\ref{ }.
+\ADComment{TODO: $\textsc{Barrier}$}
+
+\ADComment{TODO: $\textsc{Resizing-Barrier}$}
+
+ \ADComment{Local memory is not
+  explicitly modeled but can be captured inside shared state.}
+\ADComment{Semantics for intra-workgroup barriers are ignored.}
 
@@
 
-\section{Full Set of Graphs for Responsiveness Experiments}
-
-\ADComment{TODO: we need this.  Probably not the right title.}
-
 \end{document}

### 2016-11-15 write-ups/PLDI/paper.tex
Shortenings so far.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \newcommand{\TSComment}[1]{\textcolor{purple}{TS: #1}}
+
+\newcommand{\KC}{\mathit{KC}}
 
@@
 our case, the mega-kernel), must (a) track the state of all
-workgroups, i.e.\ if a workgroup is waiting or computing a kernel (b)
-maintain consistent context states for each kernel, i.e.\ the number
-of active workgroups computing a cooperative kernel and (c) provide a
-safe way for these states to be modified (according to the semantics
-of $\offerfork$ and $\offerkill$. We discuss how that is achieved
-here.
+workgroups, i.e.\ if a workgroup is waiting or computing a kernel, (b)
+maintain consistent context states for each kernel, e.g.\ the number
+of active workgroups computing a cooperative kernel, and (c) provide a
+safe way for these states to be modified in response to $\offerfork$ and $\offerkill$. We now discuss these issues in more detail.
 
@@
 be able to query a consistent context to obtain a workgroup id value
-(starting from zero up to the number of executing workgroups) and the
+and the
 number of workgroups executing the kernel. In single kernel execution,
-the OpenCL primitives {\tt get\_num\_groups()} and {\tt
-  get\_group\_id()} provide this functionality natively. For multiple
-kernel execution, we must emulate these primitives, per kernel,
-explicitly through \emph{kernel contexts}. This structure contains:
-(a) the number of active groups, (b) the total number of groups asked
-for, and (c) a mapping from the native workgroup id to a new workgroup
-id valid for a particular kernel. Our framework requires the native
-kernel context (to obtain a global base workgroup id) as well as an
-additional distinct kernel context for each kernel running
-concurrently. \TSComment{a driver implementation would likely generate
-these on the fly}.
+the OpenCL primitives $\getnumgroups()$ and $\getgroupid()$ provide this functionality natively. For multiple
+kernel execution, we emulate these primitives via a per kernel \emph{context}, containing, for a given kernel: the number of active groups $M$, the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
+id valid for the kernel.  In the case of a non-cooperative kernel, $M$ is undefined because the notion of active workgroups is not relevant.
 
@@
 
-In order to dynamically manage workgroups executing cooperative
-kernels, our framework must (a) track the state of each workgroup and
-(b) provide a channel of communication from the scheduler to
+To dynamically manage workgroups executing cooperative
+kernels, our framework must track the state of each workgroup and
+provide a channel of communication from the scheduler to
 workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
-we use a \emph{Scheduler context} structure which contains a mapping
-from primitive workgroup ids to their state i.e.\ waiting or computing
-a kernel as well as an integer value describing the number of
-workgroups in each state. The scheduler can then send cooperative
-kernels a \emph{resource message}, indicating if workgroups executing
-the cooperative kernel need to (a) exit at $\offerkill$ or (b)
-transmit their state to forking workgroups at $\offerfork$. Thus, the
+we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available}, or the id of the kernel that the workgroup is currently executing.
+The scheduler can then send cooperative
+kernels a \emph{resource message}, commanding workgroups executing
+the kernel to exit at $\offerkill$, or 
+spawn additional workgroups at $\offerfork$. Thus, the
 scheduler context needs a communication channel for each cooperative
 kernel. In our case, the communication channel is implemented using
-global atomic variables and OpenCL atomic operations to ensure
-consistency.
+global atomic variables.
 
@@
 
-As in native OpenCL, the host maintains the ability to launch kernels.
-In our framework, the host has the additional ability to request that
-a kernel be cooperative \emph{and} to specify the amount of
-\emph{resource} each kernel requires. In our implementation, the only
-resource we have explicit control over is the number of concurrently
-executing workgroups; implementations closer to metal might also
-consider local memory and registers as resource.
-
-To launch a kernel, the host sends a data packet to the GPU, which
-consists of: a kernel to execute, the inputs to the kernel, the amount
-of resource to use for the kernel (which includes the number of
-workgroups to execute the kernel), and a flag indicating if the kernel
+To launch a kernel, the host sends a data packet to the GPU
+consists of: a kernel to execute, the inputs to the kernel, the
+number of workgroups that should execute the kernel, and a flag indicating whether the kernel
 is cooperative. In our implementation, this communication channel
 between host and device is achieved through fine-grained SVM atomics.
-The kernel specified must be one of the kernels merged inside the
-mega-kernel. If there are large constant inputs to the kernel, these
-can be copied before mega-kernel launch to decrease the amount of SVM
-traffic.
 
 \paragraph{Dynamically managing workgroups}
-
-The GPU scheduler then receives the data packet from the host
-describing a kernel $K$ to be executed and now must make decisions
-about how to schedule the $K$. Consider the case where $K$ requires
-$W$ workgroups. The scheduler can query the scheduler context and if
-there is at least $W$ available workgroups, then $K$ can be scheduled
-immediately. If there are not enough available workgroups (say there
-are $W_a$ workgroups available), but a cooperative kernel $K_c$ is
-executing. The scheduler can signal $K_c$ through $K_c$'s
-communication channel in the scheduler context communication
-requesting that $K_c$ provide $W - W_a$ workgroups (via $\offerkill$)
-as soon as possible. The scheduler then waits until $W$ workgroups
-become available.
-
-When $W$ workgroups become available, the scheduler initialises a
-kernel context $KC$ for $K$, setting the number of active workgroups
-to $W$ and mapping available workgroups (located in the scheduler
-context) to their new $KC$ workgroup ids. The scheduler then sends $W$
+On receiving a data packet from the host
+describing a kernel launch $K$, the GPU scheduler must make decisions
+about how to schedule $K$. Suppose $K$ requests
+$N$ workgroups. The scheduler queries the scheduler context.  If
+there are at least $N$ available workgroups, $K$ can be scheduled
+immediately. Suppose instead that there are only $N_a < N$ available workgroups, but a cooperative kernel $K_c$ is
+executing. The scheduler can use $K_c$'s channel in the scheduler context to command $K_c$ to provide $N - N_a$ workgroups via $\offerkill$.
+Once $N$ workgroups are available, the scheduler initialises a
+kernel context $\KC$ for $K$, setting the number of active workgroups
+to $N$ and mapping available workgroups (located in the scheduler
+context) to their new $\KC$ workgroup ids. The scheduler then sends $N$
 workgroups from the available workgroups to execute kernel $K$ with
-kernel context $KC$.
-
-In the case where a cooperative kernel $K_c$ is executing and there
-are $W$ avaiable workgroups, the scheduler may decide to give $K_c$
-$W_k <= W$ additional workgroups via an $\offerfork$ command issued by
-$K_c$. To do this, scheduler signals $K_c$ through the $K_c$'s
-scheduler context communication channel to indicate the number of
-workgroups that are allowed to join at the next $\offerfork$ command.
-Then, when a workgroup $w$ of $K_c$ executes $\offerfork$, $w$
-observes $W_k$, and takes $W_k$ workgroups from the pool of available
-workgroups, provides them with transmit variables, target, and updated
-kernel context.
-\TSComment{this is sloppy}
+kernel context $\KC$.  Note that if the new kernel $K$ is itself a cooperative kernel, the scheduler could decide to provide $K$ with fewer than $N$ active workgroups initially.
+
+If a cooperative kernel $K_c$ is executing with fewer workgroups than it initially requested, the scheduler may decide to give $K_c$
+additional workgroups if $K_c$ calls $\offerfork()$. To do this, the scheduler asynchronously signals $K_c$ through $K_c$'s
+channel to indicate the number of
+workgroups that should join at the next $\offerfork$ command.
+When a workgroup $w$ of $K_c$ subsequently executes $\offerfork$, thread 0 of $w$ updates the kernel and scheduler contexts so that the given number of new workgroups are directed to the program point after the $\offerfork$ call.
+This involves selecting workgroups whose status is \emph{available}, as well as copying the values of $\transmit$-annotated variables to the new workgroups.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Add description of results in 5.5

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 color on Iris), and the game application on HD520, since they lead to
-system issues like GPU deadlock or blue screens---which we think are
-manifestations of driver bugs.
+system issues like GPU deadlock or blue screens---which we tend to think
+are manifestations of driver bugs.
 
@@
 \paragraph{Results}  \ADComment{Discuss the results.}
-explain horizontal lines
-
-missing results for all but one in bfs rmat22 since errors
-
-also, high gather time corresponds to applications and data sets where
-resize/kill is triggered less often.
-
-Each configuration was run 10 times on each GPU.
+
+\myfig~\ref{fig:fine-grained-timing} reports the measure for three
+configurations and represents the gather and execution time of the
+non-cooperative kernel w.r.t. the quantity of workgroups allocated to
+it. Times are in milliseconds, and a logarithmic scale is used since
+gather times tend to be much smaller than execution times. The
+horizontal grey lines indicates the desired rate for non-cooperative
+kernels.
+
+Responsiveness of the cooperative scheduling is measured by the gather
+time. In all examples, we observe that the gather time grows as we ask
+for more workgroups for the non-cooperative kernel: this is expected,
+since the scheduler must wait for more workgroups to call $\offerkill$
+or to reach a resizing barrier. Still, notice that the query barrier
+leads to better and stable gather times for graph applications, since it
+can collect all needed workgroups at the first time they pass through
+the barrier. Gather time also application-dependent since it is related
+to how often the cooperative kernel workgroups hit $\offerkill$ or
+$\resizingglobalbarrier$.
+
+\HEComment{missing results for all but one in bfs, discussed in expe
+setup, do we need to say it here ?}
 
@@
 \end{figure*}
+
+\HEComment{Figure gather time: replace ``O'' by ``M'' , ie max nb of active workgroups}
+
+\HEComment{Figure gather time: add ``ms'' to time coordinate}
+
+\HEComment{Figure gather time: lower legend in hd5500 bfs figure}
+
+\HEComment{Figure gather time: use consistent line styles for work stealing (ie, same style for ``heavy'' across all figures)}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Squeeeeeeeeeeeeeeze

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \newcommand{\KC}{\mathit{KC}}
+
+\newcommand{\kernelmerge}{\textsc{KernelMerge}}
 
@@
 
-\ADComment{This could be shorter.}
-However, these are not hard-and-fast rules: a scheduler might employ
-heuristics based on historical workloads to avoid thrashing.  For
-example, if graphics calls are known to arrive at a certain rate then
-the scheduler might decide that it is worth accepting an $\offerkill$
-from the cooperative kernel on the assumption that a graphics task
-will be requested momentarily.  Similarly, the scheduler might provide
-some, but not all, compute units to the cooperative kernel in response
-to an $\offerfork$ call if it has reason to believe that the remaining
-compute units are likely to be useful for other high priority tasks
-that are likely to arrive imminently.
-With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.
+Note that the above guidelines are not hard-and-fast rules.  For
+example, a smart scheduler might respond selectively to $\offerkill$
+and $\offerfork$ calls by predicting future high-priority workloads
+based on past workloads.
 
@@
 
-\ADComment{One cooperative, one non-cooperative.  Justification:
-  sufficient to evaluate the key use case of a long-running irregular
-  algorithm co-existing with a series of short-running tasks.
-  Simplifieid task of implementing prototype, which we do not want to
-  over-enginneer since it is only a mock-up.}
-
-\ADComment{Restrictions on kernel structure: variables at top of
-  kernel are implicitly $\transmit$-annotated, and $\offerkill$,
-  $\offerfork$ and $\resizingglobalbarrier$ calls must appear only in
-  kernel entry function.  Avoiding $\transmit$ allowed us to reuse the
-  clang OpenCL front-end, and restricitng starting points for fork to
-  be in main kernel allowed a ``goto''-based approach to redirecting
-  control flow; threading this through procedure calls would be
-  possible but laborious.  Restricting kill points enables kill to be
-  implemented as ``return''; exiting from a procedure call stack would
-  require stack unwinding support, akin to throwing an exception.
-  Again make point that all these issues are only relevant due to our
-  mock-up.}
+For ease of implementation, our current prototype supports
+multitasking a single cooperative kernel with a single non-cooperative
+kernel (though the non-cooperative kernel can be invoked many times).
+Rather than adding language support for $\transmit$ annotations, we
+assume that all variables declared at the root lexical scope of a
+cooperative kernel are $\transmit$-annotated; this allows our
+implementation to piggy-back on top of Clang (see below) without
+requiring front-end changes.  Finally, we require that $\offerkill$,
+$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
+function of a cooperative kernel.  This allows us to use
+$\keyword{goto}$ to direct threads joining the cooperative kernel to
+the appropriate location, and $\keyword{return}$ to force threads to
+leave the computation.  These restrictions still allow us to
+experiment with interesting irregular algorithms (see
+\mysec\ref{sec:experiments}), and could all be removed with
+engineering effort, but this effort might not be well invested since a
+full implementation of coopeartive kernels would not use the
+megakernel approach we present here.
 
@@
 
-\ADComment{Explain kernel merge tool briefly.  Explain library
-  briefly.}
-
-\TSComment{we should mention that non-cooperative kernels must be
-written in a persistent thread style}
+We have built a tool, \kernelmerge{}, that takes a cooperative kernel
+and a non-cooperative kernel and merges them to produce a megakernel
+with an embedded GPU scheduler.  We implemented \kernelmerge{} using the Clang LibTooling framework,\footnote{\url{http://clang.llvm.org/docs/LibTooling.html}}
+leveraging recent Clang support for OpenCL 2.0~\cite{ClangOpenCL20}.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Remove white lines in graph code

@@
   global nodes * out_nodes = n1;
-
   int tid = get_global_id();
   int stride = get_global_size();
-
 (*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
-
     for (int i = tid; i < nodes.size; i += stride) {
@@
     }
-
 (*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
@@
                  global nodes *n0, global nodes *n1) {
-
 (*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
@@
 (*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
-
   while(in_nodes.size > 0) {
-
 (*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
 (*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
-
     for (int i = tid; i < nodes.size; i += stride) {
@@
     }
-
     swap(&in_nodes, &out_nodes);

### 2016-11-15 write-ups/PLDI/paper.tex
Some work on experiments.

@@
 
-\subsection{Performance and Responsiveness of Cooperative Scheduling}\label{sec:responsiveness}
-
-Cooperative scheduling enables to run a non-cooperative kernel alongside
-a cooperative one, we focus here on the performance impact of this
-multitasking and on the scheduling responsiveness. The performance is
-measured with respect to the performance of each kernel run in
-sequence. Responsiveness, i.e., how fast a non-cooperative kernel can be
-scheduled, depends not only on the scheduler implementation, but also on
-how often the cooperative kernel offer workgroups to be killed.
+\subsection{Multitasking via Cooperative Scheduling}\label{sec:responsiveness}
+
+We now assess the performance impact of multitasking between a
+coopoerative and non-cooperative kernel, and on scheduling
+responsiveness.  The performance is measured with respect to the
+performance of each kernel run in sequence. Responsiveness, i.e., how
+fast a non-cooperative kernel can be scheduled, depends not only on
+the scheduler implementation, but also on how often the cooperative
+kernel offer workgroups to be killed.
 
@@
 \paragraph{Experimental setup}
-For a given cooperative kernel and its data set, we make it run while
-regularly asking to schedule a non-cooperative kernel that simulates
-graphics rendering, according to the classes listed in
+For a given cooperative kernel and its data set, we launch the kernel
+asynchronously and then repeatedly schedule a non-cooperative kernel
+that aims to simulate the intensity of a graphics rendering workload,
+according to the classes listed in
 \mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
@@
 dimensions tailored for each GPU to reach the appropriate execution
-duration.  In separate runs, we ask the scheduler to launch this
-non-cooperative kernels with different number of workgroups: either 1, a
-quarter, a half or all but one workgroups among the ones available for
-the cooperative kernel. Moreover, for graph algorithms we try both the
-regular and the query barrier.
+duration.  We conduct separate runs where we vary the number of workgroup slots requested by the non-cooperative kernel,
+considering the cases where one, a quarter, a half, and all-but-one, of the total number of workgroup slots is requested.
+Moreover, for the graph algorithms we try both the
+regular and query barrier implementations.
 
 Our experiments span 13 pairs of cooperative kernels and data sets, 3
-classes of non-cooperative kernel workloads and frequency, 4 quantities
+classes of non-cooperative kernel workloads \ADComment{cut ``and frequency''?} and frequency, 4 quantities
 of workgroups claimed for the non-cooperative kernel and 2 variations of
 resizing barriers for graph algorithms, leading to 288
-configurations. Each of this configuration is run 10 times on each
-GPU. For each run, we record the execution time of the cooperative
+configurations.  We run each configuration 10 times on each
+GPU, in order to report averaged performance numbers. For each run, we record the execution time of the cooperative
 kernel. For each scheduling of the non-cooperative kernel during the
@@
 collect enough workgroups to launch the non-cooperative kernel, and the
-non-cooperative kernel execution time. We make sure all kernels perform
-their computations by comparing their result with expected results. We
-had to ignore some configurations where the non-cooperative kernel asks
+non-cooperative kernel execution time. We sanity-check whether all kernels perform
+their computations corretly by comparing their computed result with expected reference results.
+
+To work around compiler crash and compiler hang bugs in Intel's current OpenCL drivers, we had to apply what semantics-preserving changes to the code generated by \kernelmerge{} for a few applications configurations.
+We could not obtain results for some configurations where the non-cooperative kernel asks
 for all but one workgroups (namely bc, bfs and l-sssp on all chips, and
-color on Iris), and the game application on HD520, since they lead to
-system issues like GPU deadlock or blue screens---which we tend to think
-are manifestations of driver bugs.
+color on Iris), and could not get the game application to run at all on HD520; these configurations led to 
+catastrophic issues such as display freezes or blue screens; we also attribute these to bugs in Intel's current drivers.
 
@@
 
-\paragraph{Results}  \ADComment{Discuss the results.}
-
-\myfig~\ref{fig:fine-grained-timing} reports the measure for three
+\paragraph{Results}
+
+\myfiglong\ref{fig:fine-grained-timing} reports the measure for three
 configurations and represents the gather and execution time of the

### 2016-11-15 write-ups/PLDI/paper.tex
added stuff

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \newcommand{\KC}{\mathit{KC}}
-
-\newcommand{\kernelmerge}{\textsc{KernelMerge}}
 
@@
 
-Note that the above guidelines are not hard-and-fast rules.  For
-example, a smart scheduler might respond selectively to $\offerkill$
-and $\offerfork$ calls by predicting future high-priority workloads
-based on past workloads.
+\ADComment{This could be shorter.}
+However, these are not hard-and-fast rules: a scheduler might employ
+heuristics based on historical workloads to avoid thrashing.  For
+example, if graphics calls are known to arrive at a certain rate then
+the scheduler might decide that it is worth accepting an $\offerkill$
+from the cooperative kernel on the assumption that a graphics task
+will be requested momentarily.  Similarly, the scheduler might provide
+some, but not all, compute units to the cooperative kernel in response
+to an $\offerfork$ call if it has reason to believe that the remaining
+compute units are likely to be useful for other high priority tasks
+that are likely to arrive imminently.
+With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.
 
@@
 
-For ease of implementation, our current prototype supports
-multitasking a single cooperative kernel with a single non-cooperative
-kernel (though the non-cooperative kernel can be invoked many times).
-Rather than adding language support for $\transmit$ annotations, we
-assume that all variables declared at the root lexical scope of a
-cooperative kernel are $\transmit$-annotated; this allows our
-implementation to piggy-back on top of Clang (see below) without
-requiring front-end changes.  Finally, we require that $\offerkill$,
-$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
-function of a cooperative kernel.  This allows us to use
-$\keyword{goto}$ to direct threads joining the cooperative kernel to
-the appropriate location, and $\keyword{return}$ to force threads to
-leave the computation.  These restrictions still allow us to
-experiment with interesting irregular algorithms (see
-\mysec\ref{sec:experiments}), and could all be removed with
-engineering effort, but this effort might not be well invested since a
-full implementation of coopeartive kernels would not use the
-megakernel approach we present here.
+\ADComment{One cooperative, one non-cooperative.  Justification:
+  sufficient to evaluate the key use case of a long-running irregular
+  algorithm co-existing with a series of short-running tasks.
+  Simplifieid task of implementing prototype, which we do not want to
+  over-enginneer since it is only a mock-up.}
+
+\ADComment{Restrictions on kernel structure: variables at top of
+  kernel are implicitly $\transmit$-annotated, and $\offerkill$,
+  $\offerfork$ and $\resizingglobalbarrier$ calls must appear only in
+  kernel entry function.  Avoiding $\transmit$ allowed us to reuse the
+  clang OpenCL front-end, and restricitng starting points for fork to
+  be in main kernel allowed a ``goto''-based approach to redirecting
+  control flow; threading this through procedure calls would be
+  possible but laborious.  Restricting kill points enables kill to be
+  implemented as ``return''; exiting from a procedure call stack would
+  require stack unwinding support, akin to throwing an exception.
+  Again make point that all these issues are only relevant due to our
+  mock-up.}
 
@@
 
-We have built a tool, \kernelmerge{}, that takes a cooperative kernel
-and a non-cooperative kernel and merges them to produce a megakernel
-with an embedded GPU scheduler.  We implemented \kernelmerge{} using the Clang LibTooling framework,\footnote{\url{http://clang.llvm.org/docs/LibTooling.html}}
-leveraging recent Clang support for OpenCL 2.0~\cite{ClangOpenCL20}.
+\ADComment{Explain kernel merge tool briefly.  Explain library
+  briefly.}
+
+\TSComment{we should mention that non-cooperative kernels must be
+written in a persistent thread style}
 
@@
 
-\paragraph{Efficient implementation}
-
-The interesting interactions with respect to performance and safety
-occur in the interaction of the scheduler and cooperative kernel. We
-discuss these details now.
-
-At the core of the scheduler is the workgroup pool, which maps each
-workgroup to a status and keeps track of the number of workgroups in
-each status. The workgroup pool can change in two ways: (1)
-\emph{assigning tasks}, where workgroups are changed from available to
-computing a kernel (2) \emph{becoming available}, where a workgroup
-exits a kernel computation (either through completion or $\offerkill$)
-and changes its status from computing a kernel to available.
-
-\emph{Assigning tasks} occurs either when the scheduler is launches a
-new kernel, or when a workgroup calls $\offerfork$ and observes that
-there are available resources. This creates a potential conflict
-between forking workgroups as well as the scheduler, as they may try
-to assign the same available workgroup to different tasks
-simultaneously. To fix this conflict, we introduce an
-\emph{assignment} mutex in the scheduler context.
-
-When the scheduler determines a new kernel has been launched, it spins
-waiting to acquire the assignment lock. Once acquired, it determines
-how to obtain the required number of workgroups (possibly by
-signalling a cooperative kernel to enable $\offerkill$). The scheduler
-than waits until the required number of workgroups become available.
-Because it holds the assignment mutex, the scheduler knows that the
-number of available workgroups in the workgroup pool can only
-increase.
-
-When $\offerfork$ is executed, the calling workgroup has a single
-attempt to acquire the assignment mutex.  If it fails, the call
-returns without a fork. This provides a quick return of $\offerfork$
-when another actor is doing an assigning task. If the assignment mutex
-is acquired, then the calling workgroup can check the scheduler
-context to see if its allowed to fork by the scheduler and safely
-assign available workgroups from the workgroup pool in isolation from
-other actors that which to assign tasks. \TSComment{it also updates
-  the kernel context, and sends transmit variables while holding this
-  lock}
-
-As opposed to assignment tasks, availability tasks can be performed in
-a lock-free manner without conflict. This is because a workgroup only
-modifies its own state in the task pool and can atomically increment
-and decrement the required counters. Availability tasks are also able
-to be performed lock-free with assignment tasks (although accesses to
-the workgroup state and counters must be performed
-atomically). Availability tasks will only increase the number of
-available workgroups, and do not conflict with assignment tasks
-sending available workgroups to kernels.
-
-Although assignment tasks to the workgroup pool can be performed
-concurrently, concurrent calls to $\offerkill$ must be protected
-because workgroups that exit (a) modify the kernel context which
-causes a conflict and (b) the semantics of $\offerkill$ guarantee that
-only the workgroup with the highest id can exit. Rather than introduce
-a new mutex to protect $\offerkill$, we instead use the number of
-active groups from the kernel context.  The call to $\offerkill$
-begins by checking the calling workgroup id against the number of
-active groups for the kernel. If the calling workgroup is the last
-workgroup, it enters and is then able to check if it can be
-killed. Otherwise it immediately returns. If a workgroup determines
-it should be killed, it updates
-
-From this implementation, it falls out that $\offerkill$ and
-$\offerfork$ from the same workgroup can never succeed concurrently.
-If $\offerkill$ terminates the calling workgroup, then the scheduler
-is holding the assignment lock and will continue to hold it until 
-the call to $\offerkill$ finishes. Thus, any call to $\offerfork$
-during this time will not result in forking workgroups.
+\paragraph{Safe implementation}
+
+Concurrent calls to $\offerfork()$ must be protected with a mutex,
+otherwise workgroups calling $\offerfork()$ will conflict over the
+available workgroups. The same conflict over available workgroups can
+occur at a new kernel launch between the scheduler and a workgroup
+calling $\offerfork()$ occurs between the scheduler assigning tasks.
+Thus we provide a mutex, called the \emph{assignment mutex}, in the
+scheduler context which must be held anytime a group of available
+resources is assigned to a kernel.
+
+The provides an indirect mutex protecting concurrent calls to
+$\offerfork()$ and $\offerkill()$. That is, $\offerkill()$ only
+commands a workgroup to exit when the scheduler holds the
+\emph{assignment lock}, as the scheduler is trying to obtain work
+groups to send to a task, and $\offerfork()$ cannot spawn workgroups
+without holding this lock.
+
+The semantics of $\offerkill()$, which allow only the workgroup with
+the highest id provide natural mutual exclusion for exiting workgroups
+(as part of the exit, the kernel context is updated). That is, a
+workgroup can only exit at $\offerkill()$ if it is the workgroup with
+the highest id. An exiting thread releases this natural mutex by
+decrementing the number of active workgroups in the kernel context.
 
@@
 
-\ADComment{Optimised barrier implementation.}
+In section~\ref{??}, we show that a resizing barrier is semantically
+equivalent to three normal barrier calls with calls to cfork and ckill
+in between. It is possible, however to implement the resizing with
+only one call to a global barrier with $\offerfork$ and $\offerkill$
+baked into the implementation. \TSComment{note here that these
+  implementations require no cfork or ckill outside of the barrier}
+
+We consider barriers which have the master/slave model
+(e.g. \cite{??}). That is, one workgroup (the master) collects signals
+from the other workgroups (the slaves) indicating that they have
+arrived at the barrier and are waiting for a reply indicating that
+they may leave the barrier. Once the master has recieved a signal from
+all the slaves, it replies with a signal saying that the slaves may
+leave. \TSComment{Need to discuss new active workgroup number that
+is stable across iterations}
+
+A first-attempt at incorporating $\offerfork$ and $\offerkill$ into
+such a barrier is straight forward. Upon entering the barrier, the
+slaves first execute $\offerkill$, possibly exiting. The master then
+waits for $M$ slaves (the number of active workgroups), which may
+decrease, (based on the $\offerkill$ instructions executed by the
+slaves), but will not increase. Once the master observes $M$ slaves
+have arrived, it knows all other workgroups are in the kernel are
+spinning, waiting to be released. The master executes $\offerfork$,
+where there is a conditional for the potentially newly spawned threads
+to join the other slaves waiting to be released. Finally, the master
+releases all the slaves: the original slaves and the new slaves that
+joined at $\offerfork$.
+
+While this barrier implementation works, it has its flaws. Workgroups
+only execute $\offerkill$ once per global barrier call. Depending on
+the order in which workgroups arrive at the barrier, it is possible
+that only one workgroup is killed per barrier call which could lead to
+a high time for the scheduler to gather the sufficient amount of
+workgroups to execute another kernel concurrently.
+
+We observe if we provide an additional scheduler functionality for a
+workgroup executing a cooperative kernel, we can greatly reduce the
+gather time. We call the new function {\tt query()}. The semantics
+simply return the number of workgroups that the scheduler requires to
+kill from the cooperative kernel. A new barrier can now be implented
+as follows: (1) Master waits for all slaves; (2) Master forks and sets
+new workgroups as slaves. (3) Master calls {\tt query()} and gets
+value $W$; (4) master releases slaves and broadcasts the value from
+{\tt query} to slaves. (5) Workgroups with id greater than $M-W$ spin
+on $\offerkill$, as they know from {\tt query()} that the scheduler
+needs them to exit. We show in Section~\ref{} that the barrier using
+query greatly reduces the gather time in practice.
 
@@
 color on Iris), and the game application on HD520, since they lead to
-system issues like GPU deadlock or blue screens---which we tend to think
-are manifestations of driver bugs.
+system issues like GPU deadlock or blue screens---which we think are
+manifestations of driver bugs.
 
@@
 \paragraph{Results}  \ADComment{Discuss the results.}
-
-\myfig~\ref{fig:fine-grained-timing} reports the measure for three
-configurations and represents the gather and execution time of the
-non-cooperative kernel w.r.t. the quantity of workgroups allocated to
-it. Times are in milliseconds, and a logarithmic scale is used since
-gather times tend to be much smaller than execution times. The
-horizontal grey lines indicates the desired rate for non-cooperative
-kernels.
-
-Responsiveness of the cooperative scheduling is measured by the gather
-time. In all examples, we observe that the gather time grows as we ask
-for more workgroups for the non-cooperative kernel: this is expected,
-since the scheduler must wait for more workgroups to call $\offerkill$
-or to reach a resizing barrier. Still, notice that the query barrier
-leads to better and stable gather times for graph applications, since it
-can collect all needed workgroups at the first time they pass through
-the barrier. Gather time also application-dependent since it is related
-to how often the cooperative kernel workgroups hit $\offerkill$ or
-$\resizingglobalbarrier$.
-
-\HEComment{missing results for all but one in bfs, discussed in expe
-setup, do we need to say it here ?}
+explain horizontal lines
+
+missing results for all but one in bfs rmat22 since errors
+
+also, high gather time corresponds to applications and data sets where
+resize/kill is triggered less often.
+
+Each configuration was run 10 times on each GPU.
 
@@
 \end{figure*}
-
-\HEComment{Figure gather time: replace ``O'' by ``M'' , ie max nb of active workgroups}
-
-\HEComment{Figure gather time: add ``ms'' to time coordinate}
-
-\HEComment{Figure gather time: lower legend in hd5500 bfs figure}
-
-\HEComment{Figure gather time: use consistent line styles for work stealing (ie, same style for ``heavy'' across all figures)}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Remove figure of cooperative version of work stealing, adapt text

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 (*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
-  while (more_work(queues)) {
+(*@\label{line:wksteal:mainloop}@*)  while (more_work(queues)) {
 (*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
@@
 
-\myfiglong\ref{fig:workstealing-cooperative} shows a cooperative
-version of the work stealing example of \myfig\ref{fig:workstealing} from \mysec\ref{sec:openclexamples}.
-%
 In this example there is no state to transmit since a computation is
 entirely parameterised by a task, which is retrieved from a queue
-located in global memory. In the main loop, before trying to obtain a
-task, we make the workgroup offer itself to be forked or killed
-(lines~\ref{line:cwksteal:offerfork}-\ref{line:cwksteal:offerkill}). Note
-that a workgroup may be killed even if its associated task queue is
-not empty, since remaining tasks will be stolen by other
-workgroups. Since $\offerfork$ may be the entry point of a workgroup,
-the queue id must now be computed after it (line~\ref{line:cwksteal:getgroupid}). In particular, the queue
-id cannot be transmitted since we want a newly spawned workgroup to
-read its own queue and not the one of the forking workgroup.
-
-% The queue id cannot be transmitted since it corresponds to the
-% workgroup id.
-
-\begin{figure}
-
-\begin{lstlisting}
-kernel work_stealing(global Task * queues) {
-(*@\label{line:cwksteal:morework}@*)  while (more_work(queues)) {
-(*@\label{line:cwksteal:offerfork}@*)    offer_fork();
-(*@\label{line:cwksteal:offerkill}@*)    offer_kill();
-(*@\label{line:cwksteal:getgroupid}@*)    int queue_id = get_group_id();
-(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
-    if (t) {
-      // task processing may create more work
-(*@\label{line:cwksteal:processtask}@*)      process_task(t, queues, queue_id);
-    }
-  }
-}
-\end{lstlisting}
-
-\caption{Cooperative version of the work stealing kernel of
-  \myfig~\ref{fig:workstealing}, using $\offerfork$ and
-  $\offerkill$}\label{fig:workstealing-cooperative}
-\end{figure}
+located in global memory. With respect to \myfig~\ref{fig:workstealing},
+we add $\offerfork$ and $\offerkill$ calls at the start of the main loop
+(below line~\ref{line:wksteal:mainloop}) to let a workgroup offer itself
+to be killed or forked, respectively, before it processes a task.  Note
+that a workgroup may be killed even if its associated task queue is not
+empty, since remaining tasks will be stolen by other workgroups. In
+addition, since $\offerfork$ may be the entry point of a workgroup, the
+queue id must now be computed after it, so we move
+line~\ref{line:wksteal:getgroupid} to be placed just before
+line~\ref{line:wksteal:poporsteal}. In particular, the queue id cannot
+be transmitted since we want a newly spawned workgroup to read its own
+queue and not the one of the forking workgroup.
+
+% \begin{figure}
+% \begin{lstlisting}
+% kernel work_stealing(global Task * queues) {
+% (*@\label{line:cwksteal:morework}@*)  while (more_work(queues)) {
+% (*@\label{line:cwksteal:offerfork}@*)    offer_fork();
+% (*@\label{line:cwksteal:offerkill}@*)    offer_kill();
+% (*@\label{line:cwksteal:getgroupid}@*)    int queue_id = get_group_id();
+% (*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
+%     if (t) {
+%       // task processing may create more work
+% (*@\label{line:cwksteal:processtask}@*)      process_task(t, queues, queue_id);
+%     }
+%   }
+% }
+% \end{lstlisting}
+% \caption{Cooperative version of the work stealing kernel of
+%   \myfig~\ref{fig:workstealing}, using $\offerfork$ and
+%   $\offerkill$}\label{fig:workstealing-cooperative}
+% \end{figure}
 
@@
 
-\subsection{Multitasking via Cooperative Scheduling}\label{sec:responsiveness}
-
-We now assess the performance impact of multitasking between a
-coopoerative and non-cooperative kernel, and on scheduling
-responsiveness.  The performance is measured with respect to the
-performance of each kernel run in sequence. Responsiveness, i.e., how
-fast a non-cooperative kernel can be scheduled, depends not only on
-the scheduler implementation, but also on how often the cooperative
-kernel offer workgroups to be killed.
+\subsection{Performance and Responsiveness of Cooperative Scheduling}\label{sec:responsiveness}
+
+Cooperative scheduling enables to run a non-cooperative kernel alongside
+a cooperative one, we focus here on the performance impact of this
+multitasking and on the scheduling responsiveness. The performance is
+measured with respect to the performance of each kernel run in
+sequence. Responsiveness, i.e., how fast a non-cooperative kernel can be
+scheduled, depends not only on the scheduler implementation, but also on
+how often the cooperative kernel offer workgroups to be killed.
 
@@
 \paragraph{Experimental setup}
-For a given cooperative kernel and its data set, we launch the kernel
-asynchronously and then repeatedly schedule a non-cooperative kernel
-that aims to simulate the intensity of a graphics rendering workload,
-according to the classes listed in
+For a given cooperative kernel and its data set, we make it run while
+regularly asking to schedule a non-cooperative kernel that simulates
+graphics rendering, according to the classes listed in
 \mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
@@
 dimensions tailored for each GPU to reach the appropriate execution
-duration.  We conduct separate runs where we vary the number of workgroup slots requested by the non-cooperative kernel,
-considering the cases where one, a quarter, a half, and all-but-one, of the total number of workgroup slots is requested.
-Moreover, for the graph algorithms we try both the
-regular and query barrier implementations.
+duration.  In separate runs, we ask the scheduler to launch this
+non-cooperative kernels with different number of workgroups: either 1, a
+quarter, a half or all but one workgroups among the ones available for
+the cooperative kernel. Moreover, for graph algorithms we try both the
+regular and the query barrier.
 
 Our experiments span 13 pairs of cooperative kernels and data sets, 3
-classes of non-cooperative kernel workloads \ADComment{cut ``and frequency''?} and frequency, 4 quantities
+classes of non-cooperative kernel workloads and frequency, 4 quantities
 of workgroups claimed for the non-cooperative kernel and 2 variations of
 resizing barriers for graph algorithms, leading to 288
-configurations.  We run each configuration 10 times on each
-GPU, in order to report averaged performance numbers. For each run, we record the execution time of the cooperative
+configurations. Each of this configuration is run 10 times on each
+GPU. For each run, we record the execution time of the cooperative
 kernel. For each scheduling of the non-cooperative kernel during the
@@
 collect enough workgroups to launch the non-cooperative kernel, and the
-non-cooperative kernel execution time. We sanity-check whether all kernels perform
-their computations corretly by comparing their computed result with expected reference results.
-
-To work around compiler crash and compiler hang bugs in Intel's current OpenCL drivers, we had to apply what semantics-preserving changes to the code generated by \kernelmerge{} for a few applications configurations.
-We could not obtain results for some configurations where the non-cooperative kernel asks
+non-cooperative kernel execution time. We make sure all kernels perform
+their computations by comparing their result with expected results. We
+had to ignore some configurations where the non-cooperative kernel asks
 for all but one workgroups (namely bc, bfs and l-sssp on all chips, and
-color on Iris), and could not get the game application to run at all on HD520; these configurations led to 
-catastrophic issues such as display freezes or blue screens; we also attribute these to bugs in Intel's current drivers.
+color on Iris), and the game application on HD520, since they lead to
+system issues like GPU deadlock or blue screens---which we tend to think
+are manifestations of driver bugs.
 
@@
 
-\paragraph{Results}
-
-\myfiglong\ref{fig:fine-grained-timing} reports the measure for three
+\paragraph{Results}  \ADComment{Discuss the results.}
+
+\myfig~\ref{fig:fine-grained-timing} reports the measure for three
 configurations and represents the gather and execution time of the

### 2016-11-15 write-ups/PLDI/paper.tex
Fixed undefined references.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
-Concurrent calls to $\offerfork()$ must be protected with a mutex,
-otherwise workgroups calling $\offerfork()$ will conflict over the
+Concurrent calls to $\offerfork$ must be protected by a mutex
+to avoid racing on the
 available workgroups. The same conflict over available workgroups can
 occur at a new kernel launch between the scheduler and a workgroup
-calling $\offerfork()$ occurs between the scheduler assigning tasks.
-Thus we provide a mutex, called the \emph{assignment mutex}, in the
-scheduler context which must be held anytime a group of available
-resources is assigned to a kernel.
-
-The provides an indirect mutex protecting concurrent calls to
-$\offerfork()$ and $\offerkill()$. That is, $\offerkill()$ only
-commands a workgroup to exit when the scheduler holds the
-\emph{assignment lock}, as the scheduler is trying to obtain work
-groups to send to a task, and $\offerfork()$ cannot spawn workgroups
-without holding this lock.
-
-The semantics of $\offerkill()$, which allow only the workgroup with
-the highest id provide natural mutual exclusion for exiting workgroups
-(as part of the exit, the kernel context is updated). That is, a
-workgroup can only exit at $\offerkill()$ if it is the workgroup with
-the highest id. An exiting thread releases this natural mutex by
-decrementing the number of active workgroups in the kernel context.
+calling $\offerfork()$.
+
+The scheduler context provides an \emph{assignment mutes} to protecting concurrent calls to
+$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup is only permitted to exit if the scheduler currently holds the assignment mutex; that is, the scheduler is trying to obtain work
+groups to send to a task.  Conversely, on calling $\offerfork$ a workgroup can only spawn new workgroups if the assignment mutex is not held by the scheduler.
+
+Concurrent calls to $\offerkill$ are acceptable because only the workgroup with the highest id may exit, automatically ensuring mutual exclusion for exit-related modifications.
 
@@
 
-In section~\ref{??}, we show that a resizing barrier is semantically
-equivalent to three normal barrier calls with calls to cfork and ckill
-in between. It is possible, however to implement the resizing with
+In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
+of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
+ It is possible, however to implement the resizing with
 only one call to a global barrier with $\offerfork$ and $\offerkill$
@@
 We consider barriers which have the master/slave model
-(e.g. \cite{??}). That is, one workgroup (the master) collects signals
+(e.g.\ \cite{XF10}). That is, one workgroup (the master) collects signals
 from the other workgroups (the slaves) indicating that they have
@@
 on $\offerkill$, as they know from {\tt query()} that the scheduler
-needs them to exit. We show in Section~\ref{} that the barrier using
+needs them to exit. We show in \mysec\ref{sec:responsiveness} that the barrier using
 query greatly reduces the gather time in practice.
-
-\subsection{Vendor Implementation Defects}
-
-\ADComment{Briefly rant about our pain.
-Compiler hang, compiler crashes, deadlocks, defunct processes.
-Caution: don't make it sound so terrible that you can't trust our results.
-Gently acknowledge that our implementation may be imperfect.}
 
@@
 
-Despite its apparent complexity, the $\textsc{Resizing-Barrier}$ rule simply implements the rewriting of $\resizingglobalbarrier$ in terms of $\globalbarrier$, $\offerkill$ and $\offerfork$ discussed in \mysec\ref{ }.
+Despite its apparent complexity, the $\textsc{Resizing-Barrier}$ rule simply implements the rewriting of $\resizingglobalbarrier$ in terms of $\globalbarrier$, $\offerkill$ and $\offerfork$ discussed in \mysec\ref{sec:semantics}.
 

### 2016-11-15 write-ups/PLDI/paper.tex
updated some of the tables
Shortened related work.

@@
 \section{Prototype Implementation}\label{sec:implementation}
+
+\ADComment{We have to explain in this seciton how we get fairness.}
 
@@
 algorithms using GPUs, and on the \emph{persistent threads}
-programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
-These
-approaches rely on the occupancy-bound execution model, and aim to
-take ownership of the GPU by scheduling as many workgroups as the occupancy bound allows, relying on fair scheduling between these
-workgroups.  This means the GPU is unavailable for other tasks, and
-the method is not future-proof because (as discussed throughout the
-paper) the occupancy-bound execution model is not guaranteed by GPU
-programming models and is unlikely to be guaranteed in practice by
-future GPU platforms.
-
-As we have demonstrated for a representative set of irregular
-algorithms, our cooperative kernels model allows blocking algorithms
+programming style for long-running
+kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
+These approaches rely on the occupancy-bound execution model, flooding
+available compute units with work, so that the GPU is unavailable for
+other tasks, and assuming fair scheduling between occupant workgroups,
+which is unlikely to be guaranteed on future GPU platforms.
+%
+As our experiments demonstrate, our cooperative kernels model allows blocking algorithms
 to be upgraded to run under a fair scheduler in a manner that, with
@@
 
-Hardware support for preemptive multitasking for \nvidia GPUs has been
-proposed, enabling full preemption as well as \emph{SM-draining},
+Hardware support for preemption has been proposed for \nvidia GPUs,
+as well as \emph{SM-draining}
 whereby workgroups occupying a symmetric multiprocessor (SM; a compute
-unit using our terminology) are drained until the SM becomes free for
-other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  A follow-up work adds the notion of SM \emph{flushing},
-whereby a workgroup running on an SM can be cancelled and re-scheduled
-from scratch if the workgroup has not yet committed any visible
-side-effects to shared memory~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
-using simulators, over sets of regular GPU kernels.  State-of-the-art \nvidia GPUs targeting data centers (where multi-tasking is essential) do support
+unit using our terminology) are allowed to complete until the SM becomes free for
+other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited the presence of blocking constructs, since it may not be possible to drain a blocked workgroup.
+%
+A follow-up work adds the notion of SM \emph{flushing},
+where a workgroup can be cancelled and re-scheduled
+from scratch if it has not yet committed side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
+using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs targeting data centers now support
 preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.
 
-The advantages of these approaches are that they do not require manual
-programmer effort, and the capability to fully preempt a workgroup
-guards against a rogue long-running kernel making the GPU unavailable
-for other tasks.  The drawbacks are that (a) because preemption is
-enabled by default, \emph{every} kernel takes a potential performance
-hit, and (b) the SM draining optimisation does not help in the case of
-blocking algorithms (not considered in these works), since a blocked
-workgroup cannot be drained.
-
-Our solution does not impose a performance penalty on non-cooperative
-kernels, which run in an unmodified manner, and careful placement of
-our new language primitives can exploit the structure of a
-cooperative kernel, enabling more efficient multitasking than
-kernel-oblivious preemption.
-
-There is scope for combining the approaches: if a
+The advantages of preemption are that it does not require programmer
+effort, and guards against a rogue long-running kernel making the GPU
+unavailable for other tasks.  The main drawback is that \emph{every}
+kernel takes a potential performance hit.
+%
+If implemented natively, our solution does not impact performance on regular kernels that are not scheduled with cooperative kernels.
+Furthermore, careful placement of
+our new primitives can exploit problem structure, enabling more efficient multitasking than
+application-oblivious preemption.  There is scope for combining the approaches: if a
 cooperative kernel is insufficiently cooperative, i.e.\ it does not
@@
 to best utilise GPU
-resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}.  The TimeGraph
+resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and the TimeGraph
 approach similarly aims to optimise scheduling of competing workloads,
-in the context of OpenGL, using GPU-CPU synchronisation so that the
-CPU can monitor and control the work that is scheduled on the
-GPU~\cite{DBLP:conf/usenix/KatoLRI11}.  To best utilise the compute
-units of a GPU for a particular kernel, SM-centric program
-transformations have been proposed, which aim to direct work to
-specific compute units, rather than to specific workgroups, based on
-the observation that in practice multiple workgroups may share the
-same compute unit, so that said workgrops might be in a good position
-to exploit memory caches associated with that compute
-unit~\cite{DBLP:conf/ics/WuCLSV15}.  None of these approaches tackle
-the problem of \emph{fair} scheduling on GPUs that our work addresses, thus they do not aid in safe deployment of blocking irregular algorithms.
-
-\paragraph{Dynamic parallelism} Recent versions of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
-This \emph{dynamic parallelism} can be exploited to implement a
+in the context of OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.
+None of these approaches tackle
+the problem of \emph{fair} scheduling on GPUs, thus they do not aid in safe deployment of blocking irregular algorithms.
+
+CUDA and OpenCL provide the facility for a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
+This \emph{dynamic parallelism} can be used to implement a
 GPU-based scheduler, by having an initial scheduler kernel repeatedly
 spawn further kernels as required, according to some scheduling
-policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, a kernel
-that uses dynamic parallelism is still subject to the scheduling
-limitations of current GPU programming models, and in particular the
-problem that workgroups are not fairly scheduled.  Hence this feature
-does not offer a solution to the problem of deploying blocking
-algorithms on GPUs in a safe manner that we tackle with this work.
+policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, kernels
+that uses dynamic parallelism are still prone to unfair scheduling of workgroups, so this feature
+does not help solve the problem of safely deploying blocking
+algorithms on GPUs.
 
@@
 variations of coroutines, e.g. coroutine objects in
-Python\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
-or Lua,\footnote{\url{https://www.lua.org/pil/9.1.html}} or Ruby's
-\emph{fiber}.\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}
+Python
+%\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
+and Lua,
+%\footnote{\url{https://www.lua.org/pil/9.1.html}}
+and Ruby's
+\emph{fiber}.
+%\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}
 
 In the context of GPUs, cooperative multitasking has been used to
-refer to the standard setup today, whereby the onus is on a GPU kernel
-to complete execution within a reasonable time budget and return
-control to the CPU application~\cite{adriaens2012case,CPE:CPE1722}.
-In this setting, the programmer is responsible for ensuring cooperation between the host application and GPU kernel so that a kernel does not run for too long.
-%
-This is in contrast to our notion of cooperative kernels, which is designed specifically to support the needs of long-running kernels that one might wish to multitask with other workloads.
+refer to the standard programming model, where the onus is on the GPU kernel
+to complete execution within a reasonable time budget~\cite{adriaens2012case,CPE:CPE1722}.
+%
+This is in contrast to our cooperative kernels, which specifically aim to support the needs of long-running GPU tasks.
 

### 2016-11-15 write-ups/PLDI/paper.tex

@@
 
-\ADComment{We have to explain in this seciton how we get fairness.}
-
-We envisage cooperative kernels being implemented as part of the
-runtime for a GPU implementation of OpenCL, requiring changes to the
-device driver allowing it to intercept our new primitives and schedule
-and deschedule workgroups accordingly.
-
-To experiment with our ideas on current GPUs, we have instead
-developed a prototype implementation that mocks up the required
-runtime support via a \emph{megakernel}, and exploits the
-occupancy-bound execution model that these GPUs provide.  To be clear:
-an aim of cooperative kernels is to avoid depending on the
-occupancy-bound model; our protoytpe exploits this model simply to
-allow us to experiment with current GPUs whose proprietary drivers we
-cannot change.
-
-\subsection{The megakernel approach}
+Our vision is that support for cooperative kernels will be integrated
+in the runtimes of future GPU implementations of OpenCL and related
+programming models, with driver and compiler support for our new
+primitives.  To experiment with our ideas on current GPUs, we have
+developed a prototype that mocks up the required runtime support via a
+\emph{megakernel}, and exploits the occupancy-bound execution model
+that these GPUs provide to ensure fair scheduling between workgroups.
+We emphasise that an aim of cooperative kernels is to \emph{avoid}
+depending on the occupancy-bound model.  Our protoytpe exploits this
+model simply to allow us to experiment with current GPUs whose
+proprietary drivers we cannot change.  We describe the megakernel approach (\mysec\ref{sec:megakernel})
+and detail various aspects of the scheduler component of our implementation (\mysec\ref{sec:schedulerimpl}).
+
+\subsection{The megakernel approach}\label{sec:megakernel}
 
@@
 
-\subsection{Scheduler design}\label{sec:resizingbarrier}
+\subsection{Scheduler design}\label{sec:resizingbarrier}\label{sec:schedulerimpl}
 
@@
 \begin{tabular}{ l r r r r r r}
-App. & resizing barr. & kill & fork & transmit & LoC & data sets\\
+App. & barriers & kill & fork & transmit & LoC & inputs\\
 \hline
@@
 \begin{table}
-\small
+\footnotesize
 \centering
@@
 \hline
-\# comp. units  & 24 & 24 & 47 \\
-Driver   & 20.19.15.4501 & \ADComment{...} & 20.19.15.4463 \\
-Host CPU & i3-6100U & \ADComment{...} & i3-5157U \\
+\# CU  & 24 & 24 & 47 \\
+Driver   & 20.19.15.4501 & 10.18.15.4281 & 20.19.15.4463 \\
+Host CPU & i3-6100U & i7-5600U & i3-5157U \\
 \end{tabular}

### 2016-11-15 write-ups/PLDI/paper.tex
Polishing implementation.

@@
 
-\subsection{The megakernel approach}\label{sec:megakernel}
-
-The mock-up works as follows.  Instead of multitasking multiple
-separate kernels, we merge a set of kernels into a megakernel---a
-single, monolithic kernel---and encode the scheduling logic inside
-this megakernel.  The megakernel is launched with as many workgroups
-as can run concurrently according to the occupancy-bound.  One
-workgroup takes the role of scheduler, and the remaining workgroups
-act as a pool of workers.  A worker repeatedly queries the scheduler
-to be assigned a task.  A task corresponds to executing a cooperative
-or non-cooperative kernel.  In the case where a task corresponds to a
-non-cooperative kernel, the workgroup executes the kernel function
-uninterrupted and awaits further work.  In the case of a cooperative
-kernel, the workgroup either starts from the kernel entry point,
-corresponding to the workgroup being an \emph{initial} workgroup of
-the kernel, or immediately jumps to a designated point within the
-cooperative kernel, corresponding to the workgroup being forked by an
-already executing workgroup.  In the latter case, the new workgroup
-also receives a struct containing the values of all relevant
-$\transmit$-annotated variables.
+\subsection{The megakernel mock up}\label{sec:megakernel}
+
+Instead of multitasking multiple separate kernels, we merge a set of
+kernels into a megakernel---a single, monolithic kernel.  The
+megakernel is launched with as many workgroups as can be occupant
+concurrently.  One workgroup takes the role of scheduler, and the
+scheduling logic is embedded as part of the megakernel.  The remaining
+workgroups act as a pool of workers.  A worker repeatedly queries the
+scheduler to be assigned a task.  A task corresponds to executing a
+cooperative or non-cooperative kernel.  In the non-cooperative case,
+the workgroup executes the relevant kernel function uninterrupted,
+then awaits further work.  In the cooperative case, the workgroup
+either starts from the kernel entry point or immediately jumps to a
+designated point within the kernel, depending on whether the workgroup
+is an initial workgroup of the kernel, or a forked workgroup.  In the
+latter case, the new workgroup also receives a struct containing the
+values of all relevant $\transmit$-annotated variables.
 
@@
 function of a cooperative kernel.  This allows us to use
-$\keyword{goto}$ to direct threads joining the cooperative kernel to
-the appropriate location, and $\keyword{return}$ to force threads to
-leave the computation.  These restrictions still allow us to
-experiment with interesting irregular algorithms (see
-\mysec\ref{sec:experiments}), and could all be removed with
-engineering effort, but this effort might not be well invested since a
-full implementation of coopeartive kernels would not use the
-megakernel approach we present here.
+$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
+of the kernel.  With these restrictions we can experiment
+with interesting irregular algorithms (see
+\mysec\ref{sec:experiments}).  A non-mock implementation of
+coopeartive kernels would not use the megakernel approach, so we do not deem
+the engineering effort associated with lifting these restrictions to
+be worthwhile.
 
@@
 To enable multitasking through cooperative kernels, the runtime (in
-our case, the mega-kernel), must (a) track the state of all
-workgroups, i.e.\ if a workgroup is waiting or computing a kernel, (b)
-maintain consistent context states for each kernel, e.g.\ the number
-of active workgroups computing a cooperative kernel, and (c) provide a
-safe way for these states to be modified in response to $\offerfork$ and $\offerkill$. We now discuss these issues in more detail.
-
-\paragraph{Kernel context}
+our case, the mega-kernel) must track the state of all
+workgroups, i.e.\ whether a workgroup is waiting or computing a kernel,
+maintain consistent context states for each kernel, e.g.\ tracking the number
+of active workgroups, and provide a
+safe way for these states to be modified in response to $\offerfork$/$\offerkill$. We discuss these issues,
+and describe the implementation of an efficient resizing barrier.
+
+We describe how the scheduler would handle arbitrary
+combinations of kernels, though as noted above our current
+implementation is restricted to the case of two kernels.
+
+\paragraph{Kernel and scheduler contexts}
 
@@
 and the
-number of workgroups executing the kernel. In single kernel execution,
-the OpenCL primitives $\getnumgroups()$ and $\getgroupid()$ provide this functionality natively. For multiple
-kernel execution, we emulate these primitives via a per kernel \emph{context}, containing, for a given kernel: the number of active groups $M$, the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
-id valid for the kernel.  In the case of a non-cooperative kernel, $M$ is undefined because the notion of active workgroups is not relevant.
-
-\paragraph{Scheduler context}
+number of workgroups executing the kernel. For multi-kernel
+kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
+primitives via a per kernel \emph{context}.
+For a given kernel this records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
+id valid for the kernel.
 
@@
 kernels, our framework must track the state of each workgroup and
-provide a channel of communication from the scheduler to
+provide a channel of communication from the scheduler workgroup to
 workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
-we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available}, or the id of the kernel that the workgroup is currently executing.
+we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available} or the id of the kernel that the workgroup is currently executing.
 The scheduler can then send cooperative
-kernels a \emph{resource message}, commanding workgroups executing
-the kernel to exit at $\offerkill$, or 
+kernels a \emph{resource message}, commanding workgroups to exit at $\offerkill$, or 
 spawn additional workgroups at $\offerfork$. Thus, the
 scheduler context needs a communication channel for each cooperative
-kernel. In our case, the communication channel is implemented using
-global atomic variables.
-
-\paragraph{Launching kernels}
+kernel. We implement the communication channels using atomic variables in global memory.
+
+\paragraph{Launching kernels and managing workgroups}
 
 To launch a kernel, the host sends a data packet to the GPU
-consists of: a kernel to execute, the inputs to the kernel, the
+consisting of: a kernel to execute, the inputs to the kernel, the
 number of workgroups that should execute the kernel, and a flag indicating whether the kernel
-is cooperative. In our implementation, this communication channel
-between host and device is achieved through fine-grained SVM atomics.
-
-\paragraph{Dynamically managing workgroups}
+is cooperative. In our implementation, this host-device communication channel is built using fine-grained SVM atomics.
+
 On receiving a data packet from the host

### 2016-11-15 write-ups/PLDI/paper.tex
Various edit

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \usepackage{url}                  % format URLs
+\urlstyle{same}
 \usepackage{listings}          % format code
@@
 not require reshaping of the original algorithm. Most graph come with 2
-different data sets, leading to 13 application/data set pairs in total.
+different input data sets, leading to 13 application/data set pairs in total.
 
@@
 
-We used this approach to measure the GPU availability required for
-three types of rendering: \emph{light}, whereby desktop icons are
-smoothly emphasised under the mouse pointer; \emph{medium}, whereby
-window dragging over the desktop is smoothly animated; and
-\emph{heavy}, which requires smooth animation of a WebGL shader in a
-browser.  For \emph{heavy} we used the Paper
-Planes\footnote{\url{https://paperplanes.world}}
-and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
-shaders running in a recent version of Firefox.
-
-\begin{table}
-\small
-\centering
-\begin{tabular}{ l r r }
-Workload & Duration (ms) & Rate (ms)\\
-\hline
-light & 3 & 70\\
-medium & 3 & 40\\
-heavy & 10 & 40\\
-\end{tabular}
-\caption{Computed estimates of the GPU requirements for three different classes of graphics rendering}
-\label{tab:noncooperativeworkload}
-\end{table}
-
-\mytablong\ref{tab:noncooperativeworkload} illustrates the data
-computed for the three types of workloads across our GPUs.  The
-\emph{Duration} and \emph{Rate} columns report values corresponding to
-$(X-Y)/n$ and $X/n$ in the above discussion, respectively.  For the
-medium and heavy workloads, the 40ms rate corresponds to graphics
+We used this approach to measure the GPU availability required for three
+types of rendering: \emph{light}, whereby desktop icons are smoothly
+emphasised under the mouse pointer; \emph{medium}, whereby window
+dragging over the desktop is smoothly animated; and \emph{heavy}, which
+requires smooth animation of a WebGL shader in a browser.  For
+\emph{heavy} we used WebGL demos from the Chrome
+experiments\footnote{\url{https://www.chromeexperiments.com/}} running
+in a recent version of Firefox.
+
+% the Paper Planes\footnote{\url{https://paperplanes.world}} and
+% Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
+% shaders running in a recent version of Firefox.
+
+% \begin{table}
+% \small
+% \centering
+% \begin{tabular}{ l r r }
+% Workload & Duration (ms) & Rate (ms)\\
+% \hline
+% light & 3 & 70\\
+% medium & 3 & 40\\
+% heavy & 10 & 40\\
+% \end{tabular}
+% \caption{Computed estimates of the GPU requirements for three different classes of graphics rendering}
+% \label{tab:noncooperativeworkload}
+% \end{table}
+
+Our results are the following: light workloads needs to run 3ms every
+70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
+the medium and heavy workloads, the 40ms rate corresponds to graphics
 rendering occuring 25 times per second, which coincides with the human
@@
 cheaper for basic display rendering compared with rendering that uses
-complex shaders.
-
-We use these results to guide our evaluation of the responsiveness of
-cooperative scheduling in \mysec\ref{sec:responsiveness}.
+complex shaders. We use these results to guide our evaluation of the
+responsiveness of cooperative scheduling in
+\mysec\ref{sec:responsiveness}.
+
+% The
+% \emph{Duration} and \emph{Rate} columns report values corresponding to
+% $(X-Y)/n$ and $X/n$ in the above discussion, respectively.
+
+% \mytablong\ref{tab:noncooperativeworkload} illustrates the data
+% computed for the three types of workloads across our GPUs.
 
@@
 that aims to simulate the intensity of a graphics rendering workload,
-according to the classes listed in
-\mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
-multiplication kernel as the non-cooperative workload, with matrix
-dimensions tailored for each GPU to reach the appropriate execution
-duration.  We conduct separate runs where we vary the number of workgroup slots requested by the non-cooperative kernel,
-considering the cases where one, a quarter, a half, and all-but-one, of the total number of workgroup slots is requested.
-Moreover, for the graph algorithms we try both the
-regular and query barrier implementations.
+according to the 3 classes listed in \mysec~\ref{sec:sizingnoncoop}. In
+practice, we use a matrix multiplication kernel as the non-cooperative
+workload, with matrix dimensions tailored for each GPU to reach the
+appropriate execution duration.  We conduct separate runs where we vary
+the number of workgroup slots requested by the non-cooperative kernel,
+considering the cases where one, a quarter, a half, and all-but-one, of
+the total number of workgroup slots is requested.  Moreover, for the
+graph algorithms we try both the regular and query barrier
+implementations.
 
 Our experiments span 13 pairs of cooperative kernels and data sets, 3
-classes of non-cooperative kernel workloads \ADComment{cut ``and frequency''?} and frequency, 4 quantities
-of workgroups claimed for the non-cooperative kernel and 2 variations of
-resizing barriers for graph algorithms, leading to 288
-configurations.  We run each configuration 10 times on each
-GPU, in order to report averaged performance numbers. For each run, we record the execution time of the cooperative
-kernel. For each scheduling of the non-cooperative kernel during the
-run, we also record the \emph{gather time} needed by the scheduler to
-collect enough workgroups to launch the non-cooperative kernel, and the
-non-cooperative kernel execution time. We sanity-check whether all kernels perform
-their computations corretly by comparing their computed result with expected reference results.
+classes of non-cooperative kernel workloads, 4 quantities of workgroups
+claimed for the non-cooperative kernel and 2 variations of resizing
+barriers for graph algorithms, leading to 288 configurations.  We run
+each configuration 10 times on each GPU, in order to report averaged
+performance numbers. For each run, we record the execution time of the
+cooperative kernel. For each scheduling of the non-cooperative kernel
+during the run, we also record the \emph{gather time} needed by the
+scheduler to collect enough workgroups to launch the non-cooperative
+kernel, and the non-cooperative kernel execution time. We sanity-check
+whether all kernels perform their computations corretly by comparing
+their computed result with expected reference results.
 

### 2016-11-15 write-ups/PLDI/paper.tex

@@
   columns=fullfelxible,
+  xleftmargin=2em,
   %commentstyle=\rmfamily\itshape,

### 2016-11-15 write-ups/PLDI/paper.tex
fixed lines

@@
 \usepackage{url}                  % format URLs
-\urlstyle{same}
 \usepackage{listings}          % format code
@@
 not require reshaping of the original algorithm. Most graph come with 2
-different input data sets, leading to 13 application/data set pairs in total.
+different data sets, leading to 13 application/data set pairs in total.
 
@@
 
-We used this approach to measure the GPU availability required for three
-types of rendering: \emph{light}, whereby desktop icons are smoothly
-emphasised under the mouse pointer; \emph{medium}, whereby window
-dragging over the desktop is smoothly animated; and \emph{heavy}, which
-requires smooth animation of a WebGL shader in a browser.  For
-\emph{heavy} we used WebGL demos from the Chrome
-experiments\footnote{\url{https://www.chromeexperiments.com/}} running
-in a recent version of Firefox.
-
-% the Paper Planes\footnote{\url{https://paperplanes.world}} and
-% Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
-% shaders running in a recent version of Firefox.
-
-% \begin{table}
-% \small
-% \centering
-% \begin{tabular}{ l r r }
-% Workload & Duration (ms) & Rate (ms)\\
-% \hline
-% light & 3 & 70\\
-% medium & 3 & 40\\
-% heavy & 10 & 40\\
-% \end{tabular}
-% \caption{Computed estimates of the GPU requirements for three different classes of graphics rendering}
-% \label{tab:noncooperativeworkload}
-% \end{table}
-
-Our results are the following: light workloads needs to run 3ms every
-70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
-the medium and heavy workloads, the 40ms rate corresponds to graphics
+We used this approach to measure the GPU availability required for
+three types of rendering: \emph{light}, whereby desktop icons are
+smoothly emphasised under the mouse pointer; \emph{medium}, whereby
+window dragging over the desktop is smoothly animated; and
+\emph{heavy}, which requires smooth animation of a WebGL shader in a
+browser.  For \emph{heavy} we used the Paper
+Planes\footnote{\url{https://paperplanes.world}}
+and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
+shaders running in a recent version of Firefox.
+
+\begin{table}
+\small
+\centering
+\begin{tabular}{ l r r }
+Workload & Duration (ms) & Rate (ms)\\
+\hline
+light & 3 & 70\\
+medium & 3 & 40\\
+heavy & 10 & 40\\
+\end{tabular}
+\caption{Computed estimates of the GPU requirements for three different classes of graphics rendering}
+\label{tab:noncooperativeworkload}
+\end{table}
+
+\mytablong\ref{tab:noncooperativeworkload} illustrates the data
+computed for the three types of workloads across our GPUs.  The
+\emph{Duration} and \emph{Rate} columns report values corresponding to
+$(X-Y)/n$ and $X/n$ in the above discussion, respectively.  For the
+medium and heavy workloads, the 40ms rate corresponds to graphics
 rendering occuring 25 times per second, which coincides with the human
@@
 cheaper for basic display rendering compared with rendering that uses
-complex shaders. We use these results to guide our evaluation of the
-responsiveness of cooperative scheduling in
-\mysec\ref{sec:responsiveness}.
-
-% The
-% \emph{Duration} and \emph{Rate} columns report values corresponding to
-% $(X-Y)/n$ and $X/n$ in the above discussion, respectively.
-
-% \mytablong\ref{tab:noncooperativeworkload} illustrates the data
-% computed for the three types of workloads across our GPUs.
+complex shaders.
+
+We use these results to guide our evaluation of the responsiveness of
+cooperative scheduling in \mysec\ref{sec:responsiveness}.
 
@@
 that aims to simulate the intensity of a graphics rendering workload,
-according to the 3 classes listed in \mysec~\ref{sec:sizingnoncoop}. In
-practice, we use a matrix multiplication kernel as the non-cooperative
-workload, with matrix dimensions tailored for each GPU to reach the
-appropriate execution duration.  We conduct separate runs where we vary
-the number of workgroup slots requested by the non-cooperative kernel,
-considering the cases where one, a quarter, a half, and all-but-one, of
-the total number of workgroup slots is requested.  Moreover, for the
-graph algorithms we try both the regular and query barrier
-implementations.
+according to the classes listed in
+\mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
+multiplication kernel as the non-cooperative workload, with matrix
+dimensions tailored for each GPU to reach the appropriate execution
+duration.  We conduct separate runs where we vary the number of workgroup slots requested by the non-cooperative kernel,
+considering the cases where one, a quarter, a half, and all-but-one, of the total number of workgroup slots is requested.
+Moreover, for the graph algorithms we try both the
+regular and query barrier implementations.
 
 Our experiments span 13 pairs of cooperative kernels and data sets, 3
-classes of non-cooperative kernel workloads, 4 quantities of workgroups
-claimed for the non-cooperative kernel and 2 variations of resizing
-barriers for graph algorithms, leading to 288 configurations.  We run
-each configuration 10 times on each GPU, in order to report averaged
-performance numbers. For each run, we record the execution time of the
-cooperative kernel. For each scheduling of the non-cooperative kernel
-during the run, we also record the \emph{gather time} needed by the
-scheduler to collect enough workgroups to launch the non-cooperative
-kernel, and the non-cooperative kernel execution time. We sanity-check
-whether all kernels perform their computations corretly by comparing
-their computed result with expected reference results.
+classes of non-cooperative kernel workloads \ADComment{cut ``and frequency''?} and frequency, 4 quantities
+of workgroups claimed for the non-cooperative kernel and 2 variations of
+resizing barriers for graph algorithms, leading to 288
+configurations.  We run each configuration 10 times on each
+GPU, in order to report averaged performance numbers. For each run, we record the execution time of the cooperative
+kernel. For each scheduling of the non-cooperative kernel during the
+run, we also record the \emph{gather time} needed by the scheduler to
+collect enough workgroups to launch the non-cooperative kernel, and the
+non-cooperative kernel execution time. We sanity-check whether all kernels perform
+their computations corretly by comparing their computed result with expected reference results.
 

### 2016-11-15 write-ups/PLDI/paper.tex
change data set -> input to match table

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \usepackage{url}                  % format URLs
-\urlstyle{same}
+\urlstyle{same}                   % URL use same font as regular text
+                                  %   (regular font is less wide than tt)
 \usepackage{listings}          % format code
@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{Pannotia} and Lonestar~\cite{BNP12}
-GPU application suites, using global barriers.  We indicate how many of
-the original number of barriers are changed to resizing barriers, and
-how many variables need to be transmitted.  In all cases all barriers
-were converted, except for bc which contains barriers deeper in the call
-stack, a case not supported by our prototype. Still, these barriers
-could in principle also be converted. The remaining 2 algorithms are
-workstealing applications: each required the addition of $\offerfork$
-and $\offerkill$ at the start of the main loop, and no variables needed
-to be transmitted.  For all these examples the porting to cooperative
-kernels was similarly straightforward to the porting illustrated in
+graph algorithms, based on Pannotia~\cite{Pannotia} and
+Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
+indicate how many of the original number of barriers are changed to
+resizing barriers, and how many variables need to be transmitted.  In
+all cases all barriers were converted, except for bc which contains
+barriers deeper in the call stack, a case not supported by our
+prototype. Still, these barriers could in principle also be
+converted. The remaining 2 algorithms are workstealing applications:
+each required the addition of $\offerfork$ and $\offerkill$ at the start
+of the main loop, and no variables needed to be transmitted.  For all
+these examples the porting to cooperative kernels was similarly
+straightforward to the porting illustrated in
 \mysec\ref{sec:programmingguidelines} for our running examples, and did
 not require reshaping of the original algorithm. Most graph come with 2
-different input data sets, leading to 13 application/data set pairs in total.
+different data sets as input, leading to 13 application/input pairs in
+total.
 
@@
 For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
-maximum slowdown across all applications and data sets, with averages and maxima computed over 10 runs per benchmark. For the maximum slowdowns, we indicate which
-application and data set was responsible.  The slowdown is below 2 even
-in the worst case, and closer to 1 on average. We consider these results
-as encouraging, especially since the performance of our prototype could
-clearly be improved upon if cooperative scheduling was implemented
-directly in the GPU driver.
+maximum slowdown across all applications and inputs, with averages and
+maxima computed over 10 runs per benchmark. For the maximum slowdowns,
+we indicate which application and input was responsible. The slowdown is
+below 2 even in the worst case, and closer to 1 on average. We consider
+these results as encouraging, especially since the performance of our
+prototype could clearly be improved upon if cooperative scheduling was
+implemented directly in the GPU driver.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Polishing implementation.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
   columns=fullfelxible,
-  xleftmargin=2em,
   %commentstyle=\rmfamily\itshape,
@@
 kernels, our framework must track the state of each workgroup and
-provide a channel of communication from the scheduler to
+provide a channel of communication from the scheduler workgroup to
 workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
-we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available}, or the id of the kernel that the workgroup is currently executing.
+we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available} or the id of the kernel that the workgroup is currently executing.
 The scheduler can then send cooperative
-kernels a \emph{resource message}, commanding workgroups executing
-the kernel to exit at $\offerkill$, or 
+kernels a \emph{resource message}, commanding workgroups to exit at $\offerkill$, or 
 spawn additional workgroups at $\offerfork$. Thus, the
 scheduler context needs a communication channel for each cooperative
-kernel. In our case, the communication channel is implemented using
-global atomic variables.
-
-\paragraph{Launching kernels}
+kernel. We implement the communication channels using atomic variables in global memory.
+
+\paragraph{Launching kernels and managing workgroups}
 
 To launch a kernel, the host sends a data packet to the GPU
-consists of: a kernel to execute, the inputs to the kernel, the
+scheduler consisting of: a kernel to execute, the inputs to the kernel, the
 number of workgroups that should execute the kernel, and a flag indicating whether the kernel
-is cooperative. In our implementation, this communication channel
-between host and device is achieved through fine-grained SVM atomics.
-
-\paragraph{Dynamically managing workgroups}
-On receiving a data packet from the host
+is cooperative. In our implementation, this host-device communication channel is built using fine-grained SVM atomics.
+
+On receiving a data packet
 describing a kernel launch $K$, the GPU scheduler must make decisions
@@
 workgroups from the available workgroups to execute kernel $K$ with
-kernel context $\KC$.  Note that if the new kernel $K$ is itself a cooperative kernel, the scheduler could decide to provide $K$ with fewer than $N$ active workgroups initially.
-
-If a cooperative kernel $K_c$ is executing with fewer workgroups than it initially requested, the scheduler may decide to give $K_c$
-additional workgroups if $K_c$ calls $\offerfork()$. To do this, the scheduler asynchronously signals $K_c$ through $K_c$'s
+kernel context $\KC$.  (Note that if the new kernel $K$ is itself a cooperative kernel, the scheduler would be free to provide $K$ with fewer than $N$ active workgroups initially.)
+
+If a cooperative kernel $K_c$ is executing with fewer workgroups than it initially requested, the scheduler may decide make extra workgroups available
+to $K_c$, to be obtained next time $K_c$ calls $\offerfork$.  To do this, the scheduler asynchronously signals $K_c$ through $K_c$'s
 channel to indicate the number of
@@
 
-\paragraph{Safe implementation}
+\paragraph{Accounting for concurrency}
 
@@
 occur at a new kernel launch between the scheduler and a workgroup
-calling $\offerfork()$.
-
-The scheduler context provides an \emph{assignment mutes} to protecting concurrent calls to
-$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup is only permitted to exit if the scheduler currently holds the assignment mutex; that is, the scheduler is trying to obtain work
-groups to send to a task.  Conversely, on calling $\offerfork$ a workgroup can only spawn new workgroups if the assignment mutex is not held by the scheduler.
-
-Concurrent calls to $\offerkill$ are acceptable because only the workgroup with the highest id may exit, automatically ensuring mutual exclusion for exit-related modifications.
-
-\paragraph{Example}
-
-\ADComment{Give most attention to interesting performance and safety
-  issues, e.g.\ reverse ticket lock, efficient barrier, the query
-  functions...}
-
-\ADComment{The host and device components of the scheduler, the use of
-  SVM, and the host API for scheduling both sorts of kernel.}
-
-\HEComment{I consider that the reason why we need SVM fine grain and
-  atomics is discussed in the implementation section}
-
-\ADComment{Give details of how host and device interact.}
-
-\ADComment{Host-device interaction through SVM.  Make clear that this is probably the most aggressive use of OpenCL 2.0 ever.}
+calling $\offerfork$.
+
+The scheduler context provides an \emph{assignment mutes} to protect concurrent calls to
+$\offerfork$ and $\offerkill$.
+On calling $\offerkill$, a workgroup is only permitted to exit if the scheduler workgroup currently holds the assignment mutex; that is, the scheduler is trying to obtain work
+groups to send to a task.
+Conversely, on calling $\offerfork$ a workgroup can only spawn new workgroups if the scheduler workgroup does not hold the assignment mutex.
+Thus the assignment mutex, only ever acquired by the scheduling workgroup, \emph{indirectly} provides mutual exclusion between a pairs of workgroups executing $\offerkill$ and $\offerfork$ concurrently.
+\ADComment{Please check previous -- I think I've got the coolness!}
+
+Concurrent calls to $\offerkill$ are acceptable because only the
+workgroup with the highest id may exit, automatically ensuring mutual
+exclusion for exit-related modifications.
 

### 2016-11-15 write-ups/PLDI/paper.tex
minor

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \emph{heavy} we used WebGL demos from the Chrome
-experiments\footnote{\url{https://www.chromeexperiments.com/}} running
+experiments\footnote{\url{https://www.chromeexperiments.com}} running
 in a recent version of Firefox.
@@
 % the Paper Planes\footnote{\url{https://paperplanes.world}} and
-% Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
+% Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope}}
 % shaders running in a recent version of Firefox.
@@
 We now assess the performance impact of multitasking between a
-coopoerative and non-cooperative kernel, and on scheduling
+cooperative and non-cooperative kernel, and on scheduling
 responsiveness.  The performance is measured with respect to the
@@
 
-Our experiments span 13 pairs of cooperative kernels and data sets, 3
+Our experiments span 13 pairs of cooperative kernels and inputs, 3
 classes of non-cooperative kernel workloads, 4 quantities of workgroups
@@
 
-To work around compiler crash and compiler hang bugs in Intel's current OpenCL drivers, we had to apply what semantics-preserving changes to the code generated by \kernelmerge{} for a few applications configurations.
-We could not obtain results for some configurations where the non-cooperative kernel asks
-for all but one workgroups (namely bc, bfs and l-sssp on all chips, and
-color on Iris), and could not get the game application to run at all on HD520; these configurations led to 
-catastrophic issues such as display freezes or blue screens; we also attribute these to bugs in Intel's current drivers.
+To work around compiler crash and compiler hang bugs in Intel's current
+OpenCL drivers, we had to apply some semantics-preserving changes to the
+code generated by \kernelmerge{} for a few applications configurations.
+We could not obtain results for some configurations where the
+non-cooperative kernel asks for all but one workgroups (namely bc, bfs
+and l-sssp on all chips, and color on Iris), and could not get the game
+application to run at all on HD520; these configurations led to
+catastrophic issues such as display freezes or blue screens; we also
+attribute these to bugs in Intel's current drivers.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Finished pass over implementation.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
+\vspace{-3mm}
 \begin{abstract}
@@
 
+\vspace{-1mm}
 \paragraph{A degree of fairness: occupancy-bound execution}
@@
 
+\vspace{-1mm}
 \paragraph{Resistance to guaranteeing occupancy-bound execution}
@@
 
+\vspace{-1mm}
 \paragraph{Our proposal: cooperative kernels}
@@
 In summary, our main contributions are:
-
-\vspace{-1mm}
-\begin{itemize}
-
-\item \emph{Cooperative kernels}, an extended GPU programming model that support the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}).
-
-\item A proof-of-concept portable implementation of cooperative
+\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{proof-of-concept implementation} of cooperative
   kernels on top OpenCL 2.0
-  (\mysec\ref{sec:implementation}).
-
-\item Experimental results assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).
-
-\end{itemize}
-\vspace{-1mm}
+  (\mysec\ref{sec:implementation}); \emph{experimental results} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).
 
@@
 baked into the implementation. \TSComment{note here that these
-  implementations require no cfork or ckill outside of the barrier}
-
-We consider barriers which have the master/slave model
-(e.g.\ \cite{XF10}). That is, one workgroup (the master) collects signals
+  implementations require no cfork or ckill outside of the barrier} \ADComment{Ignore until camera-ready?}
+
+We consider barriers that use the master/slave model
+(e.g.\ \cite{XF10}): one workgroup (the master) collects signals
 from the other workgroups (the slaves) indicating that they have
@@
 leave. \TSComment{Need to discuss new active workgroup number that
-is stable across iterations}
+is stable across iterations} \ADComment{Again, ignore until camera-ready?}
 
@@
 waits for $M$ slaves (the number of active workgroups), which may
-decrease, (based on the $\offerkill$ instructions executed by the
-slaves), but will not increase. Once the master observes $M$ slaves
-have arrived, it knows all other workgroups are in the kernel are
-spinning, waiting to be released. The master executes $\offerfork$,
-where there is a conditional for the potentially newly spawned threads
-to join the other slaves waiting to be released. Finally, the master
+decrease due to $\offerkill$ calls by the
+slaves, but will not increase. Once the master observes that $M$ slaves
+have arrived, it knows that all other workgroups are spinning, waiting to be released. The master executes $\offerfork$,
+and the statement immediately following this $\offerfork$ is a conditional that forces newly spawned workgroups to join the
+slaves in waiting to be released. Finally, the master
 releases all the slaves: the original slaves and the new slaves that
@@
 
-While this barrier implementation works, it has its flaws. Workgroups
-only execute $\offerkill$ once per global barrier call. Depending on
+This barrier implementation is suboptimal becuase workgroups only
+execute $\offerkill$ once per global barrier call and, depending on
 the order in which workgroups arrive at the barrier, it is possible
-that only one workgroup is killed per barrier call which could lead to
-a high time for the scheduler to gather the sufficient amount of
-workgroups to execute another kernel concurrently.
-
-We observe if we provide an additional scheduler functionality for a
-workgroup executing a cooperative kernel, we can greatly reduce the
-gather time. We call the new function {\tt query()}. The semantics
-simply return the number of workgroups that the scheduler requires to
-kill from the cooperative kernel. A new barrier can now be implented
-as follows: (1) Master waits for all slaves; (2) Master forks and sets
-new workgroups as slaves. (3) Master calls {\tt query()} and gets
-value $W$; (4) master releases slaves and broadcasts the value from
-{\tt query} to slaves. (5) Workgroups with id greater than $M-W$ spin
-on $\offerkill$, as they know from {\tt query()} that the scheduler
-needs them to exit. We show in \mysec\ref{sec:responsiveness} that the barrier using
+that only one workgroup is killed per barrier call.  This may not
+allow the scheduler to gather workgroups required to execute another
+task in a timely fashion.
+
+We can greatly reduce the gather time by providing a new $\keyword{query}()$
+function for a workgroup executing a cooperative kernel, which returns the number of workgroups that the scheduler needs to obtained
+from the cooperative kernel.  
+%
+The programmer cannot call $\keyword{query}$ directly;
+it is only invoked from within the resizing barrier implementation.
+%
+A resizing barrier can now be implemented
+as follows: (1) the master waits for all slaves to arrive; (2) the master calls $\offerfork$ and
+commands the new workgroups to be slaves; (3) the master calls $\keyword{query}$, obtaining a value $W$;
+(4) the master releases the slaves, broadcasting the value $W$ to them; (5) workgroups with ids larger than $M-W$ spin, calling 
+$\offerkill$ repeatedly until the scheduler claims them---we know from $\keyword{query}$ that the scheduler will eventually do so.
+We show in \mysec\ref{sec:responsiveness} that the barrier using
 query greatly reduces the gather time in practice.
@@
 
-We have proposed \emph{cooperative kernels}, a small set of extensions
-to the GPUP programming model that allow long-running, blocking
-kernels to be fairly scheduled and to share GPU resources with other
-workloads via multitasking.  Our prototype implementation uses a
-megakernel to mock up the environment in which a cooperative kernel
-would run if native support for our programming model was provided.
-Our experimental results over three Intel GPUs show that the
-programming model is a good fit for the needs of current irregular
-algorithms that operate on graphs and use work stealing.  The
+We have proposed \emph{cooperative kernels}, a small set of GPU
+programming extensions that allow long-running, blocking kernels to be
+fairly scheduled and to share GPU resources with other workloads.
+Experimental results using our megakernel-based prototype over three
+Intel GPUs shows that the model is a good fit for the needs of current
+GPU-accelerated irregular algorithms.  The
 performance that could be gained through a native implementation with
@@
 
-There are three main avenues for future work at present: seeking out
-additional classes of irregular algorithms that would benefit from GPU
-acceleration using cooperartive kernels, investigating providing
-native driver support for cooperative kernels in an open source
-graphics driver, and integrating cooperative kernels into the back-end
-of a compiler that generates irregular OpenCL from a higher-level DSL,
-e.g.\ an OpenCL version of a recent compiler for graph algorithms that
-targets CUDA~\cite{DBLP:conf/oopsla/PaiP16}.  Longer-term, as more
-vendors adopt the advanced features of OpenCL that our approach
-requires, we will be in a position to assess the programming model
-across a wider range of devices.
+There are three main avenues for future work at present: seeking
+additional classes of irregular algorithms to which the model might
+(be extended to) apply (to), investigating native driver support for
+cooperative kernels via open source graphics drivers, and integrating
+cooperative kernels into template- and compiler-based programming
+models for graph algorithms on
+GPUs~\cite{SreesCompetition,DBLP:conf/oopsla/PaiP16}.  As more vendors
+adopt the advanced features of OpenCL that our approach requires, we
+will be in a position to assess the programming model across a wider
+range of devices.
 

### 2016-11-15 write-ups/PLDI/paper.tex

@@
 \usepackage{url}                  % format URLs
+\urlstyle{same}                   % URL use same font as regular text
+                                  %   (regular font is less wide than tt)
 \usepackage{listings}          % format code
@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{Pannotia} and Lonestar~\cite{BNP12}
-GPU application suites, using global barriers.  We indicate how many of
-the original number of barriers are changed to resizing barriers, and
-how many variables need to be transmitted.  In all cases all barriers
-were converted, except for bc which contains barriers deeper in the call
-stack, a case not supported by our prototype. Still, these barriers
-could in principle also be converted. The remaining 2 algorithms are
-workstealing applications: each required the addition of $\offerfork$
-and $\offerkill$ at the start of the main loop, and no variables needed
-to be transmitted.  For all these examples the porting to cooperative
-kernels was similarly straightforward to the porting illustrated in
+graph algorithms, based on Pannotia~\cite{Pannotia} and
+Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
+indicate how many of the original number of barriers are changed to
+resizing barriers, and how many variables need to be transmitted.  In
+all cases all barriers were converted, except for bc which contains
+barriers deeper in the call stack, a case not supported by our
+prototype. Still, these barriers could in principle also be
+converted. The remaining 2 algorithms are workstealing applications:
+each required the addition of $\offerfork$ and $\offerkill$ at the start
+of the main loop, and no variables needed to be transmitted.  For all
+these examples the porting to cooperative kernels was similarly
+straightforward to the porting illustrated in
 \mysec\ref{sec:programmingguidelines} for our running examples, and did
 not require reshaping of the original algorithm. Most graph come with 2
-different data sets, leading to 13 application/data set pairs in total.
+different data sets as input, leading to 13 application/input pairs in
+total.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Rewrite results description for responsiveness

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
-\myfiglong\ref{fig:fine-grained-timing} reports the measure for three
-configurations and represents the gather and execution time of the
-non-cooperative kernel w.r.t. the quantity of workgroups allocated to
-it. Times are in milliseconds, and a logarithmic scale is used since
-gather times tend to be much smaller than execution times. The
-horizontal grey lines indicates the desired rate for non-cooperative
-kernels.
-
-Responsiveness of the cooperative scheduling is measured by the gather
-time. In all examples, we observe that the gather time grows as we ask
-for more workgroups for the non-cooperative kernel: this is expected,
-since the scheduler must wait for more workgroups to call $\offerkill$
-or to reach a resizing barrier. Still, notice that the query barrier
-leads to better and stable gather times for graph applications, since it
-can collect all needed workgroups at the first time they pass through
-the barrier. Gather time also application-dependent since it is related
-to how often the cooperative kernel workgroups hit $\offerkill$ or
-$\resizingglobalbarrier$.
+\myfiglong\ref{fig:fine-grained-timing} reports, on three
+configurations, the gather and execution time of the non-cooperative
+kernel w.r.t. the quantity of workgroups allocated to it. A logarithmic
+scale is used for time since gather times tend to be much smaller than
+execution times. The horizontal grey lines indicates the desired rate
+for non-cooperative kernels.
+
+Let's focus on the left figure which illustrates a work stealing
+example: when the non-cooperative kernel is given only one workgroup, it
+execution is so long that it cannot permit to terminate on time for a
+screen refresh. The gather time is very good though, since the scheduler
+only needs to collect one workgroup. As more workgroups are allocated to
+the non-cooperative kernel, it can benefits from more computational
+resources and thus lower its execution time: here the non-cooperative
+kernel becomes fast enough with a quarter (resp. half) of available
+workgroups for light (resp. heavy) graphics workload. Inversely, the
+gather time augments since the scheduler must collect more and more
+workgroups from the cooperative kernel.
+
+On the middle and right figures, which depict graph algorithms, we find
+again the same trend: as the number of workgroups allocated to the
+non-cooperative task grows, the non-cooperative kernel execution time
+lowers and the gather time raises. When the regular resizing barrier is
+used, gather time can become quite long and exceed the execution time.
+However, the scheduling responsiveness can benefits from the query
+barrier: when this barrier is used, the gather time becomes almost
+stable w.r.t. the number of workgroups that needs to be collected, since
+the scheduler collects all needed workgroups as soon as they hit a
+resizing barrier.
 

### 2016-11-15 write-ups/PLDI/paper.tex
added graph and some fixes to section 4

@@
 
-Concurrent calls to $\offerfork$ must be protected by a mutex
-to avoid racing on the
-available workgroups. The same conflict over available workgroups can
-occur at a new kernel launch between the scheduler and a workgroup
-calling $\offerfork$.
-
-The scheduler context provides an \emph{assignment mutes} to protect concurrent calls to
-$\offerfork$ and $\offerkill$.
-On calling $\offerkill$, a workgroup is only permitted to exit if the scheduler workgroup currently holds the assignment mutex; that is, the scheduler is trying to obtain work
-groups to send to a task.
-Conversely, on calling $\offerfork$ a workgroup can only spawn new workgroups if the scheduler workgroup does not hold the assignment mutex.
-Thus the assignment mutex, only ever acquired by the scheduling workgroup, \emph{indirectly} provides mutual exclusion between a pairs of workgroups executing $\offerkill$ and $\offerfork$ concurrently.
-\ADComment{Please check previous -- I think I've got the coolness!}
+Concurrent calls to $\offerfork$ must be protected by a mutex to avoid
+racing on the available workgroups. The same conflict over available
+workgroups can occur at a new kernel launch between the scheduler and
+a workgroup calling $\offerfork$. Thus, these activities are protected
+by a mutex, called the \emph{assignment mutex} provided in the scheduler
+context.
+
+The assignment mutex additionally protects concurrent calls to
+$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup
+is only permitted to exit if the scheduler workgroup currently holds
+the assignment mutex; that is, the scheduler is trying to obtain work
+groups to send to a task.  Conversely, on calling $\offerfork$ a
+workgroup can only spawn new workgroups if the scheduler workgroup
+does not hold the assignment mutex.  Thus the assignment mutex, only
+ever acquired by the scheduling workgroup, \emph{indirectly} provides
+mutual exclusion between a pairs of workgroups executing $\offerkill$
+and $\offerfork$ concurrently.
 
@@
 workgroup with the highest id may exit, automatically ensuring mutual
-exclusion for exit-related modifications.
+exclusion for exit-related state changes.
 
@@
 only one call to a global barrier with $\offerfork$ and $\offerkill$
-baked into the implementation. \TSComment{note here that these
-  implementations require no cfork or ckill outside of the barrier} \ADComment{Ignore until camera-ready?}
+baked into the implementation.
 
@@
 all the slaves, it replies with a signal saying that the slaves may
-leave. \TSComment{Need to discuss new active workgroup number that
-is stable across iterations} \ADComment{Again, ignore until camera-ready?}
+leave.
 
@@
 
+\begin{figure}
+\includegraphics[width=\columnwidth]{heavy.pdf} 
+\caption{Blah}\label{fig:blah}
+\end{figure}
+
 \section{Related Work}\label{sec:relatedwork}

### 2016-11-15 write-ups/PLDI/paper.tex
Added citation.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 models for graph algorithms on
-GPUs~\cite{SreesCompetition,DBLP:conf/oopsla/PaiP16}.  As more vendors
+GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.  As more vendors
 adopt the advanced features of OpenCL that our approach requires, we

### 2016-11-15 write-ups/PLDI/paper.tex
Edit responsiveness results

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 gather time augments since the scheduler must collect more and more
-workgroups from the cooperative kernel.
-
-On the middle and right figures, which depict graph algorithms, we find
-again the same trend: as the number of workgroups allocated to the
-non-cooperative task grows, the non-cooperative kernel execution time
-lowers and the gather time raises. When the regular resizing barrier is
-used, gather time can become quite long and exceed the execution time.
-However, the scheduling responsiveness can benefits from the query
-barrier: when this barrier is used, the gather time becomes almost
-stable w.r.t. the number of workgroups that needs to be collected, since
-the scheduler collects all needed workgroups as soon as they hit a
-resizing barrier.
-
-\HEComment{missing results for all but one in bfs, discussed in expe
-setup, do we need to say it here ?}
+workgroups.
+
+The middle and right figures, which depict graph algorithms, expose the
+same trend: as the number of non-cooperative workgroups grows, the
+execution time lowers and the gather time raises.\footnote{The right
+figure has only 3 points per line: recall that for bfs there are no runs
+with ``all but one'' non-cooperative workgroups}
+%
+The gather time is higher on the right figure: the rmat22 input graph is
+rather wide than deep, so the graph algorithm reaches resizing barriers
+less often than for the ecology input of the middle figure for
+instance. The scheduler has thus fewer opportunities to collect
+workgroups and gather time increases. Nonetheless, scheduling
+responsiveness can benefit from the query barrier: when used, this
+barrier let the scheduler collects all needed workgroups as soon as they
+hit a resizing barrier. As we can see on both the middle and right
+figure, the gather time of the query barrier is almost stable w.r.t. the
+number of workgroups that needs to be collected.
 
@@
 models for graph algorithms on
-GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.  As more vendors
+GPUs~\cite{SreesCompetition,DBLP:conf/oopsla/PaiP16}.  As more vendors
 adopt the advanced features of OpenCL that our approach requires, we

### 2016-11-15 write-ups/PLDI/paper.tex
Add discussion on performance

@@
 
-\paragraph{Results}
+\paragraph{Responsiveness}
 
@@
 \end{figure}
+
+\paragraph{Performance}
+\myfig~\ref{fig:} reports the overhead brought by the scheduling of
+non-cooperative kernels over the cooperative kernel execution time
+(geometric mean over all experiences), along with the period at which
+non-cooperative kernels can be scheduled. Non-cooperative kernels
+running on a single workgroup lead to almost no overhead, but the period
+is far too high to meet the needs of graphics workloads. As more
+workgroups are dedicated to non-cooperative kernels, they execute
+quickly enough to be scheduled at the expected period. Yet, as expected,
+this increase the execution time overhead of the cooperative
+kernel. Still, in average heavy workloads meet their period for an
+overhead less than 1.4, which we consider reasonable, while light and
+medium workloads meet their period with only a very light overhead on
+the cooperative kernel.
 

### 2016-11-15 write-ups/PLDI/paper.tex

@@
 so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
-do not mandate fair scheduling, and GPU schedulers are inherently
+do not mandate fair scheduling, and GPU schedulers are
 unfair in practice.  Current approaches avoid this issue by exploiting
-scheduling quirks particular to today's GPUs in a manner that does not
+scheduling quirks of today's GPUs in a manner that does not
 allow the GPU to be shared with other workloads (such as graphics
@@
 fairly scheduled, and multitasking is supported via a small set of
-language extensions that the kernel and scheduler use to cooperate.
+language extensions through which the kernel and scheduler cooperate.
 We describe the semantics of the cooperative kernels programming model
@@
 HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
-workgroups, and current GPUs do not
-provide fair scheduling in practice.  Roughly speaking, each workgroup
+workgroups, and current GPU scheduler are unfair in practice.  Roughly speaking, each workgroup
 executing a GPU kernel is mapped to a hardware \emph{compute
@@
 The simplest way for a GPU driver to handle 
-the case where more workgroups are launched than there are compute
+more workgroups being launched than there are compute
 units is via an \emph{occupancy-bound}
-execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
+execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where, once a workgroup has commenced execution on a
 compute unit (it has become \emph{occupant}), the workgroup has
-exclusive access to the compute unit until it has finished execution.
+exclusive access to the compute unit until it finishes execution.
 Experiments suggest that the occupancy-bound execution model is widely
@@
 throttling would require being able to suspend the execution of a
-workgroup, power down the compute unit to which the workgroup was
-bound, and schedule the workgroup for later execution on a different
-compute unit.
+workgroup, powering down the compute unit to which the workgroup was
+bound, something which the occupancy-bound execution model does not allow.
 
@@
 (including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
-and various engineers from GPU vendors), is that (1) for the above reasons, GPU vendors will
-not commit to the occupancy-bound execution model that they currently
-implement,
-yet (2) GPU vendors will not guarantee fair scheduling, due to the
-runtime overhead associated with supporting full preemption of
+and various engineers from GPU vendors), is that GPU vendors (1) will
+not commit to the occupancy-bound execution model they currently
+implement, for the above reasons,
+yet (2) will not guarantee fair scheduling due to the
+runtime cost of preempting
 workgroups.  Vendors instead wish to
@@
 sacrifice itself to the scheduler.  This indicates that the workgroup
-would ideally continue executing, but that if required the
-scheduler may destroy the workgroup for good.  The cooperative kernel
+would ideally continue executing, but that if required, the
+scheduler may destroy the workgroup for good; the cooperative kernel
 must be prepared to deal with either scenario.
@@
 \item Cooperative kernels can be implemented over the occupancy-bound
-  execution model that current GPUs provide: our proof-of-concept implementation requires
+  execution model that current GPUs provide: our prototype implementation requires
   no additional hardware or driver support.  Thus the model should be eay for vendors to implement directly.
@@
 In summary, our main contributions are:
-\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{proof-of-concept implementation} of cooperative
+\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{prototype implementation} of cooperative
   kernels on top OpenCL 2.0
-  (\mysec\ref{sec:implementation}); \emph{experimental results} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).
+  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).
 

### 2016-11-15 write-ups/PLDI/paper.tex

@@
 
-\lstset{basicstyle=\tt,numbers=none}
+\lstset{basicstyle=\small\tt,numbers=none}
 \begin{lstlisting}
@@
 resizing barrier, so that from a programmer's perspective, $M$ changes atomically.
-The middle $\globalbarrier$ ensures that forking occurs before killing, so that while the barrier may kill a workgroup with id $i$, it should not also fork a new workgroup with id $i$.
-That is, workgroups $[0, \textrm{min}(M, M') - 1]$ should be left intact.
-
-Because $\resizingglobalbarrier$ can be implemented in the above
-manner, we do not regard it \emph{conceptually} as a primitive of our
-programming model.  However, we discuss in
-\mysec\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
-implemented significantly more efficiently if it is treated as a
-primitive operation that interacts directly with the scheduler.
+The middle $\globalbarrier$ ensures that forking occurs before killing, so that workgroups $[0, \textrm{min}(M, M') - 1]$ are left intact.
+
+Because $\resizingglobalbarrier$ can be implemented as above, we do not regard it \emph{conceptually} as a primitive of our
+model.  However, in
+\mysec\ref{sec:resizingbarrier} we show how a resizing barrier can be
+implemented significantly more efficiently through direct interaction with the scheduler.
 
@@
 
-We discuss the main semantic change induced by our programming
-model---that the number of workgroups executing a kernel varies
-dynamically---and illustrate the new language constructs and the
-impact of this semantic change by adapting the work stealing and graph
-traversal algorithms of \mysec\ref{sec:openclexamples} to use
-cooperative kernels.
-
-\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
+%We discuss the main semantic change induced by our programming
+%model---that the number of workgroups executing a kernel varies
+%dynamically---and illustrate the new language constructs and the
+%impact of this semantic change by adapting the work stealing and graph
+%traversal algorithms of \mysec\ref{sec:openclexamples} to use
+%cooperative kernels.
+
+\paragraph{A changing workgroup count}  Unlike in regular OpenCL,
 the value returned by $\getnumgroups()$ is not fixed during the
-lifetime of a cooperative kernel: it corresponds to the active group count $M$ and may change due to workgroups executing
-$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Being a function of $\getnumgroups()$, the value returned by $\getglobalsize()$
-is also subject to change.
-A cooperative kernel must therefore be written in a manner that is
+lifetime of a cooperative kernel: it corresponds to the active group count $M$, which changes as workgroups execute
+$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  The value returned by $\getglobalsize()$
+is similarly subject to change.
+A cooperative kernel must thus be written in a manner that is
 robust to changes in the values returned by these functions.
@@
 In general, their volatility means that use of these functions should be avoided.
-However, the situation is more stable if a cooperative kernel does not directly
-issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
-only calls to $\resizingglobalbarrier$ can cause the number of active
-workgroups to change.  At any point
+However, the situation is more stable if a cooperative kernel does not call
+$\offerkill$ and $\offerfork$ directly, so that only $\resizingglobalbarrier$ can affect the number of active
+workgroups.  Then, at any point
 during execution, the threads of a kernel are executing between some
 pair of resizing barrier calls, which we call a \emph{resizing barrier
-  interval} (here we consider the kernel entry and exit points
-conceptually to be special cases of resizing barriers).  The number of
-workgroups executing the kernel is thus constant within each resizing
-barrier interval, i.e.\ the result returned by repeated calls to
-$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
-%
-This guarantee can be exploited by algorithms that perform strided
-data processing, which we illustrate below using the graph traversal
-example.
+  interval} (considering the kernel entry and exit points
+conceptually to be special cases of resizing barriers).  The active workgroup count is constant within each resizing
+barrier interval, so that 
+$\getnumgroups()$ and $\getglobalsize()$ return stable values during such intervals.
+%
+As we illustrate below for graph traversal, this can be exploited by algorithms that perform strided
+data processing.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Yet another pass over intro.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
-\paragraph{Responsiveness}
+\paragraph{Results}
 
@@
 gather time augments since the scheduler must collect more and more
-workgroups.
-
-The middle and right figures, which depict graph algorithms, expose the
-same trend: as the number of non-cooperative workgroups grows, the
-execution time lowers and the gather time raises.\footnote{The right
-figure has only 3 points per line: recall that for bfs there are no runs
-with ``all but one'' non-cooperative workgroups}
-%
-The gather time is higher on the right figure: the rmat22 input graph is
-rather wide than deep, so the graph algorithm reaches resizing barriers
-less often than for the ecology input of the middle figure for
-instance. The scheduler has thus fewer opportunities to collect
-workgroups and gather time increases. Nonetheless, scheduling
-responsiveness can benefit from the query barrier: when used, this
-barrier let the scheduler collects all needed workgroups as soon as they
-hit a resizing barrier. As we can see on both the middle and right
-figure, the gather time of the query barrier is almost stable w.r.t. the
-number of workgroups that needs to be collected.
+workgroups from the cooperative kernel.
+
+On the middle and right figures, which depict graph algorithms, we find
+again the same trend: as the number of workgroups allocated to the
+non-cooperative task grows, the non-cooperative kernel execution time
+lowers and the gather time raises. When the regular resizing barrier is
+used, gather time can become quite long and exceed the execution time.
+However, the scheduling responsiveness can benefits from the query
+barrier: when this barrier is used, the gather time becomes almost
+stable w.r.t. the number of workgroups that needs to be collected, since
+the scheduler collects all needed workgroups as soon as they hit a
+resizing barrier.
+
+\HEComment{missing results for all but one in bfs, discussed in expe
+setup, do we need to say it here ?}
 
@@
 \end{figure}
-
-\paragraph{Performance}
-\myfig~\ref{fig:} reports the overhead brought by the scheduling of
-non-cooperative kernels over the cooperative kernel execution time
-(geometric mean over all experiences), along with the period at which
-non-cooperative kernels can be scheduled. Non-cooperative kernels
-running on a single workgroup lead to almost no overhead, but the period
-is far too high to meet the needs of graphics workloads. As more
-workgroups are dedicated to non-cooperative kernels, they execute
-quickly enough to be scheduled at the expected period. Yet, as expected,
-this increase the execution time overhead of the cooperative
-kernel. Still, in average heavy workloads meet their period for an
-overhead less than 1.4, which we consider reasonable, while light and
-medium workloads meet their period with only a very light overhead on
-the cooperative kernel.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Trimming: losing some paragraph headings.

@@
 employ either work stealing or operate on graph data structures, and placing our new constructs follows a common, easy-to-follow pattern in each case.
-
-Like the simplified example of \myfig\ref{fig:workstealing-cooperative}, the work stealing algorithms have a transactional flavour
+%
+The work stealing algorithms have a transactional flavour
 and require little or no state to be carried between transactions.
-Thus the point at which a workgroup is ready to process a new task is a natural point to place $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.
-
-The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
+The point at which a workgroup is ready to process a new task is a natural place for $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.
+%
+\myfiglong\ref{fig:cgraphsearch} is simpler than, but representative of,
 most level-by-level graph algorithms.
@@
 the graph algorithm, the next level could be processed by more or
-fewer workgroups, something which $\resizingglobalbarrier$
-facilitates.  As in \myfig\ref{fig:cgraphsearch}, some level-specific state must be transmitted to new workgroups.
-
-Our practical experience suggests that cooperative kernels are a good
-fit for at least these two classes of irregular algorithms, which is
-of course partly because we were inspired by these algorithms in
-designing the model.
+fewer workgroups, which $\resizingglobalbarrier$
+facilitates.  Some level-specific state must be transmitted to new workgroups.
+
+Our experience is that cooperative kernels are a good
+fit for at least these two classes of irregular algorithms, in part because they inspired the model.
 
@@
 
-\paragraph{Sufficient $\offerkill$ calls}
+%\paragraph{Sufficient $\offerkill$ calls}
 
@@
 %
-Using our work stealing example: in
-\myfig\ref{fig:workstealing-cooperative} a workgroup offers itself to
+Using our work stealing example: a workgroup offers itself to
 the scheduler after processing each task.  If tasks are sufficiently
@@
 to $\offerkill$.
-
-Getting this non-functional requirement right depends on the
-constraints of the tasks to be co-scheduled with a cooperative kernel,
-on properties of the cooperative kernel itself, and on the efficiency
-of the GPU that is used for execution.  In \mysec\ref{sec:sizingnoncoop} we conduct
+%
+Getting this non-functional requirement right is GPU- and application-dependent.
+In \mysec\ref{sec:sizingnoncoop} we conduct
 experiments to understand the response rate that would be required to
-co-schedule graphics rendering with a cooperative kernel in order to
-achieve a smooth frame rate, and we evaluate responsiveness in
-practice in \mysec\ref{sec:responsiveness}.
-
-\paragraph{Sufficient workgroups for the cooperative kernel}
-
-Recall that the cooperative kernel would ideally utilise the $N$
-workgroups requested upon kernel launch, in order to conduct its
-processing efficiently.  The scheduler should thus aim to provide the
-cooperative kernel $N$ workgroups if other constraints allow it,
+co-schedule graphics rendering with a cooperative kernel, maintaining a smooth frame rate.
+
+%\paragraph{Sufficient workgroups for the cooperative kernel}
+
+Recall that, on launch, the cooperative kernel requests $N$ workgroups.
+The scheduler should thus aim to provide $N$ workgroups if other constraints allow it,
 by accepting an $\offerkill$ only if a compute unit is required for another
 task, and responding positively to $\offerfork$ calls if compute units are available.  Our
-implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
+implementation (\mysec\ref{sec:implementation}) is as
 generous as possible to the cooperative kernel in this regard.
 
-Note that the above guidelines are not hard-and-fast rules.  For
-example, a smart scheduler might respond selectively to $\offerkill$
+These guidelines are not hard-and-fast rules, and e.g.\ a smart scheduler might respond selectively to $\offerkill$
 and $\offerfork$ calls by predicting future high-priority workloads
@@
 
-A goal of coopeative kernels is to allow traditional, non-blocking GPU
-tasks (e.g.\ graphics shaders in OpenGL, Vulkan and DirectX, and regular data-parallel computational kernels)
-to be co-scheduled with blocking irregular algorithms.  It is
-important to note that our new model has no effect on the manner in
-which these traditional GPU workloads are programmed.
-
+Our new model has no effect on the manner in which traditional GPU
+workloads are programmed, such as graphics shaders in OpenGL, Vulkan
+and DirectX, and regular data-parallel computational kernels.
+%
 Furthermore, if our cooperative language extensions are
@@
 case)---then a cooperative kernel will behave just like its
-non-cooperative counterpart would behave.  As a result, a developer
+non-cooperative counterpart would behave.  A developer
 for whom today's occupancy-bound execution model suffices can upgrade

### 2016-11-15 write-ups/PLDI/paper.tex
update results together with Tyler

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 \end{figure}
+
+\paragraph{Performance}
+\myfig~\ref{fig:} reports the overhead brought by the scheduling of
+non-cooperative kernels over the cooperative kernel execution time
+(geometric mean over all applications and inputs on the Iris chip, other
+chips look similar\HEComment{We can put the graphs of ther chips in
+appendix}), along with the period (aritmetic mean over all applications
+and inputs) at which non-cooperative kernels can be scheduled. We show
+results for the three workloads listed in
+\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
+period graph correspond to the period goals of the workloads: the higher
+(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for the
+light (resp. medium and heavy) workload.
+
+Non-cooperative kernels running a workload with a single workgroup leads
+to almost no overhead, but the period is far too high to meet the needs
+of any of our three workloads. For example, a heavy workload averages a
+period of 939 ms. As more workgroups are dedicated to non-cooperative
+kernels, they execute quickly enough to be scheduled at the expected
+period. More concretely, for the light and medium workloads, a quarter
+of the workgroups executing the non-cooperative kernel are able to meet
+their goal period (70 and 40 ms resp.). However, a quarter of the
+workgroups are not sufficient to meet the goal for the heavy workload
+(giving a mean period of 88 ms). Finally, if half of the workgroups are
+allocated to the non-cooperative kernel, the heavy workload gets
+averages \~10\% over its goal period (mean of 44 ms). If asking for all
+but one, the heavy workload can reach its goal period.
+
+Yet, as expected, allocating more non-cooperative workgroups increases
+the execution time overhead of the cooperative kernel. Still the heavy
+workloads meet their period (by allocating all but one workgroup for
+non-cooperative kernels) for an overhead less than 1.4 on average. Light
+and medium workloads meet their period with only a negligable overhead
+of the cooperative kernel (less than 1.03 slowdown).
 

### 2016-11-15 write-ups/PLDI/paper.tex
Transform url footnotes into citations

@@
 and a non-cooperative kernel and merges them to produce a megakernel
-with an embedded GPU scheduler.  We implemented \kernelmerge{} using the Clang LibTooling framework,\footnote{\url{http://clang.llvm.org/docs/LibTooling.html}}
-leveraging recent Clang support for OpenCL 2.0~\cite{ClangOpenCL20}.
+with an embedded GPU scheduler.  We implemented \kernelmerge{} using the
+Clang LibTooling framework~\cite{clang}, leveraging recent Clang support
+for OpenCL 2.0~\cite{ClangOpenCL20}.
 
@@
 
-Among the GPUs available to us, four met our
-requirements: Intel HD 520, HD 5500 and Iris 6100, and AMD Radeon R7.
-However, AMD's Linux drivers for OpenCL suffer from a known defect
-whereby long-running kernels lead to defunct processes that the OS
-cannot kill.\footnote{See \url{http://developer.amd.com/download/AMD_APP_SDK_Release_Notes_Developer.pdf}}  We observed this on our AMD platform with the latest
-driver, and simultaneously obtained alarming results on this platform
-when using SVM atomics: modifying a kernel to include a simple
-host/device handshake through SVM slowed down kernel execution by an
-order of magnitude in some cases.  This behaviour, combined with the
+Among the GPUs available to us, four met our requirements: Intel HD 520,
+HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux drivers
+for OpenCL suffer from a known defect whereby long-running kernels lead
+to defunct processes that the OS cannot
+kill~\cite{amd-defunct-process}. We observed this on our AMD platform
+with the latest driver, and simultaneously obtained alarming results on
+this platform when using SVM atomics: modifying a kernel to include a
+simple host/device handshake through SVM slowed down kernel execution by
+an order of magnitude in some cases.  This behaviour, combined with the
 defunt processes issue, lead us to suspect that the support we require
@@
 \emph{heavy} we used WebGL demos from the Chrome
-experiments\footnote{\url{https://www.chromeexperiments.com}} running
+experiments~\cite{chrome-experiments} running
 in a recent version of Firefox.
@@
 operating systems, such as RISC
-OS.\footnote{\url{http://www.riscos.info/index.php/Preemptive_multitasking}}
+OS~\cite{risc-os-multitasking}.
 Many programming languages offer cooperative multitasking via

### 2016-11-15 write-ups/PLDI/paper.tex
Shortening.

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 
-The threads executing a kernel are divided into a number of \emph{workgroups}, each consisting of a fixed number of threads.  Builtin functions
-$\getlocalid()$ and $\getgroupid()$ return a thread's local id within
-its workgroup and the id of the workgroup, respectively.\footnote{For simplicity we assume 1D kernels, which captures all the
-irregular algorithms we know of, using
-e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.
+The threads of a kernel are divided into \emph{workgroups}, each consisting of a fixed number of threads.  Functions
+$\getlocalid$ and $\getgroupid$ return a thread's local id within
+its workgroup and the id of the workgroup.\footnote{For simplicity we assume 1D kernels, which captures all the
+irregular algorithms we know of, and write
+e.g.\ $\getglobalid()$ for $\getglobalid(0)$.
 }  The number
-of threads per workgroup and the number of workgroups are obtained via
-$\getlocalsize()$ and $\getnumgroups()$.  A thread's local and global
-id are related via the equation $\getglobalid() = \getgroupid() \times
+of threads per workgroup and number of workgroups are obtained via
+$\getlocalsize$ and $\getnumgroups$.  A thread's local and global
+id are related by $\getglobalid() = \getgroupid() \times
 \getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize() = \getlocalsize()\times\getnumgroups()$.
@@
 \paragraph{Memory spaces and memory model}
-An OpenCL kernel has access to four memory spaces.  \emph{Shared virtual
+A kernel has access to four memory spaces.  \emph{Shared virtual
 memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
-kernel.  Each workgroup has its own portion of \emph{local} memory for fast intra-workgroup communication.  Every thread has
-an allocation of very fast \emph{private} memory, where
-function-local variables are allocated.
+kernel.  Each workgroup has a portion of \emph{local} memory for fast intra-workgroup communication.  Every thread has
+a portion of very fast \emph{private} memory for 
+function-local variables.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Worked on experiments.

@@
 spent using the GPU to render the display between kernel calls and the
-rate at which the OS requires the GPU for display rendering,
+period at which the OS requires the GPU for display rendering,
 respectively.
@@
 70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
-the medium and heavy workloads, the 40ms rate corresponds to graphics
+the medium and heavy workloads, the 40ms period corresponds to graphics
 rendering occuring 25 times per second, which coincides with the human
@@
 
-We now assess the performance impact of multitasking between a
-cooperative and non-cooperative kernel, and on scheduling
-responsiveness.  The performance is measured with respect to the
-performance of each kernel run in sequence. Responsiveness, i.e., how
+We now assess the performance impact and responsiveness of
+multitasking between a long-running cooperative kernel and a series of
+short, non-cooperative kernel launches.  Performance is measured by
+comparing the time for multitasked execution with the time taken to
+perform all kernel launches in sequence.  Responsiveness, i.e.\ how
 fast a non-cooperative kernel can be scheduled, depends not only on
 the scheduler implementation, but also on how often the cooperative
-kernel offer workgroups to be killed.
+kernel offers workgroups to be killed.
 
@@
 that aims to simulate the intensity of a graphics rendering workload,
-according to the 3 classes listed in \mysec~\ref{sec:sizingnoncoop}. In
+according to the three classes of workload discussed in \mysec\ref{sec:sizingnoncoop}. In
 practice, we use a matrix multiplication kernel as the non-cooperative
@@
 appropriate execution duration.  We conduct separate runs where we vary
-the number of workgroup slots requested by the non-cooperative kernel,
+the number of workgroups requested by the non-cooperative kernel,
 considering the cases where one, a quarter, a half, and all-but-one, of
@@
 code generated by \kernelmerge{} for a few applications configurations.
-We could not obtain results for some configurations where the
-non-cooperative kernel asks for all but one workgroups (namely bc, bfs
+Driver bugs also meant that we could not obtain results for some configurations where the
+non-cooperative kernel asks for all-but-one workgroups (namely bc, bfs
 and l-sssp on all chips, and color on Iris), and could not get the game
 application to run at all on HD520; these configurations led to
-catastrophic issues such as display freezes or blue screens; we also
-attribute these to bugs in Intel's current drivers.
+severe failures such as machine freezes and blue screens.
 
@@
 % cooperative kernels that returned invalid results when the
-% non-cooperative kernel takes all but one of the cooperative kernel
+% non-cooperative kernel takes all-but-one of the cooperative kernel
 % workgroups, we decided to ignore these results.
@@
 configurations, the gather and execution time of the non-cooperative
-kernel w.r.t. the quantity of workgroups allocated to it. A logarithmic
+kernel w.r.t. the quantity of workgroups allocated to it. 
+\ADComment{Not clear: are the times for light and heavy averages?} A logarithmic
 scale is used for time since gather times tend to be much smaller than
-execution times. The horizontal grey lines indicates the desired rate
+execution times. The horizontal grey lines indicates the desired period
 for non-cooperative kernels.
-
-Let's focus on the left figure which illustrates a work stealing
-example: when the non-cooperative kernel is given only one workgroup, it
-execution is so long that it cannot permit to terminate on time for a
-screen refresh. The gather time is very good though, since the scheduler
+These graphs show a representative sample of our results; the full set of graphs for all configurations is provided in Appendix~\ref{appendix:extraresponsiveness} (supplementary material).
+
+Let us first focus on the left graph, which illustrates a work stealing
+example executing on the Iris GPU.  When the non-cooperative kernel is given only one workgroup, its
+execution is so long that it cannot complete within the period required for a screen refresh. The gather time is very good though, since the scheduler
 only needs to collect one workgroup. As more workgroups are allocated to
-the non-cooperative kernel, it can benefits from more computational
-resources and thus lower its execution time: here the non-cooperative
-kernel becomes fast enough with a quarter (resp. half) of available
-workgroups for light (resp. heavy) graphics workload. Inversely, the
-gather time augments since the scheduler must collect more and more
+the non-cooperative kernel, its execution time benefits from more computational
+resources: here the non-cooperative
+kernel becomes fast enough with a quarter (resp.\ half) of available
+workgroups for light (resp.\ heavy) graphics workload. Inversely, the
+gather time increases since the scheduler must collect more and more
 workgroups.
 
-The middle and right figures, which depict graph algorithms, expose the
-same trend: as the number of non-cooperative workgroups grows, the
-execution time lowers and the gather time raises.\footnote{The right
+The middle and right graphs show results for graph algorithms on the
+other two GPUs.  \ADComment{Added this; is it right?}  These
+algorithms use barriers, and we experimented with the regular and
+query barrier implementations described in \mysec\ref{ }.  The
+results shown are for executions that use the more efficient query barrier, except that we also show the gather time associated with the regular barrier for comparison.
+The graphs show a similar trend to the left-most graph: as the number of non-cooperative workgroups grows, the
+execution time decreases and the gather time increases.\footnote{The right
 figure has only 3 points per line: recall that for bfs there are no runs
-with ``all but one'' non-cooperative workgroups}
+with ``all-but-one'' non-cooperative workgroups.}
 %
@@
 less often than for the ecology input of the middle figure for
-instance. The scheduler has thus fewer opportunities to collect
+instance. The scheduler thus has fewer opportunities to collect
 workgroups and gather time increases. Nonetheless, scheduling
 responsiveness can benefit from the query barrier: when used, this
-barrier let the scheduler collects all needed workgroups as soon as they
+barrier lets the scheduler collect all needed workgroups as soon as they
 hit a resizing barrier. As we can see on both the middle and right
-figure, the gather time of the query barrier is almost stable w.r.t. the
+figure, the gather time of the query barrier is almost stable w.r.t.\ the
 number of workgroups that needs to be collected.
@@
 \includegraphics[width=\columnwidth]{heavy.pdf} 
-\caption{Blah}\label{fig:blah}
+\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative tasks execute}\label{fig:performance}
 \end{figure}
 
-\paragraph{Performance}
-\myfig~\ref{fig:} reports the overhead brought by the scheduling of
-non-cooperative kernels over the cooperative kernel execution time
-(geometric mean over all applications and inputs on the Iris chip, other
-chips look similar\HEComment{We can put the graphs of ther chips in
-appendix}), along with the period (aritmetic mean over all applications
-and inputs) at which non-cooperative kernels can be scheduled. We show
+\paragraph{Performance} \ADComment{I don't think it's clear how overhead is computed; we need to make it clear that it is a slow-down.  Let's discuss.}
+\myfiglong\ref{fig:performance} reports, for the Iris GPU, the overhead brought by the scheduling of
+non-cooperative kernels over the cooperative kernel execution time (geometric mean over all applications and inputs), along with the period (aritmetic mean over all applications
+and inputs) at which non-cooperative kernels can be scheduled.
+The results for the other GPUs are similar and are graphed in Appendix~\ref{appendix:extraperformance} (supplementary material).
+ We show
 results for the three workloads listed in
@@
 
-Non-cooperative kernels running a workload with a single workgroup leads
-to almost no overhead, but the period is far too high to meet the needs
+Co-scheduling non-cooperative kernels that compute their workload using a single workgroup leads to almost no overhead, but the period is far too high to meet the needs
 of any of our three workloads. For example, a heavy workload averages a
@@
 allocated to the non-cooperative kernel, the heavy workload gets
-averages \~10\% over its goal period (mean of 44 ms). If asking for all
-but one, the heavy workload can reach its goal period.
-
+averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
+%
 Yet, as expected, allocating more non-cooperative workgroups increases
-the execution time overhead of the cooperative kernel. Still the heavy
-workloads meet their period (by allocating all but one workgroup for
-non-cooperative kernels) for an overhead less than 1.4 on average. Light
+the execution time overhead of the cooperative kernel.
+
+Still the heavy
+workloads meet their period (by allocating all-but-one workgroup for
+non-cooperative kernels), incurring a slow down of less than 1.4$\times$ on average. Light
 and medium workloads meet their period with only a negligable overhead
-of the cooperative kernel (less than 1.03 slowdown).
+of the cooperative kernel (less than a 1.03$\times$ slowdown on average).
+
+Overall we are encouraged by our experimental findings, which provide a lower bound on potential performance of our coperative kernels model: implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.
 
@@
 
-\section{Full Set of Graphs for Responsiveness Experiments}
-
-\ADComment{TODO: we need this.  Probably not the right title.}
+\section{Graphs for Multitasking Experiments}
+
+In \mysec\ref{sec:responsiveness} we showed a selection of graphs to illustrate our
+experimental results related to performance and responsiveness.  Here
+we present the full set of graphs.
+
+\subsection{Responsiveness graphs}
+
+We present the full set of graphs exemplified by the examples in
+\myfig\ref{fig:fine-grained-timing}; for completeness we reproduce those graphs here too.
+
+\ADComment{Tyler: show all the graphs, perhaps grouped by GPU.}
+
+\subsection{Performance graphs}
+
+\myfiglong\ref{fig:performance} presents data on the performance impact and period associated with multitasking cooperative and non-cooperative workloads, for the Iris GPU.
+
+\ADComment{Tyler: say something like ``Corresponding graphs for the XXX and YYY GPUs are shown in \myfigs\ref{ } and~\ref{ }, respectively.'', adding the graphs.}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Various.

@@
 algorithms use barriers, and we experimented with the regular and
-query barrier implementations described in \mysec\ref{ }.  The
+query barrier implementations described in \mysec\ref{sec:resizingbarrier}.  The
 results shown are for executions that use the more efficient query barrier, except that we also show the gather time associated with the regular barrier for comparison.
@@
 
-\subsection{Responsiveness graphs}
+\subsection{Responsiveness graphs}\label{appendix:extraresponsiveness}
 
@@
 
-\subsection{Performance graphs}
+\subsection{Performance graphs}\label{appendix:extraperformance}
 

### 2016-11-15 write-ups/PLDI/paper.tex
Edit Sec 1 and 2 to squeeze space

--- 2016-11-15 write-ups/PLDI/paper.tex
@@
 algorithms on GPUs.  These algorithms are typically \emph{blocking},
-so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
+so they require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
 do not mandate fair scheduling, and GPU schedulers are
@@
 of work to be processed is unknown ahead of time and may be determined
-by the computation itself.  There is growing interest in accelerating such algorithms on GPUs, particularly algorithms that
+by the computation itself.  There is growing interest in accelerating
+such algorithms on GPUs, particularly the ones that
 traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
@@
 Second, \emph{energy throttling} is
-an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
-power down some compute units if the battery level is low or if
-the user puts the device into a low power state.  Implementing energy
+an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to
+power down some compute units if the battery level is low.
+Implementing energy
 throttling would require being able to suspend the execution of a
 workgroup, powering down the compute unit to which the workgroup was
-bound, something which the occupancy-bound execution model does not allow.
+bound, something that the occupancy-bound execution model does not allow.
 
@@
 sacrifice itself to the scheduler.  This indicates that the workgroup
-would ideally continue executing, but that if required, the
-scheduler may destroy the workgroup for good; the cooperative kernel
+would ideally continue executing, but that the
+scheduler may preempt the workgroup; the cooperative kernel
 must be prepared to deal with either scenario.
@@
 additional workgroups join the computation, a workgroup can execute
-$\offerfork$.  This indicates that the cooperative
+$\offerfork$ to indicates that the
 kernel is prepared to proceed with the existing set of workgroups, but
-is able to take advantage of having one or more additional workgroups
-commence execution directly after the $\offerfork$ program point.
+is able to benefit from one or more additional workgroups
+commencing execution directly after the $\offerfork$ program point.
 
 The use of $\offerfork$ and $\offerkill$ creates a contract between
-the scheduler and (the programmer of the) cooperative kernel.
+the scheduler and the (programmer of the) cooperative kernel.
 Functionally, the
@@
 kernel must ensure that $\offerkill$ is executed frequently enough
-that the scheduler can accommodate soft-real time constraints,
+such that the scheduler can accommodate soft-real time constraints,
 e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
@@
   execution model that current GPUs provide: our prototype implementation requires
-  no additional hardware or driver support.  Thus the model should be eay for vendors to implement directly.
+  no additional hardware or driver support.  Thus the model should be easy for vendors to implement directly.
 
@@
 context of this paper.  A kernel is written in OpenCL C, based on C99.
-All threads executing a kernel execute the same entry function with
-identical arguments.  A thread can call $\getglobalid()$
+All threads executing a kernel start at the same entry function with
+identical arguments.  A thread can call $\getglobalid$
 to obtain its unique id, to access distinct data or follow different control flow paths to other threads.
@@
 irregular algorithms we know of, and write
-e.g.\ $\getglobalid()$ for $\getglobalid(0)$.
+e.g.\ $\getglobalid$ for $\getglobalid(0)$.
 }  The number
@@
 $\getlocalsize$ and $\getnumgroups$.  A thread's local and global
-id are related by $\getglobalid() = \getgroupid() \times
-\getlocalsize() + \getlocalid()$.  The total number of threads executing the kernel is given by $\getglobalsize() = \getlocalsize()\times\getnumgroups()$.
+id are related by $\getglobalid = \getgroupid \times
+\getlocalsize + \getlocalid$.  The total number of threads executing the
+kernel is given by $\getglobalsize = \getlocalsize \times \getnumgroups$.
 
@@
 stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
-  for values written by work-group B without deadlock provided ... A
+  for values written by work-group B without deadlock provided ... (if) A
   comes after B in work-group flattened ID order}. This is not sufficient to support blocking algorithms that use
@@
     if (t) {
-      // task processing may create more work
 (*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
@@
 of GPUs, each workgroup has an associated queue from which it
-obtains tasks to process, and to which it stores newly created
-tasks. If a workgroup's queue is empty, the workgroup tries to \emph{steal} a task from an other
-queue. Previous work addresses the feasibility and efficiency of work
-stealing on GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}.
-
-\myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
-kernel that uses a work stealing queue. Each thread receives a pointer to the task
-queues, in global memory, initialized by the
+obtains tasks to process, and to which it stores new
+tasks. If a workgroup's queue is empty, the workgroup tries to \emph{steal} a task from another
+queue. Previous work covers work stealing feasibility and efficiency on GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}.
+
+\myfiglong\ref{fig:workstealing} shows a simplified version of a work
+stealing kernel. Each thread receives a pointer to the
+task queues, in global memory, initialized by the
 host to contain the initial tasks. Each thread identifies
-the workgroup it belongs to via $\getgroupid()$
-(line~\ref{line:wksteal:getgroupid}), which is used as a queue id and used to access the relevant task queue. The
+its workgroup via $\getgroupid()$
+(line~\ref{line:wksteal:getgroupid}), which is used as a queue id to access the relevant task queue. The
 $\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
@@
 $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
-the \emph{master thread}, will try to pop a
-task from the workgroup's queue, or steal a task from a different
+the \emph{master thread}, tries to pop a
+task from the workgroup's queue, or to steal a task from a different
 queue. The master thread uses local memory inside
 $\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
-all threads of its workgroup. If a task was obtained, then all threads
-of the workgroup participate in the processing of the task (line~\ref{line:wksteal:processtask}).
+all threads of its workgroup. If a task was obtained, then all the
+workgroup threads participate in the task processing (line~\ref{line:wksteal:processtask}).
 This may lead to the creation of further tasks, which
-the master thread will push to the workgroup's queue.
+the master thread pushes to the workgroup's queue.
 
@@
 skeleton of a frontier-based graph traversal algorithm; such algorithms have
-been shown to be execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}. 
+been shown to execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}. 
 The kernel is
 given three arguments in global memory: a graph structure, and two
-arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
+arrays of graph nodes. Initially, $\keyword{n0}$ contains the
 starting nodes to process. Private variable $\keyword{level}$ indicates the frontier level currently being
@@
 nodes to process (line~\ref{line:graph:iterate}). At each frontier,
-the nodes to be processed are evenly distributed between the executing
-threads. This is done through \emph{stride} based processing. 
+the nodes to be processed are evenly distributed between
+threads through \emph{stride} based processing. 
 %
@@
 At this point, the GPU threads must wait for all other threads to
-finishing processing the frontier. To achieve
+finish processing the frontier. To achieve
 this, we use a global barrier construct
@@
 is incremented. The threads use another global barrier to wait until the output node is
-reset (line~\ref{line:graph:gb2}), after which the threads can continue to the next frontier.
+reset (line~\ref{line:graph:gb2}), after which they continue to the next frontier.
 

### 2016-11-15 write-ups/PLDI/paper.tex
Edit

@@
 
-We now assess the performance impact and responsiveness of
-multitasking between a long-running cooperative kernel and a series of
-short, non-cooperative kernel launches.  Performance is measured by
-comparing the time for multitasked execution with the time taken to
-perform all kernel launches in sequence.  Responsiveness, i.e.\ how
-fast a non-cooperative kernel can be scheduled, depends not only on
-the scheduler implementation, but also on how often the cooperative
-kernel offers workgroups to be killed.
+We now assess the responsiveness of multitasking between a long-running cooperative kernel and a series of
+short, non-cooperative kernel launches, and the performance impact of multitasking on the cooperative kernel.
 
@@
 \myfiglong\ref{fig:fine-grained-timing} reports, on three
-configurations, the gather and execution time of the non-cooperative
+configurations, the average gather and execution times for the non-cooperative
 kernel w.r.t. the quantity of workgroups allocated to it.
-\ADComment{Not clear: are the times for light and heavy averages?} A logarithmic
+A logarithmic
 scale is used for time since gather times tend to be much smaller than

### 2016-11-15 write-ups/PLDI/paper.tex
Edit

@@
 The middle and right graphs show results for graph algorithms on the
-other two GPUs.  \ADComment{Added this; is it right?}  These
+other two GPUs.  These
 algorithms use barriers, and we experimented with the regular and
-query barrier implementations described in \mysec\ref{sec:resizingbarrier}.  The
-results shown are for executions that use the more efficient query barrier, except that we also show the gather time associated with the regular barrier for comparison.
+query barrier implementations described in \mysec\ref{sec:resizingbarrier}.
+The execution times for the non-cooperative task are averaged across all runs, including with both types of barrier.
+We show separately the average gather time associated with each type of barrier.
 The graphs show a similar trend to the left-most graph: as the number of non-cooperative workgroups grows, the

### 2016-11-15 write-ups/PLDI/paper.tex
Tweaks.

@@
 \subsection{Sizing Non-Cooperative Kernels}\label{sec:sizingnoncoop}
-
-\ADComment{Refer back to nonfunctional requirements section, \mysec\ref{sec:nonfunctional}.}
 
@@
 
-\ADComment{Can we remark a bit more on the outliers?}
-
 \subsection{Multitasking via Cooperative Scheduling}\label{sec:responsiveness}
@@
 
-\paragraph{Performance} \ADComment{I don't think it's clear how overhead is computed; we need to make it clear that it is a slow-down.  Let's discuss.}
+\paragraph{Performance} 
 \myfiglong\ref{fig:performance} reports, for the Iris GPU, the overhead brought by the scheduling of
-non-cooperative kernels over the cooperative kernel execution time (geometric mean over all applications and inputs), along with the period (aritmetic mean over all applications
-and inputs) at which non-cooperative kernels can be scheduled.
+non-cooperative kernels over the cooperative kernel execution time.
+This is the slowdown associated with running the cooperative kernel in the presence of multitasking, vs.\ running the cooperative kenrel in isolation (geometric mean over all applications and inputs).
+We also show the period at which non-cooperative kernels can be scheduled (aritmetic mean over all applications
+and inputs).
 The results for the other GPUs are similar and are graphed in Appendix~\ref{appendix:extraperformance} (supplementary material).
@@
 
-Still the heavy
+Still, the heavy
 workloads meet their period (by allocating all-but-one workgroup for
@@
 
-Overall we are encouraged by our experimental findings, which provide a lower bound on potential performance of our coperative kernels model: implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.
+Overall we are very encouraged by these experimental findings, especially since they provide a lower bound on potential performance of our coperative kernels model.
+Implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.
 

### 2016-11-16 write-ups/PLDI/paper.tex
Tweaks.

@@
 for non-cooperative kernels.
-These graphs show a representative sample of our results; the full set of graphs for all configurations is provided in Appendix~\ref{appendix:extraresponsiveness} (supplementary material).
+These graphs show a representative sample of our results; the full set of graphs for all configurations is provided in Appendix~\ref{appendix:extragraphs} (supplementary material).
 
@@
 and inputs).
-The results for the other GPUs are similar and are graphed in Appendix~\ref{appendix:extraperformance} (supplementary material).
+The results for the other GPUs are similar.
  We show
@@
 
-\section{Graphs for Multitasking Experiments}
-
-In \mysec\ref{sec:responsiveness} we showed a selection of graphs to illustrate our
-experimental results related to performance and responsiveness.  Here
-we present the full set of graphs.
-
-\subsection{Responsiveness graphs}\label{appendix:extraresponsiveness}
+\section{Graphs for Multitasking Experiments}\label{appendix:extragraphhs}
 
@@
 
-\ADComment{Tyler: show all the graphs, perhaps grouped by GPU.}
-
-\subsection{Performance graphs}\label{appendix:extraperformance}
-
-\myfiglong\ref{fig:performance} presents data on the performance impact and period associated with multitasking cooperative and non-cooperative workloads, for the Iris GPU.
-
-\ADComment{Tyler: say something like ``Corresponding graphs for the XXX and YYY GPUs are shown in \myfigs\ref{ } and~\ref{ }, respectively.'', adding the graphs.}
+\ADComment{Tyler: show all the graphs.}
 

### 2016-11-16 write-ups/PLDI/paper.tex
Space.

@@
 
-We discuss our experience porting irregular algorithms to cooperative kernels (\mysec\ref{sec:portingalgorithms}) and discuss the GPUs on which we evaluate these applications (\mysec\ref{sec:gpus}).
+We discuss our experience porting irregular algorithms to cooperative kernels and describe the GPUs on which we evaluate these applications (\mysec\ref{sec:portingalgorithms}).
 For these GPUs, we report on experiments to determine non-cooperative workloads that model the requirements of various graphics rendering tasks (\mysec\ref{sec:sizingnoncoop}).
@@
 
-\subsection{Porting Irregular GPU Algorithms}\label{sec:portingalgorithms}
+\subsection{Applications and GPUs}\label{sec:portingalgorithms}
 
@@
 interesting for our multitasking approach.
-%
-These exceptions aside, we have handled all the available OpenCL
-implemenatinos of irregular algorithms that we are aware of, plus two
-easy-to-adapt CUDA algorithms (the workstealing examples).
-
-\subsection{The GPUs we Tested}\label{sec:gpus}
 
@@
 
-\section{Graphs for Multitasking Experiments}\label{appendix:extragraphhs}
+\section{Graphs for Multitasking Experiments}\label{appendix:extragraphs}
 

### 2016-11-16 write-ups/PLDI/paper.tex
Space.

@@
 
-Many irregular algorithms require synchronisation across workgroups.
-This can be achieved via a \emph{global barrier}, as illustrated in the graph traversal example of \mysec\ref{sec:openclexamples} and \myfig\ref{fig:graphsearch}.
-Because a global barrier requires fair scheduling of workgroups it cannot be safely implemented in OpenCL, CUDA or HSA, and thus is not provided as a primitive in any of these programming models.
-
-Workgroups of a \emph{cooperative} kernel \emph{are} fairly scheduled, so a
+Because workgroups of a cooperative kernel are fairly scheduled, so a
 global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
@@
 \end{lstlisting}
-\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing global barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
+\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
 \end{figure}
@@
 kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
-primitives via a per kernel \emph{context}.
-For a given kernel this records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
+primitives via a per kernel \emph{context}, which records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
 id valid for the kernel.
@@
 In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
-of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
- It is possible, however to implement the resizing with
+of a series of calls to other primitives.  It is possible, however to implement the resizing with
 only one call to a global barrier with $\offerfork$ and $\offerkill$
@@
 This barrier implementation is suboptimal becuase workgroups only
-execute $\offerkill$ once per global barrier call and, depending on
-the order in which workgroups arrive at the barrier, it is possible
-that only one workgroup is killed per barrier call.  This may not
-allow the scheduler to gather workgroups required to execute another
-task in a timely fashion.
+execute $\offerkill$ once per barrier call and, depending on
+order of arrival, it is possible
+that only one workgroup is killed per barrier call, preventing the scheduler from gathering workgroups quickly.
 

### 2016-11-16 write-ups/PLDI/paper.tex
Space

@@
 
-For ease of implementation, our current prototype supports
+For ease of implementation, our prototype supports
 multitasking a single cooperative kernel with a single non-cooperative
 kernel (though the non-cooperative kernel can be invoked many times).
-Rather than adding language support for $\transmit$ annotations, we
-assume that all variables declared at the root lexical scope of a
-cooperative kernel are $\transmit$-annotated; this allows our
+We
+assume all variables declared at the root lexical scope of a
+cooperative kernel are implicitly $\transmit$-annotated; this allows our
 implementation to piggy-back on top of Clang (see below) without
@@
 
-\begin{table}
+\begin{table}[t]
 \small
@@
 
+\begin{table}[h]
+\footnotesize
+\centering
+\begin{tabular}{ l l l l }
+Chip & HD520 & HD5500 & Iris\\
+\hline
+\# CU  & 24 & 24 & 47 \\
+Driver   & 20.19.15.4501 & 10.18.15.4281 & 20.19.15.4463 \\
+Host CPU & i3-6100U & i7-5600U & i3-5157U \\
+\end{tabular}
+\caption{The Intel GPU platforms used in our evaluation}
+\label{tab:chipstested}
+\end{table}
+
 \mytablong\ref{tab:applications} gives an overview of the 8 irregular
@@
 We thus present experimental results for the Intel GPUs, whose characteristics are summarised in \mytab\ref{tab:chipstested}.
-
-\begin{table}
-\footnotesize
-\centering
-\begin{tabular}{ l l l l }
-Chip & HD520 & HD5500 & Iris\\
-\hline
-\# CU  & 24 & 24 & 47 \\
-Driver   & 20.19.15.4501 & 10.18.15.4281 & 20.19.15.4463 \\
-Host CPU & i3-6100U & i7-5600U & i3-5157U \\
-\end{tabular}
-\caption{The Intel GPU platforms used in our evaluation}
-\label{tab:chipstested}
-\end{table}
 

### 2016-11-16 write-ups/PLDI/paper.tex
Space

@@
 algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on Pannotia~\cite{Pannotia} and
+graph algorithms, based on the Pannotia~\cite{Pannotia} and
 Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
@@
 barriers deeper in the call stack, a case not supported by our
-prototype. Still, these barriers could in principle also be
+prototype although these barriers could in principle be
 converted. The remaining 2 algorithms are workstealing applications:
@@
 straightforward to the porting illustrated in
-\mysec\ref{sec:programmingguidelines} for our running examples, and did
-not require reshaping of the original algorithm. Most graph come with 2
+\mysec\ref{sec:programmingguidelines} for our running examples. Most graph come with 2
 different data sets as input, leading to 13 application/input pairs in

### 2016-11-16 write-ups/PLDI/paper.tex
space

--- 2016-11-16 write-ups/PLDI/paper.tex
@@
 The number of \emph{active workgroups}---workgroups executing the
-kernel---is denoted $M$.  The active workgroups
-have consecutive workgroup ids in the range $[0, M-1]$.
+kernel---is denoted $M$.  Active workgroups
+have consecutive ids in the range $[0, M-1]$.
 Initially, at least one workgroup is active;
-if necessary the scheduler must postpone execution of the kernel until
-a some compute unit becomes available.
+if necessary the scheduler must postpone the kernel until
+some compute unit becomes available.
 

### 2016-11-16 write-ups/PLDI/paper.tex
Space

--- 2016-11-16 write-ups/PLDI/paper.tex
@@
 these results as encouraging, especially since the performance of our
-prototype could clearly be improved upon if cooperative scheduling was
-implemented directly in the GPU driver.
+prototype could clearly be improved upon in a native implementation.
 
@@
 (giving a mean period of 88 ms). Finally, if half of the workgroups are
-allocated to the non-cooperative kernel, the heavy workload gets
-averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
+allocated to the non-cooperative kernel, the heavy workload averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
 %

### 2016-11-16 write-ups/PLDI/paper.tex
Cutting some non-essential paragraphs.

@@
 
-Our experience is that cooperative kernels are a good
-fit for at least these two classes of irregular algorithms, in part because they inspired the model.
+%Our experience is that cooperative kernels are a good
+%fit for at least these two classes of irregular algorithms, in part because they inspired the model.
 
@@
 
-These guidelines are not hard-and-fast rules, and e.g.\ a smart scheduler might respond selectively to $\offerkill$
-and $\offerfork$ calls by predicting future high-priority workloads
-based on past workloads.
+%These guidelines are not hard-and-fast rules, and e.g.\ a smart scheduler might respond selectively to $\offerkill$
+%and $\offerfork$ calls by predicting future high-priority workloads
+%based on past workloads.
 

### 2016-11-16 write-ups/PLDI/paper.tex
Space.

@@
 To enable multitasking through cooperative kernels, the runtime (in
-our case, the mega-kernel) must track the state of all
+our case, the megakernel) must track the state of all
 workgroups, i.e.\ whether a workgroup is waiting or computing a kernel,
@@
 Enabling rendering of smooth graphics in parallel with execution of irregular algorithms is an important use case for our approach.
-Because our prototype implementation is based on a mega-kernel that takes over the entire GPU (see \mysec\ref{sec:implementation}), we cannot assess this use case directly.
-
-As a proxy for genuine graphics rendering, we devised the following
+Because our prototype implementation is based on a megakernel that takes over the entire GPU (see \mysec\ref{sec:implementation}), we cannot assess this use case directly.
+
+We devised the following
 procedure to determine non-cooperative OpenCL workloads that simulate
@@
 We designed a synthetic kernel that occupies all workgrouops of a
-given GPU for a parameterised length of time $t$.  We wrote a host
-program to repeatedly launch this kernel in an infinite loop.  We then
-searched for a maximum value for the parameter $t$ that allowed the
+GPU for a parameterised time period $t$, invoked in an infinite loop by a host application.  We then
+searched for a maximum value for $t$ that allowed the
 synthetic kernel to execute without having an observable impact on
-graphics rendering, where ``observable'' was decided manually by one
-of the paper authors.  Having found the appropriate value for $t$, we
+graphics rendering.  Using the computed value, we
 ran the host application for $X$ seconds, measuring the time $Y < X$
-dedicated to GPU execution during this period, as well as the number
+dedicated to GPU execution during this period and the number
 of kernel launches $n$ that were issued.  We used $X \geq 10$ in all

### 2016-11-16 write-ups/PLDI/paper.tex
Crunch.
updated paper graphs to use N instead of M and added all other graphs

--- 2016-11-16 write-ups/PLDI/paper.tex
@@
 The number of \emph{active workgroups}---workgroups executing the
-kernel---is denoted $M$.  Active workgroups
-have consecutive ids in the range $[0, M-1]$.
+kernel---is denoted $M$.  The active workgroups
+have consecutive workgroup ids in the range $[0, M-1]$.
 Initially, at least one workgroup is active;
-if necessary the scheduler must postpone the kernel until
-some compute unit becomes available.
+if necessary the scheduler must postpone execution of the kernel until
+a some compute unit becomes available.
 
@@
 
-Because workgroups of a cooperative kernel are fairly scheduled, so a
+Many irregular algorithms require synchronisation across workgroups.
+This can be achieved via a \emph{global barrier}, as illustrated in the graph traversal example of \mysec\ref{sec:openclexamples} and \myfig\ref{fig:graphsearch}.
+Because a global barrier requires fair scheduling of workgroups it cannot be safely implemented in OpenCL, CUDA or HSA, and thus is not provided as a primitive in any of these programming models.
+
+Workgroups of a \emph{cooperative} kernel \emph{are} fairly scheduled, so a
 global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
@@
 \end{lstlisting}
-\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
+\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing global barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
 \end{figure}
@@
 
-For ease of implementation, our prototype supports
+For ease of implementation, our current prototype supports
 multitasking a single cooperative kernel with a single non-cooperative
 kernel (though the non-cooperative kernel can be invoked many times).
-We
-assume all variables declared at the root lexical scope of a
-cooperative kernel are implicitly $\transmit$-annotated; this allows our
+Rather than adding language support for $\transmit$ annotations, we
+assume that all variables declared at the root lexical scope of a
+cooperative kernel are $\transmit$-annotated; this allows our
 implementation to piggy-back on top of Clang (see below) without
@@
 kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
-primitives via a per kernel \emph{context}, which records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
+primitives via a per kernel \emph{context}.
+For a given kernel this records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
 id valid for the kernel.
@@
 In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
-of a series of calls to other primitives.  It is possible, however to implement the resizing with
+of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
+ It is possible, however to implement the resizing with
 only one call to a global barrier with $\offerfork$ and $\offerkill$
@@
 This barrier implementation is suboptimal becuase workgroups only
-execute $\offerkill$ once per barrier call and, depending on
-order of arrival, it is possible
-that only one workgroup is killed per barrier call, preventing the scheduler from gathering workgroups quickly.
+execute $\offerkill$ once per global barrier call and, depending on
+the order in which workgroups arrive at the barrier, it is possible
+that only one workgroup is killed per barrier call.  This may not
+allow the scheduler to gather workgroups required to execute another
+task in a timely fashion.
 
@@
 
-\begin{table}[t]
+\begin{table}
 \small
@@
 
-\begin{table}[h]
+\mytablong\ref{tab:applications} gives an overview of the 8 irregular
+algorithms that we ported to cooperative kernels. Among them, 6 are
+graph algorithms, based on Pannotia~\cite{Pannotia} and
+Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
+indicate how many of the original number of barriers are changed to
+resizing barriers, and how many variables need to be transmitted.  In
+all cases all barriers were converted, except for bc which contains
+barriers deeper in the call stack, a case not supported by our
+prototype. Still, these barriers could in principle also be
+converted. The remaining 2 algorithms are workstealing applications:
+each required the addition of $\offerfork$ and $\offerkill$ at the start
+of the main loop, and no variables needed to be transmitted.  For all
+these examples the porting to cooperative kernels was similarly
+straightforward to the porting illustrated in
+\mysec\ref{sec:programmingguidelines} for our running examples, and did
+not require reshaping of the original algorithm. Most graph come with 2
+different data sets as input, leading to 13 application/input pairs in
+total.
+
+From the Lonestar suite we omit two applications that use global barriers:
+dmr has been shown to be unstable across GPU vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially short-running so that it is not
+interesting for our multitasking approach.
+
+Our prototype implementation (\mysec\ref{sec:implementation}) requires
+two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
+atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at
+present, since these vendors do not yet support OpenCL $2.0$,
+equivalent functionality is not provided by \nvidia's CUDA. Intel
+drivers meet our requirements on Windows, but not on Linux (we
+confirmed this by trying the upstream driver that requires a patched
+Linux kernel). AMD support SVM atomics only their Kaveri family of
+chips and only on Linux, requiring to compile the latest driver.
+
+Among the GPUs available to us, four met our requirements: Intel HD 520,
+HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux drivers
+for OpenCL suffer from a known defect whereby long-running kernels lead
+to defunct processes that the OS cannot
+kill~\cite{amd-defunct-process}. We observed this on our AMD platform
+with the latest driver, and simultaneously obtained alarming results on
+this platform when using SVM atomics: modifying a kernel to include a
+simple host/device handshake through SVM slowed down kernel execution by
+an order of magnitude in some cases.  This behaviour, combined with the
+defunt processes issue, lead us to suspect that the support we require
+for cooperative kernels is not yet mature enough within AMD's drivers,
+which we plan to discuss with engineers at AMD.
+
+We thus present experimental results for the Intel GPUs, whose characteristics are summarised in \mytab\ref{tab:chipstested}.
+
+\begin{table}
 \footnotesize
@@
 \end{table}
-
-\mytablong\ref{tab:applications} gives an overview of the 8 irregular
-algorithms that we ported to cooperative kernels. Among them, 6 are
-graph algorithms, based on the Pannotia~\cite{Pannotia} and
-Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
-indicate how many of the original number of barriers are changed to
-resizing barriers, and how many variables need to be transmitted.  In
-all cases all barriers were converted, except for bc which contains
-barriers deeper in the call stack, a case not supported by our
-prototype although these barriers could in principle be
-converted. The remaining 2 algorithms are workstealing applications:
-each required the addition of $\offerfork$ and $\offerkill$ at the start
-of the main loop, and no variables needed to be transmitted.  For all
-these examples the porting to cooperative kernels was similarly
-straightforward to the porting illustrated in
-\mysec\ref{sec:programmingguidelines} for our running examples. Most graph come with 2
-different data sets as input, leading to 13 application/input pairs in
-total.
-
-From the Lonestar suite we omit two applications that use global barriers:
-dmr has been shown to be unstable across GPU vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially short-running so that it is not
-interesting for our multitasking approach.
-
-Our prototype implementation (\mysec\ref{sec:implementation}) requires
-two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
-atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at
-present, since these vendors do not yet support OpenCL $2.0$,
-equivalent functionality is not provided by \nvidia's CUDA. Intel
-drivers meet our requirements on Windows, but not on Linux (we
-confirmed this by trying the upstream driver that requires a patched
-Linux kernel). AMD support SVM atomics only their Kaveri family of
-chips and only on Linux, requiring to compile the latest driver.
-
-Among the GPUs available to us, four met our requirements: Intel HD 520,
-HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux drivers
-for OpenCL suffer from a known defect whereby long-running kernels lead
-to defunct processes that the OS cannot
-kill~\cite{amd-defunct-process}. We observed this on our AMD platform
-with the latest driver, and simultaneously obtained alarming results on
-this platform when using SVM atomics: modifying a kernel to include a
-simple host/device handshake through SVM slowed down kernel execution by
-an order of magnitude in some cases.  This behaviour, combined with the
-defunt processes issue, lead us to suspect that the support we require
-for cooperative kernels is not yet mature enough within AMD's drivers,
-which we plan to discuss with engineers at AMD.
-
-We thus present experimental results for the Intel GPUs, whose characteristics are summarised in \mytab\ref{tab:chipstested}.
 
@@
 these results as encouraging, especially since the performance of our
-prototype could clearly be improved upon in a native implementation.
+prototype could clearly be improved upon if cooperative scheduling was
+implemented directly in the GPU driver.
 
@@
 (giving a mean period of 88 ms). Finally, if half of the workgroups are
-allocated to the non-cooperative kernel, the heavy workload averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
+allocated to the non-cooperative kernel, the heavy workload gets
+averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
 %
@@
 We present the full set of graphs exemplified by the examples in
-\myfig\ref{fig:fine-grained-timing}; for completeness we reproduce those graphs here too.
-
-\ADComment{Tyler: show all the graphs.}
+\myfig\ref{fig:fine-grained-timing}; for completeness we reproduce
+those graphs here too.
+
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_color_ecology.pdf} \\
+%\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_color_G3_circuit.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_mis_ecology.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_mis_G3_circuit.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_sssp_usa_road.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_sssp_rmat20.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_sssp_usa_ny.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bc_1k_128k.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bc_2k_1M.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bfs_rmat22.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bfs_usa.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_color_ecology.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_color_G3_circuit.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_mis_ecology.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_mis_G3_circuit.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_sssp_usa_road.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_sssp_rmat20.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_sssp_usa_ny.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_bc_1k_128k.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_bc_2k_1M.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_bfs_rmat22.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/iris_bfs_usa.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_color_ecology.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_color_G3_circuit.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_mis_ecology.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_mis_G3_circuit.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_sssp_usa_road.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_sssp_rmat20.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_sssp_usa_ny.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bc_1k_128k.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bc_2k_1M.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bfs_rmat22.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bfs_usa.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/ws/hd5500_octree_NA.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/ws/hd5500_connect_four_NA.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/ws/iris_octree_NA.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/ws/iris_connect_four_NA.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/ws/hd520_octree_NA.pdf} \\
 

### 2016-11-16 write-ups/PLDI/paper.tex

--- 2016-11-16 write-ups/PLDI/paper.tex
@@
 primitives for inter-workgroup synchronisation.
-%
-We present the semantics of these constructs intuitively, using English.
-In Appendix~\ref{appendix:semantics} (see supplementary material) we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.
 

### 2016-11-16 write-ups/PLDI/paper.tex
A big cut.

--- 2016-11-16 write-ups/PLDI/paper.tex
@@
 
-\paragraph{Kernel and scheduler contexts}
-
-When a workgroup executes a kernel (cooperative or otherwise), it must
-be able to query a consistent context to obtain a workgroup id value
-and the
-number of workgroups executing the kernel. For multi-kernel
-kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
-primitives via a per kernel \emph{context}, which records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
-id valid for the kernel.
+\paragraph{Scheduler contexts}
+
+%When a workgroup executes a kernel (cooperative or otherwise), it must
+%be able to query a consistent context to obtain a workgroup id value
+%and the
+%number of workgroups executing the kernel. For multi-kernel
+%kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
+%primitives via a per kernel \emph{context}, which records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
+%id valid for the kernel.
 
@@
 executing. The scheduler can use $K_c$'s channel in the scheduler context to command $K_c$ to provide $N - N_a$ workgroups via $\offerkill$.
-Once $N$ workgroups are available, the scheduler initialises a
-kernel context $\KC$ for $K$, setting the number of active workgroups
-to $N$ and mapping available workgroups (located in the scheduler
-context) to their new $\KC$ workgroup ids. The scheduler then sends $N$
-workgroups from the available workgroups to execute kernel $K$ with
-kernel context $\KC$.  (Note that if the new kernel $K$ is itself a cooperative kernel, the scheduler would be free to provide $K$ with fewer than $N$ active workgroups initially.)
+Once $N$ workgroups are available,
+% the scheduler initialises a
+%kernel context $\KC$ for $K$, setting the number of active workgroups
+%to $N$ and mapping available workgroups (located in the scheduler
+%context) to their new $\KC$ workgroup ids. 
+the scheduler then sends $N$
+workgroups from the available workgroups to execute kernel $K$
+% with
+%kernel context $\KC$.
+(If the new kernel $K$ is itself a cooperative kernel, the scheduler would be free to provide $K$ with fewer than $N$ active workgroups initially.)
 

### 2016-11-16 write-ups/PLDI/paper.tex
Cuts.
Move table.
Bang

@@
 may contain specific program points at which a workgroup could
-\emph{gracefully} leave the computation.  We give examples of these
-for work stealing and graph traversal in \mysec\ref{sec:programmingguidelines}.
+\emph{gracefully} leave the computation.
+%We give examples of these
+%for work stealing and graph traversal in \mysec\ref{sec:programmingguidelines}.
 
@@
 executes as a no-op, or accept the offer, so that execution of the workgroup ceases and
-the number of active workgroups $M$ is atomically decremented by one.  In the second case, the scheduler can assign some other task to
-the compute unit on which the workgroup was executing.
+the number of active workgroups $M$ is atomically decremented by one.  
+%In the second case, the scheduler can assign some other task to
+%the compute unit on which the workgroup was executing.
 
@@
 
-\begin{table}[h]
+\begin{table}[t]
 \footnotesize
@@
 
-Co-scheduling non-cooperative kernels that compute their workload using a single workgroup leads to almost no overhead, but the period is far too high to meet the needs
-of any of our three workloads. For example, a heavy workload averages a
+Co-scheduling non-cooperative kernels that request a single workgroup leads to almost no overhead, but the period is far too high to meet the needs
+of any of our three workloads; e.g.\ a heavy workload averages a
 period of 939 ms. As more workgroups are dedicated to non-cooperative
 kernels, they execute quickly enough to be scheduled at the expected
-period. More concretely, for the light and medium workloads, a quarter
+period. For the light and medium workloads, a quarter
 of the workgroups executing the non-cooperative kernel are able to meet
@@
 workgroups are not sufficient to meet the goal for the heavy workload
-(giving a mean period of 88 ms). Finally, if half of the workgroups are
-allocated to the non-cooperative kernel, the heavy workload averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
+(giving a mean period of 88 ms). If half of the workgroups are
+allocated to the non-cooperative kernel, the heavy workload averages $\sim$10\% over its goal period (mean of 44 ms). If all-but-one are allocated, the heavy workload reaches its goal period.
 %
 Yet, as expected, allocating more non-cooperative workgroups increases
-the execution time overhead of the cooperative kernel.
+the execution time of the cooperative kernel.
 

### 2016-11-16 write-ups/PLDI/paper.tex
Shorter conclusion.

@@
 \end{lstlisting}
-
+\vspace{-2mm}
 \caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
+\vspace{-2mm}
 \end{figure}
@@
 \end{lstlisting}
+\vspace{-2mm}
 \caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
+\vspace{-2mm}
 \end{figure}
@@
 \end{lstlisting}
+\vspace{-2mm}
 \caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
+\vspace{-2mm}
 \end{figure}
@@
 \crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
+\vspace{-2mm}
 \caption{Blocking GPU applications investigated}
 \label{tab:applications}
+\vspace{-2mm}
 \end{table}
@@
 \end{tabular}
+\vspace{-2mm}
 \caption{The Intel GPU platforms used in our evaluation}
 \label{tab:chipstested}
+\vspace{-2mm}
 \end{table}
@@
 }
+\vspace{-2mm}
 \caption{Execution time slowdown associated with cooperative scheduling mode without multitasking}
 \label{tab:overhead}
+\vspace{-2mm}
 \end{table}
@@
 \includegraphics[width=.7\columnwidth]{hd5500_bfs_rmat22.pdf}
+\vspace{-2mm}
 \caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
+\vspace{-2mm}
 \end{figure*}
@@
 \includegraphics[width=\columnwidth]{heavy.pdf} 
+\vspace{-2mm}
 \caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative tasks execute}\label{fig:performance}
+\vspace{-2mm}
 \end{figure}
@@
 dedicated driver support would be even better.
-
-There are three main avenues for future work at present: seeking
+%
+Avenues for future work include seeking
 additional classes of irregular algorithms to which the model might
-(be extended to) apply (to), investigating native driver support for
-cooperative kernels via open source graphics drivers, and integrating
+(be extended to) apply (to), investigating implementing native support
+in open source drivers, and integrating
 cooperative kernels into template- and compiler-based programming
 models for graph algorithms on
-GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.  As more vendors
-adopt the advanced features of OpenCL that our approach requires, we
-will be in a position to assess the programming model across a wider
-range of devices.
+GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.
+%  As more vendors
+%adopt the advanced features of OpenCL that our approach requires, we
+%will be in a position to assess the programming model across a wider
+%range of devices.
 

### 2016-11-16 write-ups/PLDI/paper.tex
Space.

@@
 \caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
-\vspace{-2mm}
+\vspace{-6mm}
 \end{figure}
@@
 $\resizingglobalbarrier$ the programmer indicates that the cooperative
-kernel is prepared to proceed after the barrier with fewer workgroups,
-or with additional workgroups joining the computation.
+kernel is prepared to proceed after the barrier with more or fewer workgroups.
 

### 2016-11-16 write-ups/PLDI/paper.tex
Reached page limit.

@@
 
+\vspace{-1mm}
 \paragraph{Transmitted variables}
@@
 
+\vspace{-1mm}
 \paragraph{Active workroups}
@@
 
+\vspace{-1mm}
 \paragraph{Semantics for $\offerkill$}
@@
 
+\vspace{-1mm}
 \paragraph{Semantics for $\offerfork$}
@@
 
-Notice that $k=0$ is always a legitimate choice for the number of
-workgroups to be spawned by an $\offerfork$ call, and if the number of
-active workgroups $M$ is equal to the workgroup limit $N$, $k=0$ is
-guaranteed.
-
+Notice that $k=0$ is always a valid choice for the number of
+workgroups to be spawned by $\offerfork$, and is guarateed if $M$
+is equal to the workgroup limit $N$.
+
+\vspace{-1mm}
 \paragraph{Global barriers}

### 2016-11-16 write-ups/PLDI/paper.tex
Spelling.

@@
 Notice that $k=0$ is always a valid choice for the number of
-workgroups to be spawned by $\offerfork$, and is guarateed if $M$
+workgroups to be spawned by $\offerfork$, and is guaranteed if $M$
 is equal to the workgroup limit $N$.
@@
 We emphasise that an aim of cooperative kernels is to \emph{avoid}
-depending on the occupancy-bound model.  Our protoytpe exploits this
+depending on the occupancy-bound model.  Our prototype exploits this
 model simply to allow us to experiment with current GPUs whose
@@
 \mysec\ref{sec:experiments}).  A non-mock implementation of
-coopeartive kernels would not use the megakernel approach, so we do not deem
+cooperative kernels would not use the megakernel approach, so we do not deem
 the engineering effort associated with lifting these restrictions to
@@
 arrived at the barrier and are waiting for a reply indicating that
-they may leave the barrier. Once the master has recieved a signal from
+they may leave the barrier. Once the master has received a signal from
 all the slaves, it replies with a signal saying that the slaves may
@@
 
-This barrier implementation is suboptimal becuase workgroups only
+This barrier implementation is sub-optimal because workgroups only
 execute $\offerkill$ once per barrier call and, depending on
@@
 prototype although these barriers could in principle be
-converted. The remaining 2 algorithms are workstealing applications:
+converted. The remaining 2 algorithms are work stealing applications:
 each required the addition of $\offerfork$ and $\offerkill$ at the start
@@
 an order of magnitude in some cases.  This behaviour, combined with the
-defunt processes issue, lead us to suspect that the support we require
+defunct processes issue, lead us to suspect that the support we require
 for cooperative kernels is not yet mature enough within AMD's drivers,
@@
 %
-We designed a synthetic kernel that occupies all workgrouops of a
+We designed a synthetic kernel that occupies all workgroups of a
 GPU for a parameterised time period $t$, invoked in an infinite loop by a host application.  We then
@@
 the medium and heavy workloads, the 40ms period corresponds to graphics
-rendering occuring 25 times per second, which coincides with the human
+rendering occurring 25 times per second, which coincides with the human
 persistence of vision. The 3ms execution duration of both light and
@@
 kernel, and the non-cooperative kernel execution time. We sanity-check
-whether all kernels perform their computations corretly by comparing
+whether all kernels perform their computations correctly by comparing
 their computed result with expected reference results.
@@
 non-cooperative kernels over the cooperative kernel execution time.
-This is the slowdown associated with running the cooperative kernel in the presence of multitasking, vs.\ running the cooperative kenrel in isolation (geometric mean over all applications and inputs).
-We also show the period at which non-cooperative kernels can be scheduled (aritmetic mean over all applications
+This is the slowdown associated with running the cooperative kernel in the presence of multitasking, vs.\ running the cooperative kernel in isolation (geometric mean over all applications and inputs).
+We also show the period at which non-cooperative kernels can be scheduled (arithmetic mean over all applications
 and inputs).
@@
 non-cooperative kernels), incurring a slow down of less than 1.4$\times$ on average. Light
-and medium workloads meet their period with only a negligable overhead
+and medium workloads meet their period with only a negligible overhead
 of the cooperative kernel (less than a 1.03$\times$ slowdown on average).
 
-Overall we are very encouraged by these experimental findings, especially since they provide a lower bound on potential performance of our coperative kernels model.
+Overall we are very encouraged by these experimental findings, especially since they provide a lower bound on potential performance of our cooperative kernels model.
 Implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.
@@
 Many programming languages offer cooperative multitasking via
-variations of coroutines, e.g. coroutine objects in
+variations of co-routines, e.g. co-routine objects in
 Python
@@
 kernels relatively informally, using English, to provide the intuition
-behind our programmming model.  We now back this up with a more formal
+behind our programming model.  We now back this up with a more formal
 presentation as an operational semantics for an abstract GPU
@@
 account for this, we assume that the shared state component $\sigma$
-is not simplyl a mapping from locations to values, but instead
+is not simply a mapping from locations to values, but instead
 captures all the intricacies of GPU memory that can lead to this
@@
 on competing workloads and how the scheduler has been designed to meet
-the non-functinoal requirements discussed in
+the non-functional requirements discussed in
 \mysec\ref{sec:nonfunctional}, but in general the programmer should
-always be prepared for the possiblity that a workgroup survives after
+always be prepared for the possibility that a workgroup survives after
 calling $\offerkill$.
@@
 $\offerfork$, and the workgroup gets stuck if multiple threads in a
-workgroiup reach $\offerfork$ with different subsequent statements.
+workgroup reach $\offerfork$ with different subsequent statements.
 
@@
 a set of kernel states that can arise due to memory synchronization
-taking place.  An arbitary member of this set is selected as the next
+taking place.  An arbitrary member of this set is selected as the next
 kernel state.
@@
 scheduler requires an additional compute unit, but the active thread
-with the larget id is processing some computationally intensive work
+with the largest id is processing some computationally intensive work
 and will take a while to reach $\offerkill$.  Our semantics means that
-the scheuduler cannot take advantage of the fact that another active
+the scheduler cannot take advantage of the fact that another active
 workgroup may invoke $\offerkill$ sooner.
@@
 cooperative kernel terminating early due to the programmer
-inadvertantly allowing all workgroups to be killed, and the decision
+inadvertently allowing all workgroups to be killed, and the decision
 to keep workgroup 0 alive fits well with our choice to kill workgroups
@@
 
-We opted for transmitting only selected varibles based on the
+We opted for transmitting only selected variables based on the
 observation that many of a thread's private variables are dead at the

### 2016-11-16 write-ups/PLDI/paper.tex
Fix hyphenation

--- 2016-11-16 write-ups/PLDI/paper.tex
@@
 % known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
+
+\usepackage[shortcuts]{extdash} % hyphenation, see http://stackoverflow.com/questions/2193307/how-do-i-get-latex-to-hyphenate-a-word-that-contains-a-dash#8649145
 
@@
 
-\item The model is backwards-compatible: if $\offerkill$/$\offerfork$ are ignored, a cooperative kernel will behave
+\item The model is backwards-compatible: if $\offerkill$ and $\offerfork$ are ignored, a cooperative kernel will behave
   exactly as a regular kernel does on current GPUs.
 
-\item Cooperative kernels can be implemented over the occupancy-bound
+\item Cooperative kernels can be implemented over the occupancy\-/bound
   execution model that current GPUs provide: our prototype implementation requires
@@
 host to contain the initial tasks. Each thread identifies
-its workgroup via $\getgroupid()$
+its workgroup via $\getgroupid$
 (line~\ref{line:wksteal:getgroupid}), which is used as a queue id to access the relevant task queue. The
@@
 In this case, the stride is the total number of threads, obtained via 
-$\getglobalsize()$.  Each thread processes nodes with the following sequence of indices:
+$\getglobalsize$.  Each thread processes nodes with the following sequence of indices:
 \code{$\keyword{tid}$+$\keyword{stride}$*0},
@@
 When executed by a cooperative
-kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
-active workgroups.  This is in contrast to $\getnumgroups()$ for
+kernel, $\getnumgroups$ returns $M$, the \emph{current} number of
+active workgroups.  This is in contrast to $\getnumgroups$ for
 regular kernels, which returns the fixed number of workgroups that
@@
 threads executing the cooperative kernel.  On reaching a
-$\globalbarrier()$, a thread waits until all threads have arrived at
+$\globalbarrier$, a thread waits until all threads have arrived at
 the barrier.  Once all threads have arrived, the threads may proceed
@@
 \paragraph{A changing workgroup count}  Unlike in regular OpenCL,
-the value returned by $\getnumgroups()$ is not fixed during the
+the value returned by $\getnumgroups$ is not fixed during the
 lifetime of a cooperative kernel: it corresponds to the active group count $M$, which changes as workgroups execute
-$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  The value returned by $\getglobalsize()$
+$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  The value returned by $\getglobalsize$
 is similarly subject to change.
@@
 barrier interval, so that 
-$\getnumgroups()$ and $\getglobalsize()$ return stable values during such intervals.
+$\getnumgroups$ and $\getglobalsize$ return stable values during such intervals.
 %
@@
 directly: the value of $\keyword{stride}$ obtained from
-$\getglobalsize()$ at line~\ref{line:cgraph:rechunking2} is stable
+$\getglobalsize$ at line~\ref{line:cgraph:rechunking2} is stable
 until the next resizing barrier at line~\ref{line:cgraph:resizing1},
@@
 
-We can greatly reduce the gather time by providing a new $\keyword{query}()$
+We can greatly reduce the gather time by providing a new $\keyword{query}$
 function for a workgroup executing a cooperative kernel, which returns the number of workgroups that the scheduler needs to obtained
@@
 The $\textsc{Barrier}$ rule requires that \emph{every} thread
-executing the kernel reaches $\globalbarrier()$ an identical following
-statement.  This reflects that fact that $\globalbarrier()$ is a
+executing the kernel reaches $\globalbarrier$ an identical following
+statement.  This reflects that fact that $\globalbarrier$ is a
 kernel-level function.  Each thread then skips over the
-$\globalbarrier()$ call, and the $\sync$ function is applied to yield
+$\globalbarrier$ call, and the $\sync$ function is applied to yield
 a set of kernel states that can arise due to memory synchronization
@@
 algorithm of \myfig\ref{fig:cgraphsearch} illustrates this: the
-algorithm is prepared for $\getglobalsize()$ to change after each
-resizing barrier call, but depends on the fact that $\getglobalid()$
+algorithm is prepared for $\getglobalsize$ to change after each
+resizing barrier call, but depends on the fact that $\getglobalid$
 returns a contiguous range of thread ids.

### 2016-11-16 write-ups/PLDI/paper.tex
Line break.

@@
 thread; $\offerkill$ and $\offerfork$, the key functions that enable
-cooperative scheduling; and $\globalbarrier$ and $\resizingglobalbarrier$
+cooperative scheduling; and $\globalbarrier$ and $\mathsf{resizing\_global\_}$ $\mathsf{barrier}$
+%$\resizingglobalbarrier$
 primitives for inter-workgroup synchronisation.

### 2016-11-16 write-ups/PLDI/paper.tex
updated the paper to include the graph
added updated number for geo mean of HD 5500
Last edits to add extra paragraph about no competing methods to compare against

@@
 % 
-\documentclass[numbers,nocopyrightspace,10pt]{sigplanconf}
+\documentclass[preprint,numbers,nocopyrightspace,10pt]{sigplanconf}
 % \documentclass[pldi-cameraready]{sigplanconf-pldi16}
@@
 
-The purpose of $\offerkill$ is to provide the scheduler with an
-opportunity to destroy a workgroup in order to schedule
-higher-priority tasks.  The scheduler relies on the cooperative kernel
-to execute $\offerkill$ sufficiently frequently that soft real-time
-constraints of other workloads can be met.
+The purpose of $\offerkill$ is to let the scheduler destroy a workgroup
+in order to schedule higher-priority tasks.  The scheduler relies on the
+cooperative kernel to execute $\offerkill$ sufficiently frequently that
+soft real-time constraints of other workloads can be met.
 %
@@
 the scheduler after processing each task.  If tasks are sufficiently
-fast to process then the scheduler will have ample opportunities to
-de-schedule workgroups.  But if tasks can be very time-consuming to
+fast to process then the scheduler have ample opportunities to
+de-schedule workgroups.  But if tasks are very time-consuming to
 process then it might be necessary to rewrite the algorithm so that
@@
 On receiving a data packet
-describing a kernel launch $K$, the GPU scheduler must make decisions
-about how to schedule $K$. Suppose $K$ requests
+describing a kernel launch $K$, the GPU scheduler must decide
+ how to schedule $K$. Suppose $K$ requests
 $N$ workgroups. The scheduler queries the scheduler context.  If
@@
 
-We consider barriers that use the master/slave model
-(e.g.\ \cite{XF10}): one workgroup (the master) collects signals
-from the other workgroups (the slaves) indicating that they have
-arrived at the barrier and are waiting for a reply indicating that
-they may leave the barrier. Once the master has received a signal from
-all the slaves, it replies with a signal saying that the slaves may
-leave.
+We consider barriers that use the master/slave model~\cite{XF10}: one
+workgroup (master) collects signals from the other workgroups (slaves)
+indicating that they have arrived at the barrier and are waiting for a
+reply indicating that they may leave the barrier. Once the master has
+received a signal from all slaves, it replies with a signal saying that
+they may leave.
 
@@
 
-We discuss our experience porting irregular algorithms to cooperative kernels and describe the GPUs on which we evaluate these applications (\mysec\ref{sec:portingalgorithms}).
-For these GPUs, we report on experiments to determine non-cooperative workloads that model the requirements of various graphics rendering tasks (\mysec\ref{sec:sizingnoncoop}).
-We then examine the overhead associated with moving to cooperative kernels when multitasking is \emph{not} required (\mysec\ref{sec:overhead}), as well as the responsiveness and throughput observed when a cooperative kernel is multi-tasked with non-cooperative workloads (\mysec\ref{sec:responsiveness}).
+We discuss our experience porting irregular algorithms to cooperative
+kernels and describe the GPUs on which we evaluate these applications
+(\mysec\ref{sec:portingalgorithms}).  For these GPUs, we report on
+experiments to determine non-cooperative workloads that model the
+requirements of various graphics rendering tasks
+(\mysec\ref{sec:sizingnoncoop}).  We then examine the overhead
+associated with moving to cooperative kernels when multitasking is
+\emph{not} required (\mysec\ref{sec:overhead}), as well as the
+responsiveness and throughput observed when a cooperative kernel is
+multi-tasked with non-cooperative workloads
+(\mysec\ref{sec:responsiveness}).
+
+To our knowledge, there are no competing methods against which to
+meaningfully compare our approach.  We are not aware of any existing
+platform that supports multitasking between long-running, fully-occupant
+OpenCL kernels.  Very recent \nvidia{} GPUs, to which we do not yet have
+access, support preemption for graphics
+applications~\cite{PascalWhitepaper}, but our understanding is that this
+cannot yet be leveraged from CUDA, and certainly not from OpenCL.
 
@@
 \hline
-overall geo. mean  & $1.20$ & $1.15$ & $1.08$ \\
+overall geo. mean  & $1.20$ & $1.14$ & $1.08$ \\
 overall max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
@@
 
-Recall that our mega kernel-based implementation merges the code for a
-cooperative kernel with that of the non-cooperative task to be
-co-scheduled.  This blow-up in code size can have a negative impact on
-compilation, e.g.\ due to higher register pressure, reducing the
-occupancy for the merged kernel---the maximum number of workgroups
-that the GPU driver will schedule.  This is an artifact of our
-proof-of-concept implementation strategy, and would not be a problem
-if our cooperative scheduling approach was implemented inside the GPU
-driver.  We thus launch both the original and cooperative versions of
-a kernel with the reduced occupancy bound in order to meaningfully
-compare execution times.
+Recall that our mega kernel-based implementation merges the code of a
+cooperative and a non-cooperative kernel to be co-scheduled.  This
+blow-up in code size can have a negative impact on compilation, e.g.\
+due to higher register pressure, reducing the occupancy for the merged
+kernel---the maximum number of workgroups that the GPU driver will
+schedule.  This is an artifact of our proof-of-concept implementation
+strategy, and would not be a problem if our cooperative scheduling
+approach was implemented inside the GPU driver.  We thus launch both the
+original and cooperative versions of a kernel with the reduced occupancy
+bound in order to meaningfully compare execution times.
 
@@
 \paragraph{Experimental setup}
-For a given cooperative kernel and its data set, we launch the kernel
-asynchronously and then repeatedly schedule a non-cooperative kernel
-that aims to simulate the intensity of a graphics rendering workload,
-according to the three classes of workload discussed in \mysec\ref{sec:sizingnoncoop}. In
-practice, we use a matrix multiplication kernel as the non-cooperative
-workload, with matrix dimensions tailored for each GPU to reach the
-appropriate execution duration.  We conduct separate runs where we vary
-the number of workgroups requested by the non-cooperative kernel,
-considering the cases where one, a quarter, a half, and all-but-one, of
-the total number of workgroup slots is requested.  Moreover, for the
-graph algorithms we try both the regular and query barrier
-implementations.
+For a given cooperative kernel and its input, we launch the kernel and
+then repeatedly schedule a non-cooperative kernel that aims to simulate
+the intensity of one of the three classes of graphics rendering workload
+discussed in \mysec\ref{sec:sizingnoncoop}. In practice, we use a matrix
+multiplication kernel as the non-cooperative workload, with matrix
+dimensions tailored for each GPU to reach the appropriate execution
+duration.  We conduct separate runs where we vary the number of
+workgroups requested by the non-cooperative kernel, considering the
+cases where one, a quarter, a half, and all-but-one, of the total number
+of workgroup slots is requested.  Moreover, for the graph algorithms we
+try both regular and query barrier implementations.
 
@@
 execution times. The horizontal grey lines indicates the desired period
-for non-cooperative kernels.
-These graphs show a representative sample of our results; the full set of graphs for all configurations is provided in Appendix~\ref{appendix:extragraphs} (supplementary material).
+for non-cooperative kernels.  These graphs show a representative sample
+of our results; the full set of graphs for all configurations is
+provided in Appendix~\ref{appendix:extragraphs} (supplementary material).
 
 Let us first focus on the left graph, which illustrates a work stealing
-example executing on the Iris GPU.  When the non-cooperative kernel is given only one workgroup, its
-execution is so long that it cannot complete within the period required for a screen refresh. The gather time is very good though, since the scheduler
-only needs to collect one workgroup. As more workgroups are allocated to
-the non-cooperative kernel, its execution time benefits from more computational
-resources: here the non-cooperative
-kernel becomes fast enough with a quarter (resp.\ half) of available
-workgroups for light (resp.\ heavy) graphics workload. Inversely, the
-gather time increases since the scheduler must collect more and more
-workgroups.
+example on the Iris GPU.  When the non-cooperative kernel is given only
+one workgroup, its execution is so long that it cannot complete within
+the period required for a screen refresh. The gather time is very good
+though, since the scheduler needs to collect only one workgroup.  The
+more workgroups are allocated to the non-cooperative kernels, the faster
+it can compute: here the non-cooperative kernel becomes fast enough with
+a quarter (resp.\ half) of available workgroups for light (resp.\ heavy)
+graphics workload. Inversely, the gather time increases since the
+scheduler must collect more and more workgroups.
 
@@
 %
-The gather time is higher on the right figure: the rmat22 input graph is
+the gather time is higher on the right figure: the rmat22 input graph is
 rather wide than deep, so the graph algorithm reaches resizing barriers
@@
 \vspace{-2mm}
-\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative tasks execute}\label{fig:performance}
+\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
 \vspace{-2mm}
@@
 
-Still, the heavy
-workloads meet their period (by allocating all-but-one workgroup for
-non-cooperative kernels), incurring a slow down of less than 1.4$\times$ on average. Light
-and medium workloads meet their period with only a negligible overhead
-of the cooperative kernel (less than a 1.03$\times$ slowdown on average).
+Still, heavy workloads meet their period by allocating all-but-one
+non-cooperative workgroups, incurring a slow down of less than
+1.4$\times$ on average. Light and medium workloads meet their period
+with only a negligible overhead of the cooperative kernel (less than a
+1.03$\times$ slowdown on average).
 
@@
 As our experiments demonstrate, our cooperative kernels model allows blocking algorithms
-to be upgraded to run under a fair scheduler in a manner that, with
-appropriate care, facilitates responsive multitasking.
+to be upgraded to run in a manner that facilitates responsive multitasking.
+
+\vspace{-1mm}
 
@@
 A follow-up work adds the notion of SM \emph{flushing},
-where a workgroup can be cancelled and re-scheduled
+where a workgroup can be re-scheduled
 from scratch if it has not yet committed side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
-using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs targeting data centers now support
-preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.
+using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs support
+preemption, though it is not clear whether they guarantee fairness~\cite{PascalWhitepaper}.
 
@@
 application-oblivious preemption.  There is scope for combining the approaches: if a
-cooperative kernel is insufficiently cooperative, i.e.\ it does not
-invoke $\offerkill$ enough, then hardware support for preemption could
+cooperative kernel is insufficiently cooperative, hardware support for preemption could
 be invoked to prevent the kernel from denying other tasks access to
 the GPU indefinitely.
+
+\vspace{-1mm}
 
@@
 that uses dynamic parallelism are still prone to unfair scheduling of workgroups, so this feature
-does not help solve the problem of safely deploying blocking
+does not help in deploying blocking
 algorithms on GPUs.
@@
 
+\vspace{-1mm}
+
 \section{Conclusions and Future Work}\label{sec:conclusion}
@@
 \bibliography{tyler}
+
+\iffalse
 
@@
 \includegraphics[width=.7\columnwidth]{images/barrier/hd5500_color_ecology.pdf} \\
-%\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_color_G3_circuit.pdf} \\
+\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_color_G3_circuit.pdf} \\
 \includegraphics[width=.7\columnwidth]{images/barrier/hd5500_mis_ecology.pdf} \\
@@
 
+\fi
 \end{document}

### 2017-01-29 write-ups/PLDI/rebuttal.txt
added a rebuttal.txt

Thanks for the insightful reviews and references, they are really
helpful!

* Reviewer A: thanks for the very positive review. Could you consider
championing the paper?  You wrote: "Overall, this was a very nice
piece of work ... All I could really ask for is some additional detail
and cleanup in the writing", which sounds like you could :-) We can
certainly provide this detail, taking on board points from all the
reviewers.

* Hardware support for pre-emption (reviewers C and D):

Hardware support would actually complement what we propose. First,
efficient HW pre-emption would allow cooperative kernels to be
implemented very efficiently. Second, our method for allowing the
programmer to specify smart pre-emption points would allow the driver
to invoke hardware-based pre-emption judiciously (minimising
overhead). Third, the ability to arbitrarily pre-empt when necessary
would alleviate the valid concern that a rogue cooperative kernel
might cause starvation. Fourth and finally, just because HW
pre-emption exists does not mean it will provide fairness guarantees
to software. "Async Compute", while interesting, does not make any
fairness guarantees.

We recently presented the work to the ARM Mali GPU team, who were
positive about our ideas, since their proposed hardware pre-emption will
be very coarse-grained and thus expensive.

* Chicken and egg problem:

Until pre-emption is more widely supported in GPUs, or the primitives
necessary to implement software solutions such as ours become widely
available, it is impossible to fully understand the fairness
guarantees that future HW support might provide, and the relative
overheads. Our submission is a forward-looking paper that we believe
is high risk, in that future developments in the area may provide
other alternatives, but with potentially very high value, because we
offer a platform-independent language-based solution that could be
integrated into OpenCL. A PLDI paper on topic would be a platform for
conversations around this topic in the years to come, as hardware
solutions are actually being implemented.

* Performance (reviewers C and D):

we will clarify each issue in the final version

- 1.4x slowdown: It appears you are referring to figure 5 (overhead
for N-1) and we believe that there is a misunderstanding. This is the
overhead of only the cooperative kernel while running a *heavy*
graphics workload simultaneously. Sharing the GPU between multiple
workloads *should* provide a slowdown for an individual workload. 1.4x
for sharing resources with a heavy workload is not surprising. There
is no slowdown due to multi-tasking on the duration of the
non-cooperative; only due to the number of resources executing the
task (i.e. a non-cooperative kernel run with half as many cores will
take twice as long).

- Performance slowdowns for adding our constructs (table 3): The
highest slowdowns were on the oldest chip (520) meaning that newer
chips are not suffering as much slowdown (mean of 1.08x slowdown on
the newest chip).

- Our mega-kernel approach is not how we envision our ideas being
implemented in practice. Our software solution is an upper bound on
what a driver/runtime/hardware solution could achieve. Given this, we
believe our results are encouraging.

- Acceptable performance characteristics vary between domains. We have
made the effort to implement this (which required a great deal of
engineering) and it is surely useful to have these performance
figures, to inform future research.

* Specific questions raised by the reviews

* Reviewer A:

- Question 1: Yes, the barrier we used was based on the Xiao and Feng
implementation of [44], and subsequently examined in [33].

- Question 6: Good observation; in all of our cases offer_fork and
offer_kill are paired up. None of our use cases required them being
split, although having them split provides flexibility for future
applications.

- Question 4: We envision that the HW support for the occupancy bound
execution model will continue to be available (it is straightforward
after all), but new GPU features (multi-tasking, energy throttling,
etc.) may break the occupancy bound model fairness guarantees to
programmers. Cooperative kernels are a contract between the runtime
and programmer. The runtime must guarantee to provide cooperative
kernels the occupancy bound fairness properties either by (a) actually
implementing the occupancy bound model or (b) simulating it through
pre-emption that provides sufficient fairness properties.

* Reviewer B:

- w.r.t. only allowing 2 kernels to multitask: Our use case is
graphics and compute co-existing, thus two kernels is the simplest
form of that, which we believe provides meaningful (and easy to
understand) benchmark data. Additionally, the mega-kernel approach
requires significant engineering. Because we envision our approach to
be implemented in runtime/driver/hardware, the value of an
over-engineered software prototype is unclear.

* Reviewer C:

- Question 1: as opposed to dynamic parallelism, offer_fork does not
require new workgroups be spawned (only if they are available).
Additionally, dynamic parallelism is Nvidia/CUDA exclusive. We target a
broader range of GPUs. The OpenCL equivalent, nested parallelism,
enqueues a kernel to be executed, but only after the parent finishes.

- Question 2: We believe a compiler-based approach could offer hints
(based on some kind of live variable analysis), but in general to be
very complex. This is because the number of executing workgroups
changes with offer_fork, offer_kill. Along these lines, we expect
frameworks which generate OpenCL code (e.g. [27]) to be able to
produce these automatically.

* Reviewer D:

- w.r.t. transmitting thread 0's values: The values we needed to
transmit in our application were uniform across the workgroup so it
did not matter.  You are right about bugs and we can easily change
this in the semantics and implementation.

- w.r.t. richer query semantics: See answer below; query() has some
tricky semantics that we did not want to expose to the programmer (it
should be used only inside a resizing_barrier().

- w.r.t. to the gap between query and kill: You are right, and this is
why query() is slightly trickier. Essentially, it creates a contract
between the cooperative kernel and scheduler that the number of
workgroups it specifies *must* be killed regardless of what else
happens. Because of this complexity, we did not want to expose this as
a primitive and instead just use it inside the efficient barrier.

### 2017-02-14 write-ups/FSE/paper.tex
added directory and file for FSE

--- 2016-11-16 write-ups/PLDI/paper.tex
@@
 %
-\documentclass[preprint,numbers,nocopyrightspace,10pt]{sigplanconf}
+
+\documentclass[sigconf,review, anonymous=true, acmcopyrightmode=none]{acmart}
+\acmConference[ESEC/FSE 2017]{11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{4--8 September, 2017}{Paderborn, Germany}
+
+%\documentclass[preprint,numbers,nocopyrightspace,10pt]{sigplanconf}
 % \documentclass[pldi-cameraready]{sigplanconf-pldi16}
@@
 %
-\usepackage{SIunits}            % typset units correctly
+%\usepackage{SIunits}            % typset units correctly
 \usepackage{courier}            % standard fixed width font
@@
 \usepackage{enumitem}      % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
+%\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
 % known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
@@
 \usepackage[normalem]{ulem}
-\usepackage{amsmath}
+%\usepackage{amsmath}
 \usepackage{calc}
@@
 
-\title{Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs}
-
-%
-% any author declaration will be ignored  when using 'pldi' option (for double blind review)
-%
-
-\authorinfo{}
-{\makebox{} \\
-}
-{}
-
-%\authorinfo{Tyler Sorensen}
-%{\makebox{Imperial College London, UK} \\
-%}
-%{t.sorensen15@imperial.ac.uk}
-
-%\authorinfo{Alastair F. Donaldson}
-%{\makebox{Imperial College London, UK} \\
-%}
-%{alastair.donaldson@imperial.ac.uk}
-
-\maketitle
-
-\vspace{-3mm}
+\title{Cooperative Kernels: GPU Multitasking for Blocking Algorithms}
+
 \begin{abstract}
@@
 
+%
+% any author declaration will be ignored  when using 'pldi' option (for double blind review)
+%
+
+%\authorinfo{}
+%{\makebox{} \\
+%}
+%{}
+
+%\authorinfo{Tyler Sorensen}
+%{\makebox{Imperial College London, UK} \\
+%}
+%{t.sorensen15@imperial.ac.uk}
+
+%\authorinfo{Alastair F. Donaldson}
+%{\makebox{Imperial College London, UK} \\
+%}
+%{alastair.donaldson@imperial.ac.uk}
+
+\maketitle
+
 \section{Introduction}\label{sec:intro}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{A degree of fairness: occupancy-bound execution}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{Resistance to guaranteeing occupancy-bound execution}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{Our proposal: cooperative kernels}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \begin{enumerate}
@@
 \end{enumerate}
-\vspace{-1mm}
+%\vspace{-1mm}
 
@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
-\vspace{-6mm}
+%\vspace{-6mm}
 \end{figure}
@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{Transmitted variables}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{Active workroups}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{Semantics for $\offerkill$}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{Semantics for $\offerfork$}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 \paragraph{Global barriers}
@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 \end{tabular} \\
-\vspace{.2cm}
+%\vspace{.2cm}
 \crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
-\vspace{-2mm}
 \caption{Blocking GPU applications investigated}
 \label{tab:applications}
-\vspace{-2mm}
 \end{table}
@@
 \end{tabular}
-\vspace{-2mm}
 \caption{The Intel GPU platforms used in our evaluation}
 \label{tab:chipstested}
-\vspace{-2mm}
 \end{table}
@@
 }
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Execution time slowdown associated with cooperative scheduling mode without multitasking}
 \label{tab:overhead}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{table}
@@
 \begin{figure*}
-\includegraphics[width=.7\columnwidth]{iris_octree_NA.pdf} 
-\includegraphics[width=.7\columnwidth]{hd520_mis_ecology.pdf}
-\includegraphics[width=.7\columnwidth]{hd5500_bfs_rmat22.pdf}
-\vspace{-2mm}
+\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf} 
+\includegraphics[width=.67\columnwidth]{hd520_mis_ecology.pdf}
+\includegraphics[width=.67\columnwidth]{hd5500_bfs_rmat22.pdf}
+%\vspace{-2mm}
 \caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure*}
@@
 \includegraphics[width=\columnwidth]{heavy.pdf} 
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 
-\vspace{-1mm}
+%\vspace{-1mm}
 

### 2017-02-15 write-ups/FSE/paper.tex
almost got it into the page limit and made paragraphs look good

@@
 
-\documentclass[sigconf,review, anonymous=true, acmcopyrightmode=none]{acmart}
+\documentclass[parskip=half,sigconf,review, anonymous=true, acmcopyrightmode=none]{acmart}
 \acmConference[ESEC/FSE 2017]{11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{4--8 September, 2017}{Paderborn, Germany}
@@
 \usepackage{subfig}
+%\usepackage[T1]{fontenc}
+%\usepackage[ngerman]{babel} 
+%\usepackage[latin9]{inputenc}
+%\usepackage{blindtext}
 
@@
 
+\newcommand{\cutthree}[1]{}
+
 \newcommand\crule[3][black]{\textcolor{#1}{\rule{#2}{#3}}}
+
+\makeatletter
+
+\renewcommand\paragraph{\@startsection{paragraph}{4}{\z@}%
+  {-.5\baselineskip \@plus -2\p@ \@minus -.2\p@}%
+  {-3.5\p@}%
+  {\bfseries\@parfont}}
+
+\makeatletter
 
@@
   mathescape=true, % activates special behaviour of the dollar sign
-  basicstyle=\scriptsize\tt,
+  basicstyle=\footnotesize\tt,
   columns=fullfelxible,
@@
 
-\newcommand{\offerfork}{\mathsf{offer\_fork}}
+\newcommand{\offerfork}{\mathsf{request\_fork}}
 \newcommand{\offerkill}{\mathsf{offer\_kill}}
@@
 algorithms on GPUs.  These algorithms are typically \emph{blocking},
-so they require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
-do not mandate fair scheduling, and GPU schedulers are
+so they require fair scheduling.  But GPU programming models
+(e.g.\ OpenCL) do not mandate fair scheduling, and GPU schedulers are
 unfair in practice.  Current approaches avoid this issue by exploiting
-scheduling quirks of today's GPUs in a manner that does not
-allow the GPU to be shared with other workloads (such as graphics
-rendering tasks).  We propose \emph{cooperative kernels}, an extension
-to the traditional GPU programming model geared towards writing
-blocking algorithms.  Workgroups of a cooperative kernel \emph{are}
-fairly scheduled, and multitasking is supported via a small set of
-language extensions through which the kernel and scheduler cooperate.
-We describe the semantics of the cooperative kernels programming model
-and our prototype implementation.  We evaluate the approach by porting
-a set of irregular work stealing and graph algorithms to our
-programming model and examining their performance under multitasking
-across three GPUs.  Our implementation exploits no
-vendor-specific hardware, driver or compiler support, so our encouraging
-results provide a lower-bound on the efficiency with
-which cooperative kernels can be implemented in practice.
+scheduling quirks of today's GPUs in a manner that does not allow the
+GPU to be shared with other workloads (such as graphics rendering
+tasks).  We propose \emph{cooperative kernels}, an extension to the
+traditional GPU programming model geared towards writing blocking
+algorithms.  Workgroups of a cooperative kernel \emph{are} fairly
+scheduled, and multitasking is supported via a small set of language
+extensions through which the kernel and scheduler cooperate.  We
+describe a prototype implementation of a cooperative kernel framework
+implemented in OpenCL 2.0 and evaluate our approach by porting a set
+of blocking GPU applications to cooperative kernels and examining
+their performance under multitasking.  Our prototype exploits no
+vendor-specific hardware, driver or compiler support, thus our results
+provide a lower-bound on the efficiency with which cooperative kernels
+can be implemented in practice.
 
@@
 
+\newcommand{\myparagraph}[1]{\paragraph{#1}}
+%{\noindent{\bf\emph{#1}}\hspace{1em}}
+
 \section{Introduction}\label{sec:intro}
 
-\paragraph{The needs of irregular data-parallel algorithms}
+\myparagraph{The needs of irregular data-parallel algorithms}
 %Acceleration of general-purpose computations on graphics processing
@@
 %the workgroups that execute a GPU kernel ahead of time.  However,
-Many
-interesting data-parallel algorithms are \emph{irregular}: the amount
+Many interesting data-parallel algorithms are \emph{irregular}: the amount
 of work to be processed is unknown ahead of time and may be determined
@@
 between workgroups, to balance load and communicate intermediate
-results.  For example, many graph algorithms employ a level-by-level strategy, requiring a global barrier between each level, and 
-work stealing algorithms require each workgroup
-to maintain a queue, which is typically protected by a mutex to enable
-stealing by other workgroups.
+results.  For example, many graph algorithms employ a level-by-level
+strategy, requiring a global barrier between each level, and work
+stealing algorithms require each workgroup to maintain a queue, which
+is typically protected by a mutex to enable stealing by other
+workgroups.
 
@@
 %\vspace{-1mm}
-\paragraph{A degree of fairness: occupancy-bound execution}
-GPU schedulers are \emph{not} fair. The current GPU
-programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
-HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
-workgroups, and current GPU scheduler are unfair in practice.  Roughly speaking, each workgroup
-executing a GPU kernel is mapped to a hardware \emph{compute
-  unit}.\footnote{In practice, depending on the kernel, multiple workgroups might map to the same compute unit; we ignore this in our current discussion.}
-%
-The simplest way for a GPU driver to handle 
-more workgroups being launched than there are compute
-units is via an \emph{occupancy-bound}
-execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where, once a workgroup has commenced execution on a
-compute unit (it has become \emph{occupant}), the workgroup has
-exclusive access to the compute unit until it finishes execution.
-Experiments suggest that the occupancy-bound execution model is widely
-implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
-evidently a simple model for device vendors to implement efficiently.
+\myparagraph{A degree of fairness: occupancy-bound execution} GPU
+schedulers are \emph{not} fair. The current GPU programming
+models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
+HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding
+scheduling of workgroups, and current GPU scheduler are unfair in
+practice.  Roughly speaking, each workgroup executing a GPU kernel is
+mapped to a hardware \emph{compute unit}.\footnote{In practice,
+  depending on the kernel, multiple workgroups might map to the same
+  compute unit; we ignore this in our current discussion.}
+%
+The simplest way for a GPU driver to handle more workgroups being
+launched than there are compute units is via an \emph{occupancy-bound}
+execution
+model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16} where,
+once a workgroup has commenced execution on a compute unit (it has
+become \emph{occupant}), the workgroup has exclusive access to the
+compute unit until it finishes execution.  Experiments suggest that
+the occupancy-bound execution model is widely implemented by today's
+GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.
+It is evidently a simple model for device vendors to implement
+efficiently.
 
@@
 bound to separate compute units that operate in parallel.  Current GPU
-implementations of blocking algorithms assume the
-occupancy-bound execution model, which they exploit by launching 
-no more workgroups than there are available compute
-units~\cite{owens-persistent}. This works because \emph{today}'s GPUs seem to provide the
-occupancy-bound execution model.
+implementations of blocking algorithms assume the occupancy-bound
+execution model, which they exploit by launching no more workgroups
+than there are available compute units~\cite{owens-persistent}. This
+works because \emph{today}'s GPUs seem to provide the occupancy-bound
+execution model.
 
 %\vspace{-1mm}
-\paragraph{Resistance to guaranteeing occupancy-bound execution}
-
+\myparagraph{Resistance to occupancy-bound execution}
 Despite its practical prevalence, none of the current GPU programming
 models actually mandate occupancy-bound execution.  Further, there are
-clear reasons why this model is undesirable.
+reasons why this model is undesirable.
 % from the point-of-view of
@@
 not commit to the occupancy-bound execution model they currently
-implement, for the above reasons,
-yet (2) will not guarantee fair scheduling due to the
-runtime cost of preempting
-workgroups.  Vendors instead wish to
-retain the essence of the simple occupancy-bound model, supporting preemption
-only in key special cases.
+implement, for the above reasons, yet (2) will not guarantee fair
+scheduling due to the runtime cost of preempting workgroups in a fair
+manner.  Vendors instead wish to retain the essence of the simple
+occupancy-bound model, supporting preemption only in key special
+cases.
 
 %\vspace{-1mm}
-\paragraph{Our proposal: cooperative kernels}
+\myparagraph{Our proposal: cooperative kernels}
 %
@@
 The use of $\offerfork$ and $\offerkill$ creates a contract between
-the scheduler and the (programmer of the) cooperative kernel.
+the scheduler and the cooperative kernel.
 Functionally, the
@@
 %\vspace{-1mm}
-\begin{enumerate}
+\begin{enumerate}[leftmargin=*]
 
@@
 
-\item Our experiments with a range of applications using three GPUs show that the model can enable efficient multitasking of cooperative and non-cooperative
+\item Our experiments with a range of applications \cutthree{using three GPUs} show that the model can enable efficient multitasking of cooperative and non-cooperative
   tasks.
@@
   kernels on top OpenCL 2.0
-  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms across three GPUs (\mysec\ref{sec:experiments}).
+  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}).
 
@@
 
-\section{Background and Motivating Examples}\label{sec:background}
-
-We outline the industry-standard OpenCL programming model on which we
+\section{Background and Examples}\label{sec:background}
+
+We outline the OpenCL programming model on which we
 base cooperative kernels (\mysec\ref{sec:opencl}), and illustrate
@@
 identical arguments.  A thread can call $\getglobalid$
-to obtain its unique id, to access distinct data or follow different control flow paths to other threads.
+to obtain a unique id, to access distinct data or follow different control flow paths.
 
@@
 $\getlocalid$ and $\getgroupid$ return a thread's local id within
-its workgroup and the id of the workgroup.\footnote{For simplicity we assume 1D kernels, which captures all the
-irregular algorithms we know of, and write
-e.g.\ $\getglobalid$ for $\getglobalid(0)$.
-}  The number
+its workgroup and the id of the workgroup.
+%
+%\footnote{For simplicity we assume 1D kernels, which captures all the
+%irregular algorithms we know of, and write
+%e.g.\ $\getglobalid$ for $\getglobalid(0)$.
+%}  
+%
+The number
 of threads per workgroup and number of workgroups are obtained via
-$\getlocalsize$ and $\getnumgroups$.  A thread's local and global
-id are related by $\getglobalid = \getgroupid \times
-\getlocalsize + \getlocalid$.  The total number of threads executing the
-kernel is given by $\getglobalsize = \getlocalsize \times \getnumgroups$.
+$\getlocalsize$ and $\getnumgroups$.  
+%A thread's local and global
+%id are related by $\getglobalid = \getgroupid \times
+%\getlocalsize + \getlocalid$.  The total number of threads executing the
+%kernel is given by $\getglobalsize = \getlocalsize \times \getnumgroups$.
 
 Execution of the threads in a workgroup can be synchronised via a
-workgroup barrier, which also ensures memory consistency.  This
-barrier is a \emph{workgroup-level} function: it can only
-appear in conditional code if all threads in a workgroup that reach
-the barrier agree on the guards of all enclosing conditional
-statements.  A \emph{global} barrier
-(synchronising all threads of a kernel) is \emph{not} provided as
-a primitive.
-
-\paragraph{Memory spaces and memory model}
-A kernel has access to four memory spaces.  \emph{Shared virtual
-memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
-kernel.  Each workgroup has a portion of \emph{local} memory for fast intra-workgroup communication.  Every thread has
-a portion of very fast \emph{private} memory for 
+workgroup barrier. This barrier is a \emph{workgroup-level} function:
+it can only appear in conditional code if all threads in a workgroup
+that reach the barrier agree on the guards of all enclosing
+conditional statements.  A \emph{global} barrier (synchronising all
+threads of a kernel) is \emph{not} provided as a primitive.
+
+\myparagraph{Memory spaces and memory model} A kernel has access to
+four memory spaces.  \emph{Shared virtual memory} (SVM) is accessible
+to all device threads and the host concurrently.  \emph{Global} memory is
+shared among all device threads.  Each workgroup has a
+portion of \emph{local} memory for fast intra-workgroup communication.
+Every thread has a portion of very fast \emph{private} memory for
 function-local variables.
@@
 
-\paragraph{Execution model}
-
+\myparagraph{Execution model}
 OpenCL specifically makes no guarantees about fair scheduling between
@@
 %\vspace{-2mm}
-\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
+\caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
 %\vspace{-2mm}
@@
 
-\paragraph{Work stealing example}
+\myparagraph{Work stealing example}
 %
@@
 processing units. It is useful when the number of tasks to be
-processed is dynamic, due to one task creating an arbitrary number of new tasks.  In the context
-of GPUs, each workgroup has an associated queue from which it
-obtains tasks to process, and to which it stores new
-tasks. If a workgroup's queue is empty, the workgroup tries to \emph{steal} a task from another
-queue. Previous work covers work stealing feasibility and efficiency on GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}.
+processed is dynamic, due to one task creating an arbitrary number of
+new tasks.  Work stealing has been explored in the context of
+GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}. Each workgroup has an
+associated queue from which it obtains tasks to process, and to which
+it stores new tasks. If a workgroup's queue is empty, the workgroup
+tries to \emph{steal} a task from another queue.
 
@@
 
-\paragraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
+\myparagraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
 skeleton of a frontier-based graph traversal algorithm; such algorithms have
@@
 %\vspace{-1mm}
-\paragraph{Transmitted variables}
-
+\myparagraph{Transmitted variables}
+%
 A variable declared in the root scope of the cooperative kernel can
@@
 %\vspace{-1mm}
-\paragraph{Active workroups}
-
+\myparagraph{Active workroups}
+%
 If the host application launches a cooperative kernel requesting $N$
@@
 
-Fair scheduling \emph{is} guaranteed between active workgroups.  That is, if
-some thread in an active workgroup is enabled then eventually some
-thread in the active workgroup is guaranteed to execute an
-instruction.
+Fair scheduling \emph{is} guaranteed between active workgroups;
+i.e.\ if some thread in an active workgroup is enabled, then
+eventually this thread is guaranteed to execute an instruction.
 
@@
 %\vspace{-1mm}
-\paragraph{Semantics for $\offerkill$}
-
+\myparagraph{Semantics for $\offerkill$}
+%
 The $\offerkill$ primitive allows the cooperative kernel to return
@@
 %\vspace{-1mm}
-\paragraph{Semantics for $\offerfork$}
-
+\myparagraph{Semantics for $\offerfork$}
+%
 Recall that
@@
 %\vspace{-1mm}
-\paragraph{Global barriers}
-
+\myparagraph{Global barriers}
+%
 Because workgroups of a cooperative kernel are fairly scheduled, so a
@@
 
-The semantics of $\resizingglobalbarrier$ can be modelled via the following sequence of calls:
-
-\lstset{basicstyle=\small\tt,numbers=none}
-\begin{lstlisting}
-  $\globalbarrier()$;
-  if($\getgroupid()$ == 0) $\offerfork$();
-  $\globalbarrier()$;
-  $\offerkill()$;
-  $\globalbarrier()$;
-\end{lstlisting}
-\lstset{basicstyle=\scriptsize\tt,numbers=left}
-
+The semantics of $\resizingglobalbarrier$ can be modelled via calling $\offerfork$ and $\offerkill$
+surrounded, and seperated by calls to a $\globalbarrier$.
+%
+%\lstset{basicstyle=\small\tt,numbers=none}
+%\begin{lstlisting}
+%  $\globalbarrier()$;  $\offerfork$(); 
+%  $\globalbarrier()$; $\offerkill()$; 
+%  $\globalbarrier()$;
+%\end{lstlisting}
+%\lstset{basicstyle=\scriptsize\tt,numbers=left}
+%
 The enclosing $\globalbarrier$ calls ensure that the change in number
@@
 \mysec\ref{sec:resizingbarrier} we show how a resizing barrier can be
-implemented significantly more efficiently through direct interaction with the scheduler.
+implemented more efficiently through direct interaction with the scheduler.
 
@@
 
-\paragraph{A changing workgroup count}  Unlike in regular OpenCL,
-the value returned by $\getnumgroups$ is not fixed during the
-lifetime of a cooperative kernel: it corresponds to the active group count $M$, which changes as workgroups execute
-$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  The value returned by $\getglobalsize$
-is similarly subject to change.
-A cooperative kernel must thus be written in a manner that is
-robust to changes in the values returned by these functions.
-
-In general, their volatility means that use of these functions should be avoided.
-However, the situation is more stable if a cooperative kernel does not call
-$\offerkill$ and $\offerfork$ directly, so that only $\resizingglobalbarrier$ can affect the number of active
-workgroups.  Then, at any point
-during execution, the threads of a kernel are executing between some
-pair of resizing barrier calls, which we call a \emph{resizing barrier
-  interval} (considering the kernel entry and exit points
-conceptually to be special cases of resizing barriers).  The active workgroup count is constant within each resizing
-barrier interval, so that 
-$\getnumgroups$ and $\getglobalsize$ return stable values during such intervals.
+\myparagraph{A changing workgroup count} Unlike in regular OpenCL, the
+value returned by $\getnumgroups$ is not fixed during the lifetime of
+a cooperative kernel: it corresponds to the active group count $M$,
+which changes as workgroups execute $\offerkill$, and $\offerfork$.
+The value returned by $\getglobalsize$ is similarly subject to change.
+A cooperative kernel must thus be written in a manner that is robust
+to changes in the values returned by these functions.
+
+In general, their volatility means that use of these functions should
+be avoided.  However, the situation is more stable if a cooperative
+kernel does not call $\offerkill$ and $\offerfork$ directly, so that
+only $\resizingglobalbarrier$ can affect the number of active
+workgroups.  Then, at any point during execution, the threads of a
+kernel are executing between some pair of resizing barrier calls,
+which we call a \emph{resizing barrier interval} (considering the
+kernel entry and exit points conceptually to be special cases of
+resizing barriers).  The active workgroup count is constant within
+each resizing barrier interval, so that $\getnumgroups$ and
+$\getglobalsize$ return stable values during such intervals.
 %
@@
 
-\paragraph{Adapting work stealing}
-
+\myparagraph{Adapting work stealing}
+%
 In this example there is no state to transmit since a computation is
@@
 
-\paragraph{Adapting graph traversal} 
+\myparagraph{Adapting graph traversal} 
 \myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
@@
 $\getglobalsize$ at line~\ref{line:cgraph:rechunking2} is stable
-until the next resizing barrier at line~\ref{line:cgraph:resizing1},
-and in particular can be safely used inside the $\keyword{for}$ loop.
+until the next resizing barrier at line~\ref{line:cgraph:resizing1}.
+%and in particular can be safely used inside the $\keyword{for}$ loop.
 
@@
 
-\paragraph{Patterns for irregular algorithms}
+\myparagraph{Patterns for irregular algorithms}
 In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
@@
 %
-\myfiglong\ref{fig:cgraphsearch} is simpler than, but representative of,
+\myfiglong\ref{fig:cgraphsearch} is representative of
 most level-by-level graph algorithms.
@@
 
-The semantics presented in \mysec\ref{sec:semantics} describe the envelope of
-behaviour that a developer of a cooperative kernel should be prepared
+The semantics presented in \mysec\ref{sec:semantics} describe the 
+behaviours that a developer of a cooperative kernel should be prepared
 for.
@@
 %
-Getting this non-functional requirement right is GPU- and application-dependent.
-In \mysec\ref{sec:sizingnoncoop} we conduct
+Getting this non-functional requirement right is GPU- and
+application-dependent.  In \mysec\ref{sec:sizingnoncoop} we conduct
 experiments to understand the response rate that would be required to
-co-schedule graphics rendering with a cooperative kernel, maintaining a smooth frame rate.
+co-schedule graphics rendering with a cooperative kernel, maintaining
+a smooth frame rate.
 
@@
 by accepting an $\offerkill$ only if a compute unit is required for another
-task, and responding positively to $\offerfork$ calls if compute units are available.  Our
-implementation (\mysec\ref{sec:implementation}) is as
-generous as possible to the cooperative kernel in this regard.
+task, and responding positively to $\offerfork$ calls if compute units are available.
+
+%Our
+%implementation (\mysec\ref{sec:implementation}) is as
+%generous as possible to the cooperative kernel in this regard.
 
@@
 Our vision is that support for cooperative kernels will be integrated
-in the runtimes of future GPU implementations of OpenCL and related
-programming models, with driver and compiler support for our new
-primitives.  To experiment with our ideas on current GPUs, we have
-developed a prototype that mocks up the required runtime support via a
-\emph{megakernel}, and exploits the occupancy-bound execution model
-that these GPUs provide to ensure fair scheduling between workgroups.
-We emphasise that an aim of cooperative kernels is to \emph{avoid}
-depending on the occupancy-bound model.  Our prototype exploits this
-model simply to allow us to experiment with current GPUs whose
-proprietary drivers we cannot change.  We describe the megakernel approach (\mysec\ref{sec:megakernel})
-and detail various aspects of the scheduler component of our implementation (\mysec\ref{sec:schedulerimpl}).
+in the runtimes of future GPU implementations of OpenCL, with driver
+and compiler support for our new primitives.  To experiment with our
+ideas on current GPUs, we have developed a prototype that mocks up the
+required runtime support via a \emph{megakernel}, and exploits the
+occupancy-bound execution model that these GPUs provide to ensure fair
+scheduling between workgroups.  We emphasise that an aim of
+cooperative kernels is to \emph{avoid} depending on the
+occupancy-bound model.  Our prototype exploits this model simply to
+allow us to experiment with current GPUs whose proprietary drivers we
+cannot change.  We describe the megakernel approach
+(\mysec\ref{sec:megakernel}) and detail various aspects of the
+scheduler component of our implementation
+(\mysec\ref{sec:schedulerimpl}).
 
@@
 
-\paragraph{Simplifying assumptions}
-
+\myparagraph{Simplifying assumptions}
+%
 For ease of implementation, our prototype supports
@@
 assume all variables declared at the root lexical scope of a
-cooperative kernel are implicitly $\transmit$-annotated; this allows our
-implementation to piggy-back on top of Clang (see below) without
-requiring front-end changes.  Finally, we require that $\offerkill$,
+cooperative kernel are implicitly $\transmit$-annotated.
+%this allows our
+%implementation to piggy-back on top of Clang (see below) without
+%requiring front-end changes.  
+Finally, we require that $\offerkill$,
 $\offerfork$ and $\resizingglobalbarrier$ are called from the entry
@@
 
-\paragraph{Merging kernels}
-
-We have built a tool, \kernelmerge{}, that takes a cooperative kernel
-and a non-cooperative kernel and merges them to produce a megakernel
-with an embedded GPU scheduler.  We implemented \kernelmerge{} using the
-Clang LibTooling framework~\cite{clang}, leveraging recent Clang support
-for OpenCL 2.0~\cite{ClangOpenCL20}.
+%\myparagraph{Merging kernels}
+%
+%We have built a tool, \kernelmerge{}, that takes a cooperative kernel
+%and a non-cooperative kernel and merges them to produce a megakernel
+%with an embedded GPU scheduler.  We implemented \kernelmerge{} using the
+%Clang LibTooling framework~\cite{clang}, leveraging recent Clang support
+%for OpenCL 2.0~\cite{ClangOpenCL20}.
 
@@
 To enable multitasking through cooperative kernels, the runtime (in
-our case, the megakernel) must track the state of all
-workgroups, i.e.\ whether a workgroup is waiting or computing a kernel,
-maintain consistent context states for each kernel, e.g.\ tracking the number
-of active workgroups, and provide a
-safe way for these states to be modified in response to $\offerfork$/$\offerkill$. We discuss these issues,
-and describe the implementation of an efficient resizing barrier.
-
-We describe how the scheduler would handle arbitrary
+our case, the megakernel) must track the state of workgroups,
+i.e.\ whether a workgroup is waiting or computing a kernel, maintain
+consistent context states for each kernel, e.g.\ tracking the number
+of active workgroups, and provide a safe way for these states to be
+modified in response to $\offerfork$/$\offerkill$. We discuss these
+issues, and describe the implementation of an efficient resizing
+barrier. We describe how the scheduler would handle arbitrary
 combinations of kernels, though as noted above our current
@@
 
-\paragraph{Scheduler contexts}
-
+\myparagraph{Scheduler contexts}
+%
 %When a workgroup executes a kernel (cooperative or otherwise), it must
@@
 %id valid for the kernel.
-
+%
 To dynamically manage workgroups executing cooperative
@@
 
-\paragraph{Launching kernels and managing workgroups}
-
-To launch a kernel, the host sends a data packet to the GPU
-scheduler consisting of: a kernel to execute, the inputs to the kernel, the
-number of workgroups that should execute the kernel, and a flag indicating whether the kernel
-is cooperative. In our implementation, this host-device communication channel is built using fine-grained SVM atomics.
+\myparagraph{Launching kernels and managing workgroups}
+%
+To launch a kernel, the host sends a data packet to the GPU scheduler
+consisting of: a kernel to execute, the kernel inputs, and a
+flag indicating whether the kernel is cooperative. In our
+implementation, this host-device communication channel is built using
+fine-grained SVM atomics.
 
@@
 
-\paragraph{Accounting for concurrency}
-
+\myparagraph{Accounting for concurrency}
+%
 Concurrent calls to $\offerfork$ must be protected by a mutex to avoid
 racing on the available workgroups. The same conflict over available
-workgroups can occur at a new kernel launch between the scheduler and
-a workgroup calling $\offerfork$. Thus, these activities are protected
-by a mutex, called the \emph{assignment mutex} provided in the scheduler
-context.
+workgroups can occur when launching a new kernel between the scheduler
+and a workgroup calling $\offerfork$. Thus, $\offerfork$ is protected
+by a mutex, called the \emph{assignment mutex}. 
+%provided in the scheduler context.
 
@@
 
-\paragraph{An efficient resizing barrier}
-
-In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
-of a series of calls to other primitives.  It is possible, however to implement the resizing with
-only one call to a global barrier with $\offerfork$ and $\offerkill$
-baked into the implementation.
+\myparagraph{An efficient resizing barrier}
+%
+In \mysec\ref{sec:semantics}, we defined the semantics of a resizing
+barrier in terms of calls to other primitives.  It is possible,
+however to implement the resizing barrier with only one call to a
+global barrier with $\offerfork$ and $\offerkill$ baked inside.
 
@@
 
-A first-attempt at incorporating $\offerfork$ and $\offerkill$ into
-such a barrier is straight forward. Upon entering the barrier, the
-slaves first execute $\offerkill$, possibly exiting. The master then
-waits for $M$ slaves (the number of active workgroups), which may
-decrease due to $\offerkill$ calls by the
-slaves, but will not increase. Once the master observes that $M$ slaves
-have arrived, it knows that all other workgroups are spinning, waiting to be released. The master executes $\offerfork$,
-and the statement immediately following this $\offerfork$ is a conditional that forces newly spawned workgroups to join the
-slaves in waiting to be released. Finally, the master
+Incorporating $\offerfork$ and $\offerkill$ into such a barrier is
+straight forward. Upon entering the barrier, the slaves first execute
+$\offerkill$, possibly exiting. The master then waits for $M$ slaves
+(the number of active workgroups), which may decrease due to
+$\offerkill$ calls by the slaves, but will not increase. Once the
+master observes that $M$ slaves have arrived, it knows that all other
+workgroups are waiting to be released. The master executes
+$\offerfork$, and the statement immediately following this
+$\offerfork$ is a conditional that forces newly spawned workgroups to
+join the slaves in waiting to be released. Finally, the master
 releases all the slaves: the original slaves and the new slaves that
@@
 This barrier implementation is sub-optimal because workgroups only
-execute $\offerkill$ once per barrier call and, depending on
-order of arrival, it is possible
-that only one workgroup is killed per barrier call, preventing the scheduler from gathering workgroups quickly.
-
-We can greatly reduce the gather time by providing a new $\keyword{query}$
-function for a workgroup executing a cooperative kernel, which returns the number of workgroups that the scheduler needs to obtained
-from the cooperative kernel.  
-%
-The programmer cannot call $\keyword{query}$ directly;
-it is only invoked from within the resizing barrier implementation.
-%
-A resizing barrier can now be implemented
-as follows: (1) the master waits for all slaves to arrive; (2) the master calls $\offerfork$ and
-commands the new workgroups to be slaves; (3) the master calls $\keyword{query}$, obtaining a value $W$;
-(4) the master releases the slaves, broadcasting the value $W$ to them; (5) workgroups with ids larger than $M-W$ spin, calling 
-$\offerkill$ repeatedly until the scheduler claims them---we know from $\keyword{query}$ that the scheduler will eventually do so.
-We show in \mysec\ref{sec:responsiveness} that the barrier using
-query greatly reduces the gather time in practice.
+execute $\offerkill$ once per barrier call and, depending on order of
+arrival, it is possible that only one workgroup is killed per barrier
+call, preventing the scheduler from gathering workgroups quickly.
+
+We can reduce the gather time by providing a new
+$\keyword{query}$ function for cooperative kernels, which returns the
+number of workgroups that the scheduler needs to obtain from the
+cooperative kernel.
+%
+%The programmer cannot call $\keyword{query}$ directly; it is
+%invoked from within the resizing barrier implementation.
+%
+A resizing barrier can now be implemented as follows: (1) the master
+waits for all slaves to arrive; (2) the master calls $\offerfork$ and
+commands the new workgroups to be slaves; (3) the master calls
+$\keyword{query}$, obtaining a value $W$; (4) the master releases the
+slaves, broadcasting the value $W$ to them; (5) workgroups with ids
+larger than $M-W$ spin, calling $\offerkill$ repeatedly until the
+scheduler claims them---we know from $\keyword{query}$ that the
+scheduler will eventually do so.  We show in
+\mysec\ref{sec:responsiveness} that the barrier using $\keyword{query}$ greatly
+reduces the gather time in practice.
 
@@
 \begin{table}[t]
-\small
+\normalsize
 \centering
@@
 \label{tab:applications}
+\vspace{-8mm}
 \end{table}
 
-\begin{table}[t]
-\footnotesize
-\centering
-\begin{tabular}{ l l l l }
-Chip & HD520 & HD5500 & Iris\\
-\hline
-\# CU  & 24 & 24 & 47 \\
-Driver   & 20.19.15.4501 & 10.18.15.4281 & 20.19.15.4463 \\
-Host CPU & i3-6100U & i7-5600U & i3-5157U \\
-\end{tabular}
-\caption{The Intel GPU platforms used in our evaluation}
-\label{tab:chipstested}
-\end{table}
+%\begin{table}[t]
+%\footnotesize
+%\centering
+%\begin{tabular}{ l l l l }
+%Chip & HD520 & HD5500 & Iris\\
+%\hline
+%\# CU  & 24 & 24 & 47 \\
+%Driver   & 20.19.15.4501 & 10.18.15.4281 & 20.19.15.4463 \\
+%Host CPU & i3-6100U & i7-5600U & i3-5157U \\
+%\end{tabular}
+%\caption{The Intel GPU platforms used in our evaluation}
+%\label{tab:chipstested}
+%\end{table}
 
@@
 graph algorithms, based on the Pannotia~\cite{Pannotia} and
-Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
-indicate how many of the original number of barriers are changed to
+Lonestar~\cite{BNP12} GPU application suites, using global barriers.
+We indicate how many of the original number of barriers are changed to
 resizing barriers, and how many variables need to be transmitted.  In
@@
 barriers deeper in the call stack, a case not supported by our
-prototype although these barriers could in principle be
-converted. The remaining 2 algorithms are work stealing applications:
-each required the addition of $\offerfork$ and $\offerkill$ at the start
-of the main loop, and no variables needed to be transmitted.  For all
-these examples the porting to cooperative kernels was similarly
+prototype although these barriers could in principle be converted. The
+remaining 2 algorithms are work stealing applications: each required
+the addition of $\offerfork$ and $\offerkill$ at the start of the main
+loop, and no variables needed to be transmitted.  For all these
+examples, the porting to cooperative kernels was similarly
 straightforward to the porting illustrated in
-\mysec\ref{sec:programmingguidelines} for our running examples. Most graph come with 2
-different data sets as input, leading to 13 application/input pairs in
-total.
-
-From the Lonestar suite we omit two applications that use global barriers:
-dmr has been shown to be unstable across GPU vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially short-running so that it is not
-interesting for our multitasking approach.
+\mysec\ref{sec:programmingguidelines}. Most
+graph come with 2 different data sets as input, leading to 13
+application/input pairs in total.
+
+From the Lonestar suite we omit two applications that use global
+barriers: dmr has been shown to be unstable across GPU
+vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially
+short-running so that it is not interesting for our multitasking
+approach.
 
@@
 equivalent functionality is not provided by \nvidia's CUDA. Intel
-drivers meet our requirements on Windows, but not on Linux (we
-confirmed this by trying the upstream driver that requires a patched
-Linux kernel). AMD support SVM atomics only their Kaveri family of
-chips and only on Linux, requiring to compile the latest driver.
-
-Among the GPUs available to us, four met our requirements: Intel HD 520,
-HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux drivers
-for OpenCL suffer from a known defect whereby long-running kernels lead
-to defunct processes that the OS cannot
-kill~\cite{amd-defunct-process}. We observed this on our AMD platform
-with the latest driver, and simultaneously obtained alarming results on
-this platform when using SVM atomics: modifying a kernel to include a
-simple host/device handshake through SVM slowed down kernel execution by
-an order of magnitude in some cases.  This behaviour, combined with the
-defunct processes issue, lead us to suspect that the support we require
-for cooperative kernels is not yet mature enough within AMD's drivers,
-which we plan to discuss with engineers at AMD.
-
-We thus present experimental results for the Intel GPUs, whose characteristics are summarised in \mytab\ref{tab:chipstested}.
+drivers meet our requirements on Windows, but not on Linux. 
+%(we
+%confirmed this by trying the upstream driver that requires a patched
+%Linux kernel). 
+AMD support SVM atomics only their Kaveri family of
+chips and only on Linux.%, requiring to compile the latest driver.
+
+Among the GPUs available to us, four met our requirements: Intel HD
+520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
+drivers for OpenCL suffer from a known defect whereby long-running
+kernels lead to unkillable defunct
+processes~\cite{amd-defunct-process}. Because of this behaviour, we
+could not experiment on AMD platforms.
+
+%and simultaneously obtained alarming results on this platform when
+%using SVM atomics: modifying a kernel to include a simple host/device
+%handshake through SVM slowed down kernel execution by an order of
+%magnitude in some cases.  This behaviour, combined with the defunct
+%processes issue, lead us to suspect that the support we require for
+%cooperative kernels is not yet mature enough within AMD's drivers,
+%which we plan to discuss with engineers at AMD.
+
+We thus ran our experimental campaign on the Intel GPUs. The results
+were similar across the GPUs, so for conciseness, we report only on
+the Iris 6100 GPU (driver 20.19.15.4463) with a host CPU i3-5157U. The
+Iris has a reported 47 compute units. Results for the two other Intel
+GPUs are available in supplementary material.
 
@@
 
-\subsection{The Overhead of Moving to Cooperative Kernels}\label{sec:overhead}
+\subsection{The Overhead of Cooperative Kernels}\label{sec:overhead}
 
@@
 
-\begin{table}
-\small
-\centering
-\begin{tabular}{ l l l l }
-Chip & HD520 & HD5500 & Iris\\
-\hline
-overall geo. mean  & $1.20$ & $1.14$ & $1.08$ \\
-overall max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
-
-\hline
-barrier geo. mean & $1.18$ & $1.16$ & $1.07$ \\
-barrier max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
-
-\hline
-wk.steal. geo. mean & $1.42$ & $1.10$ & $1.12$ \\
-wk.steal. max & $1.42^{\diamond}$ & $1.18^{\diamond}$ & $1.23^{\diamond}$ \\
-
-\hline
-\end{tabular}\\
-{\footnotesize
-$^{\ast}$color ecology, $^{\dagger}$bc 1k\_128k, $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
-}
+%\begin{table}
+%\small
+%\centering
+%\begin{tabular}{ l l l l }
+%Chip & HD520 & HD5500 & Iris\\
+%\hline
+%overall geo. mean  & $1.20$ & $1.14$ & $1.08$ \\
+%overall max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
+%
+%\hline
+%barrier geo. mean & $1.18$ & $1.16$ & $1.07$ \\
+%barrier max & $1.75^{\ast}$ & $1.45^{\dagger}$ & $1.37^{\ddagger}$ \\
+%
+%\hline
+%wk.steal. geo. mean & $1.42$ & $1.10$ & $1.12$ \\
+%wk.steal. max & $1.42^{\diamond}$ & $1.18^{\diamond}$ & $1.23^{\diamond}$ \\
+%
+%\hline
+%\end{tabular}\\
+%{\footnotesize
+%$^{\ast}$color ecology, $^{\dagger}$bc 1k\_128k, $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
+%}
 %\vspace{-2mm}
-\caption{Execution time slowdown associated with cooperative scheduling mode without multitasking}
-\label{tab:overhead}
+%\caption{Execution time slowdown associated with cooperative scheduling mode without multitasking}
+%\label{tab:overhead}
 %\vspace{-2mm}
-\end{table}
-
-\paragraph{Experimental setup}
+%\end{table}
+
+\myparagraph{Experimental setup}
 Invoking the cooperative scheduling primitives incurs some overhead
@@
 
-\paragraph{Results}
-For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
+\begin{figure*}
+\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf} 
+\includegraphics[width=.67\columnwidth]{hd520_mis_ecology.pdf}
+\includegraphics[width=.67\columnwidth]{hd5500_bfs_rmat22.pdf}
+%\vspace{-2mm}
+\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
+\vspace{-3mm}
+\end{figure*}
+
+\begin{table}
+\normalsize
+\centering
+\begin{tabular}{ l l | l l | l l }
+\multicolumn{2}{c|}{overall} & \multicolumn{2}{c|}{barrier} & \multicolumn{2}{c}{wk.steal.} \\
+mean & max & mean & max & mean & max \\
+\hline
+$1.08$ & $1.37^{\ddagger}$ & $1.07$ & $1.37^{\ddagger}$ & $1.12$ & $1.23^{\diamond}$ \\
+\end{tabular}\\
+{\small
+ $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
+}
+%\vspace{-2mm}
+\caption{Cooperative kernel slowdown w/o multitasking}
+\label{tab:overhead}
+\vspace{-8mm}
+\end{table}
+
+\myparagraph{Results}
+\mytab\ref{tab:overhead} shows the geometric mean and
 maximum slowdown across all applications and inputs, with averages and
@@
 we indicate which application and input was responsible. The slowdown is
-below 2 even in the worst case, and closer to 1 on average. We consider
-these results as encouraging, especially since the performance of our
+below 1.5 even in the worst case, and closer to 1 on average. We consider
+these results encouraging, especially since the performance of our
 prototype could clearly be improved upon in a native implementation.
@@
 
-We now assess the responsiveness of multitasking between a long-running cooperative kernel and a series of
-short, non-cooperative kernel launches, and the performance impact of multitasking on the cooperative kernel.
+We now assess the responsiveness of multitasking between a
+long-running cooperative kernel and a series of short, non-cooperative
+kernel launches, and the performance impact of multitasking on the
+cooperative kernel.
 
@@
 
-\paragraph{Experimental setup}
-For a given cooperative kernel and its input, we launch the kernel and
-then repeatedly schedule a non-cooperative kernel that aims to simulate
-the intensity of one of the three classes of graphics rendering workload
-discussed in \mysec\ref{sec:sizingnoncoop}. In practice, we use a matrix
+\myparagraph{Experimental setup} For a given cooperative kernel and
+its input, we launch the kernel and then repeatedly schedule a
+non-cooperative kernel that aims to simulate the intensity of one of
+the three classes of graphics rendering workload discussed in
+\mysec\ref{sec:sizingnoncoop}. In practice, we use a matrix
 multiplication kernel as the non-cooperative workload, with matrix
-dimensions tailored for each GPU to reach the appropriate execution
-duration.  We conduct separate runs where we vary the number of
-workgroups requested by the non-cooperative kernel, considering the
-cases where one, a quarter, a half, and all-but-one, of the total number
-of workgroup slots is requested.  Moreover, for the graph algorithms we
-try both regular and query barrier implementations.
+dimensions tailored to reach the appropriate execution duration.  We
+conduct separate runs where we vary the number of workgroups requested
+by the non-cooperative kernel, considering the cases where one, a
+quarter, a half, and all-but-one, of the total number of workgroups
+are requested. For the graph algorithms we try both
+regular and query barrier implementations.
 
 Our experiments span 13 pairs of cooperative kernels and inputs, 3
-classes of non-cooperative kernel workloads, 4 quantities of workgroups
-claimed for the non-cooperative kernel and 2 variations of resizing
-barriers for graph algorithms, leading to 288 configurations.  We run
-each configuration 10 times on each GPU, in order to report averaged
+classes of non-cooperative kernel workloads, 4 quantities of
+workgroups claimed for the non-cooperative kernel and 2 variations of
+resizing barriers for graph algorithms, leading to 288 configurations.
+We run each configuration 10 times, in order to report averaged
 performance numbers. For each run, we record the execution time of the
@@
 scheduler to collect enough workgroups to launch the non-cooperative
-kernel, and the non-cooperative kernel execution time. We sanity-check
-whether all kernels perform their computations correctly by comparing
-their computed result with expected reference results.
-
-To work around compiler crash and compiler hang bugs in Intel's current
-OpenCL drivers, we had to apply some semantics-preserving changes to the
-code generated by \kernelmerge{} for a few applications configurations.
-Driver bugs also meant that we could not obtain results for some configurations where the
-non-cooperative kernel asks for all-but-one workgroups (namely bc, bfs
-and l-sssp on all chips, and color on Iris), and could not get the game
-application to run at all on HD520; these configurations led to
-severe failures such as machine freezes and blue screens.
+kernel, and the non-cooperative kernel execution time. 
+%
+%We sanity-check
+%whether all kernels perform their computations correctly by comparing
+%their result with reference results.
+
+To work around compiler crash and compiler hang bugs in Intel's
+current OpenCL drivers, we had to apply some semantics-preserving
+changes to some applications.  Driver bugs also meant that we could
+not obtain results for some configurations where the non-cooperative
+kernel asks for all-but-one workgroups (namely bc, bfs, l-sssp, and
+color).  These configurations led to severe failures such as machine
+freezes and blue screens.
 
@@
 
-\paragraph{Responsiveness}
-
+\myparagraph{Responsiveness}
+%
 \myfiglong\ref{fig:fine-grained-timing} reports, on three
-configurations, the average gather and execution times for the non-cooperative
-kernel w.r.t. the quantity of workgroups allocated to it.
-A logarithmic
-scale is used for time since gather times tend to be much smaller than
-execution times. The horizontal grey lines indicates the desired period
-for non-cooperative kernels.  These graphs show a representative sample
-of our results; the full set of graphs for all configurations is
-provided in Appendix~\ref{appendix:extragraphs} (supplementary material).
-
-Let us first focus on the left graph, which illustrates a work stealing
-example on the Iris GPU.  When the non-cooperative kernel is given only
-one workgroup, its execution is so long that it cannot complete within
-the period required for a screen refresh. The gather time is very good
+configurations, the average gather and execution times for the
+non-cooperative kernel w.r.t. the quantity of workgroups allocated to
+it.  A logarithmic scale is used for time since gather times tend to
+be much smaller than execution times. The horizontal grey lines
+indicates the desired period for non-cooperative kernels.  These
+graphs show a representative sample of our results; the full set of
+graphs for all configurations is provided in supplementary material.
+
+The left most graph illustrates a work
+stealing example.  When the non-cooperative kernel is given only one
+workgroup, its execution is so long that it cannot complete within the
+period required for a screen refresh. The gather time is very good
 though, since the scheduler needs to collect only one workgroup.  The
-more workgroups are allocated to the non-cooperative kernels, the faster
-it can compute: here the non-cooperative kernel becomes fast enough with
-a quarter (resp.\ half) of available workgroups for light (resp.\ heavy)
-graphics workload. Inversely, the gather time increases since the
-scheduler must collect more and more workgroups.
-
-The middle and right graphs show results for graph algorithms on the
-other two GPUs.  These
+more workgroups are allocated to the non-cooperative kernels, the
+faster it can compute: here the non-cooperative kernel becomes fast
+enough with a quarter (resp.\ half) of available workgroups for light
+(resp.\ heavy) graphics workload. Inversely, the gather time increases
+since the scheduler must collect more and more workgroups.
+
+The middle and right graphs show results for graph algorithms.  These
 algorithms use barriers, and we experimented with the regular and
-query barrier implementations described in \mysec\ref{sec:resizingbarrier}.
-The execution times for the non-cooperative task are averaged across all runs, including with both types of barrier.
-We show separately the average gather time associated with each type of barrier.
-The graphs show a similar trend to the left-most graph: as the number of non-cooperative workgroups grows, the
-execution time decreases and the gather time increases.\footnote{The right
-figure has only 3 points per line: recall that for bfs there are no runs
-with ``all-but-one'' non-cooperative workgroups.}
-%
-the gather time is higher on the right figure: the rmat22 input graph is
-rather wide than deep, so the graph algorithm reaches resizing barriers
-less often than for the ecology input of the middle figure for
-instance. The scheduler thus has fewer opportunities to collect
+query barrier implementations described in
+\mysec\ref{sec:resizingbarrier}.  The execution times for the
+non-cooperative task are averaged across all runs, including with both
+types of barrier.  We show separately the average gather time
+associated with each type of barrier.  The graphs show a similar trend
+to the left-most graph: as the number of non-cooperative workgroups
+grows, the execution time decreases and the gather time
+increases.\footnote{The right figure has only 3 points per line:
+  recall that for bfs there are no runs with ``all-but-one''
+  non-cooperative workgroups.}
+%
+the gather time is higher on the right figure: the rmat22 input graph
+is rather wide than deep, so the graph algorithm reaches resizing
+barriers less often than for the ecology input of the middle figure
+for instance. The scheduler thus has fewer opportunities to collect
 workgroups and gather time increases. Nonetheless, scheduling
 responsiveness can benefit from the query barrier: when used, this
-barrier lets the scheduler collect all needed workgroups as soon as they
-hit a resizing barrier. As we can see on both the middle and right
-figure, the gather time of the query barrier is almost stable w.r.t.\ the
-number of workgroups that needs to be collected.
-
-\begin{figure*}
-\includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf} 
-\includegraphics[width=.67\columnwidth]{hd520_mis_ecology.pdf}
-\includegraphics[width=.67\columnwidth]{hd5500_bfs_rmat22.pdf}
-%\vspace{-2mm}
-\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
-%\vspace{-2mm}
-\end{figure*}
+barrier lets the scheduler collect all needed workgroups as soon as
+they hit a resizing barrier. As we can see, the gather time of the
+query barrier is almost stable w.r.t.\ the number of workgroups that
+needs to be collected.
 
@@
 \caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
-%\vspace{-2mm}
+\vspace{-5mm}
 \end{figure}
 
-\paragraph{Performance} 
-\myfiglong\ref{fig:performance} reports, for the Iris GPU, the overhead brought by the scheduling of
-non-cooperative kernels over the cooperative kernel execution time.
-This is the slowdown associated with running the cooperative kernel in the presence of multitasking, vs.\ running the cooperative kernel in isolation (geometric mean over all applications and inputs).
-We also show the period at which non-cooperative kernels can be scheduled (arithmetic mean over all applications
-and inputs).
-The results for the other GPUs are similar.
- We show
-results for the three workloads listed in
+\myparagraph{Performance} \myfiglong\ref{fig:performance} reports the
+overhead brought by the scheduling of non-cooperative kernels over the
+cooperative kernel execution time.  This is the slowdown associated
+with running the cooperative kernel in the presence of multitasking,
+vs.\ running the cooperative kernel in isolation (geometric mean over
+all applications and inputs).  We also show the period at which
+non-cooperative kernels can be scheduled (arithmetic mean over all
+applications and inputs).
+We show results for the three workloads listed in
 \mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
-period graph correspond to the period goals of the workloads: the higher
-(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for the
-light (resp. medium and heavy) workload.
+period graph correspond to the period goals of the workloads: the
+higher (resp. lower) line corresponds to a period of 70ms (resp. 40ms)
+for the light (resp. medium and heavy) workload.
 
@@
 
-Overall we are very encouraged by these experimental findings, especially since they provide a lower bound on potential performance of our cooperative kernels model.
-Implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.
+We consider these results encouraging, since they provide a lower
+bound on potential performance of our cooperative kernels
+model. Implementing the model natively, with device-specific runtime
+support, could only improve performance and responsiveness compared
+with that of our megakernel-based prototype.
 
@@
 
-\paragraph{Irregular algorithms and persistent kernels}
-
+\myparagraph{Irregular algorithms and persistent kernels}
+%
 There has been a lot of work on accelerating blocking irregular
@@
 
-\paragraph{Preemptive multitasking on GPUs}
-
+\myparagraph{Preemptive multitasking on GPUs}
+%
 Hardware support for preemption has been proposed for \nvidia GPUs,
@@
 
-\paragraph{Optimizing GPU scheduling}
+\myparagraph{Optimizing GPU scheduling}
 %
@@
 
-CUDA and OpenCL provide the facility for a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
-This \emph{dynamic parallelism} can be used to implement a
-GPU-based scheduler, by having an initial scheduler kernel repeatedly
-spawn further kernels as required, according to some scheduling
-policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, kernels
-that uses dynamic parallelism are still prone to unfair scheduling of workgroups, so this feature
-does not help in deploying blocking
-algorithms on GPUs.
-
-\paragraph{Cooperative multitasking}
-
+CUDA and OpenCL provide the facility for a kernel to spawn further
+kernels~\cite{cuda-75}.  This \emph{dynamic parallelism}
+can be used to implement a GPU-based scheduler, by having an initial
+scheduler kernel repeatedly spawn further kernels as required,
+according to some scheduling
+policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, kernels that
+uses dynamic parallelism are still prone to unfair scheduling of
+workgroups, and thus does not help in deploying blocking algorithms on
+GPUs.
+
+\myparagraph{Cooperative multitasking}
+%
 Cooperative multitasking was offered in older operating systems
-(e.g. Windows before the 95 edition) and is still used by some
-operating systems, such as RISC
-OS~\cite{risc-os-multitasking}.
-Many programming languages offer cooperative multitasking via
-variations of co-routines, e.g. co-routine objects in
-Python
+(e.g. pre 1995 Windows) and is still used by some operating systems,
+such as RISC OS~\cite{risc-os-multitasking}.
+
+%Many programming
+%languages offer cooperative multitasking via variations of
+%co-routines, e.g. co-routine objects in Python
 %\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
-and Lua,
+%and Lua,
 %\footnote{\url{https://www.lua.org/pil/9.1.html}}
-and Ruby's
-\emph{fiber}.
+%and Ruby's
+%\emph{fiber}.
 %\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}
@@
 fairly scheduled and to share GPU resources with other workloads.
-Experimental results using our megakernel-based prototype over three
-Intel GPUs shows that the model is a good fit for the needs of current
+Experimental results using our megakernel-based prototype \cutthree{over three Intel GPUs} 
+shows that the model is a good fit for the needs of current
 GPU-accelerated irregular algorithms.  The
@@
 
-\paragraph{States}
-
+\myparagraph{States}
+%
 Let $L$ be a set of \emph{local states} that abstractly captures the
@@
 
-\paragraph{Thread-level transitions}
-
+\myparagraph{Thread-level transitions}
+%
 We leave the semantics for thread-level transitions abstract, assuming
@@
 
-\paragraph{Memory synchronisation}
-
+\myparagraph{Memory synchronisation}
+%
 GPUs are known to have relaxed memory models~\cite{ABDGKPSW-2015}.  To abstractly
@@
 
-\paragraph{Operational semantics rules}
-
+\myparagraph{Operational semantics rules}
+%
 \myfiglong\ref{fig:semanticrules} presents the rules of our
@@
 
-\paragraph{Killing order}
-
+\myparagraph{Killing order}
+%
 We opted for a semantics whereby only the active workgroup with the
@@
 
-\paragraph{Keeping one workgroup alive}
-
+\myparagraph{Keeping one workgroup alive}
+%
 Our semantics dictate that the workgroup with id 0 will not be killed
@@
 
-\paragraph{Transmission of partial state from a single thread}
-
+\myparagraph{Transmission of partial state from a single thread}
+%
 Recall from the semantics of $\offerfork$ that newly forked workgroups

### 2017-02-17 write-ups/FSE/paper.tex
got the paper to fit in the space for FSE

@@
 threads have arrived, which requires fair
-scheduling between workgroups for the barrier to operate correctly.
+scheduling between workgroups.% for the barrier to operate correctly.
 
@@
 \caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
-%\vspace{-2mm}
+\vspace{-4mm}
 \end{figure}
@@
 Our new model has no effect on the manner in which traditional GPU
-workloads are programmed, such as graphics shaders in OpenGL, Vulkan
-and DirectX, and regular data-parallel computational kernels.
+workloads (e.g.\ graphics shaders and and regular
+data-parallel computational kernels) are programmed.
 %
@@
 case)---then a cooperative kernel will behave just like its
-non-cooperative counterpart would behave.  A developer
-for whom today's occupancy-bound execution model suffices can upgrade
-their kernel to a cooperative form and retain backwards-compatibility
-with today's GPU platforms by pre-processing away our language
-extensions.
+non-cooperative counterpart would behave.  
+%A developer
+%for whom today's occupancy-bound execution model suffices can upgrade
+%their kernel to a cooperative form and retain backwards-compatibility
+%with today's GPU platforms by pre-processing away our language
+%extensions.
 
@@
 in the runtimes of future GPU implementations of OpenCL, with driver
-and compiler support for our new primitives.  To experiment with our
-ideas on current GPUs, we have developed a prototype that mocks up the
-required runtime support via a \emph{megakernel}, and exploits the
+support for our new primitives.  To experiment with our ideas on
+current GPUs, we have developed a prototype that mocks up the required
+runtime support via a \emph{megakernel}, and exploits the
 occupancy-bound execution model that these GPUs provide to ensure fair
@@
 %
-For ease of implementation, our prototype supports
-multitasking a single cooperative kernel with a single non-cooperative
-kernel (though the non-cooperative kernel can be invoked many times).
-We
-assume all variables declared at the root lexical scope of a
-cooperative kernel are implicitly $\transmit$-annotated.
+For ease of implementation, our prototype supports multitasking a
+single cooperative kernel with a single non-cooperative kernel (though
+the non-cooperative kernel can be invoked many times).  
+%We assume all
+%variables declared at the root lexical scope of a cooperative kernel
+%are implicitly $\transmit$-annotated.
 %this allows our
@@
 %requiring front-end changes.  
-Finally, we require that $\offerkill$,
-$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
-function of a cooperative kernel.  This allows us to use
-$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
-of the kernel.  With these restrictions we can experiment
-with interesting irregular algorithms (see
-\mysec\ref{sec:experiments}).  A non-mock implementation of
-cooperative kernels would not use the megakernel approach, so we do not deem
-the engineering effort associated with lifting these restrictions to
-be worthwhile.
+We require that $\offerkill$, $\offerfork$ and
+$\resizingglobalbarrier$ are called from the entry function of a
+cooperative kernel.  This allows us to use $\keyword{goto}$ and
+$\keyword{return}$ to direct threads into and out of the kernel.  With
+these restrictions we can experiment with interesting irregular
+algorithms (see \mysec\ref{sec:experiments}).  A non-mock
+implementation of cooperative kernels would not use the megakernel
+approach, so we do not deem the engineering effort associated with
+lifting these restrictions to be worthwhile.
 
@@
 %
-To dynamically manage workgroups executing cooperative
-kernels, our framework must track the state of each workgroup and
-provide a channel of communication from the scheduler workgroup to
-workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
-we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available} or the id of the kernel that the workgroup is currently executing.
-The scheduler can then send cooperative
-kernels a \emph{resource message}, commanding workgroups to exit at $\offerkill$, or 
-spawn additional workgroups at $\offerfork$. Thus, the
-scheduler context needs a communication channel for each cooperative
-kernel. We implement the communication channels using atomic variables in global memory.
+To dynamically manage workgroups executing cooperative kernels, our
+framework must track the state of each workgroup and provide a channel
+of communication from the scheduler workgroup to workgroups executing
+$\offerfork$ and $\offerkill$. To achieve this, we use a
+\emph{scheduler context} structure, mapping a primitive workgroup id
+the workgroup's status, which is either \emph{available} or the id of
+the kernel that the workgroup is currently executing.  The scheduler
+can then send cooperative kernels a \emph{resource message},
+commanding workgroups to exit at $\offerkill$, or spawn additional
+workgroups at $\offerfork$. Thus, the scheduler context needs a
+communication channel for each cooperative kernel. We implement the
+communication channels using atomic variables in global memory.
 
@@
 To launch a kernel, the host sends a data packet to the GPU scheduler
-consisting of: a kernel to execute, the kernel inputs, and a
-flag indicating whether the kernel is cooperative. In our
-implementation, this host-device communication channel is built using
-fine-grained SVM atomics.
-
-On receiving a data packet
-describing a kernel launch $K$, the GPU scheduler must decide
- how to schedule $K$. Suppose $K$ requests
-$N$ workgroups. The scheduler queries the scheduler context.  If
-there are at least $N$ available workgroups, $K$ can be scheduled
-immediately. Suppose instead that there are only $N_a < N$ available workgroups, but a cooperative kernel $K_c$ is
-executing. The scheduler can use $K_c$'s channel in the scheduler context to command $K_c$ to provide $N - N_a$ workgroups via $\offerkill$.
-Once $N$ workgroups are available,
-% the scheduler initialises a
-%kernel context $\KC$ for $K$, setting the number of active workgroups
-%to $N$ and mapping available workgroups (located in the scheduler
-%context) to their new $\KC$ workgroup ids. 
-the scheduler then sends $N$
-workgroups from the available workgroups to execute kernel $K$
+consisting of: a kernel to execute, the kernel inputs, and a flag
+indicating whether the kernel is cooperative. In our implementation,
+this host-device communication channel is built using fine-grained SVM
+atomics.
+
+On receiving a data packet describing a kernel launch $K$, the GPU
+scheduler must decide how to schedule $K$. Suppose $K$ requests $N$
+workgroups. The scheduler queries the scheduler context.  If there are
+at least $N$ available workgroups, $K$ can be scheduled
+immediately. Suppose instead that there are only $N_a < N$ available
+workgroups, but a cooperative kernel $K_c$ is executing. The scheduler
+can use $K_c$'s channel in the scheduler context to command $K_c$ to
+provide $N - N_a$ workgroups via $\offerkill$.  Once $N$ workgroups
+are available,
+% the scheduler initialises a kernel context $\KC$ for $K$, setting
+%the number of active workgroups to $N$ and mapping available
+%workgroups (located in the scheduler context) to their new $\KC$
+%workgroup ids.
+the scheduler then sends $N$ workgroups from the available workgroups
+to execute kernel $K$
 % with
 %kernel context $\KC$.
-(If the new kernel $K$ is itself a cooperative kernel, the scheduler would be free to provide $K$ with fewer than $N$ active workgroups initially.)
-
-If a cooperative kernel $K_c$ is executing with fewer workgroups than it initially requested, the scheduler may decide make extra workgroups available
-to $K_c$, to be obtained next time $K_c$ calls $\offerfork$.  To do this, the scheduler asynchronously signals $K_c$ through $K_c$'s
-channel to indicate the number of
-workgroups that should join at the next $\offerfork$ command.
-When a workgroup $w$ of $K_c$ subsequently executes $\offerfork$, thread 0 of $w$ updates the kernel and scheduler contexts so that the given number of new workgroups are directed to the program point after the $\offerfork$ call.
-This involves selecting workgroups whose status is \emph{available}, as well as copying the values of $\transmit$-annotated variables to the new workgroups.
+(If the new kernel $K$ is itself a cooperative kernel, the scheduler
+would be free to provide $K$ with fewer than $N$ active workgroups
+initially.)
+
+If a cooperative kernel $K_c$ is executing with fewer workgroups than
+it initially requested, the scheduler may decide make extra workgroups
+available to $K_c$, to be obtained next time $K_c$ calls $\offerfork$.
+To do this, the scheduler asynchronously signals $K_c$ through $K_c$'s
+channel to indicate the number of workgroups that should join at the
+next $\offerfork$ command.  When a workgroup $w$ of $K_c$ subsequently
+executes $\offerfork$, thread 0 of $w$ updates the kernel and
+scheduler contexts so that the given number of new workgroups are
+directed to the program point after the $\offerfork$ call.  This
+involves selecting workgroups whose status is \emph{available}, as
+well as copying the values of $\transmit$-annotated variables to the
+new workgroups.
 
@@
 scheduler claims them---we know from $\keyword{query}$ that the
-scheduler will eventually do so.  We show in
+scheduler will eventually do so.  
+We show in
 \mysec\ref{sec:responsiveness} that the barrier using $\keyword{query}$ greatly
@@
 
-To work around compiler crash and compiler hang bugs in Intel's
-current OpenCL drivers, we had to apply some semantics-preserving
-changes to some applications.  Driver bugs also meant that we could
-not obtain results for some configurations where the non-cooperative
-kernel asks for all-but-one workgroups (namely bc, bfs, l-sssp, and
-color).  These configurations led to severe failures such as machine
-freezes and blue screens.
+%To work around compiler crash and compiler hang bugs in Intel's
+%current OpenCL drivers, we had to apply some semantics-preserving
+%changes to some applications.  Driver bugs also meant that we could
+%not obtain results for some configurations where the non-cooperative
+%kernel asks for all-but-one workgroups (namely bc, bfs, l-sssp, and
+%color).  These configurations led to severe failures such as machine
+%freezes and blue screens.
 
@@
 grows, the execution time decreases and the gather time
-increases.\footnote{The right figure has only 3 points per line:
-  recall that for bfs there are no runs with ``all-but-one''
-  non-cooperative workgroups.}
+increases.%\footnote{The right figure has only 3 points per line:
+  %recall that for bfs there are no runs with ``all-but-one''
+  %non-cooperative workgroups.}
 %

### 2017-02-17 write-ups/FSE/paper.tex
Typo only: remove trailing whitespaces
Simplify description of work-stealing exemple

--- 2017-02-15 write-ups/FSE/paper.tex
@@
 $\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
-is called by all threads of a workgroup.  Inside
-$\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
-the \emph{master thread}, tries to pop a
-task from the workgroup's queue, or to steal a task from a different
-queue. The master thread uses local memory inside
-$\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
-all threads of its workgroup. If a task was obtained, then all the
-workgroup threads participate in the task processing (line~\ref{line:wksteal:processtask}).
-This may lead to the creation of further tasks, which
-the master thread pushes to the workgroup's queue.
+pops a task from the workgroup's queue or, if the queue is empty,
+tries to steal a task from other queues.
+
+% is called by all threads of a workgroup.  Inside
+% $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
+% the \emph{master thread}, tries to pop a
+% task from the workgroup's queue, or to steal a task from a different
+% queue. The master thread uses local memory inside
+% $\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
+% all threads of its workgroup.
+
+If a task was obtained, then the workgroup processes it
+(line~\ref{line:wksteal:processtask}): during this operation, new tasks
+may be created and pushed into the workgroup's queue.
 

### 2017-02-23 write-ups/FSE/paper.tex
Small changes.

@@
 
+\ADComment{Update intro to be clearer that even with preemption our approach has merit, and that preemption does not imply fairness.}
+
+\ADComment{Use NVIDIA/Pascal as motivation for our approach.}
+
 \myparagraph{The needs of irregular data-parallel algorithms}
@@
 indefinitely at a global barrier so that other workgroups do not reach the barrier.
+\ADComment{Tyler: briefly state that this still occurs on Pascal, despite it having support for preemption.}
 
@@
 
+\subsection{Comparison with Preemption on NVIDIA GPUs}\label{sec:nvidiacomparison}
+
+\ADComment{Make the point that NVIDIA GPUs now support
+  instruction-level preemption in hardware; allows arbitrary
+  preemption.}
+
+\ADComment{Note that they are doing coarse-grained, task-level
+  preemption, rather than running multiple tasks concurrently on the
+  GPU.}
+
+\ADComment{Direct comparison is not possible as we cannot implement
+  cooperative kernels on NVIDIA devices due to lack of mature OpenCL
+  2.0 support.}
+
+\ADComment{However, based on results reported by NVIDIA and a clock
+  speed calculation we can estimate...}
+
 \section{Related Work}\label{sec:relatedwork}
@@
 \myparagraph{Preemptive multitasking on GPUs}
+%
+\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
 %

### 2017-02-23 write-ups/FSE/paper.tex
Add overview figure
did a little bit of work

@@
 \ADComment{Tyler: briefly state that this still occurs on Pascal, despite it having support for preemption.}
+\TSComment{I added this but actually think it fits better a little bit later (marked with a comment). See what you think}
 
@@
 efficiently.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{overview.pdf}
+\caption{Cooperative kernels can flexibly be resized in order to let
+other tasks (e.g. graphics) access GPU ressources}
+\label{fig:overview}
+\end{figure}
 
@@
 cases.
+%
+\TSComment{added note about Pascal experiment here. I wonder if there's
+a better way to glue these paragraphs together? That is, above it provides
+two issues vendors have with forward progress and below, we show a concrete
+example of both issues in Nvidia's latest chips}
+%
+For example, Nvidia's latest architecture (Pascal) is documented to
+support instruction-level preemption~\cite{??}. Despite this, Nvidia's
+documentation still does not provide concrete forward progress
+properties (i.e. vendors will not yet commit to providing these
+properties). Furthermore, we were able to run some small experiments
+on a Pascal GPU (GTX Titan X) and observed that the GPU still adhered
+to the occupancy-bound execution model. That is, a kernel containing a
+global barrier executed successfully when run with 56 workgroups, but
+deadlocked when run with 57 workgroups. This is an example of a vendor
+supporting preemption, yet not providing full fairness properties.
 
@@
 mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
-\rowcolor{Gray1}
-bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
+%\rowcolor{Gray1}
+%bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
 \rowcolor{Gray1}
@@
 
-\subsection{Comparison with Preemption on NVIDIA GPUs}\label{sec:nvidiacomparison}
+\subsection{Preemption on NVIDIA Comparison}\label{sec:nvidiacomparison}
 
@@
 A number of works have considered how best to schedule dynamic
-workloads on GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
-Among these, the Whippletree approach employs a persistent
-megakernel to schedule multiple, dynamic tasks in a manner that aims
-to best utilise GPU
-resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and the TimeGraph
-approach similarly aims to optimise scheduling of competing workloads,
-in the context of OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.
-None of these approaches tackle
-the problem of \emph{fair} scheduling on GPUs, thus they do not aid in safe deployment of blocking irregular algorithms.
+workloads on
+GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
+Among these, the Whippletree approach employs a persistent megakernel
+to schedule multiple, dynamic tasks in a manner that aims to best
+utilise GPU resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and
+the TimeGraph approach similarly aims to optimise scheduling of
+competing workloads, in the context of
+OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.  None of these approaches
+tackle the problem of \emph{fair} scheduling on GPUs, thus they do not
+aid in safe deployment of blocking irregular algorithms.
 

### 2017-02-24 write-ups/FSE/paper.tex
Shortening with Tyler.

--- 2017-02-19 write-ups/FSE/paper.tex
@@
 
-\newcommand{\myparagraph}[1]{\paragraph{#1}}
+\newcommand{\myparagraph}[1]{\paragraph{#1.}}
 %{\noindent{\bf\emph{#1}}\hspace{1em}}
@@
 by the computation itself.  There is growing interest in accelerating
-such algorithms on GPUs, particularly the ones that
-traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
+such algorithms on GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
 
 Irregular algorithms usually require \emph{blocking synchronization}
-between workgroups, to balance load and communicate intermediate
-results.  For example, many graph algorithms employ a level-by-level
+between workgroups.  For example, many graph algorithms employ a level-by-level
 strategy, requiring a global barrier between each level, and work
@@
 
-To operate correctly, a blocking concurrent algorithm requires
-\emph{fair} scheduling of workgroups.  Without fairness, a
-workgroup may fail to make progress, leading to starvation.  For
+To avoid starvation, a blocking concurrent algorithm requires
+\emph{fair} scheduling of workgroups.  For
 example, if one workgroup holds a mutex, an unfair scheduler may cause
@@
 %\vspace{-1mm}
-\myparagraph{A degree of fairness: occupancy-bound execution} GPU
-schedulers are \emph{not} fair. The current GPU programming
+\myparagraph{A degree of fairness: occupancy-bound execution} The current GPU programming
 models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
 HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding
-scheduling of workgroups, and current GPU scheduler are unfair in
+scheduling of workgroups, and current GPU schedulers are unfair in
 practice.  Roughly speaking, each workgroup executing a GPU kernel is
@@
 become \emph{occupant}), the workgroup has exclusive access to the
-compute unit until it finishes execution.  Experiments suggest that
-the occupancy-bound execution model is widely implemented by today's
+compute unit until it finishes execution.  
+This is simple model is easy to implement
+efficiently, and experiments suggest that
+it is widely employed by today's
 GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.
-It is evidently a simple model for device vendors to implement
-efficiently.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{overview.pdf}
+\caption{Cooperative kernels can flexibly be resized in order to let
+other tasks (e.g. graphics) access GPU ressources}
+\label{fig:overview}
+\end{figure}
 
@@
 execution model, which they exploit by launching no more workgroups
-than there are available compute units~\cite{owens-persistent}. This
-works because \emph{today}'s GPUs seem to provide the occupancy-bound
-execution model.
+than there are available compute units~\cite{owens-persistent}.
 
@@
 
+For example, Nvidia's latest architecture (Pascal) is documented to
+support instruction-level preemption~\cite{??}. Despite this, Nvidia's
+most recent compute capablity does not guarantee forward progress for
+CUDA programmers (evidence that vendors will not yet commit to providing such guarantees).
+Indeed, on a GTX Titan X (Pascal) we still observe starvation:
+we found that a global barrier can be executed successfully with 56 workgroups, but
+deadlocks with 57 workgroups.  This is an example of a vendor
+supporting preemption, yet not guaranteeing fair scheduling.
+
 %\vspace{-1mm}
@@
   execution model that current GPUs provide: our prototype implementation requires
-  no additional hardware or driver support.  Thus the model should be easy for vendors to implement directly.
+  no additional hardware or driver support.
+
+\item If hardware support for preemption is available it can be leveraged to implement cooperative kernels efficiently, and cooperative kenrnels can avoid the overhead of unnecessary preemptions by allowing the programmer to communicate ``smart'' preemption points.
 
@@
 
-\subsection{Backwards Compatibility}\label{sec:backwardscompatibility}
-
-Our new model has no effect on the manner in which traditional GPU
-workloads (e.g.\ graphics shaders and and regular
-data-parallel computational kernels) are programmed.
-%
-Furthermore, if our cooperative language extensions are
-ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
-$\transmit$ annotations are ignored, and $\globalbarrier$ and
-$\resizingglobalbarrier$ calls are redirected to an existing global
-barrier implementation (so that no resizing occurs in the latter
-case)---then a cooperative kernel will behave just like its
-non-cooperative counterpart would behave.
-%A developer
-%for whom today's occupancy-bound execution model suffices can upgrade
-%their kernel to a cooperative form and retain backwards-compatibility
-%with today's GPU platforms by pre-processing away our language
-%extensions.
-
 \section{Prototype Implementation}\label{sec:implementation}
@@
 
+\subsection{Comparison with Task Level Preemption}\label{sec:nvidiacomparison}
+
+Some recent GPUs, including NVIDIA's Pascal architecture, provide
+hardware support for instruction-level preemption.  Few details of how
+this works are available, however from reading NVIDIA documentation
+and online material~\cite{ }, and running litmus tests on a Pascal
+platform, our understanding is that preemption of entire kernels, but
+not of individual workgroups running a kenrel is supported.
+Intel---the only vendor whose OpenCL implemenation is capable enough
+to implement cooperative kernels---do not provide this feature, thus
+we cannot make a direct comparison.  We present here a theoretical
+analysis of the overheads associated with sharing the GPU between
+graphics and compute tasks via kernel-level preemption.
+
+Let us assume that a graphics workload is required to be scheduled
+with period $P$ and duration $D$, and that a long-running compute
+kernel requires time $C$ to execute without interruption.  If we
+assume the cost of preemption is negligible (e.g.\ NVIDIA have
+reported preemption times of 0.1 ms for Pascal), then the overhead
+associated with switching between compute and graphics every $P$ time
+steps is $P/(P-D)$.
+
+\ADComment{Tyler to complete.}
+
 \section{Related Work}\label{sec:relatedwork}
@@
 %
+\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
+%
 Hardware support for preemption has been proposed for \nvidia GPUs,
@@
 \end{document}
+
+\subsection{Backwards Compatibility}\label{sec:backwardscompatibility}
+
+Our new model has no effect on the manner in which traditional GPU
+workloads (e.g.\ graphics shaders and and regular
+data-parallel computational kernels) are programmed.
+%
+Furthermore, if our cooperative language extensions are
+ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
+$\transmit$ annotations are ignored, and $\globalbarrier$ and
+$\resizingglobalbarrier$ calls are redirected to an existing global
+barrier implementation (so that no resizing occurs in the latter
+case)---then a cooperative kernel will behave just like its
+non-cooperative counterpart would behave.
+%A developer
+%for whom today's occupancy-bound execution model suffices can upgrade
+%their kernel to a cooperative form and retain backwards-compatibility
+%with today's GPU platforms by pre-processing away our language
+%extensions.

### 2017-02-25 write-ups/FSE/paper.tex
Update overview figure
finished writing the bit about comparisons, added text around hugues figure

@@
 \includegraphics[width=\columnwidth]{overview.pdf}
-\caption{Cooperative kernels can flexibly be resized in order to let
-other tasks (e.g. graphics) access GPU ressources}
+\caption{Cooperative kernels can flexibly resize to let run other tasks
+(e.g. graphics) concurrently}
 \label{fig:overview}
@@
 For example, Nvidia's latest architecture (Pascal) is documented to
-support instruction-level preemption~\cite{??}. Despite this, Nvidia's
-most recent compute capablity does not guarantee forward progress for
-CUDA programmers (evidence that vendors will not yet commit to providing such guarantees).
-Indeed, on a GTX Titan X (Pascal) we still observe starvation:
-we found that a global barrier can be executed successfully with 56 workgroups, but
-deadlocks with 57 workgroups.  This is an example of a vendor
-supporting preemption, yet not guaranteeing fair scheduling.
+support instruction-level preemption~\cite{PascalWhitepaper}. Despite
+this, Nvidia's most recent compute capablity does not guarantee
+forward progress for CUDA programmers (evidence that vendors will not
+yet commit to providing such guarantees).  Indeed, on a GTX Titan X
+(Pascal) we still observe starvation: we found that a global barrier
+can be executed successfully with 56 workgroups, but deadlocks with 57
+workgroups.  This is an example of a vendor supporting preemption, yet
+not guaranteeing fair scheduling.
 
@@
 The use of $\offerfork$ and $\offerkill$ creates a contract between
-the scheduler and the cooperative kernel.
-Functionally, the
-scheduler must guarantee that the workgroups executing a cooperative
-kernel are fairly scheduled, while the cooperative kernel must be
-robust to workgroups leaving and joining the computation in response
-to $\offerkill$ and $\offerfork$.  Non-functionally, a cooperative
-kernel must ensure that $\offerkill$ is executed frequently enough
-such that the scheduler can accommodate soft-real time constraints,
+the scheduler and the cooperative kernel.  Functionally, the scheduler
+must guarantee that the workgroups executing a cooperative kernel are
+fairly scheduled, while the cooperative kernel must be robust to
+workgroups leaving and joining the computation in response to
+$\offerkill$ and $\offerfork$.  Non-functionally, a cooperative kernel
+must ensure that $\offerkill$ is executed frequently enough such that
+the scheduler can accommodate soft-real time constraints,
 e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
@@
 additional workgroups when possible.
+
+\myfiglong{\ref{fig:overview}} shows an example execution of how the
+GPU compute units might be shared between a cooperative kernel and a
+graphics task. Notice that workgroups executing the cooperative kernel
+are killed at a $\offerkill$ command to make room for a graphics
+task. After the graphics task, the previously killed workgroups are
+restored to executing the cooperative kernel when a workgroup
+executing the cooperative kernel calls $\offerfork$. The order in
+which workgroups are killed is important and discussed in
+\mysec{\ref{sec:semantics}}. The \emph{gather time} is the time
+between when resources are requested and when the application has
+surrendered the resources via $\offerkill$. To satisfy soft-real time
+constraints, this time should be low; our experimental study
+(\mysec{\ref{sec:responsiveness}}) shows that in practice, the
+gather-time for our applications is acceptable for a range of graphics
+workloads.
 
@@
 (\mysec\ref{sec:semantics}), use our motivating examples to discuss
-programmability (\mysec\ref{sec:programmingguidelines}), outline
+programmability (\mysec\ref{sec:programmingguidelines}) and outline
 important nonfunctional properties that the model requires to work
-successfully (\mysec\ref{sec:nonfunctional}) and explain that the
-model is backwards-compatible
-(\mysec\ref{sec:backwardscompatibility}).
-%
-In Appendices~\ref{appendix:semantics} and~\ref{appendix:semanticalternatives} (supplementary material)
-we present a formal semantics for cooperative kernels atop an abstract GPU programming model,
-and discuss several cases where we
-might have taken different and also reasonable semantic decisions.
+successfully (\mysec\ref{sec:nonfunctional}).
+%and explain that the
+%model is backwards-compatible
+%(\mysec\ref{sec:backwardscompatibility}).
+%
+In Supplementary material we present a formal semantics for
+cooperative kernels atop an abstract GPU programming model, and
+discuss several cases where we might have taken different and also
+reasonable semantic decisions.
 
@@
 The number of \emph{active workgroups}---workgroups executing the
-kernel---is denoted $M$.  Active workgroups
-have consecutive ids in the range $[0, M-1]$.
-Initially, at least one workgroup is active;
-if necessary the scheduler must postpone the kernel until
-some compute unit becomes available.
+kernel---is denoted $M$.  Active workgroups have consecutive ids in
+the range $[0, M-1]$.  Initially, at least one workgroup is active; if
+necessary the scheduler must postpone the kernel until some compute
+unit becomes available. For example, in \myfig{\ref{fig:overview}: at the
+  beginning of the execution $M = 4$; while the graphics task is
+  executing $M = 2$; after the fork $M = 4$ again.
 
@@
 
-Suppose a workgroup with id $m$ executes $\offerkill$.
-If the workgroup has the highest largest id of all active workgroups then it can be killed by the scheduler, except that workgroup 0 can never be killed (to ensure that the kernel computation does not terminate early).
-More formally, if $m < M-1$
-or $M=1$ then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
-M-1$, the scheduler can choose to ignore the offer, so that $\offerkill$
-executes as a no-op, or accept the offer, so that execution of the workgroup ceases and
-the number of active workgroups $M$ is atomically decremented by one.
+Suppose a workgroup with id $m$ executes $\offerkill$.  If the
+workgroup has the highest largest id of all active workgroups then it
+can be killed by the scheduler, except that workgroup 0 can never be
+killed (to ensure that the kernel computation does not terminate
+early).  More formally, if $m < M-1$ or $M=1$ then $\offerkill$ is a
+no-op.  If instead $M > 1$ and $m = M-1$, the scheduler can choose to
+ignore the offer, so that $\offerkill$ executes as a no-op, or accept
+the offer, so that execution of the workgroup ceases and the number of
+active workgroups $M$ is atomically decremented by one.
+\myfiglong{\ref{fig:overview}} illustrates this, showing that
+workgroup $3$ is killed before workgroup $2$.
 %In the second case, the scheduler can assign some other task to
@@
 %
-Recall that
-a desired limit of $N$ workgroups was specified when the cooperative kernel was launched, but that the number of active workgroups, $M$, may be smaller
-than $N$, either because (due to competing workloads) the scheduler
-did not provide $N$ workgroups initially, or because the kernel has
-given up some workgroups via $\offerkill$ calls.  Through the
-$\offerfork$ primitive (also a workgroup-level function), the kernel and scheduler can collaborate to allow new
-workgroups to join the computation at an appropriate point and with
-appropriate state.
-
-Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  Then the following occurs: an
-integer $k \in [0, N-M]$ is chosen by the scheduler;
-$k$ new workgroups are spawned with consecutive ids in the range $[M,
-  M+k-1]$; the active workgroup count $M$ is atomically incremented by $k$.
+Recall that a desired limit of $N$ workgroups was specified when the
+cooperative kernel was launched, but that the number of active
+workgroups, $M$, may be smaller than $N$, either because (due to
+competing workloads) the scheduler did not provide $N$ workgroups
+initially, or because the kernel has given up some workgroups via
+$\offerkill$ calls.  Through the $\offerfork$ primitive (also a
+workgroup-level function), the kernel and scheduler can collaborate to
+allow new workgroups to join the computation at an appropriate point
+and with appropriate state.
+
+Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  Then the
+following occurs: an integer $k \in [0, N-M]$ is chosen by the
+scheduler; $k$ new workgroups are spawned with consecutive ids in the
+range $[M, M+k-1]$; the active workgroup count $M$ is atomically
+incremented by $k$.
 
@@
 
+\myfiglong{\ref{fig:overview}} illustrates the behaviour of
+$\offerfork$. After the graphics task finishes executing, there is
+space for two workgroups to join the computation of the cooperative
+kernel. Workgroup $0$ calls $\offerfork$, spawning the two new
+workgroups with ids $2$ and $3$ and broadcasting the transmit
+variables to them. Workgroups $2$ and $3$ join the computation where
+workgroup $0$ called $\offerfork$.
+
 %We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
@@
 Notice that $k=0$ is always a valid choice for the number of
-workgroups to be spawned by $\offerfork$, and is guaranteed if $M$
-is equal to the workgroup limit $N$.
+workgroups to be spawned by $\offerfork$, and is guaranteed if $M$ is
+equal to the workgroup limit $N$.
 
@@
 
-\mytablong\ref{tab:applications} gives an overview of the 8 irregular
-algorithms that we ported to cooperative kernels. Among them, 6 are
+\mytablong\ref{tab:applications} gives an overview of the 7 irregular
+algorithms that we ported to cooperative kernels. Among them, 5 are
 graph algorithms, based on the Pannotia~\cite{Pannotia} and
@@
 resizing barriers, and how many variables need to be transmitted.  In
-all cases all barriers were converted, except for bc which contains
-barriers deeper in the call stack, a case not supported by our
-prototype although these barriers could in principle be converted. The
-remaining 2 algorithms are work stealing applications: each required
-the addition of $\offerfork$ and $\offerkill$ at the start of the main
-loop, and no variables needed to be transmitted.  For all these
-examples, the porting to cooperative kernels was similarly
-straightforward to the porting illustrated in
-\mysec\ref{sec:programmingguidelines}. Most
-graph come with 2 different data sets as input, leading to 13
-application/input pairs in total.
-
-From the Lonestar suite we omit two applications that use global
-barriers: dmr has been shown to be unstable across GPU
-vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially
-short-running so that it is not interesting for our multitasking
-approach.
+all cases all barriers were converted. The remaining 2 algorithms are
+work stealing applications: each required the addition of $\offerfork$
+and $\offerkill$ at the start of the main loop, and no variables
+needed to be transmitted.  For all these examples, the porting to
+cooperative kernels was similarly straightforward to the porting
+illustrated in \mysec\ref{sec:programmingguidelines}. Most graph come
+with 2 different data sets as input, leading to 11 application/input
+pairs in total.
+
+%From the Lonestar suite we omit two applications that use global
+%barriers: dmr has been shown to be unstable across GPU
+%vendors~\cite{DBLP:conf/iwocl/SorensenD16}, and mst is trivially
+%short-running so that it is not interesting for our multitasking
+%approach.
 
@@
 %\vspace{-2mm}
-\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
+\caption{Example gather time and non-cooperative timing results \TSComment{need to get updated numbers}}\label{fig:fine-grained-timing}
 \vspace{-3mm}
@@
 \hline
-$1.08$ & $1.37^{\ddagger}$ & $1.07$ & $1.37^{\ddagger}$ & $1.12$ & $1.23^{\diamond}$ \\
+$1.09$ & $1.23^{\ddagger}$ & $1.09$ & $1.21^{\diamond}$ & $1.11$ & $1.23^{\ddagger}$ \\
 \end{tabular}\\
 {\small
- $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
+ $^{\ddagger}$octree, $^{\diamond}$sssp usa\_ny
 }
 %\vspace{-2mm}
-\caption{Cooperative kernel slowdown w/o multitasking}
+\caption{Cooperative kernel slowdown w/o multitasking \TSComment{need to get updated numbers}}
 \label{tab:overhead}
@@
 %\vspace{-2mm}
-\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
+\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute \TSComment{need to get updated numbers}}\label{fig:performance}
 \vspace{-5mm}
@@
 
-Co-scheduling non-cooperative kernels that request a single workgroup leads to almost no overhead, but the period is far too high to meet the needs
-of any of our three workloads; e.g.\ a heavy workload averages a
-period of 939 ms. As more workgroups are dedicated to non-cooperative
-kernels, they execute quickly enough to be scheduled at the expected
-period. For the light and medium workloads, a quarter
-of the workgroups executing the non-cooperative kernel are able to meet
-their goal period (70 and 40 ms resp.). However, a quarter of the
+Co-scheduling non-cooperative kernels that request a single workgroup
+leads to almost no overhead, but the period is far too high to meet
+the needs of any of our three workloads; e.g.\ a heavy workload
+averages a period of 939 ms. As more workgroups are dedicated to
+non-cooperative kernels, they execute quickly enough to be scheduled
+at the expected period. For the light and medium workloads, a quarter
+of the workgroups executing the non-cooperative kernel are able to
+meet their goal period (70 and 40 ms resp.). However, a quarter of the
 workgroups are not sufficient to meet the goal for the heavy workload
 (giving a mean period of 88 ms). If half of the workgroups are
-allocated to the non-cooperative kernel, the heavy workload averages $\sim$10\% over its goal period (mean of 44 ms). If all-but-one are allocated, the heavy workload reaches its goal period.
+allocated to the non-cooperative kernel, the heavy workload averages
+$\sim$10\% over its goal period (mean of 44 ms). If all-but-one are
+allocated, the heavy workload reaches its goal period.
 %
@@
 
+\begin{table}[t]
+\normalsize
+\centering
+\begin{tabular}{ l r r}
+g. workload &  kernel-level  & cooperative (resources) \\
+\hline
+light & 1.04 & 1.XX($N/4$)\\
+medium & 1.08 & 1.XX($N/4$)\\
+heavy & 1.33 & 1.XX($N-1$)\\
+\end{tabular} \\
+%\vspace{.2cm}
+
+\caption{Overhead of kernel level preemption vs cooperative kernels for three
+graphics workloads \TSComment{need to get updated numbers}}
+\label{tab:preemption}
+\vspace{-8mm}
+\end{table}
+
 Some recent GPUs, including NVIDIA's Pascal architecture, provide
@@
 this works are available, however from reading NVIDIA documentation
-and online material~\cite{ }, and running litmus tests on a Pascal
-platform, our understanding is that preemption of entire kernels, but
-not of individual workgroups running a kenrel is supported.
-Intel---the only vendor whose OpenCL implemenation is capable enough
-to implement cooperative kernels---do not provide this feature, thus
-we cannot make a direct comparison.  We present here a theoretical
-analysis of the overheads associated with sharing the GPU between
-graphics and compute tasks via kernel-level preemption.
+and online material~\cite{PascalWhitepaper,anandtech}, and running
+litmus tests on a Pascal platform, our understanding is that
+preemption of entire kernels, but not of individual workgroups running
+a kenrel is supported.  Intel---the only vendor whose OpenCL
+implemenation is capable enough to implement cooperative kernels---do
+not provide this feature, thus we cannot make a direct comparison.  We
+present here a theoretical analysis of the overheads associated with
+sharing the GPU between graphics and compute tasks via kernel-level
+preemption.
 
@@
 associated with switching between compute and graphics every $P$ time
-steps is $P/(P-D)$.
-
-\ADComment{Tyler to complete.}
+steps is $P/(P-D)$. We compare this task-level preemption overhead
+model with our experimental results per graphics workload in
+\mytab{\ref{tab:preemption}}.
+
+We see that our approach provides lower overhead for low and medium
+graphics workloads. Recall for the heavy graphics workload, we are
+within $10\%$ of the required deadline ($40$ ms) by using half the
+workgroups with an overhead of $1.XX$, which is less than the
+kernel-level preemption overhead of $1.33$. \TSComment{need to get
+  updated numbers}
+
+The low overhead of cooperative kernels for low and medium graphics
+workloads are due to the irregularity of the cooperative kernels and
+the property that they do not scale linearly. Thus some resources can
+be taken away during computation without proportionally reducing
+execution time.  Our poor performance for the heavy workload is
+because the graphics task requires all but one workgroup, which
+cripples the cooperative kernel enough that $\offerfork$ calls are not
+issued as frequently.
+
+These results suggest that a hybrid preemption scheme may work the
+best for multi-tasking and graphics. That is, the cooperative approach
+works well for light and medium tasks, but heavy graphics tasks
+benefit from the coarser grained, kernel-level preemption strategy.
 
@@
 
-\myparagraph{Preemptive multitasking on GPUs}
-%
-\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
-%
-Hardware support for preemption has been proposed for \nvidia GPUs,
-as well as \emph{SM-draining}
-whereby workgroups occupying a symmetric multiprocessor (SM; a compute
-unit using our terminology) are allowed to complete until the SM becomes free for
-other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited the presence of blocking constructs, since it may not be possible to drain a blocked workgroup.
-%
-A follow-up work adds the notion of SM \emph{flushing},
-where a workgroup can be re-scheduled
-from scratch if it has not yet committed side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
-using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs support
-preemption, though it is not clear whether they guarantee fairness~\cite{PascalWhitepaper}.
-
-The advantages of preemption are that it does not require programmer
-effort, and guards against a rogue long-running kernel making the GPU
-unavailable for other tasks.  The main drawback is that \emph{every}
-kernel takes a potential performance hit.
-%
-If implemented natively, our solution does not impact performance on regular kernels that are not scheduled with cooperative kernels.
-Furthermore, careful placement of
-our new primitives can exploit problem structure, enabling more efficient multitasking than
-application-oblivious preemption.  There is scope for combining the approaches: if a
-cooperative kernel is insufficiently cooperative, hardware support for preemption could
-be invoked to prevent the kernel from denying other tasks access to
-the GPU indefinitely.
+\myparagraph{GPU multitasking and scheduling}
+%
+%\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
+%
+Hardware support for preemption has been proposed for \nvidia GPUs, as
+well as \emph{SM-draining} whereby workgroups occupying a symmetric
+multiprocessor (SM; a compute unit using our terminology) are allowed
+to complete until the SM becomes free for other
+tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited
+the presence of blocking constructs, since it may not be possible to
+drain a blocked workgroup.
+%
+A follow-up work adds the notion of SM \emph{flushing}, where a
+workgroup can be re-scheduled from scratch if it has not yet committed
+side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have
+been evaluated using simulators, over sets of regular GPU kernels.
+Very recent \nvidia GPUs (i.e. the Pascal architecture) support
+preemption, though, as discussted in \mysec{sec:intro} and
+\mysec{sec:nvidiacomparison}, it is not clear whether they guarantee
+fairness or allow tasks to share GPU resources at the workgroup
+level~\cite{PascalWhitepaper}.
+
+CUDA and OpenCL provide the facility for a kernel to spawn further
+kernels~\cite{cuda-75}.  This \emph{dynamic parallelism} can be used
+to implement a GPU-based scheduler, by having an initial scheduler
+kernel repeatedly spawn further kernels as required, according to some
+scheduling policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However,
+kernels that uses dynamic parallelism are still prone to unfair
+scheduling of workgroups, and thus does not help in deploying blocking
+algorithms on GPUs.
+
+%The advantages of preemption are that it does not require programmer
+%effort, and guards against a rogue long-running kernel making the GPU
+%unavailable for other tasks.  The main drawback is that \emph{every}
+%kernel takes a potential performance hit.
+%
+%If implemented natively, our solution does not impact performance on
+%regular kernels that are not scheduled with cooperative kernels.
+%Furthermore, careful placement of our new primitives can exploit
+%problem structure, enabling more efficient multitasking than
+%application-oblivious preemption.  There is scope for combining the
+%approaches: if a cooperative kernel is insufficiently cooperative,
+%hardware support for preemption could be invoked to prevent the kernel
+%from denying other tasks access to the GPU indefinitely.
 
@@
 
-\myparagraph{Optimizing GPU scheduling}
-%
-A number of works have considered how best to schedule dynamic
-workloads on
-GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
-Among these, the Whippletree approach employs a persistent megakernel
-to schedule multiple, dynamic tasks in a manner that aims to best
-utilise GPU resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and
-the TimeGraph approach similarly aims to optimise scheduling of
-competing workloads, in the context of
-OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.  None of these approaches
-tackle the problem of \emph{fair} scheduling on GPUs, thus they do not
-aid in safe deployment of blocking irregular algorithms.
-
-CUDA and OpenCL provide the facility for a kernel to spawn further
-kernels~\cite{cuda-75}.  This \emph{dynamic parallelism}
-can be used to implement a GPU-based scheduler, by having an initial
-scheduler kernel repeatedly spawn further kernels as required,
-according to some scheduling
-policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, kernels that
-uses dynamic parallelism are still prone to unfair scheduling of
-workgroups, and thus does not help in deploying blocking algorithms on
-GPUs.
+%\myparagraph{Optimizing GPU scheduling}
+%
+%A number of works have considered how best to schedule dynamic
+%workloads on
+%GPUs~\cite{DBLP:journals/tog/SteinbergerKKHKS12,DBLP:conf/ics/WuCLSV15,DBLP:conf/usenix/KatoLRI11,DBLP:journals/tog/SteinbergerKBKDS14,DBLP:journals/rts/ElliottA12,DBLP:conf/sosp/RossbachCSRW11}.
+%Among these, the Whippletree approach employs a persistent megakernel
+%to schedule multiple, dynamic tasks in a manner that aims to best
+%utilise GPU resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}, and
+%the TimeGraph approach similarly aims to optimise scheduling of
+%competing workloads, in the context of
+%OpenGL~\cite{DBLP:conf/usenix/KatoLRI11}.  None of these approaches
+%tackle the problem of \emph{fair} scheduling on GPUs, thus they do not
+%aid in safe deployment of blocking irregular algorithms.
 

### 2017-02-26 write-ups/FSE/paper.tex
Update figure overview
Pass over new section.

@@
 \includegraphics[width=\columnwidth]{overview.pdf}
-\caption{Cooperative kernels can flexibly resize to let run other tasks
-(e.g. graphics) concurrently}
+\caption{Cooperative kernels can flexibly resize to let other tasks
+(e.g. graphics) run concurrently}
 \label{fig:overview}
@@
 
-Some recent GPUs, including NVIDIA's Pascal architecture, provide
+Some recent GPUs, including Nvidia's Pascal architecture, provide
 hardware support for instruction-level preemption.  Few details of how
-this works are available, however from reading NVIDIA documentation
+this works are available, but from reading Nvidia documentation
 and online material~\cite{PascalWhitepaper,anandtech}, and running
@@
 preemption of entire kernels, but not of individual workgroups running
-a kenrel is supported.  Intel---the only vendor whose OpenCL
-implemenation is capable enough to implement cooperative kernels---do
-not provide this feature, thus we cannot make a direct comparison.  We
-present here a theoretical analysis of the overheads associated with
+a kernel is supported.  Intel GPUs do
+not provide this feature, and out OpenCL 2.0 prototype of cooperative kernels cannot run on Nvidia GPUs, making a direct comparison impossible.
+We present here a theoretical analysis of the overheads associated with
 sharing the GPU between graphics and compute tasks via kernel-level
@@
 kernel requires time $C$ to execute without interruption.  If we
-assume the cost of preemption is negligible (e.g.\ NVIDIA have
-reported preemption times of 0.1 ms for Pascal), then the overhead
+assume the cost of preemption is negligible (e.g.\ Nvidia have
+reported preemption times of 0.1 ms for Pascal~\cite{TODO}), then the overhead
 associated with switching between compute and graphics every $P$ time
-steps is $P/(P-D)$. We compare this task-level preemption overhead
+steps is $P/(P-D)$.
+
+We compare this task-level preemption overhead
 model with our experimental results per graphics workload in
 \mytab{\ref{tab:preemption}}.
-
-We see that our approach provides lower overhead for low and medium
+%
+Based on the above assumptions, our approach provides lower overhead for low and medium
 graphics workloads. Recall for the heavy graphics workload, we are
@@
 kernel-level preemption overhead of $1.33$. \TSComment{need to get
-  updated numbers}
+  updated numbers} \ADComment{I don't quite understand the wording; sounds a bit like we still win with the heavy workload.}
 
 The low overhead of cooperative kernels for low and medium graphics
-workloads are due to the irregularity of the cooperative kernels and
-the property that they do not scale linearly. Thus some resources can
+workloads is due to sub-linear scalability of the irregular algorithms that they implement: some resources can
 be taken away during computation without proportionally reducing

### 2017-02-26 write-ups/FSE/paper.tex
Shortenings.

@@
 
-\newcommand{\myparagraph}[1]{\paragraph{#1.}}
+%\newcommand{\myparagraph}[1]{\paragraph{#1.}}
+\newcommand{\myparagraph}[1]{\vspace{1mm}\noindent\emph{\textbf{#1.}}}
 %{\noindent{\bf\emph{#1}}\hspace{1em}}
@@
 To summarise: blocking algorithms
-demand fair scheduling of workgroups, but for good reasons
+demand fair scheduling,
+% of workgroups,
+but for good reasons
 GPU vendors will not commit even to the guarantees of the
 occupancy-bound execution model.
-
+%
 We propose \emph{cooperative kernels}, an extension to the GPU
-programming model that aims to resolve this impasse.  A kernel
+programming model that aims to resolve this impasse.
+
+A kernel
 that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
 language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.
-
-At a point where the cooperative kernel could proceed with fewer
+%
+Where the cooperative kernel could proceed with fewer
 workgroups, a workgroup can execute $\offerkill$, offering to
@@
 %
-At a point where the cooperative kernel could benefit from having
-additional workgroups join the computation, a workgroup can execute
+Where the cooperative kernel could use additional resources, a workgroup can execute
 $\offerfork$ to indicates that the
@@
 the scheduler can accommodate soft-real time constraints,
-e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
-units to be powered down quickly when required.  The scheduler should
-respond to $\offerkill$ and $\offerfork$ calls in a manner that avoids
-under-utilisation of hardware resources by the cooperative kernel,
-e.g.\ the scheduler should not kill workgroups unless they are
-required for another purpose, and should facilitate forking of
-additional workgroups when possible.
-
-\myfiglong{\ref{fig:overview}} shows an example execution of how the
-GPU compute units might be shared between a cooperative kernel and a
-graphics task. Notice that workgroups executing the cooperative kernel
-are killed at a $\offerkill$ command to make room for a graphics
-task. After the graphics task, the previously killed workgroups are
-restored to executing the cooperative kernel when a workgroup
-executing the cooperative kernel calls $\offerfork$. The order in
+e.g.\ allowing a smooth frame-rate for graphics.
+%, or allowing compute
+%units to be powered down quickly when required.
+In return, the scheduler should allow the cooperative kernel to utilise
+hardware resources where possible, killing workgroups only when demanded by other tasks, and forking additional workgroups when possible.
+
+\myfiglong{\ref{fig:overview}} illustrates sharing of GPU compute units between a cooperative kernel and a
+graphics task. Workgroups 2 and 3 of the cooperative kernel
+are killed at an $\offerkill$ to make room for a graphics
+task. The workgroups are subsequently restored to the cooperative kernel when workgroup 0 calls $\offerfork$. The order in
 which workgroups are killed is important and discussed in
 \mysec{\ref{sec:semantics}}. The \emph{gather time} is the time
-between when resources are requested and when the application has
-surrendered the resources via $\offerkill$. To satisfy soft-real time
+between resources being requested and the application surrendering them via $\offerkill$. To satisfy soft-real time
 constraints, this time should be low; our experimental study
@@
 
-The cooperative kernels programming model has several appealing
+The cooperative kernels model has several appealing
 properties:
@@
 \item The model has no impact on the development of regular
-  (non-cooperative) compute and graphics kernels: these can be programmed exactly as they
-  are now.
+  (non-cooperative) compute and graphics kernels.
 %, as can graphics shading routines written using APIs such as
@@
 \item Cooperative kernels can be implemented over the occupancy\-/bound
-  execution model that current GPUs provide: our prototype implementation requires
-  no additional hardware or driver support.
-
-\item If hardware support for preemption is available it can be leveraged to implement cooperative kernels efficiently, and cooperative kenrnels can avoid the overhead of unnecessary preemptions by allowing the programmer to communicate ``smart'' preemption points.
-
-\item Our experiments with a range of applications \cutthree{using three GPUs} show that the model can enable efficient multitasking of cooperative and non-cooperative
-  tasks.
+  execution model provided by current GPUs: our prototype implementation uses no special hardware/driver support.
+
+\item If hardware support for preemption \emph{is} available, it can be leveraged to implement cooperative kernels efficiently, and cooperative kernels can avoid unnecessary preemptions by allowing the programmer to communicate ``smart'' preemption points.
 
@@
 
-Placing the new primitives requires manual effort, but our experience porting a representative set of
-GPU-accelerated irregular algorithms to use cooperative kernels
-(\mysec\ref{sec:portingalgorithms}) suggests that this is straightforward in practice.
+Placing the new primitives requires manual effort, but is straightforward for the representative set of
+GPU-accelerated irregular algorithms that we have ported to use cooperative kernels.  Our experiments show that the model can enable efficient multitasking of cooperative and non-cooperative tasks.
 
@@
   kernels on top OpenCL 2.0
-  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}).
+  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}), including a best-effort comparison with the efficiency afforded by hardware-supported preemption available on Nvidia GPUs.
 

### 2017-02-26 write-ups/FSE/paper.tex
Space saving.

@@
 
-The threads of a kernel are divided into \emph{workgroups}, each consisting of a fixed number of threads.  Functions
+The threads of a kernel are divided into \emph{workgroups}.
+%, each consisting of a fixed number of threads.
+Functions
 $\getlocalid$ and $\getgroupid$ return a thread's local id within
-its workgroup and the id of the workgroup.
+its workgroup and the workgroup id.
 %
@@
 
-Communication within a workgroup can be achieved
-using a workgroup barrier.  Finer-grained
+%Communication within a workgroup can be achieved
+%using a workgroup barrier.
+Fine-grained
 communication within a workgroup, as well as inter-workgroup
@@
 
-\begin{figure}
+\begin{figure}[t]
 
@@
 
-\begin{figure}
+\myparagraph{Work stealing}
+%
+Work stealing enables dynamic balancing of tasks across
+processing units. It is useful when the number of tasks to be
+processed is dynamic, due to one task creating an arbitrary number of
+new tasks.  Work stealing has been explored in the context of
+GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}. Each workgroup has a
+queue from which it obtains tasks to process, and to which
+it stores new tasks. If its queue is empty, a workgroup
+tries to \emph{steal} a task from another queue.
+
+\myfiglong\ref{fig:workstealing} illustrates a work
+stealing kernel. Each thread receives a pointer to the
+task queues, in global memory, initialized by the
+host to contain the initial tasks. A thread uses its workgroup id 
+(line~\ref{line:wksteal:getgroupid}) as a queue id to access the relevant task queue. The
+$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
+pops a task from the workgroup's queue or, if the queue is empty,
+tries to steal a task from other queues.
+%
+% is called by all threads of a workgroup.  Inside
+% $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
+% the \emph{master thread}, tries to pop a
+% task from the workgroup's queue, or to steal a task from a different
+% queue. The master thread uses local memory inside
+% $\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
+% all threads of its workgroup.
+%
+If a task was obtained, then the workgroup processes it
+(line~\ref{line:wksteal:processtask}), which may lead to new tasks
+being created and pushed to the workgroup's queue.
+
+Although not depicted here, concurrent accesses to queues inside
+$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
+mutex per queue, implemented using atomic compare and swap operations
+on global memory.  The kernel presents two opportunities
+for spin-waiting: spinning to obtain a mutex, and spinning in the main
+kernel loop to obtain a task.  The algorithm thus depends
+on a fair scheduler, which is not guaranteed by current GPU programming models.
+
+\begin{figure}[t]
 
@@
 
-\myparagraph{Work stealing example}
-%
-Work stealing enables dynamic balancing of tasks across several
-processing units. It is useful when the number of tasks to be
-processed is dynamic, due to one task creating an arbitrary number of
-new tasks.  Work stealing has been explored in the context of
-GPUs~\cite{DBLP:conf/egh/CedermanT08,TPO10}. Each workgroup has an
-associated queue from which it obtains tasks to process, and to which
-it stores new tasks. If a workgroup's queue is empty, the workgroup
-tries to \emph{steal} a task from another queue.
-
-\myfiglong\ref{fig:workstealing} shows a simplified version of a work
-stealing kernel. Each thread receives a pointer to the
-task queues, in global memory, initialized by the
-host to contain the initial tasks. Each thread identifies
-its workgroup via $\getgroupid$
-(line~\ref{line:wksteal:getgroupid}), which is used as a queue id to access the relevant task queue. The
-$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
-pops a task from the workgroup's queue or, if the queue is empty,
-tries to steal a task from other queues.
-
-% is called by all threads of a workgroup.  Inside
-% $\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
-% the \emph{master thread}, tries to pop a
-% task from the workgroup's queue, or to steal a task from a different
-% queue. The master thread uses local memory inside
-% $\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
-% all threads of its workgroup.
-
-If a task was obtained, then the workgroup processes it
-(line~\ref{line:wksteal:processtask}): during this operation, new tasks
-may be created and pushed into the workgroup's queue.
-
-Although not depicted here, concurrent accesses to queues inside
-$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
-mutex per queue, implemented using atomic compare and swap operations
-on global memory.  The kernel presents two opportunities
-for spin-waiting: spinning to obtain a mutex, and spinning in the main
-kernel loop to obtain a task.  The algorithm thus depends
-on a fair scheduler, which is not guaranteed by current GPU programming models.
-
-\myparagraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
-skeleton of a frontier-based graph traversal algorithm; such algorithms have
+\myparagraph{Graph traversal} \myfiglong\ref{fig:graphsearch} illustrates a frontier-based graph traversal algorithm; such algorithms have
 been shown to execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}.
@@
 arrays of graph nodes. Initially, $\keyword{n0}$ contains the
-starting nodes to process. Private variable $\keyword{level}$ indicates the frontier level currently being
-processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ point to
-distinct node arrays recording the nodes to be processed during the current and next frontier, respectively.
+starting nodes to process. Private variable $\keyword{level}$ records the current frontier level, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ point to
+distinct arrays recording the nodes to be processed during the current and next frontier, respectively.
 
@@
 In this case, the stride is the total number of threads, obtained via
-$\getglobalsize$.  Each thread processes nodes with the following sequence of indices:
+$\getglobalsize$.  Each thread processes nodes with indices
 \code{$\keyword{tid}$+$\keyword{stride}$*0},
@@
 \code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
-A thread processes a node via $\keyword{process\_node}$, which may (a) push nodes to process in the next frontier to
-$\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
-the computation. After processing the frontier, the threads swap their
+A thread calls $\keyword{process\_node}$ to process a node given the current level, with nodes to be processed next frontier being pushed to $\keyword{out\_nodes}$.
+After processing the frontier, the threads swap their
 node array pointers (line~\ref{line:graph:swap}).

### 2017-02-26 write-ups/FSE/paper.tex
A little negative spacing.
Remove mention of supplementary material for semantics and other design choices; this isn't so important.
More space.

@@
 \end{lstlisting}
-%\vspace{-2mm}
+\vspace{-2mm}
 \caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
@@
 \end{lstlisting}
-%\vspace{-2mm}
+\vspace{-2mm}
 \caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
@@
 %
-In Supplementary material we present a formal semantics for
-cooperative kernels atop an abstract GPU programming model, and
-discuss several cases where we might have taken different and also
-reasonable semantic decisions.
 
@@
 optionally be annotated with a new $\transmit$ qualifier.  Annotating
-a variable $v$ with $\transmit$ means that when a workgroup spawns new
-workgroups by calling $\offerfork$, the workgroup should transmit its
-current value for $v$ to the threads of the new workgroups, to serve
-as an initial value for $v$.  We detail the semantics for this when we
+a variable $v$ with $\transmit$ means that when a workgroup 
+uses $\offerfork$ to spawn new
+workgroups, the workgroup should transmit its
+current value for $v$ to the threads of the new workgroups.
+We detail the semantics for this when we
 describe $\offerfork$ below.
@@
 active workgroups.  This is in contrast to $\getnumgroups$ for
-regular kernels, which returns the fixed number of workgroups that
-were requested at kernel launch time (see \mysec\ref{sec:opencl}).
+regular kernels, which returns the fixed number of workgroups that execute the kernel (see \mysec\ref{sec:opencl}).
 
@@
 Suppose a workgroup with id $m$ executes $\offerkill$.  If the
-workgroup has the highest largest id of all active workgroups then it
+workgroup has the largest id among active workgroups then it
 can be killed by the scheduler, except that workgroup 0 can never be
-killed (to ensure that the kernel computation does not terminate
-early).  More formally, if $m < M-1$ or $M=1$ then $\offerkill$ is a
+killed (to avoid early termination of the kernel).  More formally, if $m < M-1$ or $M=1$ then $\offerkill$ is a
 no-op.  If instead $M > 1$ and $m = M-1$, the scheduler can choose to
@@
 To launch a kernel, the host sends a data packet to the GPU scheduler
-consisting of: a kernel to execute, the kernel inputs, and a flag
-indicating whether the kernel is cooperative. In our implementation,
+consisting of a kernel to execute, kernel inputs, and a flag
+indicating whether the kernel is cooperative. In our prototype,
 this host-device communication channel is built using fine-grained SVM
@@
 kernels lead to unkillable defunct
-processes~\cite{amd-defunct-process}. Because of this behaviour, we
-could not experiment on AMD platforms.
-
+processes~\cite{amd-defunct-process}.
+%
 %and simultaneously obtained alarming results on this platform when
@@
 %which we plan to discuss with engineers at AMD.
-
+%
 We thus ran our experimental campaign on the Intel GPUs. The results
@@
 issued as frequently.
-
+%
 These results suggest that a hybrid preemption scheme may work the

### 2017-02-27 write-ups/FSE/paper.tex
added new results and descriptions around them

@@
 \includegraphics[width=.67\columnwidth]{iris_octree_NA.pdf}
-\includegraphics[width=.67\columnwidth]{hd520_mis_ecology.pdf}
-\includegraphics[width=.67\columnwidth]{hd5500_bfs_rmat22.pdf}
+\includegraphics[width=.67\columnwidth]{iris_bfs_usa.pdf}
+\includegraphics[width=.67\columnwidth]{iris_color_G3_circuit.pdf}
 %\vspace{-2mm}
-\caption{Example gather time and non-cooperative timing results \TSComment{need to get updated numbers}}\label{fig:fine-grained-timing}
+\caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
 \vspace{-3mm}
@@
 \hline
-$1.09$ & $1.23^{\ddagger}$ & $1.09$ & $1.21^{\diamond}$ & $1.11$ & $1.23^{\ddagger}$ \\
+$1.07$ & $1.23^{\ddagger}$ & $1.06$ & $1.20^{\diamond}$ & $1.12$ & $1.23^{\ddagger}$ \\
 \end{tabular}\\
 {\small
- $^{\ddagger}$octree, $^{\diamond}$sssp usa\_ny
+ $^{\ddagger}$octree, $^{\diamond}$color G3\_circuit
 }
 %\vspace{-2mm}
-\caption{Cooperative kernel slowdown w/o multitasking \TSComment{need to get updated numbers}}
+\caption{Cooperative kernel slowdown w/o multitasking}
 \label{tab:overhead}
@@
 we indicate which application and input was responsible. The slowdown is
-below 1.5 even in the worst case, and closer to 1 on average. We consider
+below 1.25 even in the worst case, and closer to 1 on average. We consider
 these results encouraging, especially since the performance of our
@@
 
-Our experiments span 13 pairs of cooperative kernels and inputs, 3
+Our experiments span 11 pairs of cooperative kernels and inputs, 3
 classes of non-cooperative kernel workloads, 4 quantities of
 workgroups claimed for the non-cooperative kernel and 2 variations of
-resizing barriers for graph algorithms, leading to 288 configurations.
+resizing barriers for graph algorithms, leading to 240 configurations.
 We run each configuration 10 times, in order to report averaged
@@
 %
-the gather time is higher on the right figure: the rmat22 input graph
-is rather wide than deep, so the graph algorithm reaches resizing
-barriers less often than for the ecology input of the middle figure
-for instance. The scheduler thus has fewer opportunities to collect
-workgroups and gather time increases. Nonetheless, scheduling
+the gather time is higher on the right figure: the G3 circuit input
+graph is rather wide than deep, so the graph algorithm reaches
+resizing barriers less often than for the usa road input of the middle
+figure for instance. The scheduler thus has fewer opportunities to
+collect workgroups and gather time increases. Nonetheless, scheduling
 responsiveness can benefit from the query barrier: when used, this
@@
 %\vspace{-2mm}
-\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute \TSComment{need to get updated numbers}}\label{fig:performance}
+\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
 \vspace{-5mm}
@@
 with running the cooperative kernel in the presence of multitasking,
-vs.\ running the cooperative kernel in isolation (geometric mean over
-all applications and inputs).  We also show the period at which
-non-cooperative kernels can be scheduled (arithmetic mean over all
-applications and inputs).
+vs.\ running the cooperative kernel in isolation (median over all
+applications and inputs).  We also show the period at which
+non-cooperative kernels can be scheduled (median over all applications
+and inputs). We use the median as our data included some drastic
+outliers; these outliers appear to happen with applications/inputs in
+which the resizing barrier are not called very frequently and the
+graphics task asks for half or more resources. For example, a medium
+graphics workload for bfs on the rmat input has over an 8$\times$
+overhead, while the same application on the usa input has an overhead
+of 1.11$\times$. As \myfiglong\ref{fig:performance} shows, most of our
+applications/inputs are much better behaved than this. Future work
+includes examining these problem applications/inputs in more detail,
+possibly to insert more calls to $\offerfork$ and $\offerkill$.
+
 We show results for the three workloads listed in
@@
 workgroups are not sufficient to meet the goal for the heavy workload
-(giving a mean period of 88 ms). If half of the workgroups are
-allocated to the non-cooperative kernel, the heavy workload averages
-$\sim$10\% over its goal period (mean of 44 ms). If all-but-one are
-allocated, the heavy workload reaches its goal period.
+(giving a median period of 104 ms). If half of the workgroups are
+allocated to the non-cooperative kernel, the heavy workload achieves
+its goal period (median of 40 ms). 
 %
@@
 
-Still, heavy workloads meet their period by allocating all-but-one
+Still, heavy workloads meet their period by allocating half
 non-cooperative workgroups, incurring a slow down of less than
-1.4$\times$ on average. Light and medium workloads meet their period
-with only a negligible overhead of the cooperative kernel (less than a
-1.03$\times$ slowdown on average).
+1.5$\times$ (median). Light and medium workloads meet their period
+with only a small overhead of the cooperative kernel; 
+1.04$\times$ and 1.08$\times$ median slowdown respectively.
 
@@
 \centering
-\begin{tabular}{ l r r}
-g. workload &  kernel-level  & cooperative (resources) \\
+\begin{tabular}{ l r r r}
+g. workload &  kernel-level  & cooperative & resources \\
 \hline
-light & 1.04 & 1.XX($N/4$)\\
-medium & 1.08 & 1.XX($N/4$)\\
-heavy & 1.33 & 1.XX($N-1$)\\
+light & 1.04 & 1.04 & $N/4$\\
+medium & 1.08 & 1.08 & $N/4$\\
+heavy & 1.33 & 1.47 & $N/2$\\
 \end{tabular} \\
@@
 \caption{Overhead of kernel level preemption vs cooperative kernels for three
-graphics workloads \TSComment{need to get updated numbers}}
+graphics workloads}
 \label{tab:preemption}
@@
 assume the cost of preemption is negligible (e.g.\ Nvidia have
-reported preemption times of 0.1 ms for Pascal~\cite{TODO}), then the overhead
-associated with switching between compute and graphics every $P$ time
-steps is $P/(P-D)$.
-
-We compare this task-level preemption overhead
-model with our experimental results per graphics workload in
-\mytab{\ref{tab:preemption}}.
-%
-Based on the above assumptions, our approach provides lower overhead for low and medium
-graphics workloads. Recall for the heavy graphics workload, we are
-within $10\%$ of the required deadline ($40$ ms) by using half the
-workgroups with an overhead of $1.XX$, which is less than the
-kernel-level preemption overhead of $1.33$. \TSComment{need to get
-  updated numbers} \ADComment{I don't quite understand the wording; sounds a bit like we still win with the heavy workload.}
-
-The low overhead of cooperative kernels for low and medium graphics
-workloads is due to sub-linear scalability of the irregular algorithms that they implement: some resources can
-be taken away during computation without proportionally reducing
-execution time.  Our poor performance for the heavy workload is
-because the graphics task requires all but one workgroup, which
-cripples the cooperative kernel enough that $\offerfork$ calls are not
-issued as frequently.
+reported preemption times of 0.1 ms for Pascal~\cite{anandtech}), then
+the overhead associated with switching between compute and graphics
+every $P$ time steps is $P/(P-D)$.
+
+We compare this task-level preemption overhead model with our
+experimental results per graphics workload in
+\mytab{\ref{tab:preemption}}. We report the overhead of the
+configuration that allowed us to meet to deadline of the graphics
+task.
+%
+Based on the above assumptions, our approach provides similar overhead
+for low and medium graphics workloads; however, for the high workload,
+our approach has a higher overhead.
+
+Our poor performance for the heavy workload is because the graphics
+task requires half of the workgroups, which cripples the cooperative
+kernel enough that $\offerfork$ calls are not issued as frequently.
+Future work may examine how to insert more resizing calls in these
+applications to decrease this overhead.
 %
 These results suggest that a hybrid preemption scheme may work the
-best for multi-tasking and graphics. That is, the cooperative approach
-works well for light and medium tasks, but heavy graphics tasks
+well for multi-tasking and graphics. That is, the cooperative approach
+works well for light and medium tasks and requires no special hardware
+or preemption state saving; on the other hand, heavy graphics tasks
 benefit from the coarser grained, kernel-level preemption strategy.

### 2017-02-27 write-ups/FSE/paper.tex
got everything but the conclusion to fit

@@
 \label{fig:overview}
+\vspace{-2mm}
 \end{figure}
@@
 are killed at an $\offerkill$ to make room for a graphics
-task. The workgroups are subsequently restored to the cooperative kernel when workgroup 0 calls $\offerfork$. The order in
-which workgroups are killed is important and discussed in
-\mysec{\ref{sec:semantics}}. The \emph{gather time} is the time
+task. The workgroups are subsequently restored to the cooperative kernel when workgroup 0 calls $\offerfork$. The \emph{gather time} is the time
 between resources being requested and the application surrendering them via $\offerkill$. To satisfy soft-real time
@@
 (*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
-    if (t) {
+    if (t)
 (*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
-    }
   }
@@
 (*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
-    for (int i = tid; i < nodes.size; i += stride) {
+    for (int i = tid; i < nodes.size; i += stride)
       process_node(g, in_nodes[i], out_nodes, level);
-    }
 (*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
@@
 (*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
-    for (int i = tid; i < nodes.size; i += stride) {
+    for (int i = tid; i < nodes.size; i += stride)
       process_node(g, in_nodes[i], out_nodes, level);
-    }
     swap(&in_nodes, &out_nodes);
@@
 would be free to provide $K$ with fewer than $N$ active workgroups
-initially.)
+initially).
 
@@
 \label{tab:applications}
-\vspace{-8mm}
+\vspace{-10mm}
 \end{table}
@@
 %
-We thus ran our experimental campaign on the Intel GPUs. The results
+We thus ran our experiments on the Intel GPUs. The results
 were similar across the GPUs, so for conciseness, we report only on
 the Iris 6100 GPU (driver 20.19.15.4463) with a host CPU i3-5157U. The
-Iris has a reported 47 compute units. Results for the two other Intel
-GPUs are available in supplementary material.
+Iris has a reported 47 compute units. Results for the other Intel GPUs
+are available in supplementary material.
 
@@
 which the resizing barrier are not called very frequently and the
-graphics task asks for half or more resources. For example, a medium
-graphics workload for bfs on the rmat input has over an 8$\times$
-overhead, while the same application on the usa input has an overhead
-of 1.11$\times$. As \myfiglong\ref{fig:performance} shows, most of our
+graphics task asks for half or more of the resources. For example, a
+medium graphics workload for bfs on the rmat input has over an
+8$\times$ overhead when asking for all but one of the workgroups for
+graphics. As \myfiglong\ref{fig:performance} shows, most of our
 applications/inputs are much better behaved than this. Future work
@@
 
-Let us assume that a graphics workload is required to be scheduled
-with period $P$ and duration $D$, and that a long-running compute
-kernel requires time $C$ to execute without interruption.  If we
-assume the cost of preemption is negligible (e.g.\ Nvidia have
-reported preemption times of 0.1 ms for Pascal~\cite{anandtech}), then
-the overhead associated with switching between compute and graphics
-every $P$ time steps is $P/(P-D)$.
+Given a graphics workload that is required to be scheduled with period
+$P$ and duration $D$, and that a compute kernel requires time $C$ to
+execute without interruption.  If we assume the cost of preemption is
+negligible (e.g.\ Nvidia have reported preemption times of 0.1 ms for
+Pascal~\cite{anandtech}), then the overhead associated with switching
+between compute and graphics every $P$ time steps is $P/(P-D)$.
 
@@
 Based on the above assumptions, our approach provides similar overhead
-for low and medium graphics workloads; however, for the high workload,
-our approach has a higher overhead.
+for low and medium graphics workloads, however, has a higher overhead for
+the high workload.
 
@@
 Very recent \nvidia GPUs (i.e. the Pascal architecture) support
-preemption, though, as discussted in \mysec{sec:intro} and
-\mysec{sec:nvidiacomparison}, it is not clear whether they guarantee
+preemption, though, as discussted in \mysec{\ref{sec:intro}} and
+\mysec{\ref{sec:nvidiacomparison}}, it is not clear whether they guarantee
 fairness or allow tasks to share GPU resources at the workgroup

### 2017-02-27 write-ups/FSE/paper.tex
got the paper to fitmake

@@
 multi-tasked with non-cooperative workloads
-(\mysec\ref{sec:responsiveness}).
-
-To our knowledge, there are no competing methods against which to
-meaningfully compare our approach.  We are not aware of any existing
-platform that supports multitasking between long-running, fully-occupant
-OpenCL kernels.  Very recent \nvidia{} GPUs, to which we do not yet have
-access, support preemption for graphics
-applications~\cite{PascalWhitepaper}, but our understanding is that this
-cannot yet be leveraged from CUDA, and certainly not from OpenCL.
+(\mysec\ref{sec:responsiveness}). Finally, we compare
+against a performance model of \emph{kernel-level} preemption, which
+we understand to be what current Nvidia GPUs provide (\mysec\ref{sec:nvidiacomparison}).
+
+%To our knowledge, there are no competing methods against which to
+%meaningfully compare our approach.  We are not aware of any existing
+%platform that supports multitasking between long-running, fully-occupant
+%OpenCL kernels.  Very recent \nvidia{} GPUs, to which we do not yet have
+%access, support preemption for graphics
+%applications~\cite{PascalWhitepaper}, but our understanding is that this
+%cannot yet be leveraged from CUDA, and certainly not from OpenCL.
 
@@
 
-We devised the following
-procedure to determine non-cooperative OpenCL workloads that simulate
-the computational intensity of various graphics rendering workloads.
-%
-We designed a synthetic kernel that occupies all workgroups of a
-GPU for a parameterised time period $t$, invoked in an infinite loop by a host application.  We then
-searched for a maximum value for $t$ that allowed the
-synthetic kernel to execute without having an observable impact on
-graphics rendering.  Using the computed value, we
-ran the host application for $X$ seconds, measuring the time $Y < X$
-dedicated to GPU execution during this period and the number
-of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
-experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
-spent using the GPU to render the display between kernel calls and the
-period at which the OS requires the GPU for display rendering,
-respectively.
+We devised the following procedure to determine non-cooperative OpenCL
+workloads that simulate the computational intensity of various
+graphics rendering workloads.
+%
+We designed a synthetic kernel that occupies all workgroups of a GPU
+for a parameterised time period $t$, invoked in an infinite loop by a
+host application.  We then searched for a maximum value for $t$ that
+allowed the synthetic kernel to execute without having an observable
+impact on graphics rendering.  Using the computed value, we ran the
+host application for $X$ seconds, measuring the time $Y < X$ dedicated
+to GPU execution during this period and the number of kernel launches
+$n$ that were issued.  We used $X \geq 10$ in all experiments.  The
+values $(X-Y)/n$ and $X/n$ estimate the average time spent using the
+GPU to render the display between kernel calls (call this $E$) and the
+period at which the OS requires the GPU for display rendering (call
+this $P$), respectively.
 
@@
 
-Our results are the following: light workloads needs to run 3ms every
-70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
-the medium and heavy workloads, the 40ms period corresponds to graphics
-rendering occurring 25 times per second, which coincides with the human
-persistence of vision. The 3ms execution duration of both light and
-medium configurations indicates that, as expected, GPU computation is
-cheaper for basic display rendering compared with rendering that uses
-complex shaders. We use these results to guide our evaluation of the
-responsiveness of cooperative scheduling in
-\mysec\ref{sec:responsiveness}.
+Our results are the following: $P=70ms$ and $E=3ms$ for light;
+$P=40ms$, $E=3ms$ for medium; and $P=40ms$, $E=10ms$ for heavy. For
+medium and heavy, the 40ms period coincides with the human persistence
+of vision. The 3ms execution duration of both light and medium
+configurations indicates that GPU computation is cheaper for basic
+display rendering compared with more complex rendering that uses
+shaders. These results guide our evaluation of cooperative scheduling
+(\mysec\ref{sec:responsiveness}).
 
@@
 during the run, we also record the \emph{gather time} needed by the
-scheduler to collect enough workgroups to launch the non-cooperative
-kernel, and the non-cooperative kernel execution time.
+scheduler to collect workgroups to launch the non-cooperative kernel,
+and the non-cooperative kernel execution time.
 %
@@
 non-cooperative kernels can be scheduled (median over all applications
-and inputs). We use the median as our data included some drastic
-outliers; these outliers appear to happen with applications/inputs in
-which the resizing barrier are not called very frequently and the
-graphics task asks for half or more of the resources. For example, a
-medium graphics workload for bfs on the rmat input has over an
-8$\times$ overhead when asking for all but one of the workgroups for
-graphics. As \myfiglong\ref{fig:performance} shows, most of our
-applications/inputs are much better behaved than this. Future work
-includes examining these problem applications/inputs in more detail,
-possibly to insert more calls to $\offerfork$ and $\offerkill$.
+and inputs). We use the median as our data included some outliers;
+these outliers happen with applications/inputs in which the resizing
+barrier are not called very frequently and the graphics task requires
+half or more workgroups. For example, a medium graphics workload for
+bfs on the rmat input has over an 8$\times$ overhead when asking for
+all but one of the workgroups. As \myfiglong\ref{fig:performance}
+shows, most of our applications/inputs are much better behaved than
+this. Future work will examine these problem applications/inputs in
+more detail, possibly inserting more calls to $\offerfork$ and
+$\offerkill$.
 
@@
 the execution time of the cooperative kernel.
-
+%
 Still, heavy workloads meet their period by allocating half
@@
 
-\subsection{Comparison with Task Level Preemption}\label{sec:nvidiacomparison}
+\subsection{Comparison with Kernel-Level Preemption}\label{sec:nvidiacomparison}
 
@@
 hardware support for instruction-level preemption.  Few details of how
-this works are available, but from reading Nvidia documentation
-and online material~\cite{PascalWhitepaper,anandtech}, and running
-litmus tests on a Pascal platform, our understanding is that
-preemption of entire kernels, but not of individual workgroups running
-a kernel is supported.  Intel GPUs do
-not provide this feature, and out OpenCL 2.0 prototype of cooperative kernels cannot run on Nvidia GPUs, making a direct comparison impossible.
-We present here a theoretical analysis of the overheads associated with
-sharing the GPU between graphics and compute tasks via kernel-level
-preemption.
+this works are available, but from reading Nvidia documentation and
+online material~\cite{PascalWhitepaper,anandtech}, our understanding
+is that preemption of entire kernels, but not of individual workgroups
+is supported.  Intel GPUs do not provide this feature, and out OpenCL
+2.0 prototype of cooperative kernels cannot run on Nvidia GPUs, making
+a direct comparison impossible.  We present here a theoretical
+analysis of the overheads associated with sharing the GPU between
+graphics and compute tasks via kernel-level preemption.
 
@@
 fairly scheduled and to share GPU resources with other workloads.
-Experimental results using our megakernel-based prototype \cutthree{over three Intel GPUs}
-shows that the model is a good fit for the needs of current
-GPU-accelerated irregular algorithms.  The
-performance that could be gained through a native implementation with
-dedicated driver support would be even better.
-%
-Avenues for future work include seeking
-additional classes of irregular algorithms to which the model might
-(be extended to) apply (to), investigating implementing native support
-in open source drivers, and integrating
-cooperative kernels into template- and compiler-based programming
-models for graph algorithms on
+Experimental results using our megakernel-based prototype shows that
+the model is a good fit for current GPU-accelerated irregular
+algorithms.  The performance that could be gained through a native
+implementation with driver support would be even better.
+%
+Avenues for future work include seeking additional classes of
+irregular algorithms to which the model might (be extended to) apply
+(to), investigating implementing native support in open source
+drivers, and integrating cooperative kernels into template- and
+compiler-based programming models for graph algorithms on
 GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.

### 2017-02-27 write-ups/FSE/paper.tex
Fixed some spelling issues.

@@
 support instruction-level preemption~\cite{PascalWhitepaper}. Despite
-this, Nvidia's most recent compute capablity does not guarantee
+this, Nvidia's most recent compute capability does not guarantee
 forward progress for CUDA programmers (evidence that vendors will not
@@
 The semantics of $\resizingglobalbarrier$ can be modelled via calling $\offerfork$ and $\offerkill$
-surrounded, and seperated by calls to a $\globalbarrier$.
+surrounded, and separated by calls to a $\globalbarrier$.
 %
@@
 graph is rather wide than deep, so the graph algorithm reaches
-resizing barriers less often than for the usa road input of the middle
+resizing barriers less often than for the USA road input of the middle
 figure for instance. The scheduler thus has fewer opportunities to
@@
 Very recent \nvidia GPUs (i.e. the Pascal architecture) support
-preemption, though, as discussted in \mysec{\ref{sec:intro}} and
+preemption, though, as discussed in \mysec{\ref{sec:intro}} and
 \mysec{\ref{sec:nvidiacomparison}}, it is not clear whether they guarantee

### 2017-02-27 write-ups/FSE/paper.tex
Fixed some typos.

@@
 is that preemption of entire kernels, but not of individual workgroups
-is supported.  Intel GPUs do not provide this feature, and out OpenCL
+is supported.  Intel GPUs do not provide this feature, and our OpenCL
 2.0 prototype of cooperative kernels cannot run on Nvidia GPUs, making
@@
 
-Given a graphics workload that is required to be scheduled with period
+Suppose a graphics workload is required to be scheduled with period
 $P$ and duration $D$, and that a compute kernel requires time $C$ to

### 2017-06-22 write-ups/FSE/paper.tex
Getting ready for camera-ready.

@@
 
-\documentclass[parskip=half,sigconf,review, anonymous=true, acmcopyrightmode=none]{acmart}
-\acmConference[ESEC/FSE 2017]{11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{4--8 September, 2017}{Paderborn, Germany}
+\documentclass[sigconf]{acmart}
+
+%%% The following is specific to ESEC/FSE'17 and the paper
+%%% 'Cooperative Kernels: GPU Multitasking for Blocking Algorithms'
+%%% by Tyler Sorensen, Hugues Evrard, and Alastair F. Donaldson.
+%%%
+%%% If you see 'none' in the 'setcopyright' statement below,
+%%% please first submit your publishing-rights agreement with ACM (follow link on submission page),
+%%% and then visit the instructions page again (follow link in author-kit message)
+%%% in order to copy the correct 'setcopyright' statement.  Please contact us in case of questions.
+%%%
+\setcopyright{none}
+\acmDOI{10.1145/3106237.3106265}
+\acmISBN{978-1-4503-5105-8/17/09}
+\acmConference[ESEC/FSE'17]{2017 11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{September 4-8, 2017}{Paderborn, Germany}
+\acmYear{2017}
+\copyrightyear{2017}
+\acmPrice{15.00}
 
@@
   morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then,transmit},
-  escapeinside={(*@}{@*)},
+  escapeinside={//(*@}{@*)},
   numbers=left

### 2017-06-29 write-ups/FSE/paper.tex
Started working on camera-ready.

@@
 
+\ADComment{Reviewer comment: Given that the authors propose that the OpenCL (and other frameworks) be modified to incorporate their non-cooperative kernel implementation, they should explain why GPU vendors would prefer to adopt the authors' approach over a more functional preemptive multitasking one.
+Are particular properties of GPU hardware or software stack (e.g.
+size of context) hindering the use of preemptive multitasking?
+How is the effect of these properties minimized under the authors' proposed approach?}
+
+\ADComment{Get something in on NVIDIA's recent developments.}
+
+\ADComment{Possibly cite: Marc Tarpenning. Cooperative Multitasking in C++. Dr. Dobb's Journal, April 01, 1991}
+
+\ADComment{Please outline the rationale behind your choice of the particular primitives you propose.}
+
 \section{Introduction}\label{sec:intro}
@@
 \subsection{Motivating Examples}\label{sec:openclexamples}
+
+\ADComment{Reviewer comment: Section 2.2 would benefit from a short introduction explaining how the proposed examples fail in current implementations.}
 
@@
 \mysec\ref{sec:intro}, this model is not endorsed
-language standards or vendor implementations, and is unlikely to be respected in the future.
+by language standards or vendor implementations, and is unlikely to be respected in the future.
 %

### 2017-07-04 write-ups/FSE/paper.tex
some work on the paper

--- 2017-06-22 write-ups/FSE/paper.tex
@@
 %\newcommand{\TSComment}[1]{\textcolor{blue}{TS: #1}}
+\newcommand{\TSAdded}[1]{\textcolor{blue}{#1}}
+\newcommand{\TSEmph}[1]{\textcolor{red}{#1}}
 \newcommand{\HEComment}[1]{\textcolor{orange}{HE: #1}}
@@
 
+\newcommand{\transmit}{\mathsf{transmit}}
+
+\newcommand{\code}[1]{\lstset{basicstyle=\tt}\lstinline!#1!\lstset{basicstyle=\scriptsize\tt}}
+
 % lstlisting macros
@@
   morekeywords={barrier,kernel,global,__device__,__syncthreads,__global__,bool,then,transmit},
-  escapeinside={//(*@}{@*)},
+  escapeinside={(*@}{@*)},
   numbers=left
 }
-
-\newcommand{\transmit}{\mathsf{transmit}}
-
-\newcommand{\code}[1]{\lstset{basicstyle=\tt}\lstinline!#1!\lstset{basicstyle=\scriptsize\tt}}
 
@@
 \title{Cooperative Kernels: GPU Multitasking for Blocking Algorithms}
+
+\author{Tyler Sorensen}
+\affiliation{%
+  \institution{Imperial College London}
+  \city{London}
+  \country{UK}}
+\email{t.sorensen15@imperial.ac.uk}
+
+\author{Hugues Evrard}
+\affiliation{%
+  \institution{Imperial College London}
+  \city{London}
+  \country{UK}}
+\email{h.evrard@imperial.ac.uk}
+
+\author{Alastair F. Donaldson}
+\affiliation{%
+  \institution{Imperial College London}
+  \city{London}
+  \country{UK}}
+\email{alastair.donaldson@imperial.ac.uk}
 
@@
 
-%\authorinfo{Tyler Sorensen}
-%{\makebox{Imperial College London, UK} \\
-%}
-%{t.sorensen15@imperial.ac.uk}
+%\authorinfo{Tyler Sorensen} {\makebox{Imperial College London, UK} } {t.sorensen15@imperial.ac.uk }
 
@@
 
-%\newcommand{\myparagraph}[1]{\paragraph{#1.}}
-\newcommand{\myparagraph}[1]{\vspace{1mm}\noindent\emph{\textbf{#1.}}}
+\newcommand{\myparagraph}[1]{\paragraph{#1.}}
+%\newcommand{\myparagraph}[1]{\vspace{1mm}\noindent\emph{\textbf{#1.}}}
 %{\noindent{\bf\emph{#1}}\hspace{1em}}
@@
 %the workgroups that execute a GPU kernel ahead of time.  However,
-Many interesting data-parallel algorithms are \emph{irregular}: the amount
-of work to be processed is unknown ahead of time and may be determined
-by the computation itself.  There is growing interest in accelerating
-such algorithms on GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
+Many interesting data-parallel algorithms are \emph{irregular}: the
+amount of work to be processed is unknown ahead of time and may be
+determined by the computation itself.  There is growing interest in
+accelerating such algorithms on
+GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
+\TSAdded{Such
+  algorithms include octree partitioning, which has applications in
+  particle simulations; and graph traversals, which have applications
+  in speech decoding.}
 
@@
 models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
-HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding
+HSA~\cite{HSAprogramming11}, specify almost no guarantees regarding
 scheduling of workgroups, and current GPU schedulers are unfair in
@@
 compute unit until it finishes execution.
-This is simple model is easy to implement
+This simple model is easy to implement
 efficiently, and experiments suggest that
@@
 \label{fig:overview}
-\vspace{-2mm}
+%\vspace{-2mm}
 \end{figure}
@@
 implement, for the above reasons, yet (2) will not guarantee fair
-scheduling due to the runtime cost of preempting workgroups in a fair
-manner.  Vendors instead wish to retain the essence of the simple
-occupancy-bound model, supporting preemption only in key special
-cases.
+scheduling \TSAdded{using preemption. This is due to the high runtime
+  cost of preempting workgroups on GPUs, which requires managing
+  thread local state (e.g. registers, program location) for all
+  threads in a workgroup (up to 1024 on Nvidia GPUs), as well as
+  \emph{shared memory}, the workgroup local cache (up to 64 KB on
+  Nvidia GPUs)}.  Vendors instead wish to retain the essence of the
+  simple occupancy-bound model, supporting preemption only in key
+  special cases.
 
@@
 Where the cooperative kernel could use additional resources, a workgroup can execute
-$\offerfork$ to indicates that the
+$\offerfork$ to indicate that the
 kernel is prepared to proceed with the existing set of workgroups, but
@@
 %units to be powered down quickly when required.
-In return, the scheduler should allow the cooperative kernel to utilise
-hardware resources where possible, killing workgroups only when demanded by other tasks, and forking additional workgroups when possible.
+In return, the scheduler should allow the cooperative kernel to
+utilise hardware resources where possible, killing workgroups only
+when demanded by other tasks, and forking additional workgroups when
+possible.
+
+\TSAdded{Cooperative kernels allow for \emph{cooperative multitasking}
+  (see \mysec{\ref{sec:relatedwork}}), a multitasking approach used
+  historically when preemption was not available or, in the case of
+  GPUs, too costly (see previous paragraph). Our approach avoids the
+  cost of preemptive multitasking as the state of a workgroup killed
+  via $\offerkill$ does not have to be saved. Previous cooperative
+  multitasking systems often provide \emph{yield} based semantics,
+  where a processing unit would temporarily give up its hardware
+  resource. We intentionally deviate from this design as in the case
+  of a global barrier (a common idiom in the applications we study),
+  adopting yield semantics would force the cooperative kernel to block
+  \emph{completely} in the case where a single workgroup yields,
+  stalling the kernel until the given workgroup resumes.  Instead, our
+  $\offerkill$ allows a kernel to make progress with a smaller number
+  of workgroups, with workgroups potentially joining again later via
+  $\offerfork$.}
 
@@
 \emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{prototype implementation} of cooperative
-  kernels on top OpenCL 2.0
+  kernels on top of OpenCL 2.0
   (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}), including a best-effort comparison with the efficiency afforded by hardware-supported preemption available on Nvidia GPUs.
@@
 
+%\ADComment{Reviewer comment: Section 2.2 would benefit from a short introduction explaining how the proposed examples fail in current implementations.}
+
 \begin{figure}[t]
+
+% lstlisting macros
 
@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
@@
 
-\myfiglong\ref{fig:workstealing} illustrates a work
-stealing kernel. Each thread receives a pointer to the
-task queues, in global memory, initialized by the
-host to contain the initial tasks. A thread uses its workgroup id 
-(line~\ref{line:wksteal:getgroupid}) as a queue id to access the relevant task queue. The
-$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
-pops a task from the workgroup's queue or, if the queue is empty,
-tries to steal a task from other queues.
+\myfiglong\ref{fig:workstealing} illustrates a work stealing
+kernel. Each thread receives a pointer to the task queues, in global
+memory, initialized by the host to contain initial tasks. A thread
+uses its workgroup id (line~\ref{line:wksteal:getgroupid}) as a queue
+id to access the relevant task queue. The $\mathsf{pop\_or\_steal}$
+function (line~\ref{line:wksteal:poporsteal}) pops a task from the
+workgroup's queue or tries to steal a task from other queues. Although
+not depicted here, concurrent accesses to queues inside
+$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
+mutex per queue, implemented using atomic compare and swap operations
+on global memory.
+
+If a task was obtained, then the workgroup processes it
+(line~\ref{line:wksteal:processtask}), which may lead to new tasks
+being created and pushed to the workgroup's queue. The kernel presents
+two opportunities for spin-waiting: spinning to obtain a mutex, and
+spinning in the main kernel loop to obtain a task.
+
 %
@@
 %
-If a task was obtained, then the workgroup processes it
-(line~\ref{line:wksteal:processtask}), which may lead to new tasks
-being created and pushed to the workgroup's queue.
-
-Although not depicted here, concurrent accesses to queues inside
-$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
-mutex per queue, implemented using atomic compare and swap operations
-on global memory.  The kernel presents two opportunities
-for spin-waiting: spinning to obtain a mutex, and spinning in the main
-kernel loop to obtain a task.  The algorithm thus depends
-on a fair scheduler, which is not guaranteed by current GPU programming models.
 
@@
 (*@\label{line:graph:iterate}@*)  while(in_nodes.size > 0) {
-    for (int i = tid; i < nodes.size; i += stride)
+    for (int i = tid; i < in_nodes.size; i += stride)
       process_node(g, in_nodes[i], out_nodes, level);
@@
 \end{lstlisting}
-\vspace{-2mm}
+%\vspace{-2mm}
 \caption{An OpenCL graph traversal algorithm}\label{fig:graphsearch}
@@
 The global barrier used in this application is not provided as a GPU
-primitive, though previous works have shown that such a global
-barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16},
-based on CPU barrier designs~\cite[ch. 17]{HS08}.
- These barriers employ spinning to ensure threads wait at the barrier until all
-threads have arrived, which requires fair
-scheduling between workgroups.% for the barrier to operate correctly.
+primitive, though previous works have shown that such a global barrier
+can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}, based
+on CPU barrier designs~\cite[ch. 17]{HS08}.  These barriers employ
+spinning to ensure threads wait at the barrier until all threads have
+arrived, which requires fair scheduling between workgroups for the
+barrier to operate correctly.
 
@@
 shown, experimentally, to provide.  But, as discussed in
-\mysec\ref{sec:intro}, this model is not endorsed
-language standards or vendor implementations, and is unlikely to be respected in the future.
-%
+\mysec\ref{sec:intro}, this model is not endorsed by language
+standards or vendor implementations, and is unlikely to be respected
+in the future. \TSAdded{Without a fair schedule, the mutexes and
+  barriers, which use spin-waiting in their implementation, are not
+  guaranteed to be satisfied. For example, if a workgroup is not
+  fairly scheduled after obtaining a mutex, there is no guarantee that
+  the mutex will be released. Thus the kernel may hang while other
+  workgroups wait indefinitely to obtain the mutex. }
+%
+
 In \mysec\ref{sec:programmingguidelines} we show how the work stealing
@@
 %\vspace{-1mm}
-\myparagraph{Active workroups}
+\myparagraph{Active workgroups}
 %
@@
 (*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
-    for (int i = tid; i < nodes.size; i += stride)
+    for (int i = tid; i < in_nodes.size; i += stride)
       process_node(g, in_nodes[i], out_nodes, level);
@@
 \caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
-\vspace{-4mm}
+%\vspace{-4mm}
 \end{figure}
@@
 megakernel is launched with as many workgroups as can be occupant
-concurrently.  One workgroup takes the role of scheduler, and the
+concurrently.  One workgroup takes the role of the scheduler, and the
 scheduling logic is embedded as part of the megakernel.  The remaining
@@
 implementation is restricted to the case of two kernels.
+
+\TSAdded{We note that the scheduler requirements given in
+  \mysec{\ref{sec:cooperativekernels}} are agnostic to whether the
+  scheduling logic takes place on the CPU or GPU. To
+  avoid expensive communication between GPU and host, we choose to
+  implement the scheduler on the GPU.}
 
@@
 ever acquired by the scheduling workgroup, \emph{indirectly} provides
-mutual exclusion between a pairs of workgroups executing $\offerkill$
+mutual exclusion between pairs of workgroups executing $\offerkill$
 and $\offerfork$ concurrently.
@@
 \label{tab:applications}
-\vspace{-10mm}
+%\vspace{-10mm}
 \end{table}
@@
 %Linux kernel).
-AMD support SVM atomics only their Kaveri family of
+AMD supports SVM atomics only in their Kaveri family of
 chips and only on Linux.%, requiring to compile the latest driver.
@@
 \caption{Example gather time and non-cooperative timing results}\label{fig:fine-grained-timing}
-\vspace{-3mm}
+%\vspace{-3mm}
 \end{figure*}
@@
 \label{tab:overhead}
-\vspace{-8mm}
+%\vspace{-8mm}
 \end{table}
@@
 \caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative kernels execute}\label{fig:performance}
-\vspace{-5mm}
+%\vspace{-5mm}
 \end{figure}
@@
 \label{tab:preemption}
-\vspace{-8mm}
+%\vspace{-8mm}
 \end{table}
@@
 \mytab{\ref{tab:preemption}}. We report the overhead of the
-configuration that allowed us to meet to deadline of the graphics
+configuration that allowed us to meet the deadline of the graphics
 task.
@@
 %
-These results suggest that a hybrid preemption scheme may work the
-well for multi-tasking and graphics. That is, the cooperative approach
+These results suggest that a hybrid preemption scheme may work well
+for multi-tasking and graphics. That is, the cooperative approach
 works well for light and medium tasks and requires no special hardware
@@
 (e.g. pre 1995 Windows) and is still used by some operating systems,
-such as RISC OS~\cite{risc-os-multitasking}.
+such as RISC OS~\cite{risc-os-multitasking}. \TSAdded{Additionally,
+  cooperative multitasking can be efficiently implemented in today's
+  high-level languages for domains in which preemptive multitasking is
+  either too costly or not supported on legacy
+  systems~\cite{Tarpenning:1991:CMC:136810.136820}}.
 
@@
 fairly scheduled and to share GPU resources with other workloads.
-Experimental results using our megakernel-based prototype shows that
+Experimental results using our megakernel-based prototype show that
 the model is a good fit for current GPU-accelerated irregular

### 2017-07-05 write-ups/FSE/paper.tex
fixed one thing with vspace
hopefully one of the final passes through the paper

@@
 
+\usepackage{flushend}
 \begin{document}
@@
 
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10010940.10010941.10010949.10010957.10010959</concept_id>
+<concept_desc>Software and its engineering~Multiprocessing / multiprogramming / multitasking</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011039.10011311</concept_id>
+<concept_desc>Software and its engineering~Semantics</concept_desc>
+<concept_significance>100</concept_significance>
+</concept>
+<concept>
+<concept_id>10010147.10010371.10010387.10010389</concept_id>
+<concept_desc>Computing methodologies~Graphics processors</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Multiprocessing / multiprogramming / multitasking}
+\ccsdesc[100]{Software and its engineering~Semantics}
+\ccsdesc[300]{Computing methodologies~Graphics processors}
+
+\keywords{GPU, cooperative multitasking, irregular parallelism}
+
 %
@@
 
-\myparagraph{The needs of irregular data-parallel algorithms}
+\myparagraph{The Needs of Irregular Data-parallel Algorithms}
 %Acceleration of general-purpose computations on graphics processing
@@
 amount of work to be processed is unknown ahead of time and may be
-determined by the computation itself.  There is growing interest in
+determined during computation.  There is growing interest in
 accelerating such algorithms on
 GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
-\TSAdded{Such
-  algorithms include octree partitioning, which has applications in
-  particle simulations; and graph traversals, which have applications
-  in speech decoding.}
 
@@
 %\vspace{-1mm}
-\myparagraph{A degree of fairness: occupancy-bound execution} The current GPU programming
+\myparagraph{A Degree of Fairness: Occupancy-bound Execution} The current GPU programming
 models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75} and
@@
 compute unit until it finishes execution.
-This simple model is easy to implement
-efficiently, and experiments suggest that
-it is widely employed by today's
+Experiments suggest that this model
+is widely employed by today's
 GPUs~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.
@@
 %\vspace{-1mm}
-\myparagraph{Resistance to occupancy-bound execution}
+\myparagraph{Resistance to Occupancy-bound Execution}
 Despite its practical prevalence, none of the current GPU programming
@@
 unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
-tasks, such as graphics rendering.
+tasks (e.g.\ graphics rendering).
 %Precisely because
@@
 %being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
-Second, \emph{energy throttling} is
-an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to
-power down some compute units if the battery level is low.
-Implementing energy
-throttling would require being able to suspend the execution of a
-workgroup, powering down the compute unit to which the workgroup was
-bound, something that the occupancy-bound execution model does not allow.
+Second, \emph{energy throttling} is an important concern for
+battery-powered
+devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the
+future, it will be desirable for a mobile GPU driver to power down
+some compute units, and hence suspend the execution of the occupant
+workgroup, if the battery level is low.
+%Implementing energy
+%throttling would require being able to suspend the execution of a
+%workgroup, powering down the compute unit to which the workgroup was
+%bound, something that the occupancy-bound execution model does not allow.
 
@@
 standardisation efforts
-(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
-and various engineers from GPU vendors), is that GPU vendors (1) will
-not commit to the occupancy-bound execution model they currently
-implement, for the above reasons, yet (2) will not guarantee fair
-scheduling \TSAdded{using preemption. This is due to the high runtime
-  cost of preempting workgroups on GPUs, which requires managing
-  thread local state (e.g. registers, program location) for all
-  threads in a workgroup (up to 1024 on Nvidia GPUs), as well as
-  \emph{shared memory}, the workgroup local cache (up to 64 KB on
-  Nvidia GPUs)}.  Vendors instead wish to retain the essence of the
-  simple occupancy-bound model, supporting preemption only in key
-  special cases.
-
-For example, Nvidia's latest architecture (Pascal) is documented to
-support instruction-level preemption~\cite{PascalWhitepaper}. Despite
-this, Nvidia's most recent compute capability does not guarantee
-forward progress for CUDA programmers (evidence that vendors will not
-yet commit to providing such guarantees).  Indeed, on a GTX Titan X
-(Pascal) we still observe starvation: we found that a global barrier
-can be executed successfully with 56 workgroups, but deadlocks with 57
-workgroups.  This is an example of a vendor supporting preemption, yet
-not guaranteeing fair scheduling.
+(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}),
+is that GPU vendors (1) will not commit to the occupancy-bound
+execution model they currently implement, for the above reasons, yet
+(2) will not guarantee fair scheduling \TSAdded{using preemption. This
+  is due to the high runtime cost of preempting workgroups, which
+  requires managing thread local state (e.g. registers, program
+  location) for all workgroup threads (up to 1024 on Nvidia
+  GPUs), as well as \emph{shared memory}, the workgroup local cache
+  (up to 64 KB on Nvidia GPUs)}.  Vendors instead wish to retain the
+essence of the simple occupancy-bound model, supporting preemption
+only in key special cases.
+
+For example, Nvidia's Pascal architecture supports
+preemption~\cite{PascalWhitepaper}. Despite this, the device still
+does not guarantee forward progress. Indeed, on a GTX Titan X (Pascal)
+we still observe starvation: we found that a global barrier can be
+executed successfully with 56 workgroups, but deadlocks with 57
+workgroups.  Thus, although preemption is supported, there is not fair
+scheduling.
 
 %\vspace{-1mm}
-\myparagraph{Our proposal: cooperative kernels}
+\myparagraph{Our Proposal: Cooperative Kernels}
 %
@@
 but for good reasons
-GPU vendors will not commit even to the guarantees of the
+GPU vendors will not commit to the guarantees of the
 occupancy-bound execution model.
@@
   historically when preemption was not available or, in the case of
-  GPUs, too costly (see previous paragraph). Our approach avoids the
-  cost of preemptive multitasking as the state of a workgroup killed
-  via $\offerkill$ does not have to be saved. Previous cooperative
-  multitasking systems often provide \emph{yield} based semantics,
-  where a processing unit would temporarily give up its hardware
-  resource. We intentionally deviate from this design as in the case
-  of a global barrier (a common idiom in the applications we study),
-  adopting yield semantics would force the cooperative kernel to block
-  \emph{completely} in the case where a single workgroup yields,
-  stalling the kernel until the given workgroup resumes.  Instead, our
-  $\offerkill$ allows a kernel to make progress with a smaller number
-  of workgroups, with workgroups potentially joining again later via
-  $\offerfork$.}
+  GPUs, too costly. Our approach avoids the cost of arbitrary
+  preemption as the state of a workgroup killed via $\offerkill$ does
+  not have to be saved. Previous cooperative multitasking systems have
+  provided \emph{yield} semantics, where a processing unit would
+  temporarily give up its hardware resource. We deviate from this
+  design as, in the case of a global barrier, adopting yield would
+  force the cooperative kernel to block \emph{completely} in the case
+  where a single workgroup yields, stalling the kernel until the given
+  workgroup resumes.  Instead, our $\offerkill$ allows a kernel to
+  make progress with a smaller number of workgroups, with workgroups
+  potentially joining again later via $\offerfork$.}
 
@@
 
-\item The model is backwards-compatible: if $\offerkill$ and $\offerfork$ are ignored, a cooperative kernel will behave
+\item The model is backwards-compatible: $\offerkill$ and $\offerfork$ may be ignored, and a cooperative kernel will behave
   exactly as a regular kernel does on current GPUs.
@@
 %kernel is given by $\getglobalsize = \getlocalsize \times \getnumgroups$.
-
+%
 Execution of the threads in a workgroup can be synchronised via a
-workgroup barrier. This barrier is a \emph{workgroup-level} function:
-it can only appear in conditional code if all threads in a workgroup
-that reach the barrier agree on the guards of all enclosing
-conditional statements.  A \emph{global} barrier (synchronising all
+workgroup barrier. 
+%This barrier is a \emph{workgroup-level} function:
+%it can only appear in conditional code if all threads in a workgroup
+%that reach the barrier agree on the guards of all enclosing
+%conditional statements.  
+A \emph{global} barrier (synchronising all
 threads of a kernel) is \emph{not} provided as a primitive.
 
-\myparagraph{Memory spaces and memory model} A kernel has access to
+\myparagraph{Memory Spaces and Memory Model} A kernel has access to
 four memory spaces.  \emph{Shared virtual memory} (SVM) is accessible
@@
 
-\myparagraph{Execution model}
-OpenCL specifically makes no guarantees about fair scheduling between
-workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
-  conforming implementation may choose to serialize the workgroups
+\myparagraph{Execution Model}
+OpenCL~\cite[p.\ 31]{opencl2Spec} and CUDA~\cite{cuda-75} specifically makes no guarantees about fair scheduling between
+workgroups executing the same kernel. 
+%stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
+%  conforming implementation may choose to serialize the workgroups
 %so
@@
 %  parallel.
-\dots There is no safe and portable way to synchronize across
-  the independent execution of workgroups since once in the work-pool,
-  they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.
+%\dots There is no safe and portable way to synchronize across
+%  the independent execution of workgroups since once in the work-pool,
+%  they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.
 %
@@
 
-\myparagraph{Work stealing}
+\myparagraph{Work Stealing}
 %
@@
 two opportunities for spin-waiting: spinning to obtain a mutex, and
-spinning in the main kernel loop to obtain a task.
+spinning in the main kernel loop to obtain a task. \TSAdded{Without fair
+scheduling, threads waiting for the mutex might spin indefinitely,
+causing the application to hang.}
 
@@
 
-\myparagraph{Graph traversal} \myfiglong\ref{fig:graphsearch} illustrates a frontier-based graph traversal algorithm; such algorithms have
+\myparagraph{Graph Traversal} \myfiglong\ref{fig:graphsearch} illustrates a frontier-based graph traversal algorithm; such algorithms have
 been shown to execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}.
@@
 In this case, the stride is the total number of threads, obtained via
-$\getglobalsize$.  Each thread processes nodes with indices
-\code{$\keyword{tid}$+$\keyword{stride}$*0},
-\code{$\keyword{tid}$+$\keyword{stride}$*1},
-\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
+$\getglobalsize$.  
+%Each thread processes nodes with indices
+%\code{$\keyword{tid}$+$\keyword{stride}$*0},
+%\code{$\keyword{tid}$+$\keyword{stride}$*1},
+%\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
 A thread calls $\keyword{process\_node}$ to process a node given the current level, with nodes to be processed next frontier being pushed to $\keyword{out\_nodes}$.
@@
 arrived, which requires fair scheduling between workgroups for the
-barrier to operate correctly.
+barrier to operate correctly. \TSAdded{Without fair scheduling, the
+  barrier threads may wait indefinitely at the barrier, causing the
+  application to hang.}
 
@@
 \mysec\ref{sec:intro}, this model is not endorsed by language
-standards or vendor implementations, and is unlikely to be respected
-in the future. \TSAdded{Without a fair schedule, the mutexes and
-  barriers, which use spin-waiting in their implementation, are not
-  guaranteed to be satisfied. For example, if a workgroup is not
-  fairly scheduled after obtaining a mutex, there is no guarantee that
-  the mutex will be released. Thus the kernel may hang while other
-  workgroups wait indefinitely to obtain the mutex. }
+standards or vendor implementations, and may not be respected
+in the future. 
 %
@@
 We present our cooperative kernels programming model as an extension
-to OpenCL.  However, the cooperative kernels concept is more general,
-and could be applied to extend other GPU programming models, such as
-CUDA and HSA.  We describe the semantics of the model
+to OpenCL.  
+%However, the cooperative kernels concept is more general,
+%and could be applied to extend other GPU programming models, such as
+%CUDA and HSA.  
+We describe the semantics of the model
 (\mysec\ref{sec:semantics}), use our motivating examples to discuss
@@
 
-As with a regular OpenCL kernel (see \mysec\ref{sec:opencl}), a
-cooperative kernel is launched by the host application.  The host
-application passes parameters to the kernel and specifies a desired
-number of workgroups, each consisting of a specified number of
-threads.  Unlike in a regular kernel, the parameters to a cooperative kernel are immutable (though pointer
-parameters can refer to mutable data).
+As with a regular OpenCL kernel, a cooperative kernel is launched by
+the host application, passing parameters to the kernel and specifying
+a desired number of threads and workgroups. Unlike in a regular
+kernel, the parameters to a cooperative kernel are immutable (though
+pointer parameters can refer to mutable data).
 
@@
 %\vspace{-1mm}
-\myparagraph{Transmitted variables}
+\myparagraph{Transmitted Variables}
 %
@@
 %\vspace{-1mm}
-\myparagraph{Active workgroups}
+\myparagraph{Active Workgroups}
 %
@@
 $\offerkill$, is a workgroup-level function---it must be encountered
-uniformly by all threads in a workgroup (see
-\mysec\ref{sec:opencl}).
+uniformly by all threads in a workgroup.
 
@@
 \myfiglong{\ref{fig:overview}} illustrates the behaviour of
-$\offerfork$. After the graphics task finishes executing, there is
-space for two workgroups to join the computation of the cooperative
-kernel. Workgroup $0$ calls $\offerfork$, spawning the two new
-workgroups with ids $2$ and $3$ and broadcasting the transmit
-variables to them. Workgroups $2$ and $3$ join the computation where
-workgroup $0$ called $\offerfork$.
-
+$\offerfork$. After the graphics task finishes executing, workgroup
+$0$ calls $\offerfork$, spawning the two new workgroups with ids $2$
+and $3$. Workgroups $2$ and $3$ join the computation where workgroup
+$0$ called $\offerfork$.
+%
 %We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
@@
 %annotations rather than transmitting the entire thread state.
+%
 
@@
 %\vspace{-1mm}
-\myparagraph{Global barriers}
+\myparagraph{Global Barriers}
 %
@@
 the OpenCL 2.0 memory model enable a memory-safe
-implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  However, implementing such a barrier is
-involved, hence why we include this function as a primitive.
+implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  
+%However, implementing such a barrier is
+%involved, hence why we include this function as a primitive.
 
@@
 of active workgroups from $M$ to $M'$ occurs entirely within the
-resizing barrier, so that from a programmer's perspective, $M$ changes atomically.
-The middle $\globalbarrier$ ensures that forking occurs before killing, so that workgroups $[0, \textrm{min}(M, M') - 1]$ are left intact.
-
-Because $\resizingglobalbarrier$ can be implemented as above, we do not regard it \emph{conceptually} as a primitive of our
-model.  However, in
-\mysec\ref{sec:resizingbarrier} we show how a resizing barrier can be
-implemented more efficiently through direct interaction with the scheduler.
-
-\subsection{Programming With Cooperative Kernels}\label{sec:programmingguidelines}
+resizing barrier, so that from a programmer's perspective, $M$ changes
+atomically.  The middle $\globalbarrier$ ensures that forking occurs
+before killing, so that workgroups $[0, \textrm{min}(M, M') - 1]$ are
+left intact.
+
+Because $\resizingglobalbarrier$ can be implemented as above, we do
+not regard it \emph{conceptually} as a primitive of our model.
+However, in \mysec\ref{sec:resizingbarrier} we show how a resizing
+barrier can be implemented more efficiently through direct interaction
+with the scheduler.
+
+\subsection{Programming with Cooperative Kernels}\label{sec:programmingguidelines}
 
@@
 
-\myparagraph{A changing workgroup count} Unlike in regular OpenCL, the
+\myparagraph{A Changing Workgroup Count} Unlike in regular OpenCL, the
 value returned by $\getnumgroups$ is not fixed during the lifetime of
@@
 
-\myparagraph{Adapting work stealing}
+\myparagraph{Adapting Work Stealing}
 %
@@
 
-\myparagraph{Adapting graph traversal}
+\myparagraph{Adapting Graph Traversal}
 \myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
@@
 
-\myparagraph{Patterns for irregular algorithms}
+\myparagraph{Patterns for Irregular Algorithms}
 In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
@@
 
-Our vision is that support for cooperative kernels will be integrated
+Our vision is that cooperative kernel support will be integrated
 in the runtimes of future GPU implementations of OpenCL, with driver
@@
 
-\subsection{The megakernel mock up}\label{sec:megakernel}
+\subsection{The Megakernel Mock Up}\label{sec:megakernel}
 
@@
 megakernel is launched with as many workgroups as can be occupant
-concurrently.  One workgroup takes the role of the scheduler, and the
-scheduling logic is embedded as part of the megakernel.  The remaining
-workgroups act as a pool of workers.  A worker repeatedly queries the
-scheduler to be assigned a task.  A task corresponds to executing a
-cooperative or non-cooperative kernel.  In the non-cooperative case,
-the workgroup executes the relevant kernel function uninterrupted,
-then awaits further work.  In the cooperative case, the workgroup
-either starts from the kernel entry point or immediately jumps to a
-designated point within the kernel, depending on whether the workgroup
-is an initial workgroup of the kernel, or a forked workgroup.  In the
-latter case, the new workgroup also receives a struct containing the
-values of all relevant $\transmit$-annotated variables.
-
-\myparagraph{Simplifying assumptions}
+concurrently.  One workgroup takes the role of the
+scheduler\footnote{\TSAdded{We note that the scheduler requirements
+    given in \mysec{\ref{sec:cooperativekernels}} are agnostic to
+    whether the scheduling logic takes place on the CPU or GPU. To
+    avoid expensive communication between GPU and host, we choose to
+    implement the scheduler on the GPU.}  }, and the scheduling logic
+is embedded as part of the megakernel.  The remaining workgroups act
+as a pool of workers.  A worker repeatedly queries the scheduler to be
+assigned a task.  A task corresponds to executing a cooperative or
+non-cooperative kernel.  In the non-cooperative case, the workgroup
+executes the relevant kernel function uninterrupted, then awaits
+further work.  In the cooperative case, the workgroup either starts
+from the kernel entry point or immediately jumps to a designated point
+within the kernel, depending on whether the workgroup is an initial
+workgroup of the kernel, or a forked workgroup.  In the latter case,
+the new workgroup also receives a struct containing the values of all
+relevant $\transmit$-annotated variables.
+
+\myparagraph{Simplifying Assumptions}
 %
@@
 
-\subsection{Scheduler design}\label{sec:resizingbarrier}\label{sec:schedulerimpl}
+\subsection{Scheduler Design}\label{sec:resizingbarrier}\label{sec:schedulerimpl}
 
@@
 
-\TSAdded{We note that the scheduler requirements given in
-  \mysec{\ref{sec:cooperativekernels}} are agnostic to whether the
-  scheduling logic takes place on the CPU or GPU. To
-  avoid expensive communication between GPU and host, we choose to
-  implement the scheduler on the GPU.}
-
-\myparagraph{Scheduler contexts}
+\myparagraph{Scheduler Contexts}
 %
@@
 
-\myparagraph{Launching kernels and managing workgroups}
+\myparagraph{Launching Kernels and Managing Workgroups}
 %
@@
 
-\myparagraph{Accounting for concurrency}
-%
-Concurrent calls to $\offerfork$ must be protected by a mutex to avoid
-racing on the available workgroups. The same conflict over available
-workgroups can occur when launching a new kernel between the scheduler
-and a workgroup calling $\offerfork$. Thus, $\offerfork$ is protected
-by a mutex, called the \emph{assignment mutex}.
-%provided in the scheduler context.
-
-The assignment mutex additionally protects concurrent calls to
-$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup
-is only permitted to exit if the scheduler workgroup currently holds
-the assignment mutex; that is, the scheduler is trying to obtain work
-groups to send to a task.  Conversely, on calling $\offerfork$ a
-workgroup can only spawn new workgroups if the scheduler workgroup
-does not hold the assignment mutex.  Thus the assignment mutex, only
-ever acquired by the scheduling workgroup, \emph{indirectly} provides
-mutual exclusion between pairs of workgroups executing $\offerkill$
-and $\offerfork$ concurrently.
-
-Concurrent calls to $\offerkill$ are acceptable because only the
-workgroup with the highest id may exit, automatically ensuring mutual
-exclusion for exit-related state changes.
-
-\myparagraph{An efficient resizing barrier}
+\myparagraph{An Efficient Resizing Barrier}
 %
@@
 barrier in terms of calls to other primitives.  It is possible,
-however to implement the resizing barrier with only one call to a
-global barrier with $\offerfork$ and $\offerkill$ baked inside.
+however, to implement the resizing barrier with only one call to a
+global barrier with $\offerfork$ and $\offerkill$ inside.
 
@@
 \normalsize
+\caption{Blocking GPU applications investigated}
 \centering
@@
 \crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
-\caption{Blocking GPU applications investigated}
+
 \label{tab:applications}
@@
 We indicate how many of the original number of barriers are changed to
-resizing barriers, and how many variables need to be transmitted.  In
-all cases all barriers were converted. The remaining 2 algorithms are
-work stealing applications: each required the addition of $\offerfork$
-and $\offerkill$ at the start of the main loop, and no variables
-needed to be transmitted.  For all these examples, the porting to
-cooperative kernels was similarly straightforward to the porting
-illustrated in \mysec\ref{sec:programmingguidelines}. Most graph come
-with 2 different data sets as input, leading to 11 application/input
-pairs in total.
+resizing barriers (all of them), and how many variables need to be
+transmitted.  The remaining two algorithms are work stealing
+applications: each required the addition of $\offerfork$ and
+$\offerkill$ at the start of the main loop, and no variables needed to
+be transmitted (similar to example shown in \mysec\ref{sec:programmingguidelines}). 
+% For all these examples, the porting to cooperative
+%kernels was similarly straightforward to the porting illustrated in
+%\mysec\ref{sec:programmingguidelines}. 
+Most graph applications come with 2 different data sets as input,
+leading to 11 application/input pairs in total.
 
@@
 
-Our prototype implementation (\mysec\ref{sec:implementation}) requires
+\TSAdded{Our prototype implementation (\mysec\ref{sec:implementation}) requires
 two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
-atomics. This rules out evaluation on \nvidia and ARM Mali GPUs at
-present, since these vendors do not yet support OpenCL $2.0$,
-equivalent functionality is not provided by \nvidia's CUDA. Intel
-drivers meet our requirements on Windows, but not on Linux.
+atomics. Out of the GPUs available to us (ARM Mali, AMD, \nvidia, and
+Intel), only Intel GPUs provided robust support of these features.}
+
+%We thus evaluate our approach on three Intel GPUs, as GPUs from
+%other vendors (i.e. \nvidia, ARM, and AMD) do not yet provide these
+%features.
+%This rules out evaluation on \nvidia and ARM Mali GPUs at
+%present, since these vendors do not yet support OpenCL $2.0$,
+%equivalent functionality is not provided by \nvidia's CUDA. Intel
+%drivers meet our requirements on Windows, but not on Linux.
 %(we
@@
 %Linux kernel).
-AMD supports SVM atomics only in their Kaveri family of
-chips and only on Linux.%, requiring to compile the latest driver.
-
-Among the GPUs available to us, four met our requirements: Intel HD
-520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
-drivers for OpenCL suffer from a known defect whereby long-running
-kernels lead to unkillable defunct
-processes~\cite{amd-defunct-process}.
+%AMD supports SVM atomics only in their Kaveri family of
+%chips and only on Linux.%, requiring to compile the latest driver.
+
+%Among the GPUs available to us, four met our requirements: Intel HD
+%520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
+%drivers for OpenCL suffer from a known defect whereby long-running
+%kernels lead to unkillable defunct
+%processes~\cite{amd-defunct-process}.
 %
@@
 %
-We thus ran our experiments on the Intel GPUs. The results
-were similar across the GPUs, so for conciseness, we report only on
-the Iris 6100 GPU (driver 20.19.15.4463) with a host CPU i3-5157U. The
-Iris has a reported 47 compute units. Results for the other Intel GPUs
-are available in supplementary material.
-
-\subsection{Sizing Non-Cooperative Kernels}\label{sec:sizingnoncoop}
-
-Enabling rendering of smooth graphics in parallel with execution of irregular algorithms is an important use case for our approach.
-Because our prototype implementation is based on a megakernel that takes over the entire GPU (see \mysec\ref{sec:implementation}), we cannot assess this use case directly.
-
-We devised the following procedure to determine non-cooperative OpenCL
-workloads that simulate the computational intensity of various
-graphics rendering workloads.
+\TSAdded{We thus ran our experiments on the three Intel GPUs: HD 520, HD 5500
+and Iris 6100. The results were similar across the GPUs, so for
+conciseness, we report only on the Iris 6100 GPU (driver
+20.19.15.4463) with a host CPU i3-5157U. The Iris has a reported 47
+compute units. Results for the other Intel GPUs are available in
+supplementary material, found on the author's website.}
+
+\subsection{Sizing Non-cooperative Kernels}\label{sec:sizingnoncoop}
+
+Enabling rendering of smooth graphics in parallel with irregular
+algorithms is an important use case for our approach.  Because our
+prototype implementation is based on a megakernel that takes over the
+entire GPU (see \mysec\ref{sec:implementation}), we cannot assess this
+directly.
+
+We devised the following to determine OpenCL workloads that simulate
+the computational intensity of various graphics rendering workloads.
 %
@@
 \emph{heavy} we used WebGL demos from the Chrome
-experiments~\cite{chrome-experiments} running
-in a recent version of Firefox.
+experiments~\cite{chrome-experiments}.% running
+%in a recent version of Firefox.
 
@@
 configurations indicates that GPU computation is cheaper for basic
-display rendering compared with more complex rendering that uses
-shaders. These results guide our evaluation of cooperative scheduling
-(\mysec\ref{sec:responsiveness}).
+display rendering compared with more complex rendering.% that uses
+%shaders.
+
+%These results guide our evaluation of cooperative scheduling
+%(\mysec\ref{sec:responsiveness}).
 
@@
 
-\myparagraph{Experimental setup}
+\myparagraph{Experimental Setup}
 Invoking the cooperative scheduling primitives incurs some overhead
@@
 Recall that our mega kernel-based implementation merges the code of a
-cooperative and a non-cooperative kernel to be co-scheduled.  This
-blow-up in code size can have a negative impact on compilation, e.g.\
-due to higher register pressure, reducing the occupancy for the merged
-kernel---the maximum number of workgroups that the GPU driver will
-schedule.  This is an artifact of our proof-of-concept implementation
-strategy, and would not be a problem if our cooperative scheduling
-approach was implemented inside the GPU driver.  We thus launch both the
-original and cooperative versions of a kernel with the reduced occupancy
-bound in order to meaningfully compare execution times.
+cooperative and a non-cooperative kernel.
+% to be co-scheduled.  
+This can reduce the occupancy for the merged kernel, e.g.\ due to
+higher register pressure, This is an artifact of our prototype
+implementation, and would not be a problem if our approach was
+implemented inside the GPU driver.  We thus launch both the original
+and cooperative versions of a kernel with the reduced occupancy bound
+in order to meaningfully compare execution times.
 
@@
 \normalsize
+\caption{Cooperative kernel slowdown w/o multitasking}
 \centering
@@
 %\vspace{-2mm}
-\caption{Cooperative kernel slowdown w/o multitasking}
 \label{tab:overhead}
@@
 
-\myparagraph{Experimental setup} For a given cooperative kernel and
+\myparagraph{Experimental Setup} For a given cooperative kernel and
 its input, we launch the kernel and then repeatedly schedule a
@@
 the three classes of graphics rendering workload discussed in
-\mysec\ref{sec:sizingnoncoop}. In practice, we use a matrix
-multiplication kernel as the non-cooperative workload, with matrix
+\mysec\ref{sec:sizingnoncoop}. In practice, we use matrix
+multiplication as the non-cooperative workload, with matrix
 dimensions tailored to reach the appropriate execution duration.  We
@@
 barrier lets the scheduler collect all needed workgroups as soon as
-they hit a resizing barrier. As we can see, the gather time of the
+they hit a resizing barrier. 
+As we can see, the gather time of the
 query barrier is almost stable w.r.t.\ the number of workgroups that
@@
 non-cooperative kernels can be scheduled (median over all applications
-and inputs). We use the median as our data included some outliers;
-these outliers happen with applications/inputs in which the resizing
-barrier are not called very frequently and the graphics task requires
-half or more workgroups. For example, a medium graphics workload for
-bfs on the rmat input has over an 8$\times$ overhead when asking for
-all but one of the workgroups. As \myfiglong\ref{fig:performance}
-shows, most of our applications/inputs are much better behaved than
-this. Future work will examine these problem applications/inputs in
-more detail, possibly inserting more calls to $\offerfork$ and
-$\offerkill$.
+and inputs). Our data included some outliers that occur with
+benchmarks in which the resizing barrier are not called very
+frequently and the graphics task requires half or more workgroups. For
+example, a medium graphics workload for bfs on the rmat input has over
+an 8$\times$ overhead when asking for all but one of the
+workgroups. As \myfiglong\ref{fig:performance} shows, most of our
+benchmarks are much better behaved than this. Future work
+will examine these problem benchmarks in more detail,
+possibly inserting more resizing calls.
 
 We show results for the three workloads listed in
-\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
-period graph correspond to the period goals of the workloads: the
-higher (resp. lower) line corresponds to a period of 70ms (resp. 40ms)
-for the light (resp. medium and heavy) workload.
+\mysec~\ref{sec:sizingnoncoop}. The horizontal lines in the period
+graph correspond to the goals of the workloads: the higher
+(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for
+the light (resp. medium and heavy) workload.
 
@@
 of the workgroups executing the non-cooperative kernel are able to
-meet their goal period (70 and 40 ms resp.). However, a quarter of the
-workgroups are not sufficient to meet the goal for the heavy workload
-(giving a median period of 104 ms). If half of the workgroups are
-allocated to the non-cooperative kernel, the heavy workload achieves
-its goal period (median of 40 ms). 
+meet their goal period (70 and 40 ms resp.). However, this is not
+sufficient to meet the goal for the heavy workload (giving a median
+period of 104 ms). If half of the workgroups are allocated to the
+non-cooperative kernel, the heavy workload achieves its goal period
+(median of 40 ms).
 %
 Yet, as expected, allocating more non-cooperative workgroups increases
-the execution time of the cooperative kernel.
+the overhead of the cooperative kernel.
 %
 Still, heavy workloads meet their period by allocating half
-non-cooperative workgroups, incurring a slow down of less than
+of the workgroups, incurring a slow down of less than
 1.5$\times$ (median). Light and medium workloads meet their period
-with only a small overhead of the cooperative kernel; 
-1.04$\times$ and 1.08$\times$ median slowdown respectively.
-
-We consider these results encouraging, since they provide a lower
-bound on potential performance of our cooperative kernels
-model. Implementing the model natively, with device-specific runtime
-support, could only improve performance and responsiveness compared
-with that of our megakernel-based prototype.
+with only a small overhead; 1.04$\times$ and 1.08$\times$ median
+slowdown respectively.
+
+%We consider these results encouraging, since they provide a lower
+%bound on potential performance of our cooperative kernels
+%model. Implementing the model natively, with device-specific runtime
+%support, could only improve performance and responsiveness compared
+%with that of our megakernel-based prototype.
 
@@
 \normalsize
+\caption{Overhead of kernel level preemption vs cooperative kernels for three
+graphics workloads}
+
 \centering
@@
 
-\caption{Overhead of kernel level preemption vs cooperative kernels for three
-graphics workloads}
 \label{tab:preemption}
@@
 
-Some recent GPUs, including Nvidia's Pascal architecture, provide
-hardware support for instruction-level preemption.  Few details of how
-this works are available, but from reading Nvidia documentation and
-online material~\cite{PascalWhitepaper,anandtech}, our understanding
-is that preemption of entire kernels, but not of individual workgroups
-is supported.  Intel GPUs do not provide this feature, and our OpenCL
-2.0 prototype of cooperative kernels cannot run on Nvidia GPUs, making
-a direct comparison impossible.  We present here a theoretical
+Nvidia's recent Pascal architecture provides hardware support for
+instruction-level preemption~\cite{PascalWhitepaper,anandtech},
+however, preemption of entire kernels, but not of individual
+workgroups is supported.  Intel GPUs do not provide this feature, and
+our OpenCL prototype of cooperative kernels cannot run on Nvidia GPUs,
+making a direct comparison impossible.  We present here a theoretical
 analysis of the overheads associated with sharing the GPU between
@@
 
-Our poor performance for the heavy workload is because the graphics
-task requires half of the workgroups, which cripples the cooperative
-kernel enough that $\offerfork$ calls are not issued as frequently.
-Future work may examine how to insert more resizing calls in these
+Our low performance for heavy workloads is because the graphics task
+requires half of the workgroups, crippling the cooperative kernel
+enough that $\offerfork$ calls are not issued as frequently.  Future
+work may examine how to insert more resizing calls in these
 applications to decrease this overhead.
 %
-These results suggest that a hybrid preemption scheme may work well
-for multi-tasking and graphics. That is, the cooperative approach
-works well for light and medium tasks and requires no special hardware
-or preemption state saving; on the other hand, heavy graphics tasks
-benefit from the coarser grained, kernel-level preemption strategy.
+These results suggest that a hybrid preemption scheme may work
+well. That is, the cooperative approach works well for light and
+medium tasks; on the other hand, heavy graphics tasks benefit from the
+coarser grained, kernel-level preemption strategy.
 
@@
 
-\myparagraph{Irregular algorithms and persistent kernels}
+\myparagraph{Irregular Algorithms and Persistent kernels}
 %
@@
 
-\myparagraph{GPU multitasking and scheduling}
+\myparagraph{GPU Multitasking and Scheduling}
 %
@@
 
-\myparagraph{Cooperative multitasking}
+\myparagraph{Cooperative Multitasking}
 %
@@
 
-In the context of GPUs, cooperative multitasking has been used to
-refer to the standard programming model, where the onus is on the GPU kernel
-to complete execution within a reasonable time budget~\cite{adriaens2012case,CPE:CPE1722}.
-%
-This is in contrast to our cooperative kernels, which specifically aim to support the needs of long-running GPU tasks.
+%In the context of GPUs, cooperative multitasking has been used to
+%refer to the standard programming model, where the onus is on the GPU kernel
+%to complete execution within a reasonable time budget~\cite{adriaens2012case,CPE:CPE1722}.
+%
+%This is in contrast to our cooperative kernels, which specifically aim to support the needs of long-running GPU tasks.
 
@@
 GPUs~\cite{DBLP:conf/ppopp/WangDPWRO16,DBLP:conf/oopsla/PaiP16}.
+
+\section*{Acknowledgments}
+We are grateful to Lee Howes, Bernhard Kainz, Paul Kelly, Christopher
+Lidbury, Steven McDonagh, Sreepathi Pai, and Andrew Richards for
+insightful comments throughout the work. We thank the FSE reviewers
+for their thorough evaluations and feedback. This work is supported in
+part by EPSRC Fellowship EP/N026314, and a gift from Intel Corporation.
+
 %  As more vendors
@@
 %extensions.
+
+\myparagraph{Accounting for concurrency}
+%
+Concurrent calls to $\offerfork$ must be protected by a mutex to avoid
+racing on the available workgroups. The same conflict over available
+workgroups can occur when launching a new kernel between the scheduler
+and a workgroup calling $\offerfork$. Thus, $\offerfork$ is protected
+by a mutex, called the \emph{assignment mutex}.
+%provided in the scheduler context.
+
+The assignment mutex additionally protects concurrent calls to
+$\offerfork$ and $\offerkill$.  On calling $\offerkill$, a workgroup
+is only permitted to exit if the scheduler workgroup currently holds
+the assignment mutex; that is, the scheduler is trying to obtain work
+groups to send to a task.  Conversely, on calling $\offerfork$ a
+workgroup can only spawn new workgroups if the scheduler workgroup
+does not hold the assignment mutex.  Thus the assignment mutex, only
+ever acquired by the scheduling workgroup, \emph{indirectly} provides
+mutual exclusion between pairs of workgroups executing $\offerkill$
+and $\offerfork$ concurrently.
+
+Concurrent calls to $\offerkill$ are acceptable because only the
+workgroup with the highest id may exit, automatically ensuring mutual
+exclusion for exit-related state changes.

### 2017-07-06 write-ups/FSE/paper.tex
update to get the appendix working
Pass over intro.

@@
 \usepackage[normalem]{ulem}
-%\usepackage{amsmath}
+\usepackage{amsmath}
 \usepackage{calc}
@@
 Many interesting data-parallel algorithms are \emph{irregular}: the
-amount of work to be processed is unknown ahead of time and may be
-determined during computation.  There is growing interest in
+amount of work to be processed is unknown ahead of time and may change dynamically in a workload-dependent manner.
+There is growing interest in
 accelerating such algorithms on
 GPUs~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/egh/CedermanT08,TPO10,BNP12,Pannotia}.
-
+%
 Irregular algorithms usually require \emph{blocking synchronization}
@@
 \includegraphics[width=\columnwidth]{overview.pdf}
-\caption{Cooperative kernels can flexibly resize to let other tasks
-(e.g. graphics) run concurrently}
+\caption{Cooperative kernels can flexibly resize to let other tasks,
+e.g.\ graphics, run concurrently}
 \label{fig:overview}
@@
 First, the execution model does not enable
-multitasking, because a workgroup effectively \emph{owns} a compute
+multitasking, since a workgroup effectively \emph{owns} a compute
 unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
-tasks (e.g.\ graphics rendering).
+tasks (e.g.\ rendering).
 %Precisely because
@@
 future, it will be desirable for a mobile GPU driver to power down
-some compute units, and hence suspend the execution of the occupant
-workgroup, if the battery level is low.
+some compute units, suspending execution of associated occupant
+workgroups, if the battery level is low.
 %Implementing energy
@@
 
-For example, Nvidia's Pascal architecture supports
-preemption~\cite{PascalWhitepaper}. Despite this, the device still
-does not guarantee forward progress. Indeed, on a GTX Titan X (Pascal)
-we still observe starvation: we found that a global barrier can be
-executed successfully with 56 workgroups, but deadlocks with 57
-workgroups.  Thus, although preemption is supported, there is not fair
-scheduling.
+For example, preemption is supported by Nvidia's Pascal architecture~\cite{PascalWhitepaper}, but on a GTX Titan X (Pascal)
+we still observe starvation: a global barrier
+executes successfully with 56 workgroups, but deadlocks with 57
+workgroups, indicating that scheduling is still unfair.
 
@@
   (see \mysec{\ref{sec:relatedwork}}), a multitasking approach used
-  historically when preemption was not available or, in the case of
+  historically when preemption was not available or, as in the case of
   GPUs, too costly. Our approach avoids the cost of arbitrary
@@
 constraints, this time should be low; our experimental study
-(\mysec{\ref{sec:responsiveness}}) shows that in practice, the
+(\mysec{\ref{sec:responsiveness}}) shows that, in practice, the
 gather-time for our applications is acceptable for a range of graphics
@@
 \begin{figure*}
-\begin{center}
-
+\center
 \[
+$$
 \inferrule{
@@
 (\textsc{Thread-Step})
+$$
 \]
@@
 \[
+$$
 \inferrule{
@@
 (\textsc{Kill-No-Op})
+$$
 \]
@@
 (\textsc{Kill})
+$$
 \]
@@
 (\textsc{Fork})
+$$
 \]
@@
 (\textsc{Barrier})
+$$
 \]
@@
 (\textsc{Resizing-Barrier})
+$$
 \]
-
-\end{center}
 

### 2017-07-06 write-ups/FSE/paper.tex
This is such a good paper.

@@
 \myparagraph{Execution Model}
-OpenCL~\cite[p.\ 31]{opencl2Spec} and CUDA~\cite{cuda-75} specifically makes no guarantees about fair scheduling between
+OpenCL~\cite[p.\ 31]{opencl2Spec} and CUDA~\cite{cuda-75} specifically make no guarantees about fair scheduling between
 workgroups executing the same kernel. 
@@
 HSA provides limited, one-way guarantees,
-stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
+stating~\cite[p. 46]{HSAprogramming11}: \emph{``Work-group A can wait
   for values written by work-group B without deadlock provided ... (if) A
-  comes after B in work-group flattened ID order}. This is not sufficient to support blocking algorithms that use
+  comes after B in work-group flattened ID order''}. This is not sufficient to support blocking algorithms that use
 mutexes and inter-workgroup barriers, both of which require \emph{symmetric} communication between
@@
 
-If a task was obtained, then the workgroup processes it
+If a task is obtained, then the workgroup processes it
 (line~\ref{line:wksteal:processtask}), which may lead to new tasks
@@
 %\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
-A thread calls $\keyword{process\_node}$ to process a node given the current level, with nodes to be processed next frontier being pushed to $\keyword{out\_nodes}$.
+A thread calls $\keyword{process\_node}$ to process a node given the current level, with nodes to be processed during the next frontier being pushed to $\keyword{out\_nodes}$.
 After processing the frontier, the threads swap their
@@
 on CPU barrier designs~\cite[ch. 17]{HS08}.  These barriers employ
-spinning to ensure threads wait at the barrier until all threads have
-arrived, which requires fair scheduling between workgroups for the
+spinning to ensure each thread waits at the barrier until all threads have
+arrived, thus fair scheduling between workgroups is required for the
 barrier to operate correctly. \TSAdded{Without fair scheduling, the

### 2017-07-06 write-ups/FSE/paper.tex
This is such a good paper.

@@
 thread; $\offerkill$ and $\offerfork$, the key functions that enable
-cooperative scheduling; and $\globalbarrier$ and $\mathsf{resizing\_global\_}$ $\mathsf{barrier}$
+cooperative scheduling; and $\globalbarrier$ and $\mathsf{resizing\_global\_barrier}$
 %$\resizingglobalbarrier$

### 2017-07-06 write-ups/FSE/paper.tex
Done pass over Section 3.

@@
 thread; $\offerkill$ and $\offerfork$, the key functions that enable
-cooperative scheduling; and $\globalbarrier$ and $\mathsf{resizing\_global\_barrier}$
-%$\resizingglobalbarrier$
+cooperative scheduling; and $\globalbarrier$ and $\resizingglobalbarrier$
 primitives for inter-workgroup synchronisation.
@@
 workgroups, this indicates that the kernel should be executed with a
-maximum of $N$ workgroups, and that as many workgroups as possible, up
+\emph{maximum} of $N$ workgroups, and that as many workgroups as possible, up
 to this limit, are desired.  However, the scheduler may initially
@@
 %
-Because workgroups of a cooperative kernel are fairly scheduled, so a
-global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
+Because workgroups of a cooperative kernel are fairly scheduled, a
+global barrier primitive can be provided.  We specify two variants: $\globalbarrier$
 and $\resizingglobalbarrier$.
@@
 
-The semantics of $\resizingglobalbarrier$ can be modelled via calling $\offerfork$ and $\offerkill$
-surrounded, and separated by calls to a $\globalbarrier$.
+The semantics of $\resizingglobalbarrier$ can be modelled via calling $\offerfork$ and $\offerkill$,
+surrounded and separated by calls to a $\globalbarrier$.
 %
@@
 of active workgroups from $M$ to $M'$ occurs entirely within the
-resizing barrier, so that from a programmer's perspective, $M$ changes
-atomically.  The middle $\globalbarrier$ ensures that forking occurs
+resizing barrier, so that $M$ changes atomically from a programmer's perspective.  The middle $\globalbarrier$ ensures that forking occurs
 before killing, so that workgroups $[0, \textrm{min}(M, M') - 1]$ are
@@
 the scheduler after processing each task.  If tasks are sufficiently
-fast to process then the scheduler have ample opportunities to
+fast to process then the scheduler will have ample opportunities to
 de-schedule workgroups.  But if tasks are very time-consuming to

### 2017-07-06 write-ups/FSE/paper.tex
Finished pass over section 5.

@@
 concurrently.  One workgroup takes the role of the
-scheduler\footnote{\TSAdded{We note that the scheduler requirements
+scheduler,\footnote{\TSAdded{We note that the scheduler requirements
     given in \mysec{\ref{sec:cooperativekernels}} are agnostic to
@@
     avoid expensive communication between GPU and host, we choose to
-    implement the scheduler on the GPU.}  }, and the scheduling logic
+    implement the scheduler on the GPU.}  } and the scheduling logic
 is embedded as part of the megakernel.  The remaining workgroups act
@@
 implementation of cooperative kernels would not use the megakernel
-approach, so we do not deem the engineering effort associated with
-lifting these restrictions to be worthwhile.
+approach, so we did not deem the engineering effort associated with
+lifting these restrictions in our prototype to be worthwhile.
 
@@
 our case, the megakernel) must track the state of workgroups,
-i.e.\ whether a workgroup is waiting or computing a kernel, maintain
+i.e.\ whether a workgroup is waiting or computing a kernel; maintain
 consistent context states for each kernel, e.g.\ tracking the number
-of active workgroups, and provide a safe way for these states to be
+of active workgroups; and provide a safe way for these states to be
 modified in response to $\offerfork$/$\offerkill$. We discuss these
@@
 Incorporating $\offerfork$ and $\offerkill$ into such a barrier is
-straight forward. Upon entering the barrier, the slaves first execute
+straightforward. Upon entering the barrier, the slaves first execute
 $\offerkill$, possibly exiting. The master then waits for $M$ slaves

### 2017-07-06 write-ups/FSE/paper.tex
added updated licensing

--- 2017-07-06 write-ups/FSE/paper.tex
@@
 %%%
-%%% If you see 'none' in the 'setcopyright' statement below,
-%%% please first submit your publishing-rights agreement with ACM (follow link on submission page),
-%%% and then visit the instructions page again (follow link in author-kit message)
-%%% in order to copy the correct 'setcopyright' statement.  Please contact us in case of questions.
-%%%
-\setcopyright{none}
+\setcopyright{acmlicensed}
 \acmDOI{10.1145/3106237.3106265}
 \acmISBN{978-1-4503-5105-8/17/09}
-\acmConference[ESEC/FSE'17]{2017 11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{September 4-8, 2017}{Paderborn, Germany}
+\acmConference[ESEC/FSE'17]{2017 11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{September 4--8, 2017}{Paderborn, Germany}
 \acmYear{2017}

### 2017-07-06 write-ups/FSE/paper.tex
Various.

--- 2017-07-06 write-ups/FSE/paper.tex
@@
 $\offerkill$ at the start of the main loop, and no variables needed to
-be transmitted (similar to example shown in \mysec\ref{sec:programmingguidelines}). 
+be transmitted (similar to example discussed in \mysec\ref{sec:programmingguidelines}). 
 % For all these examples, the porting to cooperative
@@
 two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
-atomics. Out of the GPUs available to us (ARM Mali, AMD, \nvidia, and
-Intel), only Intel GPUs provided robust support of these features.}
+atomics. Out of the GPUs available to us, from ARM, AMD, \nvidia, and
+Intel, only Intel GPUs provided robust support of these features.}
 
@@
 %
-\TSAdded{We thus ran our experiments on the three Intel GPUs: HD 520, HD 5500
+\TSAdded{We thus ran our experiments on three Intel GPUs: HD 520, HD 5500
 and Iris 6100. The results were similar across the GPUs, so for
@@
 
-We devised the following to determine OpenCL workloads that simulate
+We devised the following method to determine OpenCL workloads that simulate
 the computational intensity of various graphics rendering workloads.
@@
 
-Our results are the following: $P=70ms$ and $E=3ms$ for light;
-$P=40ms$, $E=3ms$ for medium; and $P=40ms$, $E=10ms$ for heavy. For
-medium and heavy, the 40ms period coincides with the human persistence
-of vision. The 3ms execution duration of both light and medium
+Our results are the following: $P=70\mathit{ms}$ and $E=3\mathit{ms}$ for light;
+$P=40\mathit{ms}$, $E=3\mathit{ms}$ for medium; and $P=40\mathit{ms}$, $E=10\mathit{ms}$ for heavy. For
+medium and heavy, the $40\mathit{ms}$ period coincides with the human persistence
+of vision. The $3\mathit{ms}$ execution duration of both light and medium
 configurations indicates that GPU computation is cheaper for basic
@@
 graph correspond to the goals of the workloads: the higher
-(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for
+(resp. lower) line corresponds to a period of $70\mathit{ms}$ (resp.\ $40\mathit{ms}$) for
 the light (resp. medium and heavy) workload.
@@
 the needs of any of our three workloads; e.g.\ a heavy workload
-averages a period of 939 ms. As more workgroups are dedicated to
+averages a period of 939 $\mathit{ms}$. As more workgroups are dedicated to
 non-cooperative kernels, they execute quickly enough to be scheduled
@@
 of the workgroups executing the non-cooperative kernel are able to
-meet their goal period (70 and 40 ms resp.). However, this is not
+meet their goal period (70 and 40 $\mathit{ms}$ resp.). However, this is not
 sufficient to meet the goal for the heavy workload (giving a median
-period of 104 ms). If half of the workgroups are allocated to the
+period of 104 $\mathit{ms}$). If half of the workgroups are allocated to the
 non-cooperative kernel, the heavy workload achieves its goal period
-(median of 40 ms).
+(median of 40 $\mathit{ms}$).
 %
@@
 execute without interruption.  If we assume the cost of preemption is
-negligible (e.g.\ Nvidia have reported preemption times of 0.1 ms for
+negligible (e.g.\ Nvidia have reported preemption times of 0.1 $\mathit{ms}$ for
 Pascal~\cite{anandtech}), then the overhead associated with switching

### 2017-07-06 write-ups/FSE/paper.tex
SQUEEZE

@@
 Irregular algorithms usually require \emph{blocking synchronization}
-between workgroups.  For example, many graph algorithms employ a level-by-level
-strategy, requiring a global barrier between each level, and work
-stealing algorithms require each workgroup to maintain a queue, which
-is typically protected by a mutex to enable stealing by other
+between workgroups, e.g.\ many graph algorithms use a level-by-level
+strategy, with a global barrier between levels; work
+stealing algorithms require each workgroup to maintain a queue,
+typically mutex-protected, to enable stealing by other
 workgroups.
@@
 executes successfully with 56 workgroups, but deadlocks with 57
-workgroups, indicating that scheduling is still unfair.
+workgroups, indicating unfair scheduling.
 
@@
 \TSAdded{Cooperative kernels allow for \emph{cooperative multitasking}
-  (see \mysec{\ref{sec:relatedwork}}), a multitasking approach used
-  historically when preemption was not available or, as in the case of
-  GPUs, too costly. Our approach avoids the cost of arbitrary
+  (see \mysec{\ref{sec:relatedwork}}), used
+  historically when preemption was not available or too costly. Our approach avoids the cost of arbitrary
   preemption as the state of a workgroup killed via $\offerkill$ does
@@
   design as, in the case of a global barrier, adopting yield would
-  force the cooperative kernel to block \emph{completely} in the case
-  where a single workgroup yields, stalling the kernel until the given
+  force the cooperative kernel to block \emph{completely} when a single workgroup yields, stalling the kernel until the given
   workgroup resumes.  Instead, our $\offerkill$ allows a kernel to
@@
 
-Placing the new primitives requires manual effort, but is straightforward for the representative set of
-GPU-accelerated irregular algorithms that we have ported to use cooperative kernels.  Our experiments show that the model can enable efficient multitasking of cooperative and non-cooperative tasks.
+Placing the primitives manually is straightforward for the representative set of
+GPU-accelerated irregular algorithms we have ported so far.  Our experiments show that the model can enable efficient multitasking of cooperative and non-cooperative tasks.
 

### 2017-07-06 write-ups/FSE/paper.tex
Done.

@@
 configurations, the average gather and execution times for the
-non-cooperative kernel w.r.t. the quantity of workgroups allocated to
+non-cooperative kernel with respect to the quantity of workgroups allocated to
 it.  A logarithmic scale is used for time since gather times tend to
@@
 
-The left most graph illustrates a work
+The left-most graph illustrates a work
 stealing example.  When the non-cooperative kernel is given only one
@@
 As we can see, the gather time of the
-query barrier is almost stable w.r.t.\ the number of workgroups that
+query barrier is almost stable with respect to the number of workgroups that
 needs to be collected.
@@
 workgroups. As \myfiglong\ref{fig:performance} shows, most of our
-benchmarks are much better behaved than this. Future work
-will examine these problem benchmarks in more detail,
+benchmarks are much better behaved than this. In future work
+is required to examine the problematic benchmarks in more detail,
 possibly inserting more resizing calls.
@@
 the needs of any of our three workloads; e.g.\ a heavy workload
-averages a period of 939 $\mathit{ms}$. As more workgroups are dedicated to
+averages a period of $939\mathit{ms}$. As more workgroups are dedicated to
 non-cooperative kernels, they execute quickly enough to be scheduled
@@
 of the workgroups executing the non-cooperative kernel are able to
-meet their goal period (70 and 40 $\mathit{ms}$ resp.). However, this is not
+meet their goal period (70 and $40\mathit{ms}$ resp.). However, this is not
 sufficient to meet the goal for the heavy workload (giving a median
-period of 104 $\mathit{ms}$). If half of the workgroups are allocated to the
+period of $104\mathit{ms}$). If half of the workgroups are allocated to the
 non-cooperative kernel, the heavy workload achieves its goal period
-(median of 40 $\mathit{ms}$).
+(median of $40\mathit{ms}$).
 %

### 2017-07-06 write-ups/FSE/paper.tex
fixed Ally's comment

@@
 grows, the execution time decreases and the gather time
-increases.%\footnote{The right figure has only 3 points per line:
+increases. %\footnote{The right figure has only 3 points per line:
   %recall that for bfs there are no runs with ``all-but-one''
@@
 %
-the gather time is higher on the right figure: the G3 circuit input
+The gather time is higher on the rightmost figure as the G3 circuit input
 graph is rather wide than deep, so the graph algorithm reaches

### 2017-07-06 write-ups/FSE/paper.tex
added a note about special hardware support being required for preemption

@@
 %\newcommand{\TSComment}[1]{\textcolor{blue}{TS: #1}}
-\newcommand{\TSAdded}[1]{\textcolor{blue}{#1}}
+\newcommand{\TSAdded}[1]{#1}
 \newcommand{\TSEmph}[1]{\textcolor{red}{#1}}
@@
 execute without interruption.  If we assume the cost of preemption is
-negligible (e.g.\ Nvidia have reported preemption times of 0.1 $\mathit{ms}$ for
-Pascal~\cite{anandtech}), then the overhead associated with switching
+negligible (e.g.\ Nvidia have reported preemption times of 0.1
+$\mathit{ms}$ for Pascal~\cite{anandtech}, because of
+special hardware support), then the overhead associated with switching
 between compute and graphics every $P$ time steps is $P/(P-D)$.
@@
 work may examine how to insert more resizing calls in these
-applications to decrease this overhead.
+applications to address this.
 %
@@
 medium tasks; on the other hand, heavy graphics tasks benefit from the
-coarser grained, kernel-level preemption strategy.
+coarser grained, kernel-level preemption strategy. However, the
+preemption strategy requires specialised hardware support
+in order to be efficient.
 

### 2017-07-06 write-ups/FSE/paper.tex
Remove trailing whitespaces
Remove parenthesis

@@
 the scheduler then sends $N$ workgroups from the available workgroups
-to execute kernel $K$
+to execute kernel $K$.
 % with
 %kernel context $\KC$.
-(If the new kernel $K$ is itself a cooperative kernel, the scheduler
+If the new kernel $K$ is itself a cooperative kernel, the scheduler
 would be free to provide $K$ with fewer than $N$ active workgroups
-initially).
+initially.
 

### 2017-07-06 write-ups/FSE/extended/paper.tex
Extended version of paper.

--- 2017-07-06 write-ups/FSE/paper.tex
@@
 \setcopyright{acmlicensed}
-\acmDOI{10.1145/3106237.3106265}
-\acmISBN{978-1-4503-5105-8/17/09}
-\acmConference[ESEC/FSE'17]{2017 11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{September 4--8, 2017}{Paderborn, Germany}
+%\acmDOI{10.1145/3106237.3106265}
+%\acmISBN{978-1-4503-5105-8/17/09}
+%\acmConference[ESEC/FSE'17]{2017 11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering}{September 4--8, 2017}{Paderborn, Germany}
 \acmYear{2017}
@@
 
-\title{Cooperative Kernels: GPU Multitasking for Blocking Algorithms}
+\title{Cooperative Kernels: GPU Multitasking for Blocking Algorithms (Extended Version)}
 
@@
 We describe the semantics of the model
-(\mysec\ref{sec:semantics}), use our motivating examples to discuss
+(\mysec\ref{sec:semantics}), presenting a more formal operational semantics in Appendix~\ref{appendix:semantics}
+and discussing possible alternative semantic choices in Appendix~\ref{appendix:semanticalternatives}, use our motivating examples to discuss
 programmability (\mysec\ref{sec:programmingguidelines}) and outline
@@
 20.19.15.4463) with a host CPU i3-5157U. The Iris has a reported 47
-compute units. Results for the other Intel GPUs are available in
-supplementary material, found on the author's website.}
+compute units. Results for the other Intel GPUs are presented in Appendix~\ref{appendix:extragraphs}.}
 
@@
 graphs show a representative sample of our results; the full set of
-graphs for all configurations is provided in supplementary material.
+graphs for all configurations is provided in Appendix~\ref{appendix:extragraphs}.
 
@@
 \bibliography{tyler}
-
-\iffalse
 
@@
 
-\fi
 \end{document}

### 2017-07-06 write-ups/FSE/extended/paper.tex
Made paper cite extended version.

@@
 
-\bibliographystyle{abbrvnat}
-\bibliography{tyler}
+\begin{thebibliography}{34}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Alglave et~al.(2015)Alglave, Batty, Donaldson, Gopalakrishnan, Ketema,
+  Poetzl, Sorensen, and Wickerson]{ABDGKPSW-2015}
+J.~Alglave, M.~Batty, A.~F. Donaldson, G.~Gopalakrishnan, J.~Ketema, D.~Poetzl,
+  T.~Sorensen, and J.~Wickerson.
+\newblock {GPU} concurrency: Weak behaviours and programming assumptions.
+\newblock In \emph{ASPLOS}, pages 577--591. ACM, 2015.
+
+\bibitem[Burtscher et~al.(2012)Burtscher, Nasre, and Pingali]{BNP12}
+M.~Burtscher, R.~Nasre, and K.~Pingali.
+\newblock A quantitative study of irregular programs on {GPU}s.
+\newblock In \emph{IISWC}, pages 141--151. IEEE, 2012.
+
+\bibitem[Cederman and Tsigas(2008)]{DBLP:conf/egh/CedermanT08}
+D.~Cederman and P.~Tsigas.
+\newblock On dynamic load balancing on graphics processors.
+\newblock In \emph{EGGH}, pages 57--64, 2008.
+
+\bibitem[Che et~al.(2013)Che, Beckmann, Reinhardt, and Skadron]{Pannotia}
+S.~Che, B.~M. Beckmann, S.~K. Reinhardt, and K.~Skadron.
+\newblock Pannotia: Understanding irregular {GPGPU} graph applications.
+\newblock In \emph{IISWC}, pages 185--195, 2013.
+
+\bibitem[Davidson et~al.(2014)Davidson, Baxter, Garland, and
+  Owens]{DBLP:conf/ipps/DavidsonBGO14}
+A.~A. Davidson, S.~Baxter, M.~Garland, and J.~D. Owens.
+\newblock Work-efficient parallel {GPU} methods for single-source shortest
+  paths.
+\newblock In \emph{IPDPS}, pages 349--359, 2014.
+
+\bibitem[{Google}()]{chrome-experiments}
+{Google}.
+\newblock {C}hrome {E}xperiments.
+\newblock \url{https://www.chromeexperiments.com}.
+
+\bibitem[Gupta et~al.(2012)Gupta, Stuart, and Owens]{owens-persistent}
+K.~Gupta, J.~Stuart, and J.~D. Owens.
+\newblock A study of persistent threads style {GPU} programming for {GPGPU}
+  workloads.
+\newblock In \emph{InPar}, pages 1--14, 2012.
+
+\bibitem[Harish and Narayanan(2007)]{DBLP:conf/hipc/HarishN07}
+P.~Harish and P.~J. Narayanan.
+\newblock Accelerating large graph algorithms on the {GPU} using {CUDA}.
+\newblock In \emph{HiPC}, pages 197--208, 2007.
+
+\bibitem[Herlihy and Shavit(2008)]{HS08}
+M.~Herlihy and N.~Shavit.
+\newblock \emph{The Art of Multiprocessor Programming}.
+\newblock Morgan Kaufmann Publishers Inc., 2008.
+
+\bibitem[Howes()]{PersonalCommunicationHowes}
+L.~W. Howes.
+\newblock Personal communication.
+\newblock Editor of the OpenCL 2.0 specification. 10 September 2016.
+
+\bibitem[{HSA Foundation}(2016)]{HSAprogramming11}
+{HSA Foundation}.
+\newblock {HSAIL} virtual {ISA} and programming model, compiler writer, and
+  object format {(BRIG)}, February 2016.
+\newblock \url{http://www.hsafoundation.com/standards/}.
+
+\bibitem[Kaleem et~al.(2016)Kaleem, Venkat, Pai, Hall, and
+  Pingali]{DBLP:conf/ipps/KaleemVPHP16}
+R.~Kaleem, A.~Venkat, S.~Pai, M.~W. Hall, and K.~Pingali.
+\newblock Synchronization trade-offs in {GPU} implementations of graph
+  algorithms.
+\newblock In \emph{IPDPS}, pages 514--523, 2016.
+
+\bibitem[{Khronos Group}(2015)]{opencl2Spec}
+{Khronos Group}.
+\newblock The {OpenCL} specification version: 2.0 (rev. 29), July 2015.
+\newblock \url{https://www.khronos.org/registry/cl/specs/opencl-2.0.pdf}.
+
+\bibitem[M{\'{e}}ndez{-}Lojo et~al.(2012)M{\'{e}}ndez{-}Lojo, Burtscher, and
+  Pingali]{DBLP:conf/ppopp/Mendez-LojoBP12}
+M.~M{\'{e}}ndez{-}Lojo, M.~Burtscher, and K.~Pingali.
+\newblock A {GPU} implementation of inclusion-based points-to analysis.
+\newblock In \emph{PPoPP}, pages 107--116, 2012.
+
+\bibitem[Merrill et~al.(2015)Merrill, Garland, and
+  Grimshaw]{DBLP:journals/topc/MerrillGG15}
+D.~Merrill, M.~Garland, and A.~S. Grimshaw.
+\newblock High-performance and scalable {GPU} graph traversal.
+\newblock \emph{{TOPC}}, 1\penalty0 (2):\penalty0 14, 2015.
+
+\bibitem[Muyan{-}{\"{O}}z{\c{c}}elik and
+  Owens(2016)]{DBLP:conf/ppopp/Muyan-OzcelikO16}
+P.~Muyan{-}{\"{O}}z{\c{c}}elik and J.~D. Owens.
+\newblock Multitasking real-time embedded {GPU} computing tasks.
+\newblock In \emph{PMAM}, pages 78--87, 2016.
+
+\bibitem[Nobari et~al.(2012)Nobari, Cao, Karras, and
+  Bressan]{DBLP:conf/ppopp/NobariCKB12}
+S.~Nobari, T.~Cao, P.~Karras, and S.~Bressan.
+\newblock Scalable parallel minimum spanning forest computation.
+\newblock In \emph{PPoPPP}, pages 205--214, 2012.
+
+\bibitem[{NVIDIA}(2016)]{PascalWhitepaper}
+{NVIDIA}.
+\newblock {NVIDIA Tesla P100}, 2016.
+\newblock Whitepaper WP-08019-001\_v01.1.
+
+\bibitem[\nvidia(2016)]{cuda-75}
+\nvidia.
+\newblock {CUDA} {C} programming guide, version 7.5, July 2016.
+
+\bibitem[Pai and Pingali(2016)]{DBLP:conf/oopsla/PaiP16}
+S.~Pai and K.~Pingali.
+\newblock A compiler for throughput optimization of graph algorithms on {GPUs}.
+\newblock In \emph{OOPSLA}, pages 1--19, 2016.
+
+\bibitem[Park et~al.(2015)Park, Park, and Mahlke]{DBLP:conf/asplos/ParkPM15}
+J.~J.~K. Park, Y.~Park, and S.~A. Mahlke.
+\newblock Chimera: Collaborative preemption for multitasking on a shared {GPU}.
+\newblock In \emph{ASPLOS}, pages 593--606, 2015.
+
+\bibitem[Prabhu et~al.(2011)Prabhu, Ramalingam, Might, and
+  Hall]{DBLP:conf/popl/PrabhuRMH11}
+T.~Prabhu, S.~Ramalingam, M.~Might, and M.~W. Hall.
+\newblock {EigenCFA}: accelerating flow analysis with {GPUs}.
+\newblock In \emph{POPL}, pages 511--522, 2011.
+
+\bibitem[Richards()]{PersonalCommunicationRichards}
+A.~Richards.
+\newblock Personal communication.
+\newblock CEO of Codeplay Software Ltd. 2 September 2016.
+
+\bibitem[{RISC OS}()]{risc-os-multitasking}
+{RISC OS}.
+\newblock Preemptive multitasking.
+\newblock \url{http://www.riscos.info/index.php/Preemptive_multitasking}.
+
+\bibitem[Smith and Anandtech(2016)]{anandtech}
+R.~Smith and Anandtech.
+\newblock Preemption improved: Fine-grained preemption for time-critical tasks,
+  2016.
+\newblock
+  \url{http://www.anandtech.com/show/10325/the-nvidia-geforce-gtx-1080-and-1070-founders-edition-review/10}.
+
+\bibitem[Solomon et~al.(2010)Solomon, Thulasiraman, and
+  Thulasiram]{DBLP:conf/hpcc/SolomonTT10a}
+S.~Solomon, P.~Thulasiraman, and R.~K. Thulasiram.
+\newblock Exploiting parallelism in iterative irregular maxflow computations on
+  {GPU} accelerators.
+\newblock In \emph{HPCC}, pages 297--304, 2010.
+
+\bibitem[Sorensen et~al.(2016)Sorensen, Donaldson, Batty, Gopalakrishnan, and
+  Rakamaric]{DBLP:conf/oopsla/SorensenDBGR16}
+T.~Sorensen, A.~F. Donaldson, M.~Batty, G.~Gopalakrishnan, and Z.~Rakamaric.
+\newblock Portable inter-workgroup barrier synchronisation for {GPUs}.
+\newblock In \emph{OOPSLA}, pages 39--58, 2016.
+
+\bibitem[Tanasic et~al.(2014)Tanasic, Gelado, Cabezas, Ram{\'{\i}}rez, Navarro,
+  and Valero]{DBLP:conf/isca/TanasicGCRNV14}
+I.~Tanasic, I.~Gelado, J.~Cabezas, A.~Ram{\'{\i}}rez, N.~Navarro, and
+  M.~Valero.
+\newblock Enabling preemptive multiprogramming on {GPUs}.
+\newblock In \emph{ISCA}, pages 193--204, 2014.
+
+\bibitem[Tarpenning(1991)]{Tarpenning:1991:CMC:136810.136820}
+M.~Tarpenning.
+\newblock Cooperative multitasking in c++.
+\newblock \emph{Dr. Dobb's J.}, 16\penalty0 (4), Apr. 1991.
+
+\bibitem[Tzeng et~al.(2010)Tzeng, Patney, and Owens]{TPO10}
+S.~Tzeng, A.~Patney, and J.~D. Owens.
+\newblock Task management for irregular-parallel workloads on the {GPU}.
+\newblock In \emph{HPG}, pages 29--37, 2010.
+
+\bibitem[Vallina{-}Rodriguez and
+  Crowcroft(2013)]{DBLP:journals/comsur/Vallina-RodriguezC13}
+N.~Vallina{-}Rodriguez and J.~Crowcroft.
+\newblock Energy management techniques in modern mobile handsets.
+\newblock \emph{{IEEE} Communications Surveys and Tutorials}, 15\penalty0
+  (1):\penalty0 179--198, 2013.
+
+\bibitem[Vineet et~al.(2009)Vineet, Harish, Patidar, and
+  Narayanan]{DBLP:conf/egh/VineetHPN09}
+V.~Vineet, P.~Harish, S.~Patidar, and P.~J. Narayanan.
+\newblock Fast minimum spanning tree for large graphs on the {GPU}.
+\newblock In \emph{HPG}, pages 167--171, 2009.
+
+\bibitem[Wang et~al.(2016)Wang, Davidson, Pan, Wu, Riffel, and
+  Owens]{DBLP:conf/ppopp/WangDPWRO16}
+Y.~Wang, A.~A. Davidson, Y.~Pan, Y.~Wu, A.~Riffel, and J.~D. Owens.
+\newblock Gunrock: a high-performance graph processing library on the {GPU}.
+\newblock In \emph{PPoPP}, pages 11:1--11:12, 2016.
+
+\bibitem[Xiao and Feng(2010)]{XF10}
+S.~Xiao and W.~Feng.
+\newblock Inter-block {GPU} communication via fast barrier synchronization.
+\newblock In \emph{IPDPS}, pages 1--12, 2010.
+
+\end{thebibliography}
 

### 2017-07-06 write-ups/FSE/paper.tex
Made paper cite extended version.

@@
 
-We present our cooperative kernels programming model as an extension
-to OpenCL.
-%However, the cooperative kernels concept is more general,
-%and could be applied to extend other GPU programming models, such as
-%CUDA and HSA.
-We describe the semantics of the model
+We describe the semantics of our cooperative kernels programming model
 (\mysec\ref{sec:semantics}), use our motivating examples to discuss
 programmability (\mysec\ref{sec:programmingguidelines}) and outline
-important nonfunctional properties that the model requires to work
-successfully (\mysec\ref{sec:nonfunctional}).
+nonfunctional properties required by the model (\mysec\ref{sec:nonfunctional}).
+In the extended version of this paper we present a formal semantics, and discuss alternative semantic choices~\cite{ExtendedVersion}.
 %and explain that the
@@
 compute units. Results for the other Intel GPUs are available in
-supplementary material, found on the author's website.}
+the extended version of this paper~\cite{ExtendedVersion}.}
 
@@
 graphs show a representative sample of our results; the full set of
-graphs for all configurations is provided in supplementary material.
+graphs for all configurations is provided in the extended version of this paper~\cite{ExtendedVersion}.
 

### 2017-07-07 write-ups/FSE/paper.tex
fixed fonts but went over

@@
 \usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
+%\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
 \usepackage{url}                  % format URLs
@@
 
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{microtype}
 \usepackage[normalem]{ulem}
@@
 \usepackage[noend]{algpseudocode}
-\usepackage[T1]{fontenc}
+
 \usepackage{listings}
@@
 %
-Irregular algorithms usually require \emph{blocking synchronization}
+Irregular algorithms usually require \emph{blocking synchronisation}
 between workgroups, e.g.\ many graph algorithms use a level-by-level
@@
 kernel. Each thread receives a pointer to the task queues, in global
-memory, initialized by the host to contain initial tasks. A thread
+memory, initialised by the host to contain initial tasks. A thread
 uses its workgroup id (line~\ref{line:wksteal:getgroupid}) as a queue
